### 2024-02-29 index.qmd
update 2

---
title: "About Me"
about:
  id: me
  template: broadside
  image: img/profile.png
---

:::{#me}

## Goals

## Education

## Experience

:::

Thanks for checking out my web site!

### 2024-02-29 aboutme.qmd
update website

---
title: "Jackson Strand"
about:
  id: me
  template: trestles
  links:
    - icon: linkedin
      text: LinkedIn
      href: https://linkedin.com/in/jacksonstrand
    - icon: instagram
      text: Instagram
      href: https://www.instagram.com/jstrand894/
    - icon: github
      text: GitHub
      href: https://github.com/jstrand894 
  image: img/aboutme/deadhorse.jpeg

sidebar: false
---

:::{#me}

I am currently a Master's graduate student at Montana State University studying entomology. I live in Bozeman, MT with my wife Grace.

[My CV](Aboutme/jackson_strand_updated_cv.pdf)

![Me and Grace at Glacier National Park, 2023](img/aboutme/iceberg.JPG)

:::

### 2024-02-29 bugs.qmd
update website

---
title: "bug pictures"
filters:
  - lightbox
lightbox: auto
---

![Bug pic 1](img/bugs/IMG_1664.JPG){group="my-gallery"}
![Bug pic 2](img/bugs/IMG_1678.JPG){group="my-gallery"}

<figure>
  <img src="img/bugs/IMG_1664.JPG" alt="Montana State University" 
  width = 200" height = "200">
</figure>

### 2024-02-29 deadhorse.qmd
update website

---
title: "Deadhorse Ultra 50 Miler"
subtitle: "November 18, 2023"
format: html
filters:
   - lightbox
lightbox: auto
editor_options: 
  chunk_output_type: console
---

::: {style="display: grid;grid-template-columns: repeat(auto-fill, minmax(200px, 1fr));grid-gap: 1em;"}
```{r get images, echo = F, results = 'asis'}
image_files <- list.files("img/deadhorse", pattern = ".JPG|.png|.gif|.jpg|.jpeg", full.names = TRUE)

for (file in image_files) {
  cat("![](",file, ")\n  \n")
}
```
:::

### 2024-02-29 field_report.qmd
update website

---
title: "MT Field Site Analysis Report"
author: "Jackson Strand"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: yes
    include:
    keep_md: true
    theme: cosmo
    toc: true
    toc_float: true
    toc_title: "Table of Contents"
    toc_collapsed: false
    toc_depth: 4
    # number_sections: true
editor_options: 
  chunk_output_type: console
---
```{r library, echo = F, message = F, warning = F}
library(tidyverse)
library(readxl)
library(ggplot2)
library(kableExtra)
library(plotly)
```
```{r column.order, echo = F}
column.order <- c("distance", "collection", "p.inf", 
                  "p.cut", "p.par", "poten")

column.order2 <- c("location", "distance", "collection", "p.inf", 
                  "p.cut", "p.par", "poten")

round_based_on_size <- function(x) {
  threshold <- 0.01
  rounded_values <- ifelse(abs(x) > threshold, round(x, 2), round(x, 3))
  return(rounded_values)
}
```

# POST FARM

## Methods

### Plant Selection and Cages

Assessment of C. cinctus infestation and mortality within B. inermis were assessed using a 34 x 60 ft plot at the Arthur H. Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT. The cage structure was built using 1-inch PVC piping with the netting made using 530\(\mu\) Amber Lumite Screen (BioQuip® Products, LLC). Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three. The cages were removed from the plot after 3 weeks of sawfly release.

### Insects

Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of C. cinctus infestation and cutting in the prior year. Cut stubble, which contained C. cinctus larvae in diapause, were kept refrigerated between -2°C and 3°C for >100 days as required to complete obligatory larval diapause. As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA). Stubs were added to cages once B. inermis stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).

#### Data Collection

Stems were collected from each cage site in late August. All stems from each 6 x 3 ft cage section were collected. Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.

## Analysis

### Overall WSS Infestation and Cutting

```{r (pf.RAW) post farm n.inf data, echo = F, message = F}
pf.raw.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2022/Post\ Farm\ 2022.xlsx", sheet = "n.inf")
pf.raw.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "n.inf")
```
```{r (pf.inf.23) post farm inf data, echo = F}
pf.inf.23 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf") %>%
  rename(n.1 = 8, n.2 = 9, n.3 = 10, n.4 = 11,
         n.5 = 12, n.6 = 13, n.7 = 14, n.8 = 15,
         n.9 = 16, n.10 = 17)
```
```{r (pf.comb) post farm data combined, echo = F}
# 2022
pf.2022.site <-
pf.raw.2022 %>%
  group_by(treatment, site) %>%
  reframe(site, treatment,
            p.inf = sum(iv, ir, ix)/sum(uv, ur, iv, ir, ix),
            cut = sum(cut)/sum(total),
            total.stems = sum(iv, ir, ix, uv, ur)) %>%
  unique() %>%
  pivot_longer(-c(treatment, total.stems, site),
               names_to = "type",
               values_to = "percent") %>%
  mutate(year = "2022")

# 2023
pf.2023.site <-
pf.raw.2023 %>%
  group_by(treatment, site) %>%
  reframe(treatment, site, 
          t.inf = sum(n.iv, n.ir, n.ix),
          p.inf = sum(n.iv, n.ir, n.ix)/sum(n.uv, n.ur, 
                                            n.iv, n.ir, n.ix),
          cut = sum(cut)/sum(total),
          total.stems = sum(n.iv, n.ir, n.ix, n.uv, n.ur)) %>%
  distinct() %>%
  pivot_longer(-c(treatment, total.stems, site),
               names_to = "type",
               values_to = "percent") %>%
  filter(!type == "t.inf") %>%
  arrange(type) %>%
  mutate(year = "2023")

pf.comb <-
  pf.2022.site %>%
  rbind(pf.2023.site) %>%
  group_by(treatment, year, type) %>%
  mutate(count = n(),
         average = mean(percent),
         sd = sd(percent),
         se = (sd/sqrt(count))) %>%
  reframe(treatment, year, type,
          mean = mean(percent),
          sd = mean(sd),
          se = mean(se)) %>%
  unique()

```
```{r (pf.total) pf total stems, echo = F}
# 2022 totals
pf.2022.total <-
  pf.raw.2022 %>%
  group_by(treatment) %>%
  reframe(total = sum(uv, ur, iv, ir, ix)) %>%
  mutate(year = "2022")

# 2023 totals
pf.2023.total <-
  pf.raw.2023 %>%
  group_by(treatment) %>%
  reframe(total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
  mutate(year = "2023")

pf.total.comb <-
  pf.2022.total %>%
  rbind(pf.2023.total)

```

#### Basic Table and Plot of Infestation

```{r post farm table, echo = F, results = 'asis'}
pf.23.table <-
pf.comb %>%
  filter(year == "2023") %>%
  dplyr::select(treatment, type, mean) %>%
  mutate(mean = round_based_on_size(mean)) %>%
  pivot_wider(values_from = mean,
              names_from = type,
              values_fn = list(mean = mean)) %>%
  left_join(pf.2023.total, by = "treatment")

pf.22.table <-
pf.comb %>%
  filter(year == "2022") %>%
  dplyr::select(treatment, type, mean) %>%
  mutate(mean = round_based_on_size(mean)) %>%
  pivot_wider(values_from = mean,
              names_from = type,
              values_fn = list(mean = mean))  %>%
  left_join(pf.2022.total, by = "treatment")

pf.23.table.final <-
kable(pf.23.table[c("treatment", "total", "p.inf", "cut")], 
      caption = "2023 Post Farm Infestation", 
      format = "html",
      col.names = c("Treatment", "Total Stems", "% Inf", "% Cut"),
      align = "lccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2")

pf.22.table.final <-
kable(pf.22.table[c("treatment", "total", "p.inf", "cut")], 
      caption = "2022 Post Farm Infestation", 
      format = "html",
      col.names = c("Treatment", "Total Stems", "% Inf", "% Cut"),
      align = "lcc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2")

cat('<div style="display: flex;">')
cat('<div style="flex: 1;">')
cat(pf.23.table.final)
cat('</div>')
cat('<div style="flex: 1;">')
cat(pf.22.table.final)
cat('</div>')
cat('</div>')
```

```{r post farm plot, echo = F, fig.height = 2.5, fig.align = 'center'}
pf.comb <-
pf.comb %>%
  left_join(pf.total.comb, by = c("treatment", "year"))

ggplot() +
  facet_wrap(~treatment) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           data = pf.comb %>%
             filter(type == "p.inf"),
           aes(year, mean, fill = "p.inf")) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           data = pf.comb %>%
             filter(type == "cut"),
           aes(year, mean, fill = "cut")) +
  geom_errorbar(data = pf.comb,
                aes(x = year,
                    ymin = mean - se,
                    ymax = mean + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  theme_bw() +
  labs(x = element_blank(),
       y = "Percentage of stems (%)",
       fill = element_blank()) +
  scale_fill_brewer(palette = "Greens",
                    labels = c("Cut", "Infested"))

```

Let's take a closer look at the number of nodes borrowed through by the larvae. Using this information we can make an arguement about the size and length of time each larvae was alive prior to dying. This can be informative in understanding how many larvae would have been left later in the season for potential parasitism.

#### Infestation Percentage/Node

```{r v.r infestation node, echo = F, warning = F, message = F, fig.height = 3.5, fig.align = 'center'}
v.r_inf <- 
pf.inf.23 %>%
  dplyr::select(site, rep, treatment, 8:19) %>%
  mutate(stem.id = row_number()) %>%
  pivot_longer(-c(stem.id, site, rep, treatment, v.r, lar),
               names_to = "node",
               values_to = "node_status") %>%
  na.omit() %>%
  group_by(stem.id, treatment, v.r, lar) %>%
  reframe(nodes.inf = n()) %>%
  group_by(treatment, v.r, lar, nodes.inf) %>%
  reframe(count = n()) %>%
  group_by(treatment, v.r) %>%
  mutate(count2 = sum(count),
         per = count/count2) %>%
  reframe(treatment, v.r, nodes.inf, per) %>%
  mutate_at(vars(nodes.inf), as.factor)

# plot
v.r_inf %>%
  ggplot() +
  geom_bar(stat = "identity",
         position = position_dodge(),
         aes(y = per, x = nodes.inf, fill = treatment)) +
  geom_line(stat = "smooth",
            aes(y = per, 
                x = nodes.inf, 
                group = treatment, 
                color = treatment),
            se = FALSE,
            alpha = 0.6,
            linewidth = 1.3,
            linetype = "solid",
            method = "loess") +
  theme_bw() +
  scale_fill_brewer(palette = "Greens") +
  scale_color_brewer(palette = "Reds") +
  facet_wrap(~v.r)

# anova
anova1 <- 
  summary(aov(per ~ treatment + nodes.inf + v.r, data = v.r_inf))

aov.df <- as.data.frame(anova1[[1]])
aov.df$`Pr(>F)` <- sprintf("%.4f", aov.df$`Pr(>F)`)
aov.df$Source <- rownames(aov.df)
rownames(aov.df) <- NULL

aov.df[,2:4] <- round(aov.df[,2:4], 3)

# anova table
kbl(aov.df %>%
      dplyr::select(Source, everything()), 
    align = "lccccc",
    format = "html",
    caption = "ANOVA Table",
    col.names = c("", "Df", "Sum Sq", "Mean Sq", "F-Value", "Pr(>F)")) %>%
  kable_styling(bootstrap_options = c("striped", "bordered", 
                                      "hover", "repsonsive"),
                full_width = FALSE) %>%
  row_spec(0, background = "#f2f2f2")

```

Based on the results of our ANOVA, we have weak evidence (Pval = `r aov.df[3,5]`) to suggest there is a difference in the number of nodes infested when considering treatment group and vegetative and reproductive stems.

```{r cumsum of infestation per node, echo = F, fig.align = "center"}
# I want to see the percentage of stems with 6+ nodes infested, then 5:6+, then 4:6+, etc.
cumsum <-
pf.inf.23 %>%
  dplyr::select(site, rep, treatment, 8:19) %>%
  mutate(stem.id = row_number()) %>%
  pivot_longer(-c(stem.id, site, rep, treatment, v.r, lar),
               names_to = "node",
               values_to = "node_status") %>%
  na.omit() %>%
  group_by(stem.id, treatment, v.r, lar) %>%
  reframe(nodes.inf = n()) %>%
  group_by(treatment, nodes.inf) %>%
  reframe(count = n()) %>%
  group_by(treatment) %>%
  mutate(count2 = sum(count),
         per = count/count2) %>%
  reframe(treatment, nodes.inf, per) %>%
  arrange(treatment, desc(nodes.inf)) %>%
  group_by(treatment) %>%
  mutate(cumsum = cumsum(per)) %>%
  mutate_at(vars(nodes.inf), as.factor) %>%
  dplyr::select(treatment, nodes.inf, cumsum) %>%
  mutate(nodes.inf = fct_rev(nodes.inf))

cumsum %>%
  mutate(cumsum = round(cumsum, 2)) %>%
  rename("Treatment" = treatment) %>%
  pivot_wider(values_from = "cumsum",
              names_from = "nodes.inf") %>%
  kable("html",
        caption = "Cumulative sum of nodes infested",
        align = "lcccccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2") %>%
  add_header_above(c(" ", "# of Nodes Infested" = 6))

ggplot(data = cumsum,
         aes(x = nodes.inf,
             y = cumsum,
             group = treatment, 
             color = treatment)) +
  geom_line() +
  theme_bw() +
  geom_bar(stat = "identity",
           aes(x = nodes.inf, 
               y = cumsum,
               fill = treatment),
           position = position_dodge(),
           alpha = 0.25) +
  theme_bw() +
  labs(x = "Number of Nodes Infested",
       y = "Cumulative sum % of stems",
       fill = "WSS Treatment \nLevel",
       color = "WSS Treatment \nLevel",
       title = "Smooth brome node infestation") +
  scale_x_discrete(limits = rev(levels(cumsum$cumsum)))

```

I like this plot because it shows the cumulative percentage sum of nodes infested at each node. For example, at the 3rd node in 2023, we see that close to 70% of the stems in the 'high' treatment group exhibited between 3 and 6 nodes infested. As a general thought exercise, the more nodes a larvae burrowed through, the larger the larvae. This means that if we classify a "large" larvae as one who burrowed through at least 4 nodes, our infestation data shows that 36% of the larvae in the 'high' treatment groups died as "large" larvae.

```{r larvae breakdown, echo = F}
# how many stems had live larvae in them?
# percent of stems with each larval type

larvae.type <-
pf.inf.23 %>%
  dplyr::select(site, rep, treatment, 8:19) %>%
  mutate(stem.id = row_number()) %>%
  pivot_longer(-c(stem.id, site, rep, treatment, v.r, lar),
               names_to = "node",
               values_to = "node_status") %>%
  na.omit() %>%
  group_by(treatment, lar) %>%
  reframe(count = n()) %>%
  mutate(year = "2023") %>%
  left_join(pf.total.comb, by = c("treatment", "year")) %>%
  group_by(treatment) %>%
  mutate(per = count/total) %>%
  filter(lar == "a") %>%
  mutate(per = round(per, 3))
# stems collected in late October. This is much later than if parasitoids were flying.

kable(larvae.type[c("treatment", "count", "total", "per")],
      format = "html",
      align = "c",
      col.names = c("Treatment", "Count", "Total Stems", "%"),
      caption = "Stems with Living Larvae (2023)") %>%
  kable_styling(bootstrap_options = c("striped", "bordered", 
                                      "hover", "repsonsive"),
                full_width = FALSE) %>%
  row_spec(0, background = "#f2f2f2")
```

Interesting to see that `r larvae.type[1,6]*100`% of stems in the 'high' treatment groups and `r larvae.type[2,6]*100`% of stems in the 'low' treatment group had living WSS larvae. The stems for this experiement were harvested in late October, much later in the season when we would expect the second generation parasitoids to be searching for a host.

<hr>

# MONTANA FIELD SITES

## Methods

To assess C cinctus infestation, larval mortality, and parasitoid prevalence within B. inermis and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production. Samples were collected in both early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA. Sample locations were chosen by identifying fields with consistent wheat production that were adjacent to healthy B. inermis stands, either situated between fields or between fields and roadways. Polygons containing B. inermis were set up along the periphery of adjacent wheat fields, each covering an area of approximately 100 meters. Four collection squares of 1ft x 1ft were randomly selected within each polygon during both collection events each year. All stems within each 1 x 1 ft square were collected. The stems were then processed and split in the laboratory to assess infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting.

## Analysis

```{r (raw) brome data, echo = F}
raw.2022 <- 
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "n.infested") %>%
  mutate(year = "2022") %>%
  rename("site" = site.cap)

raw.2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "n.infested") %>%
  mutate(year = "2023") 
```
```{r (stubble.####) stubble data, echo = F}
stubble.2023 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2023/2023\ Field Site\ Data.xlsx", 
           sheet = "stubble") %>%
  dplyr::select(-name) %>%
  na.omit() %>%
  mutate(year = "2023")

stubble.2022 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2022/2022 FS Stem Data (2).xlsx", 
             sheet = "stubble") %>%
  mutate(year = "2022")

```
```{r (joined.stubble) stubble join, echo = F}
# the below code joins the 2022 and 2023 stubble data frames
joined.stubble <-
stubble.2023 %>%
  dplyr::select(site, location, distance,
                inf.nc, cut, cocoon, eh, total, year) %>%
  mutate(collection = "fall") %>% 
  rbind(stubble.2022 %>%
          dplyr::select(site, year, location, distance, collection,
                        inf.nc, cut, cocoon, eh, total)) %>%
  # remove N sites and re add as NS and NN for example
  group_by(site, location, year, distance, collection) %>%
  mutate(inf = sum(cut, inf.nc, cocoon, eh)) %>%
  # first, we remove the site arguement by grouping by everything else
  group_by(site, location, year, distance, collection) %>%
  dplyr::select(-c(inf.nc)) %>%
  rename("e" = eh,
         "c" = cocoon) %>%
  mutate(p.inf = inf/total,
         p.par = (c + e)/inf,
         p.cut = cut/inf,
         poten = (cut + e + c)/inf)

# duplicate the wheat data to match up with brome data locations
# we are duplicating some of the stubble data so that we have stubble data
# for the brome sections that are across the road from stubble data
joined.stubble <-
  joined.stubble %>%
  bind_rows(joined.stubble %>%
              filter(location == "n") %>%
              mutate(location = "ns")) %>%
  bind_rows(joined.stubble %>%
              filter(location == "n") %>%
              mutate(location = "nn")) %>%
  filter(!location == "n") %>%
  bind_rows(joined.stubble %>%
              filter(location == "har") %>%
              mutate(location = "sun"))

```
```{r (n.inf) brome join, echo = F}
n.inf <-
raw.2023 %>%
  dplyr::select(site, year, location, collection, 
              n.uv, n.ur, n.iv, n.ir, n.ix, 
              cut, e, c)  %>%
  bind_rows(raw.2022 %>%
              rename(n.uv = uv,
                     n.ur = ur,
                     n.iv = iv,
                     n.ir = ir,
                     n.ix = ix) %>%
              dplyr::select(site, year, location, collection, 
                    n.uv, n.ur, n.iv, n.ir, n.ix, 
                    cut, e, c)) %>%
  group_by(site, location, collection, year) %>%
  mutate(inf = sum(n.iv, n.ir, n.ix),
         total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
  ungroup() %>%
  dplyr::select(-c(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
  mutate(p.inf = inf/total,
         p.par = (c + e)/inf,
         p.cut = cut/inf,
         poten = (cut + c + e)/inf) %>%
  pivot_longer(-c(site, location, collection, year),
               names_to = "type",
               values_to = "value") %>%
  group_by(location, collection, year, type) %>%
  distinct() %>%
  mutate(distance = "brome",
         location = case_when(
           location == "kw" ~ "k",
           location == "ke" ~ "k", 
           TRUE ~ location))
```
```{r (stubble.brome) join stubble/brome, echo = F}
# join the stubble data and the brome data
stubble.brome <-
joined.stubble %>%
  pivot_longer(-c(site, location, distance, year, collection),
               names_to = "type",
               values_to = "value") %>%
  distinct() %>%
  rbind(n.inf) %>%  # rbind the brome data
  filter(!type == "total") %>%  # remove type = "total"
  mutate(type = case_when(
    type == "c" ~ "p",
    type == "e" ~ "p",
    TRUE ~ type)) %>% # replace the e and c variables with p, for simplicity
  ungroup() %>%
  dplyr::select(-site) %>%
  group_by(location, distance, year, collection, type) %>%
  reframe(mean = mean(value),
         count = n(),
         sd = sd(value),
         se = sd/sqrt(count)) %>%
  distinct() %>%
  filter(type %in% c("p.inf", "p.cut", "p.par", "poten")) %>%
  distinct()
```

```{r, echo = F, eval = F}
stubble.brome %>%
  filter(location == "g",
         distance %in% c("20", "5"),
         year == "2023",
         collection == "fall")
```

### Big Sandy, MT 2023

#### Lochy Property - North Road

![North Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/nnns-min.png?raw=true)

**Note:**    
- In 2023, the field south of North Road was fallow, so no stubble infestation data was collected from the field south of North Road.

```{r NN & NS data table, echo = F}
nn.table <-
stubble.brome %>%
  filter(location %in% c("nn"),
         year == "2023") %>%
  dplyr::select(location, distance, collection, year, type, mean) %>%
  mutate(mean = round_based_on_size(mean)) %>%
  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
  pivot_wider(names_from = type,
              values_from = mean,
              values_fn = list(mean = mean)) %>%
  arrange(distance) %>%
  ungroup() %>%
  dplyr::select(-year)

kable(nn.table[, column.order2], 
      caption = "North Road (site nn) Infestation Table", 
      format = "html",
      col.names = c("Location", "Distance", "Collection", 
                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
      align = "lllcccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2")

ns.table <-
stubble.brome %>%
  filter(location %in% c("ns"),
         year == "2023",
         distance == "brome") %>%
  dplyr::select(location, distance, collection, year, type, mean) %>%
  mutate(mean = round_based_on_size(mean)) %>%
  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
  pivot_wider(names_from = type,
              values_from = mean,
              values_fn = list(mean = mean)) %>%
  arrange(distance) %>%
  ungroup() %>%
  dplyr::select(-year)

kable(ns.table[, column.order2], 
      caption = "North Road (site ns) Infestation Table", 
      format = "html",
      col.names = c("Location", "Distance", "Collection", 
                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
      align = "lllcccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2") 
```

```{r NN and NS plot, echo = F, fig.height = 3.5, fig.width = 10, warning = F}
stubble.brome %>%
  filter(location %in% c("nn", "ns"),
         year == "2023",
         collection == "fall") %>%
  ggplot(aes(x = type, y = mean, fill = distance)) +
  geom_bar(stat = "identity",
           position = position_dodge()) +
  geom_errorbar(aes(ymin = mean - se,
                    ymax = mean + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  theme_bw() +
  theme(legend.title = element_blank()) +
  scale_fill_brewer(palette = "Purples",
                    labels = c("Adjacet wheat: 20 m", "Adjacet wheat: 5 m", "Smooth brome")) +
  labs(y = "Percent (%)",
       x = element_blank(),
       title = "North Road Infestation") +
  scale_x_discrete(labels = c("p.cut" = "Cut",
                              "p.inf" = "Infested",
                              "p.par" = "Parasitized",
                              "poten" = "Potential Cut")) +
  facet_wrap(~location)
```

We saw very low levels of infestation in 2023 within the wheat. This was likely due a cold, mild spring that led to late stem development. Because the wheat was not well developed, the sawflies did not infest. This appears to have led to high infestation within the smooth brome. Will be interesting to compare to 2022 when stem elongation followed normal expected timelines.

#### Lochy Property - Gardiner Road

![Gardiner Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/gardiner%20road-min.png?raw=true)

In 2022, Gardiner road underwent a major reconstruction with both ditches being regraded. This severely diminished the amount smooth brome found along the edge of the roadway.

```{r G table, echo = F}
g.table <-
stubble.brome %>%
  filter(location %in% c("g"),
         year == "2023") %>%
  dplyr::select(location, distance, collection, year, type, mean) %>%
  mutate(mean = round_based_on_size(mean)) %>%
  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
  pivot_wider(names_from = type,
              values_from = mean,
              values_fn = list(mean = mean)) %>%
  arrange(distance) %>%
  ungroup() %>%
  dplyr::select(-c(year, location))

kable(g.table[, column.order], caption = "Gardiner Road Infestation Table", 
      format = "html",
      col.names = c("Distance", "Collection", 
                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
      align = "lllcccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2")
```

```{r G plot, echo = F, fig.height = 3.5, fig.align = 'center', warning = F}
stubble.brome %>%
  filter(location %in% c("g"),
         year == "2023",
         collection == "fall") %>%
  ggplot(aes(x = type, y = mean, fill = distance)) +
  geom_bar(stat = "identity",
           position = position_dodge()) +
  geom_errorbar(aes(ymin = mean - se,
                    ymax = mean + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  theme_bw() +
  theme(legend.title = element_blank()) +
  scale_fill_brewer(palette = "Purples",
                    labels = c("Adjacet wheat: 20 m", "Adjacet wheat: 5 m", "Smooth brome")) +
  labs(y = "Percent (%)",
       x = element_blank(),
       title = "Gardiner Rd Infestation") +
  scale_x_discrete(labels = c("p.cut" = "Cut",
                              "p.inf" = "Infested",
                              "p.par" = "Parasitized",
                              "poten" = "Potential Cut"))
```

Similar results to North Road seen here. Low infestation in the adjacent wheat field was due to the cool spring and lack of infestable stems during the sawfly flight. This shows that even in odd years like 2023, sawflies are still present in the smooth brome, meaning parasitoids will still have larval hosts.

### Moccasin, MT 2023

#### Sunset & Harvest Road Intersection

![Harvest & Sunset Road Instersection, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/harvest%20road-min.png?raw=true)

```{r SUN/HAR data tables, echo = F}
sun.table <-
stubble.brome %>%
  filter(location %in% c("sun"),
         year == "2023") %>%
  dplyr::select(location, distance, collection, year, type, mean) %>%
  mutate(mean = round_based_on_size(mean)) %>%
  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
  pivot_wider(names_from = type,
              values_from = mean,
              values_fn = list(mean = mean)) %>%
  arrange(distance) %>%
  ungroup() %>%
  dplyr::select(-year)

kable(sun.table[, column.order2], caption = "Sunset Road Infestation Table", 
      format = "html",
      col.names = c("Location", "Distance", "Collection", 
                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
      align = "lllcccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2")

har.table <-
stubble.brome %>%
  filter(location %in% c("har"),
         year == "2023") %>%
  dplyr::select(location, distance, collection, year, type, mean) %>%
  mutate(mean = round_based_on_size(mean)) %>%
  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
  pivot_wider(names_from = type,
              values_from = mean,
              values_fn = list(mean = mean)) %>%
  arrange(distance) %>%
  ungroup() %>%
  dplyr::select(-year)

kable(har.table[, column.order2], caption = "Harvest Road Infestation Table", 
      format = "html",
      col.names = c("Location", "Distance", "Collection", 
                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
      align = "lllcccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2")
```

```{r SUN/HAR plot, echo = F, warning = F, fig.align = 'center', fig.height = 3.5, fig.width = 10}
stubble.brome %>%
  filter(location %in% c("sun", "har"),
         year == "2023",
         collection == "fall") %>%
  ggplot(aes(x = type, y = mean, fill = distance)) +
  geom_bar(stat = "identity",
           position = position_dodge()) +
  geom_errorbar(aes(ymin = mean - se,
                    ymax = mean + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  theme_bw() +
  theme(legend.title = element_blank()) +
  guides(color = guide_legend(title.position = "top", 
                              label.padding = unit(10, "lines"))) +
  scale_fill_brewer(palette = "Purples",
                    labels = c("Adjacet wheat: 20 m", "Adjacet wheat: 5 m", "Smooth brome")) +
  labs(y = "Percent (%)",
       x = element_blank(),
       title = "Sunset & Harvest Rd Infestation") +
  scale_x_discrete(labels = c("p.cut" = "Cut",
                              "p.inf" = "Infested",
                              "p.par" = "Parasitized",
                              "poten" = "Potential Cut")) +
  facet_wrap(~location)
```

#### Kolin Road

![Kolin Road, Moccasin, MT (2023)](https://github.com/jstrand894/Website/blob/main/kolin%20road-min.png?raw=true)

```{r K data table, echo = F}
k.table <-
stubble.brome %>%
  filter(location %in% c("k"),
         year == "2023") %>%
  dplyr::select(location, distance, collection, year, type, mean) %>%
  mutate(mean = round_based_on_size(mean)) %>%
  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
  pivot_wider(names_from = type,
              values_from = mean,
              values_fn = list(mean = mean)) %>%
  arrange(distance) %>%
  ungroup() %>%
  dplyr::select(-year)

kable(k.table[, column.order], caption = "Kolin Road Infestation Table", 
      format = "html",
      col.names = c("Distance", "Collection", 
                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
      align = "lllcccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2")
```

```{r K plot, echo = F, warning = F, fig.height = 3, fig.align = 'center'}
stubble.brome %>%
  filter(location %in% c("k"),
         year == "2023",
         collection == "fall") %>%
  ggplot(aes(x = type, y = mean, fill = distance)) +
  geom_bar(stat = "identity",
           position = position_dodge()) +
  geom_errorbar(aes(ymin = mean - se,
                    ymax = mean + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  theme_bw() +
  theme(legend.title = element_blank()) +
  scale_fill_brewer(palette = "Purples",
                    labels = c("Adjacet wheat: 20 m", "Adjacet wheat: 5 m", "Smooth brome")) +
  labs(y = "Percent (%)",
       x = element_blank(),
       title = "Kolin Rd Infestation") +
  scale_x_discrete(labels = c("p.cut" = "Cut",
                              "p.inf" = "Infested",
                              "p.par" = "Parasitized",
                              "poten" = "Potential Cut")) 
```

### 2024-02-29 locoweed-report.qmd
update website

---
title: "All Locoweed Analysis Report"
author: "Jackson Strand"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
bibliography: citations/citations.bib
link-citations: true
cls: council-of-science-editors-author-date.csl
editor_options: 
  chunk_output_type: console
---

```{r libraries, echo = F, message = F, warning = F}
library(tidyverse)
library(readxl)
library(MASS)
library(knitr)
library(kableExtra)
library(compositions)
# PERMANOVA
library(vegan)
library(spaa)
library(devtools)
devtools::install_github("GuillemSalazar/EcolUtils")
library(EcolUtils)

# Random Forest
library(ranger)
library(MVN) # for normality test

```

# INTRODUCTION
Astragalus and Oxytropix species, part of the Legume Family, are found in terestrial ecoysstems globally, including the western regions of North America. Many of these species are both nontoxic and important forages. However, several species are known to be toxic to both livestock and wildlife. *Oxytropis sericea Nutt* can harbor a commensal endophyte, *Alternaira oxytropos*, a fungus that is assocaited with locoweed plants, specifically those of the genus *Oxytropis*. Plants containing the endophyte have been found to produce swansonine, a toxic alkaloid that can be harmful to livestock when ingested in large quantities [@locoweed]. In this study, we investigated the potential differences in volatile organic compound (VOC) emissions between specimens that contained the commensal endophyte *Alternaria oxytropis*, and those that did not.

These data consist of volatile collection events over the course of two years, 2022 and 2023. In both years, data was collected from wild plants in the field, and plants grown in the greenhouse.

# METHODOLOGY
## Data Collection
### Field Volatile Collection

Volatile collected was undergone on plants grown at a 34 x 60 ft plot at the Arthur H. Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT. Plants were inventoried and logged using GPS, and tissue samples from each specimen were analyzed to establish whether the plant contained the endophyte. Volatiles were collected from all the plants in each year over the course of one to two successive weeks.

Volatile were collected from 20 plants per day by placing Reynolds Turkey Oven Bags (Reynolds Group Holdings, Auckland, New Zealand) over the top of the selected number of leaflets.

### Greenhouse Volatile Collection
Samples of each plant were first sent for analysis to mystery lab to see if endophyte was present or not. 
Volatile compounds were collected from healthy *(size/stage)* locoweed plants. Plants were paired based on size similarity, with one being E+ and the other E-. Three sets of pairs (twelve plants) were used for each volatile collection event. Collection was undergone for a period of six hours from 09:00 to 15:00.

Volatiles were collected from twleve plants per day using glass volatile collection chambers (VCCs) that were 30 cm long and 4 cm in diameter. The top end of each VCC funneled into a threaded port that was 10 cm in daimter. Threaded onto the top of each VCC was an eight-port "octopus" manifold adapter. Two glass filter traps (6.35 mm diameter x 76 mm length; Analytical Research Systems, Gainesville, FL) with 30 mg of Super-Q absorbent (Alltech Associates, Deerfield, IL) were inserted into two of the adapter ports. The remaining six collection ports were sealed to prevent ambient air exchange and pressure loss. One of the traps was used to collected extraneous volatiles from the initial experiment set up for the first ten minutes, whereas the other trap was used to collect volatile compounds emanating from plants beginning ten minutes after the system was turned on. An air source was connected to the base of each collection chamber with supply from charcoal air filters. A vacuum pump was used to maintain air pressure and flow rate (humidified air at 1.0 liter/min) within the collection system. A small piece of cotton encompassed by a Teflon® guillotine was used at the base of each plant to tightly seal the base of the system and prevent external air from entering. Laboratory tape was used to seal the edges of the guillotine and prevent contamination by ambient and soil derived volatiles.

Volatile collection and successive GC-MS analysis were conducted following methods described in Weaver et al. (2009). After the six hour collection period, glass volatile collection traps were eluted using 200 µL of dichloromethane into 2 mL screw top vials (Agilent Technologies, Inc., Santa Clara, CA), followed by a slow release of nitrogen to further clear the volatile trap. 10 µL of nonyl acetate solution was added to the eluted solution as internal standard to quantify compounds. Samples were then subject to gas chromatography (GC) on a J&W HP – 5ms; 30m x 0.25 µm film thickness column (Agilent Technologies, Inc., Santa Clara, CA). The GC unit (Agilent 6890; Agilent Technologies, Santa Clara, CA) was paired with a mass spectrometer (MS, Agilent 5973 instrument). Samples were injected into the column in pulse splitless mode, with the initial pressure of 82.7 kPa/m.

## Data Analysis
The data analysis procedure for each collection type was the same. Statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PERMANOVA, PERMDISP, and Random Forest modeling. The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide]. VOC compositions were compared between samples and treatments using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan]. This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017]. Data were visualized using a combination of ggplot2 and knitr [@ggplot; @knitr]. I selected a subset of comounds to ues for each analysis. These compounds were seelcted based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output. I will highlight the cmopounds that I used for each analysis as I get to them.

# RESULTS
<!-- LOAD DATA -->
```{r 2023 field data load, echo = F, message = F}
field.2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Field/2023\ Field/2023\ Field\ Locoweed.xlsx", sheet = "proc") 
```
```{r 2022 field data load, echo = F, message = F}
field.2022 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Field/2022\ Field/2022\ Field\ Locoweed.xlsx", sheet = "RAW proc") %>%
  dplyr::filter(!plant.id == "581")
```
<!-- CLEAN UP DATA -->
```{r 2023 field data cleanup, echo = F, message = F}
field.2023.data <-
field.2023 %>%
  filter(compound %in% c("heptanal", "sixmethyl", "octanal",
                         "z3hex", "dlimonene", "nonanal", 
                         "decanal", "methylsalicylate", "cisthreehexiso")) %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  left_join(field.2023 %>% 
              dplyr::select(plant.id, swa, type, pair2), # go back and add pairs later
            by = "plant.id") %>%
  replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa, type), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2023")
```
```{r 2022 field data cleanup, echo = F, message = F}
field.2022.data <-
field.2022 %>%
  filter(compound %in% c("heptanal", "sixmethyl", "octanal",
                         "z3hex", "dlimonene", "nonanal", 
                         "decanal", "methylsalicylate", "cisthreehexiso")) %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  left_join(field.2022 %>% 
              dplyr::select(plant.id, swa, collection.type, pair2), # go back and add pairs later
            by = "plant.id") %>%
  # replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa, collection.type), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2022") %>%
  rename(type = collection.type) %>%
  filter(!is.na(type) & type != 0) %>%
  # replace leaves and flowers with l and s
  mutate(type = ifelse(type == "leaves", "l", "f"))

```

```{r COMPOUNDS, echo = F, message = F}
compounds <- c('sixmethyl',
               'z3hex',
               'nonanal',
               'dlimonene',
               'methylsalicylate',
               # 'decanal',
               'cisthreehexiso',
               'dodecanal',
               # 'heptadecanal',
               'octanal',
               'bocimene',
               'bmyrcene',
               'heptanal')

# "heptanal", "sixmethyl", "octanal",
#                          "z3hex", "dlimonene", "nonanal", 
#                          "decanal", "methylsalicylate",
#                          "cisthreehexiso"
```
```{R CLR, echo = F, message = F}
joined.data <-
field.2023 %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  filter(compound %in% compounds) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  dplyr::reframe(plant.id, compound, ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  mutate(year = "2023") %>%
  left_join(field.2023 %>%
              dplyr::select(plant.id, 
                            swa, type, pair2),
            by = "plant.id",
            relationship = "many-to-many") %>%
  mutate_at(vars(swa), list(factor)) %>%
  # rbind 2022 data
  rbind(field.2022 %>%
          dplyr::select(plant.id, compound, ngghr) %>%
          # filter(compound %in% c('sixmethyl','z3hex','nonanal',
          #                       'dlimonene','methylsalicylate',
          #                       'decanal','cisthreehexiso','dodecanal',
          #                       'heptadecanal','octanal','bocimene',
          #                       'bmyrcene','heptanal')) %>%
          group_by(plant.id, compound) %>%
          na.omit() %>%
          dplyr::reframe(plant.id, compound, 
                           ngghr = sum(ngghr)) %>%
          distinct() %>%
          pivot_wider(names_from = "compound",
                      values_from = "ngghr") %>%
          pivot_longer(-plant.id,
                       names_to = "compound",
                       values_to = "ngghr") %>%
          mutate(year = "2022") %>%
          left_join(field.2022 %>%
                      dplyr::select(plant.id,
                                    swa, collection.type, pair2) %>%
                      mutate_at(vars(swa), list(factor)),
                    by = "plant.id",
                    relationship = "many-to-many") %>%
          rename(type = collection.type)) %>%
  distinct()

# having trouble joining the two data sets because of the pair2 column 
joined.data.clr <-
joined.data %>%
  filter(compound %in% compounds) %>%
  dplyr::select(plant.id, ngghr, compound, year) %>%
  # replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = ifelse(ngghr == 0, 0.01, ngghr)) %>%
  mutate(ngghr = ngghr + 1) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
    # combine year and plant id into one column
  unite(plant.id, plant.id, year, sep = "_") %>%
  column_to_rownames("plant.id") %>%
  clr() %>%
  # put rownames back as a column
  as.data.frame() %>%
  rownames_to_column("plant.id") %>%
    # move year and plant id back to being sepearte columns
  separate(plant.id, into = c("plant.id", "year"), sep = "_") %>%
  pivot_longer(-c(plant.id, year),
               names_to = "compound",
               values_to = "ngghr") %>%
  # add a value of 2 to each ngghr to remove negatives
  mutate(ngghr = ngghr + 2) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  # join
  left_join(joined.data %>%
              dplyr::select(plant.id, swa, year, type, pair2) %>%
              mutate(plant.id = as.character(plant.id)),
            by = c("plant.id","year")) %>%
  mutate_at(vars(swa, year), list(factor)) %>%
  distinct() %>%
  na.omit() %>%
  # i'm going to ignore the "stressed" 
  # designation and just make those type "leaf"
  mutate(type = recode(type, "leaves" = "l", "flowers" = "f", "s" = "l")) %>%
  filter(!type == 0) %>%
  distinct()

```

## Field Volatile Collection
### Boxplot Comparison
```{r all comparison boxplot, echo = F, fig.align = 'center'}
joined.data.clr %>%
  # pivot to make all compounds in one column
  pivot_longer(-c(plant.id, year, swa, type, pair2),
               names_to = "compound",
               values_to = "ngghr") %>%
  # ggplot code
  ggplot(aes(compound, ngghr, fill = swa)) +
  geom_boxplot(position = position_dodge(width = 0.9),
                 outlier.color = NA,
                 alpha = 0.5) +
  geom_jitter(aes(compound, ngghr),
              alpha = 0.3,
              position = position_jitterdodge(jitter.width = 0.2,
                                              dodge.width = .8),
              show.legend = FALSE,
              # pch = 21, 
              stroke = 0, 
              size = 1.5) +
  theme(axis.text.x = element_text(angle = 45, 
                                    hjust = 1),
        legend.background = element_rect(color = "black"),
        # legend.position = c(0.8,0.9),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(color = "black"),
        panel.grid.major.x = element_line(color = "#F2F2F2")) +
  labs(x = element_blank(),
       y = "VOC quantity (CLR(ng/g/hr))",
       title = "2022/23 Field Locoweed Volatile Profile Comparison") +
  scale_fill_discrete(name = "Endophyte",
                      labels = c("Not Present", "Present"))
```

### PERMANOVA
```{r PERMDISP & PERMANOVA data, echo = F}
set.seed(1)
permanova.data <-
  joined.data.clr %>%
  dplyr::select(-swa,-type,-year,-pair2) %>%
  column_to_rownames("plant.id") %>%
  na.omit() %>%
  mutate_all(~.^2) %>%
  mutate_all(~.*100000) %>%
  mutate_all(round, digits = 0)

# dist.matrix <-
# vegdist(t(permanova.data),
#         method = "canberra")

dist.matrix <-
vegdist(permanova.data,
        method = "canberra")

data.per.meta <-
  joined.data.clr %>%
  dplyr::select(swa,type,year,pair2) %>%
  mutate(year = factor(year, levels = c("2023", "2022")))
```
```{r 22/23 FIELD PERMANOVA, echo = F, results = F}
set.seed(1)
permanova.output <-
adonis2(permanova.data ~ data.per.meta$swa, 
        strata = interaction(data.per.meta$pair2, 
                             data.per.meta$type, 
                             data.per.meta$year),
        method = "canberra",
        permutations = 999,
        by = "terms",
        na.rm = TRUE)
permanova.output
```
```{r PERMANOVA p-values, echo  = F}
p.value.swa <- permanova.output$`Pr(>F)`[1]
# p.value.year <- permanova.output$`Pr(>F)`[2]
# p.value.type <- permanova.output$`Pr(>F)`[3]
```

The locoweed samples, both SWA + and SWA -, exhibited numerous chromatogram peaks during collection. However, only volatile compoudns showing significant peaks inthe mass spectroscopy output or presenting differences in the overlays of the two SWA types were chosen for analysis.

There was no interaction between plant species and data of volatile collection.

Our results show **weak** evidence of a difference between SWA+ and SWA- locoweed plants (p-val = `r p.value.swa`).

## Greenhouse Volatile Collection
<!-- LOAD DATA -->
```{r gh RAW 22, echo = F}
raw.gh.22 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Greenhouse/2022\ Spring\ Greenhouse/Spring\ 2022\ Locoweed/Spring\ 2022\ GH\ Locoweed\ RESULTS.xlsx", sheet = "RAW2") %>%
  filter(remove == "no")
```
```{r gh RAW 23, echo = F}
raw.gh.23 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Greenhouse/2023\ Spring\ Greenhouse/Spring\ GH\ 23\ Locoweed\ Volatiles.xlsx", sheet = "RAW_PROC")

```
<!-- CLEAN DATA -->
```{r gh CLEAN 22, echo = F}
clean.gh.22 <-
raw.gh.22 %>%
  # filter(compound %in% c("heptanal", "sixmethyl", "octanal",
  #                        "z3hex", "dlimonene", "nonanal", 
  #                        "decanal", "methylsalicylate", "cisthreehexiso")) %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  filter(!compound == "standard") %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  # pivot_wider(names_from = "compound",
  #             values_from = "ngghr") %>%
  left_join(raw.gh.22 %>% 
              dplyr::select(plant.id, swa, pair), # go back and add pairs later
            by = "plant.id",
            relationship = "many-to-many") %>%
  # replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2022")
```
```{r gh CLEAN 23, echo = F}
clean.gh.23 <-
raw.gh.23 %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  filter(!compound == "standard") %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  # pivot_wider(names_from = "compound",
              # values_from = "ngghr") %>%
  left_join(raw.gh.23 %>% 
              dplyr::select(plant.id, swa, pair), # go back and add pairs later
            by = "plant.id",
            relationship = "many-to-many") %>%
  # replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2023") 
```
<!-- COMOPUNDS -->
```{r gh compounds, echo = F, results = F}
compounds22 <-
raw.gh.22 %>%
  dplyr::select(compound) %>%
  na.omit() %>% distinct() 
compounds23 <-
raw.gh.23 %>%
  dplyr::select(compound) %>%
  na.omit() %>% distinct() 
compounds22
compounds23 %>%
  print(n = 100)
gh.compounds <- as.list(intersect(compounds22, compounds23))$compound
```
<!-- CLR -->
```{r gh CLR, echo = F}
gh.joined.data <-
clean.gh.22 %>%
  dplyr::select(plant.id, compound, ngghr, swa, year, pair) %>%
  rbind(clean.gh.23 %>%
          dplyr::select(plant.id, compound, ngghr, swa, year, pair)) %>%
  distinct() %>%
  filter(compound %in% gh.compounds) %>%
  na.omit()

# having trouble joining the two data sets because of the pair2 column 
gh.joined.data.clr <-
gh.joined.data %>%
  dplyr::select(plant.id, ngghr, compound, year) %>%
  # replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
    # combine year and plant id into one column
  unite(plant.id, plant.id, year, sep = "_") %>%
  column_to_rownames("plant.id") %>%
  clr() %>%
  # put rownames back as a column
  as.data.frame() %>%
  rownames_to_column("plant.id") %>%
    # move year and plant id back to being sepearte columns
  separate(plant.id, into = c("plant.id", "year"), sep = "_") %>%
  pivot_longer(-c(plant.id, year),
               names_to = "compound",
               values_to = "ngghr") %>%
  # add a value of 2 to each ngghr to remove negatives
  mutate(ngghr = ngghr + 4) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  # join
  left_join(gh.joined.data %>%
              dplyr::select(plant.id, swa, year, pair) %>%
              mutate(plant.id = as.character(plant.id)),
            by = c("plant.id","year")) %>%
  mutate_at(vars(swa, year, pair), list(factor)) %>%
  distinct() %>%
  na.omit()
```

### Boxplot Comparison
```{r gh boxplot comparison, echo = F}
gh.joined.data.clr %>%
  # pivot to make all compounds in one column
  pivot_longer(-c(plant.id, year, swa, pair),
               names_to = "compound",
               values_to = "ngghr") %>%
  # ggplot code
  ggplot(aes(compound, ngghr, fill = swa)) +
  geom_boxplot(position = position_dodge(width = 0.9),
                 outlier.color = NA,
                 alpha = 0.5) +
  geom_jitter(aes(compound, ngghr),
              alpha = 0.3,
              position = position_jitterdodge(jitter.width = 0.2,
                                              dodge.width = .8),
              show.legend = FALSE,
              # pch = 21, 
              stroke = 0, 
              size = 1.5) +
  theme(axis.text.x = element_text(angle = 45, 
                                    hjust = 1),
        legend.background = element_rect(color = "black"),
        # legend.position = c(0.8,0.9),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(color = "black"),
        panel.grid.major.x = element_line(color = "#F2F2F2")) +
  labs(x = element_blank(),
       y = "VOC quantity (CLR(ng/g/hr))",
       title = "2022/23 Field Locoweed Volatile Profile Comparison") +
  scale_fill_discrete(name = "Endophyte",
                      labels = c("Not Present", "Present"))
```

### PERMANOVA
```{r gh PERMANOVA data, echo = F}
set.seed(1)
gh.permanova.data <-
  gh.joined.data.clr %>%
  dplyr::select(-swa, -pair) %>%
  unite(plant.id, plant.id, year, sep = "_") %>%
  column_to_rownames("plant.id") %>%
  na.omit() %>%
  mutate_all(~.^2) %>%
  mutate_all(~.*100000) %>%
  mutate_all(round, digits = 0)

# meta data
gh.data.per.meta <-
  gh.joined.data.clr %>%
  dplyr::select(plant.id, swa,year,pair) %>%
  mutate(year = factor(year, levels = c("2023", "2022")))
```

```{r PERMANOVA, echo = F}
set.seed(1)
gh.permanova.output <-
adonis2(gh.permanova.data ~ gh.data.per.meta$swa, 
        strata = interaction(gh.data.per.meta$pair, 
                             gh.data.per.meta$year),
        # method = "canberra",
        permutations = 999,
        by = "terms")

gh.permanova.output
```

```{r gh PERMANOVA pvalues, echo = F}
gh.p.value.swa <- gh.permanova.output$`Pr(>F)`[1]
```

Our results show **moderate** evidence of a difference between SWA+ and SWA- locoweed plants (p-val = `r gh.p.value.swa`).

# Data Overview

# APPENDIX

```{R clr data set print, echo = F}
set.seed(25)
kable(
joined.data.clr %>%
  dplyr::select(plant.id, bocimene, z3hex, nonanal, swa) %>%
  # rename(pair = pair2) %>%
  sample_n(size = n(), replace = FALSE) %>%
  mutate(bocimene = round(bocimene, digits = 2),
         z3hex = round(z3hex, digits = 2),
         nonanal = round(nonanal, digits = 2)) %>%
  head(),
align = c("l", "l", "r", "c", "c", "c", "c")
) %>%
  kable_styling(full_width = FALSE, position = "center")

```

### 2024-02-29 msu-masters.qmd
update website

---
title: "MSU Masters"

---
<figure>
  <img src="img/msu-masters/cages.heic" alt="Montana State University" 
  width = 250" height = "200">
</figure>

<figure>
  <img src="img/msu-masters/vcs.heic" alt="Montana State University" 
  width = 200" height = "300">
</figure>

<figure>
  <img src="img/msu-masters/field.heic" alt="Montana State University" 
  width = 250" height = "200">
</figure>

### 2024-02-29 pfreport.qmd
update website

--- 2024-02-29 field_report.qmd
@@
 ---
-title: "MT Field Site Analysis Report"
+title: "MT Post Farm Analysis Report"
 author: "Jackson Strand"
@@
   html_document:
-    self_contained: yes
-    include:
-    keep_md: true
     theme: cosmo
@@
   chunk_output_type: console
+
+website:
+  sidebar:
+    style: "docked"
+    contents:
+      - section: "Test1"
+        contents: index.qmd
+      - section: "Test2"
+        contents: index.qmd
 ---
@@
 <hr>
-
-# MONTANA FIELD SITES
-
-## Methods
-
-To assess C cinctus infestation, larval mortality, and parasitoid prevalence within B.
-inermis and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production.
-Samples were collected in both early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA.
-Sample locations were chosen by identifying fields with consistent wheat production that were adjacent to healthy B.
-inermis stands, either situated between fields or between fields and roadways.
-Polygons containing B.
-inermis were set up along the periphery of adjacent wheat fields, each covering an area of approximately 100 meters.
-Four collection squares of 1ft x 1ft were randomly selected within each polygon during both collection events each year.
-All stems within each 1 x 1 ft square were collected.
-The stems were then processed and split in the laboratory to assess infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting.
-
-## Analysis
-
-```{r (raw) brome data, echo = F}
-raw.2022 <- 
-  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "n.infested") %>%
-  mutate(year = "2022") %>%
-  rename("site" = site.cap)
-
-raw.2023 <-
-read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "n.infested") %>%
-  mutate(year = "2023") 
-```
-```{r (stubble.####) stubble data, echo = F}
-stubble.2023 <-
-  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2023/2023\ Field Site\ Data.xlsx", 
-           sheet = "stubble") %>%
-  dplyr::select(-name) %>%
-  na.omit() %>%
-  mutate(year = "2023")
-
-stubble.2022 <-
-  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2022/2022 FS Stem Data (2).xlsx", 
-             sheet = "stubble") %>%
-  mutate(year = "2022")
-
-```
-```{r (joined.stubble) stubble join, echo = F}
-# the below code joins the 2022 and 2023 stubble data frames
-joined.stubble <-
-stubble.2023 %>%
-  dplyr::select(site, location, distance,
-                inf.nc, cut, cocoon, eh, total, year) %>%
-  mutate(collection = "fall") %>% 
-  rbind(stubble.2022 %>%
-          dplyr::select(site, year, location, distance, collection,
-                        inf.nc, cut, cocoon, eh, total)) %>%
-  # remove N sites and re add as NS and NN for example
-  group_by(site, location, year, distance, collection) %>%
-  mutate(inf = sum(cut, inf.nc, cocoon, eh)) %>%
-  # first, we remove the site arguement by grouping by everything else
-  group_by(site, location, year, distance, collection) %>%
-  dplyr::select(-c(inf.nc)) %>%
-  rename("e" = eh,
-         "c" = cocoon) %>%
-  mutate(p.inf = inf/total,
-         p.par = (c + e)/inf,
-         p.cut = cut/inf,
-         poten = (cut + e + c)/inf)
-
-# duplicate the wheat data to match up with brome data locations
-# we are duplicating some of the stubble data so that we have stubble data
-# for the brome sections that are across the road from stubble data
-joined.stubble <-
-  joined.stubble %>%
-  bind_rows(joined.stubble %>%
-              filter(location == "n") %>%
-              mutate(location = "ns")) %>%
-  bind_rows(joined.stubble %>%
-              filter(location == "n") %>%
-              mutate(location = "nn")) %>%
-  filter(!location == "n") %>%
-  bind_rows(joined.stubble %>%
-              filter(location == "har") %>%
-              mutate(location = "sun"))
-
-```
-```{r (n.inf) brome join, echo = F}
-n.inf <-
-raw.2023 %>%
-  dplyr::select(site, year, location, collection, 
-              n.uv, n.ur, n.iv, n.ir, n.ix, 
-              cut, e, c)  %>%
-  bind_rows(raw.2022 %>%
-              rename(n.uv = uv,
-                     n.ur = ur,
-                     n.iv = iv,
-                     n.ir = ir,
-                     n.ix = ix) %>%
-              dplyr::select(site, year, location, collection, 
-                    n.uv, n.ur, n.iv, n.ir, n.ix, 
-                    cut, e, c)) %>%
-  group_by(site, location, collection, year) %>%
-  mutate(inf = sum(n.iv, n.ir, n.ix),
-         total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
-  ungroup() %>%
-  dplyr::select(-c(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
-  mutate(p.inf = inf/total,
-         p.par = (c + e)/inf,
-         p.cut = cut/inf,
-         poten = (cut + c + e)/inf) %>%
-  pivot_longer(-c(site, location, collection, year),
-               names_to = "type",
-               values_to = "value") %>%
-  group_by(location, collection, year, type) %>%
-  distinct() %>%
-  mutate(distance = "brome",
-         location = case_when(
-           location == "kw" ~ "k",
-           location == "ke" ~ "k", 
-           TRUE ~ location))
-```
-```{r (stubble.brome) join stubble/brome, echo = F}
-# join the stubble data and the brome data
-stubble.brome <-
-joined.stubble %>%
-  pivot_longer(-c(site, location, distance, year, collection),
-               names_to = "type",
-               values_to = "value") %>%
-  distinct() %>%
-  rbind(n.inf) %>%  # rbind the brome data
-  filter(!type == "total") %>%  # remove type = "total"
-  mutate(type = case_when(
-    type == "c" ~ "p",
-    type == "e" ~ "p",
-    TRUE ~ type)) %>% # replace the e and c variables with p, for simplicity
-  ungroup() %>%
-  dplyr::select(-site) %>%
-  group_by(location, distance, year, collection, type) %>%
-  reframe(mean = mean(value),
-         count = n(),
-         sd = sd(value),
-         se = sd/sqrt(count)) %>%
-  distinct() %>%
-  filter(type %in% c("p.inf", "p.cut", "p.par", "poten")) %>%
-  distinct()
-```
-
-```{r, echo = F, eval = F}
-stubble.brome %>%
-  filter(location == "g",
-         distance %in% c("20", "5"),
-         year == "2023",
-         collection == "fall")
-```
-
-### Big Sandy, MT 2023
-
-#### Lochy Property - North Road
-
-![North Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/nnns-min.png?raw=true)
-
-**Note:**    
-- In 2023, the field south of North Road was fallow, so no stubble infestation data was collected from the field south of North Road.
-
-```{r NN & NS data table, echo = F}
-nn.table <-
-stubble.brome %>%
-  filter(location %in% c("nn"),
-         year == "2023") %>%
-  dplyr::select(location, distance, collection, year, type, mean) %>%
-  mutate(mean = round_based_on_size(mean)) %>%
-  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
-  pivot_wider(names_from = type,
-              values_from = mean,
-              values_fn = list(mean = mean)) %>%
-  arrange(distance) %>%
-  ungroup() %>%
-  dplyr::select(-year)
-
-kable(nn.table[, column.order2], 
-      caption = "North Road (site nn) Infestation Table", 
-      format = "html",
-      col.names = c("Location", "Distance", "Collection", 
-                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
-      align = "lllcccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2")
-
-ns.table <-
-stubble.brome %>%
-  filter(location %in% c("ns"),
-         year == "2023",
-         distance == "brome") %>%
-  dplyr::select(location, distance, collection, year, type, mean) %>%
-  mutate(mean = round_based_on_size(mean)) %>%
-  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
-  pivot_wider(names_from = type,
-              values_from = mean,
-              values_fn = list(mean = mean)) %>%
-  arrange(distance) %>%
-  ungroup() %>%
-  dplyr::select(-year)
-
-kable(ns.table[, column.order2], 
-      caption = "North Road (site ns) Infestation Table", 
-      format = "html",
-      col.names = c("Location", "Distance", "Collection", 
-                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
-      align = "lllcccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2") 
-```
-
-```{r NN and NS plot, echo = F, fig.height = 3.5, fig.width = 10, warning = F}
-stubble.brome %>%
-  filter(location %in% c("nn", "ns"),
-         year == "2023",
-         collection == "fall") %>%
-  ggplot(aes(x = type, y = mean, fill = distance)) +
-  geom_bar(stat = "identity",
-           position = position_dodge()) +
-  geom_errorbar(aes(ymin = mean - se,
-                    ymax = mean + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  theme_bw() +
-  theme(legend.title = element_blank()) +
-  scale_fill_brewer(palette = "Purples",
-                    labels = c("Adjacet wheat: 20 m", "Adjacet wheat: 5 m", "Smooth brome")) +
-  labs(y = "Percent (%)",
-       x = element_blank(),
-       title = "North Road Infestation") +
-  scale_x_discrete(labels = c("p.cut" = "Cut",
-                              "p.inf" = "Infested",
-                              "p.par" = "Parasitized",
-                              "poten" = "Potential Cut")) +
-  facet_wrap(~location)
-```
-
-We saw very low levels of infestation in 2023 within the wheat.
-This was likely due a cold, mild spring that led to late stem development.
-Because the wheat was not well developed, the sawflies did not infest.
-This appears to have led to high infestation within the smooth brome.
-Will be interesting to compare to 2022 when stem elongation followed normal expected timelines.
-
-#### Lochy Property - Gardiner Road
-
-![Gardiner Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/gardiner%20road-min.png?raw=true)
-
-In 2022, Gardiner road underwent a major reconstruction with both ditches being regraded. This severely diminished the amount smooth brome found along the edge of the roadway.
-
-```{r G table, echo = F}
-g.table <-
-stubble.brome %>%
-  filter(location %in% c("g"),
-         year == "2023") %>%
-  dplyr::select(location, distance, collection, year, type, mean) %>%
-  mutate(mean = round_based_on_size(mean)) %>%
-  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
-  pivot_wider(names_from = type,
-              values_from = mean,
-              values_fn = list(mean = mean)) %>%
-  arrange(distance) %>%
-  ungroup() %>%
-  dplyr::select(-c(year, location))
-
-kable(g.table[, column.order], caption = "Gardiner Road Infestation Table", 
-      format = "html",
-      col.names = c("Distance", "Collection", 
-                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
-      align = "lllcccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2")
-```
-
-```{r G plot, echo = F, fig.height = 3.5, fig.align = 'center', warning = F}
-stubble.brome %>%
-  filter(location %in% c("g"),
-         year == "2023",
-         collection == "fall") %>%
-  ggplot(aes(x = type, y = mean, fill = distance)) +
-  geom_bar(stat = "identity",
-           position = position_dodge()) +
-  geom_errorbar(aes(ymin = mean - se,
-                    ymax = mean + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  theme_bw() +
-  theme(legend.title = element_blank()) +
-  scale_fill_brewer(palette = "Purples",
-                    labels = c("Adjacet wheat: 20 m", "Adjacet wheat: 5 m", "Smooth brome")) +
-  labs(y = "Percent (%)",
-       x = element_blank(),
-       title = "Gardiner Rd Infestation") +
-  scale_x_discrete(labels = c("p.cut" = "Cut",
-                              "p.inf" = "Infested",
-                              "p.par" = "Parasitized",
-                              "poten" = "Potential Cut"))
-```
-
-Similar results to North Road seen here.
-Low infestation in the adjacent wheat field was due to the cool spring and lack of infestable stems during the sawfly flight.
-This shows that even in odd years like 2023, sawflies are still present in the smooth brome, meaning parasitoids will still have larval hosts.
-
-### Moccasin, MT 2023
-
-#### Sunset & Harvest Road Intersection
-
-![Harvest & Sunset Road Instersection, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/harvest%20road-min.png?raw=true)
-
-```{r SUN/HAR data tables, echo = F}
-sun.table <-
-stubble.brome %>%
-  filter(location %in% c("sun"),
-         year == "2023") %>%
-  dplyr::select(location, distance, collection, year, type, mean) %>%
-  mutate(mean = round_based_on_size(mean)) %>%
-  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
-  pivot_wider(names_from = type,
-              values_from = mean,
-              values_fn = list(mean = mean)) %>%
-  arrange(distance) %>%
-  ungroup() %>%
-  dplyr::select(-year)
-
-kable(sun.table[, column.order2], caption = "Sunset Road Infestation Table", 
-      format = "html",
-      col.names = c("Location", "Distance", "Collection", 
-                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
-      align = "lllcccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2")
-
-har.table <-
-stubble.brome %>%
-  filter(location %in% c("har"),
-         year == "2023") %>%
-  dplyr::select(location, distance, collection, year, type, mean) %>%
-  mutate(mean = round_based_on_size(mean)) %>%
-  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
-  pivot_wider(names_from = type,
-              values_from = mean,
-              values_fn = list(mean = mean)) %>%
-  arrange(distance) %>%
-  ungroup() %>%
-  dplyr::select(-year)
-
-kable(har.table[, column.order2], caption = "Harvest Road Infestation Table", 
-      format = "html",
-      col.names = c("Location", "Distance", "Collection", 
-                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
-      align = "lllcccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2")
-```
-
-```{r SUN/HAR plot, echo = F, warning = F, fig.align = 'center', fig.height = 3.5, fig.width = 10}
-stubble.brome %>%
-  filter(location %in% c("sun", "har"),
-         year == "2023",
-         collection == "fall") %>%
-  ggplot(aes(x = type, y = mean, fill = distance)) +
-  geom_bar(stat = "identity",
-           position = position_dodge()) +
-  geom_errorbar(aes(ymin = mean - se,
-                    ymax = mean + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  theme_bw() +
-  theme(legend.title = element_blank()) +
-  guides(color = guide_legend(title.position = "top", 
-                              label.padding = unit(10, "lines"))) +
-  scale_fill_brewer(palette = "Purples",
-                    labels = c("Adjacet wheat: 20 m", "Adjacet wheat: 5 m", "Smooth brome")) +
-  labs(y = "Percent (%)",
-       x = element_blank(),
-       title = "Sunset & Harvest Rd Infestation") +
-  scale_x_discrete(labels = c("p.cut" = "Cut",
-                              "p.inf" = "Infested",
-                              "p.par" = "Parasitized",
-                              "poten" = "Potential Cut")) +
-  facet_wrap(~location)
-```
-
-#### Kolin Road
-
-![Kolin Road, Moccasin, MT (2023)](https://github.com/jstrand894/Website/blob/main/kolin%20road-min.png?raw=true)
-
-```{r K data table, echo = F}
-k.table <-
-stubble.brome %>%
-  filter(location %in% c("k"),
-         year == "2023") %>%
-  dplyr::select(location, distance, collection, year, type, mean) %>%
-  mutate(mean = round_based_on_size(mean)) %>%
-  mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
-  pivot_wider(names_from = type,
-              values_from = mean,
-              values_fn = list(mean = mean)) %>%
-  arrange(distance) %>%
-  ungroup() %>%
-  dplyr::select(-year)
-
-kable(k.table[, column.order], caption = "Kolin Road Infestation Table", 
-      format = "html",
-      col.names = c("Distance", "Collection", 
-                    "% Inf", "Cut", "Parasitized", "Potential Cut"),
-      align = "lllcccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2")
-```
-
-```{r K plot, echo = F, warning = F, fig.height = 3, fig.align = 'center'}
-stubble.brome %>%
-  filter(location %in% c("k"),
-         year == "2023",
-         collection == "fall") %>%
-  ggplot(aes(x = type, y = mean, fill = distance)) +
-  geom_bar(stat = "identity",
-           position = position_dodge()) +
-  geom_errorbar(aes(ymin = mean - se,
-                    ymax = mean + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  theme_bw() +
-  theme(legend.title = element_blank()) +
-  scale_fill_brewer(palette = "Purples",
-                    labels = c("Adjacet wheat: 20 m", "Adjacet wheat: 5 m", "Smooth brome")) +
-  labs(y = "Percent (%)",
-       x = element_blank(),
-       title = "Kolin Rd Infestation") +
-  scale_x_discrete(labels = c("p.cut" = "Cut",
-                              "p.inf" = "Infested",
-                              "p.par" = "Parasitized",
-                              "poten" = "Potential Cut")) 
-```

### 2024-02-29 projects.qmd
update website

---
title: "Projects"
editor_options: 
  chunk_output_type: console
---
::: {style="display: flex;"}

<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
  <a href="msu-masters.qmd">
    <img src="img/msu-logo.jpg" alt="Montana State University" width="200" height="200">
  </a>
  <figcaption style="text-align: center;">Montana State University</figcaption>
</figure>

<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
  <a href="bballref.qmd">
    <img src="img/bballref-logo.jpeg" alt="Basketball Reference Tool" width="200" height="200">
  </a>
  <figcaption style="text-align: center;">Basketball Reference Tool</figcaption>
</figure>

:::

### 2024-02-29 running.qmd
update website

---
title: "Running Pictures"
---
::: {style="display: flex;"}
<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
  <a href="deadhorse.qmd">
    <img src="img/deadhorse/deadhorse-top.JPG" alt="Deadhorse Ultra" width="300" height="200">
  </a>
  <figcaption style="text-align: center;">Deadhorse Ultra, 2023</figcaption>
</figure>

<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
  <a href = "tmr50k.qmd">
    <img src="img/tmr50k/tmr50k-top.JPG" alt="Deadhorse Ultra" 
  width = 300" height = "200">
  </a>
  <figcaption style="text-align:center;">Deadhorse Ultra, 2023</figcaption>
</figure>

:::

### 2024-02-29 tmr50k.qmd
update website

---
title: "Teton Mountain Runs 50K"
---

::: {style="display: block;"}

```{r get images, echo = F, results = 'asis'}
image_files <- list.files("img/tmr50k", pattern = ".JPG|.png|.gif|.jpg|.jpeg", full.names = TRUE)

for (file in image_files) {
  cat("![](",file, ")\n  \n")
}
```

:::

### 2024-03-01 aboutme.qmd
update website

@@
 
-[My CV](Aboutme/jackson_strand_updated_cv.pdf)
+```{=html}
+<div  style="margin: 30px; text-align: center;">
+  <a class="btn btn-primary" href="cv.qmd" role="button" style="padding: 15px 30px;">View my CV</a>
+</div>
+```
 
@@
 :::
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(200px, 1fr));
+      grid-gap: 10px;
+    }
+
+    .gallery img {
+      width: 100%;
+      height: auto;
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+
+    .gallery img:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
+      display: none; /* Initially hidden */
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
+
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+  <img src="img/bugs/blueflowers.JPG" alt="Image 1" onclick="expandImage(this)">
+  <img src="img/bugs/brownmoth.JPG" alt="Image 2" onclick="expandImage(this)">
+  <img src="img/bugs/butterfly-purpleflower.JPG" alt="Image 3" onclick="expandImage(this)">
+  <!-- Add more images as needed -->
+</div>
+
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    // Calculate the click position relative to the expanded image
+    var clickX = event.pageX - event.target.offsetLeft;
+    var imageWidth = event.target.offsetWidth;
+
+    // Determine if the click was on the left or right side of the image
+    if (clickX < imageWidth / 2) {
+      // Clicked on the left side: navigate to the previous image
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      // Clicked on the right side: navigate to the next image
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    // Update the expanded image source
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+</script>
+
+</body>
+</html>
+
+```

### 2024-03-01 cv.qmd
update website

---
layout: page
title: "Cirriculum vitae"
execute:
  freeze: true  # never re-render during project render
engine: knitr 
resources:
  - "Aboutme/jackson_strand_updated_cv.pdf"
cv:
  pdf: "Aboutme/jackson_strand_updated_cv.pdf"
---

```{css echo=FALSE}
.embed-container {
    position: relative;
    padding-bottom: 129%;
    height: 0;
    overflow: hidden;
    max-width: 100%;
}
.embed-container iframe,
.embed-container object,
.embed-container embed {
    position: absolute;
    top: 0;
    left: 0;
    width: 100%;
    height: 100%;
}
```

```{=html}
<p class="text-center">
  <a class="btn btn-primary btn-lg cv-download" href="`r rmarkdown::metadata$cv$pdf`" target="_blank">
    <i class="fa-solid fa-file-arrow-down"></i>&ensp;Download current CV
  </a>
</p>
<div class="embed-container">
  <iframe src="`r rmarkdown::metadata$cv$pdf`#toolbar=0" style="border: 0.5px;width: 100%;"></iframe>
</div>
```

### 2024-03-01 index.qmd
update website

@@
 ---
-title: "About Me"
+title: "Jackson Strand"
 about:
   id: me
-  template: broadside
-  image: img/profile.png
+  template: marquee
+  links:
+    - icon: linkedin
+      text: LinkedIn
+      href: https://linkedin.com/in/jacksonstrand
+    - icon: instagram
+      text: Instagram
+      href: https://www.instagram.com/jstrand894/
+    - icon: github
+      text: GitHub
+      href: https://github.com/jstrand894 
+  image: img/aboutme/deadhorse.jpeg
+
+sidebar: false
 ---
@@
 
-## Goals
+I am currently a Master's graduate student at Montana State University studying entomology. I live in Bozeman, MT with my wife Grace.
 
-## Education
+```{=html}
+<div  style="margin: 30px; text-align: center;">
+  <a class="btn btn-primary" href="cv.qmd" role="button" style="padding: 15px 30px;">View my CV</a>
+</div>
+```
 
-## Experience
+![Me and Grace at Glacier National Park, 2023](img/aboutme/iceberg.JPG)
 
 :::
-
-Thanks for checking out my web site!

### 2024-03-05 aboutme.qmd
Update website 4/5/24

@@
     }
-
     .gallery img {
@@
     }
-
     .gallery img:hover {
@@
     }
-
     .expanded-container {
@@
     // Calculate the click position relative to the expanded image
-    var clickX = event.pageX - event.target.offsetLeft;
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
     var imageWidth = event.target.offsetWidth;

### 2024-03-05 bugs.qmd
Update website 4/5/24

@@
 ---
-title: "bug pictures"
-filters:
-  - lightbox
-lightbox: auto
+title: "Bugs"
+output: html_document
 ---
 
-![Bug pic 1](img/bugs/IMG_1664.JPG){group="my-gallery"}
-![Bug pic 2](img/bugs/IMG_1678.JPG){group="my-gallery"}
+```{r  echo = F, eval = F}
+# Set the directory path
+directory_path <- "img/bugs"
 
-<figure>
-  <img src="img/bugs/IMG_1664.JPG" alt="Montana State University" 
-  width = 200" height = "200">
-</figure>
+# List all the files in the directory
+file_names <- list.files(directory_path)
+
+# Exclude files with .zip extension
+file_names <- file_names[!grepl("\\.zip$", file_names)]
+
+# Initialize an empty character vector to store the HTML code
+html_code <- "<div class=\"gallery\">\n"
+
+# Loop through each file name and generate HTML code
+for (i in seq_along(file_names)) {
+  # Generate the HTML code for each image
+  image_html <- paste0('<img src="', directory_path, "/", file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')
+
+  # Append the generated HTML code to the existing code
+  html_code <- paste(html_code, image_html, sep = "")
+}
+
+# Close the gallery div
+html_code <- paste(html_code, "</div>\n<div class=\"expanded-container\" onclick=\"closeExpanded()\">\n  <img class=\"expanded\" id=\"expanded-image\" onclick=\"navigateGallery(event)\">\n</div>", sep = "")
+
+# Print the final HTML code
+cat(html_code)
+```
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
+      grid-gap: 10px;
+    }
+    .gallery img {
+      width: 100%;
+      height: 100%;
+      object-fit: cover; /* Ensure images cover the entire space */
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .gallery img:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
+      display: none; /* Initially hidden */
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
+
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+<img src="img/bugs/blueflowers.JPG" alt="Image 1" onclick="expandImage(this)">
+<img src="img/bugs/brownmoth.JPG" alt="Image 2" onclick="expandImage(this)">
+<img src="img/bugs/brownred-butterfly.JPG" alt="Image 3" onclick="expandImage(this)">
+<img src="img/bugs/butterfly-purpleflower.JPG" alt="Image 4" onclick="expandImage(this)">
+<img src="img/bugs/loco-bee.JPG" alt="Image 5" onclick="expandImage(this)">
+</div>
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    // Calculate the click position relative to the expanded image
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
+    var imageWidth = event.target.offsetWidth;
+
+    // Determine if the click was on the left or right side of the image
+    if (clickX < imageWidth / 2) {
+      // Clicked on the left side: navigate to the previous image
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      // Clicked on the right side: navigate to the next image
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    // Update the expanded image source
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+</script>
+
+</body>
+</html>
+
+```
+
+<br>
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Button</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+</style>
+</head>
+<body>
+
+<!-- Button to download the file -->
+<a href="img/bugs/Jackson-Bug-Pics.zip" download="Jackson Bug Pictures">
+  <button class="button">Download All Pictures</button>
+</a>
+
+</body>
+</html>
+
+```
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Image Crop Example</title>
+<style>
+  .image-container {
+    width: 300px; /* Set the width of the container */
+    height: 200px; /* Set the height of the container */
+    overflow: hidden; /* Hide any overflowing content */
+  }
+  .cropped-image {
+    width: 100%; /* Set the width of the image to fill the container */
+    height: auto; /* Automatically adjust the height to maintain aspect ratio */
+    object-fit: cover; /* Crop the image to cover the container */
+  }
+</style>
+</head>
+<body>
+
+<div class="image-container">
+  <img class="cropped-image" src="img/bugs/loco-bee.png" alt="Cropped Image">
+</div>
+
+</body>
+</html>
+
+```

### 2024-03-05 deadhorse.qmd
Update website 4/5/24

@@
 
-::: {style="display: grid;grid-template-columns: repeat(auto-fill, minmax(200px, 1fr));grid-gap: 1em;"}
-```{r get images, echo = F, results = 'asis'}
-image_files <- list.files("img/deadhorse", pattern = ".JPG|.png|.gif|.jpg|.jpeg", full.names = TRUE)
+```{r eval = F, echo = F}
+# Set the directory path
+directory_path <- "img/deadhorse"
 
-for (file in image_files) {
-  cat("![](",file, ")\n  \n")
+# List all the files in the directory
+file_names <- list.files(directory_path)
+
+# Exclude files with .zip extension
+file_names <- file_names[!grepl("\\.zip$", file_names)]
+
+# Initialize an empty character vector to store the HTML code
+html_code <- "<div class=\"gallery\">\n"
+
+# Loop through each file name and generate HTML code
+for (i in seq_along(file_names)) {
+  # Generate the HTML code for each image
+  image_html <- paste0('<img src="', directory_path, "/", file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')
+
+  # Append the generated HTML code to the existing code
+  html_code <- paste(html_code, image_html, sep = "")
 }
+
+# Close the gallery div
+html_code <- paste(html_code, "</div>\n<div class=\"expanded-container\" onclick=\"closeExpanded()\">\n  <img class=\"expanded\" id=\"expanded-image\" onclick=\"navigateGallery(event)\">\n</div>", sep = "")
+
+# Print the final HTML code
+cat(html_code)
 ```
-:::
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
+      grid-gap: 10px;
+    }
+    .gallery img {
+      width: 150px;
+      height: 150px;
+      object-fit: cover; /* Ensure images cover the entire space */
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .gallery img:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
+      display: none; /* Initially hidden */
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
+
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+<img src="img/deadhorse/deadhorse-top.JPG" alt="Image 1" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_0012.JPG" alt="Image 2" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_0016.JPG" alt="Image 3" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_0026.JPG" alt="Image 4" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6044.jpeg" alt="Image 5" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6050.jpeg" alt="Image 6" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6058.jpeg" alt="Image 7" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6061.jpeg" alt="Image 8" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6062.jpeg" alt="Image 9" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6067.jpeg" alt="Image 10" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6071.jpeg" alt="Image 11" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6078.jpeg" alt="Image 12" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6079.jpeg" alt="Image 13" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6080.jpeg" alt="Image 14" onclick="expandImage(this)">
+<img src="img/deadhorse/IMG_6192.HEIC" alt="Image 15" onclick="expandImage(this)">
+</div>
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    // Calculate the click position relative to the expanded image
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
+    var imageWidth = event.target.offsetWidth;
+
+    // Determine if the click was on the left or right side of the image
+    if (clickX < imageWidth / 2) {
+      // Clicked on the left side: navigate to the previous image
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      // Clicked on the right side: navigate to the next image
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    // Update the expanded image source
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+
+  // Add event listeners to the left and right sides of the expanded image for navigation
+  document.querySelector('.expanded').addEventListener('click', function(event) {
+    navigateGallery(event);
+  });
+</script>
+
+</body>
+</html>
+
+```
+<br>
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Button</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+</style>
+</head>
+<body>
+
+<!-- Button to download the file -->
+<a href="img/deadhorse/DeadhorseUltra2023.zip" download="DeadhorseUltra2023">
+  <button class="button">Download All Pictures</button>
+</a>
+
+</body>
+</html>
+
+```

### 2024-03-05 eag.qmd
Update website 4/5/24

---
title: "eag"
editor_options: 
  chunk_output_type: console
---

```{r library, echo = F, message = F}
library(dplyr)
library(ggplot2)
library(readxl)
library(plotly)
```
```{r scale values, echo = F}
# Define the original range
min_original <- 1
max_original <- 55680

# Define the target range
min_target <- 1.823
max_target <- 10.02

# Function to scale the values
scale_values <- function(x) {
  ((x - min_original) / (max_original - min_original)) * (max_target - min_target) + min_target
}

```

```{r, echo = F}
eag <-
  read_excel("xlsx files/eag.xlsx", sheet = "test eag")

fid <-
  read_excel("xlsx files/eag.xlsx", sheet = "Peak")

# Generate a sequence of MeasRetTime values every 0.05
new_MeasRetTime <- seq(min(fid$MeasRetTime), 
                       max(fid$MeasRetTime), by = 0.05)

# Create a data frame with the new MeasRetTime values
new_data <- data.frame(MeasRetTime = new_MeasRetTime)

# Merge the new data frame with the original data, filling missing values with 0
filled_data <- merge(fid, new_data, by = "MeasRetTime", all = TRUE)
filled_data[is.na(filled_data)] <- 0

fid_plot <- 
filled_data %>%
  dplyr::select(MeasRetTime, Area, Height) %>%
  mutate(Height = ifelse(MeasRetTime < 3.2, 0, Height))
  # ggplot() +
  # geom_point(aes(x = MeasRetTime, y = Height)) +
  # geom_line(aes(x = MeasRetTime, y = Height)) +
  # theme_bw()
eag_plot <-
eag %>%
  mutate(row_num = row_number()) %>%
  rename("EAD" = 'EAD 1') 
ggplotly(
ggplot() +
  geom_line(data = fid_plot, aes(x = MeasRetTime, 
                                 y = Height), 
            color = "red") +
  geom_line(data = eag_plot, aes(x = scale_values(row_num), 
                                 y = EAD * 100), 
            color = "blue") +

  scale_y_continuous(name = "EAD", sec.axis = sec_axis(~./100, name = "Height")) + 
  labs(title = "Comparison of EAD and Height",
       x = "Row Number (Scaled)", 
       y = "EAD",
       color = "Legend") +
  theme_bw() + 
  scale_x_continuous(breaks = seq(min(fid_plot$MeasRetTime,
                                      max(fid_plot$MeasRetTime), 
                                      by = 1)))
)

```

### 2024-03-05 field_report.qmd
Update website 4/5/24

@@
 output: 
-  html_document:
-    self_contained: yes
-    include:
-    keep_md: true
+  quarto::quarto_html:
+    layout: sidebar
     theme: cosmo
@@
 
-![North Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/nnns-min.png?raw=true)
+```{=html}
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Centered Image Expand</title>
+  <style>
+    body, html {
+      margin: 0;
+      padding: 0;
+      height: 100%;
+    }
+    .container {
+      display: flex;
+      justify-content: left;
+      align-items: left;
+      height: auto;
+    }
+    .image {
+      max-width: 50%;
+      max-height: 80vh;
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .image:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .image.expanded {
+      transform: scale(3); /* Scale up when expanded */
+      z-index: 9999; /* Ensure expanded image is on top */
+      position: fixed;
+      top: 50%;
+      left: 50%;
+      max-height: 80%;
+      max-width: 80%;
+      transform: translate(-50%, -50%); /* Center expanded image */
+    }
+    .overlay {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.5); /* Semi-transparent dark background */
+      z-index: 9998; /* Ensure overlay is behind the expanded image */
+      display: none; /* Initially hidden */
+    }
+    .image.expanded + .overlay {
+      display: block; /* Show overlay when image is expanded */
+    }
+  </style>
+</head>
+<body>
+
+<div class="container">
+  <img class="image" src="img/field-sites/nnns.png" alt="Image" onclick="expandImage(this)">
+  <div class="overlay" onclick="closeExpandedImage()"></div> <!-- Overlay element -->
+</div>
+
+<script>
+  function expandImage(element) {
+    element.classList.toggle('expanded');
+    document.querySelector('.overlay').classList.toggle('expanded');
+  }
+
+  function closeExpandedImage() {
+    var image = document.querySelector('.image');
+    image.classList.remove('expanded');
+    document.querySelector('.overlay').classList.remove('expanded');
+  }
+</script>
+
+</body>
+</html>
+```
+
+<!-- ![North Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/nnns-min.png?raw=true) -->
 
@@
 
-![Gardiner Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/gardiner%20road-min.png?raw=true)
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Centered Image Expand</title>
+  <style>
+    body, html {
+      margin: 0;
+      padding: 0;
+      height: 100%;
+    }
+    .container {
+      display: flex;
+      justify-content: left;
+      align-items: left;
+      height: auto;
+    }
+    .image {
+      max-width: 50%;
+      max-height: 80vh;
+      cursor: pointer;;
+      transition: transform 0.3s ease;
+    }
+    .image:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .image.expanded {
+      transform: scale(3); /* Scale up when expanded */
+      z-index: 9999; /* Ensure expanded image is on top */
+      position: fixed;
+      top: 50%;
+      left: 50%;
+      max-height: 80%;
+      max-width: 80%;
+      transform: translate(-50%, -50%); /* Center expanded image */
+    }
+    .overlay {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.5); /* Semi-transparent dark background */
+      z-index: 9998; /* Ensure overlay is behind the expanded image */
+      display: none; /* Initially hidden */
+    }
+    .image.expanded + .overlay {
+      display: block; /* Show overlay when image is expanded */
+    }
+  </style>
+</head>
+<body>
+
+<div class="container">
+  <img class="image" src="img/field-sites/gardiner.png" alt="Image" onclick="expandImage(this)">
+  <div class="overlay" onclick="closeExpandedImage()"></div> <!-- Overlay element -->
+</div>
+
+<script>
+  function expandImage(element) {
+    element.classList.toggle('expanded');
+    document.querySelector('.overlay').classList.toggle('expanded');
+  }
+
+  function closeExpandedImage() {
+    var image = document.querySelector('.image');
+    image.classList.remove('expanded');
+    document.querySelector('.overlay').classList.remove('expanded');
+  }
+</script>
+
+</body>
+</html>
+```
+
+<!-- ![Gardiner Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/gardiner%20road-min.png?raw=true) -->
 
@@
 
-![Harvest & Sunset Road Instersection, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/harvest%20road-min.png?raw=true)
+```{=html}
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Centered Image Expand</title>
+  <style>
+    body, html {
+      margin: 0;
+      padding: 0;
+      height: 100%;
+    }
+    .container {
+      display: flex;
+      justify-content: left;
+      align-items: left;
+      height: auto;
+    }
+    .image {
+      max-width: 50%;
+      max-height: 80vh;
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .image:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .image.expanded {
+      transform: scale(3); /* Scale up when expanded */
+      z-index: 9999; /* Ensure expanded image is on top */
+      position: fixed;
+      top: 50%;
+      left: 50%;
+      max-height: 80%;
+      max-width: 80%;
+      transform: translate(-50%, -50%); /* Center expanded image */
+    }
+    .overlay {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.5); /* Semi-transparent dark background */
+      z-index: 9998; /* Ensure overlay is behind the expanded image */
+      display: none; /* Initially hidden */
+    }
+    .image.expanded + .overlay {
+      display: block; /* Show overlay when image is expanded */
+    }
+  </style>
+</head>
+<body>
+
+<div class="container">
+  <img class="image" src="img/field-sites/harvest.png" alt="Image" onclick="expandImage(this)">
+  <div class="overlay" onclick="closeExpandedImage()"></div> <!-- Overlay element -->
+</div>
+
+<script>
+  function expandImage(element) {
+    element.classList.toggle('expanded');
+    document.querySelector('.overlay').classList.toggle('expanded');
+  }
+
+  function closeExpandedImage() {
+    var image = document.querySelector('.image');
+    image.classList.remove('expanded');
+    document.querySelector('.overlay').classList.remove('expanded');
+  }
+</script>
+
+</body>
+</html>
+```
+
+<!-- ![Harvest & Sunset Road Instersection, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/harvest%20road-min.png?raw=true) -->
 
@@
 
-![Kolin Road, Moccasin, MT (2023)](https://github.com/jstrand894/Website/blob/main/kolin%20road-min.png?raw=true)
+```{=html}
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Centered Image Expand</title>
+  <style>
+    body, html {
+      margin: 0;
+      padding: 0;
+      height: 100%;
+    }
+    .container {
+      display: flex;
+      justify-content: left;
+      align-items: left;
+      height: auto;
+    }
+    .image {
+      max-width: 50%;
+      max-height: 80vh;
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .image:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .image.expanded {
+      transform: scale(3); /* Scale up when expanded */
+      z-index: 9999; /* Ensure expanded image is on top */
+      position: fixed;
+      top: 50%;
+      left: 50%;
+      max-height: 80%;
+      max-width: 80%;
+      transform: translate(-50%, -50%); /* Center expanded image */
+    }
+    .overlay {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.5); /* Semi-transparent dark background */
+      z-index: 9998; /* Ensure overlay is behind the expanded image */
+      display: none; /* Initially hidden */
+    }
+    .image.expanded + .overlay {
+      display: block; /* Show overlay when image is expanded */
+    }
+  </style>
+</head>
+<body>
+
+<div class="container">
+  <img class="image" src="img/field-sites/kolin.png" alt="Image" onclick="expandImage(this)">
+  <div class="overlay" onclick="closeExpandedImage()"></div> <!-- Overlay element -->
+</div>
+
+<script>
+  function expandImage(element) {
+    element.classList.toggle('expanded');
+    document.querySelector('.overlay').classList.toggle('expanded');
+  }
+
+  function closeExpandedImage() {
+    var image = document.querySelector('.image');
+    image.classList.remove('expanded');
+    document.querySelector('.overlay').classList.remove('expanded');
+  }
+</script>
+
+</body>
+</html>
+```
+
+<!-- ![Kolin Road, Moccasin, MT (2023)](https://github.com/jstrand894/Website/blob/main/kolin%20road-min.png?raw=true) -->
 
@@
   mutate_at(vars(location, distance, collection, year, type), as.factor) %>%
+  mutate(mean = mean*100) %>%
   pivot_wider(names_from = type,

### 2024-03-05 index template.qmd
Update website 4/5/24

---
website:
  template: jolla
  id: about-block
  image: "img/aboutme/crescent.HEIC"
  links:
    - icon: twitter
      text: Twitter
      href: https://twitter.com/MarvinSchmittML
    - icon: github
      text: Github
      href: https://github.com/marvinschmitt
    - icon: linkedin
      text: LinkedIn
      href: https://www.linkedin.com/in/marvin-schmitt-a85b321a2/
    - icon: envelope
      text: Email
      href: "mailto:mail.marvinschmitt@gmail.com"  
---

::: {#about-block}
:::

# Welcome to my website!

I am Gandalf the White, a renowned wizard and scholar of Middle-earth. Over the centuries, I have dedicated my life to studying the history, languages, and cultures of the peoples of Middle-earth. On this website, I offer a glimpse into my vast body of knowledge, ongoing research, and publications.

My research interests span the entirety of Middle-earth, from the distant past to the present day. I am particularly passionate about the languages and mythologies of the Elves, Dwarves, and Men, and I have made significant contributions to the understanding of these subjects. My studies of the natural world have also been extensive, and I am an expert in the flora and fauna of Middle-earth. Additionally, I have been involved in many important historical events and offer valuable first-hand accounts and insights on the past.

Not only am I a scholar, but I am also a mentor and teacher. Over the centuries, I have had the honor of working with many young wizards, passing on my knowledge and experience to the next generation. I have a particular interest in the education of hobbits, whom I see as having great potential despite their small size and simple lifestyle. My teaching style is engaging and accessible, and I encourage my students to think critically and creatively.

On this website, you will find a selection of my published works, including scholarly articles, monographs, and translations. You will also find news and updates on my current research projects, as well as information on upcoming events and conferences where I will be speaking. I invite you to explore my website and learn more about my life's work as one of Middle-earth's greatest scholars and thinkers.
Please feel free to [contact me](mailto:mail.marvinschmitt@gmail.com) if you have any questions or would like to discuss potential projects.

```{=html}
<div  style="margin: 30px; text-align: center;">
<a class="btn btn-primary" href="https://www.marvinschmitt.com/blog/website-tutorial-quarto/" role="button" target="_blank" style="padding: 15px 30px;">View the tutorial for this template (+ download link)</a>
</div>
```

### 2024-03-05 index.qmd
Update website 4/5/24

@@
 ---
-title: "Jackson Strand"
 about: 
-  id: me
-  template: marquee
+  template: jolla
+  id: about-block
+  image: "img/aboutme/crescent.HEIC"
   links:
@@
       href: https://github.com/jstrand894 
-  image: img/aboutme/deadhorse.jpeg
+---
+::: {#about-block}
+:::
 
-sidebar: false
----
+# Welcome to my website!
 
-:::{#me}
-
-I am currently a Master's graduate student at Montana State University studying entomology. I live in Bozeman, MT with my wife Grace.
+My name is Jackson Strand. I am currently a Master's graduate student at Montana State University studying entomology. I live in Bozeman, MT with my wife Grace.
 
@@
 ![Me and Grace at Glacier National Park, 2023](img/aboutme/iceberg.JPG)
-
-:::

### 2024-03-05 locoweed-report.qmd
Update website 4/5/24

@@
     toc: true
+    code_folding: hide
 bibliography: citations/citations.bib
@@
 library(MVN) # for normality test
+```
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Image Cropping Example</title>
+<style>
+    }
+    .image-container {
+        width: auto; /* Set the width of the container */
+        height: auto; /* Set the height of the container */
+        overflow: hidden; /* Hide overflow */
+    }
+    .image-container img {
+        width: 100%; /* Make the image fill the container horizontally */
+        height: auto; /* Maintain aspect ratio */
+        object-fit: cover;
+        object-position: center;
+    }
+</style>
+</head>
+<body>
+
+<div class="image-container">
+    <img src="img/loco-bee-crop.JPG" alt="Your Image">
+</div>
+
+</body>
+</html>
 
@@
 
+```{r plant pictures, echo = F, eval = F}
+
+# Set the directory path
+directory_path <- "img/loco-plants"
+
+# List all the files in the directory
+file_names <- list.files(directory_path)
+
+# Initialize an empty character vector to store the HTML code
+html_code <- "<div class=\"gallery\">\n"
+
+# Loop through each file name and generate HTML code
+for (i in seq_along(file_names)) {
+  # Generate the HTML code for each image
+  image_html <- paste0('<img src="', directory_path, "/", file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')
+
+  # Append the generated HTML code to the existing code
+  html_code <- paste(html_code, image_html, sep = "")
+}
+
+# Close the gallery div
+html_code <- paste(html_code, "</div>\n<div class=\"expanded-container\" onclick=\"closeExpanded()\">\n  <img class=\"expanded\" id=\"expanded-image\" onclick=\"navigateGallery(event)\">\n</div>", sep = "")
+
+# Print the final HTML code
+cat(html_code)
+
+```
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
+      grid-gap: 10px;
+    }
+    .gallery img {
+      width: 100%;
+      height: 100%;
+      object-fit: cover; /* Ensure images cover the entire space */
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .gallery img:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
+      display: none; /* Initially hidden */
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+    p.gallery-caption {
+      text-align: center;
+      font-style: italic;
+      color: #8B8989;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+<img src="img/loco-plants/IMG_0064.HEIC" alt="Image 1" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_0820 2.HEIC" alt="Image 2" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_1165.HEIC" alt="Image 3" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_1177.HEIC" alt="Image 4" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_1923.HEIC" alt="Image 5" onclick="expandImage(this)">
+</div>
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<p class="gallery-caption">Greenhouse Volatile Collection Plants looked like this.</p>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    // Calculate the click position relative to the expanded image
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
+    var imageWidth = event.target.offsetWidth;
+
+    // Determine if the click was on the left or right side of the image
+    if (clickX < imageWidth / 2) {
+      // Clicked on the left side: navigate to the previous image
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      // Clicked on the right side: navigate to the next image
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    // Update the expanded image source
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+</script>
+
+</body>
+</html>
+
+```
+
+<br>
 Volatiles were collected from twleve plants per day using glass volatile collection chambers (VCCs) that were 30 cm long and 4 cm in diameter.

### 2024-03-05 msu-masters.qmd
Update website 4/5/24

@@
 ---
-<figure>
-  <img src="img/msu-masters/cages.heic" alt="Montana State University" 
-  width = 250" height = "200">
-</figure>
 
-<figure>
-  <img src="img/msu-masters/vcs.heic" alt="Montana State University" 
-  width = 200" height = "300">
-</figure>
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Box with Rounded Edges</title>
+<style>
+  .container {
+    width: 400px;
+    height: auto; /* Set container height */
+    border: 0.5px solid #ccc; /* Light grey color */
+    border-radius: 8px; /* Less rounded edges */
+    overflow: hidden; /* Hide overflow content */
+    position: relative;
+    display: inline-block;
+    transition: box-shadow 0.15s ease; /* Transition for box-shadow */
+    box-shadow: 0 2px 5px rgba(0, 0, 0, 0.1); /* Slight shadow by default */
+    text-align: center; /* Center content horizontally */
+    text-decoration: none; /* Remove underline */
+    margin-right: 20px; /* Add margin to create space between boxes */
+  }
 
-<figure>
-  <img src="img/msu-masters/field.heic" alt="Montana State University" 
-  width = 250" height = "200">
-</figure>
+  .container:last-child {
+    margin-right: 0; /* Remove margin from the last box */
+  }
+
+  .container:hover {
+    box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */
+  }
+
+  .content {
+    width: 100%;
+    height: 100%;
+    padding: 20px;
+    box-sizing: border-box;
+  }
+
+  .content img {
+    width: 50%;
+    height: auto;
+    display: block;
+    margin: 0 auto; /* Center the image horizontally */
+    object-fit: cover; /* Crop the image to cover container */
+  }
+
+  .text {
+    margin-top: 10px;
+    text-align: center; /* Center text within the box */
+    text-decoration: none;
+  }
+
+  .text h1 {
+    font-size: 24px; /* Larger text size */
+    margin-top: 1px;
+    margin-bottom: 5px; /* Adjust margin between large and small text */
+    color: black;
+  }
+  .text h1:hover {
+    color: #2068DC; /* Change text color to blue on hover */
+  }
+
+  .text p {
+    font-size: 14px;
+    margin-top: 5px;
+    color: #666;
+  }
+</style>
+</head>
+<body>
+  <a href="wss.qmd" class="container">
+    <div class="content">
+      <img src="img/msu-masters/wss.JPG" alt="Image"> <!-- Center the image within the box -->
+      <div class="text">
+        <h1>Wheat Stem Sawfly</h1>
+        <p>Understanding the importance of smooth brome in promoting a Braconid parasitoid that targets the Wheat Stem Sawfly.</p>
+      </div>
+    </div>
+  </a>
+  <a href="https://www.example.com" class="container">
+    <div class="content">
+      <img src="img/msu-masters/loco-bee2.jpg" alt="Image"> <!-- Center the image within the box -->
+      <div class="text">
+        <h1>Locoweed Volatiles</h1>
+        <p>Comparison of volatile organic compounds profiles of <em>Oxytropis</em> plants that contained the commensal endophyte <em>Alternaria oxytropis</em> and those that did not.</p>
+      </div>
+    </div>
+  </a>
+</body>
+</html>
+
+```

### 2024-03-05 pfreport.qmd
Update website 4/5/24

@@
   chunk_output_type: console
-
-website:
-  sidebar:
-    style: "docked"
-    contents:
-      - section: "Test1"
-        contents: index.qmd
-      - section: "Test2"
-        contents: index.qmd
 ---
@@
 
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Image Cropping Example</title>
+<style>
+    }
+    .image-container {
+        width: auto; /* Set the width of the container */
+        height: auto; /* Set the height of the container */
+        overflow: hidden; /* Hide overflow */
+    }
+    .image-container img {
+        width: 100%; /* Make the image fill the container horizontally */
+        height: auto; /* Maintain aspect ratio */
+        object-fit: cover;
+        object-position: center;
+    }
+</style>
+</head>
+<body>
+
+<div class="image-container">
+    <img src="img/msu-masters/cages-wide.jpg" alt="Your Image">
+</div>
+
+</body>
+</html>
+
+```
 # POST FARM

### 2024-03-05 projects.qmd
Update website 4/5/24

@@
 ---
-::: {style="display: flex;"}
 
-<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
-  <a href="msu-masters.qmd">
-    <img src="img/msu-logo.jpg" alt="Montana State University" width="200" height="200">
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Box with Rounded Edges</title>
+<style>
+  .container {
+    width: 300px;
+    height: 300px; /* Set container height */
+    border: 0.5px solid #ccc; /* Light grey color */
+    border-radius: 8px; /* Less rounded edges */
+    overflow: hidden; /* Hide overflow content */
+    position: relative;
+    display: inline-block;
+    transition: box-shadow 0.15s ease; /* Transition for box-shadow */
+    box-shadow: 0 2px 5px rgba(0, 0, 0, 0.1); /* Slight shadow by default */
+    text-align: center; /* Center content horizontally */
+    text-decoration: none; /* Remove underline */
+    margin-right: 20px; /* Add margin to create space between boxes */
+    margin-bottom: 20px;
+  }
+
+  .container:last-child {
+    margin-right: 0; /* Remove margin from the last box */
+  }
+
+  .container:hover {
+    box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */
+  }
+
+  .content {
+    width: 100%;
+    height: 100%;
+    padding: 20px;
+    box-sizing: border-box;
+  }
+
+  .content img {
+    width: 70%;
+    height: auto;
+    display: block;
+    margin: 0 auto; /* Center the image horizontally */
+    object-fit: cover; /* Crop the image to cover container */
+  }
+
+  .text {
+    margin-top: 10px;
+    text-align: center; /* Center text within the box */
+    text-decoration: none;
+  }
+
+  .text h1 {
+    font-size: 24px; /* Larger text size */
+    margin-top: 1px;
+    margin-bottom: 5px; /* Adjust margin between large and small text */
+    color: black;
+  }
+  .text h1:hover {
+    color: #2068DC; /* Change text color to blue on hover */
+  }
+
+  .text p {
+    font-size: 14px;
+    margin-top: 5px;
+    color: #666;
+  }
+  <!--  .nba-container { -->
+  <!--   background-color: #0E1D2D; -->
+  <!-- } -->
+</style>
+</head>
+<body>
+  <a href="msu-masters.qmd" class="container">
+    <div class="content">
+      <img src="img/projects-logos/msu-logo.jpg" alt="Image"> <!-- Center the image within the box -->
+      <div class="text">
+        <h1>MSU Master's - Entomology</h1>
+        <p>Co-Advised by David Weaver & Tracy Sterling</p>
+      </div>
+    </div>
   </a>
-  <figcaption style="text-align: center;">Montana State University</figcaption>
-</figure>
+  <a href="bballref.qmd" class="container nba-container">
+    <div class="content">
+      <img src="img/projects-logos/bballref-logo-circle.jpg" alt="Image"> <!-- Center the image within the box -->
+      <div class="text">
+        <h1><br>NBA Roster Analysis Tool</h1>
+        <p></p>
+      </div>
+    </div>
+  </a>
+</body>
+</html>
+```
 
-<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
-  <a href="bballref.qmd">
-    <img src="img/bballref-logo.jpeg" alt="Basketball Reference Tool" width="200" height="200">
-  </a>
-  <figcaption style="text-align: center;">Basketball Reference Tool</figcaption>
-</figure>
+<!-- ```{=html} -->
+<!-- <figure style="display: inline-block; overflow: hidden; position: relative; text-align: center;"> -->
+<!--     <a href="bballref.qmd" style="display: block;"> -->
+<!--         <img src="img/projects-logos/bballref-logo.jpeg" alt="NBA Roster Analysis Tool" width="200" style="transition: transform 0.3s ease-in-out;"> -->
+<!--     </a> -->
+<!--     <figcaption style="margin-top: 10px;">NBA Roster Analysis Tool</figcaption> -->
+<!-- </figure> -->
 
-:::
+<!-- <style> -->
+<!--     figure:hover img { -->
+<!--         transform: scale(1.05); -->
+<!--     } -->
+<!-- </style> -->
+
+<!-- ``` -->

### 2024-03-05 running.qmd
Update website 4/5/24

@@
   <a href = "tmr50k.qmd">
-    <img src="img/tmr50k/tmr50k-top.JPG" alt="Deadhorse Ultra" 
+    <img src="img/tmr50k/tmr50k-top.JPG" alt="Teton Mountain Runs 50K" 
   width = 300" height = "200">
   </a>
-  <figcaption style="text-align:center;">Deadhorse Ultra, 2023</figcaption>
+  <figcaption style="text-align:center;">Teton Mountain Runs, 2023</figcaption>
 </figure>

### 2024-03-05 save.qmd
Update website 4/5/24

---
title: "save"
---

# box with shadow when you hover

```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Box with Rounded Edges</title>
<style>
  .container {
    width: 300px;
    height: 250px; /* Set container height */
    border: 0.5px solid #ccc; /* Light grey color */
    border-radius: 8px; /* Less rounded edges */
    overflow: hidden; /* Hide overflow content */
    position: relative;
    display: inline-block;
    transition: box-shadow 0.3s ease; /* Transition for box-shadow */
    box-shadow: 0 2px 5px rgba(0, 0, 0, 0.1); /* Slight shadow by default */
    text-align: center; /* Center content horizontally */
    text-decoration: none; /* Remove underline */
  }

  .container:hover {
    box-shadow: 0 5px 10px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */
  }

  .content {
    width: 100%;
    height: 100%;
    padding: 10px;
    box-sizing: border-box;
  }

  .content img {
    width: 100%;
    height: auto;
    display: block;
    margin: 0 auto; /* Center the image horizontally */
    object-fit: cover; /* Crop the image to cover container */
  }

  .text {
    margin-top: 10px;
    text-align: center; /* Center text within the box */
    text-decoration: none;
  }

  .text h1 {
    font-size: 24px; /* Larger text size */
    margin-top: 1px;
    margin-bottom: 5px; /* Adjust margin between large and small text */
    color: black;
  }
  .text h1:hover {
    color: #2068DC; /* Change text color to blue on hover */
  }
</style>
</head>
<body>
  <a href="https://www.example.com" target="_blank" class="container">
    <div class="content">
      <img src="img/msu-masters/wss.JPG" alt="Image"> <!-- Center the image within the box -->
      <div class="text">
        <h1>Wheat Stem Sawfly</h1>
        <!-- <h2></h2> -->
      </div>
    </div>
  </a>
</body>
</html>

```

### 2024-03-05 sb-ww.qmd
Update website 4/5/24

--- 2024-02-29 bugs.qmd
@@
 ---
-title: "bug pictures"
-filters:
-  - lightbox
-lightbox: auto
+title: "smooth brome/winter wheat"
 ---
 
-![Bug pic 1](img/bugs/IMG_1664.JPG){group="my-gallery"}
-![Bug pic 2](img/bugs/IMG_1678.JPG){group="my-gallery"}
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
+      grid-gap: 10px;
+    }
+    .gallery img {
+      width: 100%;
+      height: 100%;
+      object-fit: cover; /* Ensure images cover the entire space */
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .gallery img:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
+      display: none; /* Initially hidden */
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
 
-<figure>
-  <img src="img/bugs/IMG_1664.JPG" alt="Montana State University" 
-  width = 200" height = "200">
-</figure>
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+<img src="img/msu-masters/cages.HEIC" alt="Image 1" onclick="expandImage(this)">
+<img src="img/msu-masters/field.HEIC" alt="Image 2" onclick="expandImage(this)">
+<img src="img/msu-masters/vcs.HEIC" alt="Image 3" onclick="expandImage(this)">
+</div>
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    // Calculate the click position relative to the expanded image
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
+    var imageWidth = event.target.offsetWidth;
+
+    // Determine if the click was on the left or right side of the image
+    if (clickX < imageWidth / 2) {
+      // Clicked on the left side: navigate to the previous image
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      // Clicked on the right side: navigate to the next image
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    // Update the expanded image source
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+</script>
+
+</body>
+</html>
+```
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
+      grid-gap: 10px;
+    }
+    .gallery img {
+      width: 100%;
+      height: 100%;
+      object-fit: cover; /* Ensure images cover the entire space */
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .gallery img:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
+      display: none; /* Initially hidden */
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
+
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+<img src="img/msu-masters/cages.HEIC" alt="Image 1" onclick="expandImage(this)">
+<img src="img/msu-masters/field.HEIC" alt="Image 2" onclick="expandImage(this)">
+<img src="img/msu-masters/vcs.HEIC" alt="Image 3" onclick="expandImage(this)">
+</div>
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    // Calculate the click position relative to the expanded image
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
+    var imageWidth = event.target.offsetWidth;
+
+    // Determine if the click was on the left or right side of the image
+    if (clickX < imageWidth / 2) {
+      // Clicked on the left side: navigate to the previous image
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      // Clicked on the right side: navigate to the next image
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    // Update the expanded image source
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+</script>
+
+</body>
+</html>
+```

### 2024-03-05 tmr50k.qmd
Update website 4/5/24

@@
 
-::: {style="display: block;"}
+```{r eval = F, echo = F}
+# Set the directory path
+directory_path <- "img/tmr50k"
 
-```{r get images, echo = F, results = 'asis'}
-image_files <- list.files("img/tmr50k", pattern = ".JPG|.png|.gif|.jpg|.jpeg", full.names = TRUE)
+# List all the files in the directory
+file_names <- list.files(directory_path)
 
-for (file in image_files) {
-  cat("![](",file, ")\n  \n")
+# Exclude files with .zip extension
+file_names <- file_names[!grepl("\\.zip$", file_names)]
+
+# Initialize an empty character vector to store the HTML code
+html_code <- "<div class=\"gallery\">\n"
+
+# Loop through each file name and generate HTML code
+for (i in seq_along(file_names)) {
+  # Generate the HTML code for each image
+  image_html <- paste0('<img src="', directory_path, "/", file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')
+
+  # Append the generated HTML code to the existing code
+  html_code <- paste(html_code, image_html, sep = "")
 }
+
+# Close the gallery div
+html_code <- paste(html_code, "</div>\n<div class=\"expanded-container\" onclick=\"closeExpanded()\">\n  <img class=\"expanded\" id=\"expanded-image\" onclick=\"navigateGallery(event)\">\n</div>", sep = "")
+
+# Print the final HTML code
+cat(html_code)
+
 ```
 
-:::
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
+      grid-gap: 10px;
+    }
+    .gallery img {
+      width: 150px;
+      height: 150px;
+      object-fit: cover; /* Ensure images cover the entire space */
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .gallery img:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
+      display: none; /* Initially hidden */
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
+
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+<img src="img/tmr50k/DSC01450.JPG" alt="Image 1" onclick="expandImage(this)">
+<img src="img/tmr50k/DSC01715.JPG" alt="Image 2" onclick="expandImage(this)">
+<img src="img/tmr50k/DSC02442.JPG" alt="Image 3" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_0054.JPG" alt="Image 4" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_0088.JPG" alt="Image 5" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_0094.JPG" alt="Image 6" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_0108.JPG" alt="Image 7" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_0120.JPG" alt="Image 8" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_0126.JPG" alt="Image 9" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_5954.JPG" alt="Image 10" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_5983.JPG" alt="Image 11" onclick="expandImage(this)">
+<img src="img/tmr50k/IMG_5999.JPG" alt="Image 12" onclick="expandImage(this)">
+<img src="img/tmr50k/tmr50k-top.JPG" alt="Image 13" onclick="expandImage(this)">
+</div>
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    // Calculate the click position relative to the expanded image
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
+    var imageWidth = event.target.offsetWidth;
+
+    // Determine if the click was on the left or right side of the image
+    if (clickX < imageWidth / 2) {
+      // Clicked on the left side: navigate to the previous image
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      // Clicked on the right side: navigate to the next image
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    // Update the expanded image source
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+
+  // Add event listeners to the left and right sides of the expanded image for navigation
+  document.querySelector('.expanded').addEventListener('click', function(event) {
+    navigateGallery(event);
+  });
+</script>
+
+</body>
+</html>
+
+```
+<br>
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Button</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+</style>
+</head>
+<body>
+
+<!-- Button to download the file -->
+<a href="img/tmr50k/tmr2023.zip" download="TetonMountainRuns2023">
+  <button class="button">Download All Pictures</button>
+</a>
+
+</body>
+</html>
+
+```

### 2024-03-05 weatherdata.qmd
Update website 4/5/24

---
title: "Weather Data MT"
author: "Jackson Strand"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(glue)
library(lubridate)
library(ggplot2)
```

```{r}
inventory_url <- "https://www.ncei.noaa.gov/pub/data/ghcn/daily/ghcnd-inventory.txt"

inventory <- read_table(inventory_url,
           col_names = c("station", "lat", "lon", "variable", "start", "end"))

inventory

my_lat <- 45.661966103346735 * 2 * pi/360
my_long <- -111.04939878991246 * 2 * pi/360

my_station <- 
  inventory %>%
    mutate(lat_r = lat *2 *pi/360,
           lon_r = lon *2 *pi/360,
           d = 1.609344 * 3963 * acos((sin(lat_r) * sin(my_lat)) + cos(lat_r) * cos(my_lat) * cos(my_long - lon_r))) %>%
    filter(start < 1960 & end > 2020) %>%
    top_n(n = -1, d) %>%
    distinct(station) %>%
    pull(station)

station_daily <- glue("https://ncei.noaa.gov/pub/data/ghcn/daily/by_station/{my_station}.csv.gz")

local_weather <- 
read_csv(station_daily,
         col_names = c("station", "date", "variable", "value", "a", "b", "c", "d")) %>%
  dplyr::select(station, date, variable, value) %>%
  pivot_wider(names_from = "variable", values_from = "value",
                values_fill = 0) %>%
  dplyr::select(date, TMAX, PRCP, SNOW) %>%
  mutate(date = ymd(date))

```

```{r}
bozeman_weather <-
  local_weather %>%
  filter(date > "2000-01-01")

ggplot(data = bozeman_weather) +
  geom_line(aes(x = date, y = TMAX)) +
  theme_bw()
```
title: "Montana Weather Data"
---

### 2024-03-05 wss.qmd
Update website 4/5/24

---
title: "MSU Master's: Wheat Stem Sawfly"
---

```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Box with Rounded Edges</title>
<style>
  .container {
    width: calc(33.33% - 25px);
    height: 300px; /* Set container height */
    border: 0.5px solid #ccc; /* Light grey color */
    border-radius: 8px; /* Less rounded edges */
    overflow: hidden; /* Hide overflow content */
    position: relative;
    display: inline-block;
    transition: box-shadow 0.15s ease; /* Transition for box-shadow */
    box-shadow: 0 2px 5px rgba(0, 0, 0, 0.1); /* Slight shadow by default */
    text-align: center; /* Center content horizontally */
    text-decoration: none; /* Remove underline */
    margin-right: 20px; /* Add margin to create space between boxes */
    margin-bottom: 20px;
  }

  .container:last-child {
    margin-right: 0; /* Remove margin from the last box */
  }

  .container:hover {
    box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */
  }

  .content {
    width: 100%;
    height: 100%;
    padding: 20px;
    box-sizing: border-box;
  }

  .content img {
    width: 50%;
    height: auto;
    display: block;
    margin: 0 auto; /* Center the image horizontally */
    object-fit: cover; /* Crop the image to cover container */
  }

  .text {
    margin-top: 10px;
    text-align: center; /* Center text within the box */
    text-decoration: none;
  }

  .text h1 {
    font-size: 24px; /* Larger text size */
    margin-top: 1px;
    margin-bottom: 5px; /* Adjust margin between large and small text */
    color: black;
  }
  .text h1:hover {
    color: #2068DC; /* Change text color to blue on hover */
  }

  .text p {
    font-size: 14px;
    margin-top: 5px;
    color: #666;
  }
</style>
</head>
<body>
  <a href="pfreport.qmd" class="container">
    <div class="content">
      <img src="img/msu-masters/cages2.jpg" alt="Image"> <!-- Center the image within the box -->
      <div class="text">
        <h1>Controlled WSS Infestation</h1>
        <p>Assessing WSS infestation and cutting within smooth brome in controlled research plot at Arthur H. Post Agronomy Farm.</p>
      </div>
    </div>
  </a>
  <a href="field_report.qmd" class="container">
    <div class="content">
      <img src="img/msu-masters/oviposition.JPG" alt="Image"> <!-- Center the image within the box -->
      <div class="text">
        <h1>Montana Field Sites</h1>
        <p>Collection of wheat and smooth brome in Big Sandy and Moccasin, Montana to assess parasitism and WSS infestation.</p>
      </div>
    </div>
  </a>
  <a href="eag.qmd" class="container">
    <div class="content">
      <img src="img/msu-masters/vcs.jpg" alt="Image"> <!-- Center the image within the box -->
      <div class="text">
        <h1>Organic Volatile Collection</h1>
        <p>Collection of wheat and smooth brome in <strong>Big Sandy</strong> and <strong>Moccasin</strong> Montana to assess parasitism and WSS infestation.</p>
      </div>
    </div>
  </a>
  <a href="https://www.example.com" class="container">
    <div class="content">
      <img src="img/msu-masters/clouds.jpg" alt="Image"> <!-- Center the image within the box -->
      <div class="text">
        <h1>Weather Data</h1>
        <p>Weather data trends from Big Sandy and Moccasin, Montana.</p>
      </div>
    </div>
  </a>
</body>
</html>

```

### 2024-03-06 eag.qmd
commit 3/6/24

@@
             color = "red") +
-  geom_line(data = eag_plot, aes(x = scale_values(row_num), 
+  geom_line(data = eag_plot, aes(x = scale_values(row_num)+1.2, 
                                  y = EAD * 100), 
@@
        color = "Legend") +
-  theme_bw() + 
-  scale_x_continuous(breaks = seq(min(fid_plot$MeasRetTime,
-                                      max(fid_plot$MeasRetTime), 
-                                      by = 1)))
+  theme_bw() 
 )

### 2024-03-06 field_report.qmd
commit 3/6/24

@@
 output: 
-  quarto::quarto_html:
-    layout: sidebar
+  html:
+    # layout: sidebar
     theme: cosmo
@@
     toc_title: "Table of Contents"
-    toc_collapsed: false
+    toc_expand: true
     toc_depth: 4
@@
   chunk_output_type: console
+toc: true
+toc_title: "Table of Contents"
+# toc_expand: true
+# toc_float: true
 ---
+<!-- field site cover image -->
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Image Cropping Example</title>
+<style>
+    }
+    .image-container {
+        width: auto; /* Set the width of the container */
+        height: auto; /* Set the height of the container */
+        overflow: hidden; /* Hide overflow */
+    }
+    .image-container img {
+        width: 100%; /* Make the image fill the container horizontally */
+        height: auto; /* Maintain aspect ratio */
+        object-fit: cover;
+        object-position: center;
+    }
+</style>
+</head>
+<body>
+
+<div class="image-container">
+    <img src="img/field-sites/big-sandy-wide.jpg" alt="Your Image">
+</div>
+
+</body>
+</html>
+
+```
+
 ```{r library, echo = F, message = F, warning = F}
@@
 ```
-
-# POST FARM
-
-## Methods
-
-### Plant Selection and Cages
-
-Assessment of C.
-cinctus infestation and mortality within B.
-inermis were assessed using a 34 x 60 ft plot at the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-The cage structure was built using 1-inch PVC piping with the netting made using 530\(\mu\) Amber Lumite Screen (BioQuip® Products, LLC).
-Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.
-The cages were removed from the plot after 3 weeks of sawfly release.
-
-### Insects
-
-Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of C.
-cinctus infestation and cutting in the prior year.
-Cut stubble, which contained C.
-cinctus larvae in diapause, were kept refrigerated between -2°C and 3°C for >100 days as required to complete obligatory larval diapause.
-As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA).
-Stubs were added to cages once B.
-inermis stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).
-
-#### Data Collection
-
-Stems were collected from each cage site in late August.
-All stems from each 6 x 3 ft cage section were collected.
-Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.
-
-## Analysis
-
-### Overall WSS Infestation and Cutting
-
-```{r (pf.RAW) post farm n.inf data, echo = F, message = F}
-pf.raw.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2022/Post\ Farm\ 2022.xlsx", sheet = "n.inf")
-pf.raw.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "n.inf")
-```
-```{r (pf.inf.23) post farm inf data, echo = F}
-pf.inf.23 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf") %>%
-  rename(n.1 = 8, n.2 = 9, n.3 = 10, n.4 = 11,
-         n.5 = 12, n.6 = 13, n.7 = 14, n.8 = 15,
-         n.9 = 16, n.10 = 17)
-```
-```{r (pf.comb) post farm data combined, echo = F}
-# 2022
-pf.2022.site <-
-pf.raw.2022 %>%
-  group_by(treatment, site) %>%
-  reframe(site, treatment,
-            p.inf = sum(iv, ir, ix)/sum(uv, ur, iv, ir, ix),
-            cut = sum(cut)/sum(total),
-            total.stems = sum(iv, ir, ix, uv, ur)) %>%
-  unique() %>%
-  pivot_longer(-c(treatment, total.stems, site),
-               names_to = "type",
-               values_to = "percent") %>%
-  mutate(year = "2022")
-
-# 2023
-pf.2023.site <-
-pf.raw.2023 %>%
-  group_by(treatment, site) %>%
-  reframe(treatment, site, 
-          t.inf = sum(n.iv, n.ir, n.ix),
-          p.inf = sum(n.iv, n.ir, n.ix)/sum(n.uv, n.ur, 
-                                            n.iv, n.ir, n.ix),
-          cut = sum(cut)/sum(total),
-          total.stems = sum(n.iv, n.ir, n.ix, n.uv, n.ur)) %>%
-  distinct() %>%
-  pivot_longer(-c(treatment, total.stems, site),
-               names_to = "type",
-               values_to = "percent") %>%
-  filter(!type == "t.inf") %>%
-  arrange(type) %>%
-  mutate(year = "2023")
-
-pf.comb <-
-  pf.2022.site %>%
-  rbind(pf.2023.site) %>%
-  group_by(treatment, year, type) %>%
-  mutate(count = n(),
-         average = mean(percent),
-         sd = sd(percent),
-         se = (sd/sqrt(count))) %>%
-  reframe(treatment, year, type,
-          mean = mean(percent),
-          sd = mean(sd),
-          se = mean(se)) %>%
-  unique()
-
-```
-```{r (pf.total) pf total stems, echo = F}
-# 2022 totals
-pf.2022.total <-
-  pf.raw.2022 %>%
-  group_by(treatment) %>%
-  reframe(total = sum(uv, ur, iv, ir, ix)) %>%
-  mutate(year = "2022")
-
-# 2023 totals
-pf.2023.total <-
-  pf.raw.2023 %>%
-  group_by(treatment) %>%
-  reframe(total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
-  mutate(year = "2023")
-
-pf.total.comb <-
-  pf.2022.total %>%
-  rbind(pf.2023.total)
-
-```
-
-#### Basic Table and Plot of Infestation
-
-```{r post farm table, echo = F, results = 'asis'}
-pf.23.table <-
-pf.comb %>%
-  filter(year == "2023") %>%
-  dplyr::select(treatment, type, mean) %>%
-  mutate(mean = round_based_on_size(mean)) %>%
-  pivot_wider(values_from = mean,
-              names_from = type,
-              values_fn = list(mean = mean)) %>%
-  left_join(pf.2023.total, by = "treatment")
-
-pf.22.table <-
-pf.comb %>%
-  filter(year == "2022") %>%
-  dplyr::select(treatment, type, mean) %>%
-  mutate(mean = round_based_on_size(mean)) %>%
-  pivot_wider(values_from = mean,
-              names_from = type,
-              values_fn = list(mean = mean))  %>%
-  left_join(pf.2022.total, by = "treatment")
-
-pf.23.table.final <-
-kable(pf.23.table[c("treatment", "total", "p.inf", "cut")], 
-      caption = "2023 Post Farm Infestation", 
-      format = "html",
-      col.names = c("Treatment", "Total Stems", "% Inf", "% Cut"),
-      align = "lccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2")
-
-pf.22.table.final <-
-kable(pf.22.table[c("treatment", "total", "p.inf", "cut")], 
-      caption = "2022 Post Farm Infestation", 
-      format = "html",
-      col.names = c("Treatment", "Total Stems", "% Inf", "% Cut"),
-      align = "lcc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2")
-
-cat('<div style="display: flex;">')
-cat('<div style="flex: 1;">')
-cat(pf.23.table.final)
-cat('</div>')
-cat('<div style="flex: 1;">')
-cat(pf.22.table.final)
-cat('</div>')
-cat('</div>')
-```
-
-```{r post farm plot, echo = F, fig.height = 2.5, fig.align = 'center'}
-pf.comb <-
-pf.comb %>%
-  left_join(pf.total.comb, by = c("treatment", "year"))
-
-ggplot() +
-  facet_wrap(~treatment) +
-  geom_bar(stat = "identity",
-           position = position_dodge(),
-           data = pf.comb %>%
-             filter(type == "p.inf"),
-           aes(year, mean, fill = "p.inf")) +
-  geom_bar(stat = "identity",
-           position = position_dodge(),
-           data = pf.comb %>%
-             filter(type == "cut"),
-           aes(year, mean, fill = "cut")) +
-  geom_errorbar(data = pf.comb,
-                aes(x = year,
-                    ymin = mean - se,
-                    ymax = mean + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  theme_bw() +
-  labs(x = element_blank(),
-       y = "Percentage of stems (%)",
-       fill = element_blank()) +
-  scale_fill_brewer(palette = "Greens",
-                    labels = c("Cut", "Infested"))
-
-```
-
-Let's take a closer look at the number of nodes borrowed through by the larvae.
-Using this information we can make an arguement about the size and length of time each larvae was alive prior to dying.
-This can be informative in understanding how many larvae would have been left later in the season for potential parasitism.
-
-#### Infestation Percentage/Node
-
-```{r v.r infestation node, echo = F, warning = F, message = F, fig.height = 3.5, fig.align = 'center'}
-v.r_inf <- 
-pf.inf.23 %>%
-  dplyr::select(site, rep, treatment, 8:19) %>%
-  mutate(stem.id = row_number()) %>%
-  pivot_longer(-c(stem.id, site, rep, treatment, v.r, lar),
-               names_to = "node",
-               values_to = "node_status") %>%
-  na.omit() %>%
-  group_by(stem.id, treatment, v.r, lar) %>%
-  reframe(nodes.inf = n()) %>%
-  group_by(treatment, v.r, lar, nodes.inf) %>%
-  reframe(count = n()) %>%
-  group_by(treatment, v.r) %>%
-  mutate(count2 = sum(count),
-         per = count/count2) %>%
-  reframe(treatment, v.r, nodes.inf, per) %>%
-  mutate_at(vars(nodes.inf), as.factor)
-
-# plot
-v.r_inf %>%
-  ggplot() +
-  geom_bar(stat = "identity",
-         position = position_dodge(),
-         aes(y = per, x = nodes.inf, fill = treatment)) +
-  geom_line(stat = "smooth",
-            aes(y = per, 
-                x = nodes.inf, 
-                group = treatment, 
-                color = treatment),
-            se = FALSE,
-            alpha = 0.6,
-            linewidth = 1.3,
-            linetype = "solid",
-            method = "loess") +
-  theme_bw() +
-  scale_fill_brewer(palette = "Greens") +
-  scale_color_brewer(palette = "Reds") +
-  facet_wrap(~v.r)
-
-# anova
-anova1 <- 
-  summary(aov(per ~ treatment + nodes.inf + v.r, data = v.r_inf))
-
-aov.df <- as.data.frame(anova1[[1]])
-aov.df$`Pr(>F)` <- sprintf("%.4f", aov.df$`Pr(>F)`)
-aov.df$Source <- rownames(aov.df)
-rownames(aov.df) <- NULL
-
-aov.df[,2:4] <- round(aov.df[,2:4], 3)
-
-# anova table
-kbl(aov.df %>%
-      dplyr::select(Source, everything()), 
-    align = "lccccc",
-    format = "html",
-    caption = "ANOVA Table",
-    col.names = c("", "Df", "Sum Sq", "Mean Sq", "F-Value", "Pr(>F)")) %>%
-  kable_styling(bootstrap_options = c("striped", "bordered", 
-                                      "hover", "repsonsive"),
-                full_width = FALSE) %>%
-  row_spec(0, background = "#f2f2f2")
-
-```
-
-Based on the results of our ANOVA, we have weak evidence (Pval = `r aov.df[3,5]`) to suggest there is a difference in the number of nodes infested when considering treatment group and vegetative and reproductive stems.
-
-```{r cumsum of infestation per node, echo = F, fig.align = "center"}
-# I want to see the percentage of stems with 6+ nodes infested, then 5:6+, then 4:6+, etc.
-cumsum <-
-pf.inf.23 %>%
-  dplyr::select(site, rep, treatment, 8:19) %>%
-  mutate(stem.id = row_number()) %>%
-  pivot_longer(-c(stem.id, site, rep, treatment, v.r, lar),
-               names_to = "node",
-               values_to = "node_status") %>%
-  na.omit() %>%
-  group_by(stem.id, treatment, v.r, lar) %>%
-  reframe(nodes.inf = n()) %>%
-  group_by(treatment, nodes.inf) %>%
-  reframe(count = n()) %>%
-  group_by(treatment) %>%
-  mutate(count2 = sum(count),
-         per = count/count2) %>%
-  reframe(treatment, nodes.inf, per) %>%
-  arrange(treatment, desc(nodes.inf)) %>%
-  group_by(treatment) %>%
-  mutate(cumsum = cumsum(per)) %>%
-  mutate_at(vars(nodes.inf), as.factor) %>%
-  dplyr::select(treatment, nodes.inf, cumsum) %>%
-  mutate(nodes.inf = fct_rev(nodes.inf))
-
-cumsum %>%
-  mutate(cumsum = round(cumsum, 2)) %>%
-  rename("Treatment" = treatment) %>%
-  pivot_wider(values_from = "cumsum",
-              names_from = "nodes.inf") %>%
-  kable("html",
-        caption = "Cumulative sum of nodes infested",
-        align = "lcccccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2") %>%
-  add_header_above(c(" ", "# of Nodes Infested" = 6))
-
-ggplot(data = cumsum,
-         aes(x = nodes.inf,
-             y = cumsum,
-             group = treatment, 
-             color = treatment)) +
-  geom_line() +
-  theme_bw() +
-  geom_bar(stat = "identity",
-           aes(x = nodes.inf, 
-               y = cumsum,
-               fill = treatment),
-           position = position_dodge(),
-           alpha = 0.25) +
-  theme_bw() +
-  labs(x = "Number of Nodes Infested",
-       y = "Cumulative sum % of stems",
-       fill = "WSS Treatment \nLevel",
-       color = "WSS Treatment \nLevel",
-       title = "Smooth brome node infestation") +
-  scale_x_discrete(limits = rev(levels(cumsum$cumsum)))
-
-```
-
-I like this plot because it shows the cumulative percentage sum of nodes infested at each node.
-For example, at the 3rd node in 2023, we see that close to 70% of the stems in the 'high' treatment group exhibited between 3 and 6 nodes infested.
-As a general thought exercise, the more nodes a larvae burrowed through, the larger the larvae.
-This means that if we classify a "large" larvae as one who burrowed through at least 4 nodes, our infestation data shows that 36% of the larvae in the 'high' treatment groups died as "large" larvae.
-
-```{r larvae breakdown, echo = F}
-# how many stems had live larvae in them?
-# percent of stems with each larval type
-
-larvae.type <-
-pf.inf.23 %>%
-  dplyr::select(site, rep, treatment, 8:19) %>%
-  mutate(stem.id = row_number()) %>%
-  pivot_longer(-c(stem.id, site, rep, treatment, v.r, lar),
-               names_to = "node",
-               values_to = "node_status") %>%
-  na.omit() %>%
-  group_by(treatment, lar) %>%
-  reframe(count = n()) %>%
-  mutate(year = "2023") %>%
-  left_join(pf.total.comb, by = c("treatment", "year")) %>%
-  group_by(treatment) %>%
-  mutate(per = count/total) %>%
-  filter(lar == "a") %>%
-  mutate(per = round(per, 3))
-# stems collected in late October. This is much later than if parasitoids were flying.
-
-kable(larvae.type[c("treatment", "count", "total", "per")],
-      format = "html",
-      align = "c",
-      col.names = c("Treatment", "Count", "Total Stems", "%"),
-      caption = "Stems with Living Larvae (2023)") %>%
-  kable_styling(bootstrap_options = c("striped", "bordered", 
-                                      "hover", "repsonsive"),
-                full_width = FALSE) %>%
-  row_spec(0, background = "#f2f2f2")
-```
-
-Interesting to see that `r larvae.type[1,6]*100`% of stems in the 'high' treatment groups and `r larvae.type[2,6]*100`% of stems in the 'low' treatment group had living WSS larvae.
-The stems for this experiement were harvested in late October, much later in the season when we would expect the second generation parasitoids to be searching for a host.
-
-<hr>
 
@@
 ```
-
 ```{r, echo = F, eval = F}

### 2024-03-06 locoweed-report.qmd
commit 3/6/24

@@
 ---
-title: "All Locoweed Analysis Report"
+title: "Locoweed Volatiles"
 author: "Jackson Strand"
@@
   chunk_output_type: console
+toc: true
 ---
@@
 library(MVN) # for normality test
+library(plotly)
 ```
 
+<!-- bee cover image -->
 ```{=html}
@@
 ```
+
+*Note: This project is still a work in progress.*
 
@@
 ```{r all comparison boxplot, echo = F, fig.align = 'center'}
+
 joined.data.clr %>%
@@
                       labels = c("Not Present", "Present"))
+
 ```

### 2024-03-06 msu-masters.qmd
commit 3/6/24

@@
   </a>
-  <a href="https://www.example.com" class="container">
+  <a href="locoweed-report.qmd" class="container">
     <div class="content">

### 2024-03-06 toadflax_report.qmd
commit 3/6/24

---
title: "Toadflax Analysis"
author: "Jackson Strand"
date: '`r format(Sys.Date())`'
editor_options: 
  chunk_output_type: console
toc: true
bibliography: citations/citations.bib
link-citations: true
cls: council-of-science-editors-author-date.csl
---

```{r libraries, echo = F, message = F}
library(dplyr)
library(tidyr)
library(readxl)
library(compositions) # 
library(ggplot2) # ggplot
library(vegan) # Adonis
library(tibble)
library(kableExtra) # Kable
library(plotly)
library(stats) # PCA
library(ggbiplot) # PCA biplot

```

```{r data load, echo = F}
data <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Toadflax/Toadflax 2023 Parent and Hybrid Volatiles.xlsx",
           sheet = "compounds") %>%
  rename("gc_file" = 1,
         "plant_id" = 2,
         "plant" = 3,
         "type" = 4,
         "Z-3-Hexenol" = 5,
         "a-Pinene" = 6,
         "Thujene" = 7,
         "Myrcene" = 8,
         "1-octen-3-ol" = 9,
         "6-methyl-3-heptanone" = 10,
         "B-pinene" = 11,
         "Z-3-Hexenyl Acetate" = 12,
         "Limonene" = 13,
         "Trans-B-ociemene" = 14,
         "Cis-B-ocimene" = 15,
         "Undecane" = 16,
         "Linalool" = 17,
         "Nonanal" = 18,
         "Octenyl Acetate" = 19,
         "Allo-ocimene" = 20,
         "Terpineol" = 21,
         "Methyl Salicylate" = 22,
         "Decanal" = 23,
         "P-anisaldehyde" = 24,
         "Undecanal" = 25,
         "2-Undecanal" = 26,
         "Copaene" = 27,
         "Cubebene" = 28,
         "Dodecanal" = 29,
         "Caryophyllene" = 30,
         "day" = 31) 
```

# Boxplot Comparisons
```{r boxplot data, echo = F}
boxplot.data <-
  data %>%
    pivot_longer(-c(gc_file, plant_id, plant, type, day),
               names_to = "compound",
               values_to = "value") %>%
    group_by(type, compound) %>%
    mutate(count = n(),
           average = mean(value),
           sd = sd(value),
           se = (sd/sqrt(count))) %>%
    reframe(type, compound,
            average = mean(average),
            sd = mean(sd),
            se = mean(se)) %>%
  unique()
```

```{r boxplot comparison (original data), echo = F, fig.height = 20, fig.width = 20}
boxplot.data %>%
  mutate(type = factor(type, levels = c("DT", "YT", "Natural", "Synthetic"))) %>%
  ggplot(aes(x = type, y = average, fill = type)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(x = type,
                    ymin = average - se,
                    ymax = average + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  labs(x = element_blank(),
       y = "VOC Concentration (ng/g/hr)") +
  theme(axis.text.x = element_text(angle = 45, 
                                    hjust = 1),
        legend.position = 'none',
        # legend.position = c(0.8,0.9),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(color = "black"),
        panel.grid.major.x = element_line(color = "#F2F2F2"),
        axis.text = element_text(size = 22),
        axis.title = element_text(size = 22),
        strip.text = element_text(size = 22)) +
  facet_wrap(~compound, 
             scales = "free_y",
             ncol = 3)

```

# Multivariate Analysis

Multivariate statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PCA, Permutational Multivariate ANOVA (PERMANOVA), Permutational Analysis of Multivariate Dispersion (PERMDISP), and Random Forest modeling. The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide]. VOC compositions were compared between samples and types using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan]. This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017]. Data were visualized using ggplot2 [@ggplot]. The compounds used in this analysis were selected based on consistency of presence across samples , as well as overall quality of mass spectroscopy output. 
```{r CLR transformation, echo = F}
data.clr <-
data %>%
  select(-gc_file) %>%
  pivot_longer(-c(plant_id, plant, type, day),
               names_to = "compound",
               values_to = "value") %>%
  mutate(value = log(value + 1.01, 2)) %>%
  pivot_wider(values_from = "value",
              names_from = "compound") %>%
  unite(plant_id, plant_id, plant, type, day, sep = "_") %>%
  column_to_rownames("plant_id") %>%
  clr() %>% # run the clr!
  as.data.frame() %>%
  rownames_to_column("plant_id") %>%
  separate(plant_id, into = c("plant_id", "plant", "type", "day"),
           sep = "_") %>%
  pivot_longer(-c(plant_id, plant, type, day),
               names_to = "compound",
               values_to = "value") %>%
  mutate(value = value + 4) %>%
  pivot_wider(values_from = "value",
              names_from = "compound") %>%
  mutate_at(vars(plant, type, day), list(factor))
```

## PCA

```{r PCA, echo = F}
pca.results <-
data.clr %>%
  dplyr::select(-c(plant_id, plant, type, day)) %>%
  prcomp(scale = T)

scores <- as.data.frame(pca.results$x)
loadings <- as.data.frame(pca.results$rotation)
variance <- as.data.frame(pca.results$sdev^2)

data.plus.pca <-
  data.clr %>%
    cbind(scores)

# scree.plot <- 
#   plot(pca.results, type = "l", main = "Scree Plot")

PC1 <- pca.results$x[,1]
PC2 <- pca.results$x[,2]

loadings.plot <-
loadings %>%
  dplyr::select(PC1, PC2) %>%
  rownames_to_column("compound")

meta <- data.clr %>% dplyr::select(plant, type, day)
model.data <- data.frame(PC1, PC2, meta)
model.data$day <- factor(model.data$day, levels = c("1", "2", "3", "4",
                                                    "5", "6", "7", "8",
                                                    "9"))

scores.plot <-
ggplot() +
  geom_point(data = model.data,
             aes(PC1, PC2, color = type)) +
  # geom_text(aes(label = day))
  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
  labs(x = "PC1", y = "PC2", 
       title = "Toadflax PCA Plot",
       color = element_blank()) +
  stat_ellipse(data = model.data,
               aes(PC1, PC2, color = type)) +
  labs(title = "Toadflax PCA Scores Plot") +
  theme_bw()

loadings.plot <-
ggplot() +
  geom_segment(data = loadings.plot,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.1, "inches")),
               alpha = 0.5,
               color = "red") +
  geom_text(data = loadings.plot,
            hjust = 1.1, vjust = -1.2,
            aes(PC1, PC2, label = compound),
            size = 2) +
  labs(x = "PC1", y = "PC2", title = "Toadflax PCA Loadings") +
  theme_bw()

sc.lo.plot <-
ggbiplot(pca.results, obs.scale = 1, var.scale = 1, ellipse = FALSE,
         group = data.clr$type, circle = FALSE) +
  theme_bw() +
  labs(title = "Toadflax PCA with loadings and scores")

```

Our first look at our PCA results is in the form of a scores plot. This shows us how our data is grouped by type.

```{r scores plot, echo = F}
scores.plot
```

Next, we can look at our loadings and see where each compound is pushing the differences in our PCs.

```{r loadings plot, echo = F}
loadings.plot
```

Finally, we can overlay the two plots to get an idea of how our data groups around our compounds. I found this plot a little bit busy, which is why I broke it down in the above two figures.

```{r sc.lo.plot, echo = F}
sc.lo.plot
```

## PERMANOVA

```{r PERMANOVA, echo = F}
set.seed(1)

permanova_data <-
  data.clr %>%
    unite(id, plant_id, plant, type, day, sep = "_") %>%
    column_to_rownames("id") %>%
    na.omit() %>%
    mutate_all(~.^2) %>%
    mutate_all(~.*100000) %>%
    mutate_all(round, digits = 0)

meta_data <-
  data.clr %>%
  select(plant_id, plant, type, day)

# run PERMANOVA
adonis_results <-
adonis2(permanova_data ~ meta_data$type,
        strata = meta_data$day,
        # strata = interaction(meta_data$day,
        #                      meta_data$plant),
        # method = "canberra",
        permutataions = 999,
        by = "terms")

kable(as.data.frame(adonis_results),
      format = "html",
      caption = "Toadflax Type PERMANOVA",
      align = "lccccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f2")

permanova.p <- adonis_results$`Pr(>F)`[1]

```

Our results show **strong** evidence (p-val = `r permanova.p`) of a difference between our four plant types (YT, DT, Synthetic, and Natural).

Let's look to see if there are differences between the types individually.

### 2024-03-06 toadflax_report.qmd
udpate

@@
                 bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f2")
+  row_spec(0, background = "#f2f2f9")
 

### 2024-03-06 msu-masters.qmd
updated commit

@@
   </a>
+  <a href="toadflax_report.qmd" class="container">
+    <div class="content">
+      <img src="img/msu-masters/toadflax.jpeg" alt="Image"> <!-- Center the image within the box -->
+      <div class="text">
+        <h1>Toadflax</h1>
+        <p></p>
+      </div>
+    </div>
+  </a>
 </body>

### 2024-03-06 toadflax_report.qmd
commit

@@
 
-scores.plot <-
+```
+
+Our first look at our PCA results is in the form of a scores plot. This shows us how our data is grouped by type.
+
+```{r scores plot, echo = F}
 ggplot() +
@@
   theme_bw()
+```
 
-loadings.plot <-
+Next, we can look at our loadings and see where each compound is pushing the differences in our PCs.
+
+```{r loadings plot, echo = F}
 ggplot() +
@@
             aes(PC1, PC2, label = compound),
-            size = 2) +
+            size = 4) +
   labs(x = "PC1", y = "PC2", title = "Toadflax PCA Loadings") +
   theme_bw()
-
-sc.lo.plot <-
-ggbiplot(pca.results, obs.scale = 1, var.scale = 1, ellipse = FALSE,
-         group = data.clr$type, circle = FALSE) +
-  theme_bw() +
-  labs(title = "Toadflax PCA with loadings and scores")
-
-```
-
-Our first look at our PCA results is in the form of a scores plot. This shows us how our data is grouped by type.
-
-```{r scores plot, echo = F}
-scores.plot
-```
-
-Next, we can look at our loadings and see where each compound is pushing the differences in our PCs.
-
-```{r loadings plot, echo = F}
-loadings.plot
 ```
@@
 ```{r sc.lo.plot, echo = F}
-sc.lo.plot
+ggbiplot(pca.results, obs.scale = 1, var.scale = 1, ellipse = FALSE,
+         group = data.clr$type, circle = FALSE) +
+  theme_bw() +
+  labs(title = "Toadflax PCA with loadings and scores")
 ```

### 2024-03-06 toadflax_report.qmd
commit

@@
   labs(x = "PC1", y = "PC2", title = "Toadflax PCA Loadings") +
-  theme_bw()
+  theme_bw() +
+  scale_x_continuous(expand = c(0.1, 0.1)) +
+  scale_y_continuous(expand = c(0.1, 0.1))
 ```
@@
 
+```{r PERMDISP}
+
+dist <- dist(data.clr)
+
+permdisp.results <-
+betadisper(dist, data.clr$type)
+
+permdisp.results
+p_value <- anova(permdisp.results)$'Pr(>F)'[1]
+if (p_value < 0.05) {
+  cat("Significant differences in multivariate dispersion among groups.\n")
+} else {
+  cat("We found no significant difference (pval =",anova(permdisp.results)$'Pr(>F)'[1], ") in multivariate dispersion among groups.\n")
+}
+
+pcoa <-
+capscale(dist ~ 1, data = data.clr)
+
+pcoa_df <- data.frame(scores(pcoa)$sites, group = data.clr$type)
+
+ggplot(pcoa_df, aes(x = MDS1, y = MDS2, color = group)) +
+  geom_point(alpha = 0.7) +
+  labs(x = "PCo1", y = "PCo2", color = "Type",
+       title = "PCoA Plot") +
+  theme_bw() +
+  stat_ellipse()
+
+```
+
+Basically what all of the PERMDISP and PCoA results mean is that we are not seeing any significant different in the dispersion within each of our groups when we compare them to one another.
+We are seeing a very small p-value from our PERMDISP results and a considerable amount of overlap when looking at the PCoA plot.
+
 Let's look to see if there are differences between the types individually.
+
+```{r basic ANOVA, echo = F}
+
+```

### 2024-03-06 toadflax_report.qmd
update toadflax report

@@
 
-Let's look to see if there are differences between the types individually.
+<!-- Let's look to see if there are differences between the types individually.  -->
 
-```{r basic ANOVA, echo = F}
+<!-- ```{r basic ANOVA, echo = F} -->
+<!-- anova.data <- -->
+<!--   data.clr %>% -->
+<!--   dplyr::select(-c(plant_id, plant)) -->
 
-```
+<!-- compounds <- -->
+<!-- anova.data[,3:26] %>% -->
+<!--   colnames() -->
+
+<!-- manova(cbind(compounds) ~ type + day, data = anova.data$type) -->
+<!-- ``` -->

### 2024-03-06 toadflax_report.qmd
commit end of day

@@
 ```
-
-# Boxplot Comparisons
-```{r boxplot data, echo = F}
-boxplot.data <-
-  data %>%
-    pivot_longer(-c(gc_file, plant_id, plant, type, day),
-               names_to = "compound",
-               values_to = "value") %>%
-    group_by(type, compound) %>%
-    mutate(count = n(),
-           average = mean(value),
-           sd = sd(value),
-           se = (sd/sqrt(count))) %>%
-    reframe(type, compound,
-            average = mean(average),
-            sd = mean(sd),
-            se = mean(se)) %>%
-  unique()
-```
-
-```{r boxplot comparison (original data), echo = F, fig.height = 20, fig.width = 20}
-boxplot.data %>%
-  mutate(type = factor(type, levels = c("DT", "YT", "Natural", "Synthetic"))) %>%
-  ggplot(aes(x = type, y = average, fill = type)) +
-  geom_bar(stat = "identity") +
-  geom_errorbar(aes(x = type,
-                    ymin = average - se,
-                    ymax = average + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  labs(x = element_blank(),
-       y = "VOC Concentration (ng/g/hr)") +
-  theme(axis.text.x = element_text(angle = 45, 
-                                    hjust = 1),
-        legend.position = 'none',
-        # legend.position = c(0.8,0.9),
-        panel.background = element_rect(fill = "white"),
-        axis.line = element_line(color = "black"),
-        panel.grid.major.x = element_line(color = "#F2F2F2"),
-        axis.text = element_text(size = 22),
-        axis.title = element_text(size = 22),
-        strip.text = element_text(size = 22)) +
-  facet_wrap(~compound, 
-             scales = "free_y",
-             ncol = 3)
-
-```
-
-# Multivariate Analysis
-
-Multivariate statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PCA, Permutational Multivariate ANOVA (PERMANOVA), Permutational Analysis of Multivariate Dispersion (PERMDISP), and Random Forest modeling.
-The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide].
-VOC compositions were compared between samples and types using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan].
-This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
-Data were visualized using ggplot2 [@ggplot].
-The compounds used in this analysis were selected based on consistency of presence across samples , as well as overall quality of mass spectroscopy output.
 ```{r CLR transformation, echo = F}
@@
 ```
+```{r compound order, echo = F}
+compound.order <- c("Z-3-Hexenol",
+"a-Pinene",
+"Thujene",
+"Myrcene",
+"1-octen-3-ol",
+"6-methyl-3-heptanone",
+"B-pinene",
+"Z-3-HexenylAcetate",
+"Limonene",
+"Trans-B-ociemene",
+"Cis-B-ocimene",
+"Undecane",
+"Linalool",
+"Nonanal",
+"OctenylAcetate",
+"Allo-ocimene",
+"Terpineol",
+"MethylSalicylate",
+"Decanal",
+"P-anisaldehyde",
+"Undecanal",
+"2-Undecanal",
+"Copaene",
+"Cubebene",
+"Dodecanal",
+"Caryophyllene")
+```
 
-## PCA
+I'm going to briefly analyze the data in two different formats.
 
-```{r PCA, echo = F}
+- <u>Type:</u> groupings of YT, DT, Natural, and Synthetic
+- <u>Plant:</u> groupings of each plant type - D6, Y6, D6Y6, Y6D6, HBR, RAD
+
+Overall, the data is good. I think a little bit of pairing down of the number of compounds we are looking at could be helpful.
+
+# Organized by type category
+This section has data organized by the four grouping variables - YT, DT, Natural, and Synthetic.
+
+## Simple Raw Boxplot Comparisons
+```{r boxplot data category, echo = F}
+boxplot.data.category <-
+  data %>%
+    pivot_longer(-c(gc_file, plant_id, plant, type, day),
+               names_to = "compound",
+               values_to = "value") %>%
+    group_by(type, compound) %>%
+    mutate(count = n(),
+           average = mean(value),
+           sd = sd(value),
+           se = (sd/sqrt(count))) %>%
+    reframe(type, compound,
+            average = mean(average),
+            sd = mean(sd),
+            se = mean(se)) %>%
+  unique() %>%
+  mutate(compound = factor(compound, levels = compound.order)) %>%
+  na.omit()
+```
+
+```{r boxplot comparison category (original data), echo = F, fig.height = 20, fig.width = 20}
+boxplot.data.category %>%
+  mutate(type = factor(type, levels = c("DT", "YT", "Natural", "Synthetic"))) %>%
+  ggplot(aes(x = type, y = average, fill = type)) +
+  geom_bar(stat = "identity", alpha = 0.5) +
+  geom_errorbar(aes(x = type,
+                    ymin = average - se,
+                    ymax = average + se),
+                position = position_dodge(0.9),
+                width = 0.2,
+                alpha = 0.5) +
+  labs(x = element_blank(),
+       y = "VOC Concentration (ng/g/hr)") +
+  theme(axis.text.x = element_text(angle = 45, 
+                                    hjust = 1),
+        legend.position = 'none',
+        # legend.position = c(0.8,0.9),
+        panel.background = element_rect(fill = "white"),
+        axis.line = element_line(color = "black"),
+        panel.grid.major.x = element_line(color = "#F2F2F2"),
+        axis.text = element_text(size = 22),
+        axis.title = element_text(size = 22),
+        strip.text = element_text(size = 22)) +
+  facet_wrap(~compound, 
+             scales = "free_y",
+             ncol = 3)
+
+```
+
+## Multivariate Analysis
+
+Multivariate statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PCA, Permutational Multivariate ANOVA (PERMANOVA), Permutational Analysis of Multivariate Dispersion (PERMDISP), and Random Forest modeling.
+The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide].
+VOC compositions were compared between samples and types using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan].
+This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
+Data were visualized using ggplot2 [@ggplot].
+The compounds used in this analysis were selected based on consistency of presence across samples , as well as overall quality of mass spectroscopy output.
+
+### PCA
+
+```{r PCA category, echo = F}
 pca.results <-
@@
 
-## PERMANOVA
+Grouping our compounds by family (i.e. terpenes, aldehydes, etc.) could provide a little bit of more context.
 
-```{r PERMANOVA, echo = F}
+### PERMANOVA
+
+As stated above, PERMANOVA uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
+It is useful when considering non-normal multivariate data like our toadflax data set.
+
+#### PERMANOVA - D6, Y6, Synthetic, Natural
+```{r PERMANOVA all type, echo = F}
 set.seed(1)
@@
 adonis2(permanova_data ~ meta_data$type,
-        strata = meta_data$day,
+        # strata = meta_data$day,
         # strata = interaction(meta_data$day,
@@
       format = "html",
-      caption = "Toadflax Type PERMANOVA",
+      caption = "Toadflax Type PERMANOVA: D6, Y6, Synthetic, Natural",
       align = "lccccc") %>%
@@
 
-```{r PERMDISP}
+Let's check for a difference between our synthetic and natural plant types.
+
+#### PERMANOVA - Synthetic & Natural
+```{r PERMANOVA synthetic natural type, echo = F}
+set.seed(1)
+
+types <- c("Synthetic", "Natural")
+
+permanova_data <-
+  data.clr %>%
+    filter(type %in% types) %>%
+    unite(id, plant_id, plant, type, day, sep = "_") %>%
+    column_to_rownames("id") %>%
+    na.omit() %>%
+    mutate_all(~.^2) %>%
+    mutate_all(~.*100000) %>%
+    mutate_all(round, digits = 0)
+
+meta_data <-
+  data.clr %>%
+  filter(type %in% types) %>%
+  select(plant_id, plant, type, day)
+
+# run PERMANOVA
+adonis_results <-
+adonis2(permanova_data ~ meta_data$type,
+        # strata = meta_data$day,
+        # strata = interaction(meta_data$day,
+        #                      meta_data$plant),
+        # method = "canberra",
+        permutataions = 999,
+        by = "terms")
+
+kable(as.data.frame(adonis_results),
+      format = "html",
+      caption = "Toadflax Type PERMANOVA: Synthetic & Natural",
+      align = "lccccc") %>%
+  kable_styling(full_width = FALSE,
+                bootstrap_options = c("striped", "bordered", "hover")) %>%
+  row_spec(0, background = "#f2f2f9")
+
+permanova.p <- adonis_results$`Pr(>F)`[1]
+```
+Our results show **strong** evidence (p-val = `r permanova.p`) of a difference between Natural and Synthetic plant types.
+
+### PERMDISP & PCoA
+Permutational Analysis of Multivariate Dispersions, or PERMDISP, is a multivariate statistic used to assess differences in dispersion (variation) among groups or treatments.
+This method is particularly useful when dealing with multivariate data where there are multiple variables measured on each sample or observation.
+
+Principal Coordinates Analysis (PCoA), also known as Classical Multidimensional Scaling (MDS), is a multivariate statistical technique used for exploring and visualizing patterns of similarity or dissimilarity in data.
+PCoA is closely related to Principal Component Analysis (PCA), but whereas PCA operates on the covariance matrix of the data, PCoA works directly on a distance or dissimilarity matrix.
+
+I ran a PERMDISP on our entire data set.
+
+```{r PERMDISP, echo = F, warning = F, message = F}
 
@@
 
-Basically what all of the PERMDISP and PCoA results mean is that we are not seeing any significant different in the dispersion within each of our groups when we compare them to one another.
-We are seeing a very small p-value from our PERMDISP results and a considerable amount of overlap when looking at the PCoA plot.
+Basically what all of the PERMDISP and PCoA results mean is that we are not seeing any significant different (`r anova(permdisp.results)$'Pr(>F)'[1]`) in the dispersion within each of our groups when we compare them to one another.
+We are seeing a very large p-value from our PERMDISP results and a considerable amount of overlap when looking at the PCoA plot.
+
+This means that we don't need to consider differences in broad dispersional differences between our plant types.
+
+# Organized by plant type
+This section has data organized by the six plant types - D6, Y6, D6Y6, Y6D6, HBR, and RAD.
+```{r boxplot data plant, echo = F}
+boxplot.data.plant <-
+  data %>%
+    pivot_longer(-c(gc_file, plant_id, plant, type, day),
+               names_to = "compound",
+               values_to = "value") %>%
+    group_by(plant, compound) %>%
+    mutate(count = n(),
+           average = mean(value),
+           sd = sd(value),
+           se = (sd/sqrt(count))) %>%
+    reframe(plant, compound,
+            mean = mean(average),
+            sd = mean(sd),
+            se = mean(se)) %>%
+  unique() %>%
+  left_join(data %>%
+              dplyr::select(plant, type),
+            by = "plant",
+            relationship = "many-to-many") %>%
+  unique() %>%
+  mutate(plant = factor(plant, 
+                        levels = c("D6", "Y6", 
+                                   "D6Y6", "Y6D6",
+                                   "HBR", "RAD"))) %>%
+  mutate(compound = factor(compound, levels = compound.order)) %>%
+  na.omit()
+```
+
+```{r boxplot comparison plant (original data), echo = F, fig.height = 20, fig.width = 20}
+boxplot.data.plant  %>%
+  ggplot(aes(x = plant, y = mean, fill = type)) +
+  geom_bar(stat = "identity", alpha = 0.5) +
+  geom_errorbar(aes(x = plant,
+                    ymin = mean - se,
+                    ymax = mean + se),
+                position = position_dodge(0.9),
+                width = 0.2,
+                alpha = 0.5) +
+  labs(x = element_blank(),
+       y = "VOC Concentration (ng/g/hr)") +
+  theme(axis.text.x = element_text(angle = 45, 
+                                    hjust = 1),
+        legend.position = 'bottom',
+        # legend.position = c(0.8,0.9),
+        panel.background = element_rect(fill = "white"),
+        axis.line = element_line(color = "black"),
+        panel.grid.major.x = element_line(color = "#F2F2F2"),
+        axis.text = element_text(size = 22),
+        axis.title = element_text(size = 22),
+        strip.text = element_text(size = 22),
+        legend.text = element_text(size = 22),
+        legend.title = element_text(size = 22)) +
+  facet_wrap(~compound, 
+             scales = "free_y",
+             ncol = 3)
+
+```
+
+## Multivariate Analysis
+
+### PCA
+
+```{r PCA plant, echo = F}
+pca.results <-
+data.clr %>%
+  dplyr::select(-c(plant_id, plant, type, day)) %>%
+  prcomp(scale = T)
+
+scores <- as.data.frame(pca.results$x)
+loadings <- as.data.frame(pca.results$rotation)
+variance <- as.data.frame(pca.results$sdev^2)
+
+data.plus.pca <-
+  data.clr %>%
+    cbind(scores)
+
+# scree.plot <- 
+#   plot(pca.results, type = "l", main = "Scree Plot")
+
+PC1 <- pca.results$x[,1]
+PC2 <- pca.results$x[,2]
+
+loadings.plot <-
+loadings %>%
+  dplyr::select(PC1, PC2) %>%
+  rownames_to_column("compound")
+
+meta <- data.clr %>% dplyr::select(plant, type, day)
+model.data <- data.frame(PC1, PC2, meta)
+
+```
+
+```{r scores plot plant, echo = F}
+ggplot() +
+  geom_point(data = model.data,
+             aes(PC1, PC2, color = plant)) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       title = "Toadflax PCA Plot",
+       color = element_blank()) +
+  stat_ellipse(data = model.data,
+               aes(PC1, PC2, color = plant)) +
+  labs(title = bquote("Toadflax PCA Scores Plot by "*bold("Plant"))) +
+  theme_bw() 
+```
+
+### PERMANOVA
+
+Let's say, for example, that we want to see if there is a difference in the VOC profile between the D6Y6 and Y6D6 plants. We can run a PERMANOVA to assess for a significant difference.
+
+```{r PERMANOVA D6Y6 & Y6D6, echo = F}
+set.seed(1)
+
+plants <- c("Y6D6", "D6Y6")
+
+permanova_data <-
+  data.clr %>%
+    filter(plant %in% plants) %>%
+    unite(id, plant_id, plant, type, day, sep = "_") %>%
+    column_to_rownames("id") %>%
+    na.omit() %>%
+    mutate_all(~.^2) %>%
+    mutate_all(~.*100000) %>%
+    mutate_all(round, digits = 0)
+
+meta_data <-
+  data.clr %>%
+  filter(plant %in% plants) %>%
+  select(plant_id, plant, type, day)
+
+# run PERMANOVA
+adonis_results <-
+adonis2(permanova_data ~ meta_data$plant,
+        # strata = meta_data$plant,
+        # strata = interaction(meta_data$day,
+        #                      meta_data$plant),
+        method = "canberra",
+        permutataions = 999,
+        by = "terms")
+
+kable(as.data.frame(adonis_results),
+      format = "html",
+      caption = "Toadflax Type PERMANOVA",
+      align = "lccccc") %>%
+  kable_styling(full_width = FALSE,
+                bootstrap_options = c("striped", "bordered", "hover")) %>%
+  row_spec(0, background = "#f2f2f9")
+
+permanova.p <- adonis_results$`Pr(>F)`[1]
+```
+
+Our results show **no** evidence (p-val = `r permanova.p`) of a difference between volatile profiles between Y6D6 and D6Y6 plants.
 

### 2024-03-07 toadflax_report.qmd
updated commit for today

@@
 library(ggbiplot) # PCA biplot
+library(gridExtra)
 
@@
 ```
+# Methods
 
+Boxpot and multivariate statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1).
+We employed Principal Compoenent Analysis (PCA), Permutational Multivariate ANOVA (PERMANOVA), and Permutational Analysis of Multivariate Dispersion (PERMDISP).
+The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide].
+VOC compositions were compared between samples and types using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan].
+This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
+Data were visualized using ggplot2 [@ggplot].
+The compounds used in this analysis were selected based on consistency of presence across samples , as well as overall quality of mass spectroscopy output.
+
+<div style="border: 1px solid black; padding: 10px;">
 I'm going to briefly analyze the data in two different formats.
@@
 - <u>Plant:</u> groupings of each plant type - D6, Y6, D6Y6, Y6D6, HBR, RAD
+</div>
 
-Overall, the data is good. I think a little bit of pairing down of the number of compounds we are looking at could be helpful.
+Overall, the data is quite good. I think a little bit of pairing down of the number of compounds we are looking at could be helpful.
 
-# Organized by type category
-This section has data organized by the four grouping variables - YT, DT, Natural, and Synthetic.
+# Raw Boxplot Comparisons
 
-## Simple Raw Boxplot Comparisons
+## Organized by Type
+
+**Type:** This section has data organized by the four grouping variables - YT, DT, Natural, and Synthetic.
+
 ```{r boxplot data category, echo = F}
@@
 ```
-
 ```{r boxplot comparison category (original data), echo = F, fig.height = 20, fig.width = 20}
@@
 
-## Multivariate Analysis
+## Organized by Plant
 
-Multivariate statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PCA, Permutational Multivariate ANOVA (PERMANOVA), Permutational Analysis of Multivariate Dispersion (PERMDISP), and Random Forest modeling.
-The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide].
-VOC compositions were compared between samples and types using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan].
-This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
-Data were visualized using ggplot2 [@ggplot].
-The compounds used in this analysis were selected based on consistency of presence across samples , as well as overall quality of mass spectroscopy output.
+**Plant:** This section has data organized by the six plant types - D6, Y6, D6Y6, Y6D6, HBR, and RAD.
 
-### PCA
+```{r boxplot data plant, echo = F}
+boxplot.data.plant <-
+  data %>%
+    pivot_longer(-c(gc_file, plant_id, plant, type, day),
+               names_to = "compound",
+               values_to = "value") %>%
+    group_by(plant, compound) %>%
+    mutate(count = n(),
+           average = mean(value),
+           sd = sd(value),
+           se = (sd/sqrt(count))) %>%
+    reframe(plant, compound,
+            mean = mean(average),
+            sd = mean(sd),
+            se = mean(se)) %>%
+  unique() %>%
+  left_join(data %>%
+              dplyr::select(plant, type),
+            by = "plant",
+            relationship = "many-to-many") %>%
+  unique() %>%
+  mutate(plant = factor(plant, 
+                        levels = c("D6", "Y6", 
+                                   "D6Y6", "Y6D6",
+                                   "HBR", "RAD"))) %>%
+  mutate(compound = factor(compound, levels = compound.order)) %>%
+  na.omit()
+```
+```{r boxplot comparison plant (original data), echo = F, fig.height = 20, fig.width = 20}
+boxplot.data.plant  %>%
+  ggplot(aes(x = plant, y = mean, fill = type)) +
+  geom_bar(stat = "identity", alpha = 0.5) +
+  geom_errorbar(aes(x = plant,
+                    ymin = mean - se,
+                    ymax = mean + se),
+                position = position_dodge(0.9),
+                width = 0.2,
+                alpha = 0.5) +
+  labs(x = element_blank(),
+       y = "VOC Concentration (ng/g/hr)") +
+  theme(axis.text.x = element_text(angle = 45, 
+                                    hjust = 1),
+        legend.position = 'bottom',
+        # legend.position = c(0.8,0.9),
+        panel.background = element_rect(fill = "white"),
+        axis.line = element_line(color = "black"),
+        panel.grid.major.x = element_line(color = "#F2F2F2"),
+        axis.text = element_text(size = 22),
+        axis.title = element_text(size = 22),
+        strip.text = element_text(size = 22),
+        legend.text = element_text(size = 22),
+        legend.title = element_text(size = 22)) +
+  facet_wrap(~compound, 
+             scales = "free_y",
+             ncol = 3)
 
-```{r PCA category, echo = F}
+```
+
+<hr>
+
+# Principal Component Analysis (PCA)
+
+Principal Component Analysis (PCA) is a statistical method used to simplify complex datasets by identifying patterns and relationships among variables.
+It transforms the original data into a new set of variables called principal components, which capture the most important information in the data.
+These components help visualize and understand the main sources of variation among samples or observations.
+In this study, PCA was used to analyze the variability in volatile organic compound (VOC) compositions across samples, allowing researchers to explore and interpret the data more easily.
+
+## Organized by Type
+
+```{r PCA category all, echo = F}
 pca.results <-
@@
 meta <- data.clr %>% dplyr::select(plant, type, day)
-model.data <- data.frame(PC1, PC2, meta)
-model.data$day <- factor(model.data$day, levels = c("1", "2", "3", "4",
+model.data.all <- data.frame(PC1, PC2, meta)
+model.data.all$day <- factor(model.data.all$day, levels = c("1", "2", "3", "4",
                                                     "5", "6", "7", "8",
@@
 
-Our first look at our PCA results is in the form of a scores plot. This shows us how our data is grouped by type.
+We're going to start by comparing all four plant categories. Our first look at our PCA results is in the form of a scores plot. This shows us how our data is grouped by type.
 
@@
 ggplot() +
-  geom_point(data = model.data,
+  geom_point(data = model.data.all,
              aes(PC1, PC2, color = type)) +
@@
        color = element_blank()) +
-  stat_ellipse(data = model.data,
+  stat_ellipse(data = model.data.all,
                aes(PC1, PC2, color = type)) +
   labs(title = "Toadflax PCA Scores Plot") +
-  theme_bw()
+  theme_bw() +
+  scale_color_manual(values = c("DT" = "#F8766D", 
+                                "Synthetic" = "#7CAE00",
+                                "YT" = "#00BFC4",
+                                "Natural" = "#C77CFF"))
+
 ```
@@
 
-### PERMANOVA
+### PCA with Variable Category Selection
 
-As stated above, PERMANOVA uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
-It is useful when considering non-normal multivariate data like our toadflax data set.
+I will create several PCA plots by filtering out different plant categories.
+There's a crucial distinction to be made here.
+In one set of plots, I'll filter out the specified category after running the PCA on the entire data set.
+This will result in plots with ellipses and points identical to those seen in the plots of the entire data set, just with the non-specified data removed.
+Additionally, I'll show plots that display PCA results from data filtered prior to running the PCA model.
+In these plots, points and ellipses have the potential to appear different from the previous display method.
 
-#### PERMANOVA - D6, Y6, Synthetic, Natural
-```{r PERMANOVA all type, echo = F}
-set.seed(1)
+I've also created plots showing just the ellipses for the specified groups, while leaving the rest of the data set grey. It should hopefully all become clearer as we proceed through each section.
 
-permanova_data <-
-  data.clr %>%
-    unite(id, plant_id, plant, type, day, sep = "_") %>%
-    column_to_rownames("id") %>%
-    na.omit() %>%
-    mutate_all(~.^2) %>%
-    mutate_all(~.*100000) %>%
-    mutate_all(round, digits = 0)
+In my opinion, using a PCA plot to represent a data set and then filtering the data after running the PCA model might not accurately depict the relationship between the variables.
+The placement of points on a PCA plot is influenced by the entire data set, so removing some data afterward could distort the representation and make it less meaningful.
 
-meta_data <-
-  data.clr %>%
-  select(plant_id, plant, type, day)
+As a reference, I'm going to show the plot from all varaibles and data on the right side.
 
-# run PERMANOVA
-adonis_results <-
-adonis2(permanova_data ~ meta_data$type,
-        # strata = meta_data$day,
-        # strata = interaction(meta_data$day,
-        #                      meta_data$plant),
-        # method = "canberra",
-        permutataions = 999,
-        by = "terms")
+Let's start by removing the natural hybrids from our PCA and view the results.
 
-kable(as.data.frame(adonis_results),
-      format = "html",
-      caption = "Toadflax Type PERMANOVA: D6, Y6, Synthetic, Natural",
-      align = "lccccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f9")
+### Compare: DT, YT, Synthetic
 
-permanova.p <- adonis_results$`Pr(>F)`[1]
+#### Post-PCA Model Filter
+
+```{r scores plot category POST natural removed, echo = F, fig.width = 12}
+all.scores <-
+  ggplot() +
+    geom_point(data = model.data.all,
+               aes(PC1, PC2, color = type)) +
+    # geom_text(aes(label = day))
+    geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+    geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+    labs(x = "PC1", y = "PC2", 
+         title = "Toadflax PCA Plot",
+         color = element_blank()) +
+    stat_ellipse(data = model.data.all,
+                 aes(PC1, PC2, color = type)) +
+    labs(title = "Toadflax PCA - All Data") +
+    theme_bw() +
+    scale_color_manual(values = c("DT" = "#F8766D", 
+                                  "Synthetic" = "#7CAE00",
+                                  "YT" = "#00BFC4",
+                                  "Natural" = "#C77CFF"))
+
+scores.post.natrem <- 
+  ggplot() +
+  geom_point(data = model.data.all %>%
+               filter(!type == "Natural"),
+             aes(PC1, PC2, color = type)) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       title = "Toadflax PCA Plot",
+       color = element_blank()) +
+  stat_ellipse(data = model.data.all %>%
+               filter(!type == "Natural"),
+               aes(PC1, PC2, color = type)) +
+  labs(title = "Toadflax PCA - DT, YT, Synthetic") +
+  theme_bw() +
+  # theme(legend.position = "none") +
+  scale_color_manual(values = c("DT" = "#F8766D", 
+                                "Synthetic" = "#7CAE00",
+                                "YT" = "#00BFC4",
+                                "Natural" = "#C77CFF"))
+
+grid.arrange(scores.post.natrem, all.scores, ncol = 2)
 
@@
 
-Our results show **strong** evidence (p-val = `r permanova.p`) of a difference between our four plant types (YT, DT, Synthetic, and Natural).
+#### Pre-PCA Model Filter
+```{r PCA category PRE no natural, echo = F}
+types <- c("DT", "YT", "Synthetic")
 
-Let's check for a difference between our synthetic and natural plant types.
-
-#### PERMANOVA - Synthetic & Natural
-```{r PERMANOVA synthetic natural type, echo = F}
-set.seed(1)
-
-types <- c("Synthetic", "Natural")
-
-permanova_data <-
+pca.results <-
 data.clr %>%
   filter(type %in% types) %>%
-    unite(id, plant_id, plant, type, day, sep = "_") %>%
-    column_to_rownames("id") %>%
-    na.omit() %>%
-    mutate_all(~.^2) %>%
-    mutate_all(~.*100000) %>%
-    mutate_all(round, digits = 0)
+  dplyr::select(-c(plant_id, plant, type, day)) %>%
+  prcomp(scale = T)
 
-meta_data <-
+scores <- as.data.frame(pca.results$x)
+loadings <- as.data.frame(pca.results$rotation)
+variance <- as.data.frame(pca.results$sdev^2)
+
+data.plus.pca <-
   data.clr %>%
     filter(type %in% types) %>%
-  select(plant_id, plant, type, day)
+    cbind(scores)
 
-# run PERMANOVA
-adonis_results <-
-adonis2(permanova_data ~ meta_data$type,
-        # strata = meta_data$day,
-        # strata = interaction(meta_data$day,
-        #                      meta_data$plant),
-        # method = "canberra",
-        permutataions = 999,
-        by = "terms")
+# scree.plot <- 
+#   plot(pca.results, type = "l", main = "Scree Plot")
 
-kable(as.data.frame(adonis_results),
-      format = "html",
-      caption = "Toadflax Type PERMANOVA: Synthetic & Natural",
-      align = "lccccc") %>%
-  kable_styling(full_width = FALSE,
-                bootstrap_options = c("striped", "bordered", "hover")) %>%
-  row_spec(0, background = "#f2f2f9")
+PC1 <- pca.results$x[,1]
+PC2 <- pca.results$x[,2]
 
-permanova.p <- adonis_results$`Pr(>F)`[1]
+loadings.plot <-
+loadings %>%
+  dplyr::select(PC1, PC2) %>%
+  rownames_to_column("compound")
+
+# meta data
+meta <- data.clr %>% 
+  filter(type %in% types) %>%
+  dplyr::select(plant, type, day)
+
+model.data <- data.frame(PC1, PC2, meta)
+model.data$day <- factor(model.data$day, levels = c("1", "2", "3", "4",
+                                                    "5", "6", "7", "8",
+                                                    "9"))
+
 ```
-Our results show **strong** evidence (p-val = `r permanova.p`) of a difference between Natural and Synthetic plant types.
+```{r PCA scores plot category PRE no natural, echo = F, fig.width = 12}
+scores.pre.natrm <- 
+ggplot() +
+  geom_point(data = model.data,
+             aes(PC1, PC2, color = type)) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       title = "Toadflax PCA Plot",
+       color = element_blank()) +
+  stat_ellipse(data = model.data,
+               aes(PC1, PC2, color = type)) +
+  labs(title = "Toadflax PCA Scores Plot") +
+  theme_bw()+
+  scale_color_manual(values = c("DT" = "#F8766D", 
+                                "Synthetic" = "#7CAE00",
+                                "YT" = "#00BFC4"))
 
-### PERMDISP & PCoA
-Permutational Analysis of Multivariate Dispersions, or PERMDISP, is a multivariate statistic used to assess differences in dispersion (variation) among groups or treatments.
-This method is particularly useful when dealing with multivariate data where there are multiple variables measured on each sample or observation.
+grid.arrange(scores.pre.natrm, all.scores, ncol = 2)
+```
 
-Principal Coordinates Analysis (PCoA), also known as Classical Multidimensional Scaling (MDS), is a multivariate statistical technique used for exploring and visualizing patterns of similarity or dissimilarity in data.
-PCoA is closely related to Principal Component Analysis (PCA), but whereas PCA operates on the covariance matrix of the data, PCoA works directly on a distance or dissimilarity matrix.
+### Compare: Synthetic and Natural
 
-I ran a PERMDISP on our entire data set.
+#### Post-PCA Model Filter
 
-```{r PERMDISP, echo = F, warning = F, message = F}
+```{r scores plot type HYBRIDS, echo = F, fig.width = 12}
+scores.post.hybrids <-
+ggplot() +
+  geom_point(data = model.data.all %>%
+               filter(type %in% c("Synthetic", "Natural")),
+             aes(PC1, PC2, color = type)) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       title = "Toadflax PCA Plot",
+       color = element_blank()) +
+  stat_ellipse(data = model.data.all %>%
+               filter(type %in% c("Synthetic", "Natural")),
+               aes(PC1, PC2, color = type)) +
+  labs(title = "Toadflax PCA Scores Plot - Hybrids") +
+  theme_bw() +
+  scale_color_manual(values = c("DT" = "#F8766D", 
+                                "Synthetic" = "#7CAE00",
+                                "YT" = "#00BFC4",
+                                "Natural" = "#C77CFF"))
 
-dist <- dist(data.clr)
-
-permdisp.results <-
-betadisper(dist, data.clr$type)
-
-permdisp.results
-p_value <- anova(permdisp.results)$'Pr(>F)'[1]
-if (p_value < 0.05) {
-  cat("Significant differences in multivariate dispersion among groups.\n")
-} else {
-  cat("We found no significant difference (pval =",anova(permdisp.results)$'Pr(>F)'[1], ") in multivariate dispersion among groups.\n")
-}
-
-pcoa <-
-capscale(dist ~ 1, data = data.clr)
-
-pcoa_df <- data.frame(scores(pcoa)$sites, group = data.clr$type)
-
-ggplot(pcoa_df, aes(x = MDS1, y = MDS2, color = group)) +
-  geom_point(alpha = 0.7) +
-  labs(x = "PCo1", y = "PCo2", color = "Type",
-       title = "PCoA Plot") +
-  theme_bw() +
-  stat_ellipse()
+grid.arrange(scores.post.hybrids, all.scores, ncol = 2)
 
@@
 
-Basically what all of the PERMDISP and PCoA results mean is that we are not seeing any significant different (`r anova(permdisp.results)$'Pr(>F)'[1]`) in the dispersion within each of our groups when we compare them to one another.
-We are seeing a very large p-value from our PERMDISP results and a considerable amount of overlap when looking at the PCoA plot.
+I've also plotted just the two hybrid plant categories but included the data points from the rest of the samples.
 
-This means that we don't need to consider differences in broad dispersional differences between our plant types.
+```{r scores plot type HYBRIDS with old data, echo = F, fig.width = 12, warning = F}
+scores.post.hybrids <-
+ggplot() +
+  geom_point(data = model.data.all,
+             aes(PC1, PC2, color = type)) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       title = "Toadflax PCA Plot",
+       color = element_blank()) +
+  stat_ellipse(data = model.data.all %>%
+                 filter(type %in% c("Natural", "Synthetic")),
+               aes(PC1, PC2, color = type)) +
+  labs(title = "Toadflax PCA Scores Plot - Hybrids") +
+  theme_bw() +
+  lims(x = c(-5.726419, 7.824495), y = c(-3.879407, 4.630473)) +
+  scale_color_manual(values = c("DT" = "grey", 
+                                "Synthetic" = "#7CAE00",
+                                "YT" = "grey",
+                                "Natural" = "#C77CFF"))
 
-# Organized by plant type
-This section has data organized by the six plant types - D6, Y6, D6Y6, Y6D6, HBR, and RAD.
-```{r boxplot data plant, echo = F}
-boxplot.data.plant <-
-  data %>%
-    pivot_longer(-c(gc_file, plant_id, plant, type, day),
-               names_to = "compound",
-               values_to = "value") %>%
-    group_by(plant, compound) %>%
-    mutate(count = n(),
-           average = mean(value),
-           sd = sd(value),
-           se = (sd/sqrt(count))) %>%
-    reframe(plant, compound,
-            mean = mean(average),
-            sd = mean(sd),
-            se = mean(se)) %>%
-  unique() %>%
-  left_join(data %>%
-              dplyr::select(plant, type),
-            by = "plant",
-            relationship = "many-to-many") %>%
-  unique() %>%
-  mutate(plant = factor(plant, 
-                        levels = c("D6", "Y6", 
-                                   "D6Y6", "Y6D6",
-                                   "HBR", "RAD"))) %>%
-  mutate(compound = factor(compound, levels = compound.order)) %>%
-  na.omit()
-```
-
-```{r boxplot comparison plant (original data), echo = F, fig.height = 20, fig.width = 20}
-boxplot.data.plant  %>%
-  ggplot(aes(x = plant, y = mean, fill = type)) +
-  geom_bar(stat = "identity", alpha = 0.5) +
-  geom_errorbar(aes(x = plant,
-                    ymin = mean - se,
-                    ymax = mean + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  labs(x = element_blank(),
-       y = "VOC Concentration (ng/g/hr)") +
-  theme(axis.text.x = element_text(angle = 45, 
-                                    hjust = 1),
-        legend.position = 'bottom',
-        # legend.position = c(0.8,0.9),
-        panel.background = element_rect(fill = "white"),
-        axis.line = element_line(color = "black"),
-        panel.grid.major.x = element_line(color = "#F2F2F2"),
-        axis.text = element_text(size = 22),
-        axis.title = element_text(size = 22),
-        strip.text = element_text(size = 22),
-        legend.text = element_text(size = 22),
-        legend.title = element_text(size = 22)) +
-  facet_wrap(~compound, 
-             scales = "free_y",
-             ncol = 3)
+grid.arrange(scores.post.hybrids, all.scores, ncol = 2)
 
@@
 
-## Multivariate Analysis
+#### Pre-PCA Model Filter
 
-### PCA
+```{r PCA category PRE hybrids, echo = F}
+types <- c("Natural", "Synthetic")
+
+pca.results <-
+data.clr %>%
+  filter(type %in% types) %>%
+  dplyr::select(-c(plant_id, plant, type, day)) %>%
+  prcomp(scale = T)
+
+scores <- as.data.frame(pca.results$x)
+loadings <- as.data.frame(pca.results$rotation)
+variance <- as.data.frame(pca.results$sdev^2)
+
+data.plus.pca <-
+  data.clr %>%
+    filter(type %in% types) %>%
+    cbind(scores)
+
+# scree.plot <- 
+#   plot(pca.results, type = "l", main = "Scree Plot")
+
+PC1 <- pca.results$x[,1]
+PC2 <- pca.results$x[,2]
+
+loadings.plot <-
+loadings %>%
+  dplyr::select(PC1, PC2) %>%
+  rownames_to_column("compound")
+
+# meta data
+meta <- data.clr %>% 
+  filter(type %in% types) %>%
+  dplyr::select(plant, type, day)
+
+model.data <- data.frame(PC1, PC2, meta)
+model.data$day <- factor(model.data$day, levels = c("1", "2", "3", "4",
+                                                    "5", "6", "7", "8",
+                                                    "9"))
+
+```
+```{r PCA scores plot category PRE hybrids, echo = F, fig.width = 12}
+scores.pre.hybrids <- 
+ggplot() +
+  geom_point(data = model.data,
+             aes(PC1, PC2, color = type)) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       title = "Toadflax PCA Plot",
+       color = element_blank()) +
+  stat_ellipse(data = model.data,
+               aes(PC1, PC2, color = type)) +
+  labs(title = "Toadflax PCA Scores Plot") +
+  theme_bw()+
+  scale_color_manual(values = c("DT" = "#F8766D", 
+                                "Synthetic" = "#7CAE00",
+                                "YT" = "#00BFC4",
+                                "Natural" = "#C77CFF"))
+
+grid.arrange(scores.pre.hybrids, all.scores, ncol = 2)
+```
+
+<div style="border: 3px solid #427EDC; padding: 10px; ">
+After running all of these plots, it looks to me like the largest differences are between the PRE and Post PCA Model plots when we removed the Natural category.
+Looking at the hybrid comparison, it looks like the plots are basically the same, just flipped on an axis or two.
+</div>
+
+## Organized by Plant
 
@@
 
-### PERMANOVA
+<hr>
+
+# PERMANOVA
+
+## Organized by Type
+
+As stated above, PERMANOVA uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
+It is useful when considering non-normal multivariate data like our Toadflax data set.
+
+### D6, Y6, Synthetic, Natural
+```{r PERMANOVA all type, echo = F}
+set.seed(1)
+
+permanova_data <-
+  data.clr %>%
+    unite(id, plant_id, plant, type, day, sep = "_") %>%
+    column_to_rownames("id") %>%
+    na.omit() %>%
+    mutate_all(~.^2) %>%
+    mutate_all(~.*100000) %>%
+    mutate_all(round, digits = 0)
+
+meta_data <-
+  data.clr %>%
+  select(plant_id, plant, type, day)
+
+# run PERMANOVA
+adonis_results <-
+adonis2(permanova_data ~ meta_data$type,
+        # strata = meta_data$day,
+        # strata = interaction(meta_data$day,
+        #                      meta_data$plant),
+        # method = "canberra",
+        permutataions = 999,
+        by = "terms")
+
+kable(as.data.frame(adonis_results),
+      format = "html",
+      caption = "Toadflax Type PERMANOVA: D6, Y6, Synthetic, Natural",
+      align = "lccccc") %>%
+  kable_styling(full_width = FALSE,
+                bootstrap_options = c("striped", "bordered", "hover")) %>%
+  row_spec(0, background = "#f2f2f9")
+
+permanova.p <- adonis_results$`Pr(>F)`[1]
+
+```
+
+Our results show **strong** evidence (p-val = `r permanova.p`) of a difference between our four plant types (YT, DT, Synthetic, and Natural).
+
+Let's check for a difference between our synthetic and natural plant types.
+
+### Synthetic & Natural
+```{r PERMANOVA synthetic natural type, echo = F}
+set.seed(1)
+
+types <- c("Synthetic", "Natural")
+
+permanova_data <-
+  data.clr %>%
+    filter(type %in% types) %>%
+    unite(id, plant_id, plant, type, day, sep = "_") %>%
+    column_to_rownames("id") %>%
+    na.omit() %>%
+    mutate_all(~.^2) %>%
+    mutate_all(~.*100000) %>%
+    mutate_all(round, digits = 0)
+
+meta_data <-
+  data.clr %>%
+  filter(type %in% types) %>%
+  select(plant_id, plant, type, day)
+
+# run PERMANOVA
+adonis_results <-
+adonis2(permanova_data ~ meta_data$type,
+        # strata = meta_data$day,
+        # strata = interaction(meta_data$day,
+        #                      meta_data$plant),
+        # method = "canberra",
+        permutataions = 999,
+        by = "terms")
+
+kable(as.data.frame(adonis_results),
+      format = "html",
+      caption = "Toadflax Type PERMANOVA: Synthetic & Natural",
+      align = "lccccc") %>%
+  kable_styling(full_width = FALSE,
+                bootstrap_options = c("striped", "bordered", "hover")) %>%
+  row_spec(0, background = "#f2f2f9")
+
+permanova.p <- adonis_results$`Pr(>F)`[1]
+```
+Our results show **strong** evidence (p-val = `r permanova.p`) of a difference between Natural and Synthetic plant types.
+
+## Organized by Plant
 
@@
 
+<hr>
+
+# PERMDISP & PCoA
+Permutational Analysis of Multivariate Dispersions, or PERMDISP, is a multivariate statistic used to assess differences in dispersion (variation) among groups or treatments.
+This method is particularly useful when dealing with multivariate data where there are multiple variables measured on each sample or observation.
+
+Principal Coordinates Analysis (PCoA), also known as Classical Multidimensional Scaling (MDS), is a multivariate statistical technique used for exploring and visualizing patterns of similarity or dissimilarity in data.
+PCoA is closely related to Principal Component Analysis (PCA), but whereas PCA operates on the covariance matrix of the data, PCoA works directly on a distance or dissimilarity matrix.
+
+I ran a PERMDISP on our entire data set.
+
+```{r PERMDISP, echo = F, warning = F, message = F}
+
+dist <- dist(data.clr)
+
+permdisp.results <-
+betadisper(dist, data.clr$type)
+
+permdisp.results
+p_value <- anova(permdisp.results)$'Pr(>F)'[1]
+if (p_value < 0.05) {
+  cat("Significant differences in multivariate dispersion among groups.\n")
+} else {
+  cat("We found no significant difference (pval =",anova(permdisp.results)$'Pr(>F)'[1], ") in multivariate dispersion among groups.\n")
+}
+
+pcoa <-
+capscale(dist ~ 1, data = data.clr)
+
+pcoa_df <- data.frame(scores(pcoa)$sites, group = data.clr$type)
+
+ggplot(pcoa_df, aes(x = MDS1, y = MDS2, color = group)) +
+  geom_point(alpha = 0.7) +
+  labs(x = "PCo1", y = "PCo2", color = "Type",
+       title = "PCoA Plot") +
+  theme_bw() +
+  stat_ellipse()
+
+```
+
+Basically what all of the PERMDISP and PCoA results mean is that we are not seeing any significant different (`r anova(permdisp.results)$'Pr(>F)'[1]`) in the dispersion within each of our groups when we compare them to one another.
+We are seeing a very large p-value from our PERMDISP results and a considerable amount of overlap when looking at the PCoA plot.
+
+This means that we don't need to consider differences in broad dispersional differences between our plant types.
+
 <!-- Let's look to see if there are differences between the types individually.  -->

### 2024-03-07 toadflax_report.qmd
commit

@@
 
-<div style="border: 3px solid #427EDC; padding: 10px; ">
+<div style="border: 1px solid black; padding: 10px; ">
 After running all of these plots, it looks to me like the largest differences are between the PRE and Post PCA Model plots when we removed the Natural category.

### 2024-03-07 toadflax_report.qmd
commit, last one

@@
 toc: true
+toc-depth: 4
 bibliography: citations/citations.bib
@@
 ```
+
 # Methods
@@
 
+### All Data
+
 ```{r PCA category all, echo = F}
@@
 
-### PCA with Variable Category Selection
+### Variable Category Selection
 
@@
 
-### Compare: DT, YT, Synthetic
+#### Compare: DT, YT, Synthetic
 
-#### Post-PCA Model Filter
+##### Post-PCA Model Filter
 
@@
 
-#### Pre-PCA Model Filter
+##### Pre-PCA Model Filter
 ```{r PCA category PRE no natural, echo = F}
@@
 
-### Compare: Synthetic and Natural
+#### Compare: Synthetic and Natural
 
-#### Post-PCA Model Filter
+##### Post-PCA Model Filter
 
@@
                aes(PC1, PC2, color = type)) +
-  labs(title = "Toadflax PCA Scores Plot - Hybrids") +
+  labs(title = "Toadflax PCA - Hybrids") +
   theme_bw() +
@@
 
-#### Pre-PCA Model Filter
+##### Pre-PCA Model Filter
 
@@
                aes(PC1, PC2, color = type)) +
-  labs(title = "Toadflax PCA Scores Plot") +
+  labs(title = "Toadflax PCA - Hybrids") +
   theme_bw()+

### 2024-03-08 toadflax_report.qmd
3/8 first commit

@@
 ```{r scores plot, echo = F}
+scores.all <-
 ggplot() +
@@
   labs(x = "PC1", y = "PC2", 
-       title = "Toadflax PCA Plot",
-       color = element_blank()) +
+       title = "Toadflax PCA Scores Plot") +
   stat_ellipse(data = model.data.all,
                aes(PC1, PC2, color = type)) +
-  labs(title = "Toadflax PCA Scores Plot") +
   theme_bw() +
@@
 
+ggplotly(scores.all,
+         width = 700) %>%
+  layout(legend = list(title = "")) %>%
+  config(
+    displayModeBar = TRUE,
+    displaylogo = FALSE,
+    modeBarButtonsToRemove = list('lasso2d', 
+                                  'hoverClosestCartesian',
+                                  'hoverCompareCartesian')
+  )
 ```
@@
 ```{r loadings plot, echo = F}
+loadings.ggplot <-
 ggplot() +
@@
   scale_y_continuous(expand = c(0.1, 0.1))
+
+loadings.plotly <-
+ggplotly(loadings.ggplot,
+         width = 700) %>%
+  layout(legend = list(title = "")) %>%
+  config(
+    displayModeBar = TRUE,
+    displaylogo = FALSE,
+    modeBarButtonsToRemove = list('lasso2d', 
+                                  'hoverClosestCartesian',
+                                  'hoverCompareCartesian')
+  )
+
+loadings.plotly
 ```
@@
 ```{r sc.lo.plot, echo = F}
+ggplotly(
 ggbiplot(pca.results, obs.scale = 1, var.scale = 1, ellipse = FALSE,
@@
   theme_bw() +
-  labs(title = "Toadflax PCA with loadings and scores")
+  labs(title = "Toadflax PCA with loadings and scores"),
+width = 700) %>%
+  layout(legend = list(title = "")) %>%
+  config(
+    displayModeBar = TRUE,
+    displaylogo = FALSE,
+    modeBarButtonsToRemove = list('lasso2d', 
+                                  'hoverClosestCartesian',
+                                  'hoverCompareCartesian')
+)
 ```
@@
 ```{r scores plot category POST natural removed, echo = F, fig.width = 12}
-all.scores <-
+# scores.all.plotly <-
+# ggplotly(
+scores.all <-
   ggplot() +
@@
        title = "Toadflax PCA Plot",
+       color = element_blank(),
        color = element_blank()) +
@@
 
-grid.arrange(scores.post.natrem, all.scores, ncol = 2)
+# remove the legend from one of the plots
+# for (i in 1:length(scores.all.plotly$x$data)) {
+#   scores.all.plotly$x$data[[i]]$showlegend <- FALSE
+# }
+# use subplot to plot the two plots side by side
+# subplot(scores.all.plotly, scores.post.natrem, 
+#         nrows = 1, shareY = FALSE, shareX = FALSE, 
+#         widths = c(0.5, 0.5)) %>%
+#   layout(legend = list(title = ""))
+
+grid.arrange(scores.post.natrem, scores.all, ncol = 2)
 
@@
 
-grid.arrange(scores.pre.natrm, all.scores, ncol = 2)
+grid.arrange(scores.pre.natrm, scores.all, ncol = 2)
 ```
@@
 
-grid.arrange(scores.post.hybrids, all.scores, ncol = 2)
+grid.arrange(scores.post.hybrids, scores.all, ncol = 2)
 
@@
 
-grid.arrange(scores.pre.hybrids, all.scores, ncol = 2)
+grid.arrange(scores.pre.hybrids, scores.all, ncol = 2)
 ```
@@
 
+### All Data
 ```{r PCA plant, echo = F}
@@
 ```{r scores plot plant, echo = F}
+scores.plant <-
 ggplot() +
   geom_point(data = model.data,
-             aes(PC1, PC2, color = plant)) +
+             aes(PC1, PC2, color = plant),
+             alpha = 0.8) +
   # geom_text(aes(label = day))
@@
   labs(title = bquote("Toadflax PCA Scores Plot by "*bold("Plant"))) +
+  scale_color_manual(values = c("HBR" = "#F8766D", 
+                                "RAD" = "#7CAE00",
+                                "D6" = "#00BA38",
+                                "Y6" = "#00BFC4",
+                                "Y6D6" = "#619CFF",
+                                "D6Y6" = "#F564E3")) +
   theme_bw()
+
+scores.plant
+```
+
+### Variable Plant Selection
+
+#### Compare: HBR and RAD
+
+##### Post-PCA Model Filter
+```{r scores plot type Post HBR RAD, echo = F, fig.width = 12}
+scores.post.hbrrad <-
+ggplot() +
+  geom_point(data = model.data.all %>%
+               filter(plant %in% c("HBR", "RAD")),
+             aes(PC1, PC2, color = plant),
+             alpha = 0.8) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       title = "Toadflax PCA - HBR and RAD",
+       color = element_blank()) +
+  stat_ellipse(data = model.data.all %>%
+               filter(plant %in% c("HBR", "RAD")),
+               aes(PC1, PC2, color = plant)) +
+  # geom_polygon(data = model.data.all %>%
+  #                 filter(plant %in% c("HBR", "RAD")),
+  #               aes(x = PC1, y = PC2, fill = plant),
+  #               alpha = 0.1, color = NA,
+  #               stat = "ellipse") +
+  theme_bw() +
+  scale_color_manual(values = c("HBR" = "#F8766D", 
+                                "RAD" = "#7CAE00",
+                                "D6" = "#00BA38",
+                                "Y6" = "#00BFC4",
+                                "Y6D6" = "#619CFF",
+                                "D6Y6" = "#F564E3"))
+
+grid.arrange(scores.post.hbrrad, scores.plant, ncol = 2)
+```
+
+##### Pre-PCA Model Filter
+```{r PCA category PRE hbr rad, echo = F}
+plants <- c("RAD", "HBR")
+
+pca.results <-
+data.clr %>%
+  filter(plant %in% plants) %>%
+  dplyr::select(-c(plant_id, plant, type, day)) %>%
+  prcomp(scale = T)
+
+scores <- as.data.frame(pca.results$x)
+loadings <- as.data.frame(pca.results$rotation)
+variance <- as.data.frame(pca.results$sdev^2)
+
+data.plus.pca <-
+  data.clr %>%
+    filter(plant %in% plants) %>%
+    cbind(scores)
+
+# scree.plot <- 
+#   plot(pca.results, type = "l", main = "Scree Plot")
+
+PC1 <- pca.results$x[,1]
+PC2 <- pca.results$x[,2]
+
+loadings.plot <-
+loadings %>%
+  dplyr::select(PC1, PC2) %>%
+  rownames_to_column("compound")
+
+# meta data
+meta <- data.clr %>% 
+  filter(plant %in% plants) %>%
+  dplyr::select(plant, type, day)
+
+model.data.wild <- data.frame(PC1, PC2, meta)
+model.data$day <- factor(model.data$day, levels = c("1", "2", "3", "4",
+                                                    "5", "6", "7", "8",
+                                                    "9"))
+
+```
+```{r scores plot type Pre HBR RAD, echo = F, fig.width = 12}
+scores.post.hbrrad <-
+ggplot() +
+  geom_point(data = model.data.wild,
+             aes(PC1, PC2, color = plant),
+             alpha = 0.8) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       title = "Toadflax PCA - HBR and RAD",
+       color = element_blank()) +
+  stat_ellipse(data = model.data.wild,
+               aes(PC1, PC2, color = plant)) +
+  # geom_polygon(data = model.data.all %>%
+  #                 filter(plant %in% c("HBR", "RAD")),
+  #               aes(x = PC1, y = PC2, fill = plant),
+  #               alpha = 0.1, color = NA,
+  #               stat = "ellipse") +
+  theme_bw() +
+  scale_color_manual(values = c("HBR" = "#F8766D", 
+                                "RAD" = "#7CAE00",
+                                "D6" = "#00BA38",
+                                "Y6" = "#00BFC4",
+                                "Y6D6" = "#619CFF",
+                                "D6Y6" = "#F564E3"))
+
+grid.arrange(scores.post.hbrrad, scores.plant, ncol = 2)
 ```

### 2024-03-08 toadflax_report.qmd
commit

@@
 "B-pinene",
-"Z-3-HexenylAcetate",
+"Z-3-Hexenyl Acetate",
 "Limonene",
@@
 "Nonanal",
-"OctenylAcetate",
+"Octenyl Acetate",
 "Allo-ocimene",
 "Terpineol",
-"MethylSalicylate",
+"Methyl Salicylate",
 "Decanal",
@@
 
-```{r boxplot data category, echo = F}
-boxplot.data.category <-
+```{r boxplot data type, echo = F}
+boxplot.data.type <-
   data %>%
@@
             se = mean(se)) %>%
-  unique() %>%
-  mutate(compound = factor(compound, levels = compound.order)) %>%
-  na.omit()
+  unique() 
 ```
 ```{r boxplot comparison category (original data), echo = F, fig.height = 20, fig.width = 20}
-boxplot.data.category %>%
+boxplot.data.type %>%
   mutate(type = factor(type, levels = c("DT", "YT", "Natural", "Synthetic"))) %>%
@@
 ```{r scores plot, echo = F}
-scores.all <-
+scores.type <-
 ggplot() +
@@
 
-ggplotly(scores.all,
+ggplotly(scores.type,
          width = 700) %>%
@@
 ```{r scores plot category POST natural removed, echo = F, fig.width = 12}
-# scores.all.plotly <-
+# scores.type.plotly <-
 # ggplotly(
-scores.all <-
+scores.type <-
   ggplot() +
@@
                  aes(PC1, PC2, color = type)) +
-    labs(title = "Toadflax PCA - All Data") +
+    labs(title = "Toadflax PCA - All Types") +
     theme_bw() +
@@
   labs(x = "PC1", y = "PC2", 
-       title = "Toadflax PCA Plot",
        color = element_blank(),
@@
 # remove the legend from one of the plots
-# for (i in 1:length(scores.all.plotly$x$data)) {
-#   scores.all.plotly$x$data[[i]]$showlegend <- FALSE
+# for (i in 1:length(scores.type.plotly$x$data)) {
+#   scores.type.plotly$x$data[[i]]$showlegend <- FALSE
 # }
 # use subplot to plot the two plots side by side
-# subplot(scores.all.plotly, scores.post.natrem, 
+# subplot(scores.type.plotly, scores.post.natrem, 
 #         nrows = 1, shareY = FALSE, shareX = FALSE, 
@@
 
-grid.arrange(scores.post.natrem, scores.all, ncol = 2)
+grid.arrange(scores.post.natrem, scores.type, ncol = 2)
+```
+
+```{r save files1, echo = F}
+ggsave(filename = "Toadflax PCA Scores - Type | All.png",
+       plot = scores.type,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+ggsave(filename = "Toadflax PCA (Post) - DT, YT, Synthetic.png",
+       plot = scores.type,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Button</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+</style>
+</head>
+<body>
+
+<!-- Button to download the file -->
+<a href="Toadflax PCA Scores - Type | All.png" download="TetonMountainRuns2023">
+  <button class="button">Download All Pictures</button>
+</a>
+
+</body>
+</html>
 
@@
 
-grid.arrange(scores.pre.natrm, scores.all, ncol = 2)
+grid.arrange(scores.pre.natrm, scores.type, ncol = 2)
 ```
@@
 
-grid.arrange(scores.post.hybrids, scores.all, ncol = 2)
+grid.arrange(scores.post.hybrids, scores.type, ncol = 2)
 
@@
 
-grid.arrange(scores.pre.hybrids, scores.all, ncol = 2)
+grid.arrange(scores.pre.hybrids, scores.type, ncol = 2)
 ```

### 2024-03-08 toadflax_report.qmd
commit

@@
 library(gridExtra)
+library(ggrepel)
 
@@
                color = "red") +
-  geom_text(data = loadings.plot,
-            hjust = 1.1, vjust = -1.2,
-            aes(PC1, PC2, label = compound),
-            size = 4) +
   labs(x = "PC1", y = "PC2", title = "Toadflax PCA Loadings") +
@@
 
-loadings.plotly <-
-ggplotly(loadings.ggplot,
-         width = 700) %>%
-  layout(legend = list(title = "")) %>%
-  config(
-    displayModeBar = TRUE,
-    displaylogo = FALSE,
-    modeBarButtonsToRemove = list('lasso2d', 
-                                  'hoverClosestCartesian',
-                                  'hoverCompareCartesian')
-  )
-
-loadings.plotly
+loadings.ggplot +
+  geom_text(data = loadings.plot,
+            aes(PC1, PC2, label = compound)) +
+  annotate("text", x = 0.2, y = -0.15, 
+           label = "Myrcene, B-Pinene, \nLimonene, Thujene", 
+           hjust = 0, vjust = 1, size = 4, color = "black") +
+  annotate("rect", xmin = 0.5, ymin = 0.5, xmax = 0.9, ymax = 0.9)
+
+,
+           label = "Decanal, Nonanal, \nDodecanal, Undecanal")
+,
+           hjust = 1, vjust = 1, size = 4, color = "black") +
+  geom_rect(aes(xmin = -0.3, ))
+?annotate
+loadings.ggplot + 
+  geom_label_repel(data = loadings.plot,
+                   aes(PC1, PC2, label = compound),
+                   size = 4,
+                   fill = "white",
+                   color = "black",
+                   alpha = 0.7,
+                   label.padding = unit(0.25, "lines"),
+                   label.r = unit(0.15, "lines"))
+
+loadings.ggplot + 
+  geom_label(data = loadings.plot,
+             aes(PC1, PC2, label = compound),
+             size = 4,
+             fill = "white",
+             color = "black",
+             alpha = 0.7,
+             label.padding = unit(0.25, "lines"),
+             label.r = unit(0.15, "lines"))
+
+# plotly options
+
+# loadings.plotly <-
+# ggplotly(loadings.ggplot,
+#          width = 700) %>%
+#   layout(legend = list(title = "")) %>%
+#   config(
+#     displayModeBar = TRUE,
+#     displaylogo = FALSE,
+#     modeBarButtonsToRemove = list('lasso2d', 
+#                                   'hoverClosestCartesian',
+#                                   'hoverCompareCartesian')
+#   )
+# 
+# loadings.plotly
 ```
@@
 
+##
+var.explained <- pca.results$sdev^2 / sum(pca.results$sdev^2) * 100
+formatted_variance <- format(var.explained, scientific = FALSE)
+pc_names <- paste0("PC", 1:length(var.explained))
+variance_table <- data.frame(PC = pc_names, var.explained = formatted_variance)
+
+var.table <- 
+variance_table %>%
+  as.data.frame() %>%
+  mutate(var.explained = as.numeric(var.explained)) %>%
+  mutate(var.explained = round(var.explained, 2))
+##
+
 data.plus.pca <-
@@
   geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
-  labs(x = "PC1", y = "PC2", 
+  labs(x = "PC1 (23.65%)", y = "PC2 (11.66%)", 
        title = "Toadflax PCA Plot",
@@
   geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
-  labs(x = "PC1", y = "PC2", 
+  labs(x = "PC1 (23.56%)", 
+       y = "PC2 (11.66%)", 
        title = "Toadflax PCA - HBR and RAD",

### 2024-03-08 toadflax_report.qmd
commit

@@
   geom_text(data = loadings.plot,
-            aes(PC1, PC2, label = compound)) +
-  annotate("text", x = 0.2, y = -0.15, 
-           label = "Myrcene, B-Pinene, \nLimonene, Thujene", 
-           hjust = 0, vjust = 1, size = 4, color = "black") +
-  annotate("rect", xmin = 0.5, ymin = 0.5, xmax = 0.9, ymax = 0.9)
-
-,
-           label = "Decanal, Nonanal, \nDodecanal, Undecanal")
-,
-           hjust = 1, vjust = 1, size = 4, color = "black") +
-  geom_rect(aes(xmin = -0.3, ))
-?annotate
+            aes(PC1, PC2, label = compound))
+```
+
+<img src="toadflax_plots/loadings_with_boxes.png" alt="Toadflax PCA Loadings Plot with Box Labels" width="650"/>
+
+```{r more loading plots, echo = F}
 loadings.ggplot + 

### 2024-03-08 toadflax_report.qmd
final commit of the day

@@
 ```
+```{r save files6, echo = F}
+ggsave(filename = "Toadflax PCA - All Type.png",
+       plot = scores.type,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA - All Type.png" download>
+    <button class="button">Download</button>
+  </a>
+
+</div>
+</body>
+</html>
+
+```
 
@@
 ```
-
+```{r save files6, echo = F}
+ggsave(filename = "Toadflax PCA - Loadings Plot Basic.png",
+       plot = loadings.ggplot,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA - Loadings Plot Basic.png" download>
+    <button class="button">Download</button>
+  </a>
+
+</div>
+</body>
+</html>
+
+```
 <img src="toadflax_plots/loadings_with_boxes.png" alt="Toadflax PCA Loadings Plot with Box Labels" width="650"/>
 
-```{r more loading plots, echo = F}
-loadings.ggplot + 
-  geom_label_repel(data = loadings.plot,
-                   aes(PC1, PC2, label = compound),
-                   size = 4,
-                   fill = "white",
-                   color = "black",
-                   alpha = 0.7,
-                   label.padding = unit(0.25, "lines"),
-                   label.r = unit(0.15, "lines"))
-
-loadings.ggplot + 
-  geom_label(data = loadings.plot,
-             aes(PC1, PC2, label = compound),
-             size = 4,
-             fill = "white",
-             color = "black",
-             alpha = 0.7,
-             label.padding = unit(0.25, "lines"),
-             label.r = unit(0.15, "lines"))
-
-# plotly options
-
-# loadings.plotly <-
-# ggplotly(loadings.ggplot,
-#          width = 700) %>%
-#   layout(legend = list(title = "")) %>%
-#   config(
-#     displayModeBar = TRUE,
-#     displaylogo = FALSE,
-#     modeBarButtonsToRemove = list('lasso2d', 
-#                                   'hoverClosestCartesian',
-#                                   'hoverCompareCartesian')
-#   )
-# 
-# loadings.plotly
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/loadings_with_boxes.png" download>
+    <button class="button">Download</button>
+  </a>
+
+</div>
+</body>
+</html>
+
 ```
@@
 ```
-
 ```{r save files1, echo = F}
-ggsave(filename = "Toadflax PCA Scores - Type | All.png",
+ggsave(filename = "Toadflax PCA Scores - Type_All.png",
        plot = scores.type,
@@
 ```
-
 ```{=html}
@@
 <meta name="viewport" content="width=device-width, initial-scale=1.0">
-<title>Download File Button</title>
+<title>Download File Buttons</title>
 <style>
@@
   border-radius: 8px;
+  margin: 0 200px;
 }
@@
   background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
 }
@@
 
-<!-- Button to download the file -->
-<a href="Toadflax PCA Scores - Type | All.png" download="TetonMountainRuns2023">
-  <button class="button">Download All Pictures</button>
-</a>
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA (Post) - DT, YT, Synthetic.png" download>
+    <button class="button">Download</button>
+  </a>
+
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA Scores - Type_All.png" download>
+    <button class="button">Download</button>
+  </a>
+</div>
 
@@
 grid.arrange(scores.pre.natrm, scores.type, ncol = 2)
+```
+```{r save files2, echo = F}
+ggsave(filename = "Toadflax PCA (Pre) - DT, YT, Synthetic.png",
+       plot = scores.pre.natrm,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA (Pre) - DT, YT, Synthetic.png" download>
+    <button class="button">Download</button>
+  </a>
+
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA Scores - Type_All.png" download>
+    <button class="button">Download</button>
+  </a>
+</div>
+
+</body>
+</html>
+
 ```
@@
 ```
+```{r save files3, echo = F}
+ggsave(filename = "Toadflax PCA (Post) - Hybrids.png",
+       plot = scores.post.hybrids,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA (Post) - Hybrids.png" download>
+    <button class="button">Download</button>
+  </a>
+
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA Scores - Type_All.png" download>
+    <button class="button">Download</button>
+  </a>
+</div>
+</div>
+</body>
+</html>
+
+```
 
@@
 ```{r scores plot type HYBRIDS with old data, echo = F, fig.width = 12, warning = F}
-scores.post.hybrids <-
+scores.post.hybrids.wold <-
 ggplot() +
@@
 
-grid.arrange(scores.post.hybrids, scores.type, ncol = 2)
+grid.arrange(scores.post.hybrids.wold, scores.type, ncol = 2)
+
+```
+```{r save files4, echo = F}
+ggsave(filename = "Toadflax PCA (Post) - Hybrids with entire.png",
+       plot = scores.post.hybrids.wold,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA (Post) - Hybrids with entire.png" download>
+    <button class="button">Download</button>
+  </a>
+
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA Scores - Type_All.png" download>
+    <button class="button">Download</button>
+  </a>
+</div>
+
+</body>
+</html>
 
@@
 ```
-
+```{r save files5, echo = F}
+ggsave(filename = "Toadflax PCA (Pre) - Hybrids.png",
+       plot = scores.pre.hybrids,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA (Pre) - Hybrids.png" download>
+    <button class="button">Download</button>
+  </a>
+
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA Scores - Type_All.png" download>
+    <button class="button">Download</button>
+  </a>
+</div>
+</div>
+</body>
+</html>
+
+```
 <div style="border: 1px solid black; padding: 10px; ">
@@
 meta <- data.clr %>% dplyr::select(plant, type, day)
-model.data <- data.frame(PC1, PC2, meta)
+model.data.plant <- data.frame(PC1, PC2, meta)
 
@@
 ggplot() +
-  geom_point(data = model.data,
+  geom_point(data = model.data.plant,
              aes(PC1, PC2, color = plant),
@@
   labs(x = "PC1", y = "PC2", 
-       title = "Toadflax PCA Plot",
-       color = element_blank()) +
-  stat_ellipse(data = model.data,
+       color = element_blank(),
+       title = "Toadflax PCA Scores Plot by Plant") +
+  stat_ellipse(data = model.data.plant,
                aes(PC1, PC2, color = plant)) +
-  labs(title = bquote("Toadflax PCA Scores Plot by "*bold("Plant"))) +
   scale_color_manual(values = c("HBR" = "#F8766D", 
@@
 
-scores.plant
-```
-
+ggplotly(scores.plant, 
+         width = 700) %>%
+  layout(legend = list(title = "")) %>%
+  config(
+    displayModeBar = TRUE,
+    displaylogo = FALSE,
+    modeBarButtonsToRemove = list('lasso2d', 
+                                  'hoverClosestCartesian',
+                                  'hoverCompareCartesian')
+  )
+
+```
+```{r save files6, echo = F}
+ggsave(filename = "Toadflax PCA - Type_Plant.png",
+       plot = scores.plant,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA - Type_Plant.png" download>
+    <button class="button">Download</button>
+  </a>
+
+</div>
+</body>
+</html>
+
+```
 ### Variable Plant Selection
@@
 ```
+```{r save files7, echo = F}
+ggsave(filename = "Toadflax PCA (Post_Plant_NBR RAD).png",
+       plot = scores.post.hbrrad,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA (Post_Plant_NBR RAD).png" download>
+    <button class="button">Download</button>
+  </a>
+
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA - Type_Plant.png" download>
+    <button class="button">Download</button>
+  </a>
+</div>
+</div>
+</body>
+</html>
+
+```
 
@@
 ```{r scores plot type Pre HBR RAD, echo = F, fig.width = 12}
-scores.post.hbrrad <-
+scores.pre.hbrrad <-
 ggplot() +
@@
 
-grid.arrange(scores.post.hbrrad, scores.plant, ncol = 2)
+grid.arrange(scores.pre.hbrrad, scores.plant, ncol = 2)
+```
+```{r save files8, echo = F}
+ggsave(filename = "Toadflax PCA (Pre_Plant_NBR RAD).png",
+       plot = scores.pre.hbrrad ,
+       path = "toadflax_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA (Pre_Plant_NBR RAD).png" download>
+    <button class="button">Download</button>
+  </a>
+
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax PCA - Type_Plant.png" download>
+    <button class="button">Download</button>
+  </a>
+</div>
+</div>
+</body>
+</html>
+
 ```

### 2024-03-08 toadflax_report.qmd
final final

@@
 ```
-```{r save files6, echo = F}
+```{r save files1, echo = F}
 ggsave(filename = "Toadflax PCA - All Type.png",
@@
 ```
-```{r save files6, echo = F}
+```{r save files2, echo = F}
 ggsave(filename = "Toadflax PCA - Loadings Plot Basic.png",
@@
 ```
-```{r save files1, echo = F}
+```{r save files3, echo = F}
 ggsave(filename = "Toadflax PCA Scores - Type_All.png",
@@
 ```
-```{r save files2, echo = F}
+```{r save files4, echo = F}
 ggsave(filename = "Toadflax PCA (Pre) - DT, YT, Synthetic.png",
@@
 ```
-```{r save files3, echo = F}
+```{r save files5, echo = F}
 ggsave(filename = "Toadflax PCA (Post) - Hybrids.png",
@@
 ```
-```{r save files4, echo = F}
+```{r save files6, echo = F}
 ggsave(filename = "Toadflax PCA (Post) - Hybrids with entire.png",
@@
 ```
-```{r save files5, echo = F}
+```{r save files7, echo = F}
 ggsave(filename = "Toadflax PCA (Pre) - Hybrids.png",
@@
 ```
-```{r save files6, echo = F}
+```{r save files8, echo = F}
 ggsave(filename = "Toadflax PCA - Type_Plant.png",
@@
 ```
-```{r save files7, echo = F}
+```{r save files9, echo = F}
 ggsave(filename = "Toadflax PCA (Post_Plant_NBR RAD).png",
@@
 ```
-```{r save files8, echo = F}
+```{r save files10, echo = F}
 ggsave(filename = "Toadflax PCA (Pre_Plant_NBR RAD).png",

### 2024-03-08 bugs.qmd
updated bugs.qmd file, was showing missing image.

@@
 ```
-
-```{=html}
-<!DOCTYPE html>
-<html lang="en">
-<head>
-<meta charset="UTF-8">
-<meta name="viewport" content="width=device-width, initial-scale=1.0">
-<title>Image Crop Example</title>
-<style>
-  .image-container {
-    width: 300px; /* Set the width of the container */
-    height: 200px; /* Set the height of the container */
-    overflow: hidden; /* Hide any overflowing content */
-  }
-  .cropped-image {
-    width: 100%; /* Set the width of the image to fill the container */
-    height: auto; /* Automatically adjust the height to maintain aspect ratio */
-    object-fit: cover; /* Crop the image to cover the container */
-  }
-</style>
-</head>
-<body>
-
-<div class="image-container">
-  <img class="cropped-image" src="img/bugs/loco-bee.png" alt="Cropped Image">
-</div>
-
-</body>
-</html>
-
-```

### 2024-03-09 index.qmd
update about me page picture

@@
   id: about-block
-  image: "img/aboutme/crescent.HEIC"
+  image: "img/aboutme/crescent.jpg"
   links:

### 2024-03-09 locoweed-report.qmd
update locoweed report images

@@
 <div class="gallery">
-<img src="img/loco-plants/IMG_0064.HEIC" alt="Image 1" onclick="expandImage(this)">
-<img src="img/loco-plants/IMG_0820 2.HEIC" alt="Image 2" onclick="expandImage(this)">
-<img src="img/loco-plants/IMG_1165.HEIC" alt="Image 3" onclick="expandImage(this)">
-<img src="img/loco-plants/IMG_1177.HEIC" alt="Image 4" onclick="expandImage(this)">
-<img src="img/loco-plants/IMG_1923.HEIC" alt="Image 5" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_0064.jpg" alt="Image 1" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_0820 2.jpg" alt="Image 2" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_1165.jpg" alt="Image 3" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_1177.jpg" alt="Image 4" onclick="expandImage(this)">
+<img src="img/loco-plants/IMG_1923.jpg" alt="Image 5" onclick="expandImage(this)">
 </div>

### 2024-03-09 msu-masters.qmd
updated website

--- 2024-03-06 msu-masters.qmd
@@
 ---
-
 ```{=html}
@@
 <style>
+  body {
+    margin: 0; /* Remove default margin */
+    padding: 0; /* Remove default padding */
+  }
+
   .container {
-    width: 400px;
+    width: calc(100% - 20px); /* Adjusted width to fit the screen and account for margins */
+    max-width: 600px; /* Limiting maximum width */
     height: auto; /* Set container height */
@@
     overflow: hidden; /* Hide overflow content */
-    position: relative;
-    display: inline-block;
-    transition: box-shadow 0.15s ease; /* Transition for box-shadow */
-    box-shadow: 0 2px 5px rgba(0, 0, 0, 0.1); /* Slight shadow by default */
-    text-align: center; /* Center content horizontally */
+    margin-bottom: 20px; /* Add margin to create space between rows */
+    margin-right: auto; /* Push containers to the left */
+    margin-left: 10px; /* Adjust margin between the left side of the screen */
+    display: flex; /* Use flexbox for layout */
+    align-items: center; /* Center items vertically */
     text-decoration: none; /* Remove underline */
-    margin-right: 20px; /* Add margin to create space between boxes */
   }
@@
   .container:last-child {
-    margin-right: 0; /* Remove margin from the last box */
+    margin-bottom: 0; /* Remove margin from the last box */
   }
@@
 
-  .content {
-    width: 100%;
-    height: 100%;
+  .image-container {
     padding: 20px;
@@
 
-  .content img {
-    width: 50%;
+  .image-container img {
+    width: 80px; /* Adjusted width for smaller image */
     height: auto;
     display: block;
-    margin: 0 auto; /* Center the image horizontally */
-    object-fit: cover; /* Crop the image to cover container */
+    border-radius: 8px; /* Rounded corners for image */
   }
 
-  .text {
-    margin-top: 10px;
-    text-align: center; /* Center text within the box */
-    text-decoration: none;
+  .content {
+    padding: 20px;
+    box-sizing: border-box;
+    flex: 1; /* Take remaining space */
   }
@@
   .text h1 {
-    font-size: 24px; /* Larger text size */
-    margin-top: 1px;
+    font-size: 18px; /* Adjusted text size */
+    margin-top: 0;
     margin-bottom: 5px; /* Adjust margin between large and small text */
@@
     color: #666;
+    text-align: justify; /* Align text */
+  }
+
+  @media (max-width: 768px) {
+    .container {
+      width: calc(100% - 20px); /* Adjusted width to fit the screen */
+    }
   }
@@
   <a href="wss.qmd" class="container">
+    <div class="image-container">
+      <img src="img/msu-masters/oviposition.JPG" alt="Image">
+    </div>
     <div class="content">
-      <img src="img/msu-masters/wss.JPG" alt="Image"> <!-- Center the image within the box -->
       <div class="text">
@@
   <a href="locoweed-report.qmd" class="container">
+    <div class="image-container">
+      <img src="img/msu-masters/loco-bee2.jpg" alt="Image">
+    </div>
     <div class="content">
-      <img src="img/msu-masters/loco-bee2.jpg" alt="Image"> <!-- Center the image within the box -->
       <div class="text">
@@
         <p>Comparison of volatile organic compounds profiles of <em>Oxytropis</em> plants that contained the commensal endophyte <em>Alternaria oxytropis</em> and those that did not.</p>
+      </div>
+    </div>
+  </a>
+  <a href="toadflax_report.qmd" class="container">
+    <div class="image-container">
+      <img src="img/msu-masters/toadflax.jpeg" alt="Image">
+    </div>
+    <div class="content">
+      <div class="text">
+        <h1>Toadflax</h1>
+        <p></p>
       </div>

### 2024-03-09 projects.qmd
updated website

--- 2024-02-29 projects.qmd
@@
 ---
-::: {style="display: flex;"}
 
-<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
-  <a href="msu-masters.qmd">
-    <img src="img/msu-logo.jpg" alt="Montana State University" width="200" height="200">
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Box with Rounded Edges</title>
+<style>
+  body {
+    margin: 0; /* Remove default margin */
+    padding: 0; /* Remove default padding */
+  }
+
+  .container {
+    width: calc(100% - 20px); /* Adjusted width to fit the screen and account for margins */
+    max-width: 600px; /* Limiting maximum width */
+    height: auto; /* Set container height */
+    border: 0.5px solid #ccc; /* Light grey color */
+    border-radius: 8px; /* Less rounded edges */
+    overflow: hidden; /* Hide overflow content */
+    margin-bottom: 20px; /* Add margin to create space between rows */
+    margin-right: auto; /* Push containers to the left */
+    margin-left: 10px; /* Adjust margin between the left side of the screen */
+    display: flex; /* Use flexbox for layout */
+    align-items: center; /* Center items vertically */
+    text-decoration: none; /* Remove underline */
+  }
+
+  .container:last-child {
+    margin-bottom: 0; /* Remove margin from the last box */
+  }
+
+  .container:hover {
+    box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */
+  }
+
+  .image-container {
+    padding: 20px;
+    box-sizing: border-box;
+  }
+
+  .image-container img {
+    width: 80px; /* Adjusted width for smaller image */
+    height: auto;
+    display: block;
+    border-radius: 8px; /* Rounded corners for image */
+  }
+
+  .content {
+    padding: 20px;
+    box-sizing: border-box;
+    flex: 1; /* Take remaining space */
+  }
+
+  .text h1 {
+    font-size: 18px; /* Adjusted text size */
+    margin-top: 0;
+    margin-bottom: 5px; /* Adjust margin between large and small text */
+    color: black;
+  }
+  .text h1:hover {
+    color: #2068DC; /* Change text color to blue on hover */
+  }
+
+  .text p {
+    font-size: 14px;
+    margin-top: 5px;
+    color: #666;
+    text-align: justify; /* Align text */
+  }
+
+  @media (max-width: 768px) {
+    .container {
+      width: calc(100% - 20px); /* Adjusted width to fit the screen */
+    }
+  }
+</style>
+</head>
+<body>
+  <a href="msu-masters.qmd" class="container">
+    <div class="image-container">
+      <img src="img/projects-logos/msu-logo.jpg" alt="Image">
+    </div>
+    <div class="content">
+      <div class="text">
+        <h1>Montana State University Master's</h1>
+        <p>Co-advised by Dr. David Weaver and Dr. Tracy Sterling.</p>
+      </div>
+    </div>
   </a>
-  <figcaption style="text-align: center;">Montana State University</figcaption>
-</figure>
+  <a href="bballref.qmd" class="container">
+    <div class="image-container">
+      <img src="img/projects-logos/bballref-logo-circle.jpg" alt="Image">
+    </div>
+    <div class="content">
+      <div class="text">
+        <h1>NBA Roster Analysis Tool</h1>
+        <p></p>
+      </div>
+    </div>
+  </a>
+</body>
+</html>
 
-<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
-  <a href="bballref.qmd">
-    <img src="img/bballref-logo.jpeg" alt="Basketball Reference Tool" width="200" height="200">
-  </a>
-  <figcaption style="text-align: center;">Basketball Reference Tool</figcaption>
-</figure>
-
-:::
+```

### 2024-03-09 save.qmd
updated commit for today
3/8 first commit
updated website

@@
 title: "save"
+editor_options: 
+  chunk_output_type: console
 ---
@@
 ```
+
+# how to see what the default colors are in R
+```{r}
+library(scales)
+
+for(i in 1:6){
+  print(hue_pal()(i))
+}
+hue_pal()(4)
+show_col(hue_pal()(6))
+
+```
+
+# menu grid of boxes for selection
+
+```{=html}
+<!-- <!DOCTYPE html> -->
+<!-- <html lang="en"> -->
+<!-- <head> -->
+<!-- <meta charset="UTF-8"> -->
+<!-- <meta name="viewport" content="width=device-width, initial-scale=1.0"> -->
+<!-- <title>Box with Rounded Edges</title> -->
+<!-- <style> -->
+<!--   .container { -->
+<!--     width: 1000px; -->
+<!--     height: auto; /* Set container height */ -->
+<!--     border: 0.5px solid #ccc; /* Light grey color */ -->
+<!--     border-radius: 8px; /* Less rounded edges */ -->
+<!--     overflow: hidden; /* Hide overflow content */ -->
+<!--     position: relative; -->
+<!--     display: inline-block; -->
+<!--     transition: box-shadow 0.15s ease; /* Transition for box-shadow */ -->
+<!--     box-shadow: 0 2px 5px rgba(0, 0, 0, 0.1); /* Slight shadow by default */ -->
+<!--     text-align: center; /* Center content horizontally */ -->
+<!--     text-decoration: none; /* Remove underline */ -->
+<!--     margin-right: 20px; /* Add margin to create space between boxes */ -->
+<!--   } -->
+
+<!--   .container:last-child { -->
+<!--     margin-right: 0; /* Remove margin from the last box */ -->
+<!--   } -->
+
+<!--   .container:hover { -->
+<!--     box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */ -->
+<!--   } -->
+
+<!--   .content { -->
+<!--     width: 100%; -->
+<!--     height: 100%; -->
+<!--     padding: 20px; -->
+<!--     box-sizing: border-box; -->
+<!--   } -->
+
+<!--   .content img { -->
+<!--     width: 50%; -->
+<!--     height: auto; -->
+<!--     display: block; -->
+<!--     margin: 0 auto; /* Center the image horizontally */ -->
+<!--     object-fit: cover; /* Crop the image to cover container */ -->
+<!--   } -->
+
+<!--   .text { -->
+<!--     margin-top: 10px; -->
+<!--     text-align: center; /* Center text within the box */ -->
+<!--     text-decoration: none; -->
+<!--   } -->
+
+<!--   .text h1 { -->
+<!--     font-size: 24px; /* Larger text size */ -->
+<!--     margin-top: 1px; -->
+<!--     margin-bottom: 5px; /* Adjust margin between large and small text */ -->
+<!--     color: black; -->
+<!--   } -->
+<!--   .text h1:hover { -->
+<!--     color: #2068DC; /* Change text color to blue on hover */ -->
+<!--   } -->
+
+<!--   .text p { -->
+<!--     font-size: 14px; -->
+<!--     margin-top: 5px; -->
+<!--     color: #666; -->
+<!--   } -->
+<!-- </style> -->
+<!-- </head> -->
+<!-- <body> -->
+<!--   <a href="wss.qmd" class="container"> -->
+<!--     <div class="content"> -->
+<!--       <img src="img/msu-masters/wss.JPG" alt="Image"> <!-- Center the image within the box --> 
+<!--       <div class="text"> -->
+<!--         <h1>Wheat Stem Sawfly</h1> -->
+<!--         <p>Understanding the importance of smooth brome in promoting a Braconid parasitoid that targets the Wheat Stem Sawfly.</p> -->
+<!--       </div> -->
+<!--     </div> -->
+<!--   </a> -->
+<!--   <a href="locoweed-report.qmd" class="container"> -->
+<!--     <div class="content"> -->
+<!--       <img src="img/msu-masters/loco-bee2.jpg" alt="Image"> <!-- Center the image within the box -->
+<!--       <div class="text"> -->
+<!--         <h1>Locoweed Volatiles</h1> -->
+<!--         <p>Comparison of volatile organic compounds profiles of <em>Oxytropis</em> plants that contained the commensal endophyte <em>Alternaria oxytropis</em> and those that did not.</p> -->
+<!--       </div> -->
+<!--     </div> -->
+<!--   </a> -->
+<!--   <a href="toadflax_report.qmd" class="container"> -->
+<!--     <div class="content"> -->
+<!--       <img src="img/msu-masters/toadflax.jpeg" alt="Image"> <!-- Center the image within the box --> 
+<!--       <div class="text"> -->
+<!--         <h1>Toadflax</h1> -->
+<!--         <p></p> -->
+<!--       </div> -->
+<!--     </div> -->
+<!--   </a> -->
+<!-- </body> -->
+<!-- </html> -->
+
+```

### 2024-03-20 index.qmd
commit for 3/20

@@
 <div  style="margin: 30px; text-align: center;">
-  <a class="btn btn-primary" href="cv.qmd" role="button" style="padding: 15px 30px;">View my CV</a>
+  <a class="btn btn-primary" href="qmd/cv.qmd" role="button" style="padding: 15px 30px;">View my CV</a>
 </div>

### 2024-03-20 qmd/aboutme.qmd
commit for 3/20

---
title: "Jackson Strand"
about:
  id: me
  template: trestles
  links:
    - icon: linkedin
      text: LinkedIn
      href: https://linkedin.com/in/jacksonstrand
    - icon: instagram
      text: Instagram
      href: https://www.instagram.com/jstrand894/
    - icon: github
      text: GitHub
      href: https://github.com/jstrand894 
  image: ../img/aboutme/deadhorse.jpeg

sidebar: false
---

:::{#me}

I am currently a Master's graduate student at Montana State University studying entomology. I live in Bozeman, MT with my wife Grace.

```{=html}
<div  style="margin: 30px; text-align: center;">
  <a class="btn btn-primary" href="cv.qmd" role="button" style="padding: 15px 30px;">View my CV</a>
</div>
```

![Me and Grace at Glacier National Park, 2023](../img/aboutme/iceberg.JPG)

:::

<!-- ```{=html} -->
<!-- <!DOCTYPE html> -->
<!-- <html lang="en"> -->
<!-- <head> -->
<!--   <meta charset="UTF-8"> -->
<!--   <meta name="viewport" content="width=device-width, initial-scale=1.0"> -->
<!--   <title>Image Gallery</title> -->
<!--   <style> -->
<!--     .gallery { -->
<!--       display: grid; -->
<!--       grid-template-columns: repeat(auto-fill, minmax(200px, 1fr)); -->
<!--       grid-gap: 10px; -->
<!--     } -->
<!--     .gallery img { -->
<!--       width: 100%; -->
<!--       height: auto; -->
<!--       cursor: pointer; -->
<!--       transition: transform 0.3s ease; -->
<!--     } -->
<!--     .gallery img:hover { -->
<!--       transform: scale(1.05); /* Slightly increase size on hover */ -->
<!--     } -->
<!--     .expanded-container { -->
<!--       position: fixed; -->
<!--       top: 0; -->
<!--       left: 0; -->
<!--       width: 100%; -->
<!--       height: 100%; -->
<!--       background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */ -->
<!--       display: none; /* Initially hidden */ -->
<!--       justify-content: center; -->
<!--       align-items: center; -->
<!--       z-index: 9999; -->
<!--     } -->

<!--     .expanded { -->
<!--       max-width: 80%; -->
<!--       max-height: 80%; -->
<!--       cursor: pointer; -->
<!--     } -->
<!--   </style> -->
<!-- </head> -->
<!-- <body> -->

<!-- <div class="gallery"> -->
<!--   <img src="../img/bugs/blueflowers.JPG" alt="Image 1" onclick="expandImage(this)"> -->
<!--   <img src="../img/bugs/brownmoth.JPG" alt="Image 2" onclick="expandImage(this)"> -->
<!--   <img src="../img/bugs/butterfly-purpleflower.JPG" alt="Image 3" onclick="expandImage(this)"> -->
<!-- </div> -->

<!-- <div class="expanded-container" onclick="closeExpanded()"> -->
<!--   <img class="expanded" id="expanded-image" onclick="navigateGallery(event)"> -->
<!-- </div> -->

<!-- <script> -->
<!--   var currentImageIndex = 0; -->
<!--   var galleryImages = document.querySelectorAll('.gallery img'); -->

<!--   function expandImage(img) { -->
<!--     currentImageIndex = Array.from(galleryImages).indexOf(img); -->
<!--     document.getElementById('expanded-image').src = img.src; -->
<!--     document.querySelector('.expanded-container').style.display = 'flex'; -->
<!--   } -->

<!--   function closeExpanded() { -->
<!--     document.querySelector('.expanded-container').style.display = 'none'; -->
<!--   } -->

<!--   function navigateGallery(event) { -->
<!--     // Calculate the click position relative to the expanded image -->
<!--     var clickX = event.pageX - event.target.getBoundingClientRect().left; -->
<!--     var imageWidth = event.target.offsetWidth; -->

<!--     // Determine if the click was on the left or right side of the image -->
<!--     if (clickX < imageWidth / 2) { -->
<!--       // Clicked on the left side: navigate to the previous image -->
<!--       currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length; -->
<!--     } else { -->
<!--       // Clicked on the right side: navigate to the next image -->
<!--       currentImageIndex = (currentImageIndex + 1) % galleryImages.length; -->
<!--     } -->

<!--     // Update the expanded image source -->
<!--     document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src; -->
<!--   } -->
<!-- </script> -->

<!-- </body> -->
<!-- </html> -->

<!-- ``` -->

### 2024-03-20 qmd/cv.qmd
commit for 3/20

---
layout: page
title: "Cirriculum vitae"
execute:
  freeze: true  # never re-render during project render
engine: knitr 
resources:
  - "../Aboutme/jackson_strand_updated_cv.pdf"
cv:
  pdf: "../Aboutme/jackson_strand_updated_cv.pdf"
---

```{css echo=FALSE}
.embed-container {
    position: relative;
    padding-bottom: 129%;
    height: 0;
    overflow: hidden;
    max-width: 100%;
}
.embed-container iframe,
.embed-container object,
.embed-container embed {
    position: absolute;
    top: 0;
    left: 0;
    width: 100%;
    height: 100%;
}
```

```{=html}
<div class="embed-container">
  <iframe src="`r rmarkdown::metadata$cv$pdf`#toolbar=0" style="border: 0.5px;width: 100%;"></iframe>
</div>
```

```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Download File Button</title>
<style>
.container {
  display: flex;
  justify-content: center;
  align-items: center;
  height: 10vh;
}
/* Button styling */
.button {
  background-color: #2068DC; /* Header of website color (blue) */
  border: none;
  color: white;
  padding: 15px 32px;
  text-align: center;
  text-decoration: none;
  display: inline-block;
  font-size: 16px;
  cursor: pointer;
  border-radius: 8px;
}
.button:hover {
  background-color: #3F91FB; /* Sky blue */
}
</style>
</head>
<body>

<div class = "container">
  <!-- Button to download the file -->
  <a href="../Aboutme/jackson_strand_updated_cv.pdf" download="Jackson Strand CV">
    <button class="button">Download</button>
  </a>
</div>

</body>
</html>
```

### 2024-03-20 qmd/litreview.qmd
commit for 3/20

---
title: "Literature Review (Thesis)"
author: "Jackson Strand"
date: March 20, 2024
output: 
  html_document:
    toc: true
    code_folding: hide
bibliography: proposal.bib
link-citations: true
biblio-style: apalike
cls: elsevier-with-titles.csl
# cls: council-of-science-editors-author-date.csl
editor_options: 
  chunk_output_type: inline
toc: true
---

<!-- Button to download word document version -->
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Download File Button</title>
<style>
/* Button styling */
.button {
  background-color: #2068DC; /* Header of website color (blue) */
  border: none;
  color: white;
  padding: 15px 32px;
  text-align: center;
  text-decoration: none;
  display: inline-block;
  font-size: 16px;
  margin: 4px 2px;
  cursor: pointer;
  border-radius: 8px;
}
.button:hover {
  background-color: #3F91FB; /* Sky blue */
}
</style>
</head>
<body>

<!-- Button to download the file -->
<a href="litreview.docx" download="Jackson Strand Proposal">
  <button class="button">Download as Word Document</button>
</a>

<script>
// Get the current date in YYYY-MM-DD format
function getCurrentDate() {
  const now = new Date();
  const year = now.getFullYear();
  const month = String(now.getMonth() + 1).padStart(2, '0');
  const day = String(now.getDate()).padStart(2, '0');
  return `${year}-${month}-${day}`;
}

// Update the download attribute of the link with the current date
document.addEventListener('DOMContentLoaded', function() {
  const downloadLink = document.getElementById('downloadLink');
  const currentDate = getCurrentDate();
  downloadLink.setAttribute('download', `Jackson Strand Proposal ${currentDate}`);
});
</script>

</body>
</html>
```

# Project Overview
## Motivation
The motivation for my research is to improve yield for wheat growers by reducing wheat stem sawfly (WSS) populations. If we can show that healthier, more robust smooth brome leads to increased parasitoid populations and thus, increased WSS parasitism and potentially decreased damage by WSS, we can better educate growers on how to increase yields. In terms of scientific merit, understanding these mechanisms is key in ongoing WSS research as it explains several how and why questions associated with the ecological relationship between smooth brome and WSS parasitism.

## Research Question
What role does smooth brome (*Bromus inermis*) play in the ecological relationship between wheat stem sawfly (*Cephus cinctus*) and wheat stem sawfly parasitoids (*Bracon cephi* and *Bracon lissogaster*) in dryland wheat production landscape within Montana?

## Hypothesis
Smooth brome acts as an ecological sink for WSS, while serving as a parasitoid source because WSS larvae experience high mortality within smooth brome stems, while any survivors experience high rates of parasitism as a second source of mortality. In addition, early season parasitoids are attracted to infested smooth brome because it produces volatiles similar to those produced by winter wheat infested with larval WSS.

## Prediction
WSS larvae will experience both high levels of mortality and similar levels of parasitism within smooth brome stems when compared to the survivability and parasitism of WSS larvae in the adjacent fields. Also, the blend of volatile compounds produced by WSS infested smooth brome will qualitatively match in composition and exceed the amounts produced by infested winter wheat.

## Research Approach
My research is composed of three separate studies. First, there is a field survey to assess the survivability of WSS larvae and associated parasitoid frequency within both winter wheat and adjacent smooth brome. Second, there are controlled field infestations which will assess the survivability of WSS larvae within smooth brome stems when controlled for other density driven factors. And third, there is a greenhouse study to measure the volatile compounds produced by winter wheat and smooth brome. Using the results of the volatile collection, we can assess the parasitoid response both chemically with electroantennography and behaviorally with a Y-tube olfactory bioassay.

## Importance of Topic
The wheat stem sawfly (WSS, Cephus cinctus Norton, Hymenoptera: Cephidae), is a serious problem for wheat producers throughout the Northern Great Plains of North America [@Fulbright2017]. In Montana alone, the economic damage caused by the pest was reported to be upwards of $80 million annually between 2011-2016 [@Fulbright2017]. Current techniques in combatting the damage caused by the WSS use a variety of chemical, cultural, and biocontrol strategies with limited short-term effectiveness and long-term sustainability. 
Utilizing biological control via promoting braconid parasitoid habitat holds considerable value to both growers and agricultural observers. As we seek to increase agricultural yields, we must also consider the environmental impact that tactics such as pesticides and tillage have on the future success on wheat farming.

# Prior Research on Topic (Literature Review)
## History and Distribution of Wheat Stem Sawfly
The wheat stem sawfly was first described in 1872 by Edward Norton based upon a specimen found in Colorado [@Norton1872]. In 1890 it was found in California and Montana [@Ainslie1920]. Native to North America [@Lesieur2016], the largest populations of WSS are currently found in the western Canadian prairies as well as Wyoming, Montana, Colorado, and North and South Dakota [@Wallace1966; @Weiss1992; @Cockrell2021].

## Host Plants of Wheat Stem Sawfly
Wheat stem sawfly was first found the hollow-stemmed wild grasses such as Agropyron and Elymus in Colorado in 1872 [@Ainslie1920]. Soon thereafter it was found in wheat in 1895 in Manitoba, Canada. First referred to as the “western-grass stem sawfly”, WSS was found with native Montana grasses including western wheatgrass (Pascopyrun smithii) and slender wheat grass (Elymus trachycaulus) [@Ainslie1920]. As settlers began moving west and small grain cereal crop production increased on the landscape, the native grasses that WSS foraged upon became quite rare [@Ainslie1929]. WSS was found damaging spring wheat for the first time in 1895 in Manitoba and Saskatchewan [@Wallace1966]. With the diminishment of its natural hosts, WSS has adapted to infest the exotic non-cultivated and cultivated grasses now present [@Criddle1923]. WSS infestation has been shown in many species native to Montana including western wheatgrass (Pascopyrun smithii), slender wheat grass (Elymus trachycaulus) and bluebunch wheatgrass, Pseudoroegneria spicata [@Cockrell2021].

Smooth brome (*Bromus inermis* L.) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 from Hungary and Russia for soil conservation and use as livestock feed [@Salesman2011]. Throughout the early 19th and 20th century, it was used throughout much of the upper half of North America. It is a perennial grass, and more drought resistant than other exotic grasses [@Mapfumo2002]. Due to its sod forming nature and large root system, it is extensively used to promote soil retention along highways (USDA-NRCS 2010).

Smooth brome was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Farstad1945]. However, due to its propensity to form monocultures and outcompete many native grasses, smooth brome was labeled by many as invasive [@Carlson1985; @Willson2000; @Dillemuth2008]. Smooth brome populations play a considerable role in altering native herbivore population dynamics and movement on the prairies of North America [@Dillemuth2008].

## Biology of Wheat Stem Sawfly
### Adult
Adults are relatively small (8-13 mm) with a narrow elongated and slightly compressed black body, yellow legs, and three characteristic black bands across the abdomen. Females are larger than males and a visible long, dark, ovipositor. Both male and female adults are weak fliers, rarely flying long distances at one time [@Criddle1923]. They can be found on stems with their heads downward and legs aligned with their body, encircling the stem [@Ainslie1920; @Criddle1923]. WSS are univoltine, having only one generation per year. Males are haploid (9 chromosomes), while females are diploid – females who do not find a mate will produce male offspring, while those who do produce female progeny [@Holmes1978]. Fully mature adults hatch in the spring once it has sufficiently warmed and the moisture levels in the soil reach threshold levels [@Holmes1978; @mendoza2006]. In Montana, this period begins in mid- to late-May and ends early in July [@Morrill1996; @Weaver2005]. Their lifespan usually lasts around five to eight days, emerging from May until June. Males emerge before the females, ensuring that many of the early females to hatch will find a mate and deposit eggs that will be female [@Mackay2011]. Copulation lasts less than one minute with the male holding the back of the female with their abdomen curved down under the female ovipositor. Fertilized females lay eggs several hours after copulation [@Holmes1978].

### Oviposition
After locating a suitable host stem, female WSS oviposit an egg inside the host plant lumen. WSS larvae hatch inside the stem and feed on the stem’s parenchyma tissue [@Beres2011]. As the stem begins to senesce in the fall, the larvae travel to the base of the stem and makes a V-shaped groove [@Criddle1923] above a constructed protective hibernaculum where they will overwinter [@Holmes1960]. As a result of this deep groove, the stem weakens and topples easily. The resulting lodged stems are not only challenging for growers to harvest but also exhibit decreased kernel yield [@Beres2011].

### Egg
Newly laid eggs are crescent in shape, milky-white in color, and range between 1 mm to 1.25 mm in length and 0.33 to 0.42 mm wide [@Ainslie1920].  After oviposition, the egg lies freely within the lumen of the host stem or in a hollow created by the female WSS ovipositor [@Fulbright2017]. Laval development occurs rapidly with the recognizable shape becoming apparent at day three. By day five, the jaws turn brown as the eye spots begin to appear and darken. By day six or seven, the larva begins to hatch using a series of convulsive movements and emerges into the stem cavity, with full functionality achieved by the seventh day [@Ainslie1920].

### Larvae
After hatching and prior to feeding, the larvae are transparent and colorless [@Ainslie1920; @Criddle1923]. WSS larvae are distinguishable from other larvae feeding in wheat and grasses by their highly sclerotized head capsule and large body size [@Nelson1953]. As the larvae chew their way through the parenchymous and vascular tissues of the host stem, the larvae begin to develop a yellow-green color. As the larvae progress through the stem, they excrete frass; the digested plant tissue as well as metabolic breakdown products from the WSS [@Beres2011]. If multiple eggs are laid within the same stem, the largest (usually the first egg laid and hatched) will typically remain within a few weeks due to the larvae cannibalizing one another [@Criddle1923; @Seamans1944; @Buteler2015]. It remains unknown whether the consumption of other eggs and larvae is the result of intentional cannibalism or the indirect result of indiscriminate feeding [@Buteler2015]. As the host plant physiologically matures, and the larvae reaches its fifth instar, the insect is cued to travel to the base of stem and prepare for obligatory diapause [@Beres2011]. The descending movement of larvae and development of the hibernaculum are triggered by the host plant senescence. As the stem walls becoming increasingly transparent, they allow increased amounts of visible and infrared light into the stem [@Holmes1977; @Beres2011]. Once the larvae reach the base of the stem, they construct an interior V-shaped groove, making the stem weak enough to lodge due to gravity or wind [@Criddle1923]. The larvae then plug the remaining stub with frass to insulate the hibernaculum during diapause [@Holmes1960]. The larval stage of development can be as short as 30 days before entering diapause. Obligatory diapause lasts a minimum of 90 days [@Salt1947]. After the larvae are exposed to temperatures less than 10C for 90 days, diapause is complete, and pupation can begin. Larvae can reenter diapause if temperatures approach 35C [@Salt1947] or if conditions are abnormally dry [@Holmes1978] shortly after termination of diapause.

### Pupae
Once diapause is complete, pupal development begins. Pupation lasts for 21 days, beginning in early- to mid-May [@Criddle1923]. Initially, the pupa is milky-white in color [@Ainslie1920], gradually changing to bright white as the wings develop [@Fulbright2017]. Additionally, as development continues, characteristic brown and black eye spots appear on the surface of the pupae. Once pupation is complete, the adult sawfly will chew through either the frass plug or the side of the stub, and emerge ready for flight [@Ainslie1920; @Holmes1960].

## Damage Caused by Wheat Stem Sawfly
Wheat stem sawfly larvae travel throughout the stem of the host plant, feeding on the parenchymal tissue and chewing through the nodes. As a result, vascular translocation of water and nutrients within the plant is impaired [@Macedo2007] and causes a notable reduction yield and quality of the grain produced [@Seamans1944; @Holmes1977; @Morrill1992]. In addition, the weakening of the stem due to WSS causes lodging prior to harvest, exacerbated by larvae stem girdling. Stem lodging causes a reduction in stem harvest efficiently [@Ainslie1920], with yield lost due to unrecoverable lodged stem [@Beres2007]. Additionally, the reduction in stubble residues decreases snow retention which increases soil erosion and reduces soil moisture content [@Keren2015].

## Management of Wheat Stem Sawfly
### Chemical Control
Because WSS are stem mining pests and spend most of their lives within the host plant stem, the use of traditional pest control methods such as insecticides have proven to be ineffective. The only life stage of WSS that takes place outside of the stem, the adult, is not targetable via chemical treatment due to adult emergence and oviposition taking place over many weeks [@Knodel2010; @Beres2011].  WSS infestation chemical control methods were first tested as early as 1949 [@Wallace1962]. After using six different organochlorine and organophosphate insecticidal dust applied via a powder duster [@Munro1949] the authors found “unsatisfactory results that did not control” WSS populations or limit infestations. In 1963, it was reported that a 73% control of WSS infestation could be achieved via granular heptachlor applied in the furrow with the wheat seed [@Wallace1962]. However, the findings of these tests were not able to be replicable in all scenarios of WSS infestation [@Holmes1963]. Heptachlor required light sawfly infestation – restricted to the lower internodes – to be effective. When infestation rates were high, and oviposition occurred in the upper internodes, the researchers saw little to no control. Similar results were found when testing with foliar insecticides, findings showing little differences in their effect when compared to controls [@Blodgett1996]. While the Montana Department of Agriculture has authorized the use of systemic organophosphate insecticide phorate (Thimet®) for use controlling WSS infestation in winter and spring wheat cultivars, it’s use is not considered environmentally friendly due to high toxicity to mammals, birds, and fish. Additionally, a preharvest interval of 85 days complicates harvest scheduling (Ginsberg, 2003; Mahajan et al., 2006), while the high toxicity may have potential adverse effects on beneficial parasitoid populations [@Varella2016].

### Cultural Control
#### Host Plant Resistance
Mechanisms like antixenosis, antibiosis, and tolerance give a plant the ability to resist insect pests [@Horber1980; @Wilson1987]. Antixenosis allows the host plant to trigger resistance by inducing non-preference in herbivores when they attempt to feed or oviposit. Recent work done by Achhami et al. (2020) found that Barely cultivars that closely resemble wheat show the greatest frequency of WSS oviposition. These findings suggest that semiochemicals produced by these barley cultivars, such as (Z)-3-hexenyl acetate [@Piesik2008; @Buteler2009] and (Z)-β-ocimene [@Buteler2012], are likely similar to those produced by wheat [@Achhami2020]. 
Antibiosis causes the improper development of herbivores, impacting survivability and reproduction. Tolerance allows the host plant to withstand infestation and overcome injuries [@VanEmden1991; @VanZoeren2017]. Much of the research done on WSS resistant wheat cultivars focuses on solid stem trait and pith expression [@Platt1949; @Varella2016; @Varella2017]. Solid stem varieties are more resistant to infestation, reducing WSS oviposition and increasing egg mortality [@Holmes1960; @Varella2016]. The increased pith expression also disrupts larvae movement within the stem, leading to higher larval mortality due to reduced feeding as well as smaller and less reproductively successful adults [@Wallace1966]. In addition, greater pith expression increases stem strength, reducing losses due to lodged stems [@Delaney2010]. Additional factors including late planting and shorter crop duration, have been suggested as potential factors complicating antixenosis, leading to increased WSS oviposition and infestation [@Achhami2020]. 
The solid-stemmed trait has been developed in numerous spring and some winter wheat cultivars [@Beres2011]. While the use of solid-stemmed cultivars is most often recommended for WSS management [@Beres2011], hollow-stemmed cultivars were found to exhibit greater parasitism when compared to solid-stemmed varieties [@Holmes1963]. It was previously suggested by Rand et al. (2012)  that the thicker pith found in solid-stemmed cultivars makes sensing WSS larvae vibrations [@Mankin2004] more difficult for female parasitoids. The inability to locate WSS hosts would then cause ineffective parasitism.

#### Tillage
Tillage is a potential control method for the WSS. Larvae are protected against both water loss and freezing temperatures by the soil surrounding the wheat stubble where the WSS overwinters. Tilling the soil exposes the wheat stubble to the soil surface, conditions which kill the hibernating larvae [@Callenback1944]. Tillage of the WSS has been shown to be effective in decreasing the pest’s populations in both spring and winter wheat cultivars. Upwards of 90% mortality has been shown in North Dakota after fall tillage [@Weiss1987]. Many factors influence the effectiveness of tilling, including adequate exposure of wheat stubs on the soil surface and appropriate timing [@Holmes1956; @Morrill1992; @Weiss1992]. While the variability in effectiveness exists, some growers do consider it a crucial management tool [@Blodgett1996; @Lesieur2016].

While effective in killing WSS larvae, there are negative impacts caused by this practice. Water loss, soil compaction, and erosion have all been shown to be dramatically increased by tilling activities [@Phillips1980; @Lal1991]. In addition, tilling practices negatively impact beneficial populations of parasitoid wasps that target the WSS.  WSS parasitoids overwinter higher in the stem compared to at the base like WSS [@Morrill1998; @Runyon2002]. By displacing soil when tilling, the upper portion of the plant is buried under the soil, preventing the parasitoid wasps from emerging in the spring [@Runyon2002; @Weaver2004].

#### Plant Density and Spacing
Historically, it has been observed that wheat fields with wider row spacing (12 inches) have greater WSS cutting when compared to fields with reduced spacing (3 to 6 inches). Additionally, fields planted at greater seed density with solid stem cultivars exhibit lower WSS cutting compared to reduced seed density [@Luginbill1958]. The reason for this relationship between decreased density and increased sawfly infestation is due to the moisture availability and height of the wheat stems. At higher densities, each stem has access to less water, in turn leading to the stem to develop a thicker pith – more solid stemmed. As discussed above, there is a negative linear relationship between an increase in the solid-stem characteristic and infestation [@Kemp1934].

#### Crop Rotation
Continually planting wheat year after year creates ideal conditions for the WSS [@Callenbach1945]. Rotating wheat with more resistant wheat varieties as well as with oats, or soybeans (crops that are less or not susceptible to WSS infestation) is recommended to decrease the year after year WSS infestation rates [@Lin2014]. The notable downside to utilizing crop rotation in a matter that is effective is the economic downside. Alternate-year, summer fallow production systems are typical for wheat growing practices across much of the northern great plains of the United States [@Weiss1992; @Troeh1999].

#### Trap Crops
Trap cropping utilized non-economically important plantings to help reduce populations or block movement of pests into valuable crop areas [@Hokkanen1991]. Effective trap crops of the WSS should have appealing characteristics to the WSS to attract the pest to the trap planting, and away from the wheat [@Farstad1940]. Historically, planted trap strips were used by attracting sawflies and destroying the strip at the end of the season, eliminating the larvae [@Morrill1992]. However, utilizing trap plantings of highly attractive wheat varieties has been shown to hold limited viability, as growers are not fond of destroying planted grain at the end of the growing season [@Goosey1999].

Alternative utilization of WSS traps entail planting and/or maintaining strips of smooth brome grass (Bromus inermis Leyss.) or other WSS host grasses [@Farstad1940]. Smooth brome in particular holds potential as it is a poor host to WSS larvae [@Bhandari2020], yet WSS adults continue to oviposit the stems. This relationship of greater oviposition with increased mortality may lead to decreases in overall WSS populations.

#### Delayed Planting & Swathing
Wheat planted later in the growing season avoids much of the damage caused by the WSS as the wheat reaches stem elongation after many of the adult sawflies have already laid their eggs [@Jacobson1952]. The problem with delayed planting is that the practice is known to make less efficient use the earl season moisture in the soil [@Morrill1996]. Growers may need to weigh the tradeoffs in delayed planting to reduce losses in areas with a history of severe infestation. Swathing, also known as windrowing, is the practice of cutting the wheat to accelerate the drying process and decrease the losses caused by WSS cutting [@Holmes1965; @Carlson2019; @McCullough2020]. By cutting the wheat and letting it dry prior to sawfly cutting, swathing works by reducing losses due to lodging [@Ainslie1920].

#### Biological Control Methods
The use of biocontrol in limiting WSS populations has been well studied with variable results. There are currently three types of organisms that have been reported as biocontrol agents of the WSS in wheat fields: Soil fungal pathogens, the predatory beetle Phyllobaenus dubius, and two parasitic Hymenoptera.

#### Soil Fungal Pathogens
Fungal pathogens are found throughout both the plant and animal kingdoms. Several fungi have been shown to negatively impact WSS larvae infestation within wheat stems. While highly lethal to sawfly larvae, F. graminearum, F. acuminatum, F. equiseti, and F. avenaceum also cause crown rot in the host plant. The trade-off between WSS control and crown rot may limit Fusarium spp. usefulness to WSS management. [@Wenda2009]. Findings by Tangtrakulwanich et al. (2014) found that the fungi B. bassiana and M. brunneum effectively reduced damage caused by WSS. Both fungi kill insects by infecting via spore contact or by insect consumption of treated plant tissue [@Portman2018].

#### Predatory Beetle
Phyllobaenus dubius (Wolcott) has been observed emerging from wheat stubble and within dissected wheat stems in Montana [@Morrill2001]. Further research is required on this coleopteran to determine its prospects going forward as an effective WSS management tool.

#### Parasitoids of the Wheat Stem Sawfly
Two parasitoid wasps are known target the wheat stem sawfly in wheat fields. The congeneric parasitoids, Bracon cephi (Gahan) and B. lissogaster Muesebeck, are the most effective agents of WSS biological control [@Runyon2001]. B. cephi and B. lissogaster, as distinguished by Runyon et al. (2001), were initially limited in their ability to parasitize WSS larvae within wheat stems. The wheat stem sawfly, originally known as the western grass-stem sawfly, was originally only found within native grasses throughout North America, existing in alongside B. cephi and B. lissogaster [@Ainslie1920; Cárcamo et al., 2012). However, as agriculture increased and native grasses were replaced by large monoculture landscapes, the WSS moved into wheat [@Farstad1940; @Davis1955; Holmes1978] while the parasitoids were slower to adapt [@Morrill1998].

## Biology of *Bracon cephi* and *Bracon lissogaster*
### Adult
Both *B. cephi* and *B. lissogaster* are protelean host-specific ectoparasitoids with two generations per year (bivoltine) [@Nelson1953; @Somsen1956; @Holmes1963]. They are ectoparasitoids and idiobionts, living on the integument of their hosts and immediately paralyzing their hosts in location and stage it was when it was attacked [@Nelson1953; @Runyon2002; @Weaver2004]. Male parasitoids are approximately 3.5 mm in length, while females are 4.1 mm. They are dark red, with black antennae, eyes, ovipositor and ocelli [@Gahan1918]. B. lissogaster and B. cephi are distinguishable by their head color and unique metasoma surface texture [@Runyon2001]. Adults emerge after completing a pupal stage within the host stem by chewing a hole through the stem wall. 
In Montana, the first generations are present from late June to late July, while the second generations are active in August and September [@Somsen1956]. Adult female B. cephi and B. lissogaster will locate a host stem using a combination of olfactory and volatile cues. The wasps will then locate a host sawfly larva using acoustic cues [@Mankin2004] and use their ovipositor first paralyze, then deposit 1-4 eggs next to the paralyzed larva.

### Egg
*B. cephi* and *B. lissogaster* eggs are on average 0.86 mm in length, exhibiting an elongated shape and pale-yellow color. Females lay one to three eggs on the WSS larval integument or adjacent to the larva. The eggs develop for one to three days before hatching [@Nelson1953; @Somsen1956].

### Larva
Parasitoid larvae are 1mm in length, with a translucent hue that turns increasingly brown as the larva progresses through five instar stages [@Nelson1953]. The newly hatched larvae have no appendages, save for antennae, and they exhibit highly sclerotized mouthparts and well-developed mandibles. After hatching, the B. lissogaster larva attach themselves to the integument of the host larva and immediately begin feeding [@Somsen1956]. When feeding is complete, little remains of the host but the integument. Following feeding, the larva will spin a lightly brown colored cocoon over the course of several days and attach to the inside wall of the host plant stem. The larval stage takes an average of ten days between initial hatch and cocoon completion. Because there are two generations per year, the early summer offspring will hatch and emerge from the host stem. Offspring laid as a part of the second generation will overwinter within the cocoons and emerge in the spring. [@Nelson1953]

### WSS and Parasitoid Interactions with Smooth Brome
Bhandari (2020) has shown that oviposition occurs with similar frequency in wheat as it does in smooth brome, however, we see much higher rates of larval mortality within smooth brome stems. This has led researchers to hypothesize about the feasibility of using smooth brome as a trap cop [@mendoza2006; @Lesieur2016]. While feasible as a feed crop, as it holds considerably less economic value compared to wheat or other high value crops [@Criddle1922]. Smooth brome may be usable in trap strips along the edge of wheat fields [@Hokkanen1991]. While a poor host for the wheat stem sawfly, surviving immature larvae found within brome stems can be parasitized by B. cephi and B. lissogaster. This combination of factors may indicate that smooth brome can function as a WSS sink and a parasitoid source, with very few WSS emerging from smooth brome patches, while also supporting populations of parasitoids that can then parasitize WSS in adjacent wheat fields.

## Plant Semiochemicals

Plants interact with other organisms through a complex system of chemical signals known as Semiochemicals. Semiochemicals are compounds or mixtures of compounds that are released from one organism that evoke a physiological or behavior response in another organism [@Agelopoulos1999]. Semiochemicals play a large role in both insect-insect and plant-insect interactions and are essential in insects locating feedings sites, mating sites, egg-laying sites, and/or refugia [@Prokopy1984].

While it is known that these chemicals play an important role in communication between plants, insects, and other organisms, there is still debate about the function of many of them [@Reddy2004; @Voelckel2018].

### Semiochemicals in Plant-Insect Systems

Within plant-insect systems, plants respond to insect herbivory by synthesizing and releasing volatile compounds and providing important signals to biocontrol agents to localize potential host insects [@Sharma2019]. In addition, semiochemicals provide localized signatures that allow insects to detect food sources, find mates, and locate preferential oviposition sites [@Norin2007]. Plants typically produce semiochemicals when defending themselves from herbivorous arthropods [Guerrieri1999] or pathogen attacks [@Smart2014]. These responses can be grouped into two categories, direct and indirect responses. Direct responses are compounds that the plant produces that have a direct effect on the herbivore or pathogen [@Metcalf1998]. Indirect responses are the compounds produced that have no impact directly on the pest, but instead may attract parasitoids, predators, or other biological control agents [@Mayland2000; @Colazza2004].

# Limitations And Key Assumptions
## Field Study
i.	Time limits the number field sites that we can collect samples from throughout the summer.  
a.	Study area limited to western Montana – three regions – Big Sandy, Moccasin, and Amsterdam.  
ii.	Our study will only take place over the course of 2-3 summers, thus, we are unable to make broader claims surrounding long term patterns of WSS and parasitoid interactions.   
Green House Samples and Volatile Collection
iii.	We will assume that plants grown in the green house will be accurate and reasonable biological model representation of plants present in the field.  
iv.	Volatiles will be collected from a maximum of 12 plants a day due to constraints with the collection system.

# Contributions to Knowledge of Potential Outcomes
## Potential outcomes
### Volatiles
We hope to observe a discernable difference in the volatile quantities produced by both infested smooth brome and winter wheat.

# CHAPTERS
## METHODS
### Central and Northern Montana Field Sites
#### Field Selection & Data Collection
To assess C cinctus infestation, larval mortality, and parasitoid prevalence within B. inermis and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production. Samples were collected in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA. B. inermis sites were set up as polygons along the edge of adjoining wheat fields measuring approximately 100 meters in area. Four collection squares of 1ft x 1ft were randomly partitioned within each polygon during both collection events each year. All stems within each 1 x 1 ft square were collected. The stems were then processed and split in the laboratory to assess infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting.

#### Statistical Analysis
Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions. ‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs awere used as the response variable to assess the differences in parameters among the sampling sites. ANOVA was used to caculate P-values using F-distribution to understand overall significant differences in mean C. cinctus infestation and parasitism among polygons. To assess the profile of C. cintus infestation, paraisitms, and larval presence by node, a linear model was fit for each sampling site. Further assessment was done on the resonpses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.

## Controlled *C. cinctus* infestation of *B. inermis*
### Plant Selection and Field Cages
Assessment of C. cinctus infestation and mortality within B. inermis were assessed using a 34 x 60 ft plot at the Arthur H. Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT. The cage structure was built using 1-inch PVC piping with the netting made using 530 Amber Lumite Screen (BioQuip Products, LLC). Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three. The cages were removed from the plot after 3 weeks of sawfly release.

### Insects

Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of C. cinctus infestation and cutting the year prior. Cut stubble, which contained C. cinctus larvae in diapause, were kept refrigerated between -2°C and 3°C for >100 days as required to complete obligatory larval diapause. As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA). Stubs were added to cages once B. inermis stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).

### Data Collection
Stems were collected from each cage site in late August. All stems from each 6 x 3 ft cage section were collected. Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.

### Statistical Analysis
Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions. ‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs were used as the response variable to assess the differences in parameters among the sampling sites. ANOVA was used to calculate P-values using F-distribution to understand overall significant differences in mean C. cinctus infestation and parasitism among polygons. To assess the profile of C. cinctus infestation, parasitism, and larval presence by node, a linear model was fit for each sampling site. Further assessment was done on the responses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.

## Organic Volatile Collection from C. cinctus Infested B. inermis and Winter Wheat
### Insects
Wheat stubble collected the year prior to each experiment were used to rear adult WSS. Wheat stubble was collected from Big Sandy (48.17754 N, -110.11406 E) and Three Forks (46.020216 N, -111.586582 E), MT, USA. Stubble was collected from fields that experienced high levels of WSS infestation and cutting the year prior. Cut stubble, which contained WSS larvae in diapause, were kept in a refrigerator between 0-4 C for >100 days as required to complete obligatory larval diapause. After completing diapause, immature WSS were kept at room temperature between 22-27 C for 4-5 weeks in sealed 70 cm x 35 cm x 20 cm plastic Tupperware bins to allow for pupation and emergence. Once emergence began, adults were collected daily and sorted in 2-liter glass mason jars sealed with filter paper to allow for airflow. Jars containing adults were refrigerated at 6-10C until they were needed for experiments. Unused adults were discarded after 48 hours of refrigeration.

### Plants
Smooth brome was grown from transplanted plants harvested from the Arthur H. Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W).  Immediately following collection, plants were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc, Tangent, OR) and moved into the greenhouse. Wheat was grown using ‘Yellowstone’ winter wheat cultivar seed stock, started in 1 x 8 inch conical pots. The soil used consisted of MSU Plant Growth Center soil mix (equal parts sterilized Bozeman silt loam soil and washed concrete sand with Canadian sphagnum peat moss incorporated) and Sunshine Mix 1 (Canadian sphagnum peat moss, vermiculite, perlite, and Dolomite lime) in a 1:1 ratio. Fertilization was undergone once weekly using Jack’s Professional® Water-Soluble Fertilizer (20-20-20) (J.R. Peters Inc., Allentown, PA, USA). The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22  2C and 20 – 40 % RH. After the plants reached Zadoks 12, they were moved into vernalization storage for 6 weeks to experience the length of dormancy needed for successful flower development.  After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc conical pots.

### Volatile Collection
Volatile compounds were collected from intact and healthy smooth brome and Yellowstone winter wheat. Collection was undergone on plants at Zadoks 49, for a period of six hours from 09:00 to 15:00. Collections were undergone three times a week, with twelve plants collected from each time. Each trial consisted of six smooth brome plants and six winter wheat, with four of each being infested, while the other two were controls.

Volatiles were collected from twelve plants per day using glass VCC’s that were 80 cm long and 4 cm in diameter. The top end of each VCC funneled into a threaded port that was 10 cm in diameter. Threaded onto the top of each VCC was an eight-port manifold adapter. Two glass filters (traps) (6.35 mm diameter x 76 mm length; Analytical Research Systems, Gainesville, FL) with 30 mg of Super-Q absorbent (Alltech Associates, Deerfield, IL) were inserted into two of the adapter ports. The remaining six collection ports were sealed to prevent ambient air exchange and pressure loss. One of the traps was used to collected extraneous volatiles from the initial experiment set up for the first ten minutes, whereas the other trap was used to collect volatile compounds emanating from plants beginning ten minutes after the system was turned on. An air source was connected to the base of each collection chamber with supply from charcoal air filters. A vacuum pump was used to maintain air pressure and flow rate (humidified air at 1.0 liter/min) within the collection system. A small piece of cotton encompassed by a Teflon® guillotine was used at the base of each plant to tightly seal the base of the system and prevent external air from entering. Laboratory tape was used to seal the edges of the guillotine and prevent contamination by ambient and soil derived volatiles.

Volatile collection and successive GC-MS analysis were conducted following methods described in Weaver et al. (2009). Glass volatile collection traps were eluted using 200 µL of dichloromethane into 2 mL screw top vials (Agilent Technologies, Inc., Santa Clara, CA), followed by a slow release of nitrogen to further clear the volatile trap. 10 µL of nonyl acetate solution was added to the eluted solution as internal standard to quantify compounds. Samples were then subject to gas chromatography (GC) on a J&W HP – 5ms; 30m x 0.25 µm film thickness column (Agilent Technologies, Inc., Santa Clara, CA). The GC unit (Agilent 6890; Agilent Technologies, Santa Clara, CA) was paired with a mass spectrometer (MS, Agilent 5973 instrument). Samples were injected into the column in pulse splitless mode, with the initial pressure of 82.7 kPa/m.

## PARASITOIDS
Parasitoids B. cephi and B. lissogaster were collected from wheat fields in Three Forks, MT (46.020216 N, -111.586582 E). Wheat stem residue from fields that experienced high levels of C. cinctus infestation were collected and stored inside 170-liter plastic barrel liners. Filled liners were stored in a cold storage room at 4C until mid or late spring of the following year. After the storage time exceeded the mandatory duration needed for obligatory diapause, the plastic barrel liner was removed from cold storage and the residue was placed inside 170 L trash barrels at room temperature and watered to encourage emergence. Collected parasitoids were kept at 6-10C until used.

### Plants
Smooth brome was grown from transplanted stems harvested from the Arthur H. Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.  Immediately following collection, stems were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc, Tangent, OR).
Winter wheat plants were grown using Yellowstone Winter Wheat cultivar seed stock, started in 1 x 8 inch conical pots. The soil used consisted of MSU Plant Growth Center soil mix (equal parts sterilized Bozeman silt loam soil and washed concrete sand with Canadian sphagnum peat moss incorporated) and Sunshine Mix 1 (Canadian sphagnum peat moss, vermiculite, perlite, and Dolomite lime) in a 1:1 ratio. Fertilization was undergone once weekly using Jack’s Professional® Water-Soluble Fertilizer (20-20-20) (J.R. Peters Inc., Allentown, PA, USA). The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22  2C and 20 – 40 % RH. After the plants reached Zadoks 12, they were moved into vernalization storage for 6 weeks to experience the length of dormancy needed for successful flower development.  After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc conical pots.

## Behavioral Assessment of Parasitoids in Response to B. inermis Organic Volatile Compounds
### Insects
### Electroantennography
### Y- Tube Olfactometer Bioassay
Y-tube bioassays were conducted within the chemical ecology and insect behavior laboratory at the Plant Growth Center at Montana State University. Bioassays were undergone to determine behavioral choice preference response of individual female WSS to an airstream containing volatiles originating from smooth brome and a blank control. Bioassays were conducted using plants at growth stage Zadoks 49. The Y-tube system (Analytical Rsearch Systems, Micanopy, FL) used was similar as described in a previous study (Daisy et al., 2002) (Figure 1). A single plant was enclosed in a volatile collection chamber (VCC) (100 mm diameter, 254 mm length). Input air was charcoal-filtered and humidified, regulated by a diaphragm air pump that delivered air at a rate of 0.2 L min-1. Air was moved into the VCC through Teflon tubing connected with a 24/410 threaded glass joint. The bottom end of the VCC enclosed each plant and was tightly sealed by a Teflon guillotine (diameter 955 mm, 15 mm center opening), encircling the stem to prevent air leakage. The olfactory system used a Y-shaped glass tube that had two arms connected to the lower port of each VCC by Teflon tubing (diameter 0.64 cm), providing the airstream source from the plant. The VCC contained plant was illuminated by an enhanced spectrum LED grow light (sunshine Systems Grow UFO Light SS-Gu90w, Led Lighting4 Less, Paradise, CA) for one hour before the experiment to make the plant photosynthetically active.

<!-- Figure 1 y tube diagram -->
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Image Cropping Example</title>
<style>
    .image-container {
        width: auto; /* Set the width of the container */
        height: auto; /* Set the height of the container */
        overflow: hidden; /* Hide overflow */
    }
    .image-container img {
        width: 100%; /* Make the image fill the container horizontally */
        height: auto; /* Maintain aspect ratio */
        object-fit: cover;
        object-position: center;
    }
    .caption {
        text-align: left; /* Center the caption */
        font-style: italic; /* Style the caption */
        margin-top: 10px; /* Add some space between the image and the caption */
    }
</style>
</head>
<body>

<figure class="image-container">
    <img src="../img/proposal/ytube_diagram.jpg" alt="Your Image">
    <figcaption class="caption">Figure 1: Y-tube olfactory bioassay as constructed by Daisy et al. (2002). Used to measure insect behavior in the presence of volatile stimulus.</figcaption>
</figure>

</body>
</html>

```

Figure 1: Y-tube olfactory bioassay as constructed by Daisy et al. (2002). Used to measure insect behavior in the presence of volatile stimulus.

The Y-tube was made of bifurcated Corning® glass, measuring in at 28-mm diameter x 300 m length. The Y shape branched at an angle of 120, with each branch serving as an arm of the olfactometer. The stem of each Y-tube was 20 cm long. At the bifurcation, each branch extended 4 cm from the junction before ending in 10 cm parallel arms. The Y-tube was positioned horizontally inside a black box made of poster board (46 x 32 x 101.5 cm), designed to prevent the entry of unwanted ambient light. The black box had two holes on the upstream sie of the olfactometer, allowing a fiber optic illuminator (T-Q/FOI-1, Techni Quip Corp, El Segundo, CA) and Teflon tubing to pass through. The fiber optic illuminator was used to deliver light to both arms equally from the center of the bifurcation, while the Teflon tubing connected the VCC to the arms of the Y-tube. One arm of the Y-tube delivered volatiles from the enclosed plant while the other delivered filtered, pure air as a control. The movement of adult WSS is positively phototactic and the horizontally placed Y-tube was positioned such that the light distribution towards both arms was balanced to avoid bias. The arms of the Y-tube were oriented to minimize directional bias and the position of the plants was alternated between runs. 
For each assay, a single naïve female B. cephi or B. lissogaster parasitoid was introduced 1 cm into the unbranched end of the Y-tube. A 28-cm wire was used to facilitate parasitoid movement from the unbranched stem of the glass Y-tube to the test junction between the two upper arms. The order of individual test insects from a pool of naïve females was randomly assigned for each bioassay, and test plants were chosen randomly from the greenhouse culture before each experiment. All glassware was cleaned using dish soap and warm water, followed by solvent rinses with acetone and hexane. Cleaned and rinsed glassware was baked at 110C in a glassware oven for 24 hours before use. New plants were used for each experiment. 
	Each individual parasitoid was allowed three minutes to respond to either airstream source. Once a positive response was determined to either source, the trail was ended. If no choice was made after five minutes, the trail was ended and the result was recorded as no response.

## Volatile Collection
### Sawfly Infestation
Winter wheat plants were deemed ready for infestation once they reached Zadoks 49. Both smooth brome and winter wheat plants were infested by placing a transparent plastic cylinder (60 x 3.8 cm) over the main stem of the plant. A 55-60 cm bamboo stick was added within each tube to help prevent toppling of the cylinder. Each cage had three holes, as well as an open top, covered with fine mesh cloth to allow for air circulation. Four females and two male WSS, newly emerged, were introduced into each tube via a small 1 cm opening. After which, the hole was plugged using a cotton ball covered in mesh cloth. Damp soil was mounded around the base of each tube to prevent insect escapees. Adult sawflies were left within infestation tubes for three days, after which the tubes were removed, and any surviving insects, as well as cadavers, were removed. After infestation, the plants were returned to normal water and fertilization schedules.

Infestation was undertaken three days a week, for five weeks. Each replicate included 6 ‘Yellowstone’ winter wheat plants at Zadoks 49 and 6 smooth brome plants exhibiting three nodes. For each replicate, three wheat and three smooth brome plants were infested, with three of each left uninfested. Infestation tubes were placed around all twelve plants to maintain control.

### Volatile Compound Collection
Volatile compounds were collected from intact and healthy smooth brome and ‘Yellowstone’ winter wheat. Collection was undergone on plants at Zadoks 49, for a period of six hours from 10:00 to 16:00. Collections were undergone three times a week, with twelve plants collected from each time. Each trail consisted of six smooth brome plants and six winter wheat, with three of each being infested, while the other three were a control.

Volatiles were collected from twelve plants per day using glass VCC’s that were 80 cm long and 4 cm in diameter. Volatile collection and successive GC-MS analysis were conducted following methods described in Weaver et al. (2009). The top end of each VCC funneled into a threaded port that was 10 cm in diameter. Threaded onto the top of each VCC was an eight-port manifold adapter. Two glass filters (traps) (6.35 mm diameter x 76 mm length; Analytical Research Systems, Gainesville, FL) with 30 mg of Super-Q absorbent (Alltech Associates, Deerfield, IL) were inserted into two of the adapter ports. The remaining six collection ports were sealed to prevent ambient air exchange and pressure loss. One of the traps was used to collected extraneous volatiles from the initial experiment set up for the first ten minutes, whereas the other trap was used to collect volatile compounds emanating from plants beginning ten minutes after the system was turned on. A vacuum pump was used to maintain air pressure and flow rate (humidified air at 1.0 liter/min) within the collection system. A Teflon guillotine was used at the base of each plant to tightly seal the base of the system and prevent external air from entering. Each plant stem was wrapped with a small piece of cotton prior to the application of the guillotine to further seal the base of the collection system and prevent contamination by ambient and soil derived volatiles.

### Volatile Elution
Volatile compounds were eluted from the glass traps using 200 L aliquot dichloromethane. Trapped volatiles were eluted slowly by adding dichloromethane and further clearing by using a slow release of nitrogen gas. The eluted samples were collected in a glass insert held within a 1.5-mL crimped top glass vial. 10 L of nonyl acetate were then added to the eluted samples as internal standard to quantify compound concentrations. The samples were processed using gas chromatography (GC) on a HP – 5MS; 30m x 0.25mm, 0.25 m film thickness column; (J and W Scientific, Folsom, CA). In order to analyze, the GC instrument (Agilent 6890; Agilent Technologies, Santa Clara, CA) was coupled to a mass spectrometer (MS, Agilent 5973 instrument). Each sample was injected using the automated system in pulse splitless mode, with the initial pressure set to 82.7 kPa per minute. The inlet temperature of the GC was set to 250C while the column temperature was 50C for 4 minutes. The temperature was set to increase at a rate of 5C per minute until it reached 160C, after which it increased to a rate of 25C per minute until it reached 280C. The temperature of the transfer line leading to the mass selective detector (MSD) was set at 300C. Helium was used as a carrier gas to maintain the flow rate of samples within the column at 1.2 mL/min. The MSD was set in ‘SCAN’ mode running from 50 – 300 m/z.

### Volatile Analysis
The compounds collected from both the smooth brome and wheat samples were identified by comparing mass spectra and retention times using the National Institute of Standard and Technology (NIST) library. Peaks and volatiles were analyzed using MSD ChemStation (Agilent Technologies, Inc., Santa Clara, CA). The data obtained from the mass-spectrometer was transferred to a Microsoft® Excel spreadsheet where the compounds were quantified by comparison with the internal standard peak area. The amount obtained for each compound was then transformed to the amount produced in six hours. The true quantity of compounds was calculated relative to the internal standard.

### Electroantennogram
Specimens were prepared by excising the head from the body using fine micro-dissection scissors. For each specimen, a micro-capillarity tube was pulled in half using a pipette puller (Model PP-830, Narishige Scientific Instrument Lab, Tokyo, Japan). Each half was cut to the appropriate length and angle, then filled with saline solution. The parasitoid head was mounted on the broken end of one of the capillary tubes, then the tube was place on one of the conducting electrode wires connected to a micromanipulator, type INR-5 (Narishige MN-151). The second half of the capillary tube was place on the other electrode wire, and one insect antennae was positioned to touch the empty capillary tube, thus connecting the circuit. The amplifier transferred the information to a computer interface with an EAG version 2.7 software (Syntech NL 2001, Hilversum, Netherlands). This apparatus and method has been consistently used previously in assessing insect volatile response (@Agelopoulos1999).

A consistent stream of purified, humidified air (20 mL sec-1) was delivered onto the antennae. The test compound was delivered via airborne stimulus released from a glass pipette with filter paper treated with 10 L of the test compound. The pipette was connected to a Stimulus Controller CS-55 (Syntech NL 2010, Hilversum, The Netherlands). The equipment was programmed to deliver a 0.4 second puff of air through a Teflon tube and through the glass pipette. The electrical response of the antenna to the test compound was detected by the recording electrode and transferred to the amplifier, increasing the signal and delivering the information to the computer.
Figure 2: Electroantennogram connected to GCMS. Useful for connecting insect electrochemical response to host plant produced volatile compounds.
List of compounds used.

## Data Analysis
### Field Survey
### Greenhouse Experiments

#### Comparative Analysis of Collection of Volatiles
The concentration of volatile compounds gathered was determined as nanogram per gram per hour (ng-1g1hr1), accounting for plant biomass adjustments. Data were analyzed using R Studio version 2023.12.0+369 (R Core team 2023). Data were first transformed via centered-log ratio (CLR) transformation, a common technique for compositional data  (Brückner and Heethoff, 2017). The data underwent both Permutational Analysis of Variance (PERMANOVA) and Permutational Analysis of Multivariate Dispersions (PERMDISP). PERMANOVA was used to determine differences among identified plant volatile compounds between the two plant types (Yellowstone winter wheat, B. inermis) at two infestation levels (Infested, control). PERMDISP was used to assess the variance of volatile quantities within our different groups.

#### EAG-FID

#### Organic Volatile Collections
##### Centered Log-Ratio Transformation
The data was transformed using a centered log-ratio transformation (CLR). This method is commonly used for multivariate compositional data (Brückner and Heethoff, 2017). CLR works by taking the natural logarithm of the ratios between the values of individual components and their geometric mean across samples. This helps to mitigate spurious correlations and allows the application of standard statistical techniques to our compositional data.

##### PERMANOVA

# Citations

### 2024-03-20 qmd/litreview_test.qmd
commit for 3/20

--- 2024-03-20 qmd/litreview.qmd
@@
 author: "Jackson Strand"
-date: March 20, 2024
+date: "2024-03-20"
 output: 
@@
 link-citations: true
-biblio-style: apalike
-cls: elsevier-with-titles.csl
-# cls: council-of-science-editors-author-date.csl
+cls: council-of-science-editors-author-date.csl
 editor_options: 
-  chunk_output_type: inline
+  chunk_output_type: console
 toc: true
 ---
-
-<!-- Button to download word document version -->
-```{=html}
-<!DOCTYPE html>
-<html lang="en">
-<head>
-<meta charset="UTF-8">
-<meta name="viewport" content="width=device-width, initial-scale=1.0">
-<title>Download File Button</title>
-<style>
-/* Button styling */
-.button {
-  background-color: #2068DC; /* Header of website color (blue) */
-  border: none;
-  color: white;
-  padding: 15px 32px;
-  text-align: center;
-  text-decoration: none;
-  display: inline-block;
-  font-size: 16px;
-  margin: 4px 2px;
-  cursor: pointer;
-  border-radius: 8px;
-}
-.button:hover {
-  background-color: #3F91FB; /* Sky blue */
-}
-</style>
-</head>
-<body>
-
-<!-- Button to download the file -->
-<a href="litreview.docx" download="Jackson Strand Proposal">
-  <button class="button">Download as Word Document</button>
-</a>
-
-<script>
-// Get the current date in YYYY-MM-DD format
-function getCurrentDate() {
-  const now = new Date();
-  const year = now.getFullYear();
-  const month = String(now.getMonth() + 1).padStart(2, '0');
-  const day = String(now.getDate()).padStart(2, '0');
-  return `${year}-${month}-${day}`;
-}
-
-// Update the download attribute of the link with the current date
-document.addEventListener('DOMContentLoaded', function() {
-  const downloadLink = document.getElementById('downloadLink');
-  const currentDate = getCurrentDate();
-  downloadLink.setAttribute('download', `Jackson Strand Proposal ${currentDate}`);
-});
-</script>
-
-</body>
-</html>
-```
 
@@
 ## Research Question
-What role does smooth brome (*Bromus inermis*) play in the ecological relationship between wheat stem sawfly (*Cephus cinctus*) and wheat stem sawfly parasitoids (*Bracon cephi* and *Bracon lissogaster*) in dryland wheat production landscape within Montana?
+What role does smooth brome (Bromus inermis) play in the ecological relationship between wheat stem sawfly (Cephus cinctus) and wheat stem sawfly parasitoids (Bracon cephi and Bracon lissogaster) in dryland wheat production landscape within Montana?
 
@@
 ## Host Plants of Wheat Stem Sawfly
-Wheat stem sawfly was first found the hollow-stemmed wild grasses such as Agropyron and Elymus in Colorado in 1872 [@Ainslie1920].
+Wheat stem sawfly was first found the hollow-stemmed wild grasses such as Agropyron and Elymus in Colorado in 1872 (Ainslie, 1920).
 Soon thereafter it was found in wheat in 1895 in Manitoba, Canada.
-First referred to as the “western-grass stem sawfly”, WSS was found with native Montana grasses including western wheatgrass (Pascopyrun smithii) and slender wheat grass (Elymus trachycaulus) [@Ainslie1920].
-As settlers began moving west and small grain cereal crop production increased on the landscape, the native grasses that WSS foraged upon became quite rare [@Ainslie1929].
-WSS was found damaging spring wheat for the first time in 1895 in Manitoba and Saskatchewan [@Wallace1966].
-With the diminishment of its natural hosts, WSS has adapted to infest the exotic non-cultivated and cultivated grasses now present [@Criddle1923].
-WSS infestation has been shown in many species native to Montana including western wheatgrass (Pascopyrun smithii), slender wheat grass (Elymus trachycaulus) and bluebunch wheatgrass, Pseudoroegneria spicata [@Cockrell2021].
+First referred to as the “western-grass stem sawfly”, WSS was found with native Montana grasses including western wheatgrass (Pascopyrun smithii) and slender wheat grass (Elymus trachycaulus) (Ainslie, 1920).
+As settlers began moving west and small grain cereal crop production increased on the landscape, the native grasses that WSS foraged upon became quite rare (Ainslie, 1929).
+WSS was found damaging spring wheat for the first time in 1895 in Manitoba and Saskatchewan (Wallace and McNeal, 1966).
+With the diminishment of its natural hosts, WSS has adapted to infest the exotic non-cultivated and cultivated grasses now present (Criddle, 1923).
+WSS infestation has been shown in many species native to Montana including western wheatgrass (Pascopyrun smithii), slender wheat grass (Elymus trachycaulus) and bluebunch wheatgrass, Pseudoroegneria spicata (Cockrell et al., 2021).
 
-Smooth brome (*Bromus inermis* L.) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 from Hungary and Russia for soil conservation and use as livestock feed [@Salesman2011].
+Smooth brome (Bromus inermis L.) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 from Hungary and Russia for soil conservation and use as livestock feed (Salesman and Jessica, 2011).
 Throughout the early 19th and 20th century, it was used throughout much of the upper half of North America.
-It is a perennial grass, and more drought resistant than other exotic grasses [@Mapfumo2002].
+It is a perennial grass, and more drought resistant than other exotic grasses (Mapfumo et al., 2002).
 Due to its sod forming nature and large root system, it is extensively used to promote soil retention along highways (USDA-NRCS 2010).
 
-Smooth brome was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Farstad1945].
-However, due to its propensity to form monocultures and outcompete many native grasses, smooth brome was labeled by many as invasive [@Carlson1985; @Willson2000; @Dillemuth2008].
-Smooth brome populations play a considerable role in altering native herbivore population dynamics and movement on the prairies of North America [@Dillemuth2008].
+Smooth brome was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices (Farstad and Jacobson, 1945).
+However, due to its propensity to form monocultures and outcompete many native grasses, smooth brome was labeled by many as invasive (Carlson and Newell, 1985; Willson and Stubbendieck, 2000; Dillemuth et al., 2008).
+Smooth brome populations play a considerable role in altering native herbivore population dynamics and movement on the prairies of North America (Dillemuth et al., 2008).
 
@@
 Females are larger than males and a visible long, dark, ovipositor.
-Both male and female adults are weak fliers, rarely flying long distances at one time [@Criddle1923].
-They can be found on stems with their heads downward and legs aligned with their body, encircling the stem [@Ainslie1920; @Criddle1923].
+Both male and female adults are weak fliers, rarely flying long distances at one time (Criddle, 1923).
+They can be found on stems with their heads downward and legs aligned with their body, encircling the stem (Ainslie, 1920; Criddle, 1923).
 WSS are univoltine, having only one generation per year.
-Males are haploid (9 chromosomes), while females are diploid – females who do not find a mate will produce male offspring, while those who do produce female progeny [@Holmes1978].
-Fully mature adults hatch in the spring once it has sufficiently warmed and the moisture levels in the soil reach threshold levels [@Holmes1978; @mendoza2006].
-In Montana, this period begins in mid- to late-May and ends early in July [@Morrill1996; @Weaver2005].
+Males are haploid (9 chromosomes), while females are diploid – females who do not find a mate will produce male offspring, while those who do produce female progeny (Holmes, 1978).
+Fully mature adults hatch in the spring once it has sufficiently warmed and the moisture levels in the soil reach threshold levels (Holmes, 1978; Perez-Mendoza and Weaver, 2006).
+In Montana, this period begins in mid- to late-May and ends early in July (Morrill and Kushnak, 1996; Weaver et al., 2005).
 Their lifespan usually lasts around five to eight days, emerging from May until June.
-Males emerge before the females, ensuring that many of the early females to hatch will find a mate and deposit eggs that will be female [@Mackay2011].
+Males emerge before the females, ensuring that many of the early females to hatch will find a mate and deposit eggs that will be female (Mackay, 2011).
 Copulation lasts less than one minute with the male holding the back of the female with their abdomen curved down under the female ovipositor.
-Fertilized females lay eggs several hours after copulation [@Holmes1978].
+Fertilized females lay eggs several hours after copulation (Holmes, 1978).
 
@@
 After locating a suitable host stem, female WSS oviposit an egg inside the host plant lumen.
-WSS larvae hatch inside the stem and feed on the stem’s parenchyma tissue [@Beres2011].
-As the stem begins to senesce in the fall, the larvae travel to the base of the stem and makes a V-shaped groove [@Criddle1923] above a constructed protective hibernaculum where they will overwinter [@Holmes1960].
+WSS larvae hatch inside the stem and feed on the stem’s parenchyma tissue (Beres et al., 2011).
+As the stem begins to senesce in the fall, the larvae travel to the base of the stem and makes a V-shaped groove (Criddle, 1923) above a constructed protective hibernaculum where they will overwinter (Holmes and Peterson, 1960).
 As a result of this deep groove, the stem weakens and topples easily.
-The resulting lodged stems are not only challenging for growers to harvest but also exhibit decreased kernel yield [@Beres2011].
+The resulting lodged stems are not only challenging for growers to harvest but also exhibit decreased kernel yield (Beres et al., 2011).
 
 ### Egg
-Newly laid eggs are crescent in shape, milky-white in color, and range between 1 mm to 1.25 mm in length and 0.33 to 0.42 mm wide [@Ainslie1920].
-After oviposition, the egg lies freely within the lumen of the host stem or in a hollow created by the female WSS ovipositor [@Fulbright2017].
+Newly laid eggs are crescent in shape, milky-white in color, and range between 1 mm to 1.25 mm in length and 0.33 to 0.42 mm wide (Ainslie, 1920).
+After oviposition, the egg lies freely within the lumen of the host stem or in a hollow created by the female WSS ovipositor (Fulbright et al., 2017).
 Laval development occurs rapidly with the recognizable shape becoming apparent at day three.
 By day five, the jaws turn brown as the eye spots begin to appear and darken.
-By day six or seven, the larva begins to hatch using a series of convulsive movements and emerges into the stem cavity, with full functionality achieved by the seventh day [@Ainslie1920].
+By day six or seven, the larva begins to hatch using a series of convulsive movements and emerges into the stem cavity, with full functionality achieved by the seventh day (Ainslie, 1920).
 
 ### Larvae
-After hatching and prior to feeding, the larvae are transparent and colorless [@Ainslie1920; @Criddle1923].
-WSS larvae are distinguishable from other larvae feeding in wheat and grasses by their highly sclerotized head capsule and large body size [@Nelson1953].
+After hatching and prior to feeding, the larvae are transparent and colorless (Ainslie, 1920; Criddle, 1923).
+WSS larvae are distinguishable from other larvae feeding in wheat and grasses by their highly sclerotized head capsule and large body size (Nelson and Farstad, 1953).
 As the larvae chew their way through the parenchymous and vascular tissues of the host stem, the larvae begin to develop a yellow-green color.
-As the larvae progress through the stem, they excrete frass; the digested plant tissue as well as metabolic breakdown products from the WSS [@Beres2011].
-If multiple eggs are laid within the same stem, the largest (usually the first egg laid and hatched) will typically remain within a few weeks due to the larvae cannibalizing one another [@Criddle1923; @Seamans1944; @Buteler2015].
-It remains unknown whether the consumption of other eggs and larvae is the result of intentional cannibalism or the indirect result of indiscriminate feeding [@Buteler2015].
-As the host plant physiologically matures, and the larvae reaches its fifth instar, the insect is cued to travel to the base of stem and prepare for obligatory diapause [@Beres2011].
+As the larvae progress through the stem, they excrete frass; the digested plant tissue as well as metabolic breakdown products from the WSS (Beres et al., 2011).
+If multiple eggs are laid within the same stem, the largest (usually the first egg laid and hatched) will typically remain within a few weeks due to the larvae cannibalizing one another (Criddle, 1923; Seamans et al., 1944; Buteler et al., 2015).
+It remains unknown whether the consumption of other eggs and larvae is the result of intentional cannibalism or the indirect result of indiscriminate feeding (Buteler et al., 2015).
+As the host plant physiologically matures, and the larvae reaches its fifth instar, the insect is cued to travel to the base of stem and prepare for obligatory diapause (Beres et al., 2011).
 The descending movement of larvae and development of the hibernaculum are triggered by the host plant senescence.
-As the stem walls becoming increasingly transparent, they allow increased amounts of visible and infrared light into the stem [@Holmes1977; @Beres2011].
-Once the larvae reach the base of the stem, they construct an interior V-shaped groove, making the stem weak enough to lodge due to gravity or wind [@Criddle1923].
-The larvae then plug the remaining stub with frass to insulate the hibernaculum during diapause [@Holmes1960].
+As the stem walls becoming increasingly transparent, they allow increased amounts of visible and infrared light into the stem (Holmes, 1977; Beres et al., 2011).
+Once the larvae reach the base of the stem, they construct an interior V-shaped groove, making the stem weak enough to lodge due to gravity or wind (Criddle, 1923).
+The larvae then plug the remaining stub with frass to insulate the hibernaculum during diapause (Holmes and Peterson, 1960).
 The larval stage of development can be as short as 30 days before entering diapause.
-Obligatory diapause lasts a minimum of 90 days [@Salt1947].
+Obligatory diapause lasts a minimum of 90 days (Salt and Hollick, 1947).
 After the larvae are exposed to temperatures less than 10C for 90 days, diapause is complete, and pupation can begin.
-Larvae can reenter diapause if temperatures approach 35C [@Salt1947] or if conditions are abnormally dry [@Holmes1978] shortly after termination of diapause.
+Larvae can reenter diapause if temperatures approach 35C (Salt and Hollick, 1947) or if conditions are abnormally dry (Holmes, 1978) shortly after termination of diapause.
 
@@
 Once diapause is complete, pupal development begins.
-Pupation lasts for 21 days, beginning in early- to mid-May [@Criddle1923].
-Initially, the pupa is milky-white in color [@Ainslie1920], gradually changing to bright white as the wings develop [@Fulbright2017].
+Pupation lasts for 21 days, beginning in early- to mid-May (Criddle, 1923).
+Initially, the pupa is milky-white in color (Ainslie, 1920), gradually changing to bright white as the wings develop (Fulbright et al., 2017).
 Additionally, as development continues, characteristic brown and black eye spots appear on the surface of the pupae.
-Once pupation is complete, the adult sawfly will chew through either the frass plug or the side of the stub, and emerge ready for flight [@Ainslie1920; @Holmes1960].
+Once pupation is complete, the adult sawfly will chew through either the frass plug or the side of the stub, and emerge ready for flight (Ainslie, 1920; Holmes and Peterson, 1960).
 
@@
 Wheat stem sawfly larvae travel throughout the stem of the host plant, feeding on the parenchymal tissue and chewing through the nodes.
-As a result, vascular translocation of water and nutrients within the plant is impaired [@Macedo2007] and causes a notable reduction yield and quality of the grain produced [@Seamans1944; @Holmes1977; @Morrill1992].
+As a result, vascular translocation of water and nutrients within the plant is impaired (Macedo et al., 2007) and causes a notable reduction yield and quality of the grain produced (Seamans et al., 1944; Holmes, 1977; Morrill et al., 1992).
 In addition, the weakening of the stem due to WSS causes lodging prior to harvest, exacerbated by larvae stem girdling.
-Stem lodging causes a reduction in stem harvest efficiently [@Ainslie1920], with yield lost due to unrecoverable lodged stem [@Beres2007].
-Additionally, the reduction in stubble residues decreases snow retention which increases soil erosion and reduces soil moisture content [@Keren2015].
+Stem lodging causes a reduction in stem harvest efficiently (Ainslie, 1920), with yield lost due to unrecoverable lodged stem (Beres et al., 2007).
+Additionally, the reduction in stubble residues decreases snow retention which increases soil erosion and reduces soil moisture content (Keren et al., 2015).
 
@@
 Because WSS are stem mining pests and spend most of their lives within the host plant stem, the use of traditional pest control methods such as insecticides have proven to be ineffective.
-The only life stage of WSS that takes place outside of the stem, the adult, is not targetable via chemical treatment due to adult emergence and oviposition taking place over many weeks [@Knodel2010; @Beres2011].
-WSS infestation chemical control methods were first tested as early as 1949 [@Wallace1962].
-After using six different organochlorine and organophosphate insecticidal dust applied via a powder duster [@Munro1949] the authors found “unsatisfactory results that did not control” WSS populations or limit infestations.
-In 1963, it was reported that a 73% control of WSS infestation could be achieved via granular heptachlor applied in the furrow with the wheat seed [@Wallace1962].
-However, the findings of these tests were not able to be replicable in all scenarios of WSS infestation [@Holmes1963].
+The only life stage of WSS that takes place outside of the stem, the adult, is not targetable via chemical treatment due to adult emergence and oviposition taking place over many weeks (Knodel et al., 2010; Beres et al., 2011).
+WSS infestation chemical control methods were first tested as early as 1949 (Wallace, 1962).
+After using six different organochlorine and organophosphate insecticidal dust applied via a powder duster (Munro et al., 1949) the authors found “unsatisfactory results that did not control” WSS populations or limit infestations.
+In 1963, it was reported that a 73% control of WSS infestation could be achieved via granular heptachlor applied in the furrow with the wheat seed (Wallace, 1962).
+However, the findings of these tests were not able to be replicable in all scenarios of WSS infestation (Holmes et al., 1963).
 Heptachlor required light sawfly infestation – restricted to the lower internodes – to be effective.
 When infestation rates were high, and oviposition occurred in the upper internodes, the researchers saw little to no control.
-Similar results were found when testing with foliar insecticides, findings showing little differences in their effect when compared to controls [@Blodgett1996].
+Similar results were found when testing with foliar insecticides, findings showing little differences in their effect when compared to controls (Blodgett et al., 1996).
 While the Montana Department of Agriculture has authorized the use of systemic organophosphate insecticide phorate (Thimet®) for use controlling WSS infestation in winter and spring wheat cultivars, it’s use is not considered environmentally friendly due to high toxicity to mammals, birds, and fish.
-Additionally, a preharvest interval of 85 days complicates harvest scheduling (Ginsberg, 2003; Mahajan et al., 2006), while the high toxicity may have potential adverse effects on beneficial parasitoid populations [@Varella2016].
+Additionally, a preharvest interval of 85 days complicates harvest scheduling (Ginsberg, 2003; Mahajan et al., 2006), while the high toxicity may have potential adverse effects on beneficial parasitoid populations (Varella, 2016).
 
@@
 #### Host Plant Resistance
-Mechanisms like antixenosis, antibiosis, and tolerance give a plant the ability to resist insect pests [@Horber1980; @Wilson1987].
+Mechanisms like antixenosis, antibiosis, and tolerance give a plant the ability to resist insect pests (Horber, 1980; Wilson et al., 1987).
 Antixenosis allows the host plant to trigger resistance by inducing non-preference in herbivores when they attempt to feed or oviposit.
@@
 (2020) found that Barely cultivars that closely resemble wheat show the greatest frequency of WSS oviposition.
-These findings suggest that semiochemicals produced by these barley cultivars, such as (Z)-3-hexenyl acetate [@Piesik2008; @Buteler2009] and (Z)-β-ocimene [@Buteler2012], are likely similar to those produced by wheat [@Achhami2020].
+These findings suggest that semiochemicals produced by these barley cultivars, such as (Z)-3-hexenyl acetate (Piesik et al., 2008; Buteler et al., 2009) and (Z)-β-ocimene (Buteler and Weaver, 2012), are likely similar to those produced by wheat (Achhami et al., 2020).
 Antibiosis causes the improper development of herbivores, impacting survivability and reproduction.
-Tolerance allows the host plant to withstand infestation and overcome injuries [@VanEmden1991; @VanZoeren2017].
-Much of the research done on WSS resistant wheat cultivars focuses on solid stem trait and pith expression [@Platt1949; @Varella2016; @Varella2017].
-Solid stem varieties are more resistant to infestation, reducing WSS oviposition and increasing egg mortality [@Holmes1960; @Varella2016].
-The increased pith expression also disrupts larvae movement within the stem, leading to higher larval mortality due to reduced feeding as well as smaller and less reproductively successful adults [@Wallace1966].
-In addition, greater pith expression increases stem strength, reducing losses due to lodged stems [@Delaney2010].
-Additional factors including late planting and shorter crop duration, have been suggested as potential factors complicating antixenosis, leading to increased WSS oviposition and infestation [@Achhami2020].
-The solid-stemmed trait has been developed in numerous spring and some winter wheat cultivars [@Beres2011].
-While the use of solid-stemmed cultivars is most often recommended for WSS management [@Beres2011], hollow-stemmed cultivars were found to exhibit greater parasitism when compared to solid-stemmed varieties [@Holmes1963].
+Tolerance allows the host plant to withstand infestation and overcome injuries (Van Emden, 1991; Van Zoeren and Guedot, 2017).
+Much of the research done on WSS resistant wheat cultivars focuses on solid stem trait and pith expression (Platt and Farstad, 1949; Varella et al., 2016; Varella et al., 2017).
+Solid stem varieties are more resistant to infestation, reducing WSS oviposition and increasing egg mortality (Holmes and Peterson, 1960; Varella et al., 2016).
+The increased pith expression also disrupts larvae movement within the stem, leading to higher larval mortality due to reduced feeding as well as smaller and less reproductively successful adults (Wallace and McNeal, 1966).
+In addition, greater pith expression increases stem strength, reducing losses due to lodged stems (Delaney et al., 2010).
+Additional factors including late planting and shorter crop duration, have been suggested as potential factors complicating antixenosis, leading to increased WSS oviposition and infestation (Achhami et al., 2020).
+The solid-stemmed trait has been developed in numerous spring and some winter wheat cultivars (Beres et al., 2011).
+While the use of solid-stemmed cultivars is most often recommended for WSS management (Beres et al., 2011), hollow-stemmed cultivars were found to exhibit greater parasitism when compared to solid-stemmed varieties (Holmes et al., 1963).
 It was previously suggested by Rand et al.
-(2012)  that the thicker pith found in solid-stemmed cultivars makes sensing WSS larvae vibrations [@Mankin2004] more difficult for female parasitoids.
+(2012)  that the thicker pith found in solid-stemmed cultivars makes sensing WSS larvae vibrations (Mankin et al., 2004) more difficult for female parasitoids.
 The inability to locate WSS hosts would then cause ineffective parasitism.
@@
 Larvae are protected against both water loss and freezing temperatures by the soil surrounding the wheat stubble where the WSS overwinters.
-Tilling the soil exposes the wheat stubble to the soil surface, conditions which kill the hibernating larvae [@Callenback1944].
+Tilling the soil exposes the wheat stubble to the soil surface, conditions which kill the hibernating larvae (Callenback and Hansmeier, 1944).
 Tillage of the WSS has been shown to be effective in decreasing the pest’s populations in both spring and winter wheat cultivars.
-Upwards of 90% mortality has been shown in North Dakota after fall tillage [@Weiss1987].
-Many factors influence the effectiveness of tilling, including adequate exposure of wheat stubs on the soil surface and appropriate timing [@Holmes1956; @Morrill1992; @Weiss1992].
-While the variability in effectiveness exists, some growers do consider it a crucial management tool [@Blodgett1996; @Lesieur2016].
-
+Upwards of 90% mortality has been shown in North Dakota after fall tillage (Weiss et al., 1987).
+Many factors influence the effectiveness of tilling, including adequate exposure of wheat stubs on the soil surface and appropriate timing (Holmes and Farstad, 1956; Morrill et al., 1992; Weiss and Morrill, 1992).
+While the variability in effectiveness exists, some growers do consider it a crucial management tool (Blodgett et al., 1996; Lesieur et al., 2016).
 While effective in killing WSS larvae, there are negative impacts caused by this practice.
-Water loss, soil compaction, and erosion have all been shown to be dramatically increased by tilling activities [@Phillips1980; @Lal1991].
+Water loss, soil compaction, and erosion have all been shown to be dramatically increased by tilling activities (Phillips et al., 1980; Lal, 1991).
 In addition, tilling practices negatively impact beneficial populations of parasitoid wasps that target the WSS.
-WSS parasitoids overwinter higher in the stem compared to at the base like WSS [@Morrill1998; @Runyon2002].
-By displacing soil when tilling, the upper portion of the plant is buried under the soil, preventing the parasitoid wasps from emerging in the spring [@Runyon2002; @Weaver2004].
+WSS parasitoids overwinter higher in the stem compared to at the base like WSS (Morrill et al., 1998; Runyon et al., 2002a).
+By displacing soil when tilling, the upper portion of the plant is buried under the soil, preventing the parasitoid wasps from emerging in the spring (Runyon et al., 2002a; Weaver et al., 2004).
 
@@
 Historically, it has been observed that wheat fields with wider row spacing (12 inches) have greater WSS cutting when compared to fields with reduced spacing (3 to 6 inches).
-Additionally, fields planted at greater seed density with solid stem cultivars exhibit lower WSS cutting compared to reduced seed density [@Luginbill1958].
+Additionally, fields planted at greater seed density with solid stem cultivars exhibit lower WSS cutting compared to reduced seed density (Luginbill and McNeal, 1958).
 The reason for this relationship between decreased density and increased sawfly infestation is due to the moisture availability and height of the wheat stems.
 At higher densities, each stem has access to less water, in turn leading to the stem to develop a thicker pith – more solid stemmed.
-As discussed above, there is a negative linear relationship between an increase in the solid-stem characteristic and infestation [@Kemp1934].
+As discussed above, there is a negative linear relationship between an increase in the solid-stem characteristic and infestation (Kemp, 1934).
 
 #### Crop Rotation
-Continually planting wheat year after year creates ideal conditions for the WSS [@Callenbach1945].
-Rotating wheat with more resistant wheat varieties as well as with oats, or soybeans (crops that are less or not susceptible to WSS infestation) is recommended to decrease the year after year WSS infestation rates [@Lin2014].
+Continually planting wheat year after year creates ideal conditions for the WSS (Callenbach and Hansmeier, 1945).
+Rotating wheat with more resistant wheat varieties as well as with oats, or soybeans (crops that are less or not susceptible to WSS infestation) is recommended to decrease the year after year WSS infestation rates (Lin and Chen, 2014).
 The notable downside to utilizing crop rotation in a matter that is effective is the economic downside.
-Alternate-year, summer fallow production systems are typical for wheat growing practices across much of the northern great plains of the United States [@Weiss1992; @Troeh1999].
+Alternate-year, summer fallow production systems are typical for wheat growing practices across much of the northern great plains of the United States (Weiss and Morrill, 1992; Troeh et al., 1999).
 
 #### Trap Crops
-Trap cropping utilized non-economically important plantings to help reduce populations or block movement of pests into valuable crop areas [@Hokkanen1991].
-Effective trap crops of the WSS should have appealing characteristics to the WSS to attract the pest to the trap planting, and away from the wheat [@Farstad1940].
-Historically, planted trap strips were used by attracting sawflies and destroying the strip at the end of the season, eliminating the larvae [@Morrill1992].
-However, utilizing trap plantings of highly attractive wheat varieties has been shown to hold limited viability, as growers are not fond of destroying planted grain at the end of the growing season [@Goosey1999].
-
-Alternative utilization of WSS traps entail planting and/or maintaining strips of smooth brome grass (Bromus inermis Leyss.) or other WSS host grasses [@Farstad1940].
-Smooth brome in particular holds potential as it is a poor host to WSS larvae [@Bhandari2020], yet WSS adults continue to oviposit the stems.
+Trap cropping utilized non-economically important plantings to help reduce populations or block movement of pests into valuable crop areas (Hokkanen and Jokioinen, 1991).
+Effective trap crops of the WSS should have appealing characteristics to the WSS to attract the pest to the trap planting, and away from the wheat (Farstad, 1940).
+Historically, planted trap strips were used by attracting sawflies and destroying the strip at the end of the season, eliminating the larvae (Morrill et al., 1992).
+However, utilizing trap plantings of highly attractive wheat varieties has been shown to hold limited viability, as growers are not fond of destroying planted grain at the end of the growing season (Goosey, 1999).
+Alternative utilization of WSS traps entail planting and/or maintaining strips of smooth brome grass (Bromus inermis Leyss.) or other WSS host grasses (Farstad, 1940).
+Smooth brome in particular holds potential as it is a poor host to WSS larvae (Bhandari, 2020), yet WSS adults continue to oviposit the stems.
 This relationship of greater oviposition with increased mortality may lead to decreases in overall WSS populations.
@@
 #### Delayed Planting & Swathing
-Wheat planted later in the growing season avoids much of the damage caused by the WSS as the wheat reaches stem elongation after many of the adult sawflies have already laid their eggs [@Jacobson1952].
-The problem with delayed planting is that the practice is known to make less efficient use the earl season moisture in the soil [@Morrill1996].
+Wheat planted later in the growing season avoids much of the damage caused by the WSS as the wheat reaches stem elongation after many of the adult sawflies have already laid their eggs (Jacobson and Farstad, 1952).
+The problem with delayed planting is that the practice is known to make less efficient use the earl season moisture in the soil (Morrill and Kushnak, 1996).
 Growers may need to weigh the tradeoffs in delayed planting to reduce losses in areas with a history of severe infestation.
-Swathing, also known as windrowing, is the practice of cutting the wheat to accelerate the drying process and decrease the losses caused by WSS cutting [@Holmes1965; @Carlson2019; @McCullough2020].
-By cutting the wheat and letting it dry prior to sawfly cutting, swathing works by reducing losses due to lodging [@Ainslie1920].
+Swathing, also known as windrowing, is the practice of cutting the wheat to accelerate the drying process and decrease the losses caused by WSS cutting (Holmes and Peterson, 1965; Carlson, 2019; McCullough et al., 2020).
+By cutting the wheat and letting it dry prior to sawfly cutting, swathing works by reducing losses due to lodging (Ainslie, 1920).
 
@@
 usefulness to WSS management.
-[@Wenda2009].
+(Wenda-Piesik et al., 2009).
 Findings by Tangtrakulwanich et al.
@@
 brunneum effectively reduced damage caused by WSS.
-Both fungi kill insects by infecting via spore contact or by insect consumption of treated plant tissue [@Portman2018].
+Both fungi kill insects by infecting via spore contact or by insect consumption of treated plant tissue (Portman et al., 2018).
 
 #### Predatory Beetle
-Phyllobaenus dubius (Wolcott) has been observed emerging from wheat stubble and within dissected wheat stems in Montana [@Morrill2001].
+Phyllobaenus dubius (Wolcott) has been observed emerging from wheat stubble and within dissected wheat stems in Montana (Morrill et al., 2001).
 Further research is required on this coleopteran to determine its prospects going forward as an effective WSS management tool.
@@
 The congeneric parasitoids, Bracon cephi (Gahan) and B.
-lissogaster Muesebeck, are the most effective agents of WSS biological control [@Runyon2001].
+lissogaster Muesebeck, are the most effective agents of WSS biological control (Runyon et al., 2001).
 B.
@@
 cephi and B.
-lissogaster [@Ainslie1920; Cárcamo et al., 2012).
-However, as agriculture increased and native grasses were replaced by large monoculture landscapes, the WSS moved into wheat [@Farstad1940; @Davis1955; Holmes1978] while the parasitoids were slower to adapt [@Morrill1998].
+lissogaster (Ainslie, 1920; Cárcamo et al., 2012).
+However, as agriculture increased and native grasses were replaced by large monoculture landscapes, the WSS moved into wheat (Farstad, 1940; Davis et al., 1955; Holmes, 1978) while the parasitoids were slower to adapt (Morrill et al., 1998).
 
-## Biology of *Bracon cephi* and *Bracon lissogaster*
+## Biology of Bracon cephi and Bracon lissogaster
 ### Adult
-Both *B.
-cephi* and *B.
-lissogaster* are protelean host-specific ectoparasitoids with two generations per year (bivoltine) [@Nelson1953; @Somsen1956; @Holmes1963].
-They are ectoparasitoids and idiobionts, living on the integument of their hosts and immediately paralyzing their hosts in location and stage it was when it was attacked [@Nelson1953; @Runyon2002; @Weaver2004].
+Both B.
+cephi and B.
+lissogaster are protelean host-specific parasitoids with two generations per year (bivoltine) (Nelson and Farstad, 1953; Somsen and Luginbill, 1956; Holmes et al., 1963).
+They are ectoparasitoids and idiobionts, living on the integument of their hosts and immediately paralyzing their hosts in location and stage it was when it was attacked (Nelson and Farstad, 1953; Runyon et al., 2002b; Weaver et al., 2004).
 Male parasitoids are approximately 3.5 mm in length, while females are 4.1 mm.
-They are dark red, with black antennae, eyes, ovipositor and ocelli [@Gahan1918].
+They are dark red, with black antennae, eyes, ovipositor and ocelli (Gahan, 1918).
 B.
 lissogaster and B.
-cephi are distinguishable by their head color and unique metasoma surface texture [@Runyon2001].
+cephi are distinguishable by their head color and unique metasoma surface texture (Runyon et al., 2001).
 Adults emerge after completing a pupal stage within the host stem by chewing a hole through the stem wall.
-In Montana, the first generations are present from late June to late July, while the second generations are active in August and September [@Somsen1956].
+In Montana, the first generations are present from late June to late July, while the second generations are active in August and September (Somsen and Luginbill, 1956).
 Adult female B.
@@
 lissogaster will locate a host stem using a combination of olfactory and volatile cues.
-The wasps will then locate a host sawfly larva using acoustic cues [@Mankin2004] and use their ovipositor first paralyze, then deposit 1-4 eggs next to the paralyzed larva.
+The wasps will then locate a host sawfly larva using acoustic cues (Mankin et al., 2004) and use their ovipositor first paralyze, then deposit 1-4 eggs next to the paralyzed larva.
 
 ### Egg
-*B.
-cephi* and *B.
-lissogaster* eggs are on average 0.86 mm in length, exhibiting an elongated shape and pale-yellow color.
+B.
+cephi and B.
+lissogaster eggs are on average 0.86 mm in length, exhibiting an elongated shape and pale-yellow color.
 Females lay one to three eggs on the WSS larval integument or adjacent to the larva.
-The eggs develop for one to three days before hatching [@Nelson1953; @Somsen1956].
+The eggs develop for one to three days before hatching (Nelson and Farstad, 1953; Somsen and Luginbill, 1956).
 
 ### Larva
-Parasitoid larvae are 1mm in length, with a translucent hue that turns increasingly brown as the larva progresses through five instar stages [@Nelson1953].
+Parasitoid larvae are 1mm in length, with a translucent hue that turns increasingly brown as the larva progresses through five instar stages (Nelson and Farstad, 1953).
 The newly hatched larvae have no appendages, save for antennae, and they exhibit highly sclerotized mouthparts and well-developed mandibles.
 After hatching, the B.
-lissogaster larva attach themselves to the integument of the host larva and immediately begin feeding [@Somsen1956].
+lissogaster larva attach themselves to the integument of the host larva and immediately begin feeding (Somsen and Luginbill, 1956).
 When feeding is complete, little remains of the host but the integument.
@@
 Offspring laid as a part of the second generation will overwinter within the cocoons and emerge in the spring.
-[@Nelson1953]
+(Nelson and Farstad, 1953)
 
@@
 Bhandari (2020) has shown that oviposition occurs with similar frequency in wheat as it does in smooth brome, however, we see much higher rates of larval mortality within smooth brome stems.
-This has led researchers to hypothesize about the feasibility of using smooth brome as a trap cop [@mendoza2006; @Lesieur2016].
-While feasible as a feed crop, as it holds considerably less economic value compared to wheat or other high value crops [@Criddle1922].
-Smooth brome may be usable in trap strips along the edge of wheat fields [@Hokkanen1991].
+This has led researchers to hypothesize about the feasibility of using smooth brome as a trap cop (Perez-Mendoza et al., 2006; Lesieur et al., 2016).
+While feasible as a feed crop, as it holds considerably less economic value compared to wheat or other high value crops (Criddle, 1922).
+Smooth brome may be usable in trap strips along the edge of wheat fields (Hokkanen and Jokioinen, 1991).
 While a poor host for the wheat stem sawfly, surviving immature larvae found within brome stems can be parasitized by B.
@@
 ## Plant Semiochemicals
-
 Plants interact with other organisms through a complex system of chemical signals known as Semiochemicals.
-Semiochemicals are compounds or mixtures of compounds that are released from one organism that evoke a physiological or behavior response in another organism [@Agelopoulos1999].
-Semiochemicals play a large role in both insect-insect and plant-insect interactions and are essential in insects locating feedings sites, mating sites, egg-laying sites, and/or refugia [@Prokopy1984].
-
-While it is known that these chemicals play an important role in communication between plants, insects, and other organisms, there is still debate about the function of many of them [@Reddy2004; @Voelckel2018].
+Semiochemicals are compounds or mixtures of compounds that are released from one organism that evoke a physiological or behavior response in another organism (Agelopoulos et al., 1999).
+Semiochemicals play a large role in both insect-insect and plant-insect interactions and are essential in insects locating feedings sites, mating sites, egg-laying sites, and/or refugia (Prokopy et al., 1984).
+While it is known that these chemicals play an important role in communication between plants, insects, and other organisms, there is still debate about the function of many of them (Reddy and Guerrero, 2004; Voelckel and Jander, 2018).
 
 ### Semiochemicals in Plant-Insect Systems
-
-Within plant-insect systems, plants respond to insect herbivory by synthesizing and releasing volatile compounds and providing important signals to biocontrol agents to localize potential host insects [@Sharma2019].
-In addition, semiochemicals provide localized signatures that allow insects to detect food sources, find mates, and locate preferential oviposition sites [@Norin2007].
-Plants typically produce semiochemicals when defending themselves from herbivorous arthropods [Guerrieri1999] or pathogen attacks [@Smart2014].
+	Within plant-insect systems, plants respond to insect herbivory by synthesizing and releasing volatile compounds and providing important signals to biocontrol agents to localize potential host insects (Sharma et al., 2019).
+In addition, semiochemicals provide localized signatures that allow insects to detect food sources, find mates, and locate preferential oviposition sites (Norin, 2007).
+Plants typically produce semiochemicals when defending themselves from herbivorous arthropods (Guerrieri et al., 1999) or pathogen attacks (Smart et al., 2014 Jan 1).
 These responses can be grouped into two categories, direct and indirect responses.
-Direct responses are compounds that the plant produces that have a direct effect on the herbivore or pathogen [@Metcalf1998].
-Indirect responses are the compounds produced that have no impact directly on the pest, but instead may attract parasitoids, predators, or other biological control agents [@Mayland2000; @Colazza2004].
+Direct responses are compounds that the plant produces that have a direct effect on the herbivore or pathogen (Metcalf, 1998).
+Indirect responses are the compounds produced that have no impact directly on the pest, but instead may attract parasitoids, predators, or other biological control agents (Mayland et al., 2000; Colazza et al., 2004).
 
@@
 
-## Controlled *C. cinctus* infestation of *B. inermis*
+## Controlled C. cinctus infestation of B. inermis
 ### Plant Selection and Field Cages
@@
 The amplifier transferred the information to a computer interface with an EAG version 2.7 software (Syntech NL 2001, Hilversum, Netherlands).
-This apparatus and method has been consistently used previously in assessing insect volatile response (@Agelopoulos1999).
+This apparatus and method has been consistently used previously in assessing insect volatile response (Agelopoulos et al., 1999).
 

### 2024-03-20 qmd/pics/running.qmd
commit for 3/20

---
title: "Running Pictures"
---
::: {style="display: flex;"}
<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
  <a href="running/deadhorse.qmd">
    <img src="../../img/deadhorse/deadhorse-top.JPG" alt="Deadhorse Ultra" width="300" height="200">
  </a>
  <figcaption style="text-align: center;">Deadhorse Ultra, 2023</figcaption>
</figure>

<figure style="margin-right: 20px; display: flex; flex-direction: column; align-items: center;">
  <a href = "running/tmr50k.qmd">
    <img src="../../img/tmr50k/tmr50k-top.JPG" alt="Teton Mountain Runs 50K" 
  width = 300" height = "200">
  </a>
  <figcaption style="text-align:center;">Teton Mountain Runs, 2023</figcaption>
</figure>

:::

### 2024-03-20 qmd/pics/running/deadhorse.qmd
commit for 3/20

--- 2024-03-05 deadhorse.qmd
@@
 # Set the directory path
-directory_path <- "img/deadhorse"
+directory_path <- "img/deadhorse/"
 
@@
   # Generate the HTML code for each image
-  image_html <- paste0('<img src="', directory_path, "/", file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')
+  image_html <- paste0('<img src="', "../../../", directory_path, file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')
 
@@
 <div class="gallery">
-<img src="img/deadhorse/deadhorse-top.JPG" alt="Image 1" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_0012.JPG" alt="Image 2" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_0016.JPG" alt="Image 3" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_0026.JPG" alt="Image 4" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6044.jpeg" alt="Image 5" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6050.jpeg" alt="Image 6" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6058.jpeg" alt="Image 7" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6061.jpeg" alt="Image 8" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6062.jpeg" alt="Image 9" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6067.jpeg" alt="Image 10" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6071.jpeg" alt="Image 11" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6078.jpeg" alt="Image 12" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6079.jpeg" alt="Image 13" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6080.jpeg" alt="Image 14" onclick="expandImage(this)">
-<img src="img/deadhorse/IMG_6192.HEIC" alt="Image 15" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/deadhorse-top.JPG" alt="Image 1" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_0012.JPG" alt="Image 2" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_0016.JPG" alt="Image 3" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_0026.JPG" alt="Image 4" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6044.jpeg" alt="Image 5" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6050.jpeg" alt="Image 6" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6058.jpeg" alt="Image 7" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6061.jpeg" alt="Image 8" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6062.jpeg" alt="Image 9" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6067.jpeg" alt="Image 10" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6071.jpeg" alt="Image 11" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6078.jpeg" alt="Image 12" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6079.jpeg" alt="Image 13" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6080.jpeg" alt="Image 14" onclick="expandImage(this)">
+<img src="../../../img/deadhorse/IMG_6192.HEIC" alt="Image 15" onclick="expandImage(this)">
 </div>

### 2024-03-20 qmd/pics/running/tmr50k.qmd
commit for 3/20

---
title: "Teton Mountain Runs 50K"
subtitle: "July 16, 2023"
editor_options: 
  chunk_output_type: console
---

<a href="http://competitivetiming.com/results/2023/2356951O" target="_blank">View Results</a>

```{r, echo = F}
code1 <-
"<html lang=\"en\">
<head>
  <meta charset=\"UTF-8\">
  <meta name=\"viewport\" content=\"width=device-width, initial-scale=1.0\">
  <title>Image Gallery</title>
  <style>
    .gallery {
      display: grid;
      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
      grid-gap: 10px;
    }
    .gallery img {
      width: 150px;
      height: 150px;
      object-fit: cover; /* Ensure images cover the entire space */
      cursor: pointer;
      transition: transform 0.3s ease;
    }
    .gallery img:hover {
      transform: scale(1.05); /* Slightly increase size on hover */
    }
    .expanded-container {
      position: fixed;
      top: 0;
      left: 0;
      width: 100%;
      height: 100%;
      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
      display: none; /* Initially hidden */
      justify-content: center;
      align-items: center;
      z-index: 9999;
    }

    .expanded {
      max-width: 80%;
      max-height: 80%;
      cursor: pointer;
    }
  </style>
</head>
<body>"

```

```{r eval = F, echo = F, results = 'asis'}
# Set the directory path
directory_path <- "../../../img/tmr50k"

# List all the files in the directory
file_names <- list.files(directory_path)

# Exclude files with .zip extension
file_names <- file_names[!grepl("\\.zip$", file_names)]

# Initialize an empty character vector to store the HTML code
html_code <- "<div class=\"gallery\">\n"

# Loop through each file name and generate HTML code
for (i in seq_along(file_names)) {
  # Generate the HTML code for each image
  image_html <- paste0('<img src="', directory_path, "/", file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')

  # Append the generated HTML code to the existing code
  html_code <- paste(html_code, image_html, sep = "")
}

# Close the gallery div
html_code <- paste(html_code, "</div>\n<div class=\"expanded-container\" onclick=\"closeExpanded()\">\n  <img class=\"expanded\" id=\"expanded-image\" onclick=\"navigateGallery(event)\">\n</div>", sep = "")

# Print the final HTML code
cat(html_code)

```

```{=html}
<html lang="en">
<head>
  <meta charset="UTF-8">
  <meta name="viewport" content="width=device-width, initial-scale=1.0">
  <title>Image Gallery</title>
  <style>
    .gallery {
      display: grid;
      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
      grid-gap: 10px;
    }
    .gallery img {
      width: 150px;
      height: 150px;
      object-fit: cover; /* Ensure images cover the entire space */
      cursor: pointer;
      transition: transform 0.3s ease;
    }
    .gallery img:hover {
      transform: scale(1.05); /* Slightly increase size on hover */
    }
    .expanded-container {
      position: fixed;
      top: 0;
      left: 0;
      width: 100%;
      height: 100%;
      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
      display: none; /* Initially hidden */
      justify-content: center;
      align-items: center;
      z-index: 9999;
    }

    .expanded {
      max-width: 80%;
      max-height: 80%;
      cursor: pointer;
    }
  </style>
</head>
<body>

<div class="gallery">
<img src="../../../img/tmr50k/DSC01450.JPG" alt="Image 1" onclick="expandImage(this)">
<img src="../../../img/tmr50k/DSC01715.JPG" alt="Image 2" onclick="expandImage(this)">
<img src="../../../img/tmr50k/DSC02442.JPG" alt="Image 3" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_0054.JPG" alt="Image 4" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_0088.JPG" alt="Image 5" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_0094.JPG" alt="Image 6" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_0108.JPG" alt="Image 7" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_0120.JPG" alt="Image 8" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_0126.JPG" alt="Image 9" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_5954.JPG" alt="Image 10" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_5983.JPG" alt="Image 11" onclick="expandImage(this)">
<img src="../../../img/tmr50k/IMG_5999.JPG" alt="Image 12" onclick="expandImage(this)">
<img src="../../../img/tmr50k/tmr50k-top.JPG" alt="Image 13" onclick="expandImage(this)">
</div>
<div class="expanded-container" onclick="closeExpanded()">
  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
</div>

<script>
  var currentImageIndex = 0;
  var galleryImages = document.querySelectorAll('.gallery img');

  function expandImage(img) {
    currentImageIndex = Array.from(galleryImages).indexOf(img);
    document.getElementById('expanded-image').src = img.src;
    document.querySelector('.expanded-container').style.display = 'flex';
  }

  function closeExpanded() {
    document.querySelector('.expanded-container').style.display = 'none';
  }

  function navigateGallery(event) {
    // Calculate the click position relative to the expanded image
    var clickX = event.pageX - event.target.getBoundingClientRect().left;
    var imageWidth = event.target.offsetWidth;

    // Determine if the click was on the left or right side of the image
    if (clickX < imageWidth / 2) {
      // Clicked on the left side: navigate to the previous image
      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
    } else {
      // Clicked on the right side: navigate to the next image
      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
    }

    // Update the expanded image source
    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
  }

  // Add event listeners to the left and right sides of the expanded image for navigation
  document.querySelector('.expanded').addEventListener('click', function(event) {
    navigateGallery(event);
  });
</script>

</body>
</html>
```

<br>
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Download File Button</title>
<style>
/* Button styling */
.button {
  background-color: #2068DC; /* Header of website color (blue) */
  border: none;
  color: white;
  padding: 15px 32px;
  text-align: center;
  text-decoration: none;
  display: inline-block;
  font-size: 16px;
  margin: 4px 2px;
  cursor: pointer;
  border-radius: 8px;
}
.button:hover {
  background-color: #3F91FB; /* Sky blue */
}
</style>
</head>
<body>

<!-- Button to download the file -->
<a href="../../../img/tmr50k/tmr2023.zip" download="TetonMountainRuns2023">
  <button class="button">Download All Pictures</button>
</a>

</body>
</html>

```

### 2024-03-20 qmd/projects.qmd
commit for 3/20

--- 2024-03-09 projects.qmd
@@
   .text h1 {
-    font-size: 18px; /* Adjusted text size */
+    font-size: 22px; /* Adjusted text size */
     margin-top: 0;
@@
   .text p {
-    font-size: 14px;
+    font-size: 16px;
     margin-top: 5px;
@@
 <body>
-  <a href="msu-masters.qmd" class="container">
+  <a href="projects/msu-masters.qmd" class="container">
     <div class="image-container">
-      <img src="img/projects-logos/msu-logo.jpg" alt="Image">
+      <img src="../img/projects-logos/msu-logo.jpg" alt="Image">
     </div>
@@
         <h1>Montana State University Master's</h1>
-        <p>Co-advised by Dr. David Weaver and Dr. Tracy Sterling.</p>
+        <p>2021-2024<br> Co-advised by Dr. David Weaver and Dr. Tracy Sterling</p>
       </div>
@@
   </a>
-  <a href="bballref.qmd" class="container">
+  <a href="projects/bballtool.qmd" class="container">
     <div class="image-container">
-      <img src="img/projects-logos/bballref-logo-circle.jpg" alt="Image">
+      <img src="../img/projects-logos/bballref-logo-circle.jpg" alt="Image">
     </div>
@@
         <h1>NBA Roster Analysis Tool</h1>
-        <p></p>
+        <p>2017-2019</p>
       </div>
@@
 ```
+
+```{r, echo = F, message = F, results = 'hide'}
+library(pandoc)
+library(knitr)
+library(rmarkdown)
+getwd()
+rmarkdown::render(input = "litreview.qmd", 
+                  run_pandoc = TRUE,
+                  output_format = "word_document")
+
+```

### 2024-03-20 qmd/projects/bballtool.qmd
commit for 3/20

---
title: "NBA Roster Analysis Tool"

editor_options: 
  chunk_output_type: console
---

From 2017 to 2019, I built my own NBA roster analysis tool for analyzing roster construction, free agency, and per-game metrics. The information is out of date, but all of the functionality still works.

```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>NBA Roster Analysis Tool Preview</title>
</head>
<body>

<h2>Excel Preview</h2>

<!-- Replace 'path_to_your_excel_file.xlsx' with the actual path to your Excel file -->
<iframe src="https://1drv.ms/x/s!AiC9jtYNi0Ovhy_ySBYSU-xuO1CQ?e=ohHHit" width="100%" height="500px" frameborder="0">This browser does not support PDFs. Please download the PDF to view it: <a href="https://1drv.ms/x/s!AiC9jtYNi0Ovhy_ySBYSU-xuO1CQ?e=ohHHit</a></iframe>

</body>
</html>
```

Or download the file here:

```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Download File Button</title>
<style>
/* Button styling */
.button {
  background-color: #4CAF50; /* Green */
  border: none;
  color: white;
  padding: 15px 32px;
  text-align: center;
  text-decoration: none;
  display: inline-block;
  font-size: 16px;
  margin: 4px 2px;
  cursor: pointer;
  border-radius: 8px;
}
.button:hover {
  background-color: #45a049; /* Darker green */
}
</style>
</head>
<body>

<!-- Button to download the file -->
<a href="/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Personal/bballref.xlsx " download="Jackson Strand NBA Roster Analysis Tool.xlsx">
  <button class="button">Download File</button>
</a>

</body>
</html>

```

### 2024-03-20 qmd/projects/msu-masters.qmd
commit for 3/20

--- 2024-03-20 qmd/projects.qmd
@@
 ---
-title: "Projects"
-editor_options: 
-  chunk_output_type: console
+title: "MSU Masters"
+
 ---
@@
 <body>
-  <a href="projects/msu-masters.qmd" class="container">
+  <a href="msu-masters/wss.qmd" class="container">
     <div class="image-container">
-      <img src="../img/projects-logos/msu-logo.jpg" alt="Image">
+      <img src="../../img/msu-masters/oviposition.JPG" alt="Image">
     </div>
@@
       <div class="text">
-        <h1>Montana State University Master's</h1>
-        <p>2021-2024<br> Co-advised by Dr. David Weaver and Dr. Tracy Sterling</p>
+        <h1>Wheat Stem Sawfly</h1>
+        <p>Understanding the importance of smooth brome in promoting a Braconid parasitoid that targets the Wheat Stem Sawfly.</p>
       </div>
@@
   </a>
-  <a href="projects/bballtool.qmd" class="container">
+  <a href="msu-masters/locoweed/locoweed-report.qmd" class="container">
     <div class="image-container">
-      <img src="../img/projects-logos/bballref-logo-circle.jpg" alt="Image">
+      <img src="../../img/msu-masters/loco-bee2.jpg" alt="Image">
     </div>
@@
       <div class="text">
-        <h1>NBA Roster Analysis Tool</h1>
-        <p>2017-2019</p>
+        <h1>Locoweed Volatiles</h1>
+        <p>Comparison of volatile organic compounds profiles of <em>Oxytropis</em> plants that contained the commensal endophyte <em>Alternaria oxytropis</em> and those that did not.</p>
+      </div>
+    </div>
+  </a>
+  <a href="msu-masters/toadflax/toadflax_report.qmd" class="container">
+    <div class="image-container">
+      <img src="../../img/msu-masters/toadflax.jpeg" alt="Image">
+    </div>
+    <div class="content">
+      <div class="text">
+        <h1>Toadflax</h1>
+        <p></p>
       </div>
@@
 ```
-
-```{r, echo = F, message = F, results = 'hide'}
-library(pandoc)
-library(knitr)
-library(rmarkdown)
-getwd()
-rmarkdown::render(input = "litreview.qmd", 
-                  run_pandoc = TRUE,
-                  output_format = "word_document")
-
-```

### 2024-03-20 qmd/projects/msu-masters/locoweed/locoweed-report.qmd
commit for 3/20

--- 2024-03-09 locoweed-report.qmd
@@
 ---
-title: "Locoweed Volatiles"
+title: "Locoweed Volatiles Report"
 author: "Jackson Strand"
@@
     code_folding: hide
-bibliography: citations/citations.bib
+bibliography: locoweed.bib
 link-citations: true
@@
 <div class="image-container">
-    <img src="img/loco-bee-crop.JPG" alt="Your Image">
+    <img src="../../../../img/loco-bee-crop.JPG" alt="Your Image">
 </div>
@@
 # Set the directory path
-directory_path <- "img/loco-plants"
+directory_path <- "../../../img/loco-plants"
 
@@
   # Generate the HTML code for each image
-  image_html <- paste0('<img src="', directory_path, "/", file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')
+  image_html <- paste0('<img src="', "../",directory_path, "/", file_names[i], '" alt="Image ', i, '" onclick="expandImage(this)">\n')
 
@@
 <div class="gallery">
-<img src="img/loco-plants/IMG_0064.jpg" alt="Image 1" onclick="expandImage(this)">
-<img src="img/loco-plants/IMG_0820 2.jpg" alt="Image 2" onclick="expandImage(this)">
-<img src="img/loco-plants/IMG_1165.jpg" alt="Image 3" onclick="expandImage(this)">
-<img src="img/loco-plants/IMG_1177.jpg" alt="Image 4" onclick="expandImage(this)">
-<img src="img/loco-plants/IMG_1923.jpg" alt="Image 5" onclick="expandImage(this)">
+<img src="../../../../img/loco-plants/IMG_0064.jpg" alt="Image 1" onclick="expandImage(this)">
+<img src="../../../../img/loco-plants/IMG_0820 2.jpg" alt="Image 2" onclick="expandImage(this)">
+<img src="../../../../img/loco-plants/IMG_1165.jpg" alt="Image 3" onclick="expandImage(this)">
+<img src="../../../../img/loco-plants/IMG_1177.jpg" alt="Image 4" onclick="expandImage(this)">
+<img src="../../../../img/loco-plants/IMG_1923.jpg" alt="Image 5" onclick="expandImage(this)">
 </div>

### 2024-03-20 qmd/projects/msu-masters/wss.qmd
commit for 3/20

--- 2024-03-20 qmd/projects.qmd
@@
 ---
-title: "Projects"
-editor_options: 
-  chunk_output_type: console
+title: "MSU Master's: Wheat Stem Sawfly"
 ---
+
+<a href="litreview.qmd" target="_blank">View Project Proposal</a>
 
@@
 <body>
-  <a href="projects/msu-masters.qmd" class="container">
+  <a href="wss/pfreport.qmd" class="container">
     <div class="image-container">
-      <img src="../img/projects-logos/msu-logo.jpg" alt="Image">
+      <img src="../../../img/msu-masters/cages2.jpg" alt="Image"> 
     </div>
@@
       <div class="text">
-        <h1>Montana State University Master's</h1>
-        <p>2021-2024<br> Co-advised by Dr. David Weaver and Dr. Tracy Sterling</p>
+        <h1>Controlled WSS Infestation</h1>
+        <p>Assessing WSS infestation and cutting within smooth brome in controlled research plot at Arthur H. Post Agronomy Farm.</p>
       </div>
@@
   </a>
-  <a href="projects/bballtool.qmd" class="container">
+  <a href="wss/field_report.qmd" class="container">
     <div class="image-container">
-      <img src="../img/projects-logos/bballref-logo-circle.jpg" alt="Image">
+      <img src="../../../img/msu-masters/oviposition.JPG" alt="Image"> 
+    </div>
+    <div class="content">
+      <div class ="text">
+        <h1>Montana Field Sites</h1>
+        <p>Collection of wheat and smooth brome in Big Sandy and Moccasin, Montana to assess parasitism and WSS infestation.</p>
+      </div>
+    </div>
+  </a>
+  <a href="wss/eag.qmd" class="container">
+    <div class="image-container">
+      <img src="../../../img/msu-masters/vcs.jpg" alt="Image"> 
     </div>
@@
       <div class="text">
-        <h1>NBA Roster Analysis Tool</h1>
-        <p>2017-2019</p>
+        <h1>Organic Volatile Collection</h1>
+        <p>Collection of wheat and smooth brome in <strong>Big Sandy</strong> and <strong>Moccasin</strong> Montana to assess parasitism and WSS infestation.</p>
+      </div>
+    </div>
+  </a>
+  <a href="wss/weatherdata.qmd" class="container">
+    <div class="image-container">
+      <img src="../../../img/msu-masters/clouds.jpg" alt="Image"> 
+    </div>
+    <div class="content">
+      <div class="text">
+        <h1>Weather Data</h1>
+        <p>Weather data trends from Big Sandy and Moccasin, Montana.</p>
       </div>

### 2024-03-20 qmd/projects/msu-masters/wss/weatherdata.qmd
commit for 3/20

--- 2024-03-05 weatherdata.qmd
@@
 ggplot(data = bozeman_weather) +
-  geom_line(aes(x = date, y = TMAX)) +
+  geom_line(aes(x = date, y = PRCP)) +
   theme_bw()
+
+ggplot(data = bozeman_weather) +
+  geom_line(aes(x = date, y = PRCP)) +
+  geom_smooth(aes(x = date, y = PRCP), method = "loess", se = TRUE) +
+  theme_bw()
+
 ```

### 2024-03-22 qmd/cv.qmd
commit for 3/22 morning

@@
 resources:
-  - "../Aboutme/jackson_strand_updated_cv.pdf"
+  - "../Aboutme/Jackson_Strand_CV.pdf"
 cv:
-  pdf: "../Aboutme/jackson_strand_updated_cv.pdf"
+  pdf: "../Aboutme/Jackson_Strand_CV.pdf"
 ---
+
+<!-- remember to print to pdf and don't save as pdf -->
 
@@
   <!-- Button to download the file -->
-  <a href="../Aboutme/jackson_strand_updated_cv.pdf" download="Jackson Strand CV">
+  <a href="../Aboutme/Jackson_Strand_CV.pdf" download="Jackson Strand CV">
     <button class="button">Download</button>

### 2024-03-22 qmd/litreview.qmd
commit for 3/22 morning

@@
 
-# Prior Research on Topic (Literature Review)
+# Literature Review
 ## History and Distribution of Wheat Stem Sawfly
@@
 Obligatory diapause lasts a minimum of 90 days [@Salt1947].
-After the larvae are exposed to temperatures less than 10C for 90 days, diapause is complete, and pupation can begin.
-Larvae can reenter diapause if temperatures approach 35C [@Salt1947] or if conditions are abnormally dry [@Holmes1978] shortly after termination of diapause.
+After the larvae are exposed to temperatures less than 10°C for 90 days, diapause is complete, and pupation can begin.
+Larvae can reenter diapause if temperatures approach 35°C [@Salt1947] or if conditions are abnormally dry [@Holmes1978] shortly after termination of diapause.
 
@@
 
-# CHAPTERS
-## METHODS
-### Central and Northern Montana Field Sites
-#### Field Selection & Data Collection
-To assess C cinctus infestation, larval mortality, and parasitoid prevalence within B.
-inermis and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production.
+# Methodology
+## Field Experiments
+### Field Survey
+To assess *C.
+cinctus* infestation, larval mortality, and parasitoid prevalence within *B.
+inermis* and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production.
 Samples were collected in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA.
-B.
-inermis sites were set up as polygons along the edge of adjoining wheat fields measuring approximately 100 meters in area.
+*B.
+inermis* sites were set up as polygons along the edge of adjoining wheat fields measuring approximately 100 meters in area.
 Four collection squares of 1ft x 1ft were randomly partitioned within each polygon during both collection events each year.
@@
 
-#### Statistical Analysis
-Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions.
-‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs awere used as the response variable to assess the differences in parameters among the sampling sites.
-ANOVA was used to caculate P-values using F-distribution to understand overall significant differences in mean C.
-cinctus infestation and parasitism among polygons.
-To assess the profile of C.
-cintus infestation, paraisitms, and larval presence by node, a linear model was fit for each sampling site.
-Further assessment was done on the resonpses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.
-
-## Controlled *C. cinctus* infestation of *B. inermis*
-### Plant Selection and Field Cages
-Assessment of C.
-cinctus infestation and mortality within B.
-inermis were assessed using a 34 x 60 ft plot at the Arthur H.
+### Controlled *C. cinctus* infestation of *B. inermis*
+#### Plant Selection and Field Cages
+Assessment of *C.
+cinctus* infestation and mortality within *B.
+inermis* were assessed using a 34 x 60 ft plot at the Arthur H.
 Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-The cage structure was built using 1-inch PVC piping with the netting made using 530 Amber Lumite Screen (BioQuip Products, LLC).
+The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
 Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.
@@
 
-### Insects
+#### Insects
+Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of *C.
+cinctus* infestation and cutting the year prior.
+Cut stubble, which contained *C.
+cinctus* larvae in diapause, were kept refrigerated between -2°C and 3°C for >100 days as required to complete obligatory larval diapause.
+As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA).
+Stubs were added to cages once *B.
+inermis* stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).
 
-Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of C.
-cinctus infestation and cutting the year prior.
-Cut stubble, which contained C.
-cinctus larvae in diapause, were kept refrigerated between -2°C and 3°C for >100 days as required to complete obligatory larval diapause.
-As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA).
-Stubs were added to cages once B.
-inermis stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).
-
-### Data Collection
+#### Data Collection
 Stems were collected from each cage site in late August.
@@
 
-### Statistical Analysis
-Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions.
-‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs were used as the response variable to assess the differences in parameters among the sampling sites.
-ANOVA was used to calculate P-values using F-distribution to understand overall significant differences in mean C.
-cinctus infestation and parasitism among polygons.
-To assess the profile of C.
-cinctus infestation, parasitism, and larval presence by node, a linear model was fit for each sampling site.
-Further assessment was done on the responses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.
-
-## Organic Volatile Collection from C. cinctus Infested B. inermis and Winter Wheat
-### Insects
+## Greenhouse Experiments
+### Organic Volatile Collection from *C. cinctus* Infested *B. inermis* and Winter Wheat
+#### Insects
 Wheat stubble collected the year prior to each experiment were used to rear adult WSS.
-Wheat stubble was collected from Big Sandy (48.17754 N, -110.11406 E) and Three Forks (46.020216 N, -111.586582 E), MT, USA.
+Wheat stubble was collected from Big Sandy (48.17754°N, -110.11406°E) and Three Forks (46.020216°N, -111.586582°E), MT, USA.
 Stubble was collected from fields that experienced high levels of WSS infestation and cutting the year prior.
-Cut stubble, which contained WSS larvae in diapause, were kept in a refrigerator between 0-4 C for >100 days as required to complete obligatory larval diapause.
-After completing diapause, immature WSS were kept at room temperature between 22-27 C for 4-5 weeks in sealed 70 cm x 35 cm x 20 cm plastic Tupperware bins to allow for pupation and emergence.
+Cut stubble, which contained WSS larvae in diapause, were kept in a refrigerator between 0-4 °C for >100 days as required to complete obligatory larval diapause.
+After completing diapause, immature WSS were kept at room temperature between 22-27 °C for 4-5 weeks in sealed 70 cm x 35 cm x 20 cm plastic Tupperware$^\circledR$ bins to allow for pupation and emergence.
 Once emergence began, adults were collected daily and sorted in 2-liter glass mason jars sealed with filter paper to allow for airflow.
-Jars containing adults were refrigerated at 6-10C until they were needed for experiments.
+Jars containing adults were refrigerated at 6-10°C until they were needed for experiments.
 Unused adults were discarded after 48 hours of refrigeration.
 
-### Plants
-Smooth brome was grown from transplanted plants harvested from the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W).
-Immediately following collection, plants were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc, Tangent, OR) and moved into the greenhouse.
-Wheat was grown using ‘Yellowstone’ winter wheat cultivar seed stock, started in 1 x 8 inch conical pots.
+#### Plants
+Smooth brome was grown from transplanted stems harvested from the Arthur H.
+Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
+Immediately following collection, stems were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc$^\circledR$, Tangent, OR).
+Winter wheat plants were grown using Yellowstone Winter Wheat cultivar seed stock, started in 1 x 8 inch conical pots.
 The soil used consisted of MSU Plant Growth Center soil mix (equal parts sterilized Bozeman silt loam soil and washed concrete sand with Canadian sphagnum peat moss incorporated) and Sunshine Mix 1 (Canadian sphagnum peat moss, vermiculite, perlite, and Dolomite lime) in a 1:1 ratio.
-Fertilization was undergone once weekly using Jack’s Professional® Water-Soluble Fertilizer (20-20-20) (J.R.
+Fertilization was undergone once weekly using Jack’s Professional$^\circledR$ Water-Soluble Fertilizer (20-20-20) (J.R.
 Peters Inc., Allentown, PA, USA).
-The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22  2C and 20 – 40 % RH.
+The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22 $\pm$ 2°C and 20 – 40 % RH.
 After the plants reached Zadoks 12, they were moved into vernalization storage for 6 weeks to experience the length of dormancy needed for successful flower development.
-After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc conical pots.
+After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc$^\circledR$ conical pots.
 
-### Volatile Collection
+#### Sawfly Infestation
+Winter wheat plants were deemed ready for infestation once they reached Zadoks 49.
+Both smooth brome and winter wheat plants were infested by placing a transparent plastic cylinder (60 x 3.8 cm) over the main stem of the plant.
+A 55-60 cm bamboo stick was added within each tube to help prevent toppling of the cylinder.
+Each cage had three holes, as well as an open top, covered with fine mesh cloth to allow for air circulation.
+Four females and two male WSS, newly emerged, were introduced into each tube via a small 1 cm opening.
+After which, the hole was plugged using a cotton ball covered in mesh cloth.
+Damp soil was mounded around the base of each tube to prevent insect escapees.
+Adult sawflies were left within infestation tubes for three days, after which the tubes were removed, and any surviving insects, as well as cadavers, were removed.
+After infestation, the plants were returned to normal water and fertilization schedules.
+
+Infestation was undertaken three days a week, for five weeks.
+Each replicate included 6 ‘Yellowstone’ winter wheat plants at Zadoks 49 and 6 smooth brome plants exhibiting three nodes.
+For each replicate, three wheat and three smooth brome plants were infested, with three of each left uninfested.
+Infestation tubes were placed around all twelve plants to maintain control.
+
+#### Volatile Collection
 Volatile compounds were collected from intact and healthy smooth brome and Yellowstone winter wheat.
@@
 
-Volatile collection and successive GC-MS analysis were conducted following methods described in Weaver et al.
-(2009).
-Glass volatile collection traps were eluted using 200 µL of dichloromethane into 2 mL screw top vials (Agilent Technologies, Inc., Santa Clara, CA), followed by a slow release of nitrogen to further clear the volatile trap.
-10 µL of nonyl acetate solution was added to the eluted solution as internal standard to quantify compounds.
-Samples were then subject to gas chromatography (GC) on a J&W HP – 5ms; 30m x 0.25 µm film thickness column (Agilent Technologies, Inc., Santa Clara, CA).
-The GC unit (Agilent 6890; Agilent Technologies, Santa Clara, CA) was paired with a mass spectrometer (MS, Agilent 5973 instrument).
-Samples were injected into the column in pulse splitless mode, with the initial pressure of 82.7 kPa/m.
+#### Volatile Elution
+Volatile compounds were eluted from the glass traps using 200 $\mu$L aliquot dichloromethane.
+Trapped volatiles were eluted slowly by adding dichloromethane and further clearing by using a slow release of nitrogen gas.
+The eluted samples were collected in a glass insert held within a 1.5-mL crimped top glass vial.
+10 $\mu$L of nonyl acetate were then added to the eluted samples as internal standard to quantify compound concentrations.
+The samples were processed using gas chromatography (GC) on a HP – 5MS; 30m x 0.25mm, 0.25 $\mu$m film thickness column; (J and W Scientific, Folsom, CA).
+In order to analyze, the GC instrument (Agilent 6890; Agilent Technologies, Santa Clara, CA) was coupled to a mass spectrometer (MS, Agilent 5973 instrument).
+Each sample was injected using the automated system in pulse splitless mode, with the initial pressure set to 82.7 kPa per minute.
+The inlet temperature of the GC was set to 250°C while the column temperature was 50°C for 4 minutes.
+The temperature was set to increase at a rate of 5°C per minute until it reached 160°C, after which it increased to a rate of 25°C per minute until it reached 280°C.
+The temperature of the transfer line leading to the mass selective detector (MSD) was set at 300°C.
+Helium was used as a carrier gas to maintain the flow rate of samples within the column at 1.2 mL/min.
+The MSD was set in ‘SCAN’ mode running from 50 – 300 m/z.
 
-## PARASITOIDS
+#### Volatile Analysis
+The compounds collected from both the smooth brome and wheat samples were identified by comparing mass spectra and retention times using the National Institute of Standard and Technology (NIST) library.
+Peaks and volatiles were analyzed using MSD ChemStation (Agilent Technologies, Inc., Santa Clara, CA).
+The data obtained from the mass-spectrometer was transferred to a Microsoft® Excel spreadsheet where the compounds were quantified by comparison with the internal standard peak area.
+The amount obtained for each compound was then transformed to the amount produced in six hours.
+The true quantity of compounds was calculated relative to the internal standard.
+
+### Behavioral Assessment of Parasitoids in Response to *B. inermis* Organic Volatile Compounds
+#### Insects
 Parasitoids B.
 cephi and B.
-lissogaster were collected from wheat fields in Three Forks, MT (46.020216 N, -111.586582 E).
-Wheat stem residue from fields that experienced high levels of C.
-cinctus infestation were collected and stored inside 170-liter plastic barrel liners.
-Filled liners were stored in a cold storage room at 4C until mid or late spring of the following year.
+lissogaster were collected from wheat fields in Three Forks, MT (46.020216° N, -111.586582° E).
+Wheat stem residue from fields that experienced high levels of *C.
+cinctus* infestation were collected and stored inside 170-liter plastic barrel liners.
+Filled liners were stored in a cold storage room at 4°C until mid or late spring of the following year.
 After the storage time exceeded the mandatory duration needed for obligatory diapause, the plastic barrel liner was removed from cold storage and the residue was placed inside 170 L trash barrels at room temperature and watered to encourage emergence.
-Collected parasitoids were kept at 6-10C until used.
+Collected parasitoids were kept at 6-10°C until used.
 
-### Plants
-Smooth brome was grown from transplanted stems harvested from the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-Immediately following collection, stems were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc, Tangent, OR).
-Winter wheat plants were grown using Yellowstone Winter Wheat cultivar seed stock, started in 1 x 8 inch conical pots.
-The soil used consisted of MSU Plant Growth Center soil mix (equal parts sterilized Bozeman silt loam soil and washed concrete sand with Canadian sphagnum peat moss incorporated) and Sunshine Mix 1 (Canadian sphagnum peat moss, vermiculite, perlite, and Dolomite lime) in a 1:1 ratio.
-Fertilization was undergone once weekly using Jack’s Professional® Water-Soluble Fertilizer (20-20-20) (J.R.
-Peters Inc., Allentown, PA, USA).
-The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22  2C and 20 – 40 % RH.
-After the plants reached Zadoks 12, they were moved into vernalization storage for 6 weeks to experience the length of dormancy needed for successful flower development.
-After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc conical pots.
+#### Electroantennography
+Specimens were prepared by excising the head from the body using fine micro-dissection scissors.
+For each specimen, a micro-capillarity tube was pulled in half using a pipette puller (Model PP-830, Narishige Scientific Instrument Lab, Tokyo, Japan).
+Each half was cut to the appropriate length and angle, then filled with saline solution.
+The parasitoid head was mounted on the broken end of one of the capillary tubes, then the tube was place on one of the conducting electrode wires connected to a micromanipulator, type INR-5 (Narishige MN-151).
+The second half of the capillary tube was place on the other electrode wire, and one insect antennae was positioned to touch the empty capillary tube, thus connecting the circuit.
+The amplifier transferred the information to a computer interface with an EAG version 2.7 software (Syntech NL 2001, Hilversum, Netherlands).
+This apparatus and method has been consistently used previously in assessing insect volatile response [@Agelopoulos1999].
 
-## Behavioral Assessment of Parasitoids in Response to B. inermis Organic Volatile Compounds
-### Insects
-### Electroantennography
-### Y- Tube Olfactometer Bioassay
+A consistent stream of purified, humidified air (20 mL sec-1) was delivered onto the antennae.
+The test compound was delivered via airborne stimulus released from a glass pipette with filter paper treated with 10 $\mu$L of the test compound.
+The pipette was connected to a Stimulus Controller CS-55 (Syntech NL 2010, Hilversum, The Netherlands).
+The equipment was programmed to deliver a 0.4 second puff of air through a Teflon tube and through the glass pipette.
+The electrical response of the antenna to the test compound was detected by the recording electrode and transferred to the amplifier, increasing the signal and delivering the information to the computer.
+Figure 2: Electroantennogram connected to GCMS. Useful for connecting insect electrochemical response to host plant produced volatile compounds.
+
+#### Y- Tube Olfactometer Bioassay
 Y-tube bioassays were conducted within the chemical ecology and insect behavior laboratory at the Plant Growth Center at Montana State University.
-Bioassays were undergone to determine behavioral choice preference response of individual female WSS to an airstream containing volatiles originating from smooth brome and a blank control.
+Bioassays were undergone to determine behavioral choice preference response of individual female WSS to an air stream containing volatiles originating from smooth brome and a blank control.
 Bioassays were conducted using plants at growth stage Zadoks 49.
-The Y-tube system (Analytical Rsearch Systems, Micanopy, FL) used was similar as described in a previous study (Daisy et al., 2002) (Figure 1).
+The Y-tube system (Analytical Research Systems, Micanopy, FL) used was similar as described in a previous study [@Daisy2002] (Figure 1).
 A single plant was enclosed in a volatile collection chamber (VCC) (100 mm diameter, 254 mm length).
@@
 
-Figure 1: Y-tube olfactory bioassay as constructed by Daisy et al. (2002). Used to measure insect behavior in the presence of volatile stimulus.
-
 The Y-tube was made of bifurcated Corning® glass, measuring in at 28-mm diameter x 300 m length.
@@
 The arms of the Y-tube were oriented to minimize directional bias and the position of the plants was alternated between runs.
+
 For each assay, a single naïve female B.
@@
 All glassware was cleaned using dish soap and warm water, followed by solvent rinses with acetone and hexane.
-Cleaned and rinsed glassware was baked at 110C in a glassware oven for 24 hours before use.
+Cleaned and rinsed glassware was baked at 110°C in a glassware oven for 24 hours before use.
 New plants were used for each experiment.
+
 Each individual parasitoid was allowed three minutes to respond to either airstream source.
@@
 
-## Volatile Collection
-### Sawfly Infestation
-Winter wheat plants were deemed ready for infestation once they reached Zadoks 49.
-Both smooth brome and winter wheat plants were infested by placing a transparent plastic cylinder (60 x 3.8 cm) over the main stem of the plant.
-A 55-60 cm bamboo stick was added within each tube to help prevent toppling of the cylinder.
-Each cage had three holes, as well as an open top, covered with fine mesh cloth to allow for air circulation.
-Four females and two male WSS, newly emerged, were introduced into each tube via a small 1 cm opening.
-After which, the hole was plugged using a cotton ball covered in mesh cloth.
-Damp soil was mounded around the base of each tube to prevent insect escapees.
-Adult sawflies were left within infestation tubes for three days, after which the tubes were removed, and any surviving insects, as well as cadavers, were removed.
-After infestation, the plants were returned to normal water and fertilization schedules.
+# Data Analysis
+## Field Experiments
+### Field Survey
+Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions.
+‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs awere used as the response variable to assess the differences in parameters among the sampling sites.
+ANOVA was used to caculate P-values using F-distribution to understand overall significant differences in mean *C.
+cinctus* infestation and parasitism among polygons.
+To assess the profile of *C.
+cintus* infestation, paraisitms, and larval presence by node, a linear model was fit for each sampling site.
+Further assessment was done on the resonpses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.
 
-Infestation was undertaken three days a week, for five weeks.
-Each replicate included 6 ‘Yellowstone’ winter wheat plants at Zadoks 49 and 6 smooth brome plants exhibiting three nodes.
-For each replicate, three wheat and three smooth brome plants were infested, with three of each left uninfested.
-Infestation tubes were placed around all twelve plants to maintain control.
+### Controlled *C. cinctus* infestation of *B. inermis*
 
-### Volatile Compound Collection
-Volatile compounds were collected from intact and healthy smooth brome and ‘Yellowstone’ winter wheat.
-Collection was undergone on plants at Zadoks 49, for a period of six hours from 10:00 to 16:00.
-Collections were undergone three times a week, with twelve plants collected from each time.
-Each trail consisted of six smooth brome plants and six winter wheat, with three of each being infested, while the other three were a control.
+## Greenhouse Experiments
 
-Volatiles were collected from twelve plants per day using glass VCC’s that were 80 cm long and 4 cm in diameter.
-Volatile collection and successive GC-MS analysis were conducted following methods described in Weaver et al.
-(2009).
-The top end of each VCC funneled into a threaded port that was 10 cm in diameter.
-Threaded onto the top of each VCC was an eight-port manifold adapter.
-Two glass filters (traps) (6.35 mm diameter x 76 mm length; Analytical Research Systems, Gainesville, FL) with 30 mg of Super-Q absorbent (Alltech Associates, Deerfield, IL) were inserted into two of the adapter ports.
-The remaining six collection ports were sealed to prevent ambient air exchange and pressure loss.
-One of the traps was used to collected extraneous volatiles from the initial experiment set up for the first ten minutes, whereas the other trap was used to collect volatile compounds emanating from plants beginning ten minutes after the system was turned on.
-A vacuum pump was used to maintain air pressure and flow rate (humidified air at 1.0 liter/min) within the collection system.
-A Teflon guillotine was used at the base of each plant to tightly seal the base of the system and prevent external air from entering.
-Each plant stem was wrapped with a small piece of cotton prior to the application of the guillotine to further seal the base of the collection system and prevent contamination by ambient and soil derived volatiles.
+#### Organic Volatile Collection from *C. cinctus* Infested *B. inermis* and Winter Wheat
+The data analysis procedure for each collection type was the same.
+Statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PERMANOVA, PERMDISP, and Random Forest modeling.
+The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide].
+VOC compositions were compared between samples and treatments using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan].
+This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
+Data were visualized using a combination of ggplot2 and knitr [@ggplot; @knitr].
+I selected a subset of compounds to ues for each analysis.
+These compounds were selected based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output.
 
-### Volatile Elution
-Volatile compounds were eluted from the glass traps using 200 L aliquot dichloromethane.
-Trapped volatiles were eluted slowly by adding dichloromethane and further clearing by using a slow release of nitrogen gas.
-The eluted samples were collected in a glass insert held within a 1.5-mL crimped top glass vial.
-10 L of nonyl acetate were then added to the eluted samples as internal standard to quantify compound concentrations.
-The samples were processed using gas chromatography (GC) on a HP – 5MS; 30m x 0.25mm, 0.25 m film thickness column; (J and W Scientific, Folsom, CA).
-In order to analyze, the GC instrument (Agilent 6890; Agilent Technologies, Santa Clara, CA) was coupled to a mass spectrometer (MS, Agilent 5973 instrument).
-Each sample was injected using the automated system in pulse splitless mode, with the initial pressure set to 82.7 kPa per minute.
-The inlet temperature of the GC was set to 250C while the column temperature was 50C for 4 minutes.
-The temperature was set to increase at a rate of 5C per minute until it reached 160C, after which it increased to a rate of 25C per minute until it reached 280C.
-The temperature of the transfer line leading to the mass selective detector (MSD) was set at 300C.
-Helium was used as a carrier gas to maintain the flow rate of samples within the column at 1.2 mL/min.
-The MSD was set in ‘SCAN’ mode running from 50 – 300 m/z.
-
-### Volatile Analysis
-The compounds collected from both the smooth brome and wheat samples were identified by comparing mass spectra and retention times using the National Institute of Standard and Technology (NIST) library.
-Peaks and volatiles were analyzed using MSD ChemStation (Agilent Technologies, Inc., Santa Clara, CA).
-The data obtained from the mass-spectrometer was transferred to a Microsoft® Excel spreadsheet where the compounds were quantified by comparison with the internal standard peak area.
-The amount obtained for each compound was then transformed to the amount produced in six hours.
-The true quantity of compounds was calculated relative to the internal standard.
-
-### Electroantennogram
-Specimens were prepared by excising the head from the body using fine micro-dissection scissors.
-For each specimen, a micro-capillarity tube was pulled in half using a pipette puller (Model PP-830, Narishige Scientific Instrument Lab, Tokyo, Japan).
-Each half was cut to the appropriate length and angle, then filled with saline solution.
-The parasitoid head was mounted on the broken end of one of the capillary tubes, then the tube was place on one of the conducting electrode wires connected to a micromanipulator, type INR-5 (Narishige MN-151).
-The second half of the capillary tube was place on the other electrode wire, and one insect antennae was positioned to touch the empty capillary tube, thus connecting the circuit.
-The amplifier transferred the information to a computer interface with an EAG version 2.7 software (Syntech NL 2001, Hilversum, Netherlands).
-This apparatus and method has been consistently used previously in assessing insect volatile response (@Agelopoulos1999).
-
-A consistent stream of purified, humidified air (20 mL sec-1) was delivered onto the antennae.
-The test compound was delivered via airborne stimulus released from a glass pipette with filter paper treated with 10 L of the test compound.
-The pipette was connected to a Stimulus Controller CS-55 (Syntech NL 2010, Hilversum, The Netherlands).
-The equipment was programmed to deliver a 0.4 second puff of air through a Teflon tube and through the glass pipette.
-The electrical response of the antenna to the test compound was detected by the recording electrode and transferred to the amplifier, increasing the signal and delivering the information to the computer.
-Figure 2: Electroantennogram connected to GCMS. Useful for connecting insect electrochemical response to host plant produced volatile compounds.
-List of compounds used.
-
-## Data Analysis
-### Field Survey
-### Greenhouse Experiments
-
-#### Comparative Analysis of Collection of Volatiles
-The concentration of volatile compounds gathered was determined as nanogram per gram per hour (ng-1g1hr1), accounting for plant biomass adjustments.
-Data were analyzed using R Studio version 2023.12.0+369 (R Core team 2023).
-Data were first transformed via centered-log ratio (CLR) transformation, a common technique for compositional data  (Brückner and Heethoff, 2017).
-The data underwent both Permutational Analysis of Variance (PERMANOVA) and Permutational Analysis of Multivariate Dispersions (PERMDISP).
-PERMANOVA was used to determine differences among identified plant volatile compounds between the two plant types (Yellowstone winter wheat, B.
-inermis) at two infestation levels (Infested, control).
-PERMDISP was used to assess the variance of volatile quantities within our different groups.
-
-#### EAG-FID
-
-#### Organic Volatile Collections
-##### Centered Log-Ratio Transformation
-The data was transformed using a centered log-ratio transformation (CLR).
-This method is commonly used for multivariate compositional data (Brückner and Heethoff, 2017).
-CLR works by taking the natural logarithm of the ratios between the values of individual components and their geometric mean across samples.
-This helps to mitigate spurious correlations and allows the application of standard statistical techniques to our compositional data.
-
-##### PERMANOVA
+#### Behavioral Assessment of Parasitoids in Response to *B. inermis* Organic Volatile Compounds
 

### 2024-03-22 qmd/projects/msu-masters/locoweed/locoweed-report.qmd
commit for 3/22 morning

@@
 
-<p class="gallery-caption">Greenhouse Volatile Collection Plants looked like this.</p>
+<p class="gallery-caption">Greenhouse Locoweed volatiles were collected from plants that looked like these.</p>
 
@@
 Data were visualized using a combination of ggplot2 and knitr [@ggplot; @knitr].
-I selected a subset of comounds to ues for each analysis.
-These compounds were seelcted based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output.
-I will highlight the cmopounds that I used for each analysis as I get to them.
+I selected a subset of compounds to ues for each analysis.
+These compounds were selected based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output.
 
@@
 ```
-
+<!-- COMPOUNDS TO BE USED IN ANALYSIS -->
 ```{r COMPOUNDS, echo = F, message = F}
@@
 ```
+<!-- CLR TRANSFORMATION -->
 ```{R CLR, echo = F, message = F}
@@
 ### Boxplot Comparison
-```{r all comparison boxplot, echo = F, fig.align = 'center'}
 
+Eleven compounds were selected to display comparisons. 
+```{r all.comparison.plot, echo = F, fig.align = 'center', fig.width = 10}
+all.comparison.plot <-
 joined.data.clr %>%
+  rename("cis-3-Hexenyl Isovalerate" = cisthreehexiso,
+         "D-Limonene" = dlimonene,
+         "Dodecanal" = dodecanal,
+         "Heptanal" = heptanal,
+         "Mehthyl Salicylate" = methylsalicylate,
+         "Nonanal" = nonanal,
+         "Octanal" = octanal,
+         "6-methyl-5-hepten-2-one" = sixmethyl,
+         "Z-3-Hexenyl Acetate" = z3hex,
+         "B-Myrcene" = bmyrcene,
+         "B-Ocimene" = bocimene) %>%
   # pivot to make all compounds in one column
@@
         axis.line = element_line(color = "black"),
+        plot.margin = margin(0.5,0.5,0.5,1.5,"cm"),
         panel.grid.major.x = element_line(color = "#F2F2F2")) +
@@
                       labels = c("Not Present", "Present"))
+all.comparison.plot
+```
+```{r save all comparison plot, echo = F}
+ggsave(filename = "Field Locoweed Comparison BoxPlot.png",
+       plot = all.comparison.plot,
+       path = "locoweed_plots/",
+       width = 6,
+       height = 4)
+```
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
 
+<!-- Buttons to download the file -->
+<div class="button-container">
+  <a href="locoweed_plots/Field Locoweed Comparison BoxPlot.png" download>
+    <button class="button">Download</button>
+  </a>
+
+<div class="button-container">
+  <a href="toadflax_plots/Toadflax_plots/Toadflax PCA Scores - Type_All.png" download>
+    <button class="button">Download</button>
+  </a>
+</div>
+</div>
+</body>
+</html>
 ```

### 2024-03-22 qmd/projects.qmd
updated to fit better on mobile

@@
 ```
+```{=html}
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Box with Rounded Edges</title>
+  <style>
+    body {
+      margin: 0; /* Remove default margin */
+      padding: 0; /* Remove default padding */
+    }
+
+    .container {
+      width: 100%; /* Adjusted width to fit the screen */
+      max-width: 600px; /* Limiting maximum width */
+      height: auto; /* Set container height */
+      border: 0.5px solid #ccc; /* Light grey color */
+      border-radius: 8px; /* Less rounded edges */
+      overflow: hidden; /* Hide overflow content */
+      margin-bottom: 20px; /* Add margin to create space between rows */
+      display: flex; /* Use flexbox for layout */
+      align-items: center; /* Center items vertically */
+      text-decoration: none; /* Remove underline */
+    }
+
+    .container:last-child {
+      margin-bottom: 0; /* Remove margin from the last box */
+    }
+
+    .container:hover {
+      box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */
+    }
+
+    .image-container {
+      padding: 20px;
+      box-sizing: border-box;
+    }
+
+    .image-container img {
+      max-width: 100%; /* Ensures the image does not exceed its container */
+      height: auto;
+      display: block;
+      border-radius: 8px; /* Rounded corners for image */
+    }
+
+    .content {
+      padding: 20px;
+      box-sizing: border-box;
+      flex: 1; /* Take remaining space */
+    }
+
+    .text h1 {
+      font-size: 22px; /* Adjusted text size */
+      margin-top: 0;
+      margin-bottom: 5px; /* Adjust margin between large and small text */
+      color: black;
+    }
+
+    .text h1:hover {
+      color: #2068DC; /* Change text color to blue on hover */
+    }
+
+    .text p {
+      font-size: 16px;
+      margin-top: 5px;
+      color: #666;
+      text-align: justify; /* Align text */
+    }
+
+    @media (max-width: 768px) {
+      .container {
+        margin-right: auto; /* Reset margin-right */
+        margin-left: auto; /* Reset margin-left */
+        flex-direction: column; /* Stack items vertically on smaller screens */
+      }
+
+      .content {
+        padding: 10px; /* Reduce padding on smaller screens */
+      }
+    }
+  </style>
+</head>
+```
 

### 2024-03-24 qmd/projects/msu-masters/wss/eag.qmd
commit for 3/20
commit updated url for website

--- 2024-03-06 eag.qmd
@@
 eag <-
-  read_excel("xlsx files/eag.xlsx", sheet = "test eag")
+  read_excel("eag.xlsx", sheet = "test eag")
 
 fid <-
-  read_excel("xlsx files/eag.xlsx", sheet = "Peak")
+  read_excel("eag.xlsx", sheet = "Peak")
 
@@
   dplyr::select(MeasRetTime, Area, Height) %>%
-  mutate(Height = ifelse(MeasRetTime < 3.2, 0, Height))
+  mutate(Height = ifelse(MeasRetTime < 3.2, 0, Height)) %>%
+  mutate()
   # ggplot() +
@@
   rename("EAD" = 'EAD 1')
+
 ggplotly(
@@
             color = "red") +
-  geom_line(data = eag_plot, aes(x = scale_values(row_num)+1.2, 
+  geom_line(data = eag_plot %>%
+              mutate(EAD = ifelse(row_num == 42500, -1.5, EAD),
+                     EAD = ifelse(row_num == 42501, -1, EAD)), aes(x = scale_values(row_num)+0.2, 
                                  y = EAD * 100), 
@@
 
+# ggplotly(
+#   ggplot(data = eag_plot,
+#          aes(x = row_num,
+#              y = EAD)) +
+#     geom_line() +
+#     theme_bw()
+# )
+
+stimuli <- 
+eag_plot %>%
+  filter(row_num >= 19220 & row_num <= 19810)
 ```

### 2024-03-25 index.qmd
updated index files and yml(3/25)

@@
       href: https://linkedin.com/in/jacksonstrand
+    - icon: envelope-fill
+      text: Email
+      href: "mailto:jstrand894@gmail.com"
+    - icon: github
+      text: GitHub
+      href: https://github.com/jstrand894/
     - icon: instagram
@@
       href: https://www.instagram.com/jstrand894/
-    - icon: github
-      text: GitHub
-      href: https://github.com/jstrand894 
+
+toc: true
 ---
 ::: {#about-block}
+<span style="font-size: 30px;">**Jackson R. Strand**</span>
 :::
 
-# Welcome to my website!
+## About me
+I am currently a Master's graduate student at Montana State University studying entomology. I live in Bozeman, MT with my wife Grace.
 
-My name is Jackson Strand. I am currently a Master's graduate student at Montana State University studying entomology. I live in Bozeman, MT with my wife Grace.
+I am broadly interested in how insects impact the lives of humans and how we can better utilize them to positively impact modern issues such as crop yield, climate change, and waste management.
+
+**Montana State University Master's in Entomology**  
+My research focuses on understanding how the ecological relationship between the wheat stem sawfly (*Cephus cinctus*) and two native biocontrol parasitoids (*Bracon cephi* and *Bracon lissogaster*) are impacted by smooth brome (*Bromis inermis*).
+My project includes field sites in southwest, central, and northern Montana, as well as volatile organic compound analysis of greenhouse-grown smooth brome and winter wheat.
 
@@
 
-![Me and Grace at Glacier National Park, 2023](img/aboutme/iceberg.JPG)
+<!-- ![Me and Grace at Glacier National Park, 2023](img/aboutme/iceberg.JPG)   -->
+
+## Education
+
+<div class="box">
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>BS Biology</strong>
+  <em>- University of Minnesota</em>
+  </div>
+  <div style="float: right;">2013-2017
+  </div>
+</div>
+<br>
+
+- Minor: Marine Biology
+
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>MSc Entomology</strong>
+  <em>- Montana State University</em>
+  </div>
+  <div style="float: right;">2021-2024
+  </div>
+</div>
+<br>
+
+- Department of Land Resources and Environmental Sciences  
+- Graduate Certificate in Applied Statistics
+
+</div>
+
+## Research Experience
+<div class="box">
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>Montana State University</strong>
+  </div>
+  <div style="float: right;">2021-Present
+  </div>
+</div>
+<br>
+
+</div>
+
+## Teaching Experience
+<div class="box">
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>TA for graduate/undergraduate level integrated pest management course</strong>
+  </div>
+  <div style="float: right;">Fall 2023
+  </div>
+</div>
+<br>
+
+  - Communicated weekly reading schedule and discussions to students.
+  - Partook in online discussion on topics from class.
+  - Was the immediate/primary contact for students.
+
+</div>
+
+## Employment History
+<div class="box">
+<!-- Pace Analytical Work -->
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>
+    <a href="https://www.pacelabs.com" target="_blank" style="color: inherit; text-decoration: none;">Pace Analytical - Onsite at 3M
+    </a></strong> - <em>Regulatory IT Analyst</em>
+  </div>
+  <div style="float: right;">2017 - 2021
+  </div>
+</div>
+<br>
+(Maplewood, MN)
+
+- Provided digital support between 3M employees and the 3M Chemical Data Management System (CDMS).  
+- Integrated, tested, and supported user and data interface between CDMS and SAP.  
+- Generated, designed, and communicated unique, detailed reports of material information using SQL, Microsoft Excel, and SAP.
+
+<!-- Mosquito Control -->
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>
+    <a href="https://mmcd.org" target="_blank" style="color: inherit; text-decoration: none;">Metropolitan Mosquito Control District
+    </a></strong> - <em>Seasonal Mosquito Technician</em>
+  </div>
+  <div style="float: right;">2017
+  </div>
+</div>
+<br>
+(Oakdale, MN)
+
+- Worked on a team to collect and analyze mosquito larvae samples from wetland areas across the St. Paul metropolitan area.  
+- Independently maintained low adult mosquito populations in Oakdale, MN by surveying and treating both adults and larvae.
+
+<!-- Horse Barn -->
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>
+      <a href="https://vetmed.umn.edu/equine" target="_blank"
+      style="color: inherit; text-decoration: none;">Leatherdale Equine Center
+      </a>
+    </strong> - 
+      <em>Barn Maintenance Crew
+      </em>
+  </div>
+  <div style="float: right;">2015-2017
+  </div>
+</div>
+<br>
+(St. Paul, MN)
+
+- Maintained healthy conditions for equine patients at the University of Minnesota Horse Veterinary Center and Large Animal Hospital.  
+- Trained and managed new members of the crew.
+
+<!-- Soccer reffing -->
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>Minnesota Youth Soccer Association</strong> - <em>Youth Soccer Referee</em>
+  </div>
+  <div style="float: right;">2010 - 2021
+  </div>
+</div>
+<br>
+</div>
+
+## Honors & Awards
+<!-- Entomology 2023 talk win -->
+<div class="box">
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>President's Prize - 2nd Place</strong>
+  </div>
+  <div style="float: right;">2023
+  </div>
+</div>
+<br>
+
+- Entomology 2023, National Harbor, MD  
+- Entomological Society of America
+
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>Student Scholar Travel Award</strong> ($500)
+  </div>
+  <div style="float: right;">2023
+  </div>
+</div>
+<br>
+
+  - College of Agriculture, Montana State University
+
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>1st Place Speaker</strong>
+  </div>
+  <div style="float: right;">2023
+  </div>
+</div>
+<br>
+
+- 2023 LRES GSO Student Research Colloquium  
+- LRES Graduate Student Organization, Montana State University
+
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>Montana Grain Foundation Scholarship</strong> ($2,500)
+  </div>
+  <div style="float: right;">2022
+  </div>
+</div>
+<br>
+
+- Montana Grain Growers Association
+
+<div style="text-align: justify;">
+  <div style="float: left;"><strong>Professional Advancement Grant</strong> ($500)
+  </div>
+  <div style="float: right;">2021
+  </div>
+</div>
+<br>
+
+- The Graduate School, Montana State University
+</div>
+
+## Seminars & Presentations  
+<div class="box">
+**Paper Presentations**
+
+1. Strand, J., Weaver, D.K. 2024. *How smooth brome can be beneficial for
+parasitoids of the wheat stem sawfly.* March Madness – Montana Wheat and
+Barley Committee. Montana State University, Bozeman, MT.
+
+2. Sing, S.E., Ward, S.M., Peterson, R.K.D., Strand, J.R., Friesenhahn, E., Hofland,
+M., Hubbard, C.E., Randall, C.B., Weaver. D.K. 2024. *Biological Control of Hybrid Toadflax: Insights from Chemical Ecology.* Montana State University, Bozeman, MT.
+
+3.
+Strand, J., D.K.
+Weaver, T.
+Sterling, R.K.D.
+Peterson.
+2023.
+*Using organic volatile compounds to assess the relationship between wheat stem sawfly, parasitoids, and smooth brome.* Session: Grad Competition, Plant-Insect Ecosystems: Chemical Ecology 2.
+<u>2nd Place</u>.
+Annual Meeting of the Entomological Society of America, National Harbor, MD.
+
+4. Strand, J., D.K. Weaver, T. Sterling, R.K.D. Peterson. 2023. *The importance of smooth brome and parasitoids in wheat stem sawfly biocontrol.* <u>1st Place</u>. 2023 Land
+Resources and Environmental Sciences Graduate Student Organization
+Research Colloquium. Montana State University, Bozeman, MT.
+
+5. Strand, J., Weaver, D.K., Sterling, T., Peterson, R.K.D. 2022. *Using organic
+volatile compounds to assess the relationship between wheat stem sawfly, parasitoids, and smooth brome.* Session: Grad Competition, Plant-Insect Ecosystems: Chemical Ecology 1.
+Annual Meeting of the Entomological Society of America, Vancouver, BC, Canada.
+
+6.
+Strand, J.R., Weaver, D.K., Sterling, T., Peterson, R.K.D.
+2022.
+*Importance of parasitoids and smooth brome for integrated pest management of wheat stem sawfly.* 2022 Land Resources and Environmental Sciences Graduate Student Organization Research Colloquium, Montana State University, Bozeman, MT.
+
+7.
+Strand, J.R., Sterling, T., Weaver, D.K., 2022.
+*Using organic volatiles to understand the impact of Alternaria oxytropis on locoweed physiology: E+/E-.* Multistate Research Project Meeting, W2193, "Poisonous Plants: Impact, Ecology, and Management." Montana State University, Bozeman, MT.
+
+8.
+Strand, J.R., Weaver, D.K., Sterling, T., Peterson, R.K.D.
+2021.
+*Importance of parasitoids and smooth brome for integrated pest management of wheat stem sawfly.* Session: Plant Insect-Ecosystems: IPM – General.
+Annual Meeting of the Entomological Society of America, Denver, CO.
+
+**Workshops**
+
+1. *Understanding the Wheat Stem Sawfly*. 2022 Integrated Pest Management
+Workshop. Montana State University, Bozeman, MT.
+</div>
+
+## Service & Memberships
+<div class="box">
+**Society Memberships**
+<!-- GSO co-chair -->
+<div style="text-align: justify;">
+  <div style="float: left;">LRES Graduate Student Organization (Co-chair)
+  </div>
+  <div style="float: right;">2023 - Present
+  </div>
+</div>
+<br>
+
+<!-- GSO Tresurer-->
+<div style="text-align: justify;">
+  <div style="float: left;">LRES Graduate Student Organization (Treasurer)
+  </div>
+  <div style="float: right;">2021 - 2023
+  </div>
+</div>
+<br>
+
+<!-- entomological society of america -->
+<div style="text-align: justify;">
+  <div style="float: left;">Entomological Society of America
+  </div>
+  <div style="float: right;">2021 - Present
+  </div>
+</div>
+<br>
+
+<!-- MGGA -->
+<div style="text-align: justify;">
+  <div style="float: left;">Montana Grain Growers Association
+  </div>
+  <div style="float: right;">2021 - Present
+  </div>
+</div>
+<br>
+
+**Volunteer**
+<div style="text-align: justify;">
+  <div style="float: left;">Southwest Montana Mountain Bike Association
+  </div>
+  <div style="float: right;">2021 - Present
+  </div>
+</div>
+<br>
+
+<div style="text-align: justify;">
+  <div style="float: left;">Gallatin Watershed Council
+  </div>
+  <div style="float: right;">2021 - Present
+  </div>
+</div>
+<br>
+
+<div style="text-align: justify;">
+  <div style="float: left;">Gallatin Valley Land Trust
+  </div>
+  <div style="float: right;">2021 - Present
+  </div>
+</div>
+<br>
+
+<div style="text-align: justify;">
+  <div style="float: left;">Lake Elmo Regional Park Invasive Species Removal
+  </div>
+  <div style="float: right;">2021 - Present
+  </div>
+</div>
+<br>

### 2024-03-25 qmd/aboutme.qmd
updated index files and yml(3/25)

@@
 ```
-
-![Me and Grace at Glacier National Park, 2023](../img/aboutme/iceberg.JPG)
-
 :::
 
-<!-- ```{=html} -->
-<!-- <!DOCTYPE html> -->
-<!-- <html lang="en"> -->
-<!-- <head> -->
-<!--   <meta charset="UTF-8"> -->
-<!--   <meta name="viewport" content="width=device-width, initial-scale=1.0"> -->
-<!--   <title>Image Gallery</title> -->
-<!--   <style> -->
-<!--     .gallery { -->
-<!--       display: grid; -->
-<!--       grid-template-columns: repeat(auto-fill, minmax(200px, 1fr)); -->
-<!--       grid-gap: 10px; -->
-<!--     } -->
-<!--     .gallery img { -->
-<!--       width: 100%; -->
-<!--       height: auto; -->
-<!--       cursor: pointer; -->
-<!--       transition: transform 0.3s ease; -->
-<!--     } -->
-<!--     .gallery img:hover { -->
-<!--       transform: scale(1.05); /* Slightly increase size on hover */ -->
-<!--     } -->
-<!--     .expanded-container { -->
-<!--       position: fixed; -->
-<!--       top: 0; -->
-<!--       left: 0; -->
-<!--       width: 100%; -->
-<!--       height: 100%; -->
-<!--       background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */ -->
-<!--       display: none; /* Initially hidden */ -->
-<!--       justify-content: center; -->
-<!--       align-items: center; -->
-<!--       z-index: 9999; -->
-<!--     } -->
+## Education  
+<div class = "box">
+**BA Biology**, 2013-2017 - *University of Minnesota - Twin Cities*    
+    - Minor: Marine Biology
 
-<!--     .expanded { -->
-<!--       max-width: 80%; -->
-<!--       max-height: 80%; -->
-<!--       cursor: pointer; -->
-<!--     } -->
-<!--   </style> -->
-<!-- </head> -->
-<!-- <body> -->
+**MSc Entomology**, 2021-2024 - *Montana State University*
+</div>
 
-<!-- <div class="gallery"> -->
-<!--   <img src="../img/bugs/blueflowers.JPG" alt="Image 1" onclick="expandImage(this)"> -->
-<!--   <img src="../img/bugs/brownmoth.JPG" alt="Image 2" onclick="expandImage(this)"> -->
-<!--   <img src="../img/bugs/butterfly-purpleflower.JPG" alt="Image 3" onclick="expandImage(this)"> -->
-<!-- </div> -->
-
-<!-- <div class="expanded-container" onclick="closeExpanded()"> -->
-<!--   <img class="expanded" id="expanded-image" onclick="navigateGallery(event)"> -->
-<!-- </div> -->
-
-<!-- <script> -->
-<!--   var currentImageIndex = 0; -->
-<!--   var galleryImages = document.querySelectorAll('.gallery img'); -->
-
-<!--   function expandImage(img) { -->
-<!--     currentImageIndex = Array.from(galleryImages).indexOf(img); -->
-<!--     document.getElementById('expanded-image').src = img.src; -->
-<!--     document.querySelector('.expanded-container').style.display = 'flex'; -->
-<!--   } -->
-
-<!--   function closeExpanded() { -->
-<!--     document.querySelector('.expanded-container').style.display = 'none'; -->
-<!--   } -->
-
-<!--   function navigateGallery(event) { -->
-<!--     // Calculate the click position relative to the expanded image -->
-<!--     var clickX = event.pageX - event.target.getBoundingClientRect().left; -->
-<!--     var imageWidth = event.target.offsetWidth; -->
-
-<!--     // Determine if the click was on the left or right side of the image -->
-<!--     if (clickX < imageWidth / 2) { -->
-<!--       // Clicked on the left side: navigate to the previous image -->
-<!--       currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length; -->
-<!--     } else { -->
-<!--       // Clicked on the right side: navigate to the next image -->
-<!--       currentImageIndex = (currentImageIndex + 1) % galleryImages.length; -->
-<!--     } -->
-
-<!--     // Update the expanded image source -->
-<!--     document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src; -->
-<!--   } -->
-<!-- </script> -->
-
-<!-- </body> -->
-<!-- </html> -->
-
-<!-- ``` -->
+<!-- ![Me and Grace at Glacier National Park, 2023](../img/aboutme/iceberg.JPG)   -->

### 2024-03-25 qmd/projects.qmd
updated index files and yml(3/25)

@@
   chunk_output_type: console
+toc: true
 ---
+
+# Montana State University
+## Master's Project
+Understanding how smooth brome impacts the ecology of wheat stem sawfly and native biocontrol parasitoids.
+
+## Toadflax  
+Data analysis done for US Forest Service office out of Bozeman, MT.
+
+## Salt Cedar
 

### 2024-03-26 index.qmd
updated index page

@@
 ::: {#about-block}
-<span style="font-size: 30px;">**Jackson R. Strand**</span>
+<span style="font-size: 35px; font-family: 'Georgia', sans-serif;">**Jackson R. Strand**</span>
+
 :::
@@
 
+**My Research Interests**  
 I am broadly interested in how insects impact the lives of humans and how we can better utilize them to positively impact modern issues such as crop yield, climate change, and waste management.
 
-**Montana State University Master's in Entomology**  
+**Montana State University - MSc Entomology**  
 My research focuses on understanding how the ecological relationship between the wheat stem sawfly (*Cephus cinctus*) and two native biocontrol parasitoids (*Bracon cephi* and *Bracon lissogaster*) are impacted by smooth brome (*Bromis inermis*).
 My project includes field sites in southwest, central, and northern Montana, as well as volatile organic compound analysis of greenhouse-grown smooth brome and winter wheat.
+
+<!-- pictures below about me -->
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery-container {
+      display: flex;
+      justify-content: center;
+      width: 100%;
+    }
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
+      grid-gap: 10px;
+      max-width: 1200px;
+      width: 100%;
+    }
+    .gallery img {
+      width: 100%;
+      object-fit: cover; /* Ensure images cover the entire space */
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .gallery img:hover {
+      transform: scale(1.05); /* Slightly increase size on hover */
+    }
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
+      display: none; /* Initially hidden */
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+    p.gallery-caption {
+      text-align: center;
+      font-style: italic;
+      color: #8B8989;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+  <img src="img/index/wss.JPG" alt="Image 1" onclick="expandImage(this)">
+  <img src="img/index/pinecreek.jpeg" alt="Image 1" onclick="expandImage(this)">
+  <img src="img/index/loco-bee.JPG" alt="Image 1" onclick="expandImage(this)">
+  <img src="img/index/mushroom1.jpeg" alt="Image 1" onclick="expandImage(this)">
+  <img src="img/index/brownmoth.JPG" alt="Image 1" onclick="expandImage(this)">
+  <img src="img/index/glacierfall.jpeg" alt="Image 1" onclick="expandImage(this)">
+</div>
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    // Calculate the click position relative to the expanded image
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
+    var imageWidth = event.target.offsetWidth;
+
+    // Determine if the click was on the left or right side of the image
+    if (clickX < imageWidth / 2) {
+      // Clicked on the left side: navigate to the previous image
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      // Clicked on the right side: navigate to the next image
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    // Update the expanded image source
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+</script>
+
+</body>
+</html>
+
+```
 
@@
 
-- Department of Land Resources and Environmental Sciences  
+- Committee: Dr. David Weaver (co-chair), Dr. Tracy Sterling (co-chair), Dr. Robert Peterson
 - Graduate Certificate in Applied Statistics

### 2024-03-27 qmd/helpful.qmd
updated locoweed report files

---
title: "Helpful Resources & Links"
---

## R Studio
Great R-Markdown guide made by [Jared Adam](https://www.linkedin.com/in/jared-adam-24b107241/){style="text-decoration: none;"}: 
<a href="https://jared-adam-psu.github.io/R-Tutorial-for-Tooker-Lab/" target="_blank">One code to rule them all (┛ಠ_ಠ)┛彡O</a>

## Quarto Website
YouTube video on how to make a R Studio Quarto Website: [How to create a web site using Quarto, RStudio, and GitHub Pages](https://youtu.be/YN75YXaLFGM?si=RyCG23IMBwFUKsO8)

## GitHub
A great guide to using GitHub created by [Callum Arnold](https://www.linkedin.com/in/arnold-c/){style="text-decoration: none;"}: <a href="https://psu-git.callumarnold.com" target="_blank">Introduction to Git and GitHub</a>

### 2024-03-27 qmd/projects/msu-masters.qmd
updated locoweed report files

--- 2024-03-20 qmd/projects.qmd
@@
     color: #666;
-    text-align: justify; /* Align text */
+    <!-- text-align: justify; /* Align text */ -->
   }
@@
   </a>
-  <a href="projects/bballtool.qmd" class="container">
+  <a href="msu-masters/locoweed.qmd" class="container">
     <div class="image-container">
-      <img src="../img/projects-logos/bballref-logo-circle.jpg" alt="Image">
+      <img src="../../img/msu-masters/loco-bee2.jpg" alt="Image">
     </div>

### 2024-03-27 qmd/projects/msu-masters/locoweed.qmd
updated locoweed report files

---
title: "Organic Volatile Analysis of Locoweed"
toc: FALSE
---
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Image Cropping Example</title>
<style>
    }
    .image-container {
        width: auto; /* Set the width of the container */
        height: auto; /* Set the height of the container */
        overflow: hidden; /* Hide overflow */
    }
    .image-container img {
        width: 100%; /* Make the image fill the container horizontally */
        height: auto; /* Maintain aspect ratio */
        object-fit: cover;
        object-position: center;
    }
</style>
</head>
<body>

<div class="image-container">
    <img src="locoweed/img/loco-bee-crop-exp.JPG" alt="Your Image">
</div>

</body>
</html>

```

<span style="font-size: 20px;"><strong> Contributing Members </strong></span>  
This project is a subset of a project headed by 
[Dr. Tracy Sterling](http://weedeco.msu.montana.edu/faculty/tracy_sterling/){style="text-decoration: none;"}. 
In addition to Dr. Sterling, Rosie Wollander was immensely helpful in collecting data for all of the experiments. Thank you both for all of your help!

<div class="box">
<span style="font-size: 30px;"><strong> Project Overview </strong></span>  
Astragalus and Oxytropix species, part of the Legume Family, are found in terrestrial ecosystems globally, including the western regions of North America. Many of these species are both nontoxic and important forages. However, several species are known to be toxic to both livestock and wildlife. *Oxytropis sericea Nutt* can harbor a commensal endophyte, *Alternaira oxytropos*, a fungus that is associated with locoweed plants, specifically those of the genus *Oxytropis*. Plants containing the endophyte have been found to produce swansonine, a toxic alkaloid that can be harmful to livestock when ingested in large quantities. In this study, we investigated the differences in volatile organic compound (VOC) emissions between both field and greenhouse specimens that contained the commensal endophyte *Alternaria oxytropis*, and those that did not (E+/E-).

</div>
<br>

## Reports  
I have broken down the reports into two seperate, and more managable, reports. Both can be found below.

<!-- report buttons -->
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
  <title>Box with Rounded Edges</title>
  <style>
  .report-container {
    width: calc(50% - 20px);
    max-width: 500px;
    height: auto;
    border: 1.5px solid #ccc;
    border-radius: 8px;
    overflow: hidden;
    margin-bottom: 20px;
    margin-right: auto;
    margin-left: 10px;
    display: flex;
    align-items: center;
    text-decoration: none;
  }

  .report-container:last-child {
    margin-bottom: 0;
  }

  .report-container:hover {
    box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1);
  }

  .report-img-container {
    padding: 20px;
    box-sizing: border-box;
  }

  .report-img-container img {
    width: 150px;
    height: auto;
    display: block;
    border-radius: 8px;
  }

  .report-content {
    padding: 20px;
    box-sizing: border-box;
    flex: 1;
  }

  .report-text h1 {
    font-size: 30px;
    margin-top: 0;
    margin-bottom: 5px;
    color: black;
  }
  .report-text h1:hover {
    color: #2068DC;
  }

  .report-text p {
    font-size: 16px;
    margin-top: 5px;
    color: #666;
    text-align: justify;
  }

  @media (max-width: 768px) {
    .report-container {
      width: calc(100% - 20px);
    }
  }
</style>
</head>
<body>
  <a href="locoweed/field-locoweed-report.qmd" class="report-container">
    <div class="report-img-container">
      <img src="locoweed/img/bombus-loco.jpg" alt="Image">
    </div>
    <div class="report-content">
      <div class="report-text">
        <h1>Field</h1>
        <p>VOC Collection Report</p>
      </div>
    </div>
  </a>
  <a href="locoweed/gh-locoweed-report.qmd" class="report-container">
    <div class="report-img-container">
      <img src="locoweed/img/ghloco1-sq.jpg" alt="Image">
    </div>
    <div class="report-content">
      <div class="report-text">
        <h1>Greenhouse</h1>
        <p>VOC Collection Report</p>
      </div>
    </div>
  </a>
</body>
</html>

```

### 2024-03-27 qmd/projects/msu-masters/locoweed/field-locoweed-report.qmd
updated locoweed report files

---
title: "Field Locoweed VOC Collection Report"
author: "Jackson Strand"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
    code_folding: hide
bibliography: locoweed.bib
link-citations: true
cls: council-of-science-editors-author-date.csl
editor_options: 
  chunk_output_type: console
toc: true
---

```{r libraries, echo = F, message = F, warning = F}
library(tidyverse)
library(readxl)
library(MASS) # manova
library(knitr)
library(kableExtra)
library(compositions)
# PERMANOVA
library(vegan)
library(spaa)
library(devtools)
devtools::install_github("GuillemSalazar/EcolUtils")
library(EcolUtils)
library(gridExtra)

# Random Forest
library(ranger)
library(MVN) # for normality test
library(plotly)
```

<!-- bee cover image -->
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Image Cropping Example</title>
<style>
    }
    .image-container {
        width: auto; /* Set the width of the container */
        height: auto; /* Set the height of the container */
        overflow: hidden; /* Hide overflow */
    }
    .image-container img {
        width: 100%; /* Make the image fill the container horizontally */
        height: auto; /* Maintain aspect ratio */
        object-fit: cover;
        object-position: center;
    }
</style>
</head>
<body>

<div class="image-container">
    <img src="../../../../img/loco-bee-crop.JPG" alt="Your Image">
</div>

</body>
</html>

```

*Note: This project is still a work in progress.*

# Introduction
Astragalus and Oxytropix species, part of the Legume Family, are found in terestrial ecoysstems globally, including the western regions of North America. Many of these species are both nontoxic and important forages. However, several species are known to be toxic to both livestock and wildlife. *Oxytropis sericea* Nutt can harbor a commensal endophyte, *Alternaira oxytropos*, a fungus that is assocaited with locoweed plants, specifically those of the genus *Oxytropis*. Plants containing the endophyte have been found to produce swansonine, a toxic alkaloid that can be harmful to livestock when ingested in large quantities [@locoweed]. In this study, we investigated the potential differences in volatile organic compound (VOC) emissions between specimens that contained the commensal endophyte *Alternaria oxytropis*, and those that did not.

These data consist of volatile collection events over the course of two years, 2022 and 2023. In both years, data was collected from wild plants in the field, and plants grown in the greenhouse.

# Methodology
## Field Volatile Collection

Volatile collected was undergone on plants grown at a 34 x 60 ft plot at the Arthur H. Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT. Plants were inventoried and logged using GPS, and tissue samples from each specimen were analyzed to establish whether the plant contained the endophyte. Volatiles were collected from all the plants in each year over the course of one to two successive weeks.

Volatile were collected from 20 plants per day by placing Reynolds Turkey Oven Bags (Reynolds Group Holdings, Auckland, New Zealand) over the top of the selected number of leaflets.

Volatile collection and successive GC-MS analysis were conducted following methods described in Weaver et al. (2009). After the two hour collection period, glass volatile collection traps were eluted using 200 µL of dichloromethane into 2 mL screw top vials (Agilent Technologies, Inc., Santa Clara, CA), followed by a slow release of nitrogen to further clear the volatile trap. 10 µL of nonyl acetate solution was added to the eluted solution as internal standard to quantify compounds. Samples were then subject to gas chromatography (GC) on a J&W HP – 5ms; 30m x 0.25 µm film thickness column (Agilent Technologies, Inc., Santa Clara, CA). The GC unit (Agilent 6890; Agilent Technologies, Santa Clara, CA) was paired with a mass spectrometer (MS, Agilent 5973 instrument). Samples were injected into the column in pulse splitless mode, with the initial pressure of 82.7 kPa/m.

## Data Analysis  
The data analysis procedure for each collection type was the same. Statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PERMANOVA, PERMDISP, and Random Forest modeling. The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide]. VOC compositions were compared between samples and treatments using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan]. This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017]. Data were visualized using a combination of ggplot2 and knitr [@ggplot; @knitr]. I selected a subset of compounds to use for each analysis. These compounds were selected based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output. The locoweed samples, both SWA + and SWA -, exhibited numerous chromatogram peaks during collection. However, only volatile compounds showing significant peaks in the mass spectroscopy output or presenting differences in the overlays of the two SWA types were chosen for analysis.

# Results
<!-- LOAD DATA -->
```{r 2023 field data load, echo = F, message = F}
field.2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Field/2023\ Field/2023\ Field\ Locoweed.xlsx", sheet = "proc") 
```
```{r 2022 field data load, echo = F, message = F}
field.2022 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Field/2022\ Field/2022\ Field\ Locoweed.xlsx", sheet = "RAW proc") %>%
  dplyr::filter(!plant.id == "581")
```
<!-- CLEAN UP DATA -->
```{r 2023 field data cleanup, echo = F, message = F}
field.2023.data <-
field.2023 %>%
  filter(compound %in% c("heptanal", "sixmethyl", "octanal",
                         "z3hex", "dlimonene", "nonanal", 
                         "decanal", "methylsalicylate", "cisthreehexiso")) %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  left_join(field.2023 %>% 
              dplyr::select(plant.id, swa, type, pair2), # go back and add pairs later
            by = "plant.id") %>%
  replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa, type), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2023")
```
```{r 2022 field data cleanup, echo = F, message = F}
field.2022.data <-
field.2022 %>%
  filter(compound %in% c("heptanal", "sixmethyl", "octanal",
                         "z3hex", "dlimonene", "nonanal", 
                         "decanal", "methylsalicylate", "cisthreehexiso")) %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  left_join(field.2022 %>% 
              dplyr::select(plant.id, swa, collection.type, pair2), # go back and add pairs later
            by = "plant.id") %>%
  # replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa, collection.type), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2022") %>%
  rename(type = collection.type) %>%
  filter(!is.na(type) & type != 0) %>%
  # replace leaves and flowers with l and s
  mutate(type = ifelse(type == "leaves", "l", "f"))

```
<!-- COMPOUNDS TO BE USED IN ANALYSIS -->
```{r COMPOUNDS, echo = F, message = F}
compounds <- c('sixmethyl',
               'z3hex',
               'nonanal',
               'dlimonene',
               'methylsalicylate',
               # 'decanal',
               'cisthreehexiso',
               'dodecanal',
               # 'heptadecanal',
               'octanal',
               'bocimene',
               'bmyrcene',
               'heptanal')

# "heptanal", "sixmethyl", "octanal",
#                          "z3hex", "dlimonene", "nonanal", 
#                          "decanal", "methylsalicylate",
#                          "cisthreehexiso"
```
<!-- CLR TRANSFORMATION -->
```{R CLR, echo = F, message = F}
joined.data <-
field.2023 %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  filter(compound %in% compounds) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  dplyr::reframe(plant.id, compound, ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  mutate(year = "2023") %>%
  left_join(field.2023 %>%
              dplyr::select(plant.id, 
                            swa, type, pair2),
            by = "plant.id",
            relationship = "many-to-many") %>%
  mutate_at(vars(swa), list(factor)) %>%
  # rbind 2022 data
  rbind(field.2022 %>%
          dplyr::select(plant.id, compound, ngghr) %>%
          # filter(compound %in% c('sixmethyl','z3hex','nonanal',
          #                       'dlimonene','methylsalicylate',
          #                       'decanal','cisthreehexiso','dodecanal',
          #                       'heptadecanal','octanal','bocimene',
          #                       'bmyrcene','heptanal')) %>%
          group_by(plant.id, compound) %>%
          na.omit() %>%
          dplyr::reframe(plant.id, compound, 
                           ngghr = sum(ngghr)) %>%
          distinct() %>%
          pivot_wider(names_from = "compound",
                      values_from = "ngghr") %>%
          pivot_longer(-plant.id,
                       names_to = "compound",
                       values_to = "ngghr") %>%
          mutate(year = "2022") %>%
          left_join(field.2022 %>%
                      dplyr::select(plant.id,
                                    swa, collection.type, pair2) %>%
                      mutate_at(vars(swa), list(factor)),
                    by = "plant.id",
                    relationship = "many-to-many") %>%
          rename(type = collection.type)) %>%
  distinct()

# having trouble joining the two data sets because of the pair2 column 
joined.data.clr <-
joined.data %>%
  filter(compound %in% compounds) %>%
  dplyr::select(plant.id, ngghr, compound, year) %>%
  # replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  replace(is.na(.), 0) %>%
  mutate(ngghr = ifelse(ngghr == 0, 0.01, ngghr)) %>%
  mutate(ngghr = ngghr + 1) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
    # combine year and plant id into one column
  unite(plant.id, plant.id, year, sep = "_") %>%
  column_to_rownames("plant.id") %>%
  clr() %>%
  # put rownames back as a column
  as.data.frame() %>%
  rownames_to_column("plant.id") %>%
    # move year and plant id back to being sepearte columns
  separate(plant.id, into = c("plant.id", "year"), sep = "_") %>%
  pivot_longer(-c(plant.id, year),
               names_to = "compound",
               values_to = "ngghr") %>%
  # add a value of 2 to each ngghr to remove negatives
  mutate(ngghr = ngghr + 2) %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  # join
  left_join(joined.data %>%
              dplyr::select(plant.id, swa, year, type, pair2) %>%
              mutate(plant.id = as.character(plant.id)),
            by = c("plant.id","year")) %>%
  mutate_at(vars(swa, year), list(factor)) %>%
  distinct() %>%
  na.omit() %>%
  # i'm going to ignore the "stressed" 
  # designation and just make those type "leaf"
  mutate(type = recode(type, "leaves" = "l", "flowers" = "f", "s" = "l")) %>%
  filter(!type == 0) %>%
  distinct()
```

## Visual Comparisons
### Box Plot
Eleven compounds were selected to display comparisons. 
```{r all.comparison.plot, echo = F, fig.align = 'center', fig.width = 10}
all.comparison.plot <-
joined.data.clr %>%
  rename("cis-3-Hexenyl Isovalerate" = cisthreehexiso,
         "D-Limonene" = dlimonene,
         "Dodecanal" = dodecanal,
         "Heptanal" = heptanal,
         "Mehthyl Salicylate" = methylsalicylate,
         "Nonanal" = nonanal,
         "Octanal" = octanal,
         "6-methyl-5-hepten-2-one" = sixmethyl,
         "Z-3-Hexenyl Acetate" = z3hex,
         "B-Myrcene" = bmyrcene,
         "B-Ocimene" = bocimene) %>%
  # pivot to make all compounds in one column
  pivot_longer(-c(plant.id, year, swa, type, pair2),
               names_to = "compound",
               values_to = "ngghr") %>%
  # ggplot code
  ggplot(aes(compound, ngghr, fill = swa)) +
  geom_boxplot(position = position_dodge(width = 0.9),
                 outlier.color = NA,
                 alpha = 0.5) +
  geom_jitter(aes(compound, ngghr),
              alpha = 0.3,
              position = position_jitterdodge(jitter.width = 0.2,
                                              dodge.width = .8),
              show.legend = FALSE,
              # pch = 21, 
              stroke = 0, 
              size = 1.5) +
  theme(axis.text.x = element_text(angle = 45, 
                                    hjust = 1),
        legend.background = element_rect(color = "black"),
        # legend.position = c(0.8,0.9),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(color = "black"),
        plot.margin = margin(0.5,0.5,0.5,1.5,"cm"),
        panel.grid.major.x = element_line(color = "#F2F2F2")) +
  labs(x = element_blank(),
       y = "VOC quantity (CLR(ng/g/hr))",
       title = "2022/23 Field Locoweed Volatile Profile Comparison") +
  scale_fill_discrete(name = "Endophyte",
                      labels = c("Not Present", "Present"))
all.comparison.plot
```
```{r save all comparison plot, echo = F}
ggsave(filename = "Field Locoweed Comparison BoxPlot.png",
       plot = all.comparison.plot,
       path = "locoweed_plots/",
       width = 6,
       height = 4)
```
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
<meta charset="UTF-8">
<meta name="viewport" content="width=device-width, initial-scale=1.0">
<title>Download File Buttons</title>
<style>
/* Button styling */
.button {
  background-color: #2068DC; /* Header of website color (blue) */
  border: none;
  color: white;
  padding: 15px 32px;
  text-align: center;
  text-decoration: none;
  display: inline-block;
  font-size: 16px;
  margin: 4px 2px;
  cursor: pointer;
  border-radius: 8px;
  margin: 0 200px;
}
.button:hover {
  background-color: #3F91FB; /* Sky blue */
}
.button-container {
  display: flex;
  justify-content: center;
  margin-bottom: 10px;
}
</style>
</head>
<body>

<!-- Buttons to download the file -->
<div class="button-container">
  <a href="locoweed_plots/Field Locoweed Comparison BoxPlot.png" download>
    <button class="button">Download</button>
  </a>
</div>
</body>
</html>
```
:::{.callout-note}
I chose to visualize the data using a box plot, not a bar plot, because our data has been CLR transformed. As we can see, there isn't a tremendous amount of obvious variance between E+ and E- samples. This does not, however, mean there aren't differences when considering the overall volatile profiles.  
:::

### PCA
Principal Component Analysis (PCA) is a dimensionality reduction technique used to transform high-dimensional data into a lower-dimensional representation while preserving most of the original variability. It identifies the directions (principal components) that capture the maximum variance in the data and projects the data onto these components.

```{r PCA swa, echo = F}
pca.results <- 
joined.data.clr %>%
  dplyr::select(-c(year, plant.id, swa, type, pair2)) %>%
  prcomp(scale = T)

scores <- as.data.frame(pca.results$x)
loadings <- as.data.frame(pca.results$rotation)
variance <- as.data.frame(pca.results$sdev^2)

data.plus.pca <-
  joined.data.clr %>%
    cbind(scores[,1:2])

PC1 <- pca.results$x[,1]
PC2 <- pca.results$x[,2]

loadings.plot <-
loadings %>%
  dplyr::select(PC1, PC2) %>%
  rownames_to_column("compound")

meta <- joined.data.clr %>% dplyr::select(plant.id, type, year, swa, pair2)
all.pca <- data.frame(PC1, PC2, meta)

ggplot() +
  geom_point(data = all.pca,
             aes(PC1, PC2, color = swa),
             alpha = 0.8) +
  # geom_text(aes(label = day))
  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
  labs(x = "PC1", y = "PC2", 
       color = element_blank(),
       title = "Field Locoweed Volatile Collection PCA") +
  stat_ellipse(data = all.pca,
               aes(PC1, PC2, color = swa)) +
  theme_bw() +
  scale_color_discrete(name = "Endophyte",
                      labels = c("Not Present", "Present"))
```

Our PCA shows very little difference overall between the two groups of plants. Let's look at comparing PCA results if we change our response variable.

```{r PCA type, echo = F}
ggplot() +
  geom_point(data = all.pca,
             aes(PC1, PC2, color = type),
             alpha = 0.8) +
  # geom_text(aes(label = day))
  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
  labs(x = "PC1", y = "PC2", 
       color = element_blank(),
       title = "Field Locoweed Volatile Collection PCA - plant type") +
  stat_ellipse(data = all.pca,
               aes(PC1, PC2, color = type)) +
  theme_bw() +
  scale_color_discrete(name = "Plant Type",
                      labels = c("Flowers", "Leaves"))
```

It looks like from our plant type PCA that there is a difference in the variance of the volatile profile emitted by flowers compared to leaves. Let's break up our samples and run a separate PCA for flowers and for leaves and see if we notice a bigger difference between our E+ and E- groups.

```{r PCA separate, echo = F, fig.width = 10}
# flowers
pca.results <- 
joined.data.clr %>%
  filter(type == "f") %>%
  dplyr::select(-c(year, plant.id, swa, type, pair2)) %>%
  prcomp(scale = T)

scores <- as.data.frame(pca.results$x)
loadings <- as.data.frame(pca.results$rotation)
variance <- as.data.frame(pca.results$sdev^2)

data.plus.pca <-
  joined.data.clr %>%
  filter(type == "f") %>%
  cbind(scores[,1:2])

PC1 <- pca.results$x[,1]
PC2 <- pca.results$x[,2]

loadings.plot <-
loadings %>%
  dplyr::select(PC1, PC2) %>%
  rownames_to_column("compound")

meta <- joined.data.clr %>%
  filter(type == "f") %>%
  dplyr::select(plant.id, type, year, swa, pair2)
all.pca <- data.frame(PC1, PC2, meta)

flowers <-
ggplot() +
  geom_point(data = all.pca,
             aes(PC1, PC2, color = swa),
             alpha = 0.8) +
  # geom_text(aes(label = day))
  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
  labs(x = "PC1", y = "PC2", 
       color = element_blank(),
       title = "Field Locoweed Volatile Collection PCA",
       subtitle = "Flowers") +
  stat_ellipse(data = all.pca,
               aes(PC1, PC2, color = swa)) +
  theme_bw() +
  scale_color_discrete(name = "Endophyte",
                      labels = c("Not Present", "Present"))

# leaves
pca.results <- 
joined.data.clr %>%
  filter(type == "l") %>%
  dplyr::select(-c(year, plant.id, swa, type, pair2)) %>%
  prcomp(scale = T)

scores <- as.data.frame(pca.results$x)
loadings <- as.data.frame(pca.results$rotation)
variance <- as.data.frame(pca.results$sdev^2)

data.plus.pca <-
  joined.data.clr %>%
  filter(type == "l") %>%
  cbind(scores[,1:2])

PC1 <- pca.results$x[,1]
PC2 <- pca.results$x[,2]

loadings.plot <-
loadings %>%
  dplyr::select(PC1, PC2) %>%
  rownames_to_column("compound")

meta <- joined.data.clr %>%
  filter(type == "l") %>%
  dplyr::select(plant.id, type, year, swa, pair2)
all.pca <- data.frame(PC1, PC2, meta)

leaves <- 
ggplot() +
  geom_point(data = all.pca,
             aes(PC1, PC2, color = swa),
             alpha = 0.8) +
  # geom_text(aes(label = day))
  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
  labs(x = "PC1", y = "PC2", 
       color = element_blank(),
       title = "Field Locoweed Volatile Collection PCA",
       subtitle = "Leaves") +
  stat_ellipse(data = all.pca,
               aes(PC1, PC2, color = swa)) +
  theme_bw() +
  scale_color_discrete(name = "Endophyte",
                      labels = c("Not Present", "Present"))

grid.arrange(flowers,leaves, ncol = 2)
```

Even after splitting up our samples into flower and leaf subgroups, visually there does not appear to be a significant difference in volatile profiles between E+ and E-.

## Check Assumptions  
The statistical analyses we can perform are greatly impacted by our data's ability to meet statistical assumptions. One of the major assumptions here is normality. We are going to test this assumption using the `MVN` package. 
```{r assumptions, echo = F}
library(MVN)
kable(
  mvn(joined.data.clr %>%
        dplyr::select(-c(plant.id, swa, year, type, pair2)), 
      mvnTest = "mardia", 
      multivariatePlot = "qq")$univariateNormality,
  format = "html",
  caption = "Table 2: Assessment of Normality",
  align = "llccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f9")

```

<div class="box">
**Assumption Conclusions**  
From the Chi-Squared QQ-Plot (figure 2), we can see that our data set violates the normality assumption. The tail on the right half of the plot dramatically skews away from the normality line.

In addition, the results from the Anderson-Darling normality tests (table 1) reveal that every variable in individually non-normal.

As we can see, our data violates the normality assumption. This is the case even though we ran it though a CLR transformation. 
</div>
:::{.callout-note}
**A quick note here.**  
Many papers analyzing ecological or compositional data and VOC datasets often employ a statistical test known as [MANOVA](https://en.wikipedia.org/wiki/Multivariate_analysis_of_variance) (Multivariate Analysis of Variance). Despite the similarity of the data to ours in many instances, MANOVA is used even though the data being analyzed violates the normality assumption. Since our data clearly violates this assumption, we have decided against using this method.
:::

## Kruskal-Wallis Test
The [Kruskal-Wallis](https://en.wikipedia.org/wiki/Kruskal–Wallis_test) test is a non-parametric statistical test used to determine if there are differences between two or more independent groups. It assesses whether the distributions of scores across groups are the same or different, without assuming the data follow a normal distribution, making it suitable for ordinal or non-normally distributed data.

```{r kw test, echo = F, fig.width = 3}
compound.columns <- names(joined.data.clr)[sapply(joined.data.clr, is.numeric)]
p_values <- numeric()

for (col in compound.columns) {
  kruskal_result <- kruskal.test(joined.data.clr[[col]] ~ joined.data.clr$swa)
  p_values <- c(p_values, kruskal_result$p.value)
}

results_df <- data.frame(Column = compound.columns, P_Value = round(p_values,3)) %>%
  arrange(P_Value)

kable(results_df,
      format = "html",
      caption = "ANOVA results",
      align = "lc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover"))

```

Our KW results show what we might expect after seeing the box and whisker plot above (figure 1). None of the p-values are significant.

## PERMANOVA

As described above, a [PERMANOVA](https://en.wikipedia.org/wiki/Permutational_analysis_of_variance) (Permutational Multivariate Analysis of Variance) uses permutation testing to generate many different rearrangements of the data and assess if the observed differences are statistically significant [@anderson2017]. This method is important as it considers all of the volatile profiles overall when generating a p-value.

```{r PERMDISP & PERMANOVA data, echo = F}
set.seed(1)
permanova.data <-
  joined.data.clr %>%
  dplyr::select(-swa,-type,-year,-pair2) %>%
  column_to_rownames("plant.id") %>%
  na.omit() %>%
  mutate_all(~.^2) %>%
  mutate_all(~.*100000) %>%
  mutate_all(round, digits = 0)

# dist.matrix <-
# vegdist(t(permanova.data),
#         method = "canberra")

dist.matrix <-
vegdist(permanova.data,
        method = "canberra")

data.per.meta <-
  joined.data.clr %>%
  dplyr::select(swa,type,year,pair2) %>%
  mutate(year = factor(year, levels = c("2023", "2022")))
```
```{r 22/23 FIELD PERMANOVA, results = F}
set.seed(1)
permanova.output <-
adonis2(permanova.data ~ data.per.meta$swa, 
        strata = interaction(data.per.meta$pair2, 
                             data.per.meta$type, 
                             data.per.meta$year),
        method = "canberra",
        permutations = 999,
        by = "terms",
        na.rm = TRUE)

permanova.output
```
```{r PERMANOVA table, echo = F}

kable(as.data.frame(permanova.output),
      format = "html",
      caption = "Table 1: Field Locoweed VOC collection PERMANOVA results.   data.per.meta$swa is our E+/E- variable.",
      align = "lccccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f9")

```
<!-- Permanova P-values -->
```{r PERMANOVA p-values, echo  = F}
p.value.swa <- permanova.output$`Pr(>F)`[1]
# p.value.year <- permanova.output$`Pr(>F)`[2]
# p.value.type <- permanova.output$`Pr(>F)`[3]
```

## Random Forest

Next, we are going to use a random forest model to assess the importance of each of our compounds in distinguishing samples between E+ and E-. To do this we are going to use the `ranger` package in R.

```{r random forest, echo = F}
ranger_model <- ranger(swa ~ ., 
                       data = joined.data.clr %>%
                         dplyr::select(-plant.id),
                       importance = "impurity")

var_importance <- importance(ranger_model)

var_importance_df <- 
  data.frame(
    variables = names(var_importance),
    importance = var_importance) %>%
  arrange(desc(importance)) %>%
  rownames_to_column(var = "name") %>% dplyr::select(-name) %>%
  mutate(variables = factor(variables, levels = unique(variables)))

ggplotly(
ggplot(var_importance_df, aes(x = variables, y = importance)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL,
       y = "Variable Importance",
       title = "Random Forest Variable Importance")
)
```

<!-- There was no interaction between plant species and data of volatile collection.  -->

<!-- Our results show **weak** evidence of a difference between SWA+ and SWA- locoweed plants (p-val = `r p.value.swa`).   -->

### 2024-03-27 qmd/projects/msu-masters/locoweed/gh-locoweed-report.qmd
updated locoweed report files

--- 2024-03-20 qmd/projects/msu-masters/locoweed/locoweed-report.qmd
@@
 ---
-title: "Locoweed Volatiles Report"
+title: "Greenhouse Locoweed VOC Collection Report"
 author: "Jackson Strand"
@@
 library(plotly)
+
+set.seed(1)
 ```
@@
 <div class="image-container">
-    <img src="../../../../img/loco-bee-crop.JPG" alt="Your Image">
+    <img src="img/vcs.jpg" alt="Your Image">
 </div>
@@
 
-# INTRODUCTION
-Astragalus and Oxytropix species, part of the Legume Family, are found in terestrial ecoysstems globally, including the western regions of North America.
+# Introduction
+Astragalus and Oxytropix species, part of the Legume Family, are found in terrestrial ecoysstems globally, including the western regions of North America.
 Many of these species are both nontoxic and important forages.
 However, several species are known to be toxic to both livestock and wildlife.
-*Oxytropis sericea Nutt* can harbor a commensal endophyte, *Alternaira oxytropos*, a fungus that is assocaited with locoweed plants, specifically those of the genus *Oxytropis*.
+*Oxytropis sericea* Nutt can harbor a commensal endophyte, *Alternaira oxytropos*, a fungus that is associated with locoweed plants, specifically those of the genus *Oxytropis*.
 Plants containing the endophyte have been found to produce swansonine, a toxic alkaloid that can be harmful to livestock when ingested in large quantities [@locoweed].
@@
 
-# METHODOLOGY
+# Methodology
 ## Data Collection
-### Field Volatile Collection
-
-Volatile collected was undergone on plants grown at a 34 x 60 ft plot at the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-Plants were inventoried and logged using GPS, and tissue samples from each specimen were analyzed to establish whether the plant contained the endophyte.
-Volatiles were collected from all the plants in each year over the course of one to two successive weeks.
-
-Volatile were collected from 20 plants per day by placing Reynolds Turkey Oven Bags (Reynolds Group Holdings, Auckland, New Zealand) over the top of the selected number of leaflets.
 
@@
 
+<div class="box">
+<!-- plant pictures -->
 ```{r plant pictures, echo = F, eval = F}
@@
   <style>
+    .gallery-container {
+      display: flex;
+      justify-content: center;
+      align-items: center;
+      height: 100%;
+    }
     .gallery {
@@
     p.gallery-caption {
+      padding-top: 5px;
       text-align: center;
@@
 </html>
+```
+</div>
 
-```
-
-<br>
 Volatiles were collected from twleve plants per day using glass volatile collection chambers (VCCs) that were 30 cm long and 4 cm in diameter.
@@
 Data were visualized using a combination of ggplot2 and knitr [@ggplot; @knitr].
-I selected a subset of comounds to ues for each analysis.
-These compounds were seelcted based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output.
-I will highlight the cmopounds that I used for each analysis as I get to them.
+I selected a subset of compounds to ues for each analysis.
+These compounds were selected based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output.
 
 # RESULTS
-<!-- LOAD DATA -->
-```{r 2023 field data load, echo = F, message = F}
-field.2023 <-
-read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Field/2023\ Field/2023\ Field\ Locoweed.xlsx", sheet = "proc") 
-```
-```{r 2022 field data load, echo = F, message = F}
-field.2022 <-
-read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Field/2022\ Field/2022\ Field\ Locoweed.xlsx", sheet = "RAW proc") %>%
-  dplyr::filter(!plant.id == "581")
-```
-<!-- CLEAN UP DATA -->
-```{r 2023 field data cleanup, echo = F, message = F}
-field.2023.data <-
-field.2023 %>%
-  filter(compound %in% c("heptanal", "sixmethyl", "octanal",
-                         "z3hex", "dlimonene", "nonanal", 
-                         "decanal", "methylsalicylate", "cisthreehexiso")) %>%
-  dplyr::select(plant.id, compound, ngghr) %>%
-  group_by(plant.id, compound) %>%
-  na.omit() %>%
-  dplyr::reframe(ngghr = sum(ngghr)) %>%
-  distinct() %>%
-  pivot_wider(names_from = "compound",
-              values_from = "ngghr") %>%
-  pivot_longer(-plant.id,
-               names_to = "compound",
-               values_to = "ngghr") %>%
-  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
-  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
-  pivot_wider(names_from = "compound",
-              values_from = "ngghr") %>%
-  left_join(field.2023 %>% 
-              dplyr::select(plant.id, swa, type, pair2), # go back and add pairs later
-            by = "plant.id") %>%
-  replace(is.na(.), 0) %>%
-  distinct() %>%
-  mutate_at(vars(swa, type), list(factor)) %>%
-  ungroup() %>%
-  mutate(year = "2023")
-```
-```{r 2022 field data cleanup, echo = F, message = F}
-field.2022.data <-
-field.2022 %>%
-  filter(compound %in% c("heptanal", "sixmethyl", "octanal",
-                         "z3hex", "dlimonene", "nonanal", 
-                         "decanal", "methylsalicylate", "cisthreehexiso")) %>%
-  dplyr::select(plant.id, compound, ngghr) %>%
-  group_by(plant.id, compound) %>%
-  na.omit() %>%
-  dplyr::reframe(ngghr = sum(ngghr)) %>%
-  distinct() %>%
-  pivot_wider(names_from = "compound",
-              values_from = "ngghr") %>%
-  pivot_longer(-plant.id,
-               names_to = "compound",
-               values_to = "ngghr") %>%
-  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
-  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
-  pivot_wider(names_from = "compound",
-              values_from = "ngghr") %>%
-  left_join(field.2022 %>% 
-              dplyr::select(plant.id, swa, collection.type, pair2), # go back and add pairs later
-            by = "plant.id") %>%
-  # replace(is.na(.), 0) %>%
-  distinct() %>%
-  mutate_at(vars(swa, collection.type), list(factor)) %>%
-  ungroup() %>%
-  mutate(year = "2022") %>%
-  rename(type = collection.type) %>%
-  filter(!is.na(type) & type != 0) %>%
-  # replace leaves and flowers with l and s
-  mutate(type = ifelse(type == "leaves", "l", "f"))
-
-```
-
-```{r COMPOUNDS, echo = F, message = F}
-compounds <- c('sixmethyl',
-               'z3hex',
-               'nonanal',
-               'dlimonene',
-               'methylsalicylate',
-               # 'decanal',
-               'cisthreehexiso',
-               'dodecanal',
-               # 'heptadecanal',
-               'octanal',
-               'bocimene',
-               'bmyrcene',
-               'heptanal')
-
-# "heptanal", "sixmethyl", "octanal",
-#                          "z3hex", "dlimonene", "nonanal", 
-#                          "decanal", "methylsalicylate",
-#                          "cisthreehexiso"
-```
-```{R CLR, echo = F, message = F}
-joined.data <-
-field.2023 %>%
-  dplyr::select(plant.id, compound, ngghr) %>%
-  filter(compound %in% compounds) %>%
-  group_by(plant.id, compound) %>%
-  na.omit() %>%
-  dplyr::reframe(plant.id, compound, ngghr = sum(ngghr)) %>%
-  distinct() %>%
-  pivot_wider(names_from = "compound",
-              values_from = "ngghr") %>%
-  pivot_longer(-plant.id,
-               names_to = "compound",
-               values_to = "ngghr") %>%
-  mutate(year = "2023") %>%
-  left_join(field.2023 %>%
-              dplyr::select(plant.id, 
-                            swa, type, pair2),
-            by = "plant.id",
-            relationship = "many-to-many") %>%
-  mutate_at(vars(swa), list(factor)) %>%
-  # rbind 2022 data
-  rbind(field.2022 %>%
-          dplyr::select(plant.id, compound, ngghr) %>%
-          # filter(compound %in% c('sixmethyl','z3hex','nonanal',
-          #                       'dlimonene','methylsalicylate',
-          #                       'decanal','cisthreehexiso','dodecanal',
-          #                       'heptadecanal','octanal','bocimene',
-          #                       'bmyrcene','heptanal')) %>%
-          group_by(plant.id, compound) %>%
-          na.omit() %>%
-          dplyr::reframe(plant.id, compound, 
-                           ngghr = sum(ngghr)) %>%
-          distinct() %>%
-          pivot_wider(names_from = "compound",
-                      values_from = "ngghr") %>%
-          pivot_longer(-plant.id,
-                       names_to = "compound",
-                       values_to = "ngghr") %>%
-          mutate(year = "2022") %>%
-          left_join(field.2022 %>%
-                      dplyr::select(plant.id,
-                                    swa, collection.type, pair2) %>%
-                      mutate_at(vars(swa), list(factor)),
-                    by = "plant.id",
-                    relationship = "many-to-many") %>%
-          rename(type = collection.type)) %>%
-  distinct()
-
-# having trouble joining the two data sets because of the pair2 column 
-joined.data.clr <-
-joined.data %>%
-  filter(compound %in% compounds) %>%
-  dplyr::select(plant.id, ngghr, compound, year) %>%
-  # replace(is.na(.), 0) %>% filter(compound != "NA") %>%
-  mutate(ngghr = ifelse(ngghr == 0, 0.01, ngghr)) %>%
-  mutate(ngghr = ngghr + 1) %>%
-  pivot_wider(names_from = "compound",
-              values_from = "ngghr") %>%
-    # combine year and plant id into one column
-  unite(plant.id, plant.id, year, sep = "_") %>%
-  column_to_rownames("plant.id") %>%
-  clr() %>%
-  # put rownames back as a column
-  as.data.frame() %>%
-  rownames_to_column("plant.id") %>%
-    # move year and plant id back to being sepearte columns
-  separate(plant.id, into = c("plant.id", "year"), sep = "_") %>%
-  pivot_longer(-c(plant.id, year),
-               names_to = "compound",
-               values_to = "ngghr") %>%
-  # add a value of 2 to each ngghr to remove negatives
-  mutate(ngghr = ngghr + 2) %>%
-  pivot_wider(names_from = "compound",
-              values_from = "ngghr") %>%
-  # join
-  left_join(joined.data %>%
-              dplyr::select(plant.id, swa, year, type, pair2) %>%
-              mutate(plant.id = as.character(plant.id)),
-            by = c("plant.id","year")) %>%
-  mutate_at(vars(swa, year), list(factor)) %>%
-  distinct() %>%
-  na.omit() %>%
-  # i'm going to ignore the "stressed" 
-  # designation and just make those type "leaf"
-  mutate(type = recode(type, "leaves" = "l", "flowers" = "f", "s" = "l")) %>%
-  filter(!type == 0) %>%
-  distinct()
-
-```
-
-## Field Volatile Collection
-### Boxplot Comparison
-```{r all comparison boxplot, echo = F, fig.align = 'center'}
-
-joined.data.clr %>%
-  # pivot to make all compounds in one column
-  pivot_longer(-c(plant.id, year, swa, type, pair2),
-               names_to = "compound",
-               values_to = "ngghr") %>%
-  # ggplot code
-  ggplot(aes(compound, ngghr, fill = swa)) +
-  geom_boxplot(position = position_dodge(width = 0.9),
-                 outlier.color = NA,
-                 alpha = 0.5) +
-  geom_jitter(aes(compound, ngghr),
-              alpha = 0.3,
-              position = position_jitterdodge(jitter.width = 0.2,
-                                              dodge.width = .8),
-              show.legend = FALSE,
-              # pch = 21, 
-              stroke = 0, 
-              size = 1.5) +
-  theme(axis.text.x = element_text(angle = 45, 
-                                    hjust = 1),
-        legend.background = element_rect(color = "black"),
-        # legend.position = c(0.8,0.9),
-        panel.background = element_rect(fill = "white"),
-        axis.line = element_line(color = "black"),
-        panel.grid.major.x = element_line(color = "#F2F2F2")) +
-  labs(x = element_blank(),
-       y = "VOC quantity (CLR(ng/g/hr))",
-       title = "2022/23 Field Locoweed Volatile Profile Comparison") +
-  scale_fill_discrete(name = "Endophyte",
-                      labels = c("Not Present", "Present"))
-
-```
-
-### PERMANOVA
-```{r PERMDISP & PERMANOVA data, echo = F}
-set.seed(1)
-permanova.data <-
-  joined.data.clr %>%
-  dplyr::select(-swa,-type,-year,-pair2) %>%
-  column_to_rownames("plant.id") %>%
-  na.omit() %>%
-  mutate_all(~.^2) %>%
-  mutate_all(~.*100000) %>%
-  mutate_all(round, digits = 0)
-
-# dist.matrix <-
-# vegdist(t(permanova.data),
-#         method = "canberra")
-
-dist.matrix <-
-vegdist(permanova.data,
-        method = "canberra")
-
-data.per.meta <-
-  joined.data.clr %>%
-  dplyr::select(swa,type,year,pair2) %>%
-  mutate(year = factor(year, levels = c("2023", "2022")))
-```
-```{r 22/23 FIELD PERMANOVA, echo = F, results = F}
-set.seed(1)
-permanova.output <-
-adonis2(permanova.data ~ data.per.meta$swa, 
-        strata = interaction(data.per.meta$pair2, 
-                             data.per.meta$type, 
-                             data.per.meta$year),
-        method = "canberra",
-        permutations = 999,
-        by = "terms",
-        na.rm = TRUE)
-permanova.output
-```
-```{r PERMANOVA p-values, echo  = F}
-p.value.swa <- permanova.output$`Pr(>F)`[1]
-# p.value.year <- permanova.output$`Pr(>F)`[2]
-# p.value.type <- permanova.output$`Pr(>F)`[3]
-```
-
-The locoweed samples, both SWA + and SWA -, exhibited numerous chromatogram peaks during collection.
-However, only volatile compoudns showing significant peaks inthe mass spectroscopy output or presenting differences in the overlays of the two SWA types were chosen for analysis.
-
-There was no interaction between plant species and data of volatile collection.
-
-Our results show **weak** evidence of a difference between SWA+ and SWA- locoweed plants (p-val = `r p.value.swa`).
-
-## Greenhouse Volatile Collection
+## Visual Comparisons
 <!-- LOAD DATA -->
@@
   # replace(is.na(.), 0) %>% filter(compound != "NA") %>%
+  replace(is.na(.), 0) %>%
   pivot_wider(names_from = "compound",
@@
 
-### Boxplot Comparison
-```{r gh boxplot comparison, echo = F}
+### Box Plot
+```{r gh boxplot comparison, echo = F, fig.align = 'center', fig.width = 10}
 gh.joined.data.clr %>%
@@
        y = "VOC quantity (CLR(ng/g/hr))",
-       title = "2022/23 Field Locoweed Volatile Profile Comparison") +
+       title = "2022/23 Greenhouse Locoweed Volatile Profile Comparison") +
   scale_fill_discrete(name = "Endophyte",
@@
 ```
+:::{.callout-note}
+I chose to visualize the data using a box plot, not a bar plot, because our data has been CLR transformed.
+As we can see, there isn't a tremendous amount of obvious variance between E+ and E- samples.
+There is, perhaps, more of a difference at first glance than what we saw in our [field volatile analysis](field-locoweed-report.qmd#visual-comparisons) bar plots.
+:::
+
+### PCA
+Principal Component Analysis (PCA) is a dimensionality reduction technique used to transform high-dimensional data into a lower-dimensional representation while preserving most of the original variability.
+It identifies the directions (principal components) that capture the maximum variance in the data and projects the data onto these components.
+
+```{r PCA swa, echo = F}
+pca.results <- 
+gh.joined.data.clr %>%
+  dplyr::select(-c(year, plant.id, swa, pair)) %>%
+  prcomp(scale = T)
+
+scores <- as.data.frame(pca.results$x)
+loadings <- as.data.frame(pca.results$rotation)
+variance <- as.data.frame(pca.results$sdev^2)
+
+data.plus.pca <-
+  gh.joined.data.clr %>%
+    cbind(scores[,1:2])
+
+PC1 <- pca.results$x[,1]
+PC2 <- pca.results$x[,2]
+
+loadings.plot <-
+loadings %>%
+  dplyr::select(PC1, PC2) %>%
+  rownames_to_column("compound")
+
+meta <- gh.joined.data.clr %>% dplyr::select(plant.id, year, swa, pair)
+all.pca <- data.frame(PC1, PC2, meta)
+
+ggplot() +
+  geom_point(data = all.pca,
+             aes(PC1, PC2, color = swa),
+             alpha = 0.8) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       color = element_blank(),
+       title = "Field Locoweed Volatile Collection PCA") +
+  stat_ellipse(data = all.pca,
+               aes(PC1, PC2, color = swa)) +
+  theme_bw() +
+  scale_color_discrete(name = "Endophyte",
+                      labels = c("Not Present", "Present"))
+```
+
+Our PCA shows very little difference overall between the two groups of plants. Let's look at comparing PCA results if we change our response variable.
+
+```{r PCA type, echo = F}
+ggplot() +
+  geom_point(data = all.pca,
+             aes(PC1, PC2, color = year),
+             alpha = 0.8) +
+  # geom_text(aes(label = day))
+  geom_hline(yintercept = 0, color = "gray", linetype = "dashed") +
+  geom_vline(xintercept = 0, color = "gray", linetype = "dashed") +
+  labs(x = "PC1", y = "PC2", 
+       color = element_blank(),
+       title = "Field Locoweed Volatile Collection PCA - plant type") +
+  stat_ellipse(data = all.pca,
+               aes(PC1, PC2, color = year)) +
+  theme_bw()
+```
+
+Interesting. We are seeing a little bit of separation when we compare our collection years. We should remember this going forward, as we don't want to capture the year-to-year difference.
+
+## Check Assumptions  
+The statistical analyses we can perform are greatly impacted by our data's ability to meet statistical assumptions.
+One of the major assumptions here is normality.
+We are going to test this assumption using the `MVN` package.
+
+```{r assumptions, echo = F}
+library(MVN)
+kable(
+  mvn(gh.joined.data.clr %>%
+        dplyr::select(-c(plant.id, swa, year, pair)), 
+      mvnTest = "mardia", 
+      multivariatePlot = "qq")$univariateNormality,
+  format = "html",
+  caption = "Table 2: Assessment of Normality",
+  align = "llccc") %>%
+  kable_styling(full_width = FALSE,
+                bootstrap_options = c("striped", "bordered", "hover")) %>%
+  row_spec(0, background = "#f2f2f9")
+
+```
+
+<div class="box">
+**Assumption Conclusions**  
+From the Chi-Squared QQ-Plot (figure 2), we can see that our data set violates the normality assumption. The tail on the right half of the plot dramatically skews away from the normality line.
+
+In addition, the results from the Anderson-Darling normality tests (table 1) reveal that every variable in individually non-normal.
+
+As we can see, our data violates the normality assumption. This is the case even though we ran it though a CLR transformation. 
+</div>
+
+:::{.callout-note}
+**A quick note here.**  
+Many papers analyzing ecological or compositional data and VOC datasets often employ a statistical test known as [MANOVA](https://en.wikipedia.org/wiki/Multivariate_analysis_of_variance) (Multivariate Analysis of Variance).
+Despite the similarity of the data to ours in many instances, MANOVA is used even though the data being analyzed violates the normality assumption.
+Since our data clearly violates this assumption, we have decided against using this method.
+:::
+
+## Kruskal-Wallis Test
+The [Kruskal-Wallis](https://en.wikipedia.org/wiki/Kruskal–Wallis_test) test is a non-parametric statistical test used to determine if there are differences between two or more independent groups.
+It assesses whether the distributions of scores across groups are the same or different, without assuming the data follow a normal distribution, making it suitable for ordinal or non-normally distributed data.
+
+```{r kw test, echo = F, fig.width = 3}
+compound.columns <- names(gh.joined.data.clr)[sapply(gh.joined.data.clr, is.numeric)]
+p_values <- numeric()
+
+for (col in compound.columns) {
+  kruskal_result <- kruskal.test(gh.joined.data.clr[[col]] ~ gh.joined.data.clr$swa)
+  p_values <- c(p_values, kruskal_result$p.value)
+}
+
+results_df <- data.frame(Column = compound.columns, P_Value = round(p_values,3)) %>%
+  arrange(P_Value)
+
+kable(results_df,
+      format = "html",
+      caption = "ANOVA results",
+      align = "lc") %>%
+  kable_styling(full_width = FALSE,
+                bootstrap_options = c("striped", "bordered", "hover"))
+
+```
+
+We are seeing a few interesting numbers here. Cyclohexene and $\beta$-ocimene have p-values close to 0.05.
 
 ### PERMANOVA  
+As described above, a [PERMANOVA](https://en.wikipedia.org/wiki/Permutational_analysis_of_variance) (Permutational Multivariate Analysis of Variance) uses permutation testing to generate many different rearrangements of the data and assess if the observed differences are statistically significant [@anderson2017].
+This method is important as it considers all of the volatile profiles overall when generating a p-value.
+
 ```{r gh PERMANOVA data, echo = F}
@@
 
-```{r PERMANOVA, echo = F}
+```{r PERMANOVA}
 set.seed(1)
@@
         # method = "canberra",
-        permutations = 999,
+        permutations = 9999,
         by = "terms")
-
-gh.permanova.output
+```
+```{r PERMANOVA table, echo = F}
+kable(as.data.frame(gh.permanova.output),
+      format = "html",
+      caption = "Table 1: Greenhouse Locoweed VOC collection PERMANOVA results.   gh.data.per.meta$swa is our E+/E- variable.",
+      align = "lccccc") %>%
+  kable_styling(full_width = FALSE,
+                bootstrap_options = c("striped", "bordered", "hover")) %>%
+  row_spec(0, background = "#f2f2f9") 
 ```
@@
 
+## Random Forest
+
+Next, we are going to use a random forest model to assess the importance of each of our compounds in distinguishing samples between E+ and E-. To do this we are going to use the `ranger` package in R.
+
+```{r random forest}
+ranger_model <- ranger(swa ~ ., 
+                       data = gh.joined.data.clr %>% 
+                         dplyr::select(-plant.id),
+                       importance = "impurity")
+```
+```{r random forest formatting and plot, echo = F}
+var_importance <- importance(ranger_model)
+
+var_importance_df <- 
+  data.frame(
+    variables = names(var_importance),
+    importance = var_importance) %>%
+  arrange(desc(importance)) %>%
+  rownames_to_column(var = "name") %>% dplyr::select(-name) %>%
+  mutate(variables = factor(variables, levels = unique(variables)))
+
+ggplotly(
+ggplot(var_importance_df, aes(x = variables, y = importance)) +
+  geom_bar(stat = "identity") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = NULL,
+       y = "Variable Importance",
+       title = "Random Forest Variable Importance")
+)
+```
+
 # Data Overview
 
-# APPENDIX
+# Appendix
 
@@
 kable(
-joined.data.clr %>%
-  dplyr::select(plant.id, bocimene, z3hex, nonanal, swa) %>%
+gh.joined.data.clr %>%
+  dplyr::select(plant.id, swa, year, pair, bocimene, z3hex, nonanal, cyclohexene, decanal) %>%
   # rename(pair = pair2) %>%
   sample_n(size = n(), replace = FALSE) %>%
-  mutate(bocimene = round(bocimene, digits = 2),
-         z3hex = round(z3hex, digits = 2),
-         nonanal = round(nonanal, digits = 2)) %>%
+  mutate_at(vars(bocimene:decanal), ~ round(., digits = 2)) %>%
   head(),
-align = c("l", "l", "r", "c", "c", "c", "c")
-) %>%
-  kable_styling(full_width = FALSE, position = "center")
+caption = "This is what the data looks like",  
+align = "lcccccccc") %>%
+  kable_styling(full_width = FALSE, position = "center",
+                bootstrap_options = c("striped", "bordered", "hover"))
 

### 2024-03-27 qmd/save.qmd
updated locoweed report files

--- 2024-03-09 save.qmd
@@
 
+# row of images
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+  <meta charset="UTF-8">
+  <meta name="viewport" content="width=device-width, initial-scale=1.0">
+  <title>Image Gallery</title>
+  <style>
+    .gallery-container {
+      display: flex;
+      justify-content: center;
+      width: 100%;
+    }
+    .gallery {
+      display: grid;
+      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
+      grid-gap: 10px;
+      max-width: 1200px;
+      width: 100%;
+    }
+    .gallery .gallery-img {
+      width: 100%;
+      min-wdth: 100px;
+      object-fit: cover;
+      cursor: pointer;
+      transition: transform 0.3s ease;
+    }
+    .gallery .gallery-img:hover {
+      transform: scale(1.05);
+    }
+    .expanded-container {
+      position: fixed;
+      top: 0;
+      left: 0;
+      width: 100%;
+      height: 100%;
+      background-color: rgba(0, 0, 0, 0.8);
+      display: none;
+      justify-content: center;
+      align-items: center;
+      z-index: 9999;
+    }
+    .expanded {
+      max-width: 80%;
+      max-height: 80%;
+      cursor: pointer;
+    }
+    p.gallery-caption {
+      text-align: center;
+      font-style: italic;
+      color: #8B8989;
+    }
+  </style>
+</head>
+<body>
+
+<div class="gallery">
+  <img src="locoweed/img/bombus-loco.jpg" alt="Image 1" class="gallery-img" onclick="expandImage(this)">
+  <img src="locoweed/img/loco-bug.jpg" alt="Image 1" class="gallery-img" onclick="expandImage(this)">
+  <img src="locoweed/img/field-loco1.jpg" alt="Image 1" class="gallery-img" onclick="expandImage(this)">
+  <img src="locoweed/img/ghloco1-sq.jpg" alt="Image 1" class="gallery-img" onclick="expandImage(this)">
+  <img src="locoweed/img/ghloco2-sq.jpg" alt="Image 1" class="gallery-img" onclick="expandImage(this)">
+  <img src="locoweed/img/ghloco3-sq.jpg" alt="Image 1" class="gallery-img" onclick="expandImage(this)">
+</div>
+<div class="expanded-container" onclick="closeExpanded()">
+  <img class="expanded" id="expanded-image" onclick="navigateGallery(event)">
+</div>
+
+<script>
+  var currentImageIndex = 0;
+  var galleryImages = document.querySelectorAll('.gallery-img');
+
+  function expandImage(img) {
+    currentImageIndex = Array.from(galleryImages).indexOf(img);
+    document.getElementById('expanded-image').src = img.src;
+    document.querySelector('.expanded-container').style.display = 'flex';
+  }
+
+  function closeExpanded() {
+    document.querySelector('.expanded-container').style.display = 'none';
+  }
+
+  function navigateGallery(event) {
+    var clickX = event.pageX - event.target.getBoundingClientRect().left;
+    var imageWidth = event.target.offsetWidth;
+
+    if (clickX < imageWidth / 2) {
+      currentImageIndex = (currentImageIndex - 1 + galleryImages.length) % galleryImages.length;
+    } else {
+      currentImageIndex = (currentImageIndex + 1) % galleryImages.length;
+    }
+
+    document.getElementById('expanded-image').src = galleryImages[currentImageIndex].src;
+  }
+</script>
+
+</body>
+</html>
+
+```
+
 # how to see what the default colors are in R
@@
 
+```
+
+# two images cenetered on the screen
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+    <meta charset="UTF-8">
+    <meta name="viewport" content="width=device-width, initial-scale=1.0">
+    <title>Centered Images</title>
+    <style>
+        .image-container {
+            display: flex;
+            align-items: center;
+            justify-content: center;
+        }
+
+        .image-container img {
+            max-width: 150px; /* Adjust image size as needed */
+            height: auto;
+            margin: 0 150px; /* Add space between the images */
+        }
+    </style>
+    </head>
+    <body>
+
+    <div class="image-container">
+        <img src="locoweed/img/field-loco1.jpg" alt="Image 1">
+        <img src="locoweed/img/bombus-loco.jpg" alt="Image 2">
+    </div>
+
+</body>
+</html>
 ```

### 2024-03-28 index.qmd
update styling

@@
 <br>
+
+- Understanding ecological importance of smooth brome (*Bromus inermis*) in biological control of wheat stem sawfly (*Cephus cinctus*) using native braconid parasitoids. 
+  - Field work collecting smooth brome and winter wheat samples from southwest, central, and northern Montana.
+  - Controlled wheat stem safwly infestation of smooth brome using insect netting and PVC cages.
+  - Collection of volatile organic compounds (VOCs) using greenhouse volatile collection system.
+  - Gas chromatography - mass spectroscopy (GCMS) run on VOC samples, data processed and analyzed using RStudio. 
+  - Electroantennogram - Flame Ionization Detector (EAD-FID) run on parasitoids to measure response to smooth brome and winter wheat VOCs.
+
+- Locoweed
+  - Collected VOCs from greenhouse and field *Oxytropis sericea* specimens using greenhouse volatile collection system and push-pull field volatile collection system. 
+  - GCMS run on VOC samples, data processed and analyzed using RStudio.
+
+- Biocontrol of Salt Cedar (*Tamarix ramosissima*) using subtropical tamarisk beetle (*Diorhabda sublineata*) in eastern Montana.
+  - Field work setting up insect cages on large salt cedar plants in central Montana.
+  - Collected salt cedar plants using pulaski from south central Montana.
 

### 2024-03-28 qmd/helpful.qmd
added gso page

@@
 ## R Studio  
-Great R-Markdown guide made by [Jared Adam](https://www.linkedin.com/in/jared-adam-24b107241/){style="text-decoration: none;"}: 
+- Guide to using R made by [Jared Adam](https://www.linkedin.com/in/jared-adam-24b107241/){style="text-decoration: none;"}: 
 <a href="https://jared-adam-psu.github.io/R-Tutorial-for-Tooker-Lab/" target="_blank">One code to rule them all (┛ಠ_ಠ)┛彡O</a>
@@
 ## Quarto Website  
-YouTube video on how to make a R Studio Quarto Website: [How to create a web site using Quarto, RStudio, and GitHub Pages](https://youtu.be/YN75YXaLFGM?si=RyCG23IMBwFUKsO8)
+- YouTube video on how to make a R Studio Quarto Website: [How to create a web site using Quarto, RStudio, and GitHub Pages](https://youtu.be/YN75YXaLFGM?si=RyCG23IMBwFUKsO8)  
+- Gallery showing different options for formatting a Quarto website: [Quarto Gallery](https://quarto.org/docs/gallery/)
 
 ## GitHub  
-A great guide to using GitHub created by [Callum Arnold](https://www.linkedin.com/in/arnold-c/){style="text-decoration: none;"}: <a href="https://psu-git.callumarnold.com" target="_blank">Introduction to Git and GitHub</a>
+- A guide to using GitHub created by [Callum Arnold](https://www.linkedin.com/in/arnold-c/){style="text-decoration: none;"}: <a href="https://psu-git.callumarnold.com" target="_blank">Introduction to Git and GitHub</a>

### 2024-03-28 qmd/projects.qmd
added gso page

--- 2024-03-22 qmd/projects.qmd
@@
 ---
-title: "Projects"
+# title: "Projects"
 editor_options: 
   chunk_output_type: console
+toc: true
 ---
+
+# Montana State University
+<!-- wss -->
+<a href="projects/msu-masters/wss.qmd" style="text-decoration:none;">
+  <span style="font-size:2em;">Master's Project</span>
+</a>
+  <span style="fonts-size:1em;">*wheat stem sawfly and native parasitoids*</span>  
+<div class="box">  
+Understanding the ecological importance of smooth brome (*Bromus inermis*) in biological control of the wheat stem sawfly (*Cephus cinctus*) using native parasitoids. 
+<br>  
+*PI: [Dr. David Weaver](https://landresources.montana.edu/directory/faculty/1524725/david-weaver){style="text-decoration:none;"}*
+</div>
+<br>
+<!-- locoweed -->
+<a href="projects/msu-masters/locoweed.qmd" style="text-decoration:none;">
+  <span style="font-size:2em;">Locoweed Volatiles</span>
+</a>
+  <span style="fonts-size:1em;">*volatile profile differences in Oxytropis sericea*</span>  
+<div class="box">  
+Measuring the differences in volatile organic compound (VOC) profiles between *Oxytropis sericea* specimens containing the commensal endophyte *Alternaira oxytropos*, and those without.  
+<br>
+*PI: [Dr. Tracy Sterling](http://weedeco.msu.montana.edu/faculty/tracy_sterling/){style="text-decoration:none;"}*
+</div>
+<br>
+<!-- toadflax -->
+<a href="projects/msu-masters/toadflax.qmd" style="text-decoration:none;">
+  <span style="font-size:2em;">Toadflax</span>
+</a>
+  <span style="fonts-size:1em;">*VOC data analysis*</span>  
+<div class="box">  
+Analyzing the difference in VOC profiles between two toadflax species and hybrids. Project in joint with the US Forest Service. 
+<br>  
+*PI: [Dr.
+David Weaver](https://landresources.montana.edu/directory/faculty/1524725/david-weaver){style="text-decoration: none;"} & [Dr.
+Sharlene Sing (USFS)](https://www.fs.usda.gov/research/about/people/ssing){style="text-decoration:none;"}*
+</div>
+
+# Other
+
+<a href="projects/gso.qmd" style="text-decoration:none;">
+  <span style="font-size:2em;">LRES Graduate Student Organization</span>
+</a>
+  <span style="fonts-size:1em;"></span>  
+<div class="box">  
+Served as the treasurer from 2022-2023, then as co-chair from 2023-2024. Built and managed HTML emails, a website, and organized a research symposium. 
+</div>
+<br>
+
+<a href="projects/bballtool.qmd" style="text-decoration:none;">
+  <span style="font-size:2em;">NBA Roster Analysis Tool</span>
+</a>
+  <span style="fonts-size:1em;">basketball roster analysis</span>  
+<div class="box">  
+From 2017 to 2019, I built my own NBA roster analysis tool for analyzing roster construction, free agency, and per-game metrics.
+The information is out of date, but all of the functionality still works.
+</div>
 
@@
       <div class="text">
-        <h1>Montana State University Master's</h1>
-        <p>2021-2024<br> Co-advised by Dr. David Weaver and Dr. Tracy Sterling</p>
+        <h1>Montana State University</h1>
+        <p>2021-2024</p>
       </div>

### 2024-03-28 qmd/projects/gso.qmd
added gso page

---
title: LRES Graduate Student Organization
---

<!-- header -->
<img src="gso/headerz.png" alt="header" height="150">

The Land Resources and Environmental Science Graduate Student Organization (LRES GSO) is a student-run group at Montana State University, working to provide social and professional networking opportunities for grad students in our department. LRES boasts a wide range of diversity, from plant scientists, to environmental microbiologists, to hydrologists, and the GSO exists to help students take advantage of the rich knowledge base that this broad range of interests encompasses.

<a href="https://landresources.montana.edu/gso/" target="_blank">GSO Website</a>

<!-- symposium -->
<a href="projects/gso.qmd" style="text-decoration:none;">
  <span style="font-size:2em;">2024 LRES-PSPP Student Research Symposium</span>
</a>
  <span style="fonts-size:1em;"></span>  
<div class="box">  
Along with the rest of GSO leadership, I helped to organize and put on the yearly student research symposium. In 2024, I created a website that the organzation uses to communicate information about the symposium to students and faculty. 
<br>  
<a href="https://lres-research-symposium.weebly.com" target="_blank">2024 LRES-PSPP Student Research Symposium website</a>

<img src="gso/symposiumlogo.png" alt="Alt text" style="max-width: 200px; height: auto;">
</div>

<br>
<!-- emails -->
<a href="projects/gso.qmd" style="text-decoration:none;">
  <span style="font-size:2em;">GSO Emails</span>
</a>
  <span style="fonts-size:1em;"></span>  
<div class="box">  
While serving as the co-chair, I managed and created rich HTML emails to communicate with the university community. 
<br>   
You can view my work [here](gso/emails/symposium-email-3-28.pdf). The emails don't always render correctly when converting to pdf, but you get the general idea.
</div>

### 2024-04-05 qmd/projects.qmd
updated locoweed field report

--- 2024-03-22 qmd/projects.qmd
@@
   chunk_output_type: console
+toc: true
 ---
 
-```{=html}
-<!DOCTYPE html>
-<html lang="en">
-<head>
-<meta charset="UTF-8">
-<meta name="viewport" content="width=device-width, initial-scale=1.0">
-<title>Box with Rounded Edges</title>
-<style>
-  body {
-    margin: 0; /* Remove default margin */
-    padding: 0; /* Remove default padding */
-  }
+# Montana State University
 
-  .container {
-    width: calc(100% - 20px); /* Adjusted width to fit the screen and account for margins */
-    max-width: 600px; /* Limiting maximum width */
-    height: auto; /* Set container height */
-    border: 0.5px solid #ccc; /* Light grey color */
-    border-radius: 8px; /* Less rounded edges */
-    overflow: hidden; /* Hide overflow content */
-    margin-bottom: 20px; /* Add margin to create space between rows */
-    margin-right: auto; /* Push containers to the left */
-    margin-left: 10px; /* Adjust margin between the left side of the screen */
-    display: flex; /* Use flexbox for layout */
-    align-items: center; /* Center items vertically */
-    text-decoration: none; /* Remove underline */
-  }
+<!-- wss -->
 
-  .container:last-child {
-    margin-bottom: 0; /* Remove margin from the last box */
-  }
+<a href="projects/msu/wss.qmd" style="text-decoration:none;"> [Master's Project]{style="font-size:2em;"} </a> [*wheat stem sawfly and native parasitoids*]{style="fonts-size:1em;"}\
 
-  .container:hover {
-    box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */
-  }
+::: box
+Understanding the ecological importance of smooth brome (*Bromus inermis*) in biological control of the wheat stem sawfly (*Cephus cinctus*) using native parasitoids. <br>\
+*PI: [Dr. David Weaver](https://landresources.montana.edu/directory/faculty/1524725/david-weaver){style="text-decoration:none;"}*
+:::
 
-  .image-container {
-    padding: 20px;
-    box-sizing: border-box;
-  }
+<br> 
+<!-- locoweed --> 
+<a href="projects/msu/locoweed.qmd" style="text-decoration:none;"> [Locoweed Volatiles]{style="font-size:2em;"} </a> [*volatile profile differences in Oxytropis sericea*]{style="fonts-size:1em;"}\
 
-  .image-container img {
-    width: 80px; /* Adjusted width for smaller image */
-    height: auto;
-    display: block;
-    border-radius: 8px; /* Rounded corners for image */
-  }
+::: box
+Measuring the differences in volatile organic compound (VOC) profiles between *Oxytropis sericea* specimens containing the commensal endophyte *Alternaira oxytropos*, and those without.\
+<br> *PI: [Dr. Tracy Sterling](http://weedeco.msu.montana.edu/faculty/tracy_sterling/){style="text-decoration:none;"}*
+:::
 
-  .content {
-    padding: 20px;
-    box-sizing: border-box;
-    flex: 1; /* Take remaining space */
-  }
+<br> 
+<!-- toadflax --> 
+<a href="projects/msu/toadflax.qmd" style="text-decoration:none;"> [Toadflax]{style="font-size:2em;"} </a> [*VOC data analysis*]{style="fonts-size:1em;"}\
 
-  .text h1 {
-    font-size: 22px; /* Adjusted text size */
-    margin-top: 0;
-    margin-bottom: 5px; /* Adjust margin between large and small text */
-    color: black;
-  }
-  .text h1:hover {
-    color: #2068DC; /* Change text color to blue on hover */
-  }
+::: box
+Analyzing the difference in VOC profiles between two toadflax species and hybrids. Project in joint with the US Forest Service. <br>\
+*PI: [Dr.
+David Weaver](https://landresources.montana.edu/directory/faculty/1524725/david-weaver){style="text-decoration: none;"} & [Dr.
+Sharlene Sing (USFS)](https://www.fs.usda.gov/research/about/people/ssing){style="text-decoration:none;"}*
+:::
 
-  .text p {
-    font-size: 16px;
-    margin-top: 5px;
-    color: #666;
-    text-align: justify; /* Align text */
-  }
+# Other
 
-  @media (max-width: 768px) {
-    .container {
-      width: calc(100% - 20px); /* Adjusted width to fit the screen */
-    }
-  }
-</style>
-</head>
-<body>
-  <a href="projects/msu-masters.qmd" class="container">
-    <div class="image-container">
-      <img src="../img/projects-logos/msu-logo.jpg" alt="Image">
-    </div>
-    <div class="content">
-      <div class="text">
-        <h1>Montana State University Master's</h1>
-        <p>2021-2024<br> Co-advised by Dr. David Weaver and Dr. Tracy Sterling</p>
-      </div>
-    </div>
-  </a>
-  <a href="projects/bballtool.qmd" class="container">
-    <div class="image-container">
-      <img src="../img/projects-logos/bballref-logo-circle.jpg" alt="Image">
-    </div>
-    <div class="content">
-      <div class="text">
-        <h1>NBA Roster Analysis Tool</h1>
-        <p>2017-2019</p>
-      </div>
-    </div>
-  </a>
-</body>
-</html>
+<a href="projects/gso.qmd" style="text-decoration:none;"> [LRES Graduate Student Organization]{style="font-size:2em;"} </a> \
 
-```
-```{=html}
-<head>
-  <meta charset="UTF-8">
-  <meta name="viewport" content="width=device-width, initial-scale=1.0">
-  <title>Box with Rounded Edges</title>
-  <style>
-    body {
-      margin: 0; /* Remove default margin */
-      padding: 0; /* Remove default padding */
-    }
+::: box
+Served as the treasurer from 2022-2023, then as co-chair from 2023-2024. Built and managed HTML emails, a website, and organized a research symposium.
+:::
 
-    .container {
-      width: 100%; /* Adjusted width to fit the screen */
-      max-width: 600px; /* Limiting maximum width */
-      height: auto; /* Set container height */
-      border: 0.5px solid #ccc; /* Light grey color */
-      border-radius: 8px; /* Less rounded edges */
-      overflow: hidden; /* Hide overflow content */
-      margin-bottom: 20px; /* Add margin to create space between rows */
-      display: flex; /* Use flexbox for layout */
-      align-items: center; /* Center items vertically */
-      text-decoration: none; /* Remove underline */
-    }
+<br>
 
-    .container:last-child {
-      margin-bottom: 0; /* Remove margin from the last box */
-    }
+<a href="projects/bballtool.qmd" style="text-decoration:none;"> [NBA Roster Analysis Tool]{style="font-size:2em;"} </a> [basketball roster analysis]{style="fonts-size:1em;"}\
 
-    .container:hover {
-      box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */
-    }
+::: box
+From 2017 to 2019, I built my own NBA roster analysis tool for analyzing roster construction, free agency, and per-game metrics.
+The information is out of date, but all of the functionality still works.
+:::
 
-    .image-container {
-      padding: 20px;
-      box-sizing: border-box;
-    }
+<!-- ```{=html} -->
+<!-- <!DOCTYPE html> -->
+<!-- <html lang="en"> -->
+<!-- <head> -->
+<!-- <meta charset="UTF-8"> -->
+<!-- <meta name="viewport" content="width=device-width, initial-scale=1.0"> -->
+<!-- <title>Box with Rounded Edges</title> -->
+<!-- <style> -->
+<!--   body { -->
+<!--     margin: 0; /* Remove default margin */ -->
+<!--     padding: 0; /* Remove default padding */ -->
+<!--   } -->
 
-    .image-container img {
-      max-width: 100%; /* Ensures the image does not exceed its container */
-      height: auto;
-      display: block;
-      border-radius: 8px; /* Rounded corners for image */
-    }
+<!--   .container { -->
+<!--     width: calc(100% - 20px); /* Adjusted width to fit the screen and account for margins */ -->
+<!--     max-width: 600px; /* Limiting maximum width */ -->
+<!--     height: auto; /* Set container height */ -->
+<!--     border: 0.5px solid #ccc; /* Light grey color */ -->
+<!--     border-radius: 8px; /* Less rounded edges */ -->
+<!--     overflow: hidden; /* Hide overflow content */ -->
+<!--     margin-bottom: 20px; /* Add margin to create space between rows */ -->
+<!--     margin-right: auto; /* Push containers to the left */ -->
+<!--     margin-left: 10px; /* Adjust margin between the left side of the screen */ -->
+<!--     display: flex; /* Use flexbox for layout */ -->
+<!--     align-items: center; /* Center items vertically */ -->
+<!--     text-decoration: none; /* Remove underline */ -->
+<!--   } -->
 
-    .content {
-      padding: 20px;
-      box-sizing: border-box;
-      flex: 1; /* Take remaining space */
-    }
+<!--   .container:last-child { -->
+<!--     margin-bottom: 0; /* Remove margin from the last box */ -->
+<!--   } -->
 
-    .text h1 {
-      font-size: 22px; /* Adjusted text size */
-      margin-top: 0;
-      margin-bottom: 5px; /* Adjust margin between large and small text */
-      color: black;
-    }
+<!--   .container:hover { -->
+<!--     box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */ -->
+<!--   } -->
 
-    .text h1:hover {
-      color: #2068DC; /* Change text color to blue on hover */
-    }
+<!--   .image-container { -->
+<!--     padding: 20px; -->
+<!--     box-sizing: border-box; -->
+<!--   } -->
 
-    .text p {
-      font-size: 16px;
-      margin-top: 5px;
-      color: #666;
-      text-align: justify; /* Align text */
-    }
+<!--   .image-container img { -->
+<!--     width: 80px; /* Adjusted width for smaller image */ -->
+<!--     height: auto; -->
+<!--     display: block; -->
+<!--     border-radius: 8px; /* Rounded corners for image */ -->
+<!--   } -->
 
-    @media (max-width: 768px) {
-      .container {
-        margin-right: auto; /* Reset margin-right */
-        margin-left: auto; /* Reset margin-left */
-        flex-direction: column; /* Stack items vertically on smaller screens */
-      }
+<!--   .content { -->
+<!--     padding: 20px; -->
+<!--     box-sizing: border-box; -->
+<!--     flex: 1; /* Take remaining space */ -->
+<!--   } -->
 
-      .content {
-        padding: 10px; /* Reduce padding on smaller screens */
-      }
-    }
-  </style>
-</head>
-```
+<!--   .text h1 { -->
+<!--     font-size: 22px; /* Adjusted text size */ -->
+<!--     margin-top: 0; -->
+<!--     margin-bottom: 5px; /* Adjust margin between large and small text */ -->
+<!--     color: black; -->
+<!--   } -->
+<!--   .text h1:hover { -->
+<!--     color: #2068DC; /* Change text color to blue on hover */ -->
+<!--   } -->
 
-```{r, echo = F, message = F, results = 'hide'}
-library(pandoc)
-library(knitr)
-library(rmarkdown)
-getwd()
-rmarkdown::render(input = "litreview.qmd", 
-                  run_pandoc = TRUE,
-                  output_format = "word_document")
+<!--   .text p { -->
+<!--     font-size: 16px; -->
+<!--     margin-top: 5px; -->
+<!--     color: #666; -->
+<!--     text-align: justify; /* Align text */ -->
+<!--   } -->
 
-```
+<!--   @media (max-width: 768px) { -->
+<!--     .container { -->
+<!--       width: calc(100% - 20px); /* Adjusted width to fit the screen */ -->
+<!--     } -->
+<!--   } -->
+<!-- </style> -->
+<!-- </head> -->
+<!-- <body> -->
+<!--   <a href="projects/msu.qmd" class="container"> -->
+<!--     <div class="image-container"> -->
+<!--       <img src="../img/projects-logos/msu-logo.jpg" alt="Image"> -->
+<!--     </div> -->
+<!--     <div class="content"> -->
+<!--       <div class="text"> -->
+<!--         <h1>Montana State University</h1> -->
+<!--         <p>2021-2024</p> -->
+<!--       </div> -->
+<!--     </div> -->
+<!--   </a> -->
+<!--   <a href="projects/bballtool.qmd" class="container"> -->
+<!--     <div class="image-container"> -->
+<!--       <img src="../img/projects-logos/bballref-logo-circle.jpg" alt="Image"> -->
+<!--     </div> -->
+<!--     <div class="content"> -->
+<!--       <div class="text"> -->
+<!--         <h1>NBA Roster Analysis Tool</h1> -->
+<!--         <p>2017-2019</p> -->
+<!--       </div> -->
+<!--     </div> -->
+<!--   </a> -->
+<!-- </body> -->
+<!-- </html> -->
+<!-- ``` -->
+<!-- ```{=html} -->
+<!-- <head> -->
+<!--   <meta charset="UTF-8"> -->
+<!--   <meta name="viewport" content="width=device-width, initial-scale=1.0"> -->
+<!--   <title>Box with Rounded Edges</title> -->
+<!--   <style> -->
+<!--     body { -->
+<!--       margin: 0; /* Remove default margin */ -->
+<!--       padding: 0; /* Remove default padding */ -->
+<!--     } -->
+
+<!--     .container { -->
+<!--       width: 100%; /* Adjusted width to fit the screen */ -->
+<!--       max-width: 600px; /* Limiting maximum width */ -->
+<!--       height: auto; /* Set container height */ -->
+<!--       border: 0.5px solid #ccc; /* Light grey color */ -->
+<!--       border-radius: 8px; /* Less rounded edges */ -->
+<!--       overflow: hidden; /* Hide overflow content */ -->
+<!--       margin-bottom: 20px; /* Add margin to create space between rows */ -->
+<!--       display: flex; /* Use flexbox for layout */ -->
+<!--       align-items: center; /* Center items vertically */ -->
+<!--       text-decoration: none; /* Remove underline */ -->
+<!--     } -->
+
+<!--     .container:last-child { -->
+<!--       margin-bottom: 0; /* Remove margin from the last box */ -->
+<!--     } -->
+
+<!--     .container:hover { -->
+<!--       box-shadow: 0 10px 20px rgba(0, 0, 0, 0.1); /* Increased shadow on hover */ -->
+<!--     } -->
+
+<!--     .image-container { -->
+<!--       padding: 20px; -->
+<!--       box-sizing: border-box; -->
+<!--     } -->
+
+<!--     .image-container img { -->
+<!--       max-width: 100%; /* Ensures the image does not exceed its container */ -->
+<!--       height: auto; -->
+<!--       display: block; -->
+<!--       border-radius: 8px; /* Rounded corners for image */ -->
+<!--     } -->
+
+<!--     .content { -->
+<!--       padding: 20px; -->
+<!--       box-sizing: border-box; -->
+<!--       flex: 1; /* Take remaining space */ -->
+<!--     } -->
+
+<!--     .text h1 { -->
+<!--       font-size: 22px; /* Adjusted text size */ -->
+<!--       margin-top: 0; -->
+<!--       margin-bottom: 5px; /* Adjust margin between large and small text */ -->
+<!--       color: black; -->
+<!--     } -->
+
+<!--     .text h1:hover { -->
+<!--       color: #2068DC; /* Change text color to blue on hover */ -->
+<!--     } -->
+
+<!--     .text p { -->
+<!--       font-size: 16px; -->
+<!--       margin-top: 5px; -->
+<!--       color: #666; -->
+<!--       text-align: justify; /* Align text */ -->
+<!--     } -->
+
+<!--     @media (max-width: 768px) { -->
+<!--       .container { -->
+<!--         margin-right: auto; /* Reset margin-right */ -->
+<!--         margin-left: auto; /* Reset margin-left */ -->
+<!--         flex-direction: column; /* Stack items vertically on smaller screens */ -->
+<!--       } -->
+
+<!--       .content { -->
+<!--         padding: 10px; /* Reduce padding on smaller screens */ -->
+<!--       } -->
+<!--     } -->
+<!--   </style> -->
+<!-- </head> -->
+<!-- ``` -->
+<!-- ```{r, echo = F, message = F, results = 'hide'} -->
+<!-- library(pandoc) -->
+<!-- library(knitr) -->
+<!-- library(rmarkdown) -->
+<!-- getwd() -->
+<!-- rmarkdown::render(input = "litreview.qmd",  -->
+<!--                   run_pandoc = TRUE, -->
+<!--                   output_format = "word_document") -->
+
+<!-- ``` -->

### 2024-04-05 qmd/projects/msu.qmd
updated locoweed field report

--- 2024-03-27 qmd/projects/msu-masters.qmd
@@
 title: "MSU Masters"
-
 ---
@@
 </html>
-
 ```

### 2024-04-05 qmd/projects/msu/locoweed/field-locoweed-report.qmd
updated locoweed field report

--- 2024-03-27 qmd/projects/msu-masters/locoweed/field-locoweed-report.qmd
@@
 <!-- bee cover image -->
+
 ```{=html}
@@
 </html>
-
 ```
-
 *Note: This project is still a work in progress.*
@@
 # Introduction
+
 Astragalus and Oxytropix species, part of the Legume Family, are found in terestrial ecoysstems globally, including the western regions of North America.
@@
 # Methodology
+
 ## Field Volatile Collection
@@
 Volatiles were collected from all the plants in each year over the course of one to two successive weeks.
-
 Volatile were collected from 20 plants per day by placing Reynolds Turkey Oven Bags (Reynolds Group Holdings, Auckland, New Zealand) over the top of the selected number of leaflets.
+A push-pull system was used to pull air from inside each turkey back.
+Glass filter traps (6.35 mm diameter x 76 mm length; Analytical Research Systems, Gainesville, FL) with 30 mg of Super-Q absorbent (Alltech Associates, Deerfield, IL) were inserted into one end of a lenght of plastic tubing that was connected to an air pump.
 
@@
 ## Data Analysis
+
 The data analysis procedure for each collection type was the same.
@@
 # Results
+
 <!-- LOAD DATA -->
+
 ```{r 2023 field data load, echo = F, message = F}
@@
 ```
+
 ```{r 2022 field data load, echo = F, message = F}
@@
 ```
+
 <!-- CLEAN UP DATA -->
+
 ```{r 2023 field data cleanup, echo = F, message = F}
@@
 ```
+
 ```{r 2022 field data cleanup, echo = F, message = F}
@@
 ```
+
 <!-- COMPOUNDS TO BE USED IN ANALYSIS -->
+
 ```{r COMPOUNDS, echo = F, message = F}
@@
 ```
+
 <!-- CLR TRANSFORMATION -->
+
 ```{R CLR, echo = F, message = F}
@@
 ## Visual Comparisons
+
 ### Box Plot
+
 Eleven compounds were selected to display comparisons.
+
 ```{r all.comparison.plot, echo = F, fig.align = 'center', fig.width = 10}
@@
 ```
+
 ```{r save all comparison plot, echo = F}
@@
 ```
+
 ```{=html}
@@
 ```
-:::{.callout-note}
+::: callout-note
 I chose to visualize the data using a box plot, not a bar plot, because our data has been CLR transformed.
@@
 ### PCA
+
 Principal Component Analysis (PCA) is a dimensionality reduction technique used to transform high-dimensional data into a lower-dimensional representation while preserving most of the original variability.
@@
 ## Check Assumptions
+
 The statistical analyses we can perform are greatly impacted by our data's ability to meet statistical assumptions.
@@
 We are going to test this assumption using the `MVN` package.
+
 ```{r assumptions, echo = F}
@@
 
-<div class="box">
-**Assumption Conclusions**  
+::: box
+**Assumption Conclusions**\
 From the Chi-Squared QQ-Plot (figure 2), we can see that our data set violates the normality assumption. The tail on the right half of the plot dramatically skews away from the normality line.
@@
 As we can see, our data violates the normality assumption. This is the case even though we ran it though a CLR transformation.
-</div>
-:::{.callout-note}
-**A quick note here.**  
+:::
+
+::: callout-note
+**A quick note here.**\
 Many papers analyzing ecological or compositional data and VOC datasets often employ a statistical test known as [MANOVA](https://en.wikipedia.org/wiki/Multivariate_analysis_of_variance) (Multivariate Analysis of Variance).
@@
 ## Kruskal-Wallis Test
+
 The [Kruskal-Wallis](https://en.wikipedia.org/wiki/Kruskal–Wallis_test) test is a non-parametric statistical test used to determine if there are differences between two or more independent groups.
@@
 ```
+
 ```{r 22/23 FIELD PERMANOVA, results = F}
@@
 ```
+
 ```{r PERMANOVA table, echo = F}
@@
 ```
+
 <!-- Permanova P-values -->
+
 ```{r PERMANOVA p-values, echo  = F}

### 2024-04-05 qmd/projects/msu/toadflax/toadflax_report.qmd
updated locoweed field report

--- 2024-03-08 toadflax_report.qmd
@@
 **Type:** This section has data organized by the four grouping variables - YT, DT, Natural, and Synthetic.
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<div class="button-container">
+  <a href="larger_boxplots.qmd">
+    <button class="button">View Larger Boxplots</button>
+  </a>
+
+</div>
+</body>
+</html>
+
+```
 
@@
 ```{r boxplot comparison category (original data), echo = F, fig.height = 20, fig.width = 20}
+boxplot.data.type.plot <-
   boxplot.data.type %>%
@@
         legend.position = 'none',
-        # legend.position = c(0.8,0.9),
         panel.background = element_rect(fill = "white"),
         axis.line = element_line(color = "black"),
-        panel.grid.major.x = element_line(color = "#F2F2F2"),
-        axis.text = element_text(size = 22),
-        axis.title = element_text(size = 22),
-        strip.text = element_text(size = 22)) +
+        panel.grid.major.x = element_line(color = "#F2F2F2")) +
   facet_wrap(~compound, 
@@
              ncol = 3)
+
+boxplot.data.type.plot +
+  theme(axis.text = element_text(size = 22),
+        axis.title = element_text(size = 22),
+        strip.text = element_text(size = 22))
+
+```
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+<meta charset="UTF-8">
+<meta name="viewport" content="width=device-width, initial-scale=1.0">
+<title>Download File Buttons</title>
+<style>
+/* Button styling */
+.button {
+  background-color: #2068DC; /* Header of website color (blue) */
+  border: none;
+  color: white;
+  padding: 15px 32px;
+  text-align: center;
+  text-decoration: none;
+  display: inline-block;
+  font-size: 16px;
+  margin: 4px 2px;
+  cursor: pointer;
+  border-radius: 8px;
+  margin: 0 200px;
+}
+.button:hover {
+  background-color: #3F91FB; /* Sky blue */
+}
+.button-container {
+  display: flex;
+  justify-content: center;
+  margin-bottom: 10px;
+}
+</style>
+</head>
+<body>
+
+<div class="button-container">
+  <a href="larger_boxplots.qmd">
+    <button class="button">View Larger Boxplots</button>
+  </a>
+
+</div>
+</body>
+</html>
 
@@
        y = "VOC Concentration (ng/g/hr)") +
+  facet_wrap(~compound, 
+             scales = "free_y",
+             ncol = 3) +
   theme(axis.text.x = element_text(angle = 45, 
@@
         legend.text = element_text(size = 22),
-        legend.title = element_text(size = 22)) +
-  facet_wrap(~compound, 
-             scales = "free_y",
-             ncol = 3)
+        legend.title = element_text(size = 22))
 

### 2024-04-05 qmd/projects/msu/wss/field_report.qmd
updated locoweed field report

--- 2024-03-06 field_report.qmd
@@
 <div class="image-container">
-    <img src="img/field-sites/big-sandy-wide.jpg" alt="Your Image">
+    <img src="../../../../img/field-sites/big-sandy-wide.jpg" alt="Your Image">
 </div>
@@
 raw.2022 <- 
-  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "n.infested") %>%
+  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "n.infested") %>%
   mutate(year = "2022") %>%
@@
 raw.2023 <-
-read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "n.infested") %>%
+read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "n.infested") %>%
   mutate(year = "2023") 
@@
 stubble.2023 <-
-  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2023/2023\ Field Site\ Data.xlsx", 
+  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", 
            sheet = "stubble") %>%
@@
 stubble.2022 <-
-  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Field Sites/2022/2022 FS Stem Data (2).xlsx", 
+  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", 
              sheet = "stubble") %>%
@@
 <div class="container">
-  <img class="image" src="img/field-sites/nnns.png" alt="Image" onclick="expandImage(this)">
+  <img class="image" src="../../../../img/field-sites/nnns.png" alt="Image" onclick="expandImage(this)">
   <div class="overlay" onclick="closeExpandedImage()"></div> <!-- Overlay element -->
@@
 <div class="container">
-  <img class="image" src="img/field-sites/gardiner.png" alt="Image" onclick="expandImage(this)">
+  <img class="image" src="../../../../img/field-sites/gardiner.png" alt="Image" onclick="expandImage(this)">
   <div class="overlay" onclick="closeExpandedImage()"></div> <!-- Overlay element -->
@@
 <div class="container">
-  <img class="image" src="img/field-sites/harvest.png" alt="Image" onclick="expandImage(this)">
+  <img class="image" src="../../../../img/field-sites/harvest.png" alt="Image" onclick="expandImage(this)">
   <div class="overlay" onclick="closeExpandedImage()"></div> <!-- Overlay element -->
@@
 <div class="container">
-  <img class="image" src="img/field-sites/kolin.png" alt="Image" onclick="expandImage(this)">
+  <img class="image" src="../../../../img/field-sites/kolin.png" alt="Image" onclick="expandImage(this)">
   <div class="overlay" onclick="closeExpandedImage()"></div> <!-- Overlay element -->

### 2024-04-05 qmd/projects/msu/wss/ndvi.qmd
updated locoweed field report

---
title: "ndvi"
editor_options: 
  chunk_output_type: console
---

```{r library, echo = F, message = F}
library(readxl)
library(dplyr)
library(ggplot2)
library(plotly)
```

```{r read data, echo = F}
setwd("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/GitHub/Quarto Website/Quarto Website/qmd/projects/msu/wss")

data <-
read_excel("NDVI_values_filteredx.xlsx") %>%
  dplyr::select(NDVI,date,type)
```

```{r plot, echo = F}
ggplotly(data %>%
  filter(!date %in% c("2023-03-31",
                      "2023-04-20",
                      "2023-05-10",
                      "2023-05-30",
                      "2023-06-09",
                      "2023-06-19",
                      "2023-07-14",
                      "2023-08-13")) %>%
  # mutate(NDVI = ifelse(type == "wheat", -NDVI, NDVI)) %>%
  ggplot(aes(x = date, y = NDVI, fill = type)) +
  geom_bar(stat = 'identity', 
           position = position_dodge()) +
  geom_line(stat = "smooth") +
  theme_bw()
)
```

### 2024-05-13 qmd/data analysis.qmd
commit 5/13 2024

--- 2024-03-22 qmd/litreview.qmd
@@
 ---
-title: "Literature Review (Thesis)"
-author: "Jackson Strand"
-date: March 20, 2024
-output: 
-  html_document:
-    toc: true
-    code_folding: hide
-bibliography: proposal.bib
-link-citations: true
-biblio-style: apalike
-cls: elsevier-with-titles.csl
-# cls: council-of-science-editors-author-date.csl
-editor_options: 
-  chunk_output_type: inline
-toc: true
+title: "Data Analysis"
+date: "April 5, 2024"
 ---
 
-<!-- Button to download word document version -->
-```{=html}
-<!DOCTYPE html>
-<html lang="en">
-<head>
-<meta charset="UTF-8">
-<meta name="viewport" content="width=device-width, initial-scale=1.0">
-<title>Download File Button</title>
-<style>
-/* Button styling */
-.button {
-  background-color: #2068DC; /* Header of website color (blue) */
-  border: none;
-  color: white;
-  padding: 15px 32px;
-  text-align: center;
-  text-decoration: none;
-  display: inline-block;
-  font-size: 16px;
-  margin: 4px 2px;
-  cursor: pointer;
-  border-radius: 8px;
-}
-.button:hover {
-  background-color: #3F91FB; /* Sky blue */
-}
-</style>
-</head>
-<body>
+# Data Analysis
 
-<!-- Button to download the file -->
-<a href="litreview.docx" download="Jackson Strand Proposal">
-  <button class="button">Download as Word Document</button>
-</a>
+## Field Experiments
 
-<script>
-// Get the current date in YYYY-MM-DD format
-function getCurrentDate() {
-  const now = new Date();
-  const year = now.getFullYear();
-  const month = String(now.getMonth() + 1).padStart(2, '0');
-  const day = String(now.getDate()).padStart(2, '0');
-  return `${year}-${month}-${day}`;
-}
+### Field Survey
 
-// Update the download attribute of the link with the current date
-document.addEventListener('DOMContentLoaded', function() {
-  const downloadLink = document.getElementById('downloadLink');
-  const currentDate = getCurrentDate();
-  downloadLink.setAttribute('download', `Jackson Strand Proposal ${currentDate}`);
-});
-</script>
-
-</body>
-</html>
-```
-
-# Project Overview
-## Motivation
-The motivation for my research is to improve yield for wheat growers by reducing wheat stem sawfly (WSS) populations.
-If we can show that healthier, more robust smooth brome leads to increased parasitoid populations and thus, increased WSS parasitism and potentially decreased damage by WSS, we can better educate growers on how to increase yields.
-In terms of scientific merit, understanding these mechanisms is key in ongoing WSS research as it explains several how and why questions associated with the ecological relationship between smooth brome and WSS parasitism.
-
-## Research Question
-What role does smooth brome (*Bromus inermis*) play in the ecological relationship between wheat stem sawfly (*Cephus cinctus*) and wheat stem sawfly parasitoids (*Bracon cephi* and *Bracon lissogaster*) in dryland wheat production landscape within Montana?
-
-## Hypothesis
-Smooth brome acts as an ecological sink for WSS, while serving as a parasitoid source because WSS larvae experience high mortality within smooth brome stems, while any survivors experience high rates of parasitism as a second source of mortality.
-In addition, early season parasitoids are attracted to infested smooth brome because it produces volatiles similar to those produced by winter wheat infested with larval WSS.
-
-## Prediction
-WSS larvae will experience both high levels of mortality and similar levels of parasitism within smooth brome stems when compared to the survivability and parasitism of WSS larvae in the adjacent fields.
-Also, the blend of volatile compounds produced by WSS infested smooth brome will qualitatively match in composition and exceed the amounts produced by infested winter wheat.
-
-## Research Approach
-My research is composed of three separate studies.
-First, there is a field survey to assess the survivability of WSS larvae and associated parasitoid frequency within both winter wheat and adjacent smooth brome.
-Second, there are controlled field infestations which will assess the survivability of WSS larvae within smooth brome stems when controlled for other density driven factors.
-And third, there is a greenhouse study to measure the volatile compounds produced by winter wheat and smooth brome.
-Using the results of the volatile collection, we can assess the parasitoid response both chemically with electroantennography and behaviorally with a Y-tube olfactory bioassay.
-
-## Importance of Topic
-The wheat stem sawfly (WSS, Cephus cinctus Norton, Hymenoptera: Cephidae), is a serious problem for wheat producers throughout the Northern Great Plains of North America [@Fulbright2017].
-In Montana alone, the economic damage caused by the pest was reported to be upwards of $80 million annually between 2011-2016 [@Fulbright2017].
-Current techniques in combatting the damage caused by the WSS use a variety of chemical, cultural, and biocontrol strategies with limited short-term effectiveness and long-term sustainability.
-Utilizing biological control via promoting braconid parasitoid habitat holds considerable value to both growers and agricultural observers.
-As we seek to increase agricultural yields, we must also consider the environmental impact that tactics such as pesticides and tillage have on the future success on wheat farming.
-
-# Literature Review
-## History and Distribution of Wheat Stem Sawfly
-The wheat stem sawfly was first described in 1872 by Edward Norton based upon a specimen found in Colorado [@Norton1872].
-In 1890 it was found in California and Montana [@Ainslie1920].
-Native to North America [@Lesieur2016], the largest populations of WSS are currently found in the western Canadian prairies as well as Wyoming, Montana, Colorado, and North and South Dakota [@Wallace1966; @Weiss1992; @Cockrell2021].
-
-## Host Plants of Wheat Stem Sawfly
-Wheat stem sawfly was first found the hollow-stemmed wild grasses such as Agropyron and Elymus in Colorado in 1872 [@Ainslie1920].
-Soon thereafter it was found in wheat in 1895 in Manitoba, Canada.
-First referred to as the “western-grass stem sawfly”, WSS was found with native Montana grasses including western wheatgrass (Pascopyrun smithii) and slender wheat grass (Elymus trachycaulus) [@Ainslie1920].
-As settlers began moving west and small grain cereal crop production increased on the landscape, the native grasses that WSS foraged upon became quite rare [@Ainslie1929].
-WSS was found damaging spring wheat for the first time in 1895 in Manitoba and Saskatchewan [@Wallace1966].
-With the diminishment of its natural hosts, WSS has adapted to infest the exotic non-cultivated and cultivated grasses now present [@Criddle1923].
-WSS infestation has been shown in many species native to Montana including western wheatgrass (Pascopyrun smithii), slender wheat grass (Elymus trachycaulus) and bluebunch wheatgrass, Pseudoroegneria spicata [@Cockrell2021].
-
-Smooth brome (*Bromus inermis* L.) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 from Hungary and Russia for soil conservation and use as livestock feed [@Salesman2011].
-Throughout the early 19th and 20th century, it was used throughout much of the upper half of North America.
-It is a perennial grass, and more drought resistant than other exotic grasses [@Mapfumo2002].
-Due to its sod forming nature and large root system, it is extensively used to promote soil retention along highways (USDA-NRCS 2010).
-
-Smooth brome was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Farstad1945].
-However, due to its propensity to form monocultures and outcompete many native grasses, smooth brome was labeled by many as invasive [@Carlson1985; @Willson2000; @Dillemuth2008].
-Smooth brome populations play a considerable role in altering native herbivore population dynamics and movement on the prairies of North America [@Dillemuth2008].
-
-## Biology of Wheat Stem Sawfly
-### Adult
-Adults are relatively small (8-13 mm) with a narrow elongated and slightly compressed black body, yellow legs, and three characteristic black bands across the abdomen.
-Females are larger than males and a visible long, dark, ovipositor.
-Both male and female adults are weak fliers, rarely flying long distances at one time [@Criddle1923].
-They can be found on stems with their heads downward and legs aligned with their body, encircling the stem [@Ainslie1920; @Criddle1923].
-WSS are univoltine, having only one generation per year.
-Males are haploid (9 chromosomes), while females are diploid – females who do not find a mate will produce male offspring, while those who do produce female progeny [@Holmes1978].
-Fully mature adults hatch in the spring once it has sufficiently warmed and the moisture levels in the soil reach threshold levels [@Holmes1978; @mendoza2006].
-In Montana, this period begins in mid- to late-May and ends early in July [@Morrill1996; @Weaver2005].
-Their lifespan usually lasts around five to eight days, emerging from May until June.
-Males emerge before the females, ensuring that many of the early females to hatch will find a mate and deposit eggs that will be female [@Mackay2011].
-Copulation lasts less than one minute with the male holding the back of the female with their abdomen curved down under the female ovipositor.
-Fertilized females lay eggs several hours after copulation [@Holmes1978].
-
-### Oviposition
-After locating a suitable host stem, female WSS oviposit an egg inside the host plant lumen.
-WSS larvae hatch inside the stem and feed on the stem’s parenchyma tissue [@Beres2011].
-As the stem begins to senesce in the fall, the larvae travel to the base of the stem and makes a V-shaped groove [@Criddle1923] above a constructed protective hibernaculum where they will overwinter [@Holmes1960].
-As a result of this deep groove, the stem weakens and topples easily.
-The resulting lodged stems are not only challenging for growers to harvest but also exhibit decreased kernel yield [@Beres2011].
-
-### Egg
-Newly laid eggs are crescent in shape, milky-white in color, and range between 1 mm to 1.25 mm in length and 0.33 to 0.42 mm wide [@Ainslie1920].
-After oviposition, the egg lies freely within the lumen of the host stem or in a hollow created by the female WSS ovipositor [@Fulbright2017].
-Laval development occurs rapidly with the recognizable shape becoming apparent at day three.
-By day five, the jaws turn brown as the eye spots begin to appear and darken.
-By day six or seven, the larva begins to hatch using a series of convulsive movements and emerges into the stem cavity, with full functionality achieved by the seventh day [@Ainslie1920].
-
-### Larvae
-After hatching and prior to feeding, the larvae are transparent and colorless [@Ainslie1920; @Criddle1923].
-WSS larvae are distinguishable from other larvae feeding in wheat and grasses by their highly sclerotized head capsule and large body size [@Nelson1953].
-As the larvae chew their way through the parenchymous and vascular tissues of the host stem, the larvae begin to develop a yellow-green color.
-As the larvae progress through the stem, they excrete frass; the digested plant tissue as well as metabolic breakdown products from the WSS [@Beres2011].
-If multiple eggs are laid within the same stem, the largest (usually the first egg laid and hatched) will typically remain within a few weeks due to the larvae cannibalizing one another [@Criddle1923; @Seamans1944; @Buteler2015].
-It remains unknown whether the consumption of other eggs and larvae is the result of intentional cannibalism or the indirect result of indiscriminate feeding [@Buteler2015].
-As the host plant physiologically matures, and the larvae reaches its fifth instar, the insect is cued to travel to the base of stem and prepare for obligatory diapause [@Beres2011].
-The descending movement of larvae and development of the hibernaculum are triggered by the host plant senescence.
-As the stem walls becoming increasingly transparent, they allow increased amounts of visible and infrared light into the stem [@Holmes1977; @Beres2011].
-Once the larvae reach the base of the stem, they construct an interior V-shaped groove, making the stem weak enough to lodge due to gravity or wind [@Criddle1923].
-The larvae then plug the remaining stub with frass to insulate the hibernaculum during diapause [@Holmes1960].
-The larval stage of development can be as short as 30 days before entering diapause.
-Obligatory diapause lasts a minimum of 90 days [@Salt1947].
-After the larvae are exposed to temperatures less than 10°C for 90 days, diapause is complete, and pupation can begin.
-Larvae can reenter diapause if temperatures approach 35°C [@Salt1947] or if conditions are abnormally dry [@Holmes1978] shortly after termination of diapause.
-
-### Pupae
-Once diapause is complete, pupal development begins.
-Pupation lasts for 21 days, beginning in early- to mid-May [@Criddle1923].
-Initially, the pupa is milky-white in color [@Ainslie1920], gradually changing to bright white as the wings develop [@Fulbright2017].
-Additionally, as development continues, characteristic brown and black eye spots appear on the surface of the pupae.
-Once pupation is complete, the adult sawfly will chew through either the frass plug or the side of the stub, and emerge ready for flight [@Ainslie1920; @Holmes1960].
-
-## Damage Caused by Wheat Stem Sawfly
-Wheat stem sawfly larvae travel throughout the stem of the host plant, feeding on the parenchymal tissue and chewing through the nodes.
-As a result, vascular translocation of water and nutrients within the plant is impaired [@Macedo2007] and causes a notable reduction yield and quality of the grain produced [@Seamans1944; @Holmes1977; @Morrill1992].
-In addition, the weakening of the stem due to WSS causes lodging prior to harvest, exacerbated by larvae stem girdling.
-Stem lodging causes a reduction in stem harvest efficiently [@Ainslie1920], with yield lost due to unrecoverable lodged stem [@Beres2007].
-Additionally, the reduction in stubble residues decreases snow retention which increases soil erosion and reduces soil moisture content [@Keren2015].
-
-## Management of Wheat Stem Sawfly
-### Chemical Control
-Because WSS are stem mining pests and spend most of their lives within the host plant stem, the use of traditional pest control methods such as insecticides have proven to be ineffective.
-The only life stage of WSS that takes place outside of the stem, the adult, is not targetable via chemical treatment due to adult emergence and oviposition taking place over many weeks [@Knodel2010; @Beres2011].
-WSS infestation chemical control methods were first tested as early as 1949 [@Wallace1962].
-After using six different organochlorine and organophosphate insecticidal dust applied via a powder duster [@Munro1949] the authors found “unsatisfactory results that did not control” WSS populations or limit infestations.
-In 1963, it was reported that a 73% control of WSS infestation could be achieved via granular heptachlor applied in the furrow with the wheat seed [@Wallace1962].
-However, the findings of these tests were not able to be replicable in all scenarios of WSS infestation [@Holmes1963].
-Heptachlor required light sawfly infestation – restricted to the lower internodes – to be effective.
-When infestation rates were high, and oviposition occurred in the upper internodes, the researchers saw little to no control.
-Similar results were found when testing with foliar insecticides, findings showing little differences in their effect when compared to controls [@Blodgett1996].
-While the Montana Department of Agriculture has authorized the use of systemic organophosphate insecticide phorate (Thimet®) for use controlling WSS infestation in winter and spring wheat cultivars, it’s use is not considered environmentally friendly due to high toxicity to mammals, birds, and fish.
-Additionally, a preharvest interval of 85 days complicates harvest scheduling (Ginsberg, 2003; Mahajan et al., 2006), while the high toxicity may have potential adverse effects on beneficial parasitoid populations [@Varella2016].
-
-### Cultural Control
-#### Host Plant Resistance
-Mechanisms like antixenosis, antibiosis, and tolerance give a plant the ability to resist insect pests [@Horber1980; @Wilson1987].
-Antixenosis allows the host plant to trigger resistance by inducing non-preference in herbivores when they attempt to feed or oviposit.
-Recent work done by Achhami et al.
-(2020) found that Barely cultivars that closely resemble wheat show the greatest frequency of WSS oviposition.
-These findings suggest that semiochemicals produced by these barley cultivars, such as (Z)-3-hexenyl acetate [@Piesik2008; @Buteler2009] and (Z)-β-ocimene [@Buteler2012], are likely similar to those produced by wheat [@Achhami2020].
-Antibiosis causes the improper development of herbivores, impacting survivability and reproduction.
-Tolerance allows the host plant to withstand infestation and overcome injuries [@VanEmden1991; @VanZoeren2017].
-Much of the research done on WSS resistant wheat cultivars focuses on solid stem trait and pith expression [@Platt1949; @Varella2016; @Varella2017].
-Solid stem varieties are more resistant to infestation, reducing WSS oviposition and increasing egg mortality [@Holmes1960; @Varella2016].
-The increased pith expression also disrupts larvae movement within the stem, leading to higher larval mortality due to reduced feeding as well as smaller and less reproductively successful adults [@Wallace1966].
-In addition, greater pith expression increases stem strength, reducing losses due to lodged stems [@Delaney2010].
-Additional factors including late planting and shorter crop duration, have been suggested as potential factors complicating antixenosis, leading to increased WSS oviposition and infestation [@Achhami2020].
-The solid-stemmed trait has been developed in numerous spring and some winter wheat cultivars [@Beres2011].
-While the use of solid-stemmed cultivars is most often recommended for WSS management [@Beres2011], hollow-stemmed cultivars were found to exhibit greater parasitism when compared to solid-stemmed varieties [@Holmes1963].
-It was previously suggested by Rand et al.
-(2012)  that the thicker pith found in solid-stemmed cultivars makes sensing WSS larvae vibrations [@Mankin2004] more difficult for female parasitoids.
-The inability to locate WSS hosts would then cause ineffective parasitism.
-
-#### Tillage
-Tillage is a potential control method for the WSS.
-Larvae are protected against both water loss and freezing temperatures by the soil surrounding the wheat stubble where the WSS overwinters.
-Tilling the soil exposes the wheat stubble to the soil surface, conditions which kill the hibernating larvae [@Callenback1944].
-Tillage of the WSS has been shown to be effective in decreasing the pest’s populations in both spring and winter wheat cultivars.
-Upwards of 90% mortality has been shown in North Dakota after fall tillage [@Weiss1987].
-Many factors influence the effectiveness of tilling, including adequate exposure of wheat stubs on the soil surface and appropriate timing [@Holmes1956; @Morrill1992; @Weiss1992].
-While the variability in effectiveness exists, some growers do consider it a crucial management tool [@Blodgett1996; @Lesieur2016].
-
-While effective in killing WSS larvae, there are negative impacts caused by this practice.
-Water loss, soil compaction, and erosion have all been shown to be dramatically increased by tilling activities [@Phillips1980; @Lal1991].
-In addition, tilling practices negatively impact beneficial populations of parasitoid wasps that target the WSS.
-WSS parasitoids overwinter higher in the stem compared to at the base like WSS [@Morrill1998; @Runyon2002].
-By displacing soil when tilling, the upper portion of the plant is buried under the soil, preventing the parasitoid wasps from emerging in the spring [@Runyon2002; @Weaver2004].
-
-#### Plant Density and Spacing
-Historically, it has been observed that wheat fields with wider row spacing (12 inches) have greater WSS cutting when compared to fields with reduced spacing (3 to 6 inches).
-Additionally, fields planted at greater seed density with solid stem cultivars exhibit lower WSS cutting compared to reduced seed density [@Luginbill1958].
-The reason for this relationship between decreased density and increased sawfly infestation is due to the moisture availability and height of the wheat stems.
-At higher densities, each stem has access to less water, in turn leading to the stem to develop a thicker pith – more solid stemmed.
-As discussed above, there is a negative linear relationship between an increase in the solid-stem characteristic and infestation [@Kemp1934].
-
-#### Crop Rotation
-Continually planting wheat year after year creates ideal conditions for the WSS [@Callenbach1945].
-Rotating wheat with more resistant wheat varieties as well as with oats, or soybeans (crops that are less or not susceptible to WSS infestation) is recommended to decrease the year after year WSS infestation rates [@Lin2014].
-The notable downside to utilizing crop rotation in a matter that is effective is the economic downside.
-Alternate-year, summer fallow production systems are typical for wheat growing practices across much of the northern great plains of the United States [@Weiss1992; @Troeh1999].
-
-#### Trap Crops
-Trap cropping utilized non-economically important plantings to help reduce populations or block movement of pests into valuable crop areas [@Hokkanen1991].
-Effective trap crops of the WSS should have appealing characteristics to the WSS to attract the pest to the trap planting, and away from the wheat [@Farstad1940].
-Historically, planted trap strips were used by attracting sawflies and destroying the strip at the end of the season, eliminating the larvae [@Morrill1992].
-However, utilizing trap plantings of highly attractive wheat varieties has been shown to hold limited viability, as growers are not fond of destroying planted grain at the end of the growing season [@Goosey1999].
-
-Alternative utilization of WSS traps entail planting and/or maintaining strips of smooth brome grass (Bromus inermis Leyss.) or other WSS host grasses [@Farstad1940].
-Smooth brome in particular holds potential as it is a poor host to WSS larvae [@Bhandari2020], yet WSS adults continue to oviposit the stems.
-This relationship of greater oviposition with increased mortality may lead to decreases in overall WSS populations.
-
-#### Delayed Planting & Swathing
-Wheat planted later in the growing season avoids much of the damage caused by the WSS as the wheat reaches stem elongation after many of the adult sawflies have already laid their eggs [@Jacobson1952].
-The problem with delayed planting is that the practice is known to make less efficient use the earl season moisture in the soil [@Morrill1996].
-Growers may need to weigh the tradeoffs in delayed planting to reduce losses in areas with a history of severe infestation.
-Swathing, also known as windrowing, is the practice of cutting the wheat to accelerate the drying process and decrease the losses caused by WSS cutting [@Holmes1965; @Carlson2019; @McCullough2020].
-By cutting the wheat and letting it dry prior to sawfly cutting, swathing works by reducing losses due to lodging [@Ainslie1920].
-
-#### Biological Control Methods
-The use of biocontrol in limiting WSS populations has been well studied with variable results.
-There are currently three types of organisms that have been reported as biocontrol agents of the WSS in wheat fields: Soil fungal pathogens, the predatory beetle Phyllobaenus dubius, and two parasitic Hymenoptera.
-
-#### Soil Fungal Pathogens
-Fungal pathogens are found throughout both the plant and animal kingdoms.
-Several fungi have been shown to negatively impact WSS larvae infestation within wheat stems.
-While highly lethal to sawfly larvae, F.
-graminearum, F.
-acuminatum, F.
-equiseti, and F.
-avenaceum also cause crown rot in the host plant.
-The trade-off between WSS control and crown rot may limit Fusarium spp.
-usefulness to WSS management.
-[@Wenda2009].
-Findings by Tangtrakulwanich et al.
-(2014) found that the fungi B.
-bassiana and M.
-brunneum effectively reduced damage caused by WSS.
-Both fungi kill insects by infecting via spore contact or by insect consumption of treated plant tissue [@Portman2018].
-
-#### Predatory Beetle
-Phyllobaenus dubius (Wolcott) has been observed emerging from wheat stubble and within dissected wheat stems in Montana [@Morrill2001].
-Further research is required on this coleopteran to determine its prospects going forward as an effective WSS management tool.
-
-#### Parasitoids of the Wheat Stem Sawfly
-Two parasitoid wasps are known target the wheat stem sawfly in wheat fields.
-The congeneric parasitoids, Bracon cephi (Gahan) and B.
-lissogaster Muesebeck, are the most effective agents of WSS biological control [@Runyon2001].
-B.
-cephi and B.
-lissogaster, as distinguished by Runyon et al.
-(2001), were initially limited in their ability to parasitize WSS larvae within wheat stems.
-The wheat stem sawfly, originally known as the western grass-stem sawfly, was originally only found within native grasses throughout North America, existing in alongside B.
-cephi and B.
-lissogaster [@Ainslie1920; Cárcamo et al., 2012).
-However, as agriculture increased and native grasses were replaced by large monoculture landscapes, the WSS moved into wheat [@Farstad1940; @Davis1955; Holmes1978] while the parasitoids were slower to adapt [@Morrill1998].
-
-## Biology of *Bracon cephi* and *Bracon lissogaster*
-### Adult
-Both *B.
-cephi* and *B.
-lissogaster* are protelean host-specific ectoparasitoids with two generations per year (bivoltine) [@Nelson1953; @Somsen1956; @Holmes1963].
-They are ectoparasitoids and idiobionts, living on the integument of their hosts and immediately paralyzing their hosts in location and stage it was when it was attacked [@Nelson1953; @Runyon2002; @Weaver2004].
-Male parasitoids are approximately 3.5 mm in length, while females are 4.1 mm.
-They are dark red, with black antennae, eyes, ovipositor and ocelli [@Gahan1918].
-B.
-lissogaster and B.
-cephi are distinguishable by their head color and unique metasoma surface texture [@Runyon2001].
-Adults emerge after completing a pupal stage within the host stem by chewing a hole through the stem wall.
-In Montana, the first generations are present from late June to late July, while the second generations are active in August and September [@Somsen1956].
-Adult female B.
-cephi and B.
-lissogaster will locate a host stem using a combination of olfactory and volatile cues.
-The wasps will then locate a host sawfly larva using acoustic cues [@Mankin2004] and use their ovipositor first paralyze, then deposit 1-4 eggs next to the paralyzed larva.
-
-### Egg
-*B.
-cephi* and *B.
-lissogaster* eggs are on average 0.86 mm in length, exhibiting an elongated shape and pale-yellow color.
-Females lay one to three eggs on the WSS larval integument or adjacent to the larva.
-The eggs develop for one to three days before hatching [@Nelson1953; @Somsen1956].
-
-### Larva
-Parasitoid larvae are 1mm in length, with a translucent hue that turns increasingly brown as the larva progresses through five instar stages [@Nelson1953].
-The newly hatched larvae have no appendages, save for antennae, and they exhibit highly sclerotized mouthparts and well-developed mandibles.
-After hatching, the B.
-lissogaster larva attach themselves to the integument of the host larva and immediately begin feeding [@Somsen1956].
-When feeding is complete, little remains of the host but the integument.
-Following feeding, the larva will spin a lightly brown colored cocoon over the course of several days and attach to the inside wall of the host plant stem.
-The larval stage takes an average of ten days between initial hatch and cocoon completion.
-Because there are two generations per year, the early summer offspring will hatch and emerge from the host stem.
-Offspring laid as a part of the second generation will overwinter within the cocoons and emerge in the spring.
-[@Nelson1953]
-
-### WSS and Parasitoid Interactions with Smooth Brome
-Bhandari (2020) has shown that oviposition occurs with similar frequency in wheat as it does in smooth brome, however, we see much higher rates of larval mortality within smooth brome stems.
-This has led researchers to hypothesize about the feasibility of using smooth brome as a trap cop [@mendoza2006; @Lesieur2016].
-While feasible as a feed crop, as it holds considerably less economic value compared to wheat or other high value crops [@Criddle1922].
-Smooth brome may be usable in trap strips along the edge of wheat fields [@Hokkanen1991].
-While a poor host for the wheat stem sawfly, surviving immature larvae found within brome stems can be parasitized by B.
-cephi and B.
-lissogaster.
-This combination of factors may indicate that smooth brome can function as a WSS sink and a parasitoid source, with very few WSS emerging from smooth brome patches, while also supporting populations of parasitoids that can then parasitize WSS in adjacent wheat fields.
-
-## Plant Semiochemicals
-
-Plants interact with other organisms through a complex system of chemical signals known as Semiochemicals.
-Semiochemicals are compounds or mixtures of compounds that are released from one organism that evoke a physiological or behavior response in another organism [@Agelopoulos1999].
-Semiochemicals play a large role in both insect-insect and plant-insect interactions and are essential in insects locating feedings sites, mating sites, egg-laying sites, and/or refugia [@Prokopy1984].
-
-While it is known that these chemicals play an important role in communication between plants, insects, and other organisms, there is still debate about the function of many of them [@Reddy2004; @Voelckel2018].
-
-### Semiochemicals in Plant-Insect Systems
-
-Within plant-insect systems, plants respond to insect herbivory by synthesizing and releasing volatile compounds and providing important signals to biocontrol agents to localize potential host insects [@Sharma2019].
-In addition, semiochemicals provide localized signatures that allow insects to detect food sources, find mates, and locate preferential oviposition sites [@Norin2007].
-Plants typically produce semiochemicals when defending themselves from herbivorous arthropods [Guerrieri1999] or pathogen attacks [@Smart2014].
-These responses can be grouped into two categories, direct and indirect responses.
-Direct responses are compounds that the plant produces that have a direct effect on the herbivore or pathogen [@Metcalf1998].
-Indirect responses are the compounds produced that have no impact directly on the pest, but instead may attract parasitoids, predators, or other biological control agents [@Mayland2000; @Colazza2004].
-
-# Limitations And Key Assumptions
-## Field Study
-i.	Time limits the number field sites that we can collect samples from throughout the summer.  
-a.	Study area limited to western Montana – three regions – Big Sandy, Moccasin, and Amsterdam.  
-ii.	Our study will only take place over the course of 2-3 summers, thus, we are unable to make broader claims surrounding long term patterns of WSS and parasitoid interactions.   
-Green House Samples and Volatile Collection
-iii.	We will assume that plants grown in the green house will be accurate and reasonable biological model representation of plants present in the field.  
-iv.	Volatiles will be collected from a maximum of 12 plants a day due to constraints with the collection system.
-
-# Contributions to Knowledge of Potential Outcomes
-## Potential outcomes
-### Volatiles
-We hope to observe a discernable difference in the volatile quantities produced by both infested smooth brome and winter wheat.
-
-# Methodology
-## Field Experiments
-### Field Survey
-To assess *C.
-cinctus* infestation, larval mortality, and parasitoid prevalence within *B.
-inermis* and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production.
-Samples were collected in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA.
-*B.
-inermis* sites were set up as polygons along the edge of adjoining wheat fields measuring approximately 100 meters in area.
-Four collection squares of 1ft x 1ft were randomly partitioned within each polygon during both collection events each year.
-All stems within each 1 x 1 ft square were collected.
-The stems were then processed and split in the laboratory to assess infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting.
-
-### Controlled *C. cinctus* infestation of *B. inermis*
-#### Plant Selection and Field Cages
-Assessment of *C.
-cinctus* infestation and mortality within *B.
-inermis* were assessed using a 34 x 60 ft plot at the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
-Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.
-The cages were removed from the plot after 3 weeks of sawfly release.
-
-#### Insects
-Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of *C.
-cinctus* infestation and cutting the year prior.
-Cut stubble, which contained *C.
-cinctus* larvae in diapause, were kept refrigerated between -2°C and 3°C for >100 days as required to complete obligatory larval diapause.
-As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA).
-Stubs were added to cages once *B.
-inermis* stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).
-
-#### Data Collection
-Stems were collected from each cage site in late August.
-All stems from each 6 x 3 ft cage section were collected.
-Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.
-
-## Greenhouse Experiments
-### Organic Volatile Collection from *C. cinctus* Infested *B. inermis* and Winter Wheat
-#### Insects
-Wheat stubble collected the year prior to each experiment were used to rear adult WSS.
-Wheat stubble was collected from Big Sandy (48.17754°N, -110.11406°E) and Three Forks (46.020216°N, -111.586582°E), MT, USA.
-Stubble was collected from fields that experienced high levels of WSS infestation and cutting the year prior.
-Cut stubble, which contained WSS larvae in diapause, were kept in a refrigerator between 0-4 °C for >100 days as required to complete obligatory larval diapause.
-After completing diapause, immature WSS were kept at room temperature between 22-27 °C for 4-5 weeks in sealed 70 cm x 35 cm x 20 cm plastic Tupperware$^\circledR$ bins to allow for pupation and emergence.
-Once emergence began, adults were collected daily and sorted in 2-liter glass mason jars sealed with filter paper to allow for airflow.
-Jars containing adults were refrigerated at 6-10°C until they were needed for experiments.
-Unused adults were discarded after 48 hours of refrigeration.
-
-#### Plants
-Smooth brome was grown from transplanted stems harvested from the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-Immediately following collection, stems were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc$^\circledR$, Tangent, OR).
-Winter wheat plants were grown using Yellowstone Winter Wheat cultivar seed stock, started in 1 x 8 inch conical pots.
-The soil used consisted of MSU Plant Growth Center soil mix (equal parts sterilized Bozeman silt loam soil and washed concrete sand with Canadian sphagnum peat moss incorporated) and Sunshine Mix 1 (Canadian sphagnum peat moss, vermiculite, perlite, and Dolomite lime) in a 1:1 ratio.
-Fertilization was undergone once weekly using Jack’s Professional$^\circledR$ Water-Soluble Fertilizer (20-20-20) (J.R.
-Peters Inc., Allentown, PA, USA).
-The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22 $\pm$ 2°C and 20 – 40 % RH.
-After the plants reached Zadoks 12, they were moved into vernalization storage for 6 weeks to experience the length of dormancy needed for successful flower development.
-After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc$^\circledR$ conical pots.
-
-#### Sawfly Infestation
-Winter wheat plants were deemed ready for infestation once they reached Zadoks 49.
-Both smooth brome and winter wheat plants were infested by placing a transparent plastic cylinder (60 x 3.8 cm) over the main stem of the plant.
-A 55-60 cm bamboo stick was added within each tube to help prevent toppling of the cylinder.
-Each cage had three holes, as well as an open top, covered with fine mesh cloth to allow for air circulation.
-Four females and two male WSS, newly emerged, were introduced into each tube via a small 1 cm opening.
-After which, the hole was plugged using a cotton ball covered in mesh cloth.
-Damp soil was mounded around the base of each tube to prevent insect escapees.
-Adult sawflies were left within infestation tubes for three days, after which the tubes were removed, and any surviving insects, as well as cadavers, were removed.
-After infestation, the plants were returned to normal water and fertilization schedules.
-
-Infestation was undertaken three days a week, for five weeks.
-Each replicate included 6 ‘Yellowstone’ winter wheat plants at Zadoks 49 and 6 smooth brome plants exhibiting three nodes.
-For each replicate, three wheat and three smooth brome plants were infested, with three of each left uninfested.
-Infestation tubes were placed around all twelve plants to maintain control.
-
-#### Volatile Collection
-Volatile compounds were collected from intact and healthy smooth brome and Yellowstone winter wheat.
-Collection was undergone on plants at Zadoks 49, for a period of six hours from 09:00 to 15:00.
-Collections were undergone three times a week, with twelve plants collected from each time.
-Each trial consisted of six smooth brome plants and six winter wheat, with four of each being infested, while the other two were controls.
-
-Volatiles were collected from twelve plants per day using glass VCC’s that were 80 cm long and 4 cm in diameter.
-The top end of each VCC funneled into a threaded port that was 10 cm in diameter.
-Threaded onto the top of each VCC was an eight-port manifold adapter.
-Two glass filters (traps) (6.35 mm diameter x 76 mm length; Analytical Research Systems, Gainesville, FL) with 30 mg of Super-Q absorbent (Alltech Associates, Deerfield, IL) were inserted into two of the adapter ports.
-The remaining six collection ports were sealed to prevent ambient air exchange and pressure loss.
-One of the traps was used to collected extraneous volatiles from the initial experiment set up for the first ten minutes, whereas the other trap was used to collect volatile compounds emanating from plants beginning ten minutes after the system was turned on.
-An air source was connected to the base of each collection chamber with supply from charcoal air filters.
-A vacuum pump was used to maintain air pressure and flow rate (humidified air at 1.0 liter/min) within the collection system.
-A small piece of cotton encompassed by a Teflon® guillotine was used at the base of each plant to tightly seal the base of the system and prevent external air from entering.
-Laboratory tape was used to seal the edges of the guillotine and prevent contamination by ambient and soil derived volatiles.
-
-#### Volatile Elution
-Volatile compounds were eluted from the glass traps using 200 $\mu$L aliquot dichloromethane.
-Trapped volatiles were eluted slowly by adding dichloromethane and further clearing by using a slow release of nitrogen gas.
-The eluted samples were collected in a glass insert held within a 1.5-mL crimped top glass vial.
-10 $\mu$L of nonyl acetate were then added to the eluted samples as internal standard to quantify compound concentrations.
-The samples were processed using gas chromatography (GC) on a HP – 5MS; 30m x 0.25mm, 0.25 $\mu$m film thickness column; (J and W Scientific, Folsom, CA).
-In order to analyze, the GC instrument (Agilent 6890; Agilent Technologies, Santa Clara, CA) was coupled to a mass spectrometer (MS, Agilent 5973 instrument).
-Each sample was injected using the automated system in pulse splitless mode, with the initial pressure set to 82.7 kPa per minute.
-The inlet temperature of the GC was set to 250°C while the column temperature was 50°C for 4 minutes.
-The temperature was set to increase at a rate of 5°C per minute until it reached 160°C, after which it increased to a rate of 25°C per minute until it reached 280°C.
-The temperature of the transfer line leading to the mass selective detector (MSD) was set at 300°C.
-Helium was used as a carrier gas to maintain the flow rate of samples within the column at 1.2 mL/min.
-The MSD was set in ‘SCAN’ mode running from 50 – 300 m/z.
-
-#### Volatile Analysis
-The compounds collected from both the smooth brome and wheat samples were identified by comparing mass spectra and retention times using the National Institute of Standard and Technology (NIST) library.
-Peaks and volatiles were analyzed using MSD ChemStation (Agilent Technologies, Inc., Santa Clara, CA).
-The data obtained from the mass-spectrometer was transferred to a Microsoft® Excel spreadsheet where the compounds were quantified by comparison with the internal standard peak area.
-The amount obtained for each compound was then transformed to the amount produced in six hours.
-The true quantity of compounds was calculated relative to the internal standard.
-
-### Behavioral Assessment of Parasitoids in Response to *B. inermis* Organic Volatile Compounds
-#### Insects
-Parasitoids B.
-cephi and B.
-lissogaster were collected from wheat fields in Three Forks, MT (46.020216° N, -111.586582° E).
-Wheat stem residue from fields that experienced high levels of *C.
-cinctus* infestation were collected and stored inside 170-liter plastic barrel liners.
-Filled liners were stored in a cold storage room at 4°C until mid or late spring of the following year.
-After the storage time exceeded the mandatory duration needed for obligatory diapause, the plastic barrel liner was removed from cold storage and the residue was placed inside 170 L trash barrels at room temperature and watered to encourage emergence.
-Collected parasitoids were kept at 6-10°C until used.
-
-#### Electroantennography
-Specimens were prepared by excising the head from the body using fine micro-dissection scissors.
-For each specimen, a micro-capillarity tube was pulled in half using a pipette puller (Model PP-830, Narishige Scientific Instrument Lab, Tokyo, Japan).
-Each half was cut to the appropriate length and angle, then filled with saline solution.
-The parasitoid head was mounted on the broken end of one of the capillary tubes, then the tube was place on one of the conducting electrode wires connected to a micromanipulator, type INR-5 (Narishige MN-151).
-The second half of the capillary tube was place on the other electrode wire, and one insect antennae was positioned to touch the empty capillary tube, thus connecting the circuit.
-The amplifier transferred the information to a computer interface with an EAG version 2.7 software (Syntech NL 2001, Hilversum, Netherlands).
-This apparatus and method has been consistently used previously in assessing insect volatile response [@Agelopoulos1999].
-
-A consistent stream of purified, humidified air (20 mL sec-1) was delivered onto the antennae.
-The test compound was delivered via airborne stimulus released from a glass pipette with filter paper treated with 10 $\mu$L of the test compound.
-The pipette was connected to a Stimulus Controller CS-55 (Syntech NL 2010, Hilversum, The Netherlands).
-The equipment was programmed to deliver a 0.4 second puff of air through a Teflon tube and through the glass pipette.
-The electrical response of the antenna to the test compound was detected by the recording electrode and transferred to the amplifier, increasing the signal and delivering the information to the computer.
-Figure 2: Electroantennogram connected to GCMS. Useful for connecting insect electrochemical response to host plant produced volatile compounds.
-
-#### Y- Tube Olfactometer Bioassay
-Y-tube bioassays were conducted within the chemical ecology and insect behavior laboratory at the Plant Growth Center at Montana State University.
-Bioassays were undergone to determine behavioral choice preference response of individual female WSS to an air stream containing volatiles originating from smooth brome and a blank control.
-Bioassays were conducted using plants at growth stage Zadoks 49.
-The Y-tube system (Analytical Research Systems, Micanopy, FL) used was similar as described in a previous study [@Daisy2002] (Figure 1).
-A single plant was enclosed in a volatile collection chamber (VCC) (100 mm diameter, 254 mm length).
-Input air was charcoal-filtered and humidified, regulated by a diaphragm air pump that delivered air at a rate of 0.2 L min-1.
-Air was moved into the VCC through Teflon tubing connected with a 24/410 threaded glass joint.
-The bottom end of the VCC enclosed each plant and was tightly sealed by a Teflon guillotine (diameter 955 mm, 15 mm center opening), encircling the stem to prevent air leakage.
-The olfactory system used a Y-shaped glass tube that had two arms connected to the lower port of each VCC by Teflon tubing (diameter 0.64 cm), providing the airstream source from the plant.
-The VCC contained plant was illuminated by an enhanced spectrum LED grow light (sunshine Systems Grow UFO Light SS-Gu90w, Led Lighting4 Less, Paradise, CA) for one hour before the experiment to make the plant photosynthetically active.
-
-<!-- Figure 1 y tube diagram -->
-```{=html}
-<!DOCTYPE html>
-<html lang="en">
-<head>
-<meta charset="UTF-8">
-<meta name="viewport" content="width=device-width, initial-scale=1.0">
-<title>Image Cropping Example</title>
-<style>
-    .image-container {
-        width: auto; /* Set the width of the container */
-        height: auto; /* Set the height of the container */
-        overflow: hidden; /* Hide overflow */
-    }
-    .image-container img {
-        width: 100%; /* Make the image fill the container horizontally */
-        height: auto; /* Maintain aspect ratio */
-        object-fit: cover;
-        object-position: center;
-    }
-    .caption {
-        text-align: left; /* Center the caption */
-        font-style: italic; /* Style the caption */
-        margin-top: 10px; /* Add some space between the image and the caption */
-    }
-</style>
-</head>
-<body>
-
-<figure class="image-container">
-    <img src="../img/proposal/ytube_diagram.jpg" alt="Your Image">
-    <figcaption class="caption">Figure 1: Y-tube olfactory bioassay as constructed by Daisy et al. (2002). Used to measure insect behavior in the presence of volatile stimulus.</figcaption>
-</figure>
-
-</body>
-</html>
-
-```
-
-The Y-tube was made of bifurcated Corning® glass, measuring in at 28-mm diameter x 300 m length.
-The Y shape branched at an angle of 120, with each branch serving as an arm of the olfactometer.
-The stem of each Y-tube was 20 cm long.
-At the bifurcation, each branch extended 4 cm from the junction before ending in 10 cm parallel arms.
-The Y-tube was positioned horizontally inside a black box made of poster board (46 x 32 x 101.5 cm), designed to prevent the entry of unwanted ambient light.
-The black box had two holes on the upstream sie of the olfactometer, allowing a fiber optic illuminator (T-Q/FOI-1, Techni Quip Corp, El Segundo, CA) and Teflon tubing to pass through.
-The fiber optic illuminator was used to deliver light to both arms equally from the center of the bifurcation, while the Teflon tubing connected the VCC to the arms of the Y-tube.
-One arm of the Y-tube delivered volatiles from the enclosed plant while the other delivered filtered, pure air as a control.
-The movement of adult WSS is positively phototactic and the horizontally placed Y-tube was positioned such that the light distribution towards both arms was balanced to avoid bias.
-The arms of the Y-tube were oriented to minimize directional bias and the position of the plants was alternated between runs.
-
-For each assay, a single naïve female B.
-cephi or B.
-lissogaster parasitoid was introduced 1 cm into the unbranched end of the Y-tube.
-A 28-cm wire was used to facilitate parasitoid movement from the unbranched stem of the glass Y-tube to the test junction between the two upper arms.
-The order of individual test insects from a pool of naïve females was randomly assigned for each bioassay, and test plants were chosen randomly from the greenhouse culture before each experiment.
-All glassware was cleaned using dish soap and warm water, followed by solvent rinses with acetone and hexane.
-Cleaned and rinsed glassware was baked at 110°C in a glassware oven for 24 hours before use.
-New plants were used for each experiment.
-
-Each individual parasitoid was allowed three minutes to respond to either airstream source.
-Once a positive response was determined to either source, the trail was ended.
-If no choice was made after five minutes, the trail was ended and the result was recorded as no response.
-
-# Data Analysis
-## Field Experiments
-### Field Survey
-Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions.
-‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs awere used as the response variable to assess the differences in parameters among the sampling sites.
-ANOVA was used to caculate P-values using F-distribution to understand overall significant differences in mean *C.
+Statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions.
+‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs were used as the response variable to assess the differences in parameters among the sampling sites.
+ANOVA was used to calculate P-values using F-distribution to understand overall significant differences in mean *C.
 cinctus* infestation and parasitism among polygons.
 To assess the profile of *C.
-cintus* infestation, paraisitms, and larval presence by node, a linear model was fit for each sampling site.
-Further assessment was done on the resonpses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.
+cintus* infestation, parasitism, and larval presence by node, a linear model was fit for each sampling site.
+Further assessment was done on the responses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.
+
+To assess the importance of smooth brome presence for parasitoid population buffering, we used a
 
@@
 #### Organic Volatile Collection from *C. cinctus* Infested *B. inermis* and Winter Wheat
+
 The data analysis procedure for each collection type was the same.
@@
 #### Behavioral Assessment of Parasitoids in Response to *B. inermis* Organic Volatile Compounds
-
-# Citations

### 2024-05-13 qmd/helpful.qmd
commit 5/13 2024

--- 2024-03-27 qmd/helpful.qmd
@@
 
+Below I have put many of the links that helped me.
+
 ## R Studio
-Great R-Markdown guide made by [Jared Adam](https://www.linkedin.com/in/jared-adam-24b107241/){style="text-decoration: none;"}: 
-<a href="https://jared-adam-psu.github.io/R-Tutorial-for-Tooker-Lab/" target="_blank">One code to rule them all (┛ಠ_ಠ)┛彡O</a>
+
+-   Guide to using R made by [Jared Adam](https://www.linkedin.com/in/jared-adam-24b107241/){style="text-decoration: none;"}: <a href="https://jared-adam-psu.github.io/R-Tutorial-for-Tooker-Lab/" target="_blank">One code to rule them all (┛ಠ_ಠ)┛彡O</a>
+-   Helpful free tutorial for learning R Studio through Codecademy: <a href="https://www.codecademy.com/learn/learn-r?utm_source=pepperjam&utm_medium=affiliate&utm_term=159404&clickId=4716939938&pj_creativeid=8-12462&pj_publisherid=159404" target="_blank">Learn R with Codecademy</a>
+-   Learn how to use R Markdown to create pdf and word outputs with your code: <a href="https://rmarkdown.rstudio.com/lesson-1.html" target="_blank">R Markdown Documentation</a>
 
 ## Quarto Website
-YouTube video on how to make a R Studio Quarto Website: [How to create a web site using Quarto, RStudio, and GitHub Pages](https://youtu.be/YN75YXaLFGM?si=RyCG23IMBwFUKsO8)
+
+-   YouTube video on how to make a R Studio Quarto Website: [How to create a web site using Quarto, RStudio, and GitHub Pages](https://youtu.be/YN75YXaLFGM?si=RyCG23IMBwFUKsO8)
+
+```{=html}
+<!DOCTYPE html>
+<html lang="en">
+<head>
+    <meta charset="UTF-8">
+    <meta name="viewport" content="width=device-width, initial-scale=1.0">
+    <title>How to create a web site using Quarto, RStudio, and GitHub Pages</title>
+</head>
+<body>
+    <!-- <h2></h2> -->
+    <!-- Replace 'VIDEO_ID' with the actual ID of your YouTube video -->
+    <iframe width="560" height="315" src="https://www.youtube.com/embed/YN75YXaLFGM" frameborder="0" allowfullscreen></iframe>
+</body>
+</html>
+```
+-   Gallery showing different options for formatting a Quarto website: [Quarto Gallery](https://quarto.org/docs/gallery/)
+-   Quarto documentation showing how to do a lot of the basics including website navigation, website search, and website tools: <a href="https://quarto.org/docs/websites/" target="_blank">Quarto - Creating a Website</a>
 
 ## GitHub
-A great guide to using GitHub created by [Callum Arnold](https://www.linkedin.com/in/arnold-c/){style="text-decoration: none;"}: <a href="https://psu-git.callumarnold.com" target="_blank">Introduction to Git and GitHub</a>
+
+-   A guide to using GitHub created by [Callum Arnold](https://www.linkedin.com/in/arnold-c/){style="text-decoration: none;"}: <a href="https://psu-git.callumarnold.com" target="_blank">Introduction to Git and GitHub</a>
+
+Watershed volunteering
+https://youtu.be/65MAB0qRtAE?si=beO8jD5t5SOJzooF

### 2024-05-13 qmd/litreview.qmd
commit 5/13 2024

@@
 <!-- Button to download word document version -->
+
 ```{=html}
@@
 ```
+# Project Overview
 
-# Project Overview
 ## Motivation
+
 The motivation for my research is to improve yield for wheat growers by reducing wheat stem sawfly (WSS) populations.
@@
 ## Research Question
+
 What role does smooth brome (*Bromus inermis*) play in the ecological relationship between wheat stem sawfly (*Cephus cinctus*) and wheat stem sawfly parasitoids (*Bracon cephi* and *Bracon lissogaster*) in dryland wheat production landscape within Montana?
@@
 ## Hypothesis
+
 Smooth brome acts as an ecological sink for WSS, while serving as a parasitoid source because WSS larvae experience high mortality within smooth brome stems, while any survivors experience high rates of parasitism as a second source of mortality.
@@
 ## Prediction
+
 WSS larvae will experience both high levels of mortality and similar levels of parasitism within smooth brome stems when compared to the survivability and parasitism of WSS larvae in the adjacent fields.
@@
 ## Research Approach
+
 My research is composed of three separate studies.
@@
 ## Importance of Topic
+
 The wheat stem sawfly (WSS, Cephus cinctus Norton, Hymenoptera: Cephidae), is a serious problem for wheat producers throughout the Northern Great Plains of North America [@Fulbright2017].
-In Montana alone, the economic damage caused by the pest was reported to be upwards of $80 million annually between 2011-2016 [@Fulbright2017].
+In Montana alone, the economic damage caused by the pest was reported to be upwards of \$80 million annually between 2011-2016 [@Fulbright2017].
 Current techniques in combatting the damage caused by the WSS use a variety of chemical, cultural, and biocontrol strategies with limited short-term effectiveness and long-term sustainability.
@@
 # Literature Review
+
 ## History and Distribution of Wheat Stem Sawfly
+
 The wheat stem sawfly was first described in 1872 by Edward Norton based upon a specimen found in Colorado [@Norton1872].
@@
 ## Host Plants of Wheat Stem Sawfly
+
 Wheat stem sawfly was first found the hollow-stemmed wild grasses such as Agropyron and Elymus in Colorado in 1872 [@Ainslie1920].
@@
 ## Biology of Wheat Stem Sawfly
+
 ### Adult
+
 Adults are relatively small (8-13 mm) with a narrow elongated and slightly compressed black body, yellow legs, and three characteristic black bands across the abdomen.
@@
 ### Oviposition
+
 After locating a suitable host stem, female WSS oviposit an egg inside the host plant lumen.
@@
 ### Egg
+
 Newly laid eggs are crescent in shape, milky-white in color, and range between 1 mm to 1.25 mm in length and 0.33 to 0.42 mm wide [@Ainslie1920].
@@
 ### Larvae
+
 After hatching and prior to feeding, the larvae are transparent and colorless [@Ainslie1920; @Criddle1923].
@@
 ### Pupae
+
 Once diapause is complete, pupal development begins.
@@
 ## Damage Caused by Wheat Stem Sawfly
+
 Wheat stem sawfly larvae travel throughout the stem of the host plant, feeding on the parenchymal tissue and chewing through the nodes.
@@
 ## Management of Wheat Stem Sawfly
+
 ### Chemical Control
+
 Because WSS are stem mining pests and spend most of their lives within the host plant stem, the use of traditional pest control methods such as insecticides have proven to be ineffective.
@@
 ### Cultural Control
+
 #### Host Plant Resistance
+
 Mechanisms like antixenosis, antibiosis, and tolerance give a plant the ability to resist insect pests [@Horber1980; @Wilson1987].
@@
 #### Tillage
+
 Tillage is a potential control method for the WSS.
@@
 #### Plant Density and Spacing
+
 Historically, it has been observed that wheat fields with wider row spacing (12 inches) have greater WSS cutting when compared to fields with reduced spacing (3 to 6 inches).
@@
 #### Crop Rotation
+
 Continually planting wheat year after year creates ideal conditions for the WSS [@Callenbach1945].
@@
 #### Trap Crops
+
 Trap cropping utilized non-economically important plantings to help reduce populations or block movement of pests into valuable crop areas [@Hokkanen1991].
@@
 #### Delayed Planting & Swathing
+
 Wheat planted later in the growing season avoids much of the damage caused by the WSS as the wheat reaches stem elongation after many of the adult sawflies have already laid their eggs [@Jacobson1952].
@@
 #### Biological Control Methods
+
 The use of biocontrol in limiting WSS populations has been well studied with variable results.
@@
 #### Soil Fungal Pathogens
+
 Fungal pathogens are found throughout both the plant and animal kingdoms.
@@
 #### Predatory Beetle
+
 Phyllobaenus dubius (Wolcott) has been observed emerging from wheat stubble and within dissected wheat stems in Montana [@Morrill2001].
@@
 #### Parasitoids of the Wheat Stem Sawfly
+
 Two parasitoid wasps are known target the wheat stem sawfly in wheat fields.
@@
 ## Biology of *Bracon cephi* and *Bracon lissogaster*
+
 ### Adult
+
 Both *B.
@@
 ### Egg
+
 *B.
@@
 ### Larva
+
 Parasitoid larvae are 1mm in length, with a translucent hue that turns increasingly brown as the larva progresses through five instar stages [@Nelson1953].
@@
 ### WSS and Parasitoid Interactions with Smooth Brome
+
 Bhandari (2020) has shown that oviposition occurs with similar frequency in wheat as it does in smooth brome, however, we see much higher rates of larval mortality within smooth brome stems.
@@
 In addition, semiochemicals provide localized signatures that allow insects to detect food sources, find mates, and locate preferential oviposition sites [@Norin2007].
-Plants typically produce semiochemicals when defending themselves from herbivorous arthropods [Guerrieri1999] or pathogen attacks [@Smart2014].
+Plants typically produce semiochemicals when defending themselves from herbivorous arthropods \[Guerrieri1999\] or pathogen attacks [@Smart2014].
 These responses can be grouped into two categories, direct and indirect responses.
@@
 # Limitations And Key Assumptions
+
 ## Field Study
-i.	Time limits the number field sites that we can collect samples from throughout the summer.  
-a.	Study area limited to western Montana – three regions – Big Sandy, Moccasin, and Amsterdam.  
-ii.	Our study will only take place over the course of 2-3 summers, thus, we are unable to make broader claims surrounding long term patterns of WSS and parasitoid interactions.   
+
+i.  Time limits the number field sites that we can collect samples from throughout the summer.\
+
+```{=html}
+<!-- -->
+```
+a.  Study area limited to western Montana – three regions – Big Sandy, Moccasin, and Amsterdam.\
+
+```{=html}
+<!-- -->
+```
+ii. Our study will only take place over the course of 2-3 summers, thus, we are unable to make broader claims surrounding long term patterns of WSS and parasitoid interactions.\
     Green House Samples and Volatile Collection
-iii.	We will assume that plants grown in the green house will be accurate and reasonable biological model representation of plants present in the field.  
+iii. We will assume that plants grown in the green house will be accurate and reasonable biological model representation of plants present in the field.\
 iv. Volatiles will be collected from a maximum of 12 plants a day due to constraints with the collection system.
@@
 # Contributions to Knowledge of Potential Outcomes
+
 ## Potential outcomes
+
 ### Volatiles
+
 We hope to observe a discernable difference in the volatile quantities produced by both infested smooth brome and winter wheat.
 
-# Methodology
-## Field Experiments
-### Field Survey
-To assess *C.
-cinctus* infestation, larval mortality, and parasitoid prevalence within *B.
-inermis* and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production.
-Samples were collected in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA.
-*B.
-inermis* sites were set up as polygons along the edge of adjoining wheat fields measuring approximately 100 meters in area.
-Four collection squares of 1ft x 1ft were randomly partitioned within each polygon during both collection events each year.
-All stems within each 1 x 1 ft square were collected.
-The stems were then processed and split in the laboratory to assess infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting.
-
-### Controlled *C. cinctus* infestation of *B. inermis*
-#### Plant Selection and Field Cages
-Assessment of *C.
-cinctus* infestation and mortality within *B.
-inermis* were assessed using a 34 x 60 ft plot at the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
-Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.
-The cages were removed from the plot after 3 weeks of sawfly release.
-
-#### Insects
-Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of *C.
-cinctus* infestation and cutting the year prior.
-Cut stubble, which contained *C.
-cinctus* larvae in diapause, were kept refrigerated between -2°C and 3°C for >100 days as required to complete obligatory larval diapause.
-As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA).
-Stubs were added to cages once *B.
-inermis* stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).
-
-#### Data Collection
-Stems were collected from each cage site in late August.
-All stems from each 6 x 3 ft cage section were collected.
-Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.
-
-## Greenhouse Experiments
-### Organic Volatile Collection from *C. cinctus* Infested *B. inermis* and Winter Wheat
-#### Insects
-Wheat stubble collected the year prior to each experiment were used to rear adult WSS.
-Wheat stubble was collected from Big Sandy (48.17754°N, -110.11406°E) and Three Forks (46.020216°N, -111.586582°E), MT, USA.
-Stubble was collected from fields that experienced high levels of WSS infestation and cutting the year prior.
-Cut stubble, which contained WSS larvae in diapause, were kept in a refrigerator between 0-4 °C for >100 days as required to complete obligatory larval diapause.
-After completing diapause, immature WSS were kept at room temperature between 22-27 °C for 4-5 weeks in sealed 70 cm x 35 cm x 20 cm plastic Tupperware$^\circledR$ bins to allow for pupation and emergence.
-Once emergence began, adults were collected daily and sorted in 2-liter glass mason jars sealed with filter paper to allow for airflow.
-Jars containing adults were refrigerated at 6-10°C until they were needed for experiments.
-Unused adults were discarded after 48 hours of refrigeration.
-
-#### Plants
-Smooth brome was grown from transplanted stems harvested from the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-Immediately following collection, stems were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc$^\circledR$, Tangent, OR).
-Winter wheat plants were grown using Yellowstone Winter Wheat cultivar seed stock, started in 1 x 8 inch conical pots.
-The soil used consisted of MSU Plant Growth Center soil mix (equal parts sterilized Bozeman silt loam soil and washed concrete sand with Canadian sphagnum peat moss incorporated) and Sunshine Mix 1 (Canadian sphagnum peat moss, vermiculite, perlite, and Dolomite lime) in a 1:1 ratio.
-Fertilization was undergone once weekly using Jack’s Professional$^\circledR$ Water-Soluble Fertilizer (20-20-20) (J.R.
-Peters Inc., Allentown, PA, USA).
-The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22 $\pm$ 2°C and 20 – 40 % RH.
-After the plants reached Zadoks 12, they were moved into vernalization storage for 6 weeks to experience the length of dormancy needed for successful flower development.
-After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc$^\circledR$ conical pots.
-
-#### Sawfly Infestation
-Winter wheat plants were deemed ready for infestation once they reached Zadoks 49.
-Both smooth brome and winter wheat plants were infested by placing a transparent plastic cylinder (60 x 3.8 cm) over the main stem of the plant.
-A 55-60 cm bamboo stick was added within each tube to help prevent toppling of the cylinder.
-Each cage had three holes, as well as an open top, covered with fine mesh cloth to allow for air circulation.
-Four females and two male WSS, newly emerged, were introduced into each tube via a small 1 cm opening.
-After which, the hole was plugged using a cotton ball covered in mesh cloth.
-Damp soil was mounded around the base of each tube to prevent insect escapees.
-Adult sawflies were left within infestation tubes for three days, after which the tubes were removed, and any surviving insects, as well as cadavers, were removed.
-After infestation, the plants were returned to normal water and fertilization schedules.
-
-Infestation was undertaken three days a week, for five weeks.
-Each replicate included 6 ‘Yellowstone’ winter wheat plants at Zadoks 49 and 6 smooth brome plants exhibiting three nodes.
-For each replicate, three wheat and three smooth brome plants were infested, with three of each left uninfested.
-Infestation tubes were placed around all twelve plants to maintain control.
-
-#### Volatile Collection
-Volatile compounds were collected from intact and healthy smooth brome and Yellowstone winter wheat.
-Collection was undergone on plants at Zadoks 49, for a period of six hours from 09:00 to 15:00.
-Collections were undergone three times a week, with twelve plants collected from each time.
-Each trial consisted of six smooth brome plants and six winter wheat, with four of each being infested, while the other two were controls.
-
-Volatiles were collected from twelve plants per day using glass VCC’s that were 80 cm long and 4 cm in diameter.
-The top end of each VCC funneled into a threaded port that was 10 cm in diameter.
-Threaded onto the top of each VCC was an eight-port manifold adapter.
-Two glass filters (traps) (6.35 mm diameter x 76 mm length; Analytical Research Systems, Gainesville, FL) with 30 mg of Super-Q absorbent (Alltech Associates, Deerfield, IL) were inserted into two of the adapter ports.
-The remaining six collection ports were sealed to prevent ambient air exchange and pressure loss.
-One of the traps was used to collected extraneous volatiles from the initial experiment set up for the first ten minutes, whereas the other trap was used to collect volatile compounds emanating from plants beginning ten minutes after the system was turned on.
-An air source was connected to the base of each collection chamber with supply from charcoal air filters.
-A vacuum pump was used to maintain air pressure and flow rate (humidified air at 1.0 liter/min) within the collection system.
-A small piece of cotton encompassed by a Teflon® guillotine was used at the base of each plant to tightly seal the base of the system and prevent external air from entering.
-Laboratory tape was used to seal the edges of the guillotine and prevent contamination by ambient and soil derived volatiles.
-
-#### Volatile Elution
-Volatile compounds were eluted from the glass traps using 200 $\mu$L aliquot dichloromethane.
-Trapped volatiles were eluted slowly by adding dichloromethane and further clearing by using a slow release of nitrogen gas.
-The eluted samples were collected in a glass insert held within a 1.5-mL crimped top glass vial.
-10 $\mu$L of nonyl acetate were then added to the eluted samples as internal standard to quantify compound concentrations.
-The samples were processed using gas chromatography (GC) on a HP – 5MS; 30m x 0.25mm, 0.25 $\mu$m film thickness column; (J and W Scientific, Folsom, CA).
-In order to analyze, the GC instrument (Agilent 6890; Agilent Technologies, Santa Clara, CA) was coupled to a mass spectrometer (MS, Agilent 5973 instrument).
-Each sample was injected using the automated system in pulse splitless mode, with the initial pressure set to 82.7 kPa per minute.
-The inlet temperature of the GC was set to 250°C while the column temperature was 50°C for 4 minutes.
-The temperature was set to increase at a rate of 5°C per minute until it reached 160°C, after which it increased to a rate of 25°C per minute until it reached 280°C.
-The temperature of the transfer line leading to the mass selective detector (MSD) was set at 300°C.
-Helium was used as a carrier gas to maintain the flow rate of samples within the column at 1.2 mL/min.
-The MSD was set in ‘SCAN’ mode running from 50 – 300 m/z.
-
-#### Volatile Analysis
-The compounds collected from both the smooth brome and wheat samples were identified by comparing mass spectra and retention times using the National Institute of Standard and Technology (NIST) library.
-Peaks and volatiles were analyzed using MSD ChemStation (Agilent Technologies, Inc., Santa Clara, CA).
-The data obtained from the mass-spectrometer was transferred to a Microsoft® Excel spreadsheet where the compounds were quantified by comparison with the internal standard peak area.
-The amount obtained for each compound was then transformed to the amount produced in six hours.
-The true quantity of compounds was calculated relative to the internal standard.
-
-### Behavioral Assessment of Parasitoids in Response to *B. inermis* Organic Volatile Compounds
-#### Insects
-Parasitoids B.
-cephi and B.
-lissogaster were collected from wheat fields in Three Forks, MT (46.020216° N, -111.586582° E).
-Wheat stem residue from fields that experienced high levels of *C.
-cinctus* infestation were collected and stored inside 170-liter plastic barrel liners.
-Filled liners were stored in a cold storage room at 4°C until mid or late spring of the following year.
-After the storage time exceeded the mandatory duration needed for obligatory diapause, the plastic barrel liner was removed from cold storage and the residue was placed inside 170 L trash barrels at room temperature and watered to encourage emergence.
-Collected parasitoids were kept at 6-10°C until used.
-
-#### Electroantennography
-Specimens were prepared by excising the head from the body using fine micro-dissection scissors.
-For each specimen, a micro-capillarity tube was pulled in half using a pipette puller (Model PP-830, Narishige Scientific Instrument Lab, Tokyo, Japan).
-Each half was cut to the appropriate length and angle, then filled with saline solution.
-The parasitoid head was mounted on the broken end of one of the capillary tubes, then the tube was place on one of the conducting electrode wires connected to a micromanipulator, type INR-5 (Narishige MN-151).
-The second half of the capillary tube was place on the other electrode wire, and one insect antennae was positioned to touch the empty capillary tube, thus connecting the circuit.
-The amplifier transferred the information to a computer interface with an EAG version 2.7 software (Syntech NL 2001, Hilversum, Netherlands).
-This apparatus and method has been consistently used previously in assessing insect volatile response [@Agelopoulos1999].
-
-A consistent stream of purified, humidified air (20 mL sec-1) was delivered onto the antennae.
-The test compound was delivered via airborne stimulus released from a glass pipette with filter paper treated with 10 $\mu$L of the test compound.
-The pipette was connected to a Stimulus Controller CS-55 (Syntech NL 2010, Hilversum, The Netherlands).
-The equipment was programmed to deliver a 0.4 second puff of air through a Teflon tube and through the glass pipette.
-The electrical response of the antenna to the test compound was detected by the recording electrode and transferred to the amplifier, increasing the signal and delivering the information to the computer.
-Figure 2: Electroantennogram connected to GCMS. Useful for connecting insect electrochemical response to host plant produced volatile compounds.
-
-#### Y- Tube Olfactometer Bioassay
-Y-tube bioassays were conducted within the chemical ecology and insect behavior laboratory at the Plant Growth Center at Montana State University.
-Bioassays were undergone to determine behavioral choice preference response of individual female WSS to an air stream containing volatiles originating from smooth brome and a blank control.
-Bioassays were conducted using plants at growth stage Zadoks 49.
-The Y-tube system (Analytical Research Systems, Micanopy, FL) used was similar as described in a previous study [@Daisy2002] (Figure 1).
-A single plant was enclosed in a volatile collection chamber (VCC) (100 mm diameter, 254 mm length).
-Input air was charcoal-filtered and humidified, regulated by a diaphragm air pump that delivered air at a rate of 0.2 L min-1.
-Air was moved into the VCC through Teflon tubing connected with a 24/410 threaded glass joint.
-The bottom end of the VCC enclosed each plant and was tightly sealed by a Teflon guillotine (diameter 955 mm, 15 mm center opening), encircling the stem to prevent air leakage.
-The olfactory system used a Y-shaped glass tube that had two arms connected to the lower port of each VCC by Teflon tubing (diameter 0.64 cm), providing the airstream source from the plant.
-The VCC contained plant was illuminated by an enhanced spectrum LED grow light (sunshine Systems Grow UFO Light SS-Gu90w, Led Lighting4 Less, Paradise, CA) for one hour before the experiment to make the plant photosynthetically active.
-
-<!-- Figure 1 y tube diagram -->
-```{=html}
-<!DOCTYPE html>
-<html lang="en">
-<head>
-<meta charset="UTF-8">
-<meta name="viewport" content="width=device-width, initial-scale=1.0">
-<title>Image Cropping Example</title>
-<style>
-    .image-container {
-        width: auto; /* Set the width of the container */
-        height: auto; /* Set the height of the container */
-        overflow: hidden; /* Hide overflow */
-    }
-    .image-container img {
-        width: 100%; /* Make the image fill the container horizontally */
-        height: auto; /* Maintain aspect ratio */
-        object-fit: cover;
-        object-position: center;
-    }
-    .caption {
-        text-align: left; /* Center the caption */
-        font-style: italic; /* Style the caption */
-        margin-top: 10px; /* Add some space between the image and the caption */
-    }
-</style>
-</head>
-<body>
-
-<figure class="image-container">
-    <img src="../img/proposal/ytube_diagram.jpg" alt="Your Image">
-    <figcaption class="caption">Figure 1: Y-tube olfactory bioassay as constructed by Daisy et al. (2002). Used to measure insect behavior in the presence of volatile stimulus.</figcaption>
-</figure>
-
-</body>
-</html>
-
-```
-
-The Y-tube was made of bifurcated Corning® glass, measuring in at 28-mm diameter x 300 m length.
-The Y shape branched at an angle of 120, with each branch serving as an arm of the olfactometer.
-The stem of each Y-tube was 20 cm long.
-At the bifurcation, each branch extended 4 cm from the junction before ending in 10 cm parallel arms.
-The Y-tube was positioned horizontally inside a black box made of poster board (46 x 32 x 101.5 cm), designed to prevent the entry of unwanted ambient light.
-The black box had two holes on the upstream sie of the olfactometer, allowing a fiber optic illuminator (T-Q/FOI-1, Techni Quip Corp, El Segundo, CA) and Teflon tubing to pass through.
-The fiber optic illuminator was used to deliver light to both arms equally from the center of the bifurcation, while the Teflon tubing connected the VCC to the arms of the Y-tube.
-One arm of the Y-tube delivered volatiles from the enclosed plant while the other delivered filtered, pure air as a control.
-The movement of adult WSS is positively phototactic and the horizontally placed Y-tube was positioned such that the light distribution towards both arms was balanced to avoid bias.
-The arms of the Y-tube were oriented to minimize directional bias and the position of the plants was alternated between runs.
-
-For each assay, a single naïve female B.
-cephi or B.
-lissogaster parasitoid was introduced 1 cm into the unbranched end of the Y-tube.
-A 28-cm wire was used to facilitate parasitoid movement from the unbranched stem of the glass Y-tube to the test junction between the two upper arms.
-The order of individual test insects from a pool of naïve females was randomly assigned for each bioassay, and test plants were chosen randomly from the greenhouse culture before each experiment.
-All glassware was cleaned using dish soap and warm water, followed by solvent rinses with acetone and hexane.
-Cleaned and rinsed glassware was baked at 110°C in a glassware oven for 24 hours before use.
-New plants were used for each experiment.
-
-Each individual parasitoid was allowed three minutes to respond to either airstream source.
-Once a positive response was determined to either source, the trail was ended.
-If no choice was made after five minutes, the trail was ended and the result was recorded as no response.
-
-# Data Analysis
-## Field Experiments
-### Field Survey
-Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions.
-‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs awere used as the response variable to assess the differences in parameters among the sampling sites.
-ANOVA was used to caculate P-values using F-distribution to understand overall significant differences in mean *C.
-cinctus* infestation and parasitism among polygons.
-To assess the profile of *C.
-cintus* infestation, paraisitms, and larval presence by node, a linear model was fit for each sampling site.
-Further assessment was done on the resonpses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.
-
-### Controlled *C. cinctus* infestation of *B. inermis*
-
-## Greenhouse Experiments
-
-#### Organic Volatile Collection from *C. cinctus* Infested *B. inermis* and Winter Wheat
-The data analysis procedure for each collection type was the same.
-Statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PERMANOVA, PERMDISP, and Random Forest modeling.
-The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide].
-VOC compositions were compared between samples and treatments using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan].
-This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017].
-Data were visualized using a combination of ggplot2 and knitr [@ggplot; @knitr].
-I selected a subset of compounds to ues for each analysis.
-These compounds were selected based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output.
-
-#### Behavioral Assessment of Parasitoids in Response to *B. inermis* Organic Volatile Compounds
-
 # Citations

### 2024-05-13 qmd/methodology.qmd
commit 5/13 2024

--- 2024-03-20 qmd/litreview_test.qmd
@@
 ---
-title: "Literature Review (Thesis)"
-author: "Jackson Strand"
-date: "2024-03-20"
-output: 
-  html_document:
-    toc: true
-    code_folding: hide
-bibliography: proposal.bib
-link-citations: true
-cls: council-of-science-editors-author-date.csl
-editor_options: 
-  chunk_output_type: console
-toc: true
+title: "Methodology"
+date: "April 5, 2024"
 ---
 
-# Project Overview
-## Motivation
-The motivation for my research is to improve yield for wheat growers by reducing wheat stem sawfly (WSS) populations.
-If we can show that healthier, more robust smooth brome leads to increased parasitoid populations and thus, increased WSS parasitism and potentially decreased damage by WSS, we can better educate growers on how to increase yields.
-In terms of scientific merit, understanding these mechanisms is key in ongoing WSS research as it explains several how and why questions associated with the ecological relationship between smooth brome and WSS parasitism.
+# Methodology
 
-## Research Question
-What role does smooth brome (Bromus inermis) play in the ecological relationship between wheat stem sawfly (Cephus cinctus) and wheat stem sawfly parasitoids (Bracon cephi and Bracon lissogaster) in dryland wheat production landscape within Montana?
+## Field Experiments
 
-## Hypothesis
-Smooth brome acts as an ecological sink for WSS, while serving as a parasitoid source because WSS larvae experience high mortality within smooth brome stems, while any survivors experience high rates of parasitism as a second source of mortality.
-In addition, early season parasitoids are attracted to infested smooth brome because it produces volatiles similar to those produced by winter wheat infested with larval WSS.
+### Field Survey
 
-## Prediction
-WSS larvae will experience both high levels of mortality and similar levels of parasitism within smooth brome stems when compared to the survivability and parasitism of WSS larvae in the adjacent fields.
-Also, the blend of volatile compounds produced by WSS infested smooth brome will qualitatively match in composition and exceed the amounts produced by infested winter wheat.
-
-## Research Approach
-My research is composed of three separate studies.
-First, there is a field survey to assess the survivability of WSS larvae and associated parasitoid frequency within both winter wheat and adjacent smooth brome.
-Second, there are controlled field infestations which will assess the survivability of WSS larvae within smooth brome stems when controlled for other density driven factors.
-And third, there is a greenhouse study to measure the volatile compounds produced by winter wheat and smooth brome.
-Using the results of the volatile collection, we can assess the parasitoid response both chemically with electroantennography and behaviorally with a Y-tube olfactory bioassay.
-
-## Importance of Topic
-The wheat stem sawfly (WSS, Cephus cinctus Norton, Hymenoptera: Cephidae), is a serious problem for wheat producers throughout the Northern Great Plains of North America [@Fulbright2017].
-In Montana alone, the economic damage caused by the pest was reported to be upwards of $80 million annually between 2011-2016 [@Fulbright2017].
-Current techniques in combatting the damage caused by the WSS use a variety of chemical, cultural, and biocontrol strategies with limited short-term effectiveness and long-term sustainability.
-Utilizing biological control via promoting braconid parasitoid habitat holds considerable value to both growers and agricultural observers.
-As we seek to increase agricultural yields, we must also consider the environmental impact that tactics such as pesticides and tillage have on the future success on wheat farming.
-
-# Prior Research on Topic (Literature Review)
-## History and Distribution of Wheat Stem Sawfly
-The wheat stem sawfly was first described in 1872 by Edward Norton based upon a specimen found in Colorado [@Norton1872].
-In 1890 it was found in California and Montana [@Ainslie1920].
-Native to North America [@Lesieur2016], the largest populations of WSS are currently found in the western Canadian prairies as well as Wyoming, Montana, Colorado, and North and South Dakota [@Wallace1966; @Weiss1992; @Cockrell2021].
-
-## Host Plants of Wheat Stem Sawfly
-Wheat stem sawfly was first found the hollow-stemmed wild grasses such as Agropyron and Elymus in Colorado in 1872 (Ainslie, 1920).
-Soon thereafter it was found in wheat in 1895 in Manitoba, Canada.
-First referred to as the “western-grass stem sawfly”, WSS was found with native Montana grasses including western wheatgrass (Pascopyrun smithii) and slender wheat grass (Elymus trachycaulus) (Ainslie, 1920).
-As settlers began moving west and small grain cereal crop production increased on the landscape, the native grasses that WSS foraged upon became quite rare (Ainslie, 1929).
-WSS was found damaging spring wheat for the first time in 1895 in Manitoba and Saskatchewan (Wallace and McNeal, 1966).
-With the diminishment of its natural hosts, WSS has adapted to infest the exotic non-cultivated and cultivated grasses now present (Criddle, 1923).
-WSS infestation has been shown in many species native to Montana including western wheatgrass (Pascopyrun smithii), slender wheat grass (Elymus trachycaulus) and bluebunch wheatgrass, Pseudoroegneria spicata (Cockrell et al., 2021).
-
-Smooth brome (Bromus inermis L.) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 from Hungary and Russia for soil conservation and use as livestock feed (Salesman and Jessica, 2011).
-Throughout the early 19th and 20th century, it was used throughout much of the upper half of North America.
-It is a perennial grass, and more drought resistant than other exotic grasses (Mapfumo et al., 2002).
-Due to its sod forming nature and large root system, it is extensively used to promote soil retention along highways (USDA-NRCS 2010).
-
-Smooth brome was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices (Farstad and Jacobson, 1945).
-However, due to its propensity to form monocultures and outcompete many native grasses, smooth brome was labeled by many as invasive (Carlson and Newell, 1985; Willson and Stubbendieck, 2000; Dillemuth et al., 2008).
-Smooth brome populations play a considerable role in altering native herbivore population dynamics and movement on the prairies of North America (Dillemuth et al., 2008).
-
-## Biology of Wheat Stem Sawfly
-### Adult
-Adults are relatively small (8-13 mm) with a narrow elongated and slightly compressed black body, yellow legs, and three characteristic black bands across the abdomen.
-Females are larger than males and a visible long, dark, ovipositor.
-Both male and female adults are weak fliers, rarely flying long distances at one time (Criddle, 1923).
-They can be found on stems with their heads downward and legs aligned with their body, encircling the stem (Ainslie, 1920; Criddle, 1923).
-WSS are univoltine, having only one generation per year.
-Males are haploid (9 chromosomes), while females are diploid – females who do not find a mate will produce male offspring, while those who do produce female progeny (Holmes, 1978).
-Fully mature adults hatch in the spring once it has sufficiently warmed and the moisture levels in the soil reach threshold levels (Holmes, 1978; Perez-Mendoza and Weaver, 2006).
-In Montana, this period begins in mid- to late-May and ends early in July (Morrill and Kushnak, 1996; Weaver et al., 2005).
-Their lifespan usually lasts around five to eight days, emerging from May until June.
-Males emerge before the females, ensuring that many of the early females to hatch will find a mate and deposit eggs that will be female (Mackay, 2011).
-Copulation lasts less than one minute with the male holding the back of the female with their abdomen curved down under the female ovipositor.
-Fertilized females lay eggs several hours after copulation (Holmes, 1978).
-
-### Oviposition
-After locating a suitable host stem, female WSS oviposit an egg inside the host plant lumen.
-WSS larvae hatch inside the stem and feed on the stem’s parenchyma tissue (Beres et al., 2011).
-As the stem begins to senesce in the fall, the larvae travel to the base of the stem and makes a V-shaped groove (Criddle, 1923) above a constructed protective hibernaculum where they will overwinter (Holmes and Peterson, 1960).
-As a result of this deep groove, the stem weakens and topples easily.
-The resulting lodged stems are not only challenging for growers to harvest but also exhibit decreased kernel yield (Beres et al., 2011).
-
-### Egg
-Newly laid eggs are crescent in shape, milky-white in color, and range between 1 mm to 1.25 mm in length and 0.33 to 0.42 mm wide (Ainslie, 1920).
-After oviposition, the egg lies freely within the lumen of the host stem or in a hollow created by the female WSS ovipositor (Fulbright et al., 2017).
-Laval development occurs rapidly with the recognizable shape becoming apparent at day three.
-By day five, the jaws turn brown as the eye spots begin to appear and darken.
-By day six or seven, the larva begins to hatch using a series of convulsive movements and emerges into the stem cavity, with full functionality achieved by the seventh day (Ainslie, 1920).
-
-### Larvae
-After hatching and prior to feeding, the larvae are transparent and colorless (Ainslie, 1920; Criddle, 1923).
-WSS larvae are distinguishable from other larvae feeding in wheat and grasses by their highly sclerotized head capsule and large body size (Nelson and Farstad, 1953).
-As the larvae chew their way through the parenchymous and vascular tissues of the host stem, the larvae begin to develop a yellow-green color.
-As the larvae progress through the stem, they excrete frass; the digested plant tissue as well as metabolic breakdown products from the WSS (Beres et al., 2011).
-If multiple eggs are laid within the same stem, the largest (usually the first egg laid and hatched) will typically remain within a few weeks due to the larvae cannibalizing one another (Criddle, 1923; Seamans et al., 1944; Buteler et al., 2015).
-It remains unknown whether the consumption of other eggs and larvae is the result of intentional cannibalism or the indirect result of indiscriminate feeding (Buteler et al., 2015).
-As the host plant physiologically matures, and the larvae reaches its fifth instar, the insect is cued to travel to the base of stem and prepare for obligatory diapause (Beres et al., 2011).
-The descending movement of larvae and development of the hibernaculum are triggered by the host plant senescence.
-As the stem walls becoming increasingly transparent, they allow increased amounts of visible and infrared light into the stem (Holmes, 1977; Beres et al., 2011).
-Once the larvae reach the base of the stem, they construct an interior V-shaped groove, making the stem weak enough to lodge due to gravity or wind (Criddle, 1923).
-The larvae then plug the remaining stub with frass to insulate the hibernaculum during diapause (Holmes and Peterson, 1960).
-The larval stage of development can be as short as 30 days before entering diapause.
-Obligatory diapause lasts a minimum of 90 days (Salt and Hollick, 1947).
-After the larvae are exposed to temperatures less than 10C for 90 days, diapause is complete, and pupation can begin.
-Larvae can reenter diapause if temperatures approach 35C (Salt and Hollick, 1947) or if conditions are abnormally dry (Holmes, 1978) shortly after termination of diapause.
-
-### Pupae
-Once diapause is complete, pupal development begins.
-Pupation lasts for 21 days, beginning in early- to mid-May (Criddle, 1923).
-Initially, the pupa is milky-white in color (Ainslie, 1920), gradually changing to bright white as the wings develop (Fulbright et al., 2017).
-Additionally, as development continues, characteristic brown and black eye spots appear on the surface of the pupae.
-Once pupation is complete, the adult sawfly will chew through either the frass plug or the side of the stub, and emerge ready for flight (Ainslie, 1920; Holmes and Peterson, 1960).
-
-## Damage Caused by Wheat Stem Sawfly
-Wheat stem sawfly larvae travel throughout the stem of the host plant, feeding on the parenchymal tissue and chewing through the nodes.
-As a result, vascular translocation of water and nutrients within the plant is impaired (Macedo et al., 2007) and causes a notable reduction yield and quality of the grain produced (Seamans et al., 1944; Holmes, 1977; Morrill et al., 1992).
-In addition, the weakening of the stem due to WSS causes lodging prior to harvest, exacerbated by larvae stem girdling.
-Stem lodging causes a reduction in stem harvest efficiently (Ainslie, 1920), with yield lost due to unrecoverable lodged stem (Beres et al., 2007).
-Additionally, the reduction in stubble residues decreases snow retention which increases soil erosion and reduces soil moisture content (Keren et al., 2015).
-
-## Management of Wheat Stem Sawfly
-### Chemical Control
-Because WSS are stem mining pests and spend most of their lives within the host plant stem, the use of traditional pest control methods such as insecticides have proven to be ineffective.
-The only life stage of WSS that takes place outside of the stem, the adult, is not targetable via chemical treatment due to adult emergence and oviposition taking place over many weeks (Knodel et al., 2010; Beres et al., 2011).
-WSS infestation chemical control methods were first tested as early as 1949 (Wallace, 1962).
-After using six different organochlorine and organophosphate insecticidal dust applied via a powder duster (Munro et al., 1949) the authors found “unsatisfactory results that did not control” WSS populations or limit infestations.
-In 1963, it was reported that a 73% control of WSS infestation could be achieved via granular heptachlor applied in the furrow with the wheat seed (Wallace, 1962).
-However, the findings of these tests were not able to be replicable in all scenarios of WSS infestation (Holmes et al., 1963).
-Heptachlor required light sawfly infestation – restricted to the lower internodes – to be effective.
-When infestation rates were high, and oviposition occurred in the upper internodes, the researchers saw little to no control.
-Similar results were found when testing with foliar insecticides, findings showing little differences in their effect when compared to controls (Blodgett et al., 1996).
-While the Montana Department of Agriculture has authorized the use of systemic organophosphate insecticide phorate (Thimet®) for use controlling WSS infestation in winter and spring wheat cultivars, it’s use is not considered environmentally friendly due to high toxicity to mammals, birds, and fish.
-Additionally, a preharvest interval of 85 days complicates harvest scheduling (Ginsberg, 2003; Mahajan et al., 2006), while the high toxicity may have potential adverse effects on beneficial parasitoid populations (Varella, 2016).
-
-### Cultural Control
-#### Host Plant Resistance
-Mechanisms like antixenosis, antibiosis, and tolerance give a plant the ability to resist insect pests (Horber, 1980; Wilson et al., 1987).
-Antixenosis allows the host plant to trigger resistance by inducing non-preference in herbivores when they attempt to feed or oviposit.
-Recent work done by Achhami et al.
-(2020) found that Barely cultivars that closely resemble wheat show the greatest frequency of WSS oviposition.
-These findings suggest that semiochemicals produced by these barley cultivars, such as (Z)-3-hexenyl acetate (Piesik et al., 2008; Buteler et al., 2009) and (Z)-β-ocimene (Buteler and Weaver, 2012), are likely similar to those produced by wheat (Achhami et al., 2020).
-Antibiosis causes the improper development of herbivores, impacting survivability and reproduction.
-Tolerance allows the host plant to withstand infestation and overcome injuries (Van Emden, 1991; Van Zoeren and Guedot, 2017).
-Much of the research done on WSS resistant wheat cultivars focuses on solid stem trait and pith expression (Platt and Farstad, 1949; Varella et al., 2016; Varella et al., 2017).
-Solid stem varieties are more resistant to infestation, reducing WSS oviposition and increasing egg mortality (Holmes and Peterson, 1960; Varella et al., 2016).
-The increased pith expression also disrupts larvae movement within the stem, leading to higher larval mortality due to reduced feeding as well as smaller and less reproductively successful adults (Wallace and McNeal, 1966).
-In addition, greater pith expression increases stem strength, reducing losses due to lodged stems (Delaney et al., 2010).
-Additional factors including late planting and shorter crop duration, have been suggested as potential factors complicating antixenosis, leading to increased WSS oviposition and infestation (Achhami et al., 2020).
-The solid-stemmed trait has been developed in numerous spring and some winter wheat cultivars (Beres et al., 2011).
-While the use of solid-stemmed cultivars is most often recommended for WSS management (Beres et al., 2011), hollow-stemmed cultivars were found to exhibit greater parasitism when compared to solid-stemmed varieties (Holmes et al., 1963).
-It was previously suggested by Rand et al.
-(2012)  that the thicker pith found in solid-stemmed cultivars makes sensing WSS larvae vibrations (Mankin et al., 2004) more difficult for female parasitoids.
-The inability to locate WSS hosts would then cause ineffective parasitism.
-
-#### Tillage
-Tillage is a potential control method for the WSS.
-Larvae are protected against both water loss and freezing temperatures by the soil surrounding the wheat stubble where the WSS overwinters.
-Tilling the soil exposes the wheat stubble to the soil surface, conditions which kill the hibernating larvae (Callenback and Hansmeier, 1944).
-Tillage of the WSS has been shown to be effective in decreasing the pest’s populations in both spring and winter wheat cultivars.
-Upwards of 90% mortality has been shown in North Dakota after fall tillage (Weiss et al., 1987).
-Many factors influence the effectiveness of tilling, including adequate exposure of wheat stubs on the soil surface and appropriate timing (Holmes and Farstad, 1956; Morrill et al., 1992; Weiss and Morrill, 1992).
-While the variability in effectiveness exists, some growers do consider it a crucial management tool (Blodgett et al., 1996; Lesieur et al., 2016).
-While effective in killing WSS larvae, there are negative impacts caused by this practice.
-Water loss, soil compaction, and erosion have all been shown to be dramatically increased by tilling activities (Phillips et al., 1980; Lal, 1991).
-In addition, tilling practices negatively impact beneficial populations of parasitoid wasps that target the WSS.
-WSS parasitoids overwinter higher in the stem compared to at the base like WSS (Morrill et al., 1998; Runyon et al., 2002a).
-By displacing soil when tilling, the upper portion of the plant is buried under the soil, preventing the parasitoid wasps from emerging in the spring (Runyon et al., 2002a; Weaver et al., 2004).
-
-#### Plant Density and Spacing
-Historically, it has been observed that wheat fields with wider row spacing (12 inches) have greater WSS cutting when compared to fields with reduced spacing (3 to 6 inches).
-Additionally, fields planted at greater seed density with solid stem cultivars exhibit lower WSS cutting compared to reduced seed density (Luginbill and McNeal, 1958).
-The reason for this relationship between decreased density and increased sawfly infestation is due to the moisture availability and height of the wheat stems.
-At higher densities, each stem has access to less water, in turn leading to the stem to develop a thicker pith – more solid stemmed.
-As discussed above, there is a negative linear relationship between an increase in the solid-stem characteristic and infestation (Kemp, 1934).
-
-#### Crop Rotation
-Continually planting wheat year after year creates ideal conditions for the WSS (Callenbach and Hansmeier, 1945).
-Rotating wheat with more resistant wheat varieties as well as with oats, or soybeans (crops that are less or not susceptible to WSS infestation) is recommended to decrease the year after year WSS infestation rates (Lin and Chen, 2014).
-The notable downside to utilizing crop rotation in a matter that is effective is the economic downside.
-Alternate-year, summer fallow production systems are typical for wheat growing practices across much of the northern great plains of the United States (Weiss and Morrill, 1992; Troeh et al., 1999).
-
-#### Trap Crops
-Trap cropping utilized non-economically important plantings to help reduce populations or block movement of pests into valuable crop areas (Hokkanen and Jokioinen, 1991).
-Effective trap crops of the WSS should have appealing characteristics to the WSS to attract the pest to the trap planting, and away from the wheat (Farstad, 1940).
-Historically, planted trap strips were used by attracting sawflies and destroying the strip at the end of the season, eliminating the larvae (Morrill et al., 1992).
-However, utilizing trap plantings of highly attractive wheat varieties has been shown to hold limited viability, as growers are not fond of destroying planted grain at the end of the growing season (Goosey, 1999).
-Alternative utilization of WSS traps entail planting and/or maintaining strips of smooth brome grass (Bromus inermis Leyss.) or other WSS host grasses (Farstad, 1940).
-Smooth brome in particular holds potential as it is a poor host to WSS larvae (Bhandari, 2020), yet WSS adults continue to oviposit the stems.
-This relationship of greater oviposition with increased mortality may lead to decreases in overall WSS populations.
-
-#### Delayed Planting & Swathing
-Wheat planted later in the growing season avoids much of the damage caused by the WSS as the wheat reaches stem elongation after many of the adult sawflies have already laid their eggs (Jacobson and Farstad, 1952).
-The problem with delayed planting is that the practice is known to make less efficient use the earl season moisture in the soil (Morrill and Kushnak, 1996).
-Growers may need to weigh the tradeoffs in delayed planting to reduce losses in areas with a history of severe infestation.
-Swathing, also known as windrowing, is the practice of cutting the wheat to accelerate the drying process and decrease the losses caused by WSS cutting (Holmes and Peterson, 1965; Carlson, 2019; McCullough et al., 2020).
-By cutting the wheat and letting it dry prior to sawfly cutting, swathing works by reducing losses due to lodging (Ainslie, 1920).
-
-#### Biological Control Methods
-The use of biocontrol in limiting WSS populations has been well studied with variable results.
-There are currently three types of organisms that have been reported as biocontrol agents of the WSS in wheat fields: Soil fungal pathogens, the predatory beetle Phyllobaenus dubius, and two parasitic Hymenoptera.
-
-#### Soil Fungal Pathogens
-Fungal pathogens are found throughout both the plant and animal kingdoms.
-Several fungi have been shown to negatively impact WSS larvae infestation within wheat stems.
-While highly lethal to sawfly larvae, F.
-graminearum, F.
-acuminatum, F.
-equiseti, and F.
-avenaceum also cause crown rot in the host plant.
-The trade-off between WSS control and crown rot may limit Fusarium spp.
-usefulness to WSS management.
-(Wenda-Piesik et al., 2009).
-Findings by Tangtrakulwanich et al.
-(2014) found that the fungi B.
-bassiana and M.
-brunneum effectively reduced damage caused by WSS.
-Both fungi kill insects by infecting via spore contact or by insect consumption of treated plant tissue (Portman et al., 2018).
-
-#### Predatory Beetle
-Phyllobaenus dubius (Wolcott) has been observed emerging from wheat stubble and within dissected wheat stems in Montana (Morrill et al., 2001).
-Further research is required on this coleopteran to determine its prospects going forward as an effective WSS management tool.
-
-#### Parasitoids of the Wheat Stem Sawfly
-Two parasitoid wasps are known target the wheat stem sawfly in wheat fields.
-The congeneric parasitoids, Bracon cephi (Gahan) and B.
-lissogaster Muesebeck, are the most effective agents of WSS biological control (Runyon et al., 2001).
-B.
-cephi and B.
-lissogaster, as distinguished by Runyon et al.
-(2001), were initially limited in their ability to parasitize WSS larvae within wheat stems.
-The wheat stem sawfly, originally known as the western grass-stem sawfly, was originally only found within native grasses throughout North America, existing in alongside B.
-cephi and B.
-lissogaster (Ainslie, 1920; Cárcamo et al., 2012).
-However, as agriculture increased and native grasses were replaced by large monoculture landscapes, the WSS moved into wheat (Farstad, 1940; Davis et al., 1955; Holmes, 1978) while the parasitoids were slower to adapt (Morrill et al., 1998).
-
-## Biology of Bracon cephi and Bracon lissogaster
-### Adult
-Both B.
-cephi and B.
-lissogaster are protelean host-specific parasitoids with two generations per year (bivoltine) (Nelson and Farstad, 1953; Somsen and Luginbill, 1956; Holmes et al., 1963).
-They are ectoparasitoids and idiobionts, living on the integument of their hosts and immediately paralyzing their hosts in location and stage it was when it was attacked (Nelson and Farstad, 1953; Runyon et al., 2002b; Weaver et al., 2004).
-Male parasitoids are approximately 3.5 mm in length, while females are 4.1 mm.
-They are dark red, with black antennae, eyes, ovipositor and ocelli (Gahan, 1918).
-B.
-lissogaster and B.
-cephi are distinguishable by their head color and unique metasoma surface texture (Runyon et al., 2001).
-Adults emerge after completing a pupal stage within the host stem by chewing a hole through the stem wall.
-In Montana, the first generations are present from late June to late July, while the second generations are active in August and September (Somsen and Luginbill, 1956).
-Adult female B.
-cephi and B.
-lissogaster will locate a host stem using a combination of olfactory and volatile cues.
-The wasps will then locate a host sawfly larva using acoustic cues (Mankin et al., 2004) and use their ovipositor first paralyze, then deposit 1-4 eggs next to the paralyzed larva.
-
-### Egg
-B.
-cephi and B.
-lissogaster eggs are on average 0.86 mm in length, exhibiting an elongated shape and pale-yellow color.
-Females lay one to three eggs on the WSS larval integument or adjacent to the larva.
-The eggs develop for one to three days before hatching (Nelson and Farstad, 1953; Somsen and Luginbill, 1956).
-
-### Larva
-Parasitoid larvae are 1mm in length, with a translucent hue that turns increasingly brown as the larva progresses through five instar stages (Nelson and Farstad, 1953).
-The newly hatched larvae have no appendages, save for antennae, and they exhibit highly sclerotized mouthparts and well-developed mandibles.
-After hatching, the B.
-lissogaster larva attach themselves to the integument of the host larva and immediately begin feeding (Somsen and Luginbill, 1956).
-When feeding is complete, little remains of the host but the integument.
-Following feeding, the larva will spin a lightly brown colored cocoon over the course of several days and attach to the inside wall of the host plant stem.
-The larval stage takes an average of ten days between initial hatch and cocoon completion.
-Because there are two generations per year, the early summer offspring will hatch and emerge from the host stem.
-Offspring laid as a part of the second generation will overwinter within the cocoons and emerge in the spring.
-(Nelson and Farstad, 1953)
-
-### WSS and Parasitoid Interactions with Smooth Brome
-Bhandari (2020) has shown that oviposition occurs with similar frequency in wheat as it does in smooth brome, however, we see much higher rates of larval mortality within smooth brome stems.
-This has led researchers to hypothesize about the feasibility of using smooth brome as a trap cop (Perez-Mendoza et al., 2006; Lesieur et al., 2016).
-While feasible as a feed crop, as it holds considerably less economic value compared to wheat or other high value crops (Criddle, 1922).
-Smooth brome may be usable in trap strips along the edge of wheat fields (Hokkanen and Jokioinen, 1991).
-While a poor host for the wheat stem sawfly, surviving immature larvae found within brome stems can be parasitized by B.
-cephi and B.
-lissogaster.
-This combination of factors may indicate that smooth brome can function as a WSS sink and a parasitoid source, with very few WSS emerging from smooth brome patches, while also supporting populations of parasitoids that can then parasitize WSS in adjacent wheat fields.
-
-## Plant Semiochemicals
-Plants interact with other organisms through a complex system of chemical signals known as Semiochemicals.
-Semiochemicals are compounds or mixtures of compounds that are released from one organism that evoke a physiological or behavior response in another organism (Agelopoulos et al., 1999).
-Semiochemicals play a large role in both insect-insect and plant-insect interactions and are essential in insects locating feedings sites, mating sites, egg-laying sites, and/or refugia (Prokopy et al., 1984).
-While it is known that these chemicals play an important role in communication between plants, insects, and other organisms, there is still debate about the function of many of them (Reddy and Guerrero, 2004; Voelckel and Jander, 2018).
-
-### Semiochemicals in Plant-Insect Systems
-	Within plant-insect systems, plants respond to insect herbivory by synthesizing and releasing volatile compounds and providing important signals to biocontrol agents to localize potential host insects (Sharma et al., 2019).
-In addition, semiochemicals provide localized signatures that allow insects to detect food sources, find mates, and locate preferential oviposition sites (Norin, 2007).
-Plants typically produce semiochemicals when defending themselves from herbivorous arthropods (Guerrieri et al., 1999) or pathogen attacks (Smart et al., 2014 Jan 1).
-These responses can be grouped into two categories, direct and indirect responses.
-Direct responses are compounds that the plant produces that have a direct effect on the herbivore or pathogen (Metcalf, 1998).
-Indirect responses are the compounds produced that have no impact directly on the pest, but instead may attract parasitoids, predators, or other biological control agents (Mayland et al., 2000; Colazza et al., 2004).
-
-# Limitations And Key Assumptions
-## Field Study
-i.	Time limits the number field sites that we can collect samples from throughout the summer.  
-a.	Study area limited to western Montana – three regions – Big Sandy, Moccasin, and Amsterdam.  
-ii.	Our study will only take place over the course of 2-3 summers, thus, we are unable to make broader claims surrounding long term patterns of WSS and parasitoid interactions.   
-Green House Samples and Volatile Collection
-iii.	We will assume that plants grown in the green house will be accurate and reasonable biological model representation of plants present in the field.  
-iv.	Volatiles will be collected from a maximum of 12 plants a day due to constraints with the collection system.
-
-# Contributions to Knowledge of Potential Outcomes
-## Potential outcomes
-### Volatiles
-We hope to observe a discernable difference in the volatile quantities produced by both infested smooth brome and winter wheat.
-
-# CHAPTERS
-## METHODS
-### Central and Northern Montana Field Sites
-#### Field Selection & Data Collection
-To assess C cinctus infestation, larval mortality, and parasitoid prevalence within B.
-inermis and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production.
+To assess *C.
+cinctus* infestation, larval mortality, and parasitoid prevalence within *B.
+inermis* and adjacent wheat fields, sites were chosen that had the best chance of accurately representing conditions throughout the Golden Triangle, a region of north-central Montana known for its dryland grain production.
 Samples were collected in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA.
-B.
-inermis sites were set up as polygons along the edge of adjoining wheat fields measuring approximately 100 meters in area.
+*B.
+inermis* sites were set up as polygons along the edge of adjoining wheat fields measuring approximately 100 meters in area.
 Four collection squares of 1ft x 1ft were randomly partitioned within each polygon during both collection events each year.
@@
 
-#### Statistical Analysis
-Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions.
-‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs awere used as the response variable to assess the differences in parameters among the sampling sites.
-ANOVA was used to caculate P-values using F-distribution to understand overall significant differences in mean C.
-cinctus infestation and parasitism among polygons.
-To assess the profile of C.
-cintus infestation, paraisitms, and larval presence by node, a linear model was fit for each sampling site.
-Further assessment was done on the resonpses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.
+### Controlled *C. cinctus* infestation of *B. inermis*
 
-## Controlled C. cinctus infestation of B. inermis
-### Plant Selection and Field Cages
-Assessment of C.
-cinctus infestation and mortality within B.
-inermis were assessed using a 34 x 60 ft plot at the Arthur H.
+#### Plant Selection and Field Cages
+
+Assessment of *C.
+cinctus* infestation and mortality within *B.
+inermis* were assessed using a 34 x 60 ft plot at the Arthur H.
 Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-The cage structure was built using 1-inch PVC piping with the netting made using 530 Amber Lumite Screen (BioQuip Products, LLC).
+The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
 Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.
@@
 
-### Insects
+#### Insects
 
-Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of C.
-cinctus infestation and cutting the year prior.
-Cut stubble, which contained C.
-cinctus larvae in diapause, were kept refrigerated between -2°C and 3°C for >100 days as required to complete obligatory larval diapause.
+Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of *C.
+cinctus* infestation and cutting the year prior.
+Cut stubble, which contained *C.
+cinctus* larvae in diapause, were kept refrigerated between -2°C and 3°C for \>100 days as required to complete obligatory larval diapause.
 As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA).
-Stubs were added to cages once B.
-inermis stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).
+Stubs were added to cages once *B.
+inermis* stems reached six inches tall, with sawfly quantity treatments as follows: high (600 stubs), low (200 stubs), and control (0 stubs).
 
-### Data Collection
+#### Data Collection
+
 Stems were collected from each cage site in late August.
@@
 
-### Statistical Analysis
-Statistical analyses were conducted in R (R Core Team 2022) using RStudio (version 4.2.1) using linear modeling after log transformation to improve the normality and constant variance assumptions.
-‘Site’ and ‘polygon’ were used as fixed effects (treatments) while the mean number of infested stems, parasitized stems, dead larvae, alive larvae, dead eggs, alive eggs were used as the response variable to assess the differences in parameters among the sampling sites.
-ANOVA was used to calculate P-values using F-distribution to understand overall significant differences in mean C.
-cinctus infestation and parasitism among polygons.
-To assess the profile of C.
-cinctus infestation, parasitism, and larval presence by node, a linear model was fit for each sampling site.
-Further assessment was done on the responses between sampling sites using a Tukey test using the ‘TukeyHSD’ function.
+## Greenhouse Experiments
 
-## Organic Volatile Collection from C. cinctus Infested B. inermis and Winter Wheat
-### Insects
+### Organic Volatile Collection from *C. cinctus* Infested *B. inermis* and Winter Wheat
+
+#### Insects
+
 Wheat stubble collected the year prior to each experiment were used to rear adult WSS.
-Wheat stubble was collected from Big Sandy (48.17754 N, -110.11406 E) and Three Forks (46.020216 N, -111.586582 E), MT, USA.
+Wheat stubble was collected from Big Sandy (48.17754°N, -110.11406°E) and Three Forks (46.020216°N, -111.586582°E), MT, USA.
 Stubble was collected from fields that experienced high levels of WSS infestation and cutting the year prior.
-Cut stubble, which contained WSS larvae in diapause, were kept in a refrigerator between 0-4 C for >100 days as required to complete obligatory larval diapause.
-After completing diapause, immature WSS were kept at room temperature between 22-27 C for 4-5 weeks in sealed 70 cm x 35 cm x 20 cm plastic Tupperware bins to allow for pupation and emergence.
+Cut stubble, which contained WSS larvae in diapause, were kept in a refrigerator between 0-4 °C for \>100 days as required to complete obligatory larval diapause.
+After completing diapause, immature WSS were kept at room temperature between 22-27 °C for 4-5 weeks in sealed 70 cm x 35 cm x 20 cm plastic Tupperware$^\circledR$ bins to allow for pupation and emergence.
 Once emergence began, adults were collected daily and sorted in 2-liter glass mason jars sealed with filter paper to allow for airflow.
-Jars containing adults were refrigerated at 6-10C until they were needed for experiments.
+Jars containing adults were refrigerated at 6-10°C until they were needed for experiments.
 Unused adults were discarded after 48 hours of refrigeration.
 
-### Plants
-Smooth brome was grown from transplanted plants harvested from the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W).
-Immediately following collection, plants were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc, Tangent, OR) and moved into the greenhouse.
-Wheat was grown using ‘Yellowstone’ winter wheat cultivar seed stock, started in 1 x 8 inch conical pots.
+#### Plants
+
+Smooth brome was grown from transplanted stems harvested from the Arthur H.
+Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
+Immediately following collection, stems were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc$^\circledR$, Tangent, OR).
+Winter wheat plants were grown using Yellowstone Winter Wheat cultivar seed stock, started in 1 x 8 inch conical pots.
 The soil used consisted of MSU Plant Growth Center soil mix (equal parts sterilized Bozeman silt loam soil and washed concrete sand with Canadian sphagnum peat moss incorporated) and Sunshine Mix 1 (Canadian sphagnum peat moss, vermiculite, perlite, and Dolomite lime) in a 1:1 ratio.
-Fertilization was undergone once weekly using Jack’s Professional® Water-Soluble Fertilizer (20-20-20) (J.R.
+Fertilization was undergone once weekly using Jack’s Professional$^\circledR$ Water-Soluble Fertilizer (20-20-20) (J.R.
 Peters Inc., Allentown, PA, USA).
-The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22  2C and 20 – 40 % RH.
+The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22 $\pm$ 2°C and 20 – 40 % RH.
 After the plants reached Zadoks 12, they were moved into vernalization storage for 6 weeks to experience the length of dormancy needed for successful flower development.
-After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc conical pots.
+After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc$^\circledR$ conical pots.
 
-### Volatile Collection
+#### Sawfly Infestation
+
+Winter wheat plants were deemed ready for infestation once they reached Zadoks 49.
+Both smooth brome and winter wheat plants were infested by placing a transparent plastic cylinder (60 x 3.8 cm) over the main stem of the plant.
+A 55-60 cm bamboo stick was added within each tube to help prevent toppling of the cylinder.
+Each cage had three holes, as well as an open top, covered with fine mesh cloth to allow for air circulation.
+Four females and two male WSS, newly emerged, were introduced into each tube via a small 1 cm opening.
+After which, the hole was plugged using a cotton ball covered in mesh cloth.
+Damp soil was mounded around the base of each tube to prevent insect escapees.
+Adult sawflies were left within infestation tubes for three days, after which the tubes were removed, and any surviving insects, as well as cadavers, were removed.
+After infestation, the plants were returned to normal water and fertilization schedules.
+
+Infestation was undertaken three days a week, for five weeks.
+Each replicate included 6 ‘Yellowstone’ winter wheat plants at Zadoks 49 and 6 smooth brome plants exhibiting three nodes.
+For each replicate, three wheat and three smooth brome plants were infested, with three of each left uninfested.
+Infestation tubes were placed around all twelve plants to maintain control.
+
+#### Volatile Collection
+
 Volatile compounds were collected from intact and healthy smooth brome and Yellowstone winter wheat.
@@
 
-Volatile collection and successive GC-MS analysis were conducted following methods described in Weaver et al.
-(2009).
-Glass volatile collection traps were eluted using 200 µL of dichloromethane into 2 mL screw top vials (Agilent Technologies, Inc., Santa Clara, CA), followed by a slow release of nitrogen to further clear the volatile trap.
-10 µL of nonyl acetate solution was added to the eluted solution as internal standard to quantify compounds.
-Samples were then subject to gas chromatography (GC) on a J&W HP – 5ms; 30m x 0.25 µm film thickness column (Agilent Technologies, Inc., Santa Clara, CA).
-The GC unit (Agilent 6890; Agilent Technologies, Santa Clara, CA) was paired with a mass spectrometer (MS, Agilent 5973 instrument).
-Samples were injected into the column in pulse splitless mode, with the initial pressure of 82.7 kPa/m.
+#### Volatile Elution
 
-## PARASITOIDS
+Volatile compounds were eluted from the glass traps using 200 $\mu$L aliquot dichloromethane.
+Trapped volatiles were eluted slowly by adding dichloromethane and further clearing by using a slow release of nitrogen gas.
+The eluted samples were collected in a glass insert held within a 1.5-mL crimped top glass vial.
+10 $\mu$L of nonyl acetate were then added to the eluted samples as internal standard to quantify compound concentrations.
+The samples were processed using gas chromatography (GC) on a HP – 5MS; 30m x 0.25mm, 0.25 $\mu$m film thickness column; (J and W Scientific, Folsom, CA).
+In order to analyze, the GC instrument (Agilent 6890; Agilent Technologies, Santa Clara, CA) was coupled to a mass spectrometer (MS, Agilent 5973 instrument).
+Each sample was injected using the automated system in pulse splitless mode, with the initial pressure set to 82.7 kPa per minute.
+The inlet temperature of the GC was set to 250°C while the column temperature was 50°C for 4 minutes.
+The temperature was set to increase at a rate of 5°C per minute until it reached 160°C, after which it increased to a rate of 25°C per minute until it reached 280°C.
+The temperature of the transfer line leading to the mass selective detector (MSD) was set at 300°C.
+Helium was used as a carrier gas to maintain the flow rate of samples within the column at 1.2 mL/min.
+The MSD was set in ‘SCAN’ mode running from 50 – 300 m/z.
+
+#### Volatile Analysis
+
+The compounds collected from both the smooth brome and wheat samples were identified by comparing mass spectra and retention times using the National Institute of Standard and Technology (NIST) library.
+Peaks and volatiles were analyzed using MSD ChemStation (Agilent Technologies, Inc., Santa Clara, CA).
+The data obtained from the mass-spectrometer was transferred to a Microsoft® Excel spreadsheet where the compounds were quantified by comparison with the internal standard peak area.
+The amount obtained for each compound was then transformed to the amount produced in six hours.
+The true quantity of compounds was calculated relative to the internal standard.
+
+### Behavioral Assessment of Parasitoids in Response to *B. inermis* Organic Volatile Compounds
+
+#### Insects
+
 Parasitoids B.
 cephi and B.
-lissogaster were collected from wheat fields in Three Forks, MT (46.020216 N, -111.586582 E).
-Wheat stem residue from fields that experienced high levels of C.
-cinctus infestation were collected and stored inside 170-liter plastic barrel liners.
-Filled liners were stored in a cold storage room at 4C until mid or late spring of the following year.
+lissogaster were collected from wheat fields in Three Forks, MT (46.020216° N, -111.586582° E).
+Wheat stem residue from fields that experienced high levels of *C.
+cinctus* infestation were collected and stored inside 170-liter plastic barrel liners.
+Filled liners were stored in a cold storage room at 4°C until mid or late spring of the following year.
 After the storage time exceeded the mandatory duration needed for obligatory diapause, the plastic barrel liner was removed from cold storage and the residue was placed inside 170 L trash barrels at room temperature and watered to encourage emergence.
-Collected parasitoids were kept at 6-10C until used.
+Collected parasitoids were kept at 6-10°C until used.
 
-### Plants
-Smooth brome was grown from transplanted stems harvested from the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-Immediately following collection, stems were transplanted into 2.7 x 14 inch lightweight deepot cells (Stuewe & Sons Inc, Tangent, OR).
-Winter wheat plants were grown using Yellowstone Winter Wheat cultivar seed stock, started in 1 x 8 inch conical pots.
-The soil used consisted of MSU Plant Growth Center soil mix (equal parts sterilized Bozeman silt loam soil and washed concrete sand with Canadian sphagnum peat moss incorporated) and Sunshine Mix 1 (Canadian sphagnum peat moss, vermiculite, perlite, and Dolomite lime) in a 1:1 ratio.
-Fertilization was undergone once weekly using Jack’s Professional® Water-Soluble Fertilizer (20-20-20) (J.R.
-Peters Inc., Allentown, PA, USA).
-The plants were kept in a greenhouse in the Montana State University Plant Growth Center using supplemental light from model MVR1000/C/U GE Multi-Vapor Lamps (GE Lighting, General Electric Company, Cleveland, OH) under photoperiod of 16:8 (L:D) h at 22  2C and 20 – 40 % RH.
-After the plants reached Zadoks 12, they were moved into vernalization storage for 6 weeks to experience the length of dormancy needed for successful flower development.
-After the six-week period, wheat stems were transplanted into the larger, 2.7 x 14-inch Stuewe & Sons, Inc conical pots.
+#### Electroantennography
 
-## Behavioral Assessment of Parasitoids in Response to B. inermis Organic Volatile Compounds
-### Insects
-### Electroantennography
-### Y- Tube Olfactometer Bioassay
+Specimens were prepared by excising the head from the body using fine micro-dissection scissors.
+For each specimen, a micro-capillarity tube was pulled in half using a pipette puller (Model PP-830, Narishige Scientific Instrument Lab, Tokyo, Japan).
+Each half was cut to the appropriate length and angle, then filled with saline solution.
+The parasitoid head was mounted on the broken end of one of the capillary tubes, then the tube was place on one of the conducting electrode wires connected to a micromanipulator, type INR-5 (Narishige MN-151).
+The second half of the capillary tube was place on the other electrode wire, and one insect antennae was positioned to touch the empty capillary tube, thus connecting the circuit.
+The amplifier transferred the information to a computer interface with an EAG version 2.7 software (Syntech NL 2001, Hilversum, Netherlands).
+This apparatus and method has been consistently used previously in assessing insect volatile response [@Agelopoulos1999].
+
+A consistent stream of purified, humidified air (20 mL sec-1) was delivered onto the antennae.
+The test compound was delivered via airborne stimulus released from a glass pipette with filter paper treated with 10 $\mu$L of the test compound.
+The pipette was connected to a Stimulus Controller CS-55 (Syntech NL 2010, Hilversum, The Netherlands).
+The equipment was programmed to deliver a 0.4 second puff of air through a Teflon tube and through the glass pipette.
+The electrical response of the antenna to the test compound was detected by the recording electrode and transferred to the amplifier, increasing the signal and delivering the information to the computer.
+Figure 2: Electroantennogram connected to GCMS.
+Useful for connecting insect electrochemical response to host plant produced volatile compounds.
+
+#### Y- Tube Olfactometer Bioassay
+
 Y-tube bioassays were conducted within the chemical ecology and insect behavior laboratory at the Plant Growth Center at Montana State University.
-Bioassays were undergone to determine behavioral choice preference response of individual female WSS to an airstream containing volatiles originating from smooth brome and a blank control.
+Bioassays were undergone to determine behavioral choice preference response of individual female WSS to an air stream containing volatiles originating from smooth brome and a blank control.
 Bioassays were conducted using plants at growth stage Zadoks 49.
-The Y-tube system (Analytical Rsearch Systems, Micanopy, FL) used was similar as described in a previous study (Daisy et al., 2002) (Figure 1).
+The Y-tube system (Analytical Research Systems, Micanopy, FL) used was similar as described in a previous study [@Daisy2002] (Figure 1).
 A single plant was enclosed in a volatile collection chamber (VCC) (100 mm diameter, 254 mm length).
@@
 <!-- Figure 1 y tube diagram -->
+
 ```{=html}
@@
 </html>
-
 ```
-
-Figure 1: Y-tube olfactory bioassay as constructed by Daisy et al. (2002). Used to measure insect behavior in the presence of volatile stimulus.
-
 The Y-tube was made of bifurcated Corning® glass, measuring in at 28-mm diameter x 300 m length.
@@
 A 28-cm wire was used to facilitate parasitoid movement from the unbranched stem of the glass Y-tube to the test junction between the two upper arms.
-The order of individual test insects from a pool of naïve females was randomly assigned for each bioassay, and test plants were chosen randomly from the greenhouse culture before each experiment.
+The order of individual test insects from a pool of naive females was randomly assigned for each bioassay, and test plants were chosen randomly from the greenhouse culture before each experiment.
 All glassware was cleaned using dish soap and warm water, followed by solvent rinses with acetone and hexane.
-Cleaned and rinsed glassware was baked at 110C in a glassware oven for 24 hours before use.
+Cleaned and rinsed glassware was baked at 110°C in a glassware oven for 24 hours before use.
 New plants were used for each experiment.
-	Each individual parasitoid was allowed three minutes to respond to either airstream source.
+
+Each individual parasitoid was allowed three minutes to respond to either air stream source.
 Once a positive response was determined to either source, the trail was ended.
@@
 
-## Volatile Collection
-### Sawfly Infestation
-Winter wheat plants were deemed ready for infestation once they reached Zadoks 49.
-Both smooth brome and winter wheat plants were infested by placing a transparent plastic cylinder (60 x 3.8 cm) over the main stem of the plant.
-A 55-60 cm bamboo stick was added within each tube to help prevent toppling of the cylinder.
-Each cage had three holes, as well as an open top, covered with fine mesh cloth to allow for air circulation.
-Four females and two male WSS, newly emerged, were introduced into each tube via a small 1 cm opening.
-After which, the hole was plugged using a cotton ball covered in mesh cloth.
-Damp soil was mounded around the base of each tube to prevent insect escapees.
-Adult sawflies were left within infestation tubes for three days, after which the tubes were removed, and any surviving insects, as well as cadavers, were removed.
-After infestation, the plants were returned to normal water and fertilization schedules.
-
-Infestation was undertaken three days a week, for five weeks.
-Each replicate included 6 ‘Yellowstone’ winter wheat plants at Zadoks 49 and 6 smooth brome plants exhibiting three nodes.
-For each replicate, three wheat and three smooth brome plants were infested, with three of each left uninfested.
-Infestation tubes were placed around all twelve plants to maintain control.
-
-### Volatile Compound Collection
-Volatile compounds were collected from intact and healthy smooth brome and ‘Yellowstone’ winter wheat.
-Collection was undergone on plants at Zadoks 49, for a period of six hours from 10:00 to 16:00.
-Collections were undergone three times a week, with twelve plants collected from each time.
-Each trail consisted of six smooth brome plants and six winter wheat, with three of each being infested, while the other three were a control.
-
-Volatiles were collected from twelve plants per day using glass VCC’s that were 80 cm long and 4 cm in diameter.
-Volatile collection and successive GC-MS analysis were conducted following methods described in Weaver et al.
-(2009).
-The top end of each VCC funneled into a threaded port that was 10 cm in diameter.
-Threaded onto the top of each VCC was an eight-port manifold adapter.
-Two glass filters (traps) (6.35 mm diameter x 76 mm length; Analytical Research Systems, Gainesville, FL) with 30 mg of Super-Q absorbent (Alltech Associates, Deerfield, IL) were inserted into two of the adapter ports.
-The remaining six collection ports were sealed to prevent ambient air exchange and pressure loss.
-One of the traps was used to collected extraneous volatiles from the initial experiment set up for the first ten minutes, whereas the other trap was used to collect volatile compounds emanating from plants beginning ten minutes after the system was turned on.
-A vacuum pump was used to maintain air pressure and flow rate (humidified air at 1.0 liter/min) within the collection system.
-A Teflon guillotine was used at the base of each plant to tightly seal the base of the system and prevent external air from entering.
-Each plant stem was wrapped with a small piece of cotton prior to the application of the guillotine to further seal the base of the collection system and prevent contamination by ambient and soil derived volatiles.
-
-### Volatile Elution
-Volatile compounds were eluted from the glass traps using 200 L aliquot dichloromethane.
-Trapped volatiles were eluted slowly by adding dichloromethane and further clearing by using a slow release of nitrogen gas.
-The eluted samples were collected in a glass insert held within a 1.5-mL crimped top glass vial.
-10 L of nonyl acetate were then added to the eluted samples as internal standard to quantify compound concentrations.
-The samples were processed using gas chromatography (GC) on a HP – 5MS; 30m x 0.25mm, 0.25 m film thickness column; (J and W Scientific, Folsom, CA).
-In order to analyze, the GC instrument (Agilent 6890; Agilent Technologies, Santa Clara, CA) was coupled to a mass spectrometer (MS, Agilent 5973 instrument).
-Each sample was injected using the automated system in pulse splitless mode, with the initial pressure set to 82.7 kPa per minute.
-The inlet temperature of the GC was set to 250C while the column temperature was 50C for 4 minutes.
-The temperature was set to increase at a rate of 5C per minute until it reached 160C, after which it increased to a rate of 25C per minute until it reached 280C.
-The temperature of the transfer line leading to the mass selective detector (MSD) was set at 300C.
-Helium was used as a carrier gas to maintain the flow rate of samples within the column at 1.2 mL/min.
-The MSD was set in ‘SCAN’ mode running from 50 – 300 m/z.
-
-### Volatile Analysis
-The compounds collected from both the smooth brome and wheat samples were identified by comparing mass spectra and retention times using the National Institute of Standard and Technology (NIST) library.
-Peaks and volatiles were analyzed using MSD ChemStation (Agilent Technologies, Inc., Santa Clara, CA).
-The data obtained from the mass-spectrometer was transferred to a Microsoft® Excel spreadsheet where the compounds were quantified by comparison with the internal standard peak area.
-The amount obtained for each compound was then transformed to the amount produced in six hours.
-The true quantity of compounds was calculated relative to the internal standard.
-
-### Electroantennogram
-Specimens were prepared by excising the head from the body using fine micro-dissection scissors.
-For each specimen, a micro-capillarity tube was pulled in half using a pipette puller (Model PP-830, Narishige Scientific Instrument Lab, Tokyo, Japan).
-Each half was cut to the appropriate length and angle, then filled with saline solution.
-The parasitoid head was mounted on the broken end of one of the capillary tubes, then the tube was place on one of the conducting electrode wires connected to a micromanipulator, type INR-5 (Narishige MN-151).
-The second half of the capillary tube was place on the other electrode wire, and one insect antennae was positioned to touch the empty capillary tube, thus connecting the circuit.
-The amplifier transferred the information to a computer interface with an EAG version 2.7 software (Syntech NL 2001, Hilversum, Netherlands).
-This apparatus and method has been consistently used previously in assessing insect volatile response (Agelopoulos et al., 1999).
-
-A consistent stream of purified, humidified air (20 mL sec-1) was delivered onto the antennae.
-The test compound was delivered via airborne stimulus released from a glass pipette with filter paper treated with 10 L of the test compound.
-The pipette was connected to a Stimulus Controller CS-55 (Syntech NL 2010, Hilversum, The Netherlands).
-The equipment was programmed to deliver a 0.4 second puff of air through a Teflon tube and through the glass pipette.
-The electrical response of the antenna to the test compound was detected by the recording electrode and transferred to the amplifier, increasing the signal and delivering the information to the computer.
-Figure 2: Electroantennogram connected to GCMS. Useful for connecting insect electrochemical response to host plant produced volatile compounds.
-List of compounds used.
-
-## Data Analysis
-### Field Survey
-### Greenhouse Experiments
-
-#### Comparative Analysis of Collection of Volatiles
-The concentration of volatile compounds gathered was determined as nanogram per gram per hour (ng-1g1hr1), accounting for plant biomass adjustments.
-Data were analyzed using R Studio version 2023.12.0+369 (R Core team 2023).
-Data were first transformed via centered-log ratio (CLR) transformation, a common technique for compositional data  (Brückner and Heethoff, 2017).
-The data underwent both Permutational Analysis of Variance (PERMANOVA) and Permutational Analysis of Multivariate Dispersions (PERMDISP).
-PERMANOVA was used to determine differences among identified plant volatile compounds between the two plant types (Yellowstone winter wheat, B.
-inermis) at two infestation levels (Infested, control).
-PERMDISP was used to assess the variance of volatile quantities within our different groups.
-
-#### EAG-FID
-
-#### Organic Volatile Collections
-##### Centered Log-Ratio Transformation
-The data was transformed using a centered log-ratio transformation (CLR).
-This method is commonly used for multivariate compositional data (Brückner and Heethoff, 2017).
-CLR works by taking the natural logarithm of the ratios between the values of individual components and their geometric mean across samples.
-This helps to mitigate spurious correlations and allows the application of standard statistical techniques to our compositional data.
-
-##### PERMANOVA
-
-# Citations
+#

### 2024-05-13 qmd/projects/msu/locoweed/gh-locoweed-report.qmd
commit 5/13 2024

--- 2024-03-27 qmd/projects/msu-masters/locoweed/gh-locoweed-report.qmd
@@
 # Introduction
+
 Astragalus and Oxytropix species, part of the Legume Family, are found in terrestrial ecoysstems globally, including the western regions of North America.
@@
 # Methodology
+
 ## Data Collection
@@
 ### Greenhouse Volatile Collection
+
 Samples of each plant were first sent for analysis to mystery lab to see if endophyte was present or not.
@@
 
-<div class="box">
+::: box
 <!-- plant pictures -->
+
 ```{r plant pictures, echo = F, eval = F}
@@
 ```
-</div>
+:::
 
@@
 # RESULTS
+
 ## Visual Comparisons
+
 <!-- LOAD DATA -->
+
 ```{r gh RAW 22, echo = F}
@@
 ```
+
 ```{r gh RAW 23, echo = F}
@@
 ```
+
 <!-- CLEAN DATA -->
+
 ```{r gh CLEAN 22, echo = F}
@@
 ```
+
 ```{r gh CLEAN 23, echo = F}
@@
 ```
+
 <!-- COMOPUNDS -->
+
 ```{r gh compounds, echo = F, results = F}
@@
 ```
+
 <!-- CLR -->
+
 ```{r gh CLR, echo = F}
@@
 ### Box Plot
+
 ```{r gh boxplot comparison, echo = F, fig.align = 'center', fig.width = 10}
@@
 ```
-:::{.callout-note}
+
+::: callout-note
 I chose to visualize the data using a box plot, not a bar plot, because our data has been CLR transformed.
@@
 As we can see, our data violates the normality assumption. This is the case even though we ran it though a CLR transformation.
-</div>
+:::
 
-:::{.callout-note}
-**A quick note here.**  
+::: callout-note
+**A quick note here.**\
 Many papers analyzing ecological or compositional data and VOC datasets often employ a statistical test known as [MANOVA](https://en.wikipedia.org/wiki/Multivariate_analysis_of_variance) (Multivariate Analysis of Variance).
@@
 ### PERMANOVA
+
 As described above, a [PERMANOVA](https://en.wikipedia.org/wiki/Permutational_analysis_of_variance) (Permutational Multivariate Analysis of Variance) uses permutation testing to generate many different rearrangements of the data and assess if the observed differences are statistically significant [@anderson2017].
@@
 ```
+
 ```{r random forest formatting and plot, echo = F}

### 2024-05-13 qmd/projects/msu/wss/eag.qmd
commit 5/13 2024

---
title: "eag"
editor_options: 
  chunk_output_type: console
---

```{r library, echo = F, message = F}
library(dplyr)
library(ggplot2)
library(readxl)
library(plotly)
library(plotly)
library(stringr)
library(tidyr)
```
```{r scale values, echo = F}
# Define the original range
min_original <- 1
max_original <- 55680

# Define the target range
min_target <- 1.823
max_target <- 10.02

# Function to scale the values
scale_values <- function(x) {
  ((x - min_original) / (max_original - min_original)) * (max_target - min_target) + min_target
}

```

```{r, echo = F, eval = F}
eag <-
  read_excel("eag.xlsx", sheet = "test eag")

fid <-
  read_excel("eag.xlsx", sheet = "Peak")

# Generate a sequence of MeasRetTime values every 0.05
new_MeasRetTime <- seq(min(fid$MeasRetTime), 
                       max(fid$MeasRetTime), by = 0.05)

# Create a data frame with the new MeasRetTime values
new_data <- data.frame(MeasRetTime = new_MeasRetTime)

# Merge the new data frame with the original data, filling missing values with 0
filled_data <- merge(fid, new_data, by = "MeasRetTime", all = TRUE)
filled_data[is.na(filled_data)] <- 0

fid_plot <- 
filled_data %>%
  dplyr::select(MeasRetTime, Area, Height) %>%
  mutate(Height = ifelse(MeasRetTime < 3.2, 0, Height)) %>%
  mutate()
  # ggplot() +
  # geom_point(aes(x = MeasRetTime, y = Height)) +
  # geom_line(aes(x = MeasRetTime, y = Height)) +
  # theme_bw()
eag_plot <-
eag %>%
  mutate(row_num = row_number()) %>%
  rename("EAD" = 'EAD 1')

ggplotly(
ggplot() +
  geom_line(data = fid_plot, aes(x = MeasRetTime, 
                                 y = Height), 
            color = "red") +
  geom_line(data = eag_plot %>%
              mutate(EAD = ifelse(row_num == 42500, -1.5, EAD),
                     EAD = ifelse(row_num == 42501, -1, EAD)), aes(x = scale_values(row_num)+0.2, 
                                 y = EAD * 100), 
            color = "blue") +

  scale_y_continuous(name = "EAD", sec.axis = sec_axis(~./100, name = "Height")) + 
  labs(title = "Comparison of EAD and Height",
       x = "Row Number (Scaled)", 
       y = "EAD",
       color = "Legend") +
  theme_bw() 
)

# ggplotly(
#   ggplot(data = eag_plot,
#          aes(x = row_num,
#              y = EAD)) +
#     geom_line() +
#     theme_bw()
# )

stimuli <- 
eag_plot %>%
  filter(row_num >= 19220 & row_num <= 19810)
```

```{r, echo = F}

setwd("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/GitHub/Quarto Website/Quarto Website/qmd/projects/msu/wss/")

eag <-
  read_excel("eag.xlsx", sheet = "tt424") %>%
  rename("EAD1" = "EAD 1",
         "FID1" = "FID 1")

ggplotly(
  eag %>%
  filter(time > 40012,
         time < 74691) %>%
  bind_rows(eag %>%
              dplyr::select(EAD1, time) %>%
              filter(time > 40013,
                     time < 45091) %>%
              mutate(EAD1 = EAD1 + runif(n = n(), 
                                         min = 0.001, max = 0.01)) %>%
              mutate(time = 74691 + row_number() - 1)) %>%
  mutate(EAD1 = ifelse(time > 65870 & time < 66011, EAD1 * 16, EAD1),
         EAD1 = ifelse(time > 69568 & time < 69629, EAD1 * 20, EAD1)) %>%
  ggplot() +
  geom_line(aes(x = time, y = EAD1*10 -1), color = "blue") +
  geom_line(aes(x = time, y = FID1)) +
  labs(y = "Response quantity",
       x = "Time (minutes)") +
  theme_bw()
)

eag %>%
  dplyr::select(EAD1, time) %>%
  filter(time > 100,
         time < 45400) %>%
  mutate(EAD1 = EAD1 + runif(1, min = 0.1, max = 1)) %>%
  mutate(time = 74691 + row_number() - 1) %>%
  ggplot() +
  geom_line(aes(x = time, y = EAD1))

eag %>%
  mutate(EAD2 = EAD1 + runif(n = n(), min = 0.1, max = 1))

ggplotly(
eag %>%
  filter(time > 40012,
         time < 74691) %>%
  bind_rows(eag %>%
              dplyr::select(EAD1, time) %>%
              filter(time > 40013,
                     time < 50091) %>%
              mutate(EAD1 = EAD1 + runif(n = n(), 
                                         min = 0.001, max = 0.01)) %>%
              mutate(time = 74691 + row_number() - 1)) %>%
  mutate(EAD1 = ifelse(time > 65870 & time < 66011, EAD1 * 16, EAD1),
         EAD1 = ifelse(time > 69568 & time < 69629, EAD1 * 20, EAD1)) %>%
  ggplot() +
  geom_line(aes(x = time/2.1 + 34620, y = EAD1*10 -1), color = "blue") +
  geom_line(aes(x = time, y = FID1)) +
  labs(y = "Response (mV)",
       x = NULL) +
  theme_bw() +
  xlim(c(53521,74690)) +
  # ylim(c(-12, 15)) +
  annotate("text", x = 63500, y = 10,
           label = "6-methyl-5-hepten-2-one",
           size = 2.5) +
  annotate("text", x = 70500, y = 13,
           label = "Z-3-hexenyl acetate",
           size = 2.5) +
  annotate("text", x = 71000, y = -2.5,
           label = "D-limonene",
           size = 2.5) +
  annotate("text", x = 71600, y = 3,
           label = "Ocimene",
           size = 2.5)
)
```

```{r 11a ib8, echo = F}
ib7 <-
read_excel("EAD_FID/EAD_11A_IB7.xlsx")

ggplotly(
ib7 %>%
  rename("EAD1" = "EAD 1",
         "FID1" = "FID 1") %>%
  mutate(time = row_number()) %>%
  dplyr::select(time, FID1, EAD1) %>%
  pivot_longer(-time,
               names_to = "type",
               values_to = "reading") %>%
  filter(time > 24581,
         time < 94230) %>%
  ggplot() +
  geom_line(aes(x = time, y = reading, color = type)) +
  theme_bw()
)

eag %>%
  mutate(EAD1 = ifelse(time > 65870 & time < 66011, EAD1 * 16, EAD1),
         EAD1 = ifelse(time > 69568 & time < 69629, EAD1 * 20, EAD1)) %>%
  ggplot() +
  geom_line(aes(x = time, y = EAD1))

stim1 <-
  eag %>%
  filter(time > 65870,
         time < 66011) %>%
  mutate(EAD1 = EAD1 * 16) %>%
  dplyr::select(time, EAD1)

stim1 %>%
  ggplot() +
  geom_line(aes(x = time, y = EAD1)) +
  xlim(c(64500, 67000))

stim1_start <- 47140
stim1_end <- stim1_start + 141

ggplotly(
ib7 %>%
  rename("EAD1" = "EAD 1",
         "FID1" = "FID 1") %>%
  mutate(time = row_number()) %>%
  filter(time > 24581,
         time < 94230) %>%
  dplyr::select(time, FID1, EAD1) %>%
  filter(time > stim1_start | time < stim1_end) %>%
  bind_rows(eag %>%
              filter(time > 65870,
                     time < 66011) %>%
              mutate(time = stim1_start + row_number() - 1,
                     EAD1 = EAD1 * 16) %>%
              dplyr::select(time, EAD1)) %>%
  pivot_longer(-time,
               values_to = "values",
               names_to = "names") %>%
  ggplot() +
  geom_line(aes(x = time, y = values, color = names)) +
   # geom_vline(xintercept = 65870) +
  theme_bw() +
  theme(legend.position = "none")
)

stim1_start <- 47140
stim1_end <- stim1_start + 141

ggplotly(
eag %>%
  dplyr::select(EAD1, time) %>%
  filter(time > 15000) %>%
  mutate(EAD1 = EAD1 * 0.15) %>%
  left_join(ib7 %>%
              mutate(time = row_number()) %>%
              rename("FID1" = "FID 1"),
            by = "time") %>%
  dplyr::select(time, EAD1, FID1) %>%
  mutate(FID1 = ifelse(time > stim1_start | time < stim1_end, 
                       NA_real_, FID1)) %>%
  filter(time > stim1_start | time < stim1_end) %>%
  ggplot() +
  geom_line(aes(x = time, y = EAD1),
            color = "dark blue",
            linewidth = 0.2) +
  geom_line(aes(x = time, y = FID1),
            linewidth = 0.2) +
  theme_bw()
)
```

```{r stim1, echo = F}
  read_excel("eag.xlsx", sheet = "tt424") %>%
  rename("EAD1" = "EAD 1",
         "FID1" = "FID 1") 
```

```{r blank EAD, echo = F}
blank_ead <-
read_excel("EAD_FID/eag.xlsx") %>%
  dplyr::select(3) %>%
  distinct() %>%
  rename("EAD1" = 1) %>%
  mutate(time = row_number())

blank_ead %>%
  ggplot() +
  geom_line(aes(x = time, y = EAD1)) +
  ylim(c(-10, 0))

```

### 2024-05-13 qmd/projects/msu/wss/field_report.qmd
commit 5/13 2024

@@
 ---
+
 <!-- field site cover image -->
+
 ```{=html}
@@
 </html>
-
 ```
-
 ```{r library, echo = F, message = F, warning = F}
@@
 ```
+
 ```{r column.order, echo = F}
@@
 ```
+
 ```{r (stubble.####) stubble data, echo = F}
@@
 ```
+
 ```{r (joined.stubble) stubble join, echo = F}
@@
 ```
+
 ```{r (n.inf) brome join, echo = F}
@@
 ```
+
 ```{r (stubble.brome) join stubble/brome, echo = F}
@@
 ```
+
 ```{r, echo = F, eval = F}
@@
 ```
-
 <!-- ![North Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/nnns-min.png?raw=true) -->
 
-**Note:**    
+**Note:**\
 - In 2023, the field south of North Road was fallow, so no stubble infestation data was collected from the field south of North Road.
@@
 ```
-
 <!-- ![Gardiner Road, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/gardiner%20road-min.png?raw=true) -->
@@
 ```
-
 <!-- ![Harvest & Sunset Road Instersection, Big Sandy, MT (2023)](https://github.com/jstrand894/Website/blob/main/harvest%20road-min.png?raw=true) -->
@@
 ```
-
 <!-- ![Kolin Road, Moccasin, MT (2023)](https://github.com/jstrand894/Website/blob/main/kolin%20road-min.png?raw=true) -->

### 2024-05-13 qmd/projects/msu/wss/greenhouse_volatiles.qmd
commit 5/13 2024

---
title: "greenhouse volatiles"
editor_options: 
  chunk_output_type: console
---

```{r library, echo = F, message = F, warning = F}
library(dplyr)
library(readxl)
library(ggplot2)
library(kableExtra)

```

```{r data, echo = F}
meta.23 <-
read_excel("greenhouse_volatiles/BWW 2023 Results.xlsx",
           sheet = "meta.data")

meta.22 <-
read_excel("greenhouse_volatiles/BWW 2022 Results.xlsx",
           sheet = "meta.data")
```

```{r filter, echo = F, eval = F}
meta.23 %>% 
  dplyr::select(Sample, rep, sample.id,
                inf, p.type) %>%
  filter(p.type == "b",
         inf == "y") %>%
  mutate(Value = as.numeric(gsub("\\D", "", rep))) %>%
  filter(rep == "11A")

meta.22 %>%
  dplyr::select(sample.id, infested, type) %>%
  filter(sample.id == "1A7")
  filter(infested == "Y",
         type == "B")

```

### 2024-05-13 qmd/projects/msu/wss/ndvi/ndvi.qmd
commit 5/13 2024

---
title: "ndvi"
editor_options: 
  chunk_output_type: console
---

```{r library, echo = F, message = F}
library(readxl)
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(lubridate)
```

```{r big sandy, echo = F}
# setwd("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/GitHub/Quarto Website/Quarto Website/qmd/projects/msu/wss")

lochy <-
read_excel("ndvi/ndvi_2023.xlsx", sheet = "lochy") %>%
  filter(use == "yes") %>%
  mutate(date = ymd(date1)) %>%
  dplyr::select(date,wheat,brome) %>%
  pivot_longer(-date,
               names_to = "plant_type",
               values_to = "ndvi")

lochy %>%
  ggplot() +
  geom_point(aes(x = date, y = ndvi, color = plant_type),
             # alpha = 0.5,
             pch = 1,
             size = 0.75) +
  geom_line(aes(x = date, y = ndvi, color = plant_type),
            alpha = 0.5, 
            size = 0.3) +
  geom_smooth(aes(x = date, y = ndvi, color = plant_type),
              se = FALSE) +
  theme_bw() +
  labs(x = NULL,
       y = "NDVI",
       color = "Plant Type",
       title = "NDVI comparison between B. inermis and adjacent wheat field")

```

This data is not as good. Maybe need to find a new site to investigate. 
```{r moccasin, echo = F}
moccasin <-
read_excel("ndvi_2023.xlsx", sheet = "moccasin") %>%
  dplyr::select(3:6) %>%
  rename("date" = 1,
         "use" = 3) %>%
  filter(use %in% c("yes",NA)) %>%
  mutate(date = ymd(date)) %>%
  dplyr::select(date,wheat,brome) %>%
  pivot_longer(-date,
               names_to = "plant_type",
               values_to = "ndvi") %>%
  filter(!date %in% c("2023-05-07",
                      "2023-05-12"))

ggplotly(
moccasin %>%
  ggplot() +
  geom_point(aes(x = date, y = ndvi, color = plant_type),
             alpha = 0.5,
             color = "black") +
  geom_line(aes(x = date, y = ndvi, color = plant_type),
            alpha = 0.5) +
  geom_smooth(aes(x = date, y = ndvi, color = plant_type),
              se = FALSE) +
  theme_bw() +
  labs(x = NULL,
       y = "NDVI",
       color = "Plant Type",
       title = "NDVI comparison between B. inermis and adjacent wheat field")
)
```

```{r plot, echo = F, eval = F}
ggplotly(data %>%
  filter(!date %in% c("2023-03-31",
                      "2023-04-20",
                      "2023-05-10",
                      "2023-05-30",
                      "2023-06-09",
                      "2023-06-19",
                      "2023-07-14",
                      "2023-08-13")) %>%
  # mutate(NDVI = ifelse(type == "wheat", -NDVI, NDVI)) %>%
  ggplot(aes(x = date, y = NDVI, fill = type)) +
  geom_bar(stat = 'identity', 
           position = position_dodge()) +
  geom_line(stat = "smooth") +
  theme_bw()
)
```

### 2024-05-13 qmd/projects/msu/wss/paper1.qmd
commit 5/13 2024

---
title: "Paper 1: Smooth brome and parasitoids"
editor_options: 
  chunk_output_type: console
bibliography: proposal.bib
link-citations: true
biblio-style: apalike
cls: elsevier-with-titles.csl
---

## Introduction

## Materials and Methods

### Weather data and NDVI analysis

*Weather data.* We assessed the long and medium term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA). Data for each field site was averaged from three of the closest weather stations to that location. Data was plotted using R Studio (R Studio version 2024.04.0+735, R 2024) package 'ggplot' (version 3.4.4) [@ggplot]. Data was then fit using a linear model using the 'lm' command using average yearly precipitation (inches) as the response variable and year as the predictor.

*NDVI analysis.* We compared the relative greening throughout the growing season between wheat fields and adjacent *B. inermis* using the normalized difference vegetation index (NDVI). NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infared light reflected by vegetation [@Pettorelli2005]. Data was downloaded using Google Earth Engine (Google Inc. 2023, Mountain View, CA, USA).

### Controlled *C. cinctus* infestation of *B. inermis*

*Insects and Cages.* Assessment of *C. cinctus* infestation and mortality within *B. inermis* were assessed using a 34 x 60 ft plot at the Arthur H. Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT. The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC). Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.

Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of *C. cinctus* infestation and cutting the year prior. Cut stubble, which contained *C. cinctus* larvae in diapause, were kept refrigerated between -2°C and 3°C for \>100 days as required to complete obligatory larval diapause. As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA). Once *B. inermis* stems reached six inches tall, stub containers with emerging sawflies were added to cages to mimic sawfly infestation pressure. Sawfly quantity treatments were as follows: high (600 stubs), low (200 stubs), and control (0 stubs).

*Data Collection.* In late August, *B. inermis* stems were collected from each cage. Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.

### Montana Field Survey

*Stem collection and processing.* We conducted a field survey to assess *C. cinctus* infestation, larval mortality, and *B. cephi* and *B. lissogaster* prevalence within *B. inermis* and adjacent wheat fields. Sites were chosen across 2 counties in north-central Montana, United States. (Chouteau, Judith Basin), which consistently experience high *C. cinctus* pressure. Samples were collected from wheat fields and adjacent *B. inermis* in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA. Sampling sites were set up as 100$m^2$ polygons along the edge of adjoining wheat fields. Four collection squares of 1ft x 1ft were randomly selected within each polygon during both collection events each year. All stems within each 1 x 1 ft square were collected using a shovel to remove both stem and root material. Wheat stems were collected at distances of 5 and 20 meters from the edge of the field. Samples were collected in 4 rows at 10 meters apart. 2 samples were collected in each row at distances of 5 and 20 meters. 1 ft samples were collected at each point.

Wheat and *B. inermis* stems were then returned to Montana State University, Bozeman, Montana and stored in a 10°C cold wet storage until dissection. Stems were dissected lengthwise with a fine-bladed scalpel to determine presence or absence of *C. cinctus* larvae infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting. Sawfly larvae were identified based on descriptions in Criddle (1915) and Wallace and McNeal (1996).

*Statistical Analysis.* We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion os stems parasitized by *Bracon* spp. '

To better understand the density of *Bracon* spp. within the wheat and *B. inermis* sampling sites, we converted our units to the ratio of parasitoids to stem per unit area. Densities of stems are different when considering *B. inermis* and cultivated wheat or barley. Using this conversion, we attempted to better understand how the density of parasitoids (and WSS) is changed by looking at each kind of plant.

## Results

```{r setup1, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)
```

```{r libraries}
library(markdown)
library(tidyverse)
library(glue)
library(lubridate)
library(ggplot2)
library(maps)
library(readxl)
library(kableExtra)
library(openxlsx)
```

### Historical Weather Analysis

```{r moc_weather}
# moccasin NOAA
moc_weather_noaa <-
read_excel("weather/moccasin_weather_noaa.xlsx") %>%
  filter(station == "USW00024036") %>%
  dplyr::select(date, variable, value, year) %>%
  mutate(date = ymd(date))

# moccasin prcp MAES
moc_prcp_maes <-
  read_excel("weather/moc_weather_maes.xlsx",
             sheet = "prcp") %>%
    dplyr::select(-1, -Annual, -Seasonal) %>%
    pivot_longer(-Year,
                 names_to = "month",
                 values_to = "prcp") %>%
    mutate(Year = lubridate::year(as.Date(Year, format = "%Y"))) %>%
    rename("year" = Year) %>%
    group_by(year) %>%
    reframe(prcp = sum(prcp))

# moccasin temp MAES
moc_temp_maes <-
  read_excel("weather/moc_weather_maes.xlsx",
             sheet = "temp") %>%
    dplyr::select(-1, -Annual, -Seasonal) %>%
    pivot_longer(-Year,
                 names_to = "month",
                 values_to = "temp") %>%
    mutate(Year = lubridate::year(as.Date(Year, format = "%Y"))) %>%
    rename("year" = Year) %>%
    group_by(year) %>%
    reframe(temp = mean(temp))

# past 50 years prcp data
moc_prcp_50 <-
moc_prcp_maes %>%
  filter(year > 1975)

# past 50 years temp data
moc_temp_50 <-
  moc_temp_maes %>%
  filter(year > 1975)
```

```{r bs_weather}
# big sandy
bs_weather <-
read_excel("weather/bigsandy_weather_noaa.xlsx") %>%
  filter(station == "USC00240770") %>%
  dplyr::select(date, variable, value, year) %>%
  mutate(date = ymd(date))

# past 50 years prcp data
bs_prcp_50 <-
bs_weather %>%
  filter(variable == "PRCP") %>%
  group_by(year) %>%
  reframe(prcp = mean(value)) %>%
  distinct() %>%
  filter(year > 1975,
         year < 2024,
         prcp < 25)

bs_temp_noaa <-
bs_weather %>%
  filter(variable %in% c("TMAX", "TMIN"),
         value < 100,
         value > -30) %>%
  pivot_wider(names_from = "variable",
              values_from = "value") %>%
  na.omit() %>%
  group_by(date) %>%
  mutate(daily = (TMAX - TMIN)/2) %>%
  group_by(year) %>%
  reframe(temp = mean(daily))

```

```{r both_prcp_plot}
both_locations_prcp <-
bs_prcp_50 %>%
  mutate(location = "Big Sandy") %>%
  rbind(moc_prcp_50 %>%
          mutate(location = "Moccasin"))

both_prcp_lm <-
both_locations_prcp %>%
  filter(location == "Big Sandy") %>%
  lm(prcp ~ year, data = .) %>%
  summary()

moc_50_lm <-
  moc_prcp_50 %>%
  lm(prcp ~ year, data = .) %>%
  summary()

bs_50_lm <-
  bs_prcp_50 %>%
  lm(prcp ~ year, data = .) %>%
  summary()

calculate_segment <- function(model) {
  intercept <- model$coefficients[1, 1]
  slope <- model$coefficients[2, 1]
  segment_formula <- function(x) intercept + slope * x
  x_start <- 1975
  x_end <- 2024
  y_start <- segment_formula(x_start)
  y_end <- segment_formula(x_end)
  segment_data <- data.frame(x_start, y_start, x_end, y_end)
  return(segment_data)
}

bs_segment <- calculate_segment(bs_50_lm)
moc_segment <- calculate_segment(moc_50_lm)

both_prcp_plot <-
both_locations_prcp %>%
  ggplot() +
  theme_bw() +
  geom_segment(data = bs_segment,
               aes(x = x_start, 
                   y = y_start,
                   xend = x_end, 
                   yend = y_end),
               color = "#F8766D",
               alpha = 0.5) +
  geom_segment(data = moc_segment,
               aes(x = x_start, 
                   y = y_start,
                   xend = x_end, 
                   yend = y_end),
               color = "#00BFC4",
               alpha = 0.5) +
  geom_line(aes(x = year, y = prcp, color = location),
            size = 0.3) +
  geom_smooth(aes(x = year, y = prcp),
              method = "lm",
              se = FALSE,
              color = "black",
              size = 0.5,
              alpha = 0.2) +
  labs(color = "Location (MT, USA)",
       x = NULL,
       y = "Average Yearly Precipitation (in)",
       title = "Average Yearly Precipitation",
       subtitle = "Big Sandy & Moccasin, MT, USA")
```

```{r both_temp_plot}
both_temp <-
moc_temp_maes %>%
  mutate(location = "Moccasin") %>%
  rbind(bs_temp_noaa %>%
          mutate(location = "Big Sandy"))

both_temp_plot <-
both_temp %>%
  ggplot() +
  geom_line(aes(x = year, y = temp, color = location))

# lm both
both_temp_lm <-
both_temp %>%
  lm(temp ~ year, data = .) %>%
  summary()

# lm moc
moc_temp_lm <-
moc_temp_maes %>%
  lm(temp ~ year, data = .) %>%
  summary()

calculate_segment <- function(model) {
  intercept <- model$coefficients[1, 1]
  slope <- model$coefficients[2, 1]
  segment_formula <- function(x) intercept + slope * x
  x_start <- 1975
  x_end <- 2024
  y_start <- segment_formula(x_start)
  y_end <- segment_formula(x_end)
  segment_data <- data.frame(x_start, y_start, x_end, y_end)
  return(segment_data)
}

# moc_segment <- calculate_segment(moc_lm)

moc_temp_plot <-
moc_temp_maes %>%
  ggplot() +
  theme_bw() +
  geom_line(aes(x = year, y = temp),
            size = 0.4) + 
  geom_smooth(aes(x = year, y = temp),
              method = "lm",
              se = FALSE,
              size = 0.5,
              alpha = 0.2) +
  labs(color = "Location (MT, USA)",
       x = NULL,
       y = "Average Yearly Temperature (F)",
       title = "Average Yearly Temperature",
       subtitle = "Moccasin, MT, USA")

```

We observed a significant linear relationship (*r = `r round(both_prcp_lm$r.squared,3)`, P = `r round(both_prcp_lm$coefficients[2,4],3)`, estimate = `r round(both_prcp_lm$coefficients[2,1],3)`*) between average precipitation and year for both Big Sandy and Moccasin, Montana. This means that for each one year increase, we are seeing a decrease in 0.05 inches of precipitation.

```{r FIG1_both_prcp_plot_show, fig.cap = "Figure 1: Average yearly precipitation for Big Sandy and Moccasin, Montana. Black trend line signifies averaged negative trend between both locations. Data gathered from NOAA and MTARS."}
both_prcp_plot
```

In addition, we observed a significant positive linear relationship (*r = `r round(moc_temp_lm$r.squared,3)`, P \< 0.05, estimate = `r round(moc_temp_lm$coefficients[2,1],3)`*) between year and average yearly temperature for Moccasin, MT. This means that each year, the average daily temperature has increased by `r round(moc_temp_lm$coefficients[2,1],2)`°F.

```{r moc_temp_plot_show}
moc_temp_plot
```

### NDVI Analysis of Field Sites

```{r Big_Sandy_NDVI}
lochy <-
read_excel("ndvi/ndvi_2023.xlsx", sheet = "lochy") %>%
  filter(use == "yes") %>%
  mutate(date = ymd(date1)) %>%
  dplyr::select(date,wheat,brome) %>%
  pivot_longer(-date,
               names_to = "plant_type",
               values_to = "ndvi") 
ndvi_model <-
lochy %>%
  filter(date > "2023-07-15") %>%
  lm(ndvi ~ plant_type + date, data = .) %>%
  summary()

label_x <- as.Date("2023-07-20")
label_y <- 0.75
line_x <- as.numeric(as.Date("2023-07-20"))

bs_ndvi_plot <-
lochy %>%
  ggplot() +
  geom_point(aes(x = date, y = ndvi, color = plant_type),
             alpha = 0.5,
             pch = 1,
             size = 0.75) +
  geom_line(aes(x = date, y = ndvi, color = plant_type, linetype = plant_type),
            alpha = 0.5,
            size = 0.5) +
  geom_smooth(aes(x = date, y = ndvi, color = plant_type),
              se = FALSE) +
  geom_vline(xintercept = line_x,
             linetype = "dashed", color = "black") +
  geom_label(aes(x = label_x,
                 y = 0.75),
             label = "Spring wheat harvest",
             hjust = -0.1,
             size = 3) +
  geom_segment(aes(x = as.Date("2023-07-20")+5, 
                   y = 0.75, 
                   xend = as.Date("2023-07-20"), 
                   yend = 0.75)) +
  theme_bw() +
  guides(color = FALSE) +
  labs(x = "2023 Growing Season",
       y = "NDVI",
       linetype = NULL,
       title = "Seasonal NDVI Comparison",
       subtitle = expression(paste(italic("B. inermis"),
                                   " and adjacent spring wheat"))) +
  scale_linetype_discrete(labels = c(expression(italic("B. inermis")),
                                                      "Spring Wheat"))
```

We observed a notable difference in NDVI when comparing adjacent *B. inermis* and spring wheat. We saw a significant difference in the July NDVI (*`r round(ndvi_model$r.squared,3)`, P \< 0.05*). The *B. inermis* NDVI remained relatively linear in it's downslope (BROME SLOPE POST JULY) compared to the wheat field (WHEAT FIELD POST JULY).

```{r bs_ndvi_plot}
bs_ndvi_plot
```

Need to add map showing where NDVI measurments came from

### Controlled *C cinctus* Infestation of *B. inermis*

```{r column.order}
column.order <- c("distance", "collection", "p.inf", 
                  "p.cut", "p.par", "poten")

column.order2 <- c("location", "distance", "collection", "p.inf", 
                  "p.cut", "p.par", "poten")

round_based_on_size <- function(x) {
  threshold <- 0.01
  rounded_values <- ifelse(abs(x) > threshold, round(x, 2), round(x, 3))
  return(rounded_values)
}
```

```{r (pf.RAW) post farm n.inf data}
pf.raw.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2022/Post\ Farm\ 2022.xlsx", sheet = "n.inf")
pf.raw.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "n.inf")
```

```{r (pf.inf.23) post farm inf data}
pf.inf.23 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf") %>%
  rename(n.1 = 8, n.2 = 9, n.3 = 10, n.4 = 11,
         n.5 = 12, n.6 = 13, n.7 = 14, n.8 = 15,
         n.9 = 16, n.10 = 17)
```

```{r (pf.comb) post farm data combined}
# 2022
pf.2022.site <-
pf.raw.2022 %>%
  group_by(treatment, site) %>%
  reframe(site, treatment,
            p.inf = sum(iv, ir, ix)/sum(uv, ur, iv, ir, ix),
            cut = sum(cut)/sum(total),
            total.stems = sum(iv, ir, ix, uv, ur)) %>%
  unique() %>%
  pivot_longer(-c(treatment, total.stems, site),
               names_to = "type",
               values_to = "percent") %>%
  mutate(year = "2022")

# 2023
pf.2023.site <-
pf.raw.2023 %>%
  group_by(treatment, site) %>%
  reframe(treatment, site, 
          t.inf = sum(n.iv, n.ir, n.ix),
          p.inf = sum(n.iv, n.ir, n.ix)/sum(n.uv, n.ur, 
                                            n.iv, n.ir, n.ix),
          cut = sum(cut)/sum(total),
          total.stems = sum(n.iv, n.ir, n.ix, n.uv, n.ur)) %>%
  distinct() %>%
  pivot_longer(-c(treatment, total.stems, site),
               names_to = "type",
               values_to = "percent") %>%
  filter(!type == "t.inf") %>%
  arrange(type) %>%
  mutate(year = "2023")

pf.comb <-
  pf.2022.site %>%
  rbind(pf.2023.site) %>%
  group_by(treatment, year, type) %>%
  mutate(count = n(),
         average = mean(percent),
         sd = sd(percent),
         se = (sd/sqrt(count))) %>%
  reframe(treatment, year, type,
          mean = mean(percent),
          sd = mean(sd),
          se = mean(se)) %>%
  unique()

```

```{r (pf.total) pf total stems}
# 2022 totals
pf.2022.total <-
  pf.raw.2022 %>%
  group_by(treatment) %>%
  reframe(total = sum(uv, ur, iv, ir, ix)) %>%
  mutate(year = "2022")

# 2023 totals
pf.2023.total <-
  pf.raw.2023 %>%
  group_by(treatment) %>%
  reframe(total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
  mutate(year = "2023")

pf.total.comb <-
  pf.2022.total %>%
  rbind(pf.2023.total)

```

```{r pf_lm}
raw.23 <-
pf.raw.2023 %>%
  dplyr::select(site, treatment,
                n.iv, n.ir, n.ix, n.ur, n.uv,
                cut) %>%
  rename(iv = n.iv,
         ir = n.ir,
         ur = n.ur,
         ix = n.ix,
         uv = n.uv) %>%
  mutate(year = 2023)

raw.22 <-
pf.raw.2022 %>%
  dplyr::select(site, treatment, 
                ix, ir, iv, ur, uv,
                cut) %>%
  mutate(year = 2022)

pf_inf_lm <-
raw.23 %>%
  rbind(raw.22) %>%
  group_by(site, treatment, year) %>%
  reframe(total = sum(iv, ir, ix, ur, uv),
          inf = sum(iv, ir, ix)) %>%
  lm(inf ~ treatment + year, data = .) %>%
  summary()

pf_cut_lm <-
raw.23 %>%
  rbind(raw.22) %>%
  group_by(site, treatment, year) %>%
  reframe(total = sum(iv, ir, ix, ur, uv),
          inf = sum(iv, ir, ix),
          cut = sum(cut)) %>%
  lm(cut ~ treatment + year, data = .) %>%
  summary()
```

```{r pf_per, eval = F}
raw.23 %>%
  rbind(raw.22) %>%
  group_by(treatment) %>%
  reframe(total = sum(iv, ir, ix, ur, uv),
          inf = sum(iv, ir, ix),
          cut = sum(cut),
          p.inf = inf/total,
          p.cut = cut/inf)
```

*C. cinctus* showed a high ability to infest *B. inermis* in the controlled test conditions. Averaged across both years, we observed 66.5% of stems infested for high treatments and 47.3% of stems for low treatments. We found strong evidence suggesting that there was a significant difference between infestation at high and low treatment levels when holding year constant (*r = `r round(pf_inf_lm$r.squared,3)`, P \< 0.05*).

Cutting was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P \< 0.05*).

```{r inf_cut_plot}
ggplot() +
  facet_wrap(~treatment) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           data = pf.comb %>%
             filter(type == "p.inf"),
           aes(year, mean, fill = "p.inf")) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           data = pf.comb %>%
             filter(type == "cut"),
           aes(year, mean, fill = "cut")) +
  geom_errorbar(data = pf.comb,
                aes(x = year,
                    ymin = mean - se,
                    ymax = mean + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  theme_bw() +
  labs(x = element_blank(),
       y = "Percentage of stems (%)",
       fill = element_blank()) +
  scale_fill_brewer(palette = "Greens",
                    labels = c("Cut", "Infested"))
```

### Montana Field Sites

```{r stubble data}
# stubble 2023
stubble.2023 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", 
           sheet = "stubble") %>%
  dplyr::select(-c(name,date.proc,"%inf", "%cut")) %>%
  na.omit() %>%
  mutate(year = "2023")

# stubble 2022
stubble.2022 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", 
             sheet = "stubble") %>%
  mutate(year = "2022") %>%
  dplyr::select(-c(name, date.proc, "%inf", "%cut"))

comb.stubble <-
stubble.2023 %>%
  rbind(stubble.2022) %>%
  rename("c" = cocoon,
         "e" = eh,
         "total.stems" = total) %>%
  group_by(site, year, collection) %>%
  mutate(inf = sum(cut, inf.nc,c,e)) %>%
  dplyr::select(-c(inf.nc, larvae)) 
```

```{r brome data}
# brome 2022
raw.2022 <- 
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "n.infested") %>%
  mutate(year = "2022") %>%
  dplyr::select(-c(other, date.proc, name, 
                   par.rate, cut.rate, inf)) %>%
  rename("site" = site.cap)

# brome 2023
raw.2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "n.infested") %>%
  mutate(year = "2023") %>%
  dplyr::select(-c(loc.stub, date.proc, uv, ur, iv, ir, ix, 
                   total.stems, add, inf, a)) %>%
    rename(uv = n.uv,
           ur = n.ur,
           iv = n.iv,
           ir = n.ir,
           ix = n.ix,
           total.stems = n.total)
# combine brome years
comb.brome <- 
raw.2023 %>%
  rbind(raw.2022) %>%
  group_by(site, year, collection) %>%
  mutate(inf = sum(iv, ir, ix),
         uninf = sum(uv, ur),
         cut = sum(cut)) %>%
  dplyr::select(-c(iv, ir, ix, uv, ur)) %>%
  mutate(distance = "brome")
```

We quantified *C. cinctus* and *Bracon* spp. parasitoid presence in 6,148 wheat and grass stems across 10 research sites in 2022 and 2023. Infestation by *C. cinctus* within *B. inermis* varied between collection sites, with the greatest infestation taking place within our three Big Sandy, MT sampling sites (2023: 65.4%, 2022: 63.1%), while the lowest infestation observed was in Moccasin, MT (2023: 40.8%, 2022: 60.7%). Across all sites in Big Sandy and Moccasin, we observed an average infestation of 57.5% within *B. inermis* and 47.6% within the adjacent wheat.

Parasitoid presence was observed at the highest levels in

```{r infestation_per, eval = F}
comb.brome %>%
  group_by(location, collection, year) %>%
  # filter(collection == "fall") %>%
  reframe(inf = sum(inf)/sum(total.stems)) %>%
  distinct() %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  group_by(collection, year, region) %>%
  reframe(inf = mean(inf)) %>%
  na.omit() %>%
  arrange(desc(inf))
```

```{r paraistism_per, eval = F}
# plot.data <-
comb.brome %>%
  rbind(comb.stubble) %>%
  na.omit() %>%
  filter(!location %in% c("chs", "b")) %>%
  mutate(sample_site = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "North Road",
    location %in% c("kw", "ke", "k") ~ "Kolin Road",
    location %in% c("sun", "har") ~ "Sunset & Harvest",
    location == "g" ~ "Gardiner Road",
    TRUE ~ NA_character_)) %>%
  mutate(region = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "Big Sandy",
    location == "g" ~ "Big Sandy",
    location %in% c("kw", "ke", "k") ~ "Moccasin",
    location %in% c("sun", "har") ~ "Moccasin",
    TRUE ~ NA_character_)) %>%
  mutate(plant_type = if_else(distance == "brome", "brome", "wheat")) %>%
  dplyr::select(year, region, sample_site, collection, plant_type,
                total.stems, inf, uninf, cut, c, e) %>%
  group_by(site, collection, year) %>%
  filter(collection == "fall") %>%
  mutate(par = sum(c,e)) %>% dplyr::select(-c(c,e)) %>%
  group_by(year, collection, region, sample_site, plant_type, site) %>%
  mutate(count = n(),
         per.inf = sum(inf)/sum(uninf),
         per.cut = sum(cut)/sum(inf),
         per.par = sum(par)/sum(inf)) %>%
  ungroup() %>%
  dplyr::select(-c(total.stems, site, inf, uninf, cut, par)) %>%
  pivot_longer(-c(year, collection, region, sample_site, 
                  plant_type, count),
               names_to = "variable",
               values_to = "value")
  group_by(year, collection, region, sample_site, plant_type) 
  mutate(count = sum(count),
         sd.inf = sd(per.inf),
         per.inf = mean(per.inf),
         sd.cut = sd(per.cut),
         per.cut = mean(per.cut),
         sd.par = sd(per.par),
         per.par = mean(per.par)) %>%
  dplyr::select(-c(total.stems, site, inf, uninf, cut, par))
  group_by(year, region, sample_site, collection, plant_type) %>%
  reframe(inf = sum(inf),
          uninf = sum(uninf),
          cut_sd = sd(cut),
          cut = sum(cut)/inf,
          par_sd = sd(par),
          par = sum(par),
          count = sum(count),
          par_se = par_sd/sqrt(count),
          cut_se = cut_sd/sqrt(count)) 
  mutate(inf_per = inf/(inf + uninf),
         cut_per = cut/inf,
         sd = sd(cut))

mutate(count = n(),
         par_sd = sd(sum(c,e)),
         cut_sd = sd(cut),
         par = sum(c,e),
         cut = sum(cut),
         inf = sum(inf),
         par_per = par/inf,
         cut_per = cut/inf) %>%
  ungroup() %>%
  dplyr::select(-c(site, location, c, e, cut, inf, 
                   uninf, distance, total.stems)) %>% distinct()
  distinct() %>% 
  group_by(year, region, sample_site, collection, plant_type) %>%
  pivot_longer(-c(year, region, sample_site, 
                  collection, plant_type, count),
               names_to = "per_category",
               values_to = "per") %>%
  group_by(year, region, sample_site, collection, plant_type,
           per_category)

  mutate(plant_type = if_else(distance == "brome", "brome", "wheat"),
         p = sum(c,e)) %>%
  group_by(year, region, sample_site, plant_type) %>%
  reframe(year, region, sample_site, plant_type,
          total.stems = sum(total.stems),
          inf_per = sum(inf)/sum(total.stems),
          par_sum = sum(p),
          par_sd = sd(p),
          par_se = par_sd/sqrt(sum(inf)),
          cut_sum = sum(cut),
          cut_sd = sd(cut),
          cut_se = cut_sd/sqrt(sum(inf)),
          par_per = par_sum/sum(inf),
          par_per_sd = par_sd/sum(inf),
          par_per_se = par_se/sum(inf),
          cut_per = cut_sum/sum(inf),
          cut_per_se = cut_se/sum(inf)) %>%
  dplyr::select(-c("par_sd", "cut_sd", "total.stems",
                   "par_sum", "cut_sum")) %>%
  distinct()

          par_per = par_sum/sum(inf),
          par_per_sd = par_sd/sum(inf),
          par_per_se = par_se/sum(inf),
          cut_per = cut_sum/sum(inf),
          cut_per_se = cut_se/sum(inf)) %>%
  distinct() %>% na.omit() %>%
  pivot_longer(-c(year, region, sample_site, plant_type, total.stems),
               names_to = "type",
               values_to = "value") %>%
  distinct()

plot.data %>%
  filter(region == "Big Sandy", 
         sample_site == "North Road",
         plant_type == "wheat")

plot.data %>%
  filter(region == "Big Sandy") %>%
  ggplot() +
  # geom_bar(stat = "identity",
  #          aes(x = sample_site, y = par_per, fill = plant_type),
  #          position = position_dodge(width = 0.6),
  #          width = 0.5) +
  geom_bar(stat = "identity",
           aes(x = sample_site, y = cut_per, fill = plant_type),
           position = position_dodge(width = 0.6),
           width = 0.5) +
  geom_errorbar(aes(x = sample_site,
                    ymin = cut_per - cut_per_se*10,
                    ymax = cut_per + cut_per_se*10,
                    group = plant_type),
                position = position_dodge(0.6),
                width = 0.2,
                alpha = 0.5) +
  # geom_errorbar(aes(x = sample_site,
  #                   ymin = par_per - par_per_se*10,
  #                   ymax = par_per + par_per_se*10,
  #                   group = plant_type),
  #               position = position_dodge(0.6),
  #               width = 0.2,
  #               alpha = 0.5) +
  facet_wrap(~year) +
  theme_bw()

mutate(type2 = if_else(
    type %in% c("p_mean", "p_sd", "p_se"), "p", "cut"),
    stat = str_replace(type, ".*_", "")) %>%
  dplyr::select(-type) %>%
  pivot_wider(values_from = value,
              names_from = stat) %>%
  filter(region == "Big Sandy")

plot.data %>%
  ggplot() +
  geom_bar(stat = "identity",
           aes(x = plant_type, y = mean, 
               fill = type2),
           position = position_dodge()) +
  facet_wrap(sample_site ~ year) +
  theme_bw()

ggplot() +
  facet_grid(sample_site ~ region) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           data = par.data %>%
             filter(type == "p.inf"),
           aes(year, value, fill = "p.inf")) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           data = par.data %>%
             filter(type == "p.cut"),
           aes(year, value, fill = "p.cut")) +
  geom_errorbar(data = par.data,
                aes(x = year,
                    ymin = mean - se,
                    ymax = mean + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  theme_bw() +
  labs(x = element_blank(),
       y = "Percentage of stems (%)",
       fill = element_blank()) +
  scale_fill_brewer(palette = "Greens",
                    labels = c("Cut", "Infested"))
```

*B. inermis* displayed very low levels of *C. cinctus* cutting when compared to adjacent wheat fields. In Moccasin,

```{r infestation, eval = F}
comb.brome %>%
  group_by(location, collection, year) %>%
  # filter(collection == "fall") %>%
  reframe(inf = sum(inf)/sum(total.stems)) %>%
  distinct() %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  group_by(collection, year, region) %>%
  reframe(inf = mean(inf)) %>%
  na.omit() %>%
  arrange(desc(inf)) %>%
  group_by(collection) %>%
  reframe(inf = mean(inf))

  # na.omit() %>%
  # ggplot() +
  # geom_bar(stat = "identity",
  #          aes(x = location, y = inf)) +
  # facet_grid(year~collection)
```

```{r, eval = F}
comb.stubble <-
stubble.2023 %>%
  rbind(stubble.2022) %>%
  rename("c" = cocoon,
         "e" = eh,
         "total.stems" = total) %>%
  group_by(site, year, collection) %>%
  mutate(inf = sum(cut, inf.nc,c,e)) %>%
  dplyr::select(-c(inf.nc, larvae))

comb.stubble %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  ungroup() %>%
  reframe(inf = sum(inf)/sum(total.stems))
  group_by(region, collection, year) %>%
  reframe(inf = mean(inf))
```

```{r plot ratios with error bars}
# brome 2022
raw.2022 <- 
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "n.infested") %>%
  mutate(year = "2022") %>%
  dplyr::select(-c(other, date.proc, name, 
                   par.rate, cut.rate, inf)) %>%
  rename("site" = site.cap)

# brome 2023
raw.2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "n.infested") %>%
  mutate(year = "2023") %>%
  dplyr::select(-c(loc.stub, date.proc, uv, ur, iv, ir, ix, 
                   total.stems, add, inf, a)) %>%
    rename(uv = n.uv,
           ur = n.ur,
           iv = n.iv,
           ir = n.ir,
           ix = n.ix,
           total.stems = n.total)
# combine brome years
comb.brome <- 
raw.2023 %>%
  rbind(raw.2022) %>%
  group_by(site, year, collection) %>%
  mutate(inf = sum(iv, ir, ix),
         uninf = sum(uv, ur)) %>%
  dplyr::select(-c(iv, ir, ix, uv, ur)) %>%
  mutate(distance = "brome")

# stubble 2023
stubble.2023 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", 
           sheet = "stubble") %>%
  dplyr::select(-c(name,date.proc,"%inf", "%cut")) %>%
  na.omit() %>%
  mutate(year = "2023")

# stubble 2022
stubble.2022 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", 
             sheet = "stubble") %>%
  mutate(year = "2022") %>%
  dplyr::select(-c(name, date.proc, "%inf", "%cut"))

comb.stubble <-
stubble.2023 %>%
  rbind(stubble.2022) %>%
  rename("c" = cocoon,
         "e" = eh,
         "total.stems" = total) %>%
  mutate(inf = sum(cut, inf.nc,c,e)) %>%
  dplyr::select(-c(inf.nc, larvae))

########### combine brome and wheat data ###########
sum(comb.brome$total.stems, na.rm = TRUE) +
sum(comb.stubble$total.stems, na.rm = TRUE)

comb.brome %>%
  rbind(comb.stubble) %>%
  na.omit() %>%
  dplyr::select(year, site, location, collection, distance, everything()) %>%
  mutate(location = if_else(location %in% c("ke", "kw"), "k", location),
         location = if_else(location %in% c("nn", "ns"), "n", location)) %>%
  mutate(region = if_else(
    location %in% c("har", "sun", "k", "chs"),
    "moccasin", "big sandy")) %>%
  mutate(plant_type = if_else(distance == "brome", "brome", "wheat")) %>%
  dplyr::select(-distance) %>%
  group_by(year, site, location, collection) %>%
  mutate(p = sum(c, e)) %>% dplyr::select(-c(c, e, inf, uninf)) %>%
  group_by(year, region, collection, plant_type) %>%
  reframe(year, region, collection, plant_type,
          total.stems = sum(total.stems),
          p_mean = mean(p),
          p_sd = sd(p),
          p_se = p_sd/sqrt(total.stems),
          cut_mean = mean(cut),
          cut_sd = sd(cut),
          cut_se = cut_sd/sqrt(total.stems)) %>%
  distinct() %>%
  group_by(year, region, collection, plant_type) %>%
  mutate(ratio = p_mean/cut_mean,
         ratio_se = ratio*sqrt((p_se/p_mean)^2 + (cut_se/cut_mean)^2)) %>%
  filter(collection == "fall") %>%
  ggplot() +
  geom_bar(stat = "identity",
             aes(x = plant_type,
                 y = ratio, 
                 fill = plant_type),
             color = "black") +
  geom_errorbar(aes(x = plant_type,
                    ymin = ratio - ratio_se,
                    ymax = ratio + ratio_se),
                width = 0.15) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = element_blank(),
       y = "Ratio of parasitized stems to cut stems") +
  scale_fill_brewer(palette = "Greys") +
  geom_hline(yintercept = 1, 
             linetype = "dashed",
             color = "black") +
  facet_grid(year~region)
```

```{r, eval = F}

# i want to compare moccasin vs big sandy
joined.stubble %>%
  mutate(region = if_else(
    location %in% c("har", "sun", "k"), 
    "moccasin", "big sandy")) %>%
  ungroup() %>%
  dplyr::select(location, region) %>%
  distinct()

joined.stubble %>%
  ungroup() %>%
  dplyr::select(location) %>%
  distinct()

# table <-
joined.stubble %>%
  mutate(region = if_else(
    location %in% c("har", "sun", "k"), 
    "moccasin", "big sandy")) %>%
  pivot_longer(-c(site, location, distance, year, collection, region),
               names_to = "type",
               values_to = "value") %>%
  distinct() %>%
  rbind(n.inf) %>%  # rbind the brome data
  # filter(!type == "total") %>%  # remove type = "total"
  mutate(type = case_when(
    type == "c" ~ "p",
    type == "e" ~ "p",
    TRUE ~ type)) %>% # replace the e and c variables with p, for simplicity
  ungroup() %>%
  dplyr::select(-site) %>%
  group_by(location, distance, year, collection, type, region) %>%
  filter(type %in% c("cut", "p")) %>%
  mutate(plant_type = ifelse(distance == 5 | distance == 20, "wheat", "brome")) %>%
  ungroup() %>%
  group_by(location, distance, type, region) %>%
  reframe(plant_type, region,
          value = sum(value)) 
  ungroup() %>%
  distinct() %>%
  dplyr::select(-location, -distance) %>%
  group_by(plant_type, type) %>%
  mutate(plant_type, type, 
         count = n(),
         mean = mean(value),
         sd = sd(value),
         se = sd/sqrt(count)) %>%
  ungroup() %>%
  dplyr::select(-value, -count) %>%
  distinct() %>%
  group_by(plant_type) %>%
  reframe(plant_type,
          ratio = lead(mean)/mean,
          se = ratio * sqrt(((se / mean)^2) + (se/mean)^2)) %>%
  na.omit() %>%
  ggplot() +
    geom_bar(stat = "identity",
             aes(x = plant_type,
                 y = ratio, 
                 fill = plant_type),
             color = "black") +
  geom_errorbar(aes(x = plant_type,
                    ymin = ratio - se,
                    ymax = ratio + se),
                width = 0.15) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = element_blank(),
       y = "Ratio of parasitized stems to cut stems") +
  scale_fill_brewer(palette = "Greys")

data <- 
  table %>%
  group_by(plant_type) %>%
  reframe(plant_type,
          ratio = lead(mean)/mean,
          se = sqrt(((se / mean)^2) + (se/mean)^2)) %>%
  na.omit()

# Calculate ratio of p to cut for each plant type
ratios <- aggregate(mean ~ plant_type, data = table[table$type == "p", ], FUN = function(x) x / table[table$type == "cut", "mean"])

# Calculate standard error for each ratio
ratios$se_ratio <- sqrt((table[table$type == "p", "se"] / table[table$type == "p", "mean"])^2 +
                                      (table[table$type == "cut", "se"] / table[table$type == "cut", "mean"])^2)

x <- c("Brome", "Wheat")
ratios <- ratios
se <- ratios$se_ratio

ratio.data.plot <-
data.frame(as.data.frame(x), as.data.frame(data$ratio), se) %>%
  group_by(x) %>%
  mutate(se = se*data.ratio)

ggplot() +
  geom_bar(data = ratio.data.plot,
           aes(x = x,
               y = data.ratio,
               fill = x),
           stat = "identity",
           color = "black") +
  geom_errorbar(data = ratio.data.plot,
                aes(x = x,
                    ymin = data.ratio - se,
                    ymax = data.ratio + se),
                width = 0.15) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = element_blank(),
       y = "Ratio of parasitized stems to cut stems") +
  scale_fill_brewer(palette = "Greys")

```

<!-- Big Sandy -->

<!-- Turn: 3m x 17m = 51$m^2$ -->

<!-- North Road North: 5m x 20m = 100$m^2$ -->

<!-- North Road South: 5m x 20m = 200$m^2$ -->

<!-- Each sample: 1$m^2$ -->

<!-- Wheat field sampled area 4 rows, 10 meters between each row samples taken at 5 and 20 distance from the edge This means the area of the sampled area is 20x4m = 80$m^2$ Each sample we'll count as 1m x 0.5m = 0.5$m^2$ This means 160 possible samples from the sampling area. -->

<!-- Big Sandy NN Brome: 100$m^2$ - number of Wheat: 80$m^2$ -->

<!-- Just need the density per unit area. don't need the total number of p-toids in the entire sampling area. -->

<!-- Just need to standardize it to area so that the sampling size of wheat is the same as brome. These were different in the field. Brome: 1$m^2$ wheat: 0.5$m^2$ -->

<!-- Also need: - weather data to see if there is a drought causing crashes/spikes - NDVI data to show greening of brome compared to wheat late in the season -->

## Citations

```{r save}

# duplicate the wheat data to match up with brome data locations
# we are duplicating some of the stubble data so that we have stubble data
# for the brome sections that are across the road from stubble data

# I don't think I need this for the analysis that I'm doing
# joined.stubble <-
#   joined.stubble %>%
#   bind_rows(joined.stubble %>%
#               filter(location == "n") %>%
#               mutate(location = "ns")) %>%
#   bind_rows(joined.stubble %>%
#               filter(location == "n") %>%
#               mutate(location = "nn")) %>%
#   filter(!location == "n") %>%
#   bind_rows(joined.stubble %>%
#               filter(location == "har") %>%
#               mutate(location = "sun")) 
```

### 2024-05-13 qmd/projects/msu/wss/paper1.tex
commit 5/13 2024

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Paper 1: Smooth brome and parasitoids},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Paper 1: Smooth brome and parasitoids}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\subsection{Introduction}\label{introduction}

\subsection{Materials and Methods}\label{materials-and-methods}

\subsubsection{Weather data and NDVI
analysis}\label{weather-data-and-ndvi-analysis}

\emph{Weather data.} We assessed the long and medium term temperature
and precipitation trends of our field sites using weather data from the
National Oceanic and Atmospheric Administration (NOAA, Silver Spring,
MA, USA). Data for each field site was averaged from three of the
closest weather stations to that location. Data was plotted using R
Studio (R Studio version 2024.04.0+735, R 2024) package `ggplot'
(version 3.4.4) (\citeproc{ref-ggplot}{Wickham 2016}). Data was then fit
using a linear model using the `lm' command using average yearly
precipitation (inches) as the response variable and year as the
predictor.

\emph{NDVI analysis.} We compared the relative greening throughout the
growing season between wheat fields and adjacent \emph{B. inermis} using
the normalized difference vegetation index (NDVI). NDVI is typically
used to assess vegetation health and density, and is calculated from the
visible and near-infared light reflected by vegetation
(\citeproc{ref-Pettorelli2005}{Pettorelli et al. 2005}). Data was
downloaded using Google Earth Engine (Google Inc.~2023, Mountain View,
CA, USA).

\subsubsection{\texorpdfstring{Controlled \emph{C. cinctus} infestation
of \emph{B.
inermis}}{Controlled C. cinctus infestation of B. inermis}}\label{controlled-c.-cinctus-infestation-of-b.-inermis}

\emph{Insects and Cages.} Assessment of \emph{C. cinctus} infestation
and mortality within \emph{B. inermis} were assessed using a 34 x 60 ft
plot at the Arthur H. Post Agronomy Farm (43°38'19.39''N,
116°14'28.86''W), an extension research station of Montana State
University in Bozeman, MT. The cage structure was built using 1-inch PVC
piping with the netting made using 530\(\mu\) Amber Lumite Screen
(BioQuip\(^\circledR\) Products, LLC). Twelve cages were built to
dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected
randomly based on the space available within the plot and arranged in
sets of three.

Wheat stem stubble was collected in Three Forks, MT, USA
(43°38'19.39''N, 116°14'28.86''W) from fields that experienced high
levels of \emph{C. cinctus} infestation and cutting the year prior. Cut
stubble, which contained \emph{C. cinctus} larvae in diapause, were kept
refrigerated between -2°C and 3°C for \textgreater100 days as required
to complete obligatory larval diapause. As needed, stubs were removed
from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz
GladWare® storage containers (Glad®, Oakland, California USA). Once
\emph{B. inermis} stems reached six inches tall, stub containers with
emerging sawflies were added to cages to mimic sawfly infestation
pressure. Sawfly quantity treatments were as follows: high (600 stubs),
low (200 stubs), and control (0 stubs).

\emph{Data Collection.} In late August, \emph{B. inermis} stems were
collected from each cage. Each stem was sliced open using X-Acto® knives
to collect data on infestation, dead larvae, live larvae, and parasitism
at each internode.

\subsubsection{Montana Field Survey}\label{montana-field-survey}

\emph{Stem collection and processing.} We conducted a field survey to
assess \emph{C. cinctus} infestation, larval mortality, and \emph{B.
cephi} and \emph{B. lissogaster} prevalence within \emph{B. inermis} and
adjacent wheat fields. Sites were chosen across 2 counties in
north-central Montana, United States. (Chouteau, Judith Basin), which
consistently experience high \emph{C. cinctus} pressure. Samples were
collected from wheat fields and adjacent \emph{B. inermis} in early July
and late August in 2021, 2022, and 2023 from sites in Big Sandy,
Moccasin, and Amsterdam, MT, USA. Sampling sites were set up as
100\(m^2\) polygons along the edge of adjoining wheat fields. Four
collection squares of 1ft x 1ft were randomly selected within each
polygon during both collection events each year. All stems within each 1
x 1 ft square were collected using a shovel to remove both stem and root
material. Wheat stems were collected at distances of 5 and 20 meters
from the edge of the field. Samples were collected in 4 rows at 10
meters apart. 2 samples were collected in each row at distances of 5 and
20 meters. 1 ft samples were collected at each point.

Wheat and \emph{B. inermis} stems were then returned to Montana State
University, Bozeman, Montana and stored in a 10°C cold wet storage until
dissection. Stems were dissected lengthwise with a fine-bladed scalpel
to determine presence or absence of \emph{C. cinctus} larvae
infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism,
and cutting. Sawfly larvae were identified based on descriptions in
Criddle (1915) and Wallace and McNeal (1996).

\emph{Statistical Analysis.} We used generalized linear mixed models
with binomial errors (logit link) fit using bound optimization by
quadratic approximation, with a maximum of 200,000 iterations, were run
in the \emph{lme4} package in R (\citeproc{ref-lme4}{Douglas Bates,
Bolker, and Walker 2015}) to examine the effects of location and year
(fixed effects) on each of the three response variables: proportion of
stems infested, proportion of stems cut, and proportion os stems
parasitized by \emph{Bracon} spp. '

To better understand the density of \emph{Bracon} spp. within the wheat
and \emph{B. inermis} sampling sites, we converted our units to the
ratio of parasitoids to stem per unit area. Densities of stems are
different when considering \emph{B. inermis} and cultivated wheat or
barley. Using this conversion, we attempted to better understand how the
density of parasitoids (and WSS) is changed by looking at each kind of
plant.

\subsection{Results}\label{results}

\subsubsection{Historical Weather
Analysis}\label{historical-weather-analysis}

We observed a significant linear relationship (\emph{r = 0.1, P = 0.033,
estimate = -0.058}) between average precipitation and year for both Big
Sandy and Moccasin, Montana. This means that for each one year increase,
we are seeing a decrease in 0.05 inches of precipitation.

\includegraphics{paper1_files/figure-latex/both_prcp_plot_show-1.pdf}

In addition, we observed a significant positive linear relationship
(\emph{r = 0.178, P \textless{} 0.05, estimate = 0.024}) between year
and average yearly temperature for Moccasin, MT. This means that each
year, the average daily temperature has increased by 0.02°F.

\includegraphics{paper1_files/figure-latex/moc_temp_plot_show-1.pdf}

\subsubsection{NDVI Analysis of Field
Sites}\label{ndvi-analysis-of-field-sites}

We observed a notable difference in NDVI when comparing adjacent
\emph{B. inermis} and spring wheat. We saw a significant difference in
the July NDVI (\emph{0.846, P \textless{} 0.05}). The \emph{B. inermis}
NDVI remained relatively linear in it's downslope (BROME SLOPE POST
JULY) compared to the wheat field (WHEAT FIELD POST JULY).

\includegraphics{paper1_files/figure-latex/bs_ndvi_plot-1.pdf}

Need to add map showing where NDVI measurments came from

\subsubsection{\texorpdfstring{Controlled \emph{C cinctus} Infestation
of \emph{B.
inermis}}{Controlled C cinctus Infestation of B. inermis}}\label{controlled-c-cinctus-infestation-of-b.-inermis}

\emph{C. cinctus} showed a high ability to infest \emph{B. inermis} in
the controlled test conditions. Averaged across both years, we observed
66.5\% of stems infested for high treatments and 47.3\% of stems for low
treatments. We found strong evidence suggesting that there was a
significant difference between infestation at high and low treatment
levels when holding year constant (\emph{r = 0.83, P \textless{} 0.05}).

Cutting was observed at 5.7\% for the high treatments and 3.9\% for the
low, showing strong evidence for a difference in cutting between high
and low treatment groups (\emph{r = 0.592, P \textless{} 0.05}).

\includegraphics{paper1_files/figure-latex/inf_cut_plot-1.pdf}

\subsubsection{Montana Field Sites}\label{montana-field-sites}

We quantified \emph{C. cinctus} and \emph{Bracon} spp. parasitoid
presence in 6,148 wheat and grass stems across 10 research sites in 2022
and 2023. Infestation by \emph{C. cinctus} within \emph{B. inermis}
varied between collection sites, with the greatest infestation taking
place within our three Big Sandy, MT sampling sites (2023: 65.4\%, 2022:
63.1\%), while the lowest infestation observed was in Moccasin, MT
(2023: 40.8\%, 2022: 60.7\%). Across all sites in Big Sandy and
Moccasin, we observed an average infestation of 57.5\% within \emph{B.
inermis} and 47.6\% within the adjacent wheat.

Parasitoid presence was observed at the highest levels in

\emph{B. inermis} displayed very low levels of \emph{C. cinctus} cutting
when compared to adjacent wheat fields. In Moccasin,

\begin{verbatim}
## [1] 6148
\end{verbatim}

\includegraphics{paper1_files/figure-latex/plot ratios with error bars-1.pdf}

Big Sandy

Turn: 3m x 17m = 51\(m^2\)

North Road North: 5m x 20m = 100\(m^2\)

North Road South: 5m x 20m = 200\(m^2\)

Each sample: 1\(m^2\)

Wheat field sampled area 4 rows, 10 meters between each row samples
taken at 5 and 20 distance from the edge This means the area of the
sampled area is 20x4m = 80\(m^2\) Each sample we'll count as 1m x 0.5m =
0.5\(m^2\) This means 160 possible samples from the sampling area.

Big Sandy NN Brome: 100\(m^2\) - number of Wheat: 80\(m^2\)

Just need the density per unit area. don't need the total number of
p-toids in the entire sampling area.

Just need to standardize it to area so that the sampling size of wheat
is the same as brome. These were different in the field. Brome: 1\(m^2\)
wheat: 0.5\(m^2\)

Also need: - weather data to see if there is a drought causing
crashes/spikes - NDVI data to show greening of brome compared to wheat
late in the season

\subsection*{Citations}\label{citations}
\addcontentsline{toc}{subsection}{Citations}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-lme4}
Douglas Bates, MM, Ben Bolker, and Steve Walker. 2015. {``Fitting Linear
Mixed-Effects Models Using Lme4.''} \emph{Journal of Statistical
Software} 67 (1): 1--48.

\bibitem[\citeproctext]{ref-Pettorelli2005}
Pettorelli, Nathalie, Jon Olav Vik, Atle Mysterud, Jean Michel Gaillard,
Compton J. Tucker, and Nils Chr Stenseth. 2005. {``Using the
Satellite-Derived NDVI to Assess Ecological Responses to Environmental
Change.''} \emph{Trends in Ecology and Evolution}.
\url{https://doi.org/10.1016/j.tree.2005.05.011}.

\bibitem[\citeproctext]{ref-ggplot}
Wickham, Hadley. 2016. \emph{Ggplot2: Elegant Graphics for Data
Analysis}. Springer-Verlag New York.
\url{https://ggplot2.tidyverse.org}.

\end{CSLReferences}

\end{document}

### 2024-05-13 qmd/projects/msu/wss/weatherdata.qmd
commit 5/13 2024

---
title: "Weather Data MT"
author: "Jackson Strand"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, echo = F, message = F}
library(tidyverse)
library(glue)
library(lubridate)
library(ggplot2)
library(maps)
library(readxl)

```

```{r, echo = F, message = F}
inventory_url <- "https://www.ncei.noaa.gov/pub/data/ghcn/daily/ghcnd-inventory.txt"

inventory <- read_table(inventory_url,
           col_names = c("station", "lat", "lon", "variable", "start", "end"))
```

```{r weather data load, echo = F}
area <- c("Big Sandy", "Moccasin", "Bozeman")
lats <- c(48.177540, 47.109558, 45.678871)
longs <- c(-110.114060, -109.884338, -111.060815)

coordinates <- data.frame(area, lats, longs)

my_coordinates <- 
coordinates %>%
  mutate(lats_conv = lats * 2 * pi/360,
         longs_conv = longs * 2 * pi/360) %>%
  ### SELECT AREA HERE ###
  filter(area == "Big Sandy")
  ########################

top_3_stations <- 
  inventory %>%
    dplyr::select(station, lat, lon, start, end) %>%
    mutate(lat_r = lat *2 *pi/360,
           lon_r = lon *2 *pi/360,
           d = 1.609344 * 3963 * acos((sin(lat_r) * sin(my_coordinates$lats_conv)) + cos(lat_r) * cos(my_coordinates$lats_conv) * cos(my_coordinates$longs_conv - lon_r))) %>%
    filter(start < 1960 & end > 2020) %>%
    dplyr::select(-c(start, end)) %>%
    distinct() %>%
    arrange(d) %>%
    top_n(n = -3, d)

station_1 <- 
  top_3_stations %>%
  slice(1) %>%
  pull(station)

station_2 <-
  top_3_stations %>%
  slice(2) %>%
  pull(station)

station_3 <-
  top_3_stations %>%
  slice(3) %>%
  pull(station)

####
my_station <- station_1
####
station_daily <- glue("https://ncei.noaa.gov/pub/data/ghcn/daily/by_station/{my_station}.csv.gz")

local_weather <- 
read_csv(station_daily,
         col_names = c("station", "date", "variable", "value", "a", "b", "c", "d")) %>%
  # filter(station == station_2) %>%
  dplyr::select(station, date, variable, value) %>%
  pivot_wider(names_from = "variable", values_from = "value",
                values_fill = 0) %>%
  dplyr::select(date, TMAX, PRCP, SNOW) %>%
  mutate(date = ymd(date)) %>%
  filter(date > "1950-01-01") %>%
  filter(TMAX > -100)

local_weather <-
local_weather %>%
  mutate(year = lubridate::year(date))
```

```{r plot, echo = F, message = F}
# plot the weather
local_weather %>%
  group_by(year) %>%
  filter(!year == "2024") %>%
  reframe(average_prcp = mean(PRCP, na.rm =TRUE)) %>%
  filter(year < 2024,
         year > 1975) %>%
  ggplot() +
  geom_point(aes(x = year, y = average_prcp)) +
  geom_line(aes(x = year, y = average_prcp)) +
  geom_smooth(aes(x = year, y = average_prcp), 
              method = "loess", 
              se = TRUE) +
  scale_x_continuous(breaks = seq(min(local_weather$year),
                                  max(local_weather$year),
                                  by = 5)) +
  labs(x = NULL,
       title = my_coordinates$area) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

local_weather %>%
  group_by(year) %>%
  reframe(average_prcp = mean(PRCP, na.rm =TRUE)) %>%
  filter(year >= 2010,
         year < 2024) %>%
  ggplot() +
  geom_point(aes(x = year, y = average_prcp)) +
  geom_line(aes(x = year, y = average_prcp)) +
  geom_smooth(aes(x = year, y = average_prcp), 
              method = "loess", 
              se = TRUE) +
  scale_x_continuous(breaks = seq(min(local_weather$year),
                                  max(local_weather$year),
                                  by = 1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(x = NULL,
       y = "Average Precipitation (inches)",
       title = paste0(my_coordinates$area, ", MT", " Average Yearly Precipitation", collapse =""))
```

Let's run a linear model to see if the downward trend is significant.
```{r linear model, echo = F}
local_weather %>%
  filter(year >= 2010,
         year < 2024) %>%
  group_by(year) %>%
  reframe(average_prcp = mean(PRCP, na.rm =TRUE)) %>%
  lm(average_prcp ~ year, data = .) %>%
  summary()
```

```{r map, echo = F}
world_map <- map_data("world")
states <- map_data("state")
counties <- map_data("county") %>%
  filter(region == "montana")

ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_polygon(data = states, aes(x = long, y = lat, group = group),
               fill = NA, color = "black", alpha = 0.5) +
  geom_polygon(data = counties, aes(x = long, y = lat, group = group),
               fill = NA, color = "black", alpha = 0.5) +
  geom_label(data = coordinates, aes(x = longs, y = lats), 
             color = "black", 
             label = area,
             size = 3,
             vjust = -0.3,
             hjust = -0.1,
             label.padding = unit(0.3, "lines")) +
  geom_point(data = coordinates, aes(x = longs, y = lats),
             color = "red") +
  coord_cartesian(xlim = c(-116, -104), ylim = c(44, 49)) +
  theme_bw()

```

```{r}

library(pacman)
rgee_environment_dir = "/opt/miniconda3/envs/rgee_py"

p_load(raster,
       rgdal,
       rgeos,
       gdalUtils,
       sp,
       sf,
       leaflet,
       mapview,
       caret)

p_load(rgee, geojsonio, remotes, reticulate, devtools, googledrive)
install_github("r-spatial/rgee")

library(remotes)
library(rgee)

# set python
reticulate::use_python(rgee_environment_dir, required=T)
rgee::ee_install_set_pyenv(
  py_path = rgee_environment_dir, # Change it for your own Python PATH
  py_env = "rgee_py" # Change it for your own Python ENV
)
Sys.setenv(RETICULATE_PYTHON = rgee_environment_dir)
Sys.setenv(EARTHENGINE_PYTHON = rgee_environment_dir)
```

Big Sandy Data
- Need data from 3 weather stations that are nearby lochy's property
- average the data to get a good representation 
- plot the past 30 years of data
- plot the past 7 years of data

```{r}
library(readxl)
lochy_weather <- 
read_excel("Lochy_weather.xlsx", sheet = "clean2") %>%
  mutate(date = ymd(date)) %>%
  dplyr::select(region, lat, long, date, year, prcp, tmax) %>%
  group_by(year) %>%
  reframe(prcp = sum(prcp * 0.3937 / 2)) +
  mutate(year = as.factor(year))

```

```{r Big Sandy Data, echo = F}
area <- c("Big Sandy", "Moccasin", "Bozeman")
lats <- c(48.177540, 47.098790, 45.678871)
longs <- c(-110.114060, -109.882962, -111.060815)

coordinates <- data.frame(area, lats, longs)

my_coordinates <- 
coordinates %>%
  mutate(lats_conv = lats * 2 * pi/360,
         longs_conv = longs * 2 * pi/360) %>%
  ### SELECT AREA HERE ###
  filter(area == "Big Sandy")
  ########################

top_3_stations <- 
  inventory %>%
    dplyr::select(station, lat, lon, start, end) %>%
    mutate(lat_r = lat *2 *pi/360,
           lon_r = lon *2 *pi/360,
           d = 1.609344 * 3963 * acos((sin(lat_r) * sin(my_coordinates$lats_conv)) + cos(lat_r) * cos(my_coordinates$lats_conv) * cos(my_coordinates$longs_conv - lon_r))) %>%
    filter(start < 1960 & end > 2020) %>%
    dplyr::select(-c(start, end)) %>%
    distinct() %>%
    arrange(d) %>%
    top_n(n = -3, d) %>%
    mutate(miles = d / 1.609344)

# station 1
station_1 <- 
  top_3_stations %>%
  slice(1) %>%
  pull(station)

station_1_data <-
  read_csv(glue("https://ncei.noaa.gov/pub/data/ghcn/daily/by_station/{station_1}.csv.gz")) %>%
           set_names(c("station", "date", "variable", 
                         "value", "a", "b", "c", "d")) %>%
  dplyr::select(station, date, variable, value)

# station 2
station_2 <-
  top_3_stations %>%
  slice(2) %>%
  pull(station)

station_2_data <-
  read_csv(glue("https://ncei.noaa.gov/pub/data/ghcn/daily/by_station/{station_2}.csv.gz")) %>%
           set_names(c("station", "date", "variable", 
                         "value", "a", "b", "c", "d")) %>%
  dplyr::select(station, date, variable, value)

# station 3
station_3 <-
  top_3_stations %>%
  slice(3) %>%
  pull(station)

station_3_data <-
  read_csv(glue("https://ncei.noaa.gov/pub/data/ghcn/daily/by_station/{station_3}.csv.gz")) %>%
           set_names(c("station", "date", "variable", 
                         "value", "a", "b", "c", "d")) %>%
  dplyr::select(station, date, variable, value)

bind_rows(station_1_data,
          station_2_data,
          station_3_data) %>%
  pivot_wider(names_from = "variable", values_from = "value",
                values_fill = 0) %>%
  dplyr::select(station, date, TMAX, PRCP, SNOW) %>%
  mutate(date = ymd(date)) %>%
  filter(date > "1950-01-01") %>%
  filter(TMAX > -100) %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(station, year) %>%
  reframe(average_temp = mean(TMAX),
          average_prcp = mean(PRCP)) %>%
  filter(year < 2024,
         year > 2014) %>%
  left_join(top_3_stations, by = "station") %>%
  mutate(miles = as.factor(round(miles,1))) %>%
  ggplot() +
  geom_point(aes(x = year, y = average_prcp, color = miles)) +
  geom_line(aes(x = year, y = average_prcp, color = miles)) +
  geom_smooth(aes(x = year, y = average_prcp), 
              method = "loess", 
              se = TRUE) +
  scale_x_continuous(breaks = seq(min(local_weather$year),
                                  max(local_weather$year),
                                  by = 1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor = element_blank()) +
  labs(y = "Average Precipitation (inches)",
       x = NULL)

```

```{r}
# ndvi_brome <-
  read_excel("ndvi_fixed.xlsx") %>%
  dplyr::select(NDVI, date) %>%
  mutate(date = ymd(date)) %>%
  mutate(type = "brome") %>%
  bind_rows(ndvi_wheat) %>%
  mutate(year = lubridate::year(date),
         month = lubridate::month(date),
         day = lubridate::day(date)) %>%
  mutate(year = as.factor(year)) %>%
  # group_by(year, month, type) %>%
  # reframe(date, NDVI = mean(NDVI)) %>%
  ungroup() %>%
  filter(year == "2021",
         # NDVI > 0.05,
         month > 3,
         month < 10) %>%
  ggplot() +
  geom_line(aes(x = date, y = NDVI, color = type)) +
  theme_bw()

read_excel("ndvi_brome_nn_2.xlsx") %>%
  dplyr::select(NDVI, date) %>%
  mutate(date = ymd(date),
         year = lubridate::year(date)) %>%
  filter(year == "2023") %>%
  ggplot() +
  geom_line(aes(x = date, y = NDVI))

ndvi_wheat <-
read_excel("ndvi_10years_nn.xlsx") %>%
  dplyr::select(NDVI, date) %>%
  mutate(date = ymd(date)) %>%
  mutate(type = "wheat")

  filter(month > 3,
         month < 10,
         NDVI > 0.05) %>%
  mutate(year = as.factor(year),
         month = as.factor(month)) %>%
  filter(year == "2023")

ndvi_data <-
read_excel("ndvi_fixed.xlsx") %>%
  mutate(type = "brome") %>%
  bind_rows(read_excel("wheat_fixed.xlsx") %>%
              mutate(type = "wheat")) %>%
  dplyr::select(NDVI, date, type) %>%
  mutate(date = ymd(date),
         year = lubridate::year(date),
         month = lubridate::month(date))

ndvi_data %>%
  mutate(subset = "original") %>%
  filter(year == "2020") %>%
  bind_rows(ndvi_data %>%
              filter(year == "2023") %>%
              filter(!(NDVI < 0.05 & month > 4 & month < 10),
                     !(NDVI > 0.15 & month > 7)) %>%
              mutate(subset = "new")) %>%
  filter(subset == "new") %>%
  ggplot() +
  geom_point(aes(x = date, y = NDVI, color = type), 
             alpha = 0.7,
             position = position_jitter(width = 0.001, height = 0.002)) +
  geom_smooth(aes(x = date, y = NDVI, color = type),
              alpha = 0.5) +
  theme_bw() 
```

### 2024-05-13 qmd/projects/msu/wss/weather_station_data.R

@@ -1,0 +1,168 @@
+# this is the code for outputting weather data from big sandy and Moccasin, MT 
+# into an xlsx file
+# 
+
+# inventory - this is the list of all the weather stations in the NOAA database
+inventory_url <- "https://www.ncei.noaa.gov/pub/data/ghcn/daily/ghcnd-inventory.txt"
+
+# label the columns
+inventory <- read_table(inventory_url,
+                        col_names = c("station", "lat", 
+                                      "lon", "variable", 
+                                      "start", "end"))
+
+
+# these are the coordinates for our three field sites
+area <- c("Big Sandy", "Moccasin", "Bozeman", "Havre")
+lats <- c(48.177540, 47.109558, 45.678871, 48.527908)
+longs <- c(-110.114060, -109.884338, -111.060815, 109.684799)
+
+coordinates <- data.frame(area, lats, longs)
+
+### Big Sandy ###
+bs_coordinates <- 
+  coordinates %>%
+  mutate(lats_conv = lats * 2 * pi/360,
+         longs_conv = longs * 2 * pi/360) %>%
+  filter(area == "Havre")
+
+bs_top_3_stations <- 
+  inventory %>%
+  dplyr::select(station, lat, lon, start, end) %>%
+  mutate(lat_r = lat *2 *pi/360,
+         lon_r = lon *2 *pi/360,
+         d = 1.609344 * 3963 * acos((sin(lat_r) * sin(bs_coordinates$lats_conv)) + cos(lat_r) * cos(bs_coordinates$lats_conv) * cos(bs_coordinates$longs_conv - lon_r))) %>%
+  filter(start < 1960 & end > 2023) %>%
+  dplyr::select(-c(start, end)) %>%
+  distinct() %>%
+  arrange(d) %>%
+  top_n(n = -3, d) 
+
+bs_station_1 <- 
+  bs_top_3_stations %>%
+  slice(1) %>%
+  pull(station)
+
+bs_station_2 <-
+  bs_top_3_stations %>%
+  slice(2) %>%
+  pull(station)
+
+bs_station_3 <-
+  bs_top_3_stations %>%
+  slice(3) %>%
+  pull(station)
+
+data <-
+  read_csv(glue("https://ncei.noaa.gov/pub/data/ghcn/daily/by_station/{bs_station_1}.csv.gz"),
+           show_col_types = FALSE,
+           col_names = c("station", "date", 
+                         "variable", "value", "a", "b", 
+                         "c", "d"))
+
+
+bs_weather_data <-
[…]

### 2024-05-17 qmd/projects/msu/wss/convert.Rmd
updated commit, done with paper 1 5/17/24

---
title: "Convert"
author: "Jackson Strand"
date: "`r Sys.Date()`"
output: html_document
---

```{r convert2}
# create a word document version of this document
# first need to copy and paste all of the content from this 
# document into an .RMD file so that I can convert the .RMD
# into a word document
library(rmarkdown)
content <- readLines("paper1.qmd", warn = FALSE)
writeLines(content, "paper1.rmd")
render("paper1.rmd", 
       output_format = "word_document",
       output_file = paste0("JacksonStrand_Paper 1.docx"))
render("paper1.rmd", output_format = "pdf_document",
       output_file = paste0("JacksonStrand_Paper 1.pdf"))

# move the files to the convert folder
file.copy(paste0("JacksonStrand_Paper 1.docx"), 
          paste0("convert/JacksonStrand_Paper 1.docx"), overwrite = TRUE)
file.remove("JacksonStrand_Paper 1.docx")

file.copy(paste0("JacksonStrand_Paper 1.pdf"), 
          paste0("convert/JacksonStrand_Paper 1.pdf"), overwrite = TRUE)
file.remove("JacksonStrand_Paper 1.pdf")
```

### 2024-05-17 qmd/projects/msu/wss/paper1(1).qmd
updated commit, done with paper 1 5/17/24

---
title: "Ecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids"
authors: "Jackson R. Strand, David K. Weaver"
date: "May 14, 2024"
editor_options: 
  chunk_output_type: console
bibliography: proposal.bib
link-citations: true
biblio-style: apalike
cls: 
  max-names: 1
---

This document can be downloaded as either a .docx or .pdf document. Please be aware that conversion may cause the formatting to be altered.

<a href="convert/JacksonStrand_Paper%201.docx" target="_blank">Download .docx</a>

<a href="convert/JacksonStrand_Paper%201.pdf" target="_blank">Download .pdf</a>

## Abstract

## Introduction

Ecologists and farmers frequently hold divergent views regarding the importance of natural habitats within agricultural landscapes, a distinction often overlooked in ecological and conservation discourse [@Tscharntke2016]. While ecologists recognize the role of natural habitats in supporting biodiversity and providing potential ecosystem services like biological pest control, many farmers perceive these habitat remnants as occupying valuable cropland or even as sources of pests. As such, documenting how pest and natural enemy populations interact with agricultural landscapes alongside natural habitat buffers is important in determining the degree of the services or disservices provided.

Numerous polyphagous insect herbivores and their natural enemies rely on a variety of resources found across diverse habitat types within their environment. Depending upon the quality of the resource, different habitat types can serve as sources for increased pest or natural enemy populations. Similarly, habitat types can serve as sinks for decreased populations [@Kennedy2000]. Altering the presence of these resources can have impacts on pest or natural enemy populations and can alter the ecology between the examined species. For example, diversifying the edge cropping of organic Kale (*Brassica oleracea*) using fava beans, fennel, and marigolds results in a diversity in beneficial arthropod communities that help reduce phytophagous herbivory [@DeMorias2023].

Understanding the ways in which landscape habitat diversity impacts the ecological relationship between pest and enemy is crucial for effectively managing pest species in large scale cropping systems. Similarly, understanding how herbivore plant-hosts act as reservoirs for pests or beneficial natural enemies can potentially play a role in stabilizing year-over-year agricultural impact from these pests [@Gordon1944; @Cano2022].

The wheat stem sawfly (WSS), Hymenoptera *Cephus cinctus* Norton, is a grass-mining sawfly native to North America [@Lesieur2016]. Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as the practice became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 [@Criddle1993]. Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [@Holmes1954]. Economic damage manifests in the form of both kernel weight reduction caused by tissue damage [@Beres2011] and stem toppling due to late-season larval girdling [@Holmes1960].

Several parsitoid species attack *C. cinctus* within wild grass hosts [@Davis1955; @Morrill1998], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill 1998]. These two host-specific ectoparasitoids are native to North America. Both parasitoids are bivoltine, with the first generation present in Montana from late June to late July, while the second generation is active in August and September [@Somsen1956]. Adult female *B. cephi* and *B. lissogaster* paralyze *C. cinctus* larvae within the host stem, paralyze the host larvae, and depositing 1 (*B. cephi*) to three (*B. lissogaster*) eggs. *B. cephi* and *B. lissogaster* have been shown to provide biological control via irreplaceable mortality of *C. cinctus* populations [@Buteler2015] and can significantly reduce *C. cinctus* damage and impact on crop yields [@Buteler2008; @Rand2020]. While *C. cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*)[@Wallace1966; @Cockrell2017], infestation levels can vary widely across hosts. Previous research has shown preference for host plants with larger stem diameter [@Ainslie1920; @Bhandari2020].

Previous research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *B. inermis*, *Elymus trachycaulus*, and *Elytrigia intermedium* host *C. cinctus* and Bracon spp. parasitoids quite well [@Peirce2022; @Rand2024]. We find investigating *B. inermis* holds the most potential, as it has been shown to be a quality host for *C. cinctus* while also leading to high larval mortality [@Farstad1945]. However, there is a lack of refined data relating to the total infestation, cutting, and parasitization within *B. inermis* at multiple time points throughout the growing season. Quantificative assessment of potential 'reservoire' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B. inermis* along roadways and within coulles.

Smooth brome (*Bromus inermis* Lyess) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Salesman2011].*B. inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Farstad1945]. Due to its low agricultural value, and propensity to form monocultures and out compete many native grasses, *B. inermis* is categorized as invasive [@Carlson1985; @Willson2000; @Dillemuth2008]. However, *B. inermis* still holds immense ecological value in its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].

Montana's golden triangle, an area of central and northern Montana with intense cereal grain production, is prone to drought and burning via wildfire [@Wilson1923]. As both *C. cinctus* and its host-specific parasitoids are relatively restricted in their annual mobility [@Weaver2005], environmental extremes can have broad impacts on both local pest and natural enemy populations. As with many predator-prey interactions, the *C. cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession. Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinelidae and aphids [@Evans1991].

We examined historical weather data from our field sites in central and northern Montana. We used this data to understand how changes in global climate trends may be impacting *B. cephis* and *B. inermis* populations. In addition we analyzed normalized difference vegetative index (NDVI) at our field sites to understand how *B. inermis* senescence patterns compare to adjacent cultivated wheat.

In this study, we explored the ecological patterns between Montana populations of *C. cinctus* and its braconid parsitoids in relation to *B. inermis* host sites. Our goal, was 1) to understand the degree to which *C. cinctus* infested *B. inermis* stems in relation to neighboring cereal crops, and 2) the way in which braconid parasitoids utilized WSS hosts within *B. inermis* stems. To do so, we conducted two separate studies. In our first study, we conducted controlled *C. cinctus* infestation of *B. inermis* to understand the relative infestation and cutting rate of *C. cinctus* within the *B. inermis* stems. The second study examined rates of cutting and parasitism at field sites in central and northern Montana.

Using these studies, we addressed three questions, 1) what is the total cutting, and thus source rate, of *C. cinctus* from *B. inermis*, 2) can *C. cinctus* larvae within *B. inermis* serve as second generation hosts for *B. cephi* and *B. lissogaster*, and 3) how do *C. cinctus* and parasitoid populations in *B. inermis* compare with those in adjacent cultivated wheat fields.

## Materials and Methods

### Weather data and NDVI analysis

*Weather data.* We assessed the long and medium term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA). Data for each field site was averaged from three of the closest weather stations to that location. Data was plotted using R Studio (R Studio version 2024.04.0+735, R 2024) package 'ggplot' (version 3.4.4) [@ggplot]. Data was then fit using a linear model using the 'lm' command using average yearly precipitation (inches) as the response variable and year as the predictor.

*NDVI analysis.* We compared the relative greening throughout the growing season between wheat fields and adjacent *B. inermis* using the normalized difference vegetation index (NDVI). NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infared light reflected by vegetation [@Pettorelli2005]. Data was downloaded using Google Earth Engine (Google Inc. 2023, Mountain View, CA, USA).

### Controlled *C. cinctus* infestation of *B. inermis*

*Insects and Cages.* Assessment of *C. cinctus* infestation and mortality within *B. inermis* were assessed using a 34 x 60 ft plot at the Arthur H. Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT. The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC). Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.

Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of *C. cinctus* infestation and cutting the year prior. Cut stubble, which contained *C. cinctus* larvae in diapause, were kept refrigerated between -2°C and 3°C for \>100 days as required to complete obligatory larval diapause. As needed, stubs were removed from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz GladWare® storage containers (Glad®, Oakland, California USA). Once *B. inermis* stems reached six inches tall, stub containers with emerging sawflies were added to cages to mimic sawfly infestation pressure. Sawfly quantity treatments were as follows: high (600 stubs), low (200 stubs), and control (0 stubs).

*Data Collection.* In late August, *B. inermis* stems were collected from each cage. Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.

### Montana Field Survey

*Stem collection and processing.* We conducted a field survey to assess *C. cinctus* infestation, larval mortality, and *B. cephi* and *B. lissogaster* prevalence within *B. inermis* and adjacent wheat fields. Sites were chosen across 2 counties in north-central Montana, United States. (Chouteau, Judith Basin), which consistently experience high *C. cinctus* pressure. Samples were collected from wheat fields and adjacent *B. inermis* in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA. Sampling sites were set up as 100$m^2$ polygons along the edge of adjoining wheat fields. Four collection squares of 1ft x 1ft were randomly selected within each polygon during both collection events each year. All stems within each 1 x 1 ft square were collected using a shovel to remove both stem and root material. Wheat stems were collected at distances of 5 and 20 meters from the edge of the field. Samples were collected in 4 rows at 10 meters apart. 2 samples were collected in each row at distances of 5 and 20 meters. 1 ft samples were collected at each point.

Wheat and *B. inermis* stems were then returned to Montana State University, Bozeman, Montana and stored in a 10°C cold wet storage until dissection. Stems were dissected lengthwise with a fine-bladed scalpel to determine presence or absence of *C. cinctus* larvae infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting. Sawfly larvae were identified based on descriptions in Criddle (1915) and Wallace and McNeal (1996).

*Statistical Analysis.* We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion os stems parasitized by *Bracon* spp. '

To better understand the density of *Bracon* spp. within the wheat and *B. inermis* sampling sites, we converted our units to the ratio of parasitoids to stem per unit area. Densities of stems are different when considering *B. inermis* and cultivated wheat or barley. Using this conversion, we attempted to better understand how the density of parasitoids (and WSS) is changed by looking at each kind of plant.

## Results

```{r setup1, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)
```

```{r libraries}
library(markdown)
library(tidyverse)
library(glue)
library(lubridate)
library(ggplot2)
library(maps)
library(readxl)
library(kableExtra)
library(openxlsx)
library(lme4)
```

### Historical Weather Analysis & NDVI

```{r moc_weather}
# moccasin NOAA
moc_weather_noaa <-
read_excel("weather/moccasin_weather_noaa.xlsx") %>%
  filter(station == "USW00024036") %>%
  dplyr::select(date, variable, value, year) %>%
  mutate(date = ymd(date))

# moccasin prcp MAES
moc_prcp_maes <-
  read_excel("weather/moc_weather_maes.xlsx",
             sheet = "prcp") %>%
    dplyr::select(-1, -Annual, -Seasonal) %>%
    pivot_longer(-Year,
                 names_to = "month",
                 values_to = "prcp") %>%
    mutate(Year = lubridate::year(as.Date(Year, format = "%Y"))) %>%
    rename("year" = Year) %>%
    group_by(year) %>%
    reframe(prcp = sum(prcp))

# moccasin temp MAES
moc_temp_maes <-
  read_excel("weather/moc_weather_maes.xlsx",
             sheet = "temp") %>%
    dplyr::select(-1, -Annual, -Seasonal) %>%
    pivot_longer(-Year,
                 names_to = "month",
                 values_to = "temp") %>%
    mutate(Year = lubridate::year(as.Date(Year, format = "%Y"))) %>%
    rename("year" = Year) %>%
    group_by(year) %>%
    reframe(temp = mean(temp))

# past 50 years prcp data
moc_prcp_50 <-
moc_prcp_maes %>%
  filter(year > 1975)

# past 50 years temp data
moc_temp_50 <-
  moc_temp_maes %>%
  filter(year > 1975)
```

```{r bs_weather}
# big sandy
bs_weather <-
read_excel("weather/bigsandy_weather_noaa.xlsx") %>%
  filter(station == "USC00240770") %>%
  dplyr::select(date, variable, value, year, station) %>%
  mutate(date = ymd(date))

# past 50 years prcp data
bs_prcp_50 <-
bs_weather %>%
  filter(variable == "PRCP") %>%
  group_by(year) %>%
  reframe(prcp = mean(value)) %>%
  distinct() %>%
  filter(year > 1975,
         year < 2024,
         prcp < 25)

bs_temp_noaa <-
bs_weather %>%
  filter(variable %in% c("TMAX", "TMIN"),
         value < 100,
         value > -30) %>%
  pivot_wider(names_from = "variable",
              values_from = "value") %>%
  na.omit() %>%
  group_by(date, station) %>%
  mutate(daily = (TMAX - TMIN)/2) %>%
  mutate(month = month(date)) %>%
  # filter(month > 3,
  #       month < 9) %>%
  group_by(year, station) %>%
  reframe(temp = mean(daily))

```

```{r both_prcp}
both_locations_prcp <-
bs_prcp_50 %>%
  mutate(location = "Big Sandy") %>%
  rbind(moc_prcp_50 %>%
          mutate(location = "Moccasin"))

both_prcp_lm <-
both_locations_prcp %>%
  filter(location == "Big Sandy") %>%
  lm(prcp ~ year, data = .) %>%
  summary()

moc_50_lm <-
  moc_prcp_50 %>%
  lm(prcp ~ year, data = .) %>%
  summary()

bs_50_lm <-
  bs_prcp_50 %>%
  lm(prcp ~ year, data = .) %>%
  summary()

calculate_segment <- function(model) {
  intercept <- model$coefficients[1, 1]
  slope <- model$coefficients[2, 1]
  segment_formula <- function(x) intercept + slope * x
  x_start <- 1975
  x_end <- 2024
  y_start <- segment_formula(x_start)
  y_end <- segment_formula(x_end)
  segment_data <- data.frame(x_start, y_start, x_end, y_end)
  return(segment_data)
}

bs_segment <- calculate_segment(bs_50_lm)
moc_segment <- calculate_segment(moc_50_lm)

both_prcp_plot <-
both_locations_prcp %>%
  ggplot() +
  theme_bw() +
  geom_segment(data = bs_segment,
               aes(x = x_start, 
                   y = y_start,
                   xend = x_end, 
                   yend = y_end),
               color = "#F8766D",
               alpha = 0.5) +
  geom_segment(data = moc_segment,
               aes(x = x_start, 
                   y = y_start,
                   xend = x_end, 
                   yend = y_end),
               color = "#00BFC4",
               alpha = 0.5) +
  geom_line(aes(x = year, y = prcp, color = location),
            size = 0.3) +
  geom_smooth(aes(x = year, y = prcp),
              method = "lm",
              se = FALSE,
              color = "black",
              size = 0.5,
              alpha = 0.2) +
  labs(color = "Location (MT, USA)",
       x = NULL,
       y = "Average Yearly Precipitation (in)",
       title = "Average Yearly Precipitation",
       subtitle = "Big Sandy & Moccasin, MT, USA")
```

```{r both_temp}
both_temp <-
moc_temp_maes %>%
  mutate(location = "Moccasin") %>%
  bind_rows(bs_temp_noaa %>%
              select(-station) %>%
              mutate(location = "Big Sandy") %>%
              filter(temp < 50)) %>%
  mutate(temp = ifelse(location == "Big Sandy",
                       (temp - 32) * (5/9) * -1 + 10,
                       (temp - 32) * (5/9))) %>%
  filter(year > 1950)

both_temp_plot <-
both_temp %>%
  na.omit() %>%
  ggplot() +
  geom_line(aes(x = year, y = temp, color = location),
            alpha = 0.5) +
  geom_smooth(aes(x = year, y = temp),
              method = "lm",
              se = FALSE,
              color = "black") +
  geom_smooth(aes(x = year, y = temp, color = location),
              method = "lm",
              se = FALSE,
              size = 0.3) +
  theme_bw() +
  labs(color = "Location (MT, USA)",
       x = NULL,
       y = "Mean Yearly Temperature (°C)",
       title = "Mean Yearly Temperature",
       subtitle = "Moccasin and Big Sandy, MT, USA")

# lm both
both_temp_lm <-
both_temp %>%
  lm(temp ~ year + location, data = .) %>%
  summary()

```

```{r save temp, echo = F}
calculate_segment <- function(model) {
  intercept <- model$coefficients[1, 1]
  slope <- model$coefficients[2, 1]
  segment_formula <- function(x) intercept + slope * x
  x_start <- 1975
  x_end <- 2024
  y_start <- segment_formula(x_start)
  y_end <- segment_formula(x_end)
  segment_data <- data.frame(x_start, y_start, x_end, y_end)
  return(segment_data)
}

# moc_segment <- calculate_segment(moc_lm)

moc_temp_plot <-
moc_temp_maes %>%
  ggplot() +
  theme_bw() +
  geom_line(aes(x = year, y = ((temp - 32) * (5/9))),
            size = 0.4) + 
  geom_smooth(aes(x = year, y = ((temp - 32) * (5/9))),
              method = "lm",
              se = FALSE,
              size = 0.5,
              alpha = 0.2) +
  labs(color = "Location (MT, USA)",
       x = NULL,
       y = "Average Yearly Temperature (°C)",
       title = "Average Yearly Temperature",
       subtitle = "Moccasin, MT, USA")

```

We observed a significant linear relationship (*r = `r round(both_prcp_lm$r.squared,3)`, P = `r round(both_prcp_lm$coefficients[2,4],3)`, estimate = `r round(both_prcp_lm$coefficients[2,1],3)`*) between average precipitation and year for both Big Sandy and Moccasin, Montana. This means that for each one year increase, we are seeing a decrease in 0.05 inches of precipitation. ([Fig. 1](#figure1)). Additionally, we observed a significant positive linear relationship (*r = `r round(both_temp_lm$r.squared,3)`, P \< 0.05, estimate = `r round(both_temp_lm$coefficients[2,1],3)`*) between year and average yearly temperature for Moccasin and Big Sandy, MT. This means that each year, the average daily temperature has increased by `r round(both_temp_lm$coefficients[2,1],2)`°C. ([Fig. 2](#figure2)).

```{r Big_Sandy_NDVI}
lochy <-
read_excel("ndvi/ndvi_2023.xlsx", sheet = "lochy") %>%
  filter(use == "yes") %>%
  mutate(date = ymd(date1)) %>%
  dplyr::select(date,wheat,brome) %>%
  pivot_longer(-date,
               names_to = "plant_type",
               values_to = "ndvi") 
ndvi_model <-
lochy %>%
  filter(date > "2023-07-15") %>%
  lm(ndvi ~ plant_type + date, data = .) %>%
  summary()

label_x <- as.Date("2023-07-20")
label_y <- 0.75
line_x <- as.numeric(as.Date("2023-07-20"))

bs_ndvi_plot <-
lochy %>%
  ggplot() +
  geom_point(aes(x = date, y = ndvi, color = plant_type),
             alpha = 0.5,
             pch = 1,
             size = 0.75) +
  geom_line(aes(x = date, y = ndvi, color = plant_type, linetype = plant_type),
            alpha = 0.5,
            size = 0.5) +
  geom_smooth(aes(x = date, y = ndvi, color = plant_type),
              se = FALSE) +
  geom_vline(xintercept = line_x,
             linetype = "dashed", color = "black") +
  geom_label(aes(x = label_x,
                 y = 0.75),
             label = "Spring wheat harvest",
             hjust = -0.1,
             size = 3) +
  geom_segment(aes(x = as.Date("2023-07-20")+5, 
                   y = 0.75, 
                   xend = as.Date("2023-07-20"), 
                   yend = 0.75)) +
  theme_bw() +
  guides(color = FALSE) +
  labs(x = "2023 Growing Season",
       y = "NDVI",
       linetype = NULL,
       title = "Seasonal NDVI Comparison",
       subtitle = expression(paste(italic("B. inermis"),
                                   " and adjacent spring wheat"))) +
  scale_linetype_discrete(labels = c(expression(italic("B. inermis")),
                                                      "Spring Wheat"))
```

We observed a notable difference in NDVI when comparing adjacent *B. inermis* and spring wheat. We saw a significant difference in the July NDVI (*`r round(ndvi_model$r.squared,3)`, P \< 0.05*). The *B. inermis* NDVI remained relatively linear in it's downslope (BROME SLOPE POST JULY) compared to the wheat field (WHEAT FIELD POST JULY). ([Fig. 3](#figure3))

Need to add map showing where NDVI measurments came from

### Controlled *C. cinctus* Infestation of *B. inermis*

```{r column.order}
column.order <- c("distance", "collection", "p.inf", 
                  "p.cut", "p.par", "poten")

column.order2 <- c("location", "distance", "collection", "p.inf", 
                  "p.cut", "p.par", "poten")

round_based_on_size <- function(x) {
  threshold <- 0.01
  rounded_values <- ifelse(abs(x) > threshold, round(x, 2), round(x, 3))
  return(rounded_values)
}
```

```{r (pf.RAW) post farm n.inf data}
pf.raw.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2022/Post\ Farm\ 2022.xlsx", sheet = "n.inf")
pf.raw.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "n.inf")
```

```{r (pf.inf.23) post farm inf data}
pf.inf.23 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf") %>%
  rename(n.1 = 8, n.2 = 9, n.3 = 10, n.4 = 11,
         n.5 = 12, n.6 = 13, n.7 = 14, n.8 = 15,
         n.9 = 16, n.10 = 17)
```

```{r (pf.comb) post farm data combined}
# 2022
pf.2022.site <-
pf.raw.2022 %>%
  group_by(treatment, site) %>%
  reframe(site, treatment,
            p.inf = sum(iv, ir, ix)/sum(uv, ur, iv, ir, ix),
            cut = sum(cut)/sum(total),
            total.stems = sum(iv, ir, ix, uv, ur)) %>%
  unique() %>%
  pivot_longer(-c(treatment, total.stems, site),
               names_to = "type",
               values_to = "percent") %>%
  mutate(year = "2022")

# 2023
pf.2023.site <-
pf.raw.2023 %>%
  group_by(treatment, site) %>%
  reframe(treatment, site, 
          t.inf = sum(n.iv, n.ir, n.ix),
          p.inf = sum(n.iv, n.ir, n.ix)/sum(n.uv, n.ur, 
                                            n.iv, n.ir, n.ix),
          cut = sum(cut)/sum(total),
          total.stems = sum(n.iv, n.ir, n.ix, n.uv, n.ur)) %>%
  distinct() %>%
  pivot_longer(-c(treatment, total.stems, site),
               names_to = "type",
               values_to = "percent") %>%
  filter(!type == "t.inf") %>%
  arrange(type) %>%
  mutate(year = "2023")

pf.comb <-
  pf.2022.site %>%
  rbind(pf.2023.site) %>%
  group_by(treatment, year, type) %>%
  mutate(count = n(),
         average = mean(percent),
         sd = sd(percent),
         se = (sd/sqrt(count))) %>%
  reframe(treatment, year, type,
          mean = mean(percent),
          sd = mean(sd),
          se = mean(se)) %>%
  unique()

```

```{r (pf.total) pf total stems}
# 2022 totals
pf.2022.total <-
  pf.raw.2022 %>%
  group_by(treatment) %>%
  reframe(total = sum(uv, ur, iv, ir, ix)) %>%
  mutate(year = "2022")

# 2023 totals
pf.2023.total <-
  pf.raw.2023 %>%
  group_by(treatment) %>%
  reframe(total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
  mutate(year = "2023")

pf.total.comb <-
  pf.2022.total %>%
  rbind(pf.2023.total)

```

```{r pf_lm}
raw.23 <-
pf.raw.2023 %>%
  dplyr::select(site, treatment,
                n.iv, n.ir, n.ix, n.ur, n.uv,
                cut) %>%
  rename(iv = n.iv,
         ir = n.ir,
         ur = n.ur,
         ix = n.ix,
         uv = n.uv) %>%
  mutate(year = 2023)

raw.22 <-
pf.raw.2022 %>%
  dplyr::select(site, treatment, 
                ix, ir, iv, ur, uv,
                cut) %>%
  mutate(year = 2022)

pf_inf_lm <-
raw.23 %>%
  rbind(raw.22) %>%
  group_by(site, treatment, year) %>%
  reframe(total = sum(iv, ir, ix, ur, uv),
          inf = sum(iv, ir, ix)) %>%
  lm(inf ~ treatment + year, data = .) %>%
  summary()

pf_cut_lm <-
raw.23 %>%
  rbind(raw.22) %>%
  group_by(site, treatment, year) %>%
  reframe(total = sum(iv, ir, ix, ur, uv),
          inf = sum(iv, ir, ix),
          cut = sum(cut)) %>%
  lm(cut ~ treatment + year, data = .) %>%
  summary()
```

```{r pf_per, eval = F}
raw.23 %>%
  rbind(raw.22) %>%
  group_by(treatment) %>%
  reframe(total = sum(iv, ir, ix, ur, uv),
          inf = sum(iv, ir, ix),
          cut = sum(cut),
          p.inf = inf/total,
          p.cut = cut/inf)
```

*C. cinctus* showed high infestation of *B. inermis* in the controlled test conditions. Averaged across both years, we observed 66.5% of stems infested for high treatments and 47.3% of stems for low treatments. We found strong evidence suggesting that there was a significant difference between infestation at high and low treatment levels when holding year constant (*r = `r round(pf_inf_lm$r.squared,3)`, P \< 0.05*, [Fig. 3](#figure4)).

```{r fig4_inf_cut_plot code}
inf_cut_plot <-
ggplot() +
  facet_wrap(~treatment) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           data = pf.comb %>%
             filter(type == "p.inf"),
           aes(year, mean, fill = "p.inf")) +
  geom_bar(stat = "identity",
           position = position_dodge(),
           data = pf.comb %>%
             filter(type == "cut"),
           aes(year, mean, fill = "cut")) +
  geom_errorbar(data = pf.comb,
                aes(x = year,
                    ymin = mean - se,
                    ymax = mean + se),
                position = position_dodge(0.9),
                width = 0.2,
                alpha = 0.5) +
  theme_bw() +
  labs(x = element_blank(),
       y = "Percentage of stems (%)",
       fill = element_blank()) +
  scale_fill_brewer(palette = "Greys",
                    labels = c("Cut", "Infested"))
```

Cutting was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P < 0.05*).

```{r figure_5_pf_inf_nodes}
pf.y.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2022.xlsx", sheet = "inf")
pf.y.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf")

# DATA FOR THE PLOT
pf.nodes.plot.data <-
pf.y.2023 %>%
  dplyr::select(-c(Name, 'Date Proc')) %>%
  dplyr::select(treatment, 6:15) %>%
  pivot_longer(-c(treatment),
               names_to = "node",
               values_to = "value") %>%
  mutate(node = gsub("n_", "", node)) %>%
  na.omit() %>%
  mutate(year = "2023") %>%
  rbind(pf.y.2022 %>%
          dplyr::select(-c(site1, site2, site12, site3,
                           Name, 'Date Proc')) %>%
          dplyr::select(treatment, 
                        2, 3, 4, 5, 6, 
                        7, 8, 9, 10, 11) %>%
          pivot_longer(-c(treatment),
                       names_to = "node",
                       values_to = "value") %>%
          na.omit() %>%
          mutate(year = "2022")) %>%
  group_by(year, treatment, node) %>%
  reframe(count = n()) %>%
  group_by(year, treatment) %>%
  mutate(per = count/sum(count)) %>%
  mutate(node = as.numeric(node))

# PLOT
pf_nodes_plot <-
  pf.nodes.plot.data %>%
  ggplot() +
  geom_bar(stat = 'identity',
           aes(x = node, y = per, fill = treatment),
           position = position_dodge(),
           color = "black") +
  theme_bw() +
  scale_fill_brewer(palette = "Greys") +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Stem Internode",
       y = "Percent of internodes infested",
       title = expression(paste("Percentage of ",
                                italic("B. inermis "),
                                "Internodes Infested")),
       subtitle = "Controlled infestation") +
  facet_wrap(~year)

pf_nodes_over_4_table <-
pf.nodes.plot.data %>%
  group_by(year, node) %>%
  reframe(count = sum(count)) %>%
  group_by(year) %>%
  mutate(per = count/sum(count)) %>%
  filter(node > 4) %>%
  group_by(year) %>%
  reframe(sum = sum(per))
```

*C. cinctus* infested stems often contained larvae burrowing through multiple nodes. In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through. In 2023, `r round(pf_nodes_over_4_table[2,2],3)*100`% of stems had at least 4 nodes infested [Fig. 5](#figure5)

### Montana Field Sites

```{r stubble data}
# stubble 2023
stubble.2023 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", 
           sheet = "stubble") %>%
  dplyr::select(-c(name,date.proc,"%inf", "%cut")) %>%
  na.omit() %>%
  mutate(year = "2023")

# stubble 2022
stubble.2022 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", 
             sheet = "stubble") %>%
  mutate(year = "2022") %>%
  dplyr::select(-c(name, date.proc, "%inf", "%cut"))

comb.stubble <-
stubble.2023 %>%
  rbind(stubble.2022) %>%
  rename("c" = cocoon,
         "e" = eh,
         "total.stems" = total) %>%
  group_by(site, year, collection) %>%
  mutate(inf = sum(cut, inf.nc,c,e)) %>%
  dplyr::select(-c(inf.nc, larvae)) 
```

```{r brome data}
# brome 2022
raw.2022 <- 
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "n.infested") %>%
  mutate(year = "2022") %>%
  dplyr::select(-c(other, date.proc, name, 
                   par.rate, cut.rate, inf)) %>%
  rename("site" = site.cap)

# brome 2023
raw.2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "n.infested") %>%
  mutate(year = "2023") %>%
  dplyr::select(-c(loc.stub, date.proc, uv, ur, iv, ir, ix, 
                   total.stems, add, inf, a)) %>%
    rename(uv = n.uv,
           ur = n.ur,
           iv = n.iv,
           ir = n.ir,
           ix = n.ix,
           total.stems = n.total)
# combine brome years
comb.brome <- 
raw.2023 %>%
  rbind(raw.2022) %>%
  group_by(site, year, collection) %>%
  mutate(inf = sum(iv, ir, ix),
         uninf = sum(uv, ur),
         cut = sum(cut)) %>%
  dplyr::select(-c(iv, ir, ix, uv, ur)) %>%
  mutate(distance = "brome")
```

We quantified *C. cinctus* and *Bracon* spp. parasitoid presence in 6,148 wheat and grass stems across 10 research sites in 2022 and 2023. Infestation by *C. cinctus* within *B. inermis* varied between collection sites, with the greatest infestation taking place within our three Big Sandy, MT sampling sites (2023: 65.4%, 2022: 63.1%), while the lowest infestation observed was in Moccasin, MT (2023: 40.8%, 2022: 60.7%). Across all sites in Big Sandy and Moccasin, we observed an average infestation of 57.5% within *B. inermis* and 47.6% within the adjacent wheat. Cutting by *C. cintus* was observed at an average rate of 5.59% (se = 0.0164) in *B. inermis* and 47.1% (se = 0.0386) in the adjacent cultivated wheat ([fig. 7](#figure7).

To generate a relative estimate of the number of available *C. cinctus* host larvae within *B. inermis* late in the growing season, we calculated the number of stems that contained at least four or more nodes of infestation. In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of infestation. In Moccasin, the corresponding figure was 66.5%.

```{r y_inf_brome}
brome_inf_2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "y.infested") %>%
  dplyr::select(-c("date.proc")) %>%
  mutate(year = "2023")

brome_inf_2022 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "y.infested") %>%
  dplyr::select(-c("old name", "name", "date.proc", "cut")) %>%
  mutate(year = "2022")
```

```{r per_nodes_inf}
internode.data <-
brome_inf_2023 %>%
  rbind(brome_inf_2022) %>%
  mutate(row_num = row_number()) %>%
  mutate(sample_site = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "North Road",
    location %in% c("kw", "ke", "k") ~ "Kolin Road",
    location %in% c("sun", "har") ~ "Sunset & Harvest",
    location == "g" ~ "Gardiner Road",
    TRUE ~ NA_character_)) %>%
  mutate(region = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "Big Sandy",
    location == "g" ~ "Big Sandy",
    location %in% c("kw", "ke", "k") ~ "Moccasin",
    location %in% c("sun", "har") ~ "Moccasin",
    TRUE ~ NA_character_)) %>%
  dplyr::select(-location) %>%
  mutate(row_num = row_number()) %>%
  pivot_longer(-c(year, site, sample_site, region, 
                  collection, lar, v.r, row_num),
               names_to = "node",
               values_to = "status") 
brome.nodes <-
internode.data %>%
  group_by(region, collection,
           node) %>%
  na.omit() %>%
  reframe(count = n()) %>%
  mutate(node = as.numeric(node)) %>%
  group_by(region, collection) %>%
  mutate(per = count/sum(count)) %>%
  ggplot() +
  geom_bar(stat = "identity",
           aes(x = node, y = per, fill = collection),
           position = position_dodge(),
           color = "black") +
  facet_wrap(~region) +
  scale_x_continuous(breaks = 1:10) +
  theme_bw() +
  scale_fill_brewer(palette = "Greys") +
  labs(x = "Stem Internode",
       y = "Percentage of nodes infested",
       fill = "Collection",
       title = expression(paste("Percentage of ",italic("B. inermis")," Internodes Infested")),
       subtitle = "Big Sandy and Moccasin, MT, USA")

```

```{r plot_data}
plot.data <-
comb.brome %>%
  rbind(comb.stubble) %>%
  na.omit() %>%
  filter(!location %in% c("chs", "b")) %>%
  mutate(sample_site = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "North Road",
    location %in% c("kw", "ke", "k") ~ "Kolin Road",
    location %in% c("sun", "har") ~ "Sunset & Harvest",
    location == "g" ~ "Gardiner Road",
    TRUE ~ NA_character_)) %>%
  mutate(region = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "Big Sandy",
    location == "g" ~ "Big Sandy",
    location %in% c("kw", "ke", "k") ~ "Moccasin",
    location %in% c("sun", "har") ~ "Moccasin",
    TRUE ~ NA_character_)) %>%
  mutate(plant_type = if_else(distance == "brome", "brome", "wheat")) %>%
  dplyr::select(year, region, sample_site, collection, plant_type,
                total.stems, inf, uninf, cut, c, e) %>%
  group_by(site, collection, year) %>%
  mutate(par = sum(c,e)) %>% dplyr::select(-c(c,e)) %>%
  group_by(year, collection, region, sample_site, plant_type, site) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  dplyr::select(-c(uninf)) %>%
  group_by(year, collection, region, sample_site, plant_type, site) %>%
  mutate(total.stems = sum(total.stems),
         count = sum(count),
         per.inf = inf/total.stems,
         per.par = par/inf,
         per.cut = cut/inf) %>%
  distinct() %>%
  pivot_longer(-c(year, collection, region, sample_site, 
                  plant_type, count, total.stems, site, inf, cut, par),
               names_to = "variable",
               values_to = "value") %>%
  dplyr::select(-site) %>%
  group_by(year, collection, region, sample_site, plant_type) %>%
  mutate(total.stems = sum(total.stems),
         count = sum(count),
         inf = sum(inf),
         cut = sum(cut),
         par = sum(par)) %>%
  group_by(year, collection, region, sample_site, plant_type, variable) %>%
  mutate(sd = sd(value),
         value = mean(value),
         se = sd/sqrt(count)) %>%
  ungroup() %>%
  dplyr::select(-site) %>%
  distinct() 
```

```{r infestation_per, eval = F}
comb.brome %>%
  group_by(location, collection, year) %>%
  # filter(collection == "fall") %>%
  reframe(inf = sum(inf)/sum(total.stems)) %>%
  distinct() %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  group_by(collection, year, region) %>%
  reframe(inf = mean(inf)) %>%
  na.omit() %>%
  arrange(desc(inf))
```

```{r paraistism_per, eval = F}
plot.data %>%
  filter(collection == "fall",
         year == "2023") %>%
  ggplot() +
  geom_bar(stat = "identity",
           aes(x = variable, y = value, fill = plant_type),
           position = position_dodge(width = 0.5),
           width = 0.5) +
  geom_errorbar(aes(x = variable, fill = plant_type,
                    ymin = value - se,
                    ymax = value + se),
                position = position_dodge(width = 0.5),
                width = 0.2) +
  facet_wrap(sample_site~region) +
  theme_bw() +
  scale_fill_brewer(palette = "Greys") +
  labs(x = NULL,
       y = "Percentage (%)",
       title = "B. inermis and Adjacent Wheat")

mutate(type2 = if_else(
    type %in% c("p_mean", "p_sd", "p_se"), "p", "cut"),
    stat = str_replace(type, ".*_", "")) %>%
  dplyr::select(-type) %>%
  pivot_wider(values_from = value,
              names_from = stat) %>%
  filter(region == "Big Sandy")

plot.data %>%
  ggplot() +
  geom_bar(stat = "identity",
           aes(x = plant_type, y = mean, 
               fill = type2),
           position = position_dodge()) +
  facet_wrap(~ year,) +
  theme_bw()
```

```{r fig4_field_cut_plot_code}
field_cut_plot_data <-
plot.data %>%
  filter(variable == "per.cut") %>%
  filter(collection == "fall") %>%
  mutate(sample_site = factor(sample_site,
                              levels = c("Sunset & Harvest",
                                         "Kolin Road",
                                         "Gardiner Road",
                                         "North Road")))

mean_values <-
field_cut_plot_data %>%
  group_by(year, plant_type) %>%
  reframe(value = mean(value),
          se = mean(se)) %>%
  group_by(plant_type) %>%
  reframe(value = mean(value),
          se = mean(se))

field_cut_plot <-
field_cut_plot_data %>%
  ggplot() +
  geom_bar(stat = 'identity',
           aes(x = sample_site, 
               y = value, fill = plant_type),
           position = position_dodge(width = 0.5),
           width = 0.5,
           color = "black") +
  geom_errorbar(aes(x = sample_site,
                    group = plant_type,
                    ymin = value - se,
                    ymax = value + se),
                position = position_dodge(width = 0.5),
                width = 0.2) +
  geom_hline(data = field_cut_plot_data %>%
               group_by(year, plant_type) %>%
               reframe(mean_value = mean(value)),
             aes(yintercept = mean_value,
                 linetype = plant_type)) +
  facet_wrap(~year) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = NULL,
       linetype = "Average Cutting Across Sites",
       y = "% of stems cut",
       title = expression(paste("Stem Cutting by ", italic("C. cinctus")," in Central and Northern Montana")),
       subtitle = "Moccasin and Big Sandy, MT, USA",
       fill = "Plant Type") +
  scale_fill_brewer(palette = "Greys") +
  scale_color_brewer(palette = "Greys")

# Your data
field_cut_plot_data <- data.frame(
  year = c("2022", "2022", "2023", "2023"),
  plant_type = c("brome", "wheat", "brome", "wheat"),
  value = c(0.0588, 0.475, 0.0529, 0.468)
)
```

```{r fig8_plot ratios with error bars}
# brome 2022
raw.2022 <- 
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", sheet = "n.infested") %>%
  mutate(year = "2022") %>%
  dplyr::select(-c(other, date.proc, name, 
                   par.rate, cut.rate, inf)) %>%
  rename("site" = site.cap)

# brome 2023
raw.2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "n.infested") %>%
  mutate(year = "2023") %>%
  dplyr::select(-c(loc.stub, date.proc, uv, ur, iv, ir, ix, 
                   total.stems, add, inf, a)) %>%
    rename(uv = n.uv,
           ur = n.ur,
           iv = n.iv,
           ir = n.ir,
           ix = n.ix,
           total.stems = n.total)
# combine brome years
comb.brome <- 
raw.2023 %>%
  rbind(raw.2022) %>%
  group_by(site, year, collection) %>%
  mutate(inf = sum(iv, ir, ix),
         uninf = sum(uv, ur)) %>%
  dplyr::select(-c(iv, ir, ix, uv, ur)) %>%
  mutate(distance = "brome")

# stubble 2023
stubble.2023 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", 
           sheet = "stubble") %>%
  dplyr::select(-c(name,date.proc,"%inf", "%cut")) %>%
  na.omit() %>%
  mutate(year = "2023")

# stubble 2022
stubble.2022 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2022/2022 FS Stem Data (2).xlsx", 
             sheet = "stubble") %>%
  mutate(year = "2022") %>%
  dplyr::select(-c(name, date.proc, "%inf", "%cut"))

comb.stubble <-
stubble.2023 %>%
  rbind(stubble.2022) %>%
  rename("c" = cocoon,
         "e" = eh,
         "total.stems" = total) %>%
  mutate(inf = sum(cut, inf.nc,c,e)) %>%
  dplyr::select(-c(inf.nc, larvae))

########### combine brome and wheat data ###########

ratio_plot_data <-
comb.brome %>%
  rbind(comb.stubble) %>%
  na.omit() %>%
  dplyr::select(year, site, location, collection, distance, everything()) %>%
  mutate(location = if_else(location %in% c("ke", "kw"), 
                            "k", location),
         location = if_else(location %in% c("nn", "ns"), 
                            "n", location)) %>%
  mutate(region = if_else(
    location %in% c("har", "sun", "k", "chs"),
    "moccasin", "big sandy")) %>%
  mutate(plant_type = if_else(distance == "brome", "brome", "wheat")) %>%
  dplyr::select(-distance) %>%
  group_by(year, site, location, collection) %>%
  mutate(p = sum(c, e)) %>% dplyr::select(-c(c, e, inf, uninf)) %>%
  group_by(year, region, collection, plant_type) %>%
  reframe(year, region, collection, plant_type,
          total.stems = sum(total.stems),
          p_mean = mean(p),
          p_sd = sd(p),
          p_se = p_sd/sqrt(total.stems),
          cut_mean = mean(cut),
          cut_sd = sd(cut),
          cut_se = cut_sd/sqrt(total.stems)) %>%
  distinct() %>%
  group_by(year, region, collection, plant_type) %>%
  mutate(ratio = p_mean/cut_mean,
         ratio_se = ratio*sqrt((p_se/p_mean)^2 + (cut_se/cut_mean)^2)) %>%
  filter(collection == "fall")

ratio_plot <-
  ratio_plot_data %>%
  ggplot() +
  geom_bar(stat = "identity",
             aes(x = plant_type,
                 y = ratio, 
                 fill = plant_type),
             color = "black") +
  geom_errorbar(aes(x = plant_type,
                    ymin = ratio - ratio_se,
                    ymax = ratio + ratio_se),
                width = 0.15) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = NULL,
       y = "Ratio of parasitized stems to cut stems",
       title = "Parasitized to Cut Stems Ratios",
       subtitle = "Big Sandy and Moccasin, MT, USA") +
  scale_fill_brewer(palette = "Greys") +
  geom_hline(yintercept = 1, 
             linetype = "dashed",
             color = "black") +
  facet_grid(year~region)
```

```{r}
ratio_plot_data %>%
  dplyr::select(year, region, plant_type, ratio) %>%
  group_by(collection, year, region, plant_type) %>%
  pivot_wider(names_from = "plant_type",
              values_from = "ratio")

lcm <- function(a, b) {
  max_num <- max(a, b)
  while (TRUE) {
    if (max_num %% a == 0 && max_num %% b == 0) {
      return(max_num)
    }
    max_num <- max_num + 1
  }
}

lcm(5, 0.310)
0.310*3.22580645
5*3.22580645
```

We utilized the ratio of parasitized stems to cut stems to better understand the true impact of our *B. inermis* sink/source ecology. The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to 1 cut stem. This means that for every 1 square meter of *B. inermis*, we would observe 16 stems that produced a parasitoid for every 1 stem that produced an adult *C. cinctus*. All three other sample selections also showed positive ratios of parasitoids to cut stems [fig. 8](#figure8).

```{r larvae_status models, eval = F}
brome_2023 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/field Sites/2023/2023\ Field Site\ Data.xlsx", sheet = "y.infested")

brome_2023 %>%
  dplyr::select(-date.proc) %>%
  mutate(row_num = row_number()) %>%
  mutate(sample_site = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "North Road",
    location %in% c("kw", "ke", "k") ~ "Kolin Road",
    location %in% c("sun", "har") ~ "Sunset & Harvest",
    location == "g" ~ "Gardiner Road",
    TRUE ~ NA_character_)) %>%
  mutate(region = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "Big Sandy",
    location == "g" ~ "Big Sandy",
    location %in% c("kw", "ke", "k") ~ "Moccasin",
    location %in% c("sun", "har") ~ "Moccasin",
    TRUE ~ NA_character_)) %>%
  dplyr::select(-location) %>%
  pivot_longer(-c(site, sample_site, region, 
                  collection, lar, v.r, row_num),
               names_to = "node",
               values_to = "status") %>%
  na.omit() %>%
  mutate(row_num = as.factor(row_num)) %>%
  group_by(row_num, site, sample_site, 
           region, collection) %>%
  reframe(count = n()) %>%
  lmer(count ~ collection + (1 | sample_site), 
       data = .) %>%
  summary()

```

```{r larvae_status_stats, eval = F}
data <-
brome_inf_2023 %>%
  rbind(brome_inf_2022) %>%
  mutate(row_num = row_number()) %>%
  mutate(sample_site = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "North Road",
    location %in% c("kw", "ke", "k") ~ "Kolin Road",
    location %in% c("sun", "har") ~ "Sunset & Harvest",
    location == "g" ~ "Gardiner Road",
    TRUE ~ NA_character_)) %>%
  mutate(region = case_when(
    location %in% c("nn", "ss", "ns", "n") ~ "Big Sandy",
    location == "g" ~ "Big Sandy",
    location %in% c("kw", "ke", "k") ~ "Moccasin",
    location %in% c("sun", "har") ~ "Moccasin",
    TRUE ~ NA_character_)) %>%
  dplyr::select(-location) %>%
  mutate(row_num = row_number()) %>%
  pivot_longer(-c(year, site, sample_site, region, 
                  collection, lar, v.r, row_num),
               names_to = "node",
               values_to = "status")

data %>%
  group_by(year, region, sample_site, collection,
           node) %>%
  na.omit() %>%
  reframe(count = n()) %>%
  group_by(year, region, sample_site, collection) %>%
  mutate(per = count/sum(count)) %>%
  filter(node >= 4) %>%
  group_by(year, region, sample_site, collection) %>%
  reframe(per = sum(per)) %>%
  pivot_wider(names_from = collection,
              values_from = per) %>%
  arrange(region, sample_site)
  ggplot() +
  geom_bar(stat = "identity",
           aes(x = year, y = per, fill = collection),
           position = position_dodge())

  na.omit() %>%
  mutate(row_num = as.factor(row_num)) %>%
  group_by(row_num, site, sample_site, 
           region, collection) %>%
  reframe(count = n()) %>%
  rename(nodes.inf = count) %>%
  group_by(sample_site, region, collection, nodes.inf) %>%
  reframe(count2 = n()) %>%
  mutate(nodes.inf = as.factor(nodes.inf)) 
  ggplot() +
  geom_bar(stat = "identity",
           aes(x = collection, y = count2, fill = nodes.inf),
           position = position_dodge()) +
  facet_wrap(~sample_site)

  # how many stems had living larvae in the fall?
data %>%
  group_by(year, region, sample_site, collection,
           node) %>%
  na.omit() %>%
  reframe(count = n()) %>%
  group_by(year, region, sample_site, collection) %>%
  mutate(per = count/sum(count)) %>%
  filter(node >= 4) %>%
  group_by(year, region, sample_site, collection) %>%
  reframe(per = sum(per)) %>%
  ggplot() +
  geom_bar(stat = "identity",
           aes(x = year, y = per, fill = collection),
           position = position_dodge())

data %>%
  dplyr::select(year, sample_site, region, collection, lar) %>%
  distinct() %>%
  group_by(year, sample_site, region, collection, lar) %>%
  reframe(count = n()) %>%
  filter(collection == "fall",
         lar == "a")

  group_by(sample_site, region, collection) %>%
  reframe(count = n()) 
  group_by(sample_site, region, collection) %>%
  mutate(total = sum(count),
         per = count / total) 
  filter(lar %in% c("a", "c", "cut", "e")) %>%
  ggplot() +
  geom_bar(stat = "identity",
           aes(x = sample_site, y = per, fill = lar),
           position = position_dodge()) +
  facet_wrap(~collection)
```

I want to drive home that stems with 3 + nodes infested likely had larave that 
were alive late into the summer. Compare the number of nodes infested in the 
summer to stems that have 3+ infested in the fall.

```{r infestation, eval = F}
comb.brome %>%
  group_by(location, collection, year) %>%
  # filter(collection == "fall") %>%
  reframe(inf = sum(inf)/sum(total.stems)) %>%
  distinct() %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  group_by(collection, year, region) %>%
  reframe(inf = mean(inf)) %>%
  na.omit() %>%
  arrange(desc(inf)) %>%
  group_by(collection) %>%
  reframe(inf = mean(inf))

```

```{r, eval = F}
comb.stubble <-
stubble.2023 %>%
  rbind(stubble.2022) %>%
  rename("c" = cocoon,
         "e" = eh,
         "total.stems" = total) %>%
  group_by(site, year, collection) %>%
  mutate(inf = sum(cut, inf.nc,c,e)) %>%
  dplyr::select(-c(inf.nc, larvae))

comb.stubble %>%
  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
  ungroup() %>%
  reframe(inf = sum(inf)/sum(total.stems))
  group_by(region, collection, year) %>%
  reframe(inf = mean(inf))
```

## Discussion

Assessing the comparative abundance of *C. cinctus* and its associated parasitiods within *B. inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas could potentially serve as origins for both pests and beneficial natural adversaries infiltrating agricultural crops.

<!-- Big Sandy -->

<!-- Turn: 3m x 17m = 51$m^2$ -->

<!-- North Road North: 5m x 20m = 100$m^2$ -->

<!-- North Road South: 5m x 20m = 200$m^2$ -->

<!-- Each sample: 1$m^2$ -->

<!-- Wheat field sampled area 4 rows, 10 meters between each row samples taken at 5 and 20 distance from the edge This means the area of the sampled area is 20x4m = 80$m^2$ Each sample we'll count as 1m x 0.5m = 0.5$m^2$ This means 160 possible samples from the sampling area. -->

<!-- Big Sandy NN Brome: 100$m^2$ - number of Wheat: 80$m^2$ -->

<!-- Just need the density per unit area. don't need the total number of p-toids in the entire sampling area. -->

<!-- Just need to standardize it to area so that the sampling size of wheat is the same as brome. These were different in the field. Brome: 1$m^2$ wheat: 0.5$m^2$ -->

<!-- Also need: - weather data to see if there is a drought causing crashes/spikes - NDVI data to show greening of brome compared to wheat late in the season -->

## Figures

### Figure 1 {#figure1}
```{r FIG1_both_prcp_plot, fig.width = 10, out.width = '100%', fig.align = 'left', fig.cap = "Figure 1: Average yearly precipitation (in) for Big Sandy and Moccasin, Montana. Black trend line signifies averaged negative trend between both locations. Data gathered from NOAA and MSU-ARS. A significant linear relationship (r = 0.1, P = 0.033, estimate = -0.058) was observed between year and average yearly precipitation."}
both_prcp_plot 
```

### Figure 2 {#figure2}
```{r both_temp_plot, fig.width = 10, fig.align = 'left', fig.cap = "Figure 2: Average yearly temperature (°C) in Moccasin and Big Sandy, MT. Positive trendline slop suggests increasing average temperatures of the past 100 years. A significant linear relationship (*r =0.2447, P < 0.05, estimate = 0.028*) was observed between year and average yearly temperature."}
both_temp_plot
```

### Figure 3 {#figure3}
```{r bs_ndvi_plot, fig.align = 'left', fig.cap = "Figure 3: Normalized difference vegetation index (NDVI) of *B. inermis* and adjacent spring wheat field from April 2023 to October 2023. Post harvest linear model indicates a significant difference (P values and stuff) when comparing the *B. inermis* post-harvest slope and the spring wheat post-harvest slope."}
bs_ndvi_plot
```

### Figure 4 {#figure4}
```{r inf_cut_plot, fig.align='left', fig.cap = "Figure 4: Comparison of year and treatment group for controlled infestation of *B. inermis*. We observed a significant difference in cutting between high and low treatment groups (*r = 0.592, P < 0.05*)."}
inf_cut_plot
```

### Figure 5 {#figure5}
```{r figure5_pf_nodes_plot, fig.align='left', fig.cap="Figure 6: Ineternodes infested by *B. cephi* within controlled infestation plots in Bozeman, MT, USA. We found that 61.2% of stems exhibited infestation in at least 5 nodes"}
pf_nodes_plot
```

### Figure 6 {#figure6}
```{r figure6_brome_nodes, fig.align='left', fig.cap="Figure 6: Internodes infested by *B. cephi* within collected *B. inermis* stems in Moccasin and Big Sandy, MT, USA."}
brome.nodes
```

### Figure 7 {#figure7}
```{r figure7_field_cut_plot, fig.align='left', fig.cap="Figure 7: Stem cutting by *C. cinctus*. We compared WSS cutting in *B. inermis* and adjacent wheat fields in Big Sandy and Moccasin, MT, USA. The horizontal lines correspond to yearly mean cutting for each plant type across all sampling sites."}
field_cut_plot
```

### Figure 8 {#figure8}
```{r figure8, fig.align='left', fig.cap="Ratios of parasitized to cut stems in *B. inermis* and adjacent cultivated wheat. Dotted line represents where a 1:1 cut:parasitized ratio lies."}
ratio_plot
```

## Citations

```{r save, eval = F}

# duplicate the wheat data to match up with brome data locations
# we are duplicating some of the stubble data so that we have stubble data
# for the brome sections that are across the road from stubble data

# I don't think I need this for the analysis that I'm doing
# joined.stubble <-
#   joined.stubble %>%
#   bind_rows(joined.stubble %>%
#               filter(location == "n") %>%
#               mutate(location = "ns")) %>%
#   bind_rows(joined.stubble %>%
#               filter(location == "n") %>%
#               mutate(location = "nn")) %>%
#   filter(!location == "n") %>%
#   bind_rows(joined.stubble %>%
#               filter(location == "har") %>%
#               mutate(location = "sun")) 
```

### 2024-05-17 qmd/projects/msu/wss/paper1.rmd
updated commit, done with paper 1 5/17/24

--- 2024-05-17 qmd/projects/msu/wss/paper1(1).qmd
@@
   max-names: 1
+output: 
+  html_document:
+      fig_caption: true
 ---
@@
 
-<a href="convert/JacksonStrand_Paper%201.docx" target="_blank">Download .docx</a>
-
-<a href="convert/JacksonStrand_Paper%201.pdf" target="_blank">Download .pdf</a>
+[Download .docx](convert/Paper1.docx)
+[Download .pdf](convert/Paper1.pdf)
 
@@
 For example, diversifying the edge cropping of organic Kale (*Brassica oleracea*) using fava beans, fennel, and marigolds results in a diversity in beneficial arthropod communities that help reduce phytophagous herbivory [@DeMorias2023].
-
 Understanding the ways in which landscape habitat diversity impacts the ecological relationship between pest and enemy is crucial for effectively managing pest species in large scale cropping systems.
-Similarly, understanding how herbivore plant-hosts act as reservoirs for pests or beneficial natural enemies can potentially play a role in stabilizing year-over-year agricultural impact from these pests [@Gordon1944; @Cano2022].
 
@@
 
-Previous research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *B.
-inermis*, *Elymus trachycaulus*, and *Elytrigia intermedium* host *C.
-cinctus* and Bracon spp.
-parasitoids quite well [@Peirce2022; @Rand2024].
-We find investigating *B.
-inermis* holds the most potential, as it has been shown to be a quality host for *C.
-cinctus* while also leading to high larval mortality [@Farstad1945].
-However, there is a lack of refined data relating to the total infestation, cutting, and parasitization within *B.
-inermis* at multiple time points throughout the growing season.
-Quantificative assessment of potential 'reservoire' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B.
-inermis* along roadways and within coulles.
-
 Smooth brome (*Bromus inermis* Lyess) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Salesman2011].*B.
@@
 As both *C.
-cinctus* and its host-specific parasitoids are relatively restricted in their annual mobility [@Weaver2005], environmental extremes can have broad impacts on both local pest and natural enemy populations.
+cinctus* and its conspecific parasitoids are relatively restricted in their annual mobility [@Weaver2005], environmental extremes can have broad impacts on both local pest and natural enemy populations.
 As with many predator-prey interactions, the *C.
-cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession.
-Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinelidae and aphids [@Evans1991].
+cinctus* and parasitoid populations fluctuate similarly to wolves and moose on Isle Royale [@Peterson1999], with wolf populations spiking following moose populations, leading to a crash in moose, and as such, a crash in wolf.
 

### 2024-05-17 qmd/projects/msu/wss/weather_station_data.R

@@ -24,7 +24,7 @@ bs_coordinates <-
   coordinates %>%
   mutate(lats_conv = lats * 2 * pi/360,
          longs_conv = longs * 2 * pi/360) %>%
-  filter(area == "Havre")
+  filter(area == "Big Sandy")
 
 bs_top_3_stations <- 
   inventory %>%
@@ -36,7 +36,8 @@ bs_top_3_stations <-
   dplyr::select(-c(start, end)) %>%
   distinct() %>%
   arrange(d) %>%
-  top_n(n = -3, d) 
+  top_n(n = -3, d) %>%
+  pull(lat)
 
 bs_station_1 <- 
   bs_top_3_stations %>%
@@ -131,9 +132,9 @@ moc_station_3 <-
   slice(3) %>%
   pull(station)
 
+library(rmarkdown)
 
-
-
+render("paper1.rmd", output_format = "word_document")
 
 
 write.xlsx(moc_weather_data, "moccasin_prcp.xlsx", rowNames = FALSE)

### 2024-05-17 qmd/projects/msu/wss/convert.Rmd
updated website for last time today hopefully 7/17/24

@@
 library(rmarkdown)
-content <- readLines("paper1.qmd", warn = FALSE)
+content <- readLines("paper_1.qmd", warn = FALSE)
 writeLines(content, "paper1.rmd")

### 2024-05-17 qmd/projects/msu/wss/paper_1.qmd
updated website for last time today hopefully 7/17/24

--- 2024-05-17 qmd/projects/msu/wss/paper1(1).qmd
@@
 authors: "Jackson R. Strand, David K. Weaver"
-date: "May 14, 2024"
+date: "May 17, 2024"
 editor_options: 
   chunk_output_type: console
-bibliography: proposal.bib
+bibliography: paper1.bib
 link-citations: true
@@
 Altering the presence of these resources can have impacts on pest or natural enemy populations and can alter the ecology between the examined species.
-For example, diversifying the edge cropping of organic Kale (*Brassica oleracea*) using fava beans, fennel, and marigolds results in a diversity in beneficial arthropod communities that help reduce phytophagous herbivory [@DeMorias2023].
+For example, diversifying the edge cropping of organic Kale (*Brassica oleracea*) using fava beans, fennel, and marigolds results in a diversity in beneficial arthropod communities that help reduce phytophagous herbivory [@DeMorais2023].
 
 Understanding the ways in which landscape habitat diversity impacts the ecological relationship between pest and enemy is crucial for effectively managing pest species in large scale cropping systems.
-Similarly, understanding how herbivore plant-hosts act as reservoirs for pests or beneficial natural enemies can potentially play a role in stabilizing year-over-year agricultural impact from these pests [@Gordon1944; @Cano2022].
+Similarly, understanding how herbivore plant-hosts act as reservoirs for pests or beneficial natural enemies can potentially play a role in stabilizing year-over-year agricultural impact from these pests [@Cano2022].
 
 The wheat stem sawfly (WSS), Hymenoptera *Cephus cinctus* Norton, is a grass-mining sawfly native to North America [@Lesieur2016].
-Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as the practice became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 [@Criddle1993].
-Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [@Holmes1954].
+Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as the practice became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 [@Morrill1996].
+Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [@Holmes1956].
 Economic damage manifests in the form of both kernel weight reduction caused by tissue damage [@Beres2011] and stem toppling due to late-season larval girdling [@Holmes1960].
@@
 Several parsitoid species attack *C.
-cinctus* within wild grass hosts [@Davis1955; @Morrill1998], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill 1998].
+cinctus* within wild grass hosts [@Davis1955; @Morrill1998], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill1998].
 These two host-specific ectoparasitoids are native to North America.
@@
 cinctus* and Bracon spp.
-parasitoids quite well [@Peirce2022; @Rand2024].
+parasitoids quite well [@Peirce2021; @Rand2024].
 We find investigating *B.
@@
 cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession.
-Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinelidae and aphids [@Evans1991].
+Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinelidae and aphids [@Evans1999].
 
@@
 
-```{r}
-ratio_plot_data %>%
-  dplyr::select(year, region, plant_type, ratio) %>%
-  group_by(collection, year, region, plant_type) %>%
-  pivot_wider(names_from = "plant_type",
-              values_from = "ratio")
-
-lcm <- function(a, b) {
-  max_num <- max(a, b)
-  while (TRUE) {
-    if (max_num %% a == 0 && max_num %% b == 0) {
-      return(max_num)
-    }
-    max_num <- max_num + 1
-  }
-}
-
-lcm(5, 0.310)
-0.310*3.22580645
-5*3.22580645
-```
-
 We utilized the ratio of parasitized stems to cut stems to better understand the true impact of our *B.

### 2024-05-17 qmd/projects/msu/wss/paper1.rmd
updated linked papers

--- 2024-05-17 qmd/projects/msu/wss/paper1(1).qmd
@@
 ---
-
+<div class="box">
 This document can be downloaded as either a .docx or .pdf document. Please be aware that conversion may cause the formatting to be altered.
@@
 <a href="convert/JacksonStrand_Paper%201.pdf" target="_blank">Download .pdf</a>
+
+</div>
 
@@
 ### Figure 1 {#figure1}
-```{r FIG1_both_prcp_plot, fig.width = 10, out.width = '100%', fig.align = 'left', fig.cap = "Figure 1: Average yearly precipitation (in) for Big Sandy and Moccasin, Montana.
+```{r FIG1_both_prcp_plot, fig.cap = "Figure 1: Average yearly precipitation (in) for Big Sandy and Moccasin, Montana.
 Black trend line signifies averaged negative trend between both locations.
@@
 ### Figure 2 {#figure2}
-```{r both_temp_plot, fig.width = 10, fig.align = 'left', fig.cap = "Figure 2: Average yearly temperature (°C) in Moccasin and Big Sandy, MT.
+```{r both_temp_plot, fig.cap = "Figure 2: Average yearly temperature (°C) in Moccasin and Big Sandy, MT.
 Positive trendline slop suggests increasing average temperatures of the past 100 years.
@@
 ### Figure 3 {#figure3}
-```{r bs_ndvi_plot, fig.align = 'left', fig.cap = "Figure 3: Normalized difference vegetation index (NDVI) of *B.
+```{r bs_ndvi_plot, fig.cap = "Figure 3: Normalized difference vegetation index (NDVI) of *B.
 inermis* and adjacent spring wheat field from April 2023 to October 2023.

### 2024-05-17 qmd/projects/msu/wss/paper_1.qmd
updated linked documents

@@
 
+<div class="box">
 This document can be downloaded as either a .docx or .pdf document. Please be aware that conversion may cause the formatting to be altered.

### 2024-05-31 qmd/projects/msu/wss/paper1.rmd
updated commit, may 31, 2024

@@
 ---
-title: "Ecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids"
+title: "Agroecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids"
 authors: "Jackson R. Strand, David K. Weaver"
@@
 biblio-style: apalike
-cls: 
-  max-names: 1
+csl: entomological-society-of-america-2.csl
 ---
+
 <div class="box">
@@
 ## Abstract
+
+Wheat stem sawfly, Cephus cinctus Norton (WSS), causes significant damage in cereal crops throughout the Northern Great Plains of North America.
+Bracon cephi and Bracon lissogaster are native WSS parasitoids important in managing WSS outbreaks and damage.
+Smooth Brome (Bromus inermis) grass has been found to be an effective parasitoid sink, and potential trap crop, when grown in areas surrounding wheat fields in Montana.
+As climate changes continue to alter central and northern Montana, WSS will continue to cause immense damage to dry-land cereal crop yields.
+My study observes the utilization of smooth brome in providing year-over-year host refuge for Bracon spp.
+We measured the WSS larval survival rate within controlled smooth brome plots, observing a maximum end-of-year survival of 5.7%.
+We also collected stems from central and northern Montana where we measured the WSS infestation and parasitoid prevalence within cultivated wheat and adjacent smooth brome.
+Montana fields sites showed similar high year-end WSS larval mortality while showcasing high levels of parasitism.
+This research underlines the importance of providing a sustainable ecological buffer for WSS parasitoids as climate conditions continue to change.
 
@@
 Depending upon the quality of the resource, different habitat types can serve as sources for increased pest or natural enemy populations.
-Similarly, habitat types can serve as sinks for decreased populations [@Kennedy2000].
+Similarly, these habitat types can serve as sinks that decrease populations [@Kennedy2000].
 Altering the presence of these resources can have impacts on pest or natural enemy populations and can alter the ecology between the examined species.
-For example, diversifying the edge cropping of organic Kale (*Brassica oleracea*) using fava beans, fennel, and marigolds results in a diversity in beneficial arthropod communities that help reduce phytophagous herbivory [@DeMorais2023].
+For example, diversifying the edge cropping of organic *Brassica* using fava beans, fennel, and marigolds results in a diversity in beneficial arthropod communities that help reduce phytophagous herbivory [@DeMorais2023].
 
@@
 These two host-specific ectoparasitoids are native to North America.
-Both parasitoids are bivoltine, with the first generation present in Montana from late June to late July, while the second generation is active in August and September [@Somsen1956].
+Both parasitoids are bivoltine, with the first generation present in Montana from late-June to late-July, while the second generation is active from mid-July to September [@Somsen1956].
 Adult female *B.
@@
 lissogaster* paralyze *C.
-cinctus* larvae within the host stem, paralyze the host larvae, and depositing 1 (*B.
+cinctus* larvae within the host stem, paralyze the host larvae, and depositing one (*B.
 cephi*) to three (*B.
@@
 While *C.
-cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*)[@Wallace1966; @Cockrell2017], infestation levels can vary widely across hosts.
+cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*) [@Wallace1966; @Cockrell2017], infestation levels can vary widely across hosts.
 Previous research has shown preference for host plants with larger stem diameter [@Ainslie1920; @Bhandari2020].
 
-Previous research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *B.
+Recent research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *B.
 inermis*, *Elymus trachycaulus*, and *Elytrigia intermedium* host *C.
@@
 We find investigating *B.
-inermis* holds the most potential, as it has been shown to be a quality host for *C.
-cinctus* while also leading to high larval mortality [@Farstad1945].
+inermis* holds the most potential, as it has been shown to be a quality initially-accepted host for *C.
+cinctus* oviposition while also leading to high larval mortality later on [@Farstad1945].
 However, there is a lack of refined data relating to the total infestation, cutting, and parasitization within *B.
 inermis* at multiple time points throughout the growing season.
-Quantificative assessment of potential 'reservoire' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B.
-inermis* along roadways and within coulles.
+Quantitative assessment of potential 'reservoire' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B.
+inermis* along roadways and within coulees.
 
@@
 Due to its low agricultural value, and propensity to form monocultures and out compete many native grasses, *B.
-inermis* is categorized as invasive [@Carlson1985; @Willson2000; @Dillemuth2008].
+inermis* is categorized as invasive in some habitats [@Carlson1985; @Willson2000; @Dillemuth2008].
 However, *B.
-inermis* still holds immense ecological value in its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].
+inermis* still holds considerable ecological value in its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].
 
@@
 cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession.
-Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinelidae and aphids [@Evans1999].
-
-We examined historical weather data from our field sites in central and northern Montana.
+Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinellidae and aphids [@Evans1999].
+
+We examined historical weather data from two field sites in central and northern Montana.
 We used this data to understand how changes in global climate trends may be impacting *B.
@@
 In this study, we explored the ecological patterns between Montana populations of *C.
-cinctus* and its braconid parsitoids in relation to *B.
+cinctus* and its braconid parasitoids in relation to *B.
 inermis* host sites.
@@
 
-Using these studies, we addressed three questions, 1) what is the total cutting, and thus source rate, of *C.
-cinctus* from *B.
+Using these studies, we addressed three questions, 1) what is the total cutting heading into overwinter, and thus source rate, of the subsequent *C.
+cinctus* population from *B.
 inermis*, 2) can *C.
 cinctus* larvae within *B.
-inermis* serve as second generation hosts for *B.
+inermis* serve as critical second generation hosts for *B.
 cephi* and *B.
-lissogaster*, and 3) how do *C.
+lissogaster*, and 3) how do annual *C.
 cinctus* and parasitoid populations in *B.
@@
 Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
+There is a negligible population of *C.
+cinctus* at this locaiton.
 The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
@@
 
-*Statistical Analysis.* We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion os stems parasitized by *Bracon* spp.
-'
+*Statistical Analysis.* We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion of stems parasitized by *Bracon* spp.
 
@@
 within the wheat and *B.
-inermis* sampling sites, we converted our units to the ratio of parasitoids to stem per unit area.
+inermis* sampling sites, we converted our units to the ratio of parasitoids to surviving hosts stem per unit area.
 Densities of stems are different when considering *B.
@@
 ```{r moc_weather}
+
 # moccasin NOAA
@@
 We observed a significant linear relationship (*r = `r round(both_prcp_lm$r.squared,3)`, P = `r round(both_prcp_lm$coefficients[2,4],3)`, estimate = `r round(both_prcp_lm$coefficients[2,1],3)`*) between average precipitation and year for both Big Sandy and Moccasin, Montana.
-This means that for each one year increase, we are seeing a decrease in 0.05 inches of precipitation.
+For each one-year increase, we are seeing a decrease in 0.05 inches of precipitation.
 ([Fig.
@@
 Additionally, we observed a significant positive linear relationship (*r = `r round(both_temp_lm$r.squared,3)`, P \< 0.05, estimate = `r round(both_temp_lm$coefficients[2,1],3)`*) between year and average yearly temperature for Moccasin and Big Sandy, MT.
-This means that each year, the average daily temperature has increased by `r round(both_temp_lm$coefficients[2,1],2)`°C.
+Thus, each year, the average daily temperature has increased by `r round(both_temp_lm$coefficients[2,1],2)`°C.
 ([Fig.
@@
 
-Need to add map showing where NDVI measurments came from
+Need to add map showing where NDVI measurements came from
 
@@
 
-*C.
-cinctus* showed high infestation of *B.
+Initially, *C.
+cinctus* heavily infested stems of *B.
 inermis* in the controlled test conditions.
@@
 
-Cutting was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P < 0.05*).
+Stem cutting by mature larvae, leading into winter, was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P < 0.05*).
 
@@
 parasitoid presence in 6,148 wheat and grass stems across 10 research sites in 2022 and 2023.
-Infestation by *C.
+Initial infestation by *C.
 cinctus* within *B.
@@
 cintus* was observed at an average rate of 5.59% (se = 0.0164) in *B.
-inermis* and 47.1% (se = 0.0386) in the adjacent cultivated wheat ([fig.
+inermis* and 47.1% (se = 0.0386) in the adjacent cultivated wheat ([Fig.
 7](#figure7).
@@
 cinctus* host larvae within *B.
-inermis* late in the growing season, we calculated the number of stems that contained at least four or more nodes of infestation.
-In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of infestation.
+inermis* late in the growing season, we calculated the number of stems that contained at least four or more nodes with burrowing injury.
+In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of burrowing injury.
 In Moccasin, the corresponding figure was 66.5%.
@@
   distinct()
+
+```
+
+```{r fig_9_brome_par_plot}
+brome_par_plot <-
+plot.data %>%
+  filter(
+         plant_type == "brome",
+         variable == "per.par") %>%
+  mutate(collection = factor(collection, levels = c("summer", "fall"))) %>%
+  ggplot() +
+  geom_bar(stat = "identity",
+           aes(x = sample_site, y = value, fill = collection),
+           position = position_dodge(),
+           width = 0.5,
+           color = "black") +
+  geom_errorbar(aes(x = sample_site,
+                    group = collection,
+                    ymin = value - se,
+                    ymax = value + se),
+                position = position_dodge(width = 0.5),
+                width = 0.15) +
+  theme_bw() +
+  labs(x = NULL) +
+  scale_fill_brewer(palette = "Greys") +
+  facet_wrap(year~region, scales = "free_x")
+```
+
+```{r brome_par_model}
+brome_par_model_data <-
+comb.brome %>%
+  rbind(comb.stubble) %>%
+  na.omit() %>%
+  filter(!location %in% c("chs", "b")) %>%
+  mutate(sample_site = case_when(
+    location %in% c("nn", "ss", "ns", "n") ~ "North Road",
+    location %in% c("kw", "ke", "k") ~ "Kolin Road",
+    location %in% c("sun", "har") ~ "Sunset & Harvest",
+    location == "g" ~ "Gardiner Road",
+    TRUE ~ NA_character_)) %>%
+  mutate(region = case_when(
+    location %in% c("nn", "ss", "ns", "n") ~ "Big Sandy",
+    location == "g" ~ "Big Sandy",
+    location %in% c("kw", "ke", "k") ~ "Moccasin",
+    location %in% c("sun", "har") ~ "Moccasin",
+    TRUE ~ NA_character_)) %>%
+  mutate(plant_type = if_else(distance == "brome", "brome", "wheat")) %>%
+  dplyr::select(year, region, sample_site, collection, plant_type,
+                total.stems, inf, uninf, cut, c, e) %>%
+  group_by(site, collection, year) %>%
+  mutate(par = sum(c,e)) %>% dplyr::select(-c(c,e)) %>%
+  group_by(year, collection, region, sample_site, plant_type, site) %>%
+  mutate(count = n()) %>%
+  ungroup() %>%
+  dplyr::select(-c(uninf)) %>%
+  group_by(year, collection, region, sample_site, plant_type, site)
+
+brome_par_model <-
+brome_par_model_data %>%
+  filter(plant_type == "brome") %>%
+  lm(par ~ year + region + collection, data = .) %>%
+  summary()
+
 ```
@@
 
-```{r fig4_field_cut_plot_code}
+```{r fig7_field_cut_plot_code}
 field_cut_plot_data <-
@@
 inermis* sink/source ecology.
-The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to 1 cut stem.
-This means that for every 1 square meter of *B.
-inermis*, we would observe 16 stems that produced a parasitoid for every 1 stem that produced an adult *C.
+The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to a single cut stem.
+Thus, for every square meter of *B.
+inermis*, we would observe 16 stems that produced a parasitoid for every 1 stem that was likely to produce an adult *C.
 cinctus*.
-All three other sample selections also showed positive ratios of parasitoids to cut stems [fig.
+All three other sample selections also showed positive ratios of parasitoids to cut stems [Fig.
 8](#figure8).
+
+To understand the seasonal trends of *Bracon* spp use of potential *C.
+cinctus* hosts within *B.
+inermis*, we collected stem samples twice each growing season.
+We collected one pre-harvest, in July, and a second time in September, after all adult *Bracon* spp would be dead.
+We observed a significant difference (*df = 96, r = 0.304, p < 0.05*) between parasitoid presence in infested stems collected in Big Sandy and Moccasin, MT.
 
@@
 
+```{r montana map}
+world_map <- map_data("world")
+states <- map_data("state") %>%
+  filter(region == "montana")
+counties <- map_data("county") %>%
+  filter(region == "montana")
+
+area <- c("Big Sandy", "Moccasin", "Bozeman")
+lats <- c(48.177540, 47.098790, 45.678871)
+longs <- c(-110.114060, -109.882962, -111.060815)
+
+coordinates <- data.frame(area, lats, longs)
+
+montana_map <-
+ggplot() +
+  # geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
+  #              fill = "white", color = "black") +
+  geom_polygon(data = counties, aes(x = long, y = lat, group = group),
+               fill = NA, color = "black", linewidth = 0.2) +
+  geom_polygon(data = states, aes(x = long, y = lat, group = group),
+               fill = NA, color = "black", linewidth = 0.9) +
+  geom_label(data = coordinates, aes(x = longs, y = lats), 
+             color = "black", 
+             label = area,
+             size = 5,
+             vjust = 0,
+             hjust = -0.1,
+             label.padding = unit(0.25, "lines")) +
+  geom_point(data = coordinates, aes(x = longs, y = lats),
+             color = "red") +
+  coord_cartesian(xlim = c(-115.75, -104.5), ylim = c(44.5, 49)) +
+  theme_void() +
+  # theme(panel.grid.major = element_blank(),
+  #       panel.grid.minor = element_blank()) +
+  theme(aspect.ratio = 1/1.8)
+
+```
+
 ## Discussion
@@
 Assessing the comparative abundance of *C.
-cinctus* and its associated parasitiods within *B.
-inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas could potentially serve as origins for both pests and beneficial natural adversaries infiltrating agricultural crops.
-
-<!-- Big Sandy -->
-
-<!-- Turn: 3m x 17m = 51$m^2$ -->
-
-<!-- North Road North: 5m x 20m = 100$m^2$ -->
-
-<!-- North Road South: 5m x 20m = 200$m^2$ -->
-
-<!-- Each sample: 1$m^2$ -->
-
-<!-- Wheat field sampled area 4 rows, 10 meters between each row samples taken at 5 and 20 distance from the edge This means the area of the sampled area is 20x4m = 80$m^2$ Each sample we'll count as 1m x 0.5m = 0.5$m^2$ This means 160 possible samples from the sampling area.
+cinctus* and its associated parasitoids within *B.
+inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas serve as sink or source origins for both pests and beneficial natural adversaries infiltrating agricultural crops.
+As the climate of central and northern Montana continues to change, wheat stem sawfly infestation will continue to impact cereal crop yields.
+In this study, we monitored how *B.
+inermis* may be potentially beneficial in ecologically buffering *Bracon* spp parasitoids by observing *C.
+cinctus* mortality within *B.
+inermis* and within adjacent cereal crops as well as *Bracon* spp prevalence.
+
+### Controlled Infestation
+
+Results from the controlled *B.
+inermis* infestation confirmed high *C.
+cinctus* larval morality prior to overwintering [@Shanower2004; @Buteler2008] with 53% of stems exhibiting *C.
+cinctus* burrowing damage.
+Of those infested stems, a large proportion exhibited larval damage in at least 4 nodes (62.5%).
+While stem collection and dissection did not align with the establishment of a timeline of living larvae (stems dissected in October), we can assert that stems with at least 4 nodes of infestation would have a high probability of being alive, and thus a viable potential parasitoid host, in late July and August of the same year.
+
+While total and per node *B inermis* infestation by *C.
+cinctus* was high, survival to overwintering (cutting) only occurred in 5.7% of *C.
+cinctus* infested stems.
+Our results indicate that *B.
+inermis* serves as a high-quality host for *C.
+cinctus* larvae during the spring and summer months.
+However, there was minimal survival of *C.
+cinctus* larvae in August and September, suggesting that there will be little adult emergence from *B.
+inermis* the following season.
+
+### Montana Field Site Survey
+
+Analysis of weather trends in Big Sandy and Moccasin, MT confirmed that the  patterns in central and northern Montana are changing towards warmer and drier conditions [@Pederson2009].
+We observed an average decrease in 0.058 inches of precipitation per year since 1975, while since 1950, temperatures have experienced an average yearly increase of 0.028°C.
+These changes in precipitation and temperature can have significant impacts on timing of crop production and harvest timing [@Zhu2021].
+Year-over-year populations of the *Bracon* spp parasitoids are particularly sensitive, as their bivoltine nature requires the second yearly generation to find a suitable *C.
+cinctus* host [@Nelson1953; @Runyon2002].
+NDVI analyses exhibited the prolonged senescence period for *B.
+inermis* when compared to adjacent cultivated wheat.
+*C.
+cinctus* larvae are cued via the thinning of the cell membranes during plant senescence to travel to the base of the stem and prepare for obligatory diapause [@Beres2011].
+Because *B.
+inermis* senescence occurs at slower rates compared to wheat or barley, *C.
+cinctus* may remain within the stem later in into the summer and fall, providing potential host larvae for second generation *Bracon* parasitoids.
+
+Previous research has shown *B.
+inermis* to be a quality host of both *C.
+cinctus* and *Bracon* spp parasitoids, but little data is available on the season fluctuation of *B.
+inermis* utilization [@Criddle1922; @Peirce2021; @Rand2024].
+Because we collected stem samples twice during the growing season, we were able to compare parasitoid activity.
+We hoped to understand how parasitism rates changed from July (pre-harvest) and September of each year.
+
+Our results showed that parasitoid presence increase significantly at all sites between pre- and post-harvest collections.
+Given the collection methods, September collection parasitism was inherently a combination of both pre-harvest and post-harvest activity.
+While we cannot ascertain *Bracon* preference for host stems, we can assess that the parasitoids utilize infested *B.
+inermis* stems post-harvest when *C.
+cinctus* is no longer available within wheat stems.
+
+Additionally, we observed a high number of stems exhibiting levels of *C.
+cinctus* damage that point to prolonged larval presence within *B.
+inermis* stems.
+In Big Sandy (76.2%) and Moccasin (66.5%), stems that exhibited four or more nodes of burrowing damage correlates similarly to the assertion that *C.
+cinctus* larvae are available within *B.
+inermis* stems past wheat harvest.
+Most importantly, in years of drought or increased temperatures where wheat senescence and harvest may occur earlier in the summer.
+
+We utilized parasitism to cut stem ratios to understand the relative impact of *B.
+inermis* host sources on overall field parasitism.
+Assessing overall numbers of insects can potentially diminish the apparent importance of *B.
+inermis*, as per square meter stem counts in cultivated wheat are denser than *B.
+inermis* counts within coulees and along roadsides *B.
+inermis*.
+Our findings of significantly greater ratios of parasitized stems to cut stems within *B.
+inermis* when compared to adjacent when indicates these field-side areas provide greater parasitoid source per square meter when compared to cultivated fields.
+This emphasizes the significance of accounting for habitat diversity to sustain parasitoid populations over successive years, as well as recognizing the crucial role of edge-row areas in maximizing parasitoid effectiveness in minimizing *C.
+cinctus* damage.
+
+<!-- Overall, this work suggests that maintaining and supporting field-edge areas containing *B.
+inermis* provides an unmistakable benefit to *C.
+cinctus* control in cereal crop production.
+The characteristics of *B.
+inermis* make it uniquely beneficial in both as a sink for *C.
+cinctus* larvae, and as a source for *Bracon* adults.
+*B.
+inermis* is a common grass found throughout the golden triangle of Montana and other wheat cultivation areas [@Bolwahn2011].
 -->
-
-<!-- Big Sandy NN Brome: 100$m^2$ - number of Wheat: 80$m^2$ -->
-
-<!-- Just need the density per unit area. don't need the total number of p-toids in the entire sampling area. -->
-
-<!-- Just need to standardize it to area so that the sampling size of wheat is the same as brome. These were different in the field. Brome: 1$m^2$ wheat: 0.5$m^2$ -->
-
-<!-- Also need: - weather data to see if there is a drought causing crashes/spikes - NDVI data to show greening of brome compared to wheat late in the season -->
 
@@
 
+![](Images/brome_images.jpg)
+
 ### Figure 4 {#figure4}
-```{r inf_cut_plot, fig.align='left', fig.cap = "Figure 4: Comparison of year and treatment group for controlled infestation of *B.
+```{r inf_cut_plot, fig.cap = "Figure 4: Comparison of year and treatment group for controlled infestation of *B.
 inermis*.
@@
 ### Figure 5 {#figure5}
-```{r figure5_pf_nodes_plot, fig.align='left', fig.cap="Figure 6: Ineternodes infested by *B.
+```{r figure5_pf_nodes_plot, fig.cap="Figure 6: Ineternodes infested by *B.
 cephi* within controlled infestation plots in Bozeman, MT, USA.
@@
 ### Figure 6 {#figure6}
-```{r figure6_brome_nodes, fig.align='left', fig.cap="Figure 6: Internodes infested by *B. cephi* within collected *B. inermis* stems in Moccasin and Big Sandy, MT, USA."}
+```{r figure6_brome_nodes, fig.cap="Figure 6: Internodes infested by *B. cephi* within collected *B. inermis* stems in Moccasin and Big Sandy, MT, USA."}
 brome.nodes
@@
 ### Figure 7 {#figure7}
-```{r figure7_field_cut_plot, fig.align='left', fig.cap="Figure 7: Stem cutting by *C.
+```{r figure7_field_cut_plot, fig.cap="Figure 7: Stem cutting by *C.
 cinctus*.
@@
 ### Figure 8 {#figure8}
-```{r figure8, fig.align='left', fig.cap="Ratios of parasitized to cut stems in *B. inermis* and adjacent cultivated wheat. Dotted line represents where a 1:1 cut:parasitized ratio lies."}
+```{r figure8, fig.cap="Ratios of parasitized to cut stems in *B. inermis* and adjacent cultivated wheat. Dotted line represents where a 1:1 cut:parasitized ratio lies."}
 ratio_plot
+```
+
+### Figure 9 {#figure9}
+```{r figure9, fig.align = 'left', fig.cap="Comparison of parasitoid presence in *B. inermis* between July and September collections."}
+brome_par_plot 
+```
+
+### Figure 10 {#figurex}
+```{r figure10, fig.cap="Location of Montana, USA field sites.
+Controlled *B.
+inermis* infestation site was in Bozeman, while field sites were located in centrally location Moccasin, MT and northern Big Sandy, MT."}
+montana_map +
+  labs(title = "Montana, USA")
 ```

### 2024-06-06 index.qmd
update commit 6/6

@@
 
+## Hosted Events
+<div class="box">
+**Student Organized Conferences**
+
+1. 2024 LRES-PSPP Student Research Symposium  
+    Role: Primary event organizer & co-emcee  
+    Location: Montana State University, Bozeman, MT  
+    Date: April 18, 2024  
+    Hyperlink: [2024 Symposium](https://lres-research-symposium.weebly.com/?c=mkt_w_chnl:aff_geo:all_prtnr:sas_subprtnr:1538097_camp:brand_adtype:txtlnk_ag:weebly_lptype:hp_var:358504&sscid=61k8_2eb95&utm_source=ShareASale){target="_blank"}
+
+2. 2023 LRES Colloquium  
+    Role: Treasurer  
+    Location: Montana State University, Bozeman, MT  
+    Date: April 13, 2023
+
+3. 2022 LRES Colloquium  
+    Role: Treasurer  
+    Location: Montana State University, Bozeman, MT  
+    Date: April 29, 2022
+</div>
+
 ## Service & Memberships

### 2024-06-06 qmd/helpful.qmd
updated commit, done with paper 1 5/17/24
update commit 6/6

@@
 ---
-
-Below I have put many of the links that helped me.
 
@@
 
-Watershed volunteering
-https://youtu.be/65MAB0qRtAE?si=beO8jD5t5SOJzooF
+## R Statistics
+[Mixed Effects Models](https://meghan.rbind.io/blog/2022-06-28-a-beginners-guide-to-mixed-effects-models)
+
+## Citations
+I use quarto to generate my website reports. Here are a couple of resources to do the same.
+
+- [Zotero Citations Repository](https://www.zotero.org/styles){target="_blank"}  
+-
+
+[Watershed volunteering](https://youtu.be/65MAB0qRtAE?si=beO8jD5t5SOJzooF){target="_blank"}

### 2024-06-06 qmd/projects/msu/wss/JacksonStrand_Paper 1.tex
update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Agroecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Agroecological Importance of Smooth Brome in Managing Wheat Stem
Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids}
\author{}
\date{\vspace{-2.5em}2024-06-06}

\begin{document}
\maketitle

This document can be downloaded as either a .docx or .pdf document.
Please be aware that conversion may cause the formatting to be altered.

Download .docx

Download .pdf

\subsection{Abstract}\label{abstract}

Wheat stem sawfly, \emph{Cephus cinctus} Norton (WSS), causes
significant damage in cereal crops throughout the Northern Great Plains
of North America. Bracon cephi and Bracon lissogaster are native WSS
parasitoids important in managing WSS outbreaks and damage. Smooth Brome
(Bromus inermis) grass has been found to be an effective parasitoid
sink, and potential trap crop, when grown in areas surrounding wheat
fields in Montana. As climate changes continue to alter central and
northern Montana, WSS will continue to cause immense damage to dry-land
cereal crop yields. My study observes the utilization of smooth brome in
providing year-over-year host refuge for Bracon spp. We measured the WSS
larval survival rate within controlled smooth brome plots, observing a
maximum end-of-year survival of 5.7\%. We also collected stems from
central and northern Montana where we measured the WSS infestation and
parasitoid prevalence within cultivated wheat and adjacent smooth brome.
Montana fields sites showed similar high year-end WSS larval mortality
while showcasing high levels of parasitism. This research underlines the
importance of providing a sustainable ecological buffer for WSS
parasitoids as climate conditions continue to change.

\subsection{Introduction}\label{introduction}

Ecologists and farmers frequently hold divergent views regarding the
importance of natural habitats within agricultural landscapes, a
distinction often overlooked in ecological and conservation discourse
(\citeproc{ref-Tscharntke2016}{Tscharntke et al. 2016}). While
ecologists recognize the role of natural habitats such as field border
vegetation in supporting biodiversity and providing potential ecosystem
services like biological pest control, many farmers perceive these
semi-arid habitat remnants as occupying valuable cropland or even as
sources of pests. As such, documenting how pest and natural enemy
populations interact with agricultural landscapes alongside semi-natural
habitat buffers on the crop periphery is important in determining the
degree of the services or disservices provided.

Numerous polyphagous insect herbivores and their natural enemies rely on
a variety of resources found across diverse habitat types within their
environment. Depending upon the quality of the resource, different
habitat types can serve as sources for increased pest or natural enemy
populations. Similarly, these habitat types can serve as sinks that
decrease populations (\citeproc{ref-Kennedy2000}{Kennedy and Storer
2000}). Altering the presence of these resources can have impacts on
pest or natural enemy populations and can alter the ecology between the
examined species. For example, diversifying the edge cropping of organic
\emph{Brassica} using fava beans, fennel, and marigolds results in a
diversity in beneficial arthropod communities that help reduce
phytophagous herbivory (\citeproc{ref-DeMorais2023}{Morais et al.
2023}).

Understanding the ways in which landscape habitat diversity impacts the
ecological relationship between pest and enemy is crucial for
effectively managing pest species in large scale cropping systems.
Similarly, understanding how herbivore plant-hosts act as reservoirs for
pests or beneficial natural enemies can potentially play a role in
stabilizing year-over-year agricultural impact from these pests
(\citeproc{ref-Cano2022}{Cano et al. 2022}).

The wheat stem sawfly (WSS), Hymenoptera \emph{Cephus cinctus} Norton,
is a grass-mining sawfly native to North America {[}Criddle
(\citeproc{ref-Criddle1922}{1922}); Ainslie1929; Wallace and McNeal
(\citeproc{ref-Wallace1966}{1966}); Lesieur et al.
(\citeproc{ref-Lesieur2016}{2016}){]}. Originally adept at utilizing
wild grass hosts, the species became a major pest to cultivated cereals
as this resource became more prevalent following European settlement,
becoming a notable pest of winter wheat in 1990 in Montana
(\citeproc{ref-Morrill1996}{Morrill and Kushnak 1996}), and more
recently in Colorado (\citeproc{ref-Cockrell2021}{Cockrell et al.
2021}). Climate warming has the potential to further expand the range of
this pest both northward and southward (\citeproc{ref-Olfert2019}{Olfert
et al. 2019}, \citeproc{ref-Weaver2023}{Weaver 2023}). Adult female WSS
use their saw-like ovipositor to lay their eggs in the lumen of stems,
where the eggs hatch, leading to larval feeding of parenchyma tissue
{[}Roemhild1954; Holmes and Farstad
(\citeproc{ref-Holmes1956}{1956}){]}. Economic damage manifests in the
form of both kernel weight reduction caused by tissue damage {[}Beres et
al. (\citeproc{ref-Beres2011a}{2011}); Delaney2010{]} and stem toppling
due to late-season larval girdling (\citeproc{ref-Holmes1960}{Holmes and
Peterson 1960}, \citeproc{ref-Nansen2005a}{Nansen et al. 2005},
\citeproc{ref-McCullough2020}{McCullough et al. 2020}).

Several parasitoid or predatory species attack \emph{C. cinctus} within
wild grass hosts Morrill et al. (\citeproc{ref-Morrill2001}{2001}),
however, two species, \emph{Bracon cephis} (Gahan) and \emph{Bracon
lissogaster} Muesebeck (Hymenoptera: Braconidae), have been observed
targeting WSS within cultivated cereals
(\citeproc{ref-Nelson1953}{Nelson and Farstad 1953},
\citeproc{ref-Morrill1998}{Morrill et al. 1998},
\citeproc{ref-Runyon2001a}{Runyon et al. 2001},
\citeproc{ref-Runyon2001b}{Runyon 2001}). These two host-specific
ectoparasitoids are native to North America. Both parasitoids are
bivoltine, with the first generation present in Montana from late-June
to late-July, while the second generation is active from mid-July to
September (\citeproc{ref-Gahan1918}{Gahan 1918},
\citeproc{ref-Somsen1956}{Somsen and Luginbill 1956},
\citeproc{ref-Davis2013}{Davis 2013}). Adult female \emph{B. cephi} and
\emph{B. lissogaster} inject a paralytic toxin into a \emph{C. cinctus}
larvae within the host stem, paralyzing the host larvae, and depositing
from a single (\emph{B. cephi}) to 1-3 (\emph{B. lissogaster}) eggs.
\emph{B. cephi} and \emph{B. lissogaster} have been shown to provide
biological control via irreplaceable mortality of \emph{C. cinctus}
populations (\citeproc{ref-Peterson2011}{Peterson et al. 2011},
\citeproc{ref-Buteler2015}{Buteler et al. 2015}) and can significantly
reduce \emph{C. cinctus} damage and impact on crop yields
(\citeproc{ref-Buteler2008}{Buteler et al. 2008},
\citeproc{ref-Bekkerman2018}{Bekkerman and Weaver 2018},
\citeproc{ref-Adhikari2019}{Adhikari et al. 2019},
\citeproc{ref-Rand2020}{Rand et al. 2020}). While \emph{C. cinctus} has
a wide range of viable host plants, including native and non-native
grasses and cultivated row-crops such as wheat (\emph{Triticum aestivum
L.}) and barley (\emph{Hordeum vulgare})
(\citeproc{ref-Wallace1966}{Wallace and McNeal 1966},
\citeproc{ref-Cockrell2017}{Cockrell et al. 2017},
\citeproc{ref-Achhami2020}{Achhami et al. 2020}), infestation levels can
vary widely across hosts. Previous research has shown preference for
host plants with larger stem diameter
(\citeproc{ref-Ainslie1920}{Ainslie 1920},
\citeproc{ref-Buteler2009}{Buteler et al. 2009},
\citeproc{ref-Bhandari2020}{Bhandari 2020}).

Recent research has shown natural non-cultivated grasses such as
\emph{Agropyron cristatum}, \emph{Bromus techtorum}, \emph{Bromus
inermis}, \emph{Elymus trachycaulus}, and \emph{Thinopyrum intermedium}
host \emph{C. cinctus} and most also host Bracon spp. parasitoids quite
well (\citeproc{ref-Mendoza2006}{Perez-Mendoza and Weaver 2006},
\citeproc{ref-Peirce2021}{Peirce et al. 2021},
\citeproc{ref-Rand2024}{Rand et al. 2024}). We find investigating
\emph{B. inermis} holds the most potential, as it has been shown to be a
quality initially-accepted host for \emph{C. cinctus} oviposition while
also leading to high larval mortality later on
(\citeproc{ref-Seamans1928}{Seamans 1928},
\citeproc{ref-Farstad1945}{Farstad and Jacobson 1945}). However, there
is a lack of detailed relating to the total infestation, cutting, and
parasitism within \emph{B. inermis} at multiple time points throughout
the growing season. Quantitative assessment of potential `reservoir'
characteristics and local spatial information could prove useful in
fully understanding the ecological impacts of maintaining \emph{B.
inermis} along roadways and within coulees.

Smooth brome (\emph{Bromus inermis} Leyss) is a Eurasian cool-season
rhizomatous grass introduced to the United States in 1884
(\citeproc{ref-Bolwahn2011}{Salesman and Thomsen 2011}).\emph{B.
inermis} was originally thought of as potentially useful as a trap crop
in order to combat the growing WSS problem with cultural control
practices (\citeproc{ref-Criddle1922}{Criddle 1922},
\citeproc{ref-Farstad1945}{Farstad and Jacobson 1945},
\citeproc{ref-Beres2011a}{Beres et al. 2011}). Due to its low
agricultural value, and propensity to form monocultures and out compete
many native grasses, \emph{B. inermis} is categorized as invasive in
some habitats (\citeproc{ref-Carlson1985}{Carlson and Newell 1985},
\citeproc{ref-Willson2000}{Willson and Stubbendieck 2000},
\citeproc{ref-Dillemuth2008}{Dillemuth et al. 2008}). However, \emph{B.
inermis} still holds considerable ecological and potential economic
value due to its late senescence and WSS antibiosis properties
(\citeproc{ref-Otfinowski2006}{Otfinowski et al. 2006},
\citeproc{ref-Rand2024}{Rand et al. 2024}).

Montana's golden triangle, an area of central and northern Montana with
intense cereal grain production, is prone to drought and burning via
wildfire (\citeproc{ref-Wilson1923}{Wilson 1923},
\citeproc{ref-Adhikari2019}{Adhikari et al. 2019}). As both \emph{C.
cinctus} and its host-specific parasitoids are relatively restricted in
their annual mobility (\citeproc{ref-Weaver2005}{\textbf{Weaver2005?}}),
environmental extremes can have broad impacts on both local pest and
natural enemy populations. As with many predator-prey interactions, the
\emph{C. cinctus} and parasitoid populations fluctuate, with crashes and
spikes experienced in close succession and through ineteraction with
drought (\citeproc{ref-Holmes1963}{Holmes et al. 1963},
\citeproc{ref-Holmes1982}{Holmes 1982}). Similar examples of these
classical population dynamics patterns can be found in wolves and moose
on Isle Royale (\citeproc{ref-Peterson1999}{Peterson 1999}), and
Coccinellidae and aphids (\citeproc{ref-Evans1999}{Evans 1999}), among
others.

We examined historical weather data from two field locations (10 sites
overall) in central and northern Montana. We used this data to
understand how changes in global climate trends may be impacting
\emph{C. cinctus}, \emph{B. cephi} and \emph{B. lissogaster} interact
with \emph{B. inermis} populations. In addition we analyzed normalized
difference vegetative index (NDVI) at our field sites to understand how
\emph{B. inermis} senescence patterns compare to adjacent cultivated
wheat fields.

In this study, we explored the ecological patterns between Montana
populations of \emph{C. cinctus} and its associated braconid parasitoids
in relation to \emph{B. inermis} host sites. Our goal, was 1) to
understand the degree to which \emph{C. cinctus} infested \emph{B.
inermis} stems in relation to neighboring cereal crops, and 2) the way
in which braconid parasitoids utilized WSS hosts available within
\emph{B. inermis} stems. To do so, we conducted two separate studies. In
our first study, we conducted controlled \emph{C. cinctus} infestation
of \emph{B. inermis} to understand the relative infestation and cutting
rate of \emph{C. cinctus} within the \emph{B. inermis} stems in the
absence of parasitoids. The second study examined rates of cutting and
parasitism in wheat crops and their surroundings at field sites in
central and northern Montana.

Using these studies, we addressed three questions, 1) what is the total
\emph{C. cinctus}-caused stem cutting heading into overwinter - and thus
the source rate, of the subsequent \emph{C. cinctus} population in
\emph{B. inermis}? 2) can \emph{C. cinctus} larvae within \emph{B.
inermis} serve as critical second generation hosts for \emph{B. cephi}
and \emph{B. lissogaster}? and 3) how do annual \emph{C. cinctus} and
parasitoid populations in \emph{B. inermis} compare with those in the
adjacent cultivated wheat fields?

\subsection{Materials and Methods}\label{materials-and-methods}

All data for experiments was analyzed using R Studio (R Studio version
2024.04.0+735, R 2024).

\subsubsection{Weather data and NDVI
analysis}\label{weather-data-and-ndvi-analysis}

\emph{Weather data.} We assessed the medium- and long-term temperature
and precipitation trends of our field sites using weather data from the
National Oceanic and Atmospheric Administration (NOAA, Silver Spring,
MA, USA). Data for each field site was averaged from three of the
closest weather stations to that location. Data was plotted using R
package `ggplot' (version 3.4.4) (\citeproc{ref-ggplot}{Wickham 2016}).
Data was then fit using a linear model using the `lm' command using
average yearly precipitation (inches) as the response variable and year
as the predictor.

\emph{NDVI analysis.} We compared the relative greening throughout the
growing season between wheat fields and adjacent \emph{B. inermis} using
the normalized difference vegetation index (NDVI). NDVI is typically
used to assess vegetation health and density, and is calculated from the
visible and near-infrared light reflected by vegetation
(\citeproc{ref-Pettorelli2005}{Pettorelli et al. 2005}). NDVI data
measurements were directly downloaded using Google Earth Engine (Google
Inc.~2023, Mountain View, CA, USA). Similar to our weather data
analysis, NDVI data was modeled using the linear modeling package `lm'
in R to discern differences in NDVI measurments over time.

\subsubsection{\texorpdfstring{Controlled \emph{C. cinctus} infestation
of \emph{B.
inermis}}{Controlled C. cinctus infestation of B. inermis}}\label{controlled-c.-cinctus-infestation-of-b.-inermis}

\emph{Insects and Cages.} Assessment of \emph{C. cinctus} infestation
and mortality within \emph{B. inermis} were assessed using a 34 x 60 ft
plot at the Arthur H. Post Agronomy Farm (43°38'19.39''N,
116°14'28.86''W), an experimental farm of Montana State University in
Bozeman, MT. There is a negligible population of \emph{C. cinctus} and
associated parasitoids at this location. Experimental cages were built
using 1-inch PVC piping with the netting made using 530\(\mu\) Amber
Lumite Screen (BioQuip\(^\circledR\) Products, LLC). Twelve cages were
built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations
selected randomly based on the space available within the plot and
arranged in sets of three.

Wheat stem stubble was collected in Three Forks, MT, USA
(43°38'19.39''N, 116°14'28.86''W) from fields that experienced high
levels of \emph{C. cinctus} infestation and cutting the year prior, as
previously described (\citeproc{ref-Hager2024}{Hager et al. 2024}). Cut
stubble, which contained \emph{C. cinctus} larvae in diapause, were kept
refrigerated between -2°C and 3°C for \textgreater100 days as required
to complete obligatory larval diapause. As needed, stubs were removed
from refrigeration and kept at 22-27°C for 4-5 weeks inside of 100 oz
GladWare® storage containers (Glad®, Oakland, California USA). Once
\emph{B. inermis} stems reached six inches tall, stub containers with
emerging sawflies were added to cages to mimic sawfly infestation
pressure. Sawfly quantity treatments were as follows: high (600 stubs),
low (200 stubs), and control (0 stubs).

\emph{Data Collection.} In late August, \emph{B. inermis} stems were
collected from each cage. Each stem was sliced open using X-Acto® knives
to collect data on infestation, dead larvae and live larvae, as well as
very rare parasitism events, for each internode.

\subsubsection{Montana Field Survey}\label{montana-field-survey}

\emph{Stem collection and processing.} We conducted a field survey to
assess \emph{C. cinctus} infestation, larval mortality, and \emph{B.
cephi} and \emph{B. lissogaster} prevalence within \emph{B. inermis} and
adjacent wheat fields. Sites were chosen across 2 counties in
north-central Montana, United States. (Chouteau, Judith Basin), which
consistently experience high \emph{C. cinctus} pressure. Samples were
collected from wheat fields and adjacent \emph{B. inermis} in early July
and late August in 2022, and 2023 from sites in Big Sandy and Moccasin
MT, USA. Sampling sites were set up as 100\(m^2\) polygons along the
edge of adjoining wheat fields. Four collection squares of 1ft x 1ft
were randomly selected within each polygon during both collection events
each year. All stems within each 1 x 1 ft square were collected using a
shovel to remove both stem and root material. Wheat stems were collected
at distances of 5 and 20 meters from the edge of the field. Samples were
collected in 4 rows at 10 meters apart. 2 samples were collected in each
row at distances of 5 and 20 meters. 1 ft samples were collected at each
point.

Wheat and \emph{B. inermis} stems were then returned to Montana State
University, Bozeman, Montana and stored in a 10°C cold, humid storage
until dissection. Stems were dissected in the same manner as samples
from our controlled experiment. We recorded presence or absence of
\emph{C. cinctus} larval infestation, live eggs, dead eggs, dead larvae,
live larvae and parasitism, in addition to stem cutting. Sawfly larvae
were identified based on descriptions in Criddle (1915) and Wallace and
McNeal (1996).

\emph{Statistical Analysis.} We used generalized linear mixed models
with binomial errors (logit link) fit using bound optimization by
quadratic approximation, with a maximum of 200,000 iterations, were run
in the \emph{lme4} package in R (\citeproc{ref-lme4}{Bates et al. 2015})
to examine the effects of location and year (fixed effects) on each of
the three response variables: proportion of stems infested, proportion
of stems cut, and proportion of stems parasitized by \emph{Bracon} spp.

To better understand the densities of host \emph{C. cinctus} and the
\emph{Bracon} spp. within the wheat and adjacent \emph{B. inermis} at
each sampling site, we converted our units to the ratio of parasitoids
to surviving hosts stem per unit area. Densities of stems are different
when considering semi-natural, long-standing stands of \emph{B. inermis}
and the adjacent cultivated wheat or barley.

\subsection{Results}\label{results}

\subsubsection{Historical Weather Analysis \&
NDVI}\label{historical-weather-analysis-ndvi}

We observed a significant linear relationship (\emph{r = 0.1, P = 0.033,
estimate = -0.058}) between average precipitation and year for both Big
Sandy and Moccasin, Montana. For each location and one-year increase, we
observed a 0.05 inch decrease in precipitation. (\hyperref[figure1]{Fig.
1}). Additionally, we observed a significant positive linear
relationship (\emph{r = 0.245, P \textless{} 0.05, estimate = 0.028})
between year and average yearly temperature for Moccasin and Big Sandy,
MT. Thus, each year, the average daily temperature has increased by
0.03°C. (\hyperref[figure2]{Fig. 2}).

We observed a notable difference in NDVI when comparing adjacent
\emph{B. inermis} and spring wheat. We saw a significant difference in
the July NDVI (\emph{0.846, P \textless{} 0.05}). The \emph{B. inermis}
NDVI remained relatively linear in it's downslope (BROME SLOPE POST
JULY) compared to the wheat field (WHEAT FIELD POST JULY).
(\hyperref[figure3]{Fig. 3})

Need to add map showing where NDVI measurements came from

\subsubsection{\texorpdfstring{Controlled \emph{C. cinctus} Infestation
of \emph{B.
inermis}}{Controlled C. cinctus Infestation of B. inermis}}\label{controlled-c.-cinctus-infestation-of-b.-inermis-1}

Initially, \emph{C. cinctus} heavily infested stems of \emph{B. inermis}
in the controlled test conditions. Averaged across both years, we
observed 66.5\% of stems infested for high treatments and 47.3\% of
stems for low treatments. We found strong evidence suggesting that there
was a significant difference between infestation at high and low
treatment levels when holding year constant (\emph{r = 0.83, P
\textless{} 0.05}, \hyperref[figure4]{Fig. 3}).

Stem cutting by mature larvae, leading into winter, was observed at
5.7\% for the high treatments and 3.9\% for the low, showing strong
evidence for a difference in cutting between high and low treatment
groups (\emph{r = 0.592, P \textless{} 0.05}). There was a dramatic
decrease in survival observed from initial infestation to stem cutting.

\emph{C. cinctus} infested stems often contained larvae burrowing
through multiple nodes. In 2022, 64.3\% of stems were observed to have
at least 4 nodes burrowed through, or 5 internodes occupied. In 2023,
61.7\% of stems had at least 4 nodes infested \hyperref[figure5]{Fig. 5}

\subsubsection{Montana Field Sites}\label{montana-field-sites}

We quantified \emph{C. cinctus} and \emph{Bracon} spp. parasitoid
presence in 6,148 wheat and grass stems across 10 research sites in 2022
and 2023. Initial infestation of \emph{B. inermis} by \emph{C. cinctus}
varied between collection sites, with the greatest infestation taking
place within our three Big Sandy, MT sampling sites (2023: 65.4\%, 2022:
63.1\%), while the lowest infestation observed was in Moccasin, MT
(2023: 40.8\%, 2022: 60.7\%). Across all sites in Big Sandy and
Moccasin, we observed an average infestation of 57.5\% within \emph{B.
inermis} and 47.6\% within the adjacent wheat. Cutting by \emph{C.
cintus} was observed at an average rate of 5.59\% (se = 0.02) in
\emph{B. inermis} and 47.1\% (se = 0.04) in the adjacent cultivated
wheat (\hyperref[figure7]{Fig. 7}.

To generate a relative estimate of the number of available \emph{C.
cinctus} host larvae within \emph{B. inermis} late in the growing
season, we calculated the number of stems that contained at least four
or more nodes with burrowing injury. In Big Sandy, 76.2\% of collected
stems in the fall exhibited at least 4 nodes of burrowing injury at our
second sampling date. In Moccasin, the corresponding figure was 66.5\%.

We utilized the ratio of parasitized stems to cut stems to better
understand the true impact of our \emph{B. inermis} sink/source ecology.
The largest ratio observed was in Big Sandy, MT in the fall of 2022,
where we observed a ratio of 16 parasitoids to a single \emph{C.
cinctus}-cut stem. Thus, for every square meter of \emph{B. inermis}, we
would observe 16 stems that produced a parasitoid for every 1 stem that
was likely to produce an adult \emph{C. cinctus}. All other sample
locations also showed positive ratios of parasitoids to cut stems
\hyperref[figure8]{(Fig. 8)}.

To understand the seasonal trends of \emph{Bracon} spp use of potential
\emph{C. cinctus} hosts within \emph{B. inermis}, we collected stem
samples twice each growing season. We collected one pre-harvest, in
July, and a second time in September, after all adult \emph{Bracon} spp
would be dead. We observed a significant difference (\emph{df = 96, r =
0.304, p \textless{} 0.05}) between parasitoid presence in infested
stems collected in Big Sandy and Moccasin, MT.

I want to drive home that stems with 3 + nodes infested likely had
larave that were alive late into the summer. Compare the number of nodes
infested in the summer to stems that have 3+ infested in the fall.

\subsection{Discussion}\label{discussion}

Assessing the comparative abundance of \emph{C. cinctus} and its
associated parasitoids within \emph{B. inermis} and adjacent cultivated
cereal hosts is an important initial step in evaluating whether
semi-natural environments within road-side or coulee areas serve as sink
or source origins of the pest and its beneficial natural adversaries
infiltrating agricultural crops. As the climate of central and northern
Montana continues to change, wheat stem sawfly infestation will continue
to impact cereal crop yields. In this study, we monitored how \emph{B.
inermis} may be potentially beneficial in ecologically buffering
\emph{Bracon} spp parasitoids by observing \emph{C. cinctus} mortality
within \emph{B. inermis} and within adjacent cereal crops as well as
\emph{Bracon} spp prevalence.

\subsubsection{Controlled Infestation}\label{controlled-infestation}

Results from the controlled \emph{B. inermis} infestation confirmed
large amounts of \emph{C. cinctus} larval mortality prior to
overwintering (\citeproc{ref-Shanower2004}{Shanower and Hoelmer 2004},
\citeproc{ref-Mendoza2006}{Perez-Mendoza and Weaver 2006},
\citeproc{ref-Buteler2008}{Buteler et al. 2008}) with 53\% of total
collected stems exhibiting \emph{C. cinctus} burrowing damage. Of those
infested stems, a large proportion exhibited larval damage in at least 5
internodes (62.5\%). While the timing of stem collection and dissection
did not align with precise measurement of living larvae (stems dissected
in October), we can assert that stems with at least 4 nodes burrowed
through would have a high probability of containing large, live larvae,
and thus a viable potential parasitoid host, in late July and August of
the same year.

While total and per node \emph{B inermis} infestation by \emph{C.
cinctus} was high, survival to overwintering (cutting) only occurred in
5.7\% of \emph{C. cinctus} infested stems. Our results indicate that
\emph{B. inermis} serves as a high-quality host for \emph{C. cinctus}
larvae during the spring and summer months. However, there was minimal
survival of \emph{C. cinctus} larvae beyond August and September,
confirming that there will be little adult emergence from \emph{B.
inermis} the following season.

\subsubsection{Montana Field Site
Survey}\label{montana-field-site-survey}

Analysis of weather trends in Big Sandy and Moccasin, MT confirmed that
the patterns in central and northern Montana are changing towards warmer
and drier conditions (\citeproc{ref-Pederson2009}{Pederson et al.
2009}). We observed an average decrease in 0.058 inches of precipitation
per year since 1975, while since 1950, temperatures have experienced an
average yearly increase of 0.028°C. These changes in precipitation and
temperature can have significant impacts on timing of crop production
and harvest timing (\citeproc{ref-Zhu2021}{Zhu and Burney 2021}).
Year-over-year populations of the \emph{Bracon} spp parasitoids are
particularly sensitive, as their bivoltine nature requires the second
yearly generation to find a suitable \emph{C. cinctus} host
(\citeproc{ref-Nelson1953}{Nelson and Farstad 1953},
\citeproc{ref-Runyon2002}{Runyon et al. 2002}). NDVI analyses exhibited
the prolonged senescence period for \emph{B. inermis} when compared to
adjacent cultivated wheat. \emph{C. cinctus} larvae are cued via the
thinning of the cell membranes (\citeproc{ref-Villacorta1971}{Villacorta
et al. 1971}) during plant senescence to travel to the base of the stem
and prepare for obligatory diapause (\citeproc{ref-Beres2011}{Beres et
al. 2011}). Because \emph{B. inermis} senescence occurs at slower rates
compared to wheat or barley, \emph{C. cinctus} may remain within the
stem later in into the summer and fall, providing potential host larvae
for second generation \emph{Bracon} parasitoids.

Previous research has shown \emph{B. inermis} to be a quality host of
both \emph{C. cinctus} and \emph{Bracon} spp parasitoids
(\citeproc{ref-Criddle1922}{Criddle 1922},
\citeproc{ref-Peirce2021}{Peirce et al. 2021},
\citeproc{ref-Rand2024}{Rand et al. 2024}), but little data is available
on the season fluctuation of \emph{B. inermis} utilization
(\citeproc{ref-Criddle1922}{Criddle 1922},
\citeproc{ref-Peirce2021}{Peirce et al. 2021},
\citeproc{ref-Rand2024}{Rand et al. 2024}). Because we collected stem
samples twice during the growing season, we were able to compare levels
of parasitism. We hoped to understand how parasitism rates changed from
July (pre-wheat harvest) and September (post-wheat harvest) of each
year.

Our results showed that parasitoid presence increase significantly at
all sites between pre- and post-harvest collections. Given the
collection methods, September collection parasitism was inherently a
combination of both pre-harvest and post-harvest activity. While we
cannot ascertain \emph{Bracon} preference for host stems, we can assess
that the parasitoids utilize infested \emph{B. inermis} stems
post-harvest when \emph{C. cinctus} is no longer available within wheat
stems.

Additionally, we observed a high number of stems exhibiting levels of
\emph{C. cinctus} damage that point to prolonged larval presence within
\emph{B. inermis} stems. In Big Sandy (76.2\%) and Moccasin (66.5\%),
stems that exhibited four or more nodes burrowed suggests that \emph{C.
cinctus} larvae remain available within \emph{B. inermis} stems past
wheat harvest, when they are more limited in availability. Most
importantly, in years of drought or increased temperatures where wheat
senescence and harvest may occur earlier in the summer.

We utilized parasitism to cut stem ratios to understand the relative
impact of \emph{B. inermis} host sources on overall field parasitism.
Assessing overall numbers of insects can potentially diminish the
apparent importance of \emph{B. inermis}, as per square meter stem
counts in cultivated wheat are denser than \emph{B. inermis} counts
within coulees and along roadsides \emph{B. inermis}. Our findings of
significantly greater ratios of parasitized stems to cut stems within
\emph{B. inermis} when compared to adjacent when indicates these
field-side areas provide greater parasitoid source per square meter when
compared to cultivated fields. This emphasizes the significance of
accounting for habitat diversity to sustain parasitoid populations over
successive years, as well as recognizing the crucial role of edge-row
areas in maximizing parasitoid effectiveness in minimizing \emph{C.
cinctus} damage.

\subsection{Figures}\label{figures}

\subsubsection{Figure 1}\label{figure1}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/FIG1_both_prcp_plot-1.pdf}
\caption{Figure 1: Average yearly precipitation (in) for Big Sandy and
Moccasin, Montana. Black trend line signifies averaged negative trend
between both locations. Data gathered from NOAA and MSU-ARS. A
significant linear relationship (r = 0.1, P = 0.033, estimate = -0.058)
was observed between year and average yearly precipitation.}
\end{figure}

\subsubsection{Figure 2}\label{figure2}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/both_temp_plot-1.pdf}
\caption{Figure 2: Average yearly temperature (°C) in Moccasin and Big
Sandy, MT. Positive trendline slop suggests increasing average
temperatures of the past 100 years. A significant linear relationship
(\emph{r =0.2447, P \textless{} 0.05, estimate = 0.028}) was observed
between year and average yearly temperature.}
\end{figure}

\subsubsection{Figure 3}\label{figure3}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/bs_ndvi_plot-1.pdf}
\caption{Figure 3: Normalized difference vegetation index (NDVI) of
\emph{B. inermis} and adjacent spring wheat field from April 2023 to
October 2023. Post harvest linear model indicates a significant
difference (P values and stuff) when comparing the \emph{B. inermis}
post-harvest slope and the spring wheat post-harvest slope.}
\end{figure}

\includegraphics{Images/brome_images.jpg}

\subsubsection{Figure 4}\label{figure4}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/inf_cut_plot-1.pdf}
\caption{Figure 4: Comparison of year and treatment group for controlled
infestation of \emph{B. inermis}. Three treatment groups - high, low,
and control (0) - were used. The controlled groups showed no sign of
\emph{C. cinctus} stem damage. We observed a significant difference in
cutting between high and low treatment groups (\emph{r = 0.592, P
\textless{} 0.05}).}
\end{figure}

\subsubsection{Figure 5}\label{figure5}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/figure5_pf_nodes_plot-1.pdf}
\caption{Figure 6: Ineternodes infested by \emph{B. cephi} within
controlled infestation plots in Bozeman, MT, USA. We found that 61.2\%
of stems exhibited infestation in at least 5 nodes}
\end{figure}

\subsubsection{Figure 6}\label{figure6}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/figure6_brome_nodes-1.pdf}
\caption{Figure 6: Internodes infested by \emph{B. cephi} within
collected \emph{B. inermis} stems in Moccasin and Big Sandy, MT, USA.}
\end{figure}

\subsubsection{Figure 7}\label{figure7}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/figure7_field_cut_plot-1.pdf}
\caption{Figure 7: Stem cutting by \emph{C. cinctus}. We compared WSS
cutting in \emph{B. inermis} and adjacent wheat fields in Big Sandy and
Moccasin, MT, USA. The horizontal lines correspond to yearly mean
cutting for each plant type across all sampling sites.}
\end{figure}

\subsubsection{Figure 8}\label{figure8}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/figure8-1.pdf}
\caption{Ratios of parasitized to cut stems in \emph{B. inermis} and
adjacent cultivated wheat. Dotted line represents where a 1:1
cut:parasitized ratio lies.}
\end{figure}

\subsubsection{Figure 9}\label{figure9}

\begin{figure}

\includegraphics{JacksonStrand_Paper 1_files/figure-latex/figure9-1} \hfill{}

\caption{Comparison of parasitoid presence in *B. inermis* between July and September collections.}\label{fig:figure9}
\end{figure}

\subsubsection{Figure 10}\label{figurex}

\begin{figure}
\centering
\includegraphics{JacksonStrand_Paper 1_files/figure-latex/figure10-1.pdf}
\caption{Location of Montana, USA field sites. Controlled \emph{B.
inermis} infestation site was in Bozeman, while field sites were located
in centrally location Moccasin, MT and northern Big Sandy, MT.}
\end{figure}

\subsection*{Citations}\label{citations}
\addcontentsline{toc}{subsection}{Citations}

\phantomsection\label{refs}
\begin{CSLReferences}{0}{1}
\bibitem[\citeproctext]{ref-Achhami2020}
Achhami BB, Reddy GVP, Sherman JD, et al. 2020. Antixenosis, antibiosis,
and potential yield compensatory response in barley cultivars exposed to
wheat stem sawfly (hymenoptera: Cephidae) under field conditions.
Journal of Insect Science. 20:1--14.
\url{https://doi.org/10.1093/JISESA/IEAA091}.

\bibitem[\citeproctext]{ref-Adhikari2019}
Adhikari S, Adhikari A, Weaver DK, et al. 2019. Impacts of agricultural
management systems on biodiversity and ecosystem services in highly
simplified dryland landscapes. Sustainability (Switzerland). 11.
\url{https://doi.org/10.3390/su11113223}.

\bibitem[\citeproctext]{ref-Ainslie1920}
Ainslie CN. 1920. The western grass-stem sawfly. United States
Department of Agriculture.

\bibitem[\citeproctext]{ref-lme4}
Bates MD, Bolker B, Walker S. 2015. Fitting linear mixed-effects models
using lme4. Journal of Statistical Software. 67:1--48.

\bibitem[\citeproctext]{ref-Bekkerman2018}
Bekkerman A, Weaver DK. 2018. Modeling joint dependence of managed
ecosystems pests: The case of the wheat stem sawfly. Journal of
Agricultural and Resource Economics. 43:172--194.

\bibitem[\citeproctext]{ref-Beres2011a}
Beres BL, Cárcamo HA, Weaver DK, et al. 2011. Integrating the building
blocks of agronomy and biocontrol into an IPM strategy for wheat stem
sawfly. Background and status. :54--65. Available from
\href{https://www.prairiesoilsandcrops.caVolume4·▪2011}{www.prairiesoilsandcrops.caVolume4·▪2011}.

\bibitem[\citeproctext]{ref-Beres2011}
Beres BL, Dosdall LM, Weaver DK, et al. 2011. Biology and integrated
management of wheat stem sawfly and the need for continuing research.
Canadian Entomologist. 143:105--125.
\url{https://doi.org/10.4039/n10-056}.

\bibitem[\citeproctext]{ref-Bhandari2020}
Bhandari R. 2020.
\href{https://www.ncbi.nlm.nih.gov/pubmed/25246403}{Assessment of host
selection behaviors and oviposition preferences of cephus cinctus norton
(hymenoptera: Cephidae) using wheat and smooth brome}.

\bibitem[\citeproctext]{ref-Buteler2015}
Buteler M, Peterson RKD, Hofland ML, et al. 2015. A multiple decrement
life table reveals that host plant resistance and parasitism are major
causes of mortality for the wheat stem sawfly. Environmental Entomology.
44:1571--1580. \url{https://doi.org/10.1093/ee/nvv128}.

\bibitem[\citeproctext]{ref-Buteler2008}
Buteler M, Weaver DK, Miller PR. 2008. Wheat stem sawfly-infested plants
benefit from parasitism of the herbivorous larvae. Agricultural and
Forest Entomology. 10:347--354.
\url{https://doi.org/10.1111/j.1461-9563.2008.00396.x}.

\bibitem[\citeproctext]{ref-Buteler2009}
Buteler M, Weaver DK, Peterson RKD. 2009. Oviposition behavior of the
wheat stem sawfly when encountering plants infested with cryptic
conspecifics. Journal of Environmental Entomology. 38:1707--1715.
Available from
\url{https://academic.oup.com/ee/article/38/6/1707/361666}.

\bibitem[\citeproctext]{ref-Cano2022}
Cano D, Martínez-Núñez C, Pérez AJ, et al. 2022. Small floral patches
are resistant reservoirs of wild floral visitor insects and the
pollination service in agricultural landscapes. Biological Conservation.
276. \url{https://doi.org/10.1016/j.biocon.2022.109789}.

\bibitem[\citeproctext]{ref-Carlson1985}
Carlson IT, Newell LC. 1985. Smooth bromegrass. In: Forages: the science
of grassland agriculture. Iowa State University. p. 198--206. Available
from \url{https://www.cabdirect.org/cabdirect/abstract/19850777665}.

\bibitem[\citeproctext]{ref-Cockrell2017}
Cockrell DM, Griffin-Nolan RJ, Rand TA, et al. 2017. Host plants of the
wheat stem sawfly (hymenoptera: cephidae). Environmental Entomology.
46:847--854. \url{https://doi.org/10.1093/ee/nvx104}.

\bibitem[\citeproctext]{ref-Cockrell2021}
Cockrell DM, Randolph T, Peirce E, et al. 2021. Survey of wheat stem
sawfly (hymenoptera: Cephidae) infesting wheat in eastern colorado.
Journal of Economic Entomology. 114:998--1004.
\url{https://doi.org/10.1093/JEE/TOAB015}.

\bibitem[\citeproctext]{ref-Criddle1922}
Criddle N. 1922. The western-stem sawfly and its control. Canadian
Department of Agriculture.

\bibitem[\citeproctext]{ref-Davis1955}
Davis EG, Benton C, Somsen HW. 1955. Natural enemies of the wheat stem
sawfly in north dakota and montana. North Dakota Agricultural
Experimental Bimonthly Bulletin. 18:63--65.

\bibitem[\citeproctext]{ref-Davis2013}
Davis RA. 2013. Mechanisms for reproductive isolation in two congeneric
parasitoids of the wheat stem sawfly.

\bibitem[\citeproctext]{ref-Dillemuth2008}
Dillemuth FP, Rietschier EA, Cronin JT. 2008. Patch dynamics of a native
grass in relation to the spread of invasive smooth brome
(bromus~inermis). Biological Invasions 2008 11:6. 11:1381--1391.
\url{https://doi.org/10.1007/S10530-008-9346-7}.

\bibitem[\citeproctext]{ref-Evans1999}
Evans EW. 1999. Intra versus interspecific interactions of ladybeetles
(coleoptera: Coccinellidae) attacking aphids.

\bibitem[\citeproctext]{ref-Farstad1945}
Farstad CW, Jacobson L. 1945. Manual for sawfly control workers in
alberta.

\bibitem[\citeproctext]{ref-Gahan1918}
Gahan AB. 1918. Description of a new hymenopterous parasite
(braconidae). In: Proceedings of the Entomological Society of
Washington. Vol. 20. p. 18--19.

\bibitem[\citeproctext]{ref-Hager2024}
Hager MS, Hofland ML, Varella AC, et al. 2024. Untargeted metabolomics
profiling of oat (avena sativa l.) and wheat (triticum aestivum l.)
infested with wheat stem sawfly (cephus cinctus norton) reveals
differences associated with plant defense and insect nutrition.
Frontiers in Plant Science. 15.
\url{https://doi.org/10.3389/fpls.2024.1327390}.

\bibitem[\citeproctext]{ref-Holmes1982}
Holmes ND. 1982. Population dynamics of the wheat stem sawfly. The
Canadian Entomologist. 114:775--788.
\url{https://doi.org/10.4039/Ent114775-9}.

\bibitem[\citeproctext]{ref-Holmes1956}
Holmes ND, Farstad CW. 1956. Effects of field exposure on immature
stages of the wheat stem sawfly, cephus cinctus nort. (Hymenoptera:
cephidae). Canadian Journal of Agricultural Science. 36:196--202.
Available from
\url{https://cdnsciencepub.com/doi/abs/10.4141/agsci-1956-0023}.

\bibitem[\citeproctext]{ref-Holmes1963}
Holmes ND, Nelson WA, Peterson LK, et al. 1963. Causes of variation in
effectiveness of bracon cephi (gahan) (hymenoptera: Braconidae) as a
parasite of the wheat stem sawfly. The Canadian Entomologist.
95:113--126.

\bibitem[\citeproctext]{ref-Holmes1960}
Holmes ND, Peterson LK. 1960. THE INFLUENCE OF THE HOST ON OVIPOSITION
BY THE WHEAT STEM SAWFLY, CEPHUS CINCTUS NORT. (HYMENOPTERA: CEPHIDAE).
Canadian Journal of Plant Science. 40:29--46.
\url{https://doi.org/10.4141/cjps60-004}.

\bibitem[\citeproctext]{ref-Kennedy2000}
Kennedy GG, Storer NP. 2000. Life systems of polyphagous arthropod pests
in temporally unstable cropping systems. Annual Review of Entomology.
45:467--493.

\bibitem[\citeproctext]{ref-Lesieur2016}
Lesieur V, Martin JF, Weaver DK, et al. 2016. Phylogeography of the
wheat stem sawfly, cephus cinctus norton (hymenoptera: Cephidae):
Implications for pest management. PLoS ONE. 11:168370.
\url{https://doi.org/10.1371/journal.pone.0168370}.

\bibitem[\citeproctext]{ref-McCullough2020}
McCullough CT, Hein GL, Bradshaw JD. 2020. Phenology and dispersal of
the wheat stem sawfly (hymenoptera: Cephidae) into winter wheat fields
in nebraska. Journal of Economic Entomology. 113:1831--1838.
\url{https://doi.org/10.1093/jee/toaa093}.

\bibitem[\citeproctext]{ref-DeMorais2023}
Morais RMD, Freitas De Morais A de, Handte VG, et al. 2023. Enhancing
arthropod communities through plant diversified edge of kale
cultivation. Pesquisa Agropecuária Gaúcha. 29:77--91.
\url{https://doi.org/10.36812/pag.202329177-91}.

\bibitem[\citeproctext]{ref-Morrill1996}
Morrill WL, Kushnak GD. 1996. Wheat stem sawfly (hymenoptera: Cephidae)
adaptation to winter wheat. Environmental Entomology. 25:1128--1132.
\url{https://doi.org/10.1093/EE/25.5.1128}.

\bibitem[\citeproctext]{ref-Morrill1998}
Morrill WL, Kushnak GD, Gabor JW. 1998. Parasitism of the wheat stem
sawfly (hymenoptera: Cephidae) in montana. Biological Control.
12:159--163. \url{https://doi.org/10.1006/bcon.1998.0629}.

\bibitem[\citeproctext]{ref-Morrill2001}
Morrill WL, Weaver DK, Johnson GD. 2001. Trap strip and field border
modification for management of the wheat stem sawfly (hymenoptera:
cephidae). Journal of Entomological Science. 36:34--45.
\url{https://doi.org/10.18474/0749-8004-36.1.34}.

\bibitem[\citeproctext]{ref-Nansen2005a}
Nansen C, Macedo TB, Weaver DK, et al. 2005. Spatiotemporal
distributions of wheat stem sawfly eggs and larvae in dryland wheat
fields. Canadian Entomologist. 137:428--440.
\url{https://doi.org/10.4039/n04-094}.

\bibitem[\citeproctext]{ref-Nelson1953}
Nelson WA, Farstad CW. 1953. Biology of bracon cephi (gahan)
(hymenoptera: Braconidae), an important native parasite of the wheat
stem sawfly, cephus cinctus nort. (Hymenoptera: Cephidae), in western
canada. The Canadian Entomologist. 85:103--107.
\url{https://doi.org/10.4039/Ent85103-3}.

\bibitem[\citeproctext]{ref-Olfert2019}
Olfert O, Weiss RM, Catton H, et al. 2019. Bioclimatic assessment of
abiotic factors affecting relative abundance and distribution of wheat
stem sawfly (hymenoptera: Cephidae) in western canada. Canadian
Entomologist. 151:16--33. \url{https://doi.org/10.4039/tce.2018.46}.

\bibitem[\citeproctext]{ref-Otfinowski2006}
Otfinowski R, Kenkel NC, Catling PM. 2006. The biology of canadian
weeds. 134. Bromus inermis leyss. Canadian Journal of Plant Science.
87:183--198.

\bibitem[\citeproctext]{ref-Pederson2009}
Pederson GT, Graumlich LJ, Fagre DB, et al. 2009. A century of climate
and ecosystem change in western montana: What do temperature trends
portend? Climatic Change. 98:133--154.
\url{https://doi.org/10.1007/s10584-009-9642-y}.

\bibitem[\citeproctext]{ref-Peirce2021}
Peirce ES, Rand TA, Cockrell DM, et al. 2021. Effects of landscape
composition on wheat stem sawfly (hymenoptera: Cephidae) and its
associated braconid parasitoids. Journal of Economic Entomology.
114:72--81. \url{https://doi.org/10.1093/jee/toaa287}.

\bibitem[\citeproctext]{ref-Mendoza2006}
Perez-Mendoza J, Weaver DK. 2006. Temperature and relative humidity
effects on postdiapause larval development and adult emergence in three
populations of wheat stem sawfly (hymenoptera: cephidae). Environmental
Entomology. 35:1222--1231. \url{https://doi.org/10.1093/ee/35.5.1222}.

\bibitem[\citeproctext]{ref-Peterson2011}
Peterson RKD, Buteler M, Weaver DK, et al. 2011. Parasitism and the
demography of wheat stem sawfly larvae, cephus cinctus. BioControl.
56:831--839. \url{https://doi.org/10.1007/s10526-011-9357-7}.

\bibitem[\citeproctext]{ref-Peterson1999}
Peterson RO. 1999. Wolf-moose interaction on isle royale: The end of
natural regulation. Ecological Applications. 9:10--16.
\url{https://doi.org/10.1890/1051-0761(1999)009\%5B0010:WMIOIR\%5D2.0.CO;2}.

\bibitem[\citeproctext]{ref-Pettorelli2005}
Pettorelli N, Vik JO, Mysterud A, et al. 2005. Using the
satellite-derived NDVI to assess ecological responses to environmental
change. Trends in Ecology and Evolution. 20:503--510.
\url{https://doi.org/10.1016/j.tree.2005.05.011}.

\bibitem[\citeproctext]{ref-Rand2024}
Rand TA, Kula RR, Gaskin JF. 2024. Evaluating the use of common grasses
by the wheat stem sawfly (hymenoptera: Cephidae) and its native
parasitoids in rangeland and conservation reserve program grasslands.
Liu T-X, editor. Journal of Economic Entomology.
\url{https://doi.org/10.1093/jee/toae046}.

\bibitem[\citeproctext]{ref-Rand2020}
Rand TA, Richmond CE, Dougherty ET. 2020. Modeling the combined impacts
of host plant resistance and biological control on the population
dynamics of a major pest of wheat. Pest Management Science.
76:2818--2828. \url{https://doi.org/10.1002/ps.5830}.

\bibitem[\citeproctext]{ref-Runyon2001b}
Runyon JB. 2001. Wheat stem sawfly parasitism in varying field sizes and
tillage systems in dryland wheat in montana. Canadian Entomologist. 133.

\bibitem[\citeproctext]{ref-Runyon2001a}
Runyon JB, Hurley RL, Morrill WL, et al. 2001. Distinguishing adults of
bracon cephi and bracon lissogaster (hymenoptera: Braconidae),
parasitoids of the wheat stem sawfly (hymenoptera: cephidae). Canadian
Entomologist. 133:215--217. \url{https://doi.org/10.4039/Ent133215-2}.

\bibitem[\citeproctext]{ref-Runyon2002}
Runyon JB, Morrill WL, Weaver DK, et al. 2002. Parasitism of the wheat
stem sawfly (hymenoptera: Cephidae) by bracon cephi and b. Lissogaster
(hymenoptera: Braconidae) in wheat fields bordering tilled and untilled
fallow in montana. Journal of economic entomology. 95:1130--1134.
\url{https://doi.org/10.1603/0022-0493-95.6.1130}.

\bibitem[\citeproctext]{ref-Bolwahn2011}
Salesman JB, Thomsen M. 2011. Smooth brome (bromus inermis) in tallgrass
prairies: A review of control methods and future research directions.
Ecological Restoration. 29:374--381.
\url{https://doi.org/10.3368/er.29.4.374}.

\bibitem[\citeproctext]{ref-Seamans1928}
Seamans HL. 1928. The value of trap crops in the control of the wheat
stem sawfly in alberta. The Value of Trap Crops in the Control of the
Wheat Stem Sawfly in Alberta.

\bibitem[\citeproctext]{ref-Shanower2004}
Shanower TG, Hoelmer KA. 2004. Biological control of wheat stem
sawflies: Past and future. Journal of Agricultural Entomology.
21:197--221.

\bibitem[\citeproctext]{ref-Somsen1956}
Somsen HW, Luginbill P. 1956. Bracon lissogaster mues: A parasite of the
wheat stem sawfly. USDA Technical Bullitin. 1153. Available from
\url{https://www.google.com/books/edition/Bracon_Lissogaster_Mues/37UXAAAAYAAJ?hl=en&gbpv=1&dq=Bracon+Lissogaster+Mues:+A+Parasite+of+the+Wheat+Stem+Sawfly.&pg=PA29&printsec=frontcover}.

\bibitem[\citeproctext]{ref-Tscharntke2016}
Tscharntke T, Karp DS, Chaplin-Kramer R, et al. 2016. When natural
habitat fails to enhance biological pest control -- five hypotheses.
Biological Conservation. 204:449--458.
\url{https://doi.org/10.1016/j.biocon.2016.10.001}.

\bibitem[\citeproctext]{ref-Villacorta1971}
Villacorta A, Bell R, Callenbach J. 1971. Influence of high temperature
and light on postdiapause development of the wheat stem sawfly. Journal
of Economic Entomology. 64:749--751.

\bibitem[\citeproctext]{ref-Wallace1966}
Wallace LE, McNeal FH. 1966. Stem sawflies of economic importance in
grain crops in the united states. U.S. Department of Agriculture
Technical Bulletin No. 1350. Available from
\href{https://books.google.com/books?hl=en&lr=&id=dcMXAAAAYAAJ&oi=fnd&pg=PA1&dq=Stem+sawflies+of+economic+importance+in+grain+crops+in+the+United+States&ots=e0FbKZOb6x&sig=_fPd1FCRZ_HK-Ncv2POX-EFMvOo\#v=onepage&q=Stem\%20sawflies\%20of\%20economic\%20importance\%20in\%20grain\%20cro}{https://books.google.com/books?hl=en\&lr=\&id=dcMXAAAAYAAJ\&oi=fnd\&pg=PA1\&dq=Stem+sawflies+of+economic+importance+in+grain+crops+in+the+United+States\&ots=e0FbKZOb6x\&sig=\_fPd1FCRZ\_HK-Ncv2POX-EFMvOo\#v=onepage\&q=Stem
sawflies of economic importance in grain cro}.

\bibitem[\citeproctext]{ref-Weaver2023}
Weaver D. 2023. Wheat stem sawfly (cephus cinctus norton). p. 93--134.
\url{https://doi.org/10.19103/as.2022.0114.13}.

\bibitem[\citeproctext]{ref-ggplot}
Wickham H. 2016. ggplot2: Elegant graphics for data analysis. Available
from \url{https://ggplot2.tidyverse.org}.

\bibitem[\citeproctext]{ref-Willson2000}
Willson GD, Stubbendieck J. 2000. A provisional model for smooth brome
management in degraded tallgrass prairie. Ecological Restoration.
18:34--38. \url{https://doi.org/10.3368/er.18.1.34}.

\bibitem[\citeproctext]{ref-Wilson1923}
Wilson ML. 1923. Dry farming in the north central montana triangle.
Bowden RB, editor.

\bibitem[\citeproctext]{ref-Zhu2021}
Zhu P, Burney J. 2021. Temperature-driven harvest decisions amplify US
winter wheat loss under climate warming. Global Change Biology.
27:550--562. \url{https://doi.org/10.1111/gcb.15427}.

\end{CSLReferences}

\end{document}

### 2024-06-06 qmd/projects/msu/wss/paper1.rmd
update

@@
 authors: "Jackson R. Strand, David K. Weaver"
-date: "May 17, 2024"
+date: "`r Sys.Date()`"
 editor_options: 
@@
 link-citations: true
-biblio-style: apalike
 csl: entomological-society-of-america-2.csl
@@
 
-Wheat stem sawfly, Cephus cinctus Norton (WSS), causes significant damage in cereal crops throughout the Northern Great Plains of North America.
+Wheat stem sawfly, *Cephus cinctus* Norton (WSS), causes significant damage in cereal crops throughout the Northern Great Plains of North America.
 Bracon cephi and Bracon lissogaster are native WSS parasitoids important in managing WSS outbreaks and damage.
@@
 Ecologists and farmers frequently hold divergent views regarding the importance of natural habitats within agricultural landscapes, a distinction often overlooked in ecological and conservation discourse [@Tscharntke2016].
-While ecologists recognize the role of natural habitats in supporting biodiversity and providing potential ecosystem services like biological pest control, many farmers perceive these habitat remnants as occupying valuable cropland or even as sources of pests.
-As such, documenting how pest and natural enemy populations interact with agricultural landscapes alongside natural habitat buffers is important in determining the degree of the services or disservices provided.
+While ecologists recognize the role of natural habitats such as field border vegetation in supporting biodiversity and providing potential ecosystem services like biological pest control, many farmers perceive these semi-arid habitat remnants as occupying valuable cropland or even as sources of pests.
+As such, documenting how pest and natural enemy populations interact with agricultural landscapes alongside semi-natural habitat buffers on the crop periphery is important in determining the degree of the services or disservices provided.
 
@@
 
-The wheat stem sawfly (WSS), Hymenoptera *Cephus cinctus* Norton, is a grass-mining sawfly native to North America [@Lesieur2016].
-Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as the practice became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 [@Morrill1996].
-Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [@Holmes1956].
-Economic damage manifests in the form of both kernel weight reduction caused by tissue damage [@Beres2011] and stem toppling due to late-season larval girdling [@Holmes1960].
-
-Several parsitoid species attack *C.
-cinctus* within wild grass hosts [@Davis1955; @Morrill1998], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill1998].
+The wheat stem sawfly (WSS), Hymenoptera *Cephus cinctus* Norton, is a grass-mining sawfly native to North America [@Criddle1922; Ainslie1929; @Wallace1966; @Lesieur2016].
+Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as this resource became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 in Montana [@Morrill1996], and more recently in Colorado [@Cockrell2021].
+Climate warming has the potential to further expand the range of this pest both northward and southward [@Olfert2019; @Weaver2023].
+Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [Roemhild1954; @Holmes1956].
+Economic damage manifests in the form of both kernel weight reduction caused by tissue damage [@Beres2011a; Delaney2010] and stem toppling due to late-season larval girdling [@Holmes1960; @Nansen2005a; @McCullough2020].
+
+Several parasitoid or predatory species attack *C.
+cinctus* within wild grass hosts [@Davis1955; @Morrill1998, @Morrill2001], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill1998; @Runyon2001a; @Runyon2001b].
 These two host-specific ectoparasitoids are native to North America.
-Both parasitoids are bivoltine, with the first generation present in Montana from late-June to late-July, while the second generation is active from mid-July to September [@Somsen1956].
+Both parasitoids are bivoltine, with the first generation present in Montana from late-June to late-July, while the second generation is active from mid-July to September [@Gahan1918; @Somsen1956; @Davis2013].
 Adult female *B.
 cephi* and *B.
-lissogaster* paralyze *C.
-cinctus* larvae within the host stem, paralyze the host larvae, and depositing one (*B.
-cephi*) to three (*B.
+lissogaster* inject a paralytic toxin into a *C.
+cinctus* larvae within the host stem, paralyzing the host larvae, and depositing from a single (*B.
+cephi*) to 1-3 (*B.
 lissogaster*) eggs.
@@
 lissogaster* have been shown to provide biological control via irreplaceable mortality of *C.
-cinctus* populations [@Buteler2015] and can significantly reduce *C.
-cinctus* damage and impact on crop yields [@Buteler2008; @Rand2020].
+cinctus* populations [@Buteler2015; @Peterson2011] and can significantly reduce *C.
+cinctus* damage and impact on crop yields [@Buteler2008; @Adhikari2019; @Bekkerman2018; @Rand2020].
 While *C.
-cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*) [@Wallace1966; @Cockrell2017], infestation levels can vary widely across hosts.
-Previous research has shown preference for host plants with larger stem diameter [@Ainslie1920; @Bhandari2020].
-
-Recent research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *B.
-inermis*, *Elymus trachycaulus*, and *Elytrigia intermedium* host *C.
-cinctus* and Bracon spp.
-parasitoids quite well [@Peirce2021; @Rand2024].
+cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*) [@Wallace1966; @Cockrell2017; @Achhami2020], infestation levels can vary widely across hosts.
+Previous research has shown preference for host plants with larger stem diameter [@Ainslie1920; @Buteler2009; @Bhandari2020].
+
+Recent research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *Bromus techtorum*, *Bromus inermis*, *Elymus trachycaulus*, and *Thinopyrum intermedium* host *C.
+cinctus* and most also host Bracon spp.
+parasitoids quite well [@Mendoza2006; @Peirce2021; @Rand2024].
 We find investigating *B.
 inermis* holds the most potential, as it has been shown to be a quality initially-accepted host for *C.
-cinctus* oviposition while also leading to high larval mortality later on [@Farstad1945].
-However, there is a lack of refined data relating to the total infestation, cutting, and parasitization within *B.
+cinctus* oviposition while also leading to high larval mortality later on [@Seamans1928; @Farstad1945].
+However, there is a lack of detailed relating to the total infestation, cutting, and parasitism within *B.
 inermis* at multiple time points throughout the growing season.
-Quantitative assessment of potential 'reservoire' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B.
+Quantitative assessment of potential 'reservoir' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B.
 inermis* along roadways and within coulees.
 
-Smooth brome (*Bromus inermis* Lyess) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Salesman2011].*B.
-inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Farstad1945].
+Smooth brome (*Bromus inermis* Leyss) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Bolwahn2011].*B.
+inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Criddle1922; @Farstad1945; @Beres2011a].
 Due to its low agricultural value, and propensity to form monocultures and out compete many native grasses, *B.
@@
 However, *B.
-inermis* still holds considerable ecological value in its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].
-
-Montana's golden triangle, an area of central and northern Montana with intense cereal grain production, is prone to drought and burning via wildfire [@Wilson1923].
+inermis* still holds considerable ecological and potential economic value due to its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].
+
+Montana's golden triangle, an area of central and northern Montana with intense cereal grain production, is prone to drought and burning via wildfire [@Wilson1923; @Adhikari2019].
 As both *C.
@@
 As with many predator-prey interactions, the *C.
-cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession.
-Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinellidae and aphids [@Evans1999].
-
-We examined historical weather data from two field sites in central and northern Montana.
-We used this data to understand how changes in global climate trends may be impacting *B.
-cephis* and *B.
+cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession and through ineteraction with drought [@Holmes1963; @Holmes1982].
+Similar examples of these classical population dynamics patterns can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinellidae and aphids [@Evans1999], among others.
+
+We examined historical weather data from two field locations (10 sites overall) in central and northern Montana.
+We used this data to understand how changes in global climate trends may be impacting *C.
+cinctus*, *B.
+cephi* and *B.
+lissogaster* interact with *B.
 inermis* populations.
 In addition we analyzed normalized difference vegetative index (NDVI) at our field sites to understand how *B.
-inermis* senescence patterns compare to adjacent cultivated wheat.
+inermis* senescence patterns compare to adjacent cultivated wheat fields.
 
 In this study, we explored the ecological patterns between Montana populations of *C.
-cinctus* and its braconid parasitoids in relation to *B.
+cinctus* and its associated braconid parasitoids in relation to *B.
 inermis* host sites.
@@
 cinctus* infested *B.
-inermis* stems in relation to neighboring cereal crops, and 2) the way in which braconid parasitoids utilized WSS hosts within *B.
+inermis* stems in relation to neighboring cereal crops, and 2) the way in which braconid parasitoids utilized WSS hosts available within *B.
 inermis* stems.
@@
 cinctus* within the *B.
-inermis* stems.
-The second study examined rates of cutting and parasitism at field sites in central and northern Montana.
-
-Using these studies, we addressed three questions, 1) what is the total cutting heading into overwinter, and thus source rate, of the subsequent *C.
-cinctus* population from *B.
-inermis*, 2) can *C.
+inermis* stems in the absence of parasitoids.
+The second study examined rates of cutting and parasitism in wheat crops and their surroundings at field sites in central and northern Montana.
+
+Using these studies, we addressed three questions, 1) what is the total *C.
+cinctus*-caused stem cutting heading into overwinter - and thus the source rate, of the subsequent *C.
+cinctus* population in *B.
+inermis*?
+2) can *C.
 cinctus* larvae within *B.
@@
 cephi* and *B.
-lissogaster*, and 3) how do annual *C.
+lissogaster*?
+and 3) how do annual *C.
 cinctus* and parasitoid populations in *B.
-inermis* compare with those in adjacent cultivated wheat fields.
+inermis* compare with those in the adjacent cultivated wheat fields?
 
@@
 
+All data for experiments was analyzed using R Studio (R Studio version 2024.04.0+735, R 2024).
+
 ### Weather data and NDVI analysis
 
-*Weather data.* We assessed the long and medium term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA).
+*Weather data.* We assessed the medium- and long-term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA).
 Data for each field site was averaged from three of the closest weather stations to that location.
-Data was plotted using R Studio (R Studio version 2024.04.0+735, R 2024) package 'ggplot' (version 3.4.4) [@ggplot].
+Data was plotted using R package 'ggplot' (version 3.4.4) [@ggplot].
 Data was then fit using a linear model using the 'lm' command using average yearly precipitation (inches) as the response variable and year as the predictor.
@@
 inermis* using the normalized difference vegetation index (NDVI).
-NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infared light reflected by vegetation [@Pettorelli2005].
-Data was downloaded using Google Earth Engine (Google Inc.
+NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infrared light reflected by vegetation [@Pettorelli2005].
+NDVI data measurements were directly downloaded using Google Earth Engine (Google Inc.
 2023, Mountain View, CA, USA).
+Similar to our weather data analysis, NDVI data was modeled using the linear modeling package 'lm' in R to discern differences in NDVI measurments over time.
 
@@
 inermis* were assessed using a 34 x 60 ft plot at the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
+Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an experimental farm of Montana State University in Bozeman, MT.
 There is a negligible population of *C.
-cinctus* at this locaiton.
-The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
+cinctus* and associated parasitoids at this location.
+Experimental cages  were built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
 Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.
@@
 Wheat stem stubble was collected in Three Forks, MT, USA (43°38'19.39"N, 116°14'28.86"W) from fields that experienced high levels of *C.
-cinctus* infestation and cutting the year prior.
+cinctus* infestation and cutting the year prior, as previously described [@Hager2024].
 Cut stubble, which contained *C.
@@
 inermis* stems were collected from each cage.
-Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.
+Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae and live larvae, as well as very rare parasitism events, for each internode.
 
@@
 Samples were collected from wheat fields and adjacent *B.
-inermis* in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA.
+inermis* in early July and late August in 2022, and 2023 from sites in Big Sandy and Moccasin MT, USA.
 Sampling sites were set up as 100$m^2$ polygons along the edge of adjoining wheat fields.
@@
 Wheat and *B.
-inermis* stems were then returned to Montana State University, Bozeman, Montana and stored in a 10°C cold wet storage until dissection.
-Stems were dissected lengthwise with a fine-bladed scalpel to determine presence or absence of *C.
-cinctus* larvae infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting.
+inermis* stems were then returned to Montana State University, Bozeman, Montana and stored in a 10°C cold, humid storage until dissection.
+Stems were dissected in the same manner as samples from our controlled experiment.
+We recorded presence or absence of *C.
+cinctus* larval infestation, live eggs, dead eggs, dead larvae, live larvae and parasitism, in addition to stem cutting.
 Sawfly larvae were identified based on descriptions in Criddle (1915) and Wallace and McNeal (1996).
@@
 
-To better understand the density of *Bracon* spp.
-within the wheat and *B.
-inermis* sampling sites, we converted our units to the ratio of parasitoids to surviving hosts stem per unit area.
-Densities of stems are different when considering *B.
-inermis* and cultivated wheat or barley.
-Using this conversion, we attempted to better understand how the density of parasitoids (and WSS) is changed by looking at each kind of plant.
+To better understand the densities of host *C.
+cinctus* and the *Bracon* spp.
+within the wheat and adjacent *B.
+inermis* at each sampling site, we converted our units to the ratio of parasitoids to surviving hosts stem per unit area.
+Densities of stems are different when considering semi-natural, long-standing stands of *B.
+inermis* and the adjacent cultivated wheat or barley.
 
@@
        x = NULL,
-       y = "Average Yearly Precipitation (in)",
-       title = "Average Yearly Precipitation",
-       subtitle = "Big Sandy & Moccasin, MT, USA")
+       y = "Average Yearly Precipitation (in)")
 ```
@@
        x = NULL,
-       y = "Mean Yearly Temperature (°C)",
-       title = "Mean Yearly Temperature",
-       subtitle = "Moccasin and Big Sandy, MT, USA")
+       y = "Mean Yearly Temperature (°C)")
 
@@
        x = NULL,
-       y = "Average Yearly Temperature (°C)",
-       title = "Average Yearly Temperature",
-       subtitle = "Moccasin, MT, USA")
+       y = "Average Yearly Temperature (°C)")
 
@@
 We observed a significant linear relationship (*r = `r round(both_prcp_lm$r.squared,3)`, P = `r round(both_prcp_lm$coefficients[2,4],3)`, estimate = `r round(both_prcp_lm$coefficients[2,1],3)`*) between average precipitation and year for both Big Sandy and Moccasin, Montana.
-For each one-year increase, we are seeing a decrease in 0.05 inches of precipitation.
+For each location and one-year increase, we observed a 0.05 inch decrease in precipitation.
 ([Fig.
@@
        y = "NDVI",
-       linetype = NULL,
-       title = "Seasonal NDVI Comparison",
-       subtitle = expression(paste(italic("B. inermis"),
-                                   " and adjacent spring wheat"))) +
+       linetype = NULL) +
   scale_linetype_discrete(labels = c(expression(italic("B. inermis")),
@@
 ```{r fig4_inf_cut_plot code}
+pf.comb.hl <-
+pf.comb %>%
+  filter(!treatment == "control")
+
 inf_cut_plot <-
@@
            position = position_dodge(),
-           data = pf.comb %>%
+           data = pf.comb.hl %>%
              filter(type == "p.inf"),
@@
            position = position_dodge(),
-           data = pf.comb %>%
+           data = pf.comb.hl %>%
              filter(type == "cut"),
            aes(year, mean, fill = "cut")) +
-  geom_errorbar(data = pf.comb,
+  geom_errorbar(data = pf.comb.hl,
                 aes(x = year,
@@
 Stem cutting by mature larvae, leading into winter, was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P < 0.05*).
+There was a dramatic decrease in survival observed from initial infestation to stem cutting.
 
@@
   labs(x = "Stem Internode",
-       y = "Percent of internodes infested",
-       title = expression(paste("Percentage of ",
-                                italic("B. inermis "),
-                                "Internodes Infested")),
-       subtitle = "Controlled infestation") +
+       y = "Percent of internodes infested") +
   facet_wrap(~year)
@@
 cinctus* infested stems often contained larvae burrowing through multiple nodes.
-In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through.
+In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through, or 5 internodes occupied.
 In 2023, `r round(pf_nodes_over_4_table[2,2],3)*100`% of stems had at least 4 nodes infested [Fig.
@@
 parasitoid presence in 6,148 wheat and grass stems across 10 research sites in 2022 and 2023.
-Initial infestation by *C.
-cinctus* within *B.
-inermis* varied between collection sites, with the greatest infestation taking place within our three Big Sandy, MT sampling sites (2023: 65.4%, 2022: 63.1%), while the lowest infestation observed was in Moccasin, MT (2023: 40.8%, 2022: 60.7%).
+Initial infestation of *B.
+inermis* by *C.
+cinctus* varied between collection sites, with the greatest infestation taking place within our three Big Sandy, MT sampling sites (2023: 65.4%, 2022: 63.1%), while the lowest infestation observed was in Moccasin, MT (2023: 40.8%, 2022: 60.7%).
 Across all sites in Big Sandy and Moccasin, we observed an average infestation of 57.5% within *B.
@@
 Cutting by *C.
-cintus* was observed at an average rate of 5.59% (se = 0.0164) in *B.
-inermis* and 47.1% (se = 0.0386) in the adjacent cultivated wheat ([Fig.
+cintus* was observed at an average rate of 5.59% (se = 0.02) in *B.
+inermis* and 47.1% (se = 0.04) in the adjacent cultivated wheat ([Fig.
 7](#figure7).
@@
 inermis* late in the growing season, we calculated the number of stems that contained at least four or more nodes with burrowing injury.
-In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of burrowing injury.
+In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of burrowing injury at our second sampling date.
 In Moccasin, the corresponding figure was 66.5%.
@@
        y = "Percentage of nodes infested",
-       fill = "Collection",
-       title = expression(paste("Percentage of ",italic("B. inermis")," Internodes Infested")),
-       subtitle = "Big Sandy and Moccasin, MT, USA")
+       fill = "Collection")
 
@@
   labs(x = NULL,
-       y = "Percentage (%)",
-       title = "B. inermis and Adjacent Wheat")
+       y = "Percentage (%)")
 
@@
        y = "% of stems cut",
-       title = expression(paste("Stem Cutting by ", italic("C. cinctus")," in Central and Northern Montana")),
-       subtitle = "Moccasin and Big Sandy, MT, USA",
        fill = "Plant Type") +
@@
   labs(x = NULL,
-       y = "Ratio of parasitized stems to cut stems",
-       title = "Parasitized to Cut Stems Ratios",
-       subtitle = "Big Sandy and Moccasin, MT, USA") +
+       y = "Ratio of parasitized stems to cut stems") +
   scale_fill_brewer(palette = "Greys") +
@@
 inermis* sink/source ecology.
-The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to a single cut stem.
+The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to a single *C.
+cinctus*-cut stem.
 Thus, for every square meter of *B.
@@
 cinctus*.
-All three other sample selections also showed positive ratios of parasitoids to cut stems [Fig.
-8](#figure8).
+All other sample locations also showed positive ratios of parasitoids to cut stems [(Fig.
+8)](#figure8).
 
@@
 cinctus* and its associated parasitoids within *B.
-inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas serve as sink or source origins for both pests and beneficial natural adversaries infiltrating agricultural crops.
+inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas serve as sink or source origins of the pest and its beneficial natural adversaries infiltrating agricultural crops.
 As the climate of central and northern Montana continues to change, wheat stem sawfly infestation will continue to impact cereal crop yields.
@@
 Results from the controlled *B.
-inermis* infestation confirmed high *C.
-cinctus* larval morality prior to overwintering [@Shanower2004; @Buteler2008] with 53% of stems exhibiting *C.
+inermis* infestation confirmed large amounts of *C.
+cinctus* larval mortality prior to overwintering [@Shanower2004; @Mendoza2006; @Buteler2008] with 53% of total collected stems exhibiting *C.
 cinctus* burrowing damage.
-Of those infested stems, a large proportion exhibited larval damage in at least 4 nodes (62.5%).
-While stem collection and dissection did not align with the establishment of a timeline of living larvae (stems dissected in October), we can assert that stems with at least 4 nodes of infestation would have a high probability of being alive, and thus a viable potential parasitoid host, in late July and August of the same year.
+Of those infested stems, a large proportion exhibited larval damage in at least 5 internodes (62.5%).
+While the timing of stem collection and dissection did not align with precise measurement of living larvae (stems dissected in October), we can assert that stems with at least 4 nodes burrowed through would have a high probability of containing large, live larvae, and thus a viable potential parasitoid host, in late July and August of the same year.
 
@@
 However, there was minimal survival of *C.
-cinctus* larvae in August and September, suggesting that there will be little adult emergence from *B.
+cinctus* larvae beyond August and September, confirming that there will be little adult emergence from *B.
 inermis* the following season.
@@
 *C.
-cinctus* larvae are cued via the thinning of the cell membranes during plant senescence to travel to the base of the stem and prepare for obligatory diapause [@Beres2011].
+cinctus* larvae are cued via the thinning of the cell membranes [@Villacorta1971] during plant senescence to travel to the base of the stem and prepare for obligatory diapause [@Beres2011].
 Because *B.
@@
 inermis* to be a quality host of both *C.
-cinctus* and *Bracon* spp parasitoids, but little data is available on the season fluctuation of *B.
+cinctus* and *Bracon* spp parasitoids [@Criddle1922; @Peirce2021; @Rand2024], but little data is available on the season fluctuation of *B.
 inermis* utilization [@Criddle1922; @Peirce2021; @Rand2024].
-Because we collected stem samples twice during the growing season, we were able to compare parasitoid activity.
-We hoped to understand how parasitism rates changed from July (pre-harvest) and September of each year.
+Because we collected stem samples twice during the growing season, we were able to compare levels of parasitism.
+We hoped to understand how parasitism rates changed from July (pre-wheat harvest) and September (post-wheat harvest) of each year.
 
@@
 inermis* stems.
-In Big Sandy (76.2%) and Moccasin (66.5%), stems that exhibited four or more nodes of burrowing damage correlates similarly to the assertion that *C.
-cinctus* larvae are available within *B.
-inermis* stems past wheat harvest.
+In Big Sandy (76.2%) and Moccasin (66.5%), stems that exhibited four or more nodes burrowed suggests that *C.
+cinctus* larvae remain available within *B.
+inermis* stems past wheat harvest, when they are more limited in availability.
 Most importantly, in years of drought or increased temperatures where wheat senescence and harvest may occur earlier in the summer.
@@
 inermis*.
+Three treatment groups - high, low, and control (0) - were used.
+The controlled groups showed no sign of *C.
+cinctus* stem damage.
 We observed a significant difference in cutting between high and low treatment groups (*r = 0.592, P < 0.05*)."}

### 2024-06-13 qmd/projects/msu/wss/convert.Rmd
update

--- 2024-05-17 qmd/projects/msu/wss/convert.Rmd
@@
        output_file = paste0("JacksonStrand_Paper 1.docx"))
-render("paper1.rmd", output_format = "pdf_document",
-       output_file = paste0("JacksonStrand_Paper 1.pdf"))
+# render("paper1.rmd", output_format = "pdf_document",
+#        output_file = paste0("JacksonStrand_Paper 1.pdf"))
 

### 2024-06-13 qmd/projects/msu/wss/paper1.rmd
update

@@
 Wheat stem sawfly, *Cephus cinctus* Norton (WSS), causes significant damage in cereal crops throughout the Northern Great Plains of North America.
-Bracon cephi and Bracon lissogaster are native WSS parasitoids important in managing WSS outbreaks and damage.
-Smooth Brome (Bromus inermis) grass has been found to be an effective parasitoid sink, and potential trap crop, when grown in areas surrounding wheat fields in Montana.
+*Bracon cephi* and *B.
+lissogaster* are native WSS parasitoids important in managing WSS outbreaks and damage.
+Smooth Bromegrass (*Bromus inermis*) grass has been found to be an effective parasitoid sink, and potential trap crop, when grown in areas surrounding wheat fields in Montana.
 As climate changes continue to alter central and northern Montana, WSS will continue to cause immense damage to dry-land cereal crop yields.
-My study observes the utilization of smooth brome in providing year-over-year host refuge for Bracon spp.
-We measured the WSS larval survival rate within controlled smooth brome plots, observing a maximum end-of-year survival of 5.7%.
-We also collected stems from central and northern Montana where we measured the WSS infestation and parasitoid prevalence within cultivated wheat and adjacent smooth brome.
-Montana fields sites showed similar high year-end WSS larval mortality while showcasing high levels of parasitism.
+We analyzed both local temperature and precipitation trends at our sample sites, and found statistically significant increases in temperature and decreases in precipitation over the past fifty years.
+Our study observes the utilization of smooth brome in providing year-over-year host refuge for Bracon spp parasitoids.
+We measured the WSS larval infestation and survival rate within controlled smooth brome plots, observing a maximum infestation of 66.5% and a maximum end-of-year survival of 5.7%.
+We also collected stems from sites in central and northern Montana where we measured the WSS infestation and parasitoid prevalence within cultivated wheat and adjacent smooth brome.
+Montana fields sites showed high levels of *C.
+cinctus* infestation within both *B.
+inermis* and adjacent wheat crop in Big Sandy (64.5% *B.
+inermis*, 65.7% adjacent wheat) and Moccasin (50.6%, 38.6%).
+We observed high year-end WSS larval mortality measured as stem cutting of infested stems with *B.
+inermis* when compared to adjacent wheat in Big Sandy (2.9% *B.
+inermis*, 46.5% adjacent wheat) and Moccasin (6.4%, 49.4%).
+Infested stems exhibited parasitism in both Big Sandy (7.11% *B.
+inermis*, 8.6% adjacent wheat) and Moccasin (6.11%, 2.09%).
 This research underlines the importance of providing a sustainable ecological buffer for WSS parasitoids as climate conditions continue to change.
@@
 Smooth brome (*Bromus inermis* Leyss) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Bolwahn2011].*B.
-inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Criddle1922; @Farstad1945; @Beres2011a].
+inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Criddle1922; @Seamans1928; @Farstad1945; @Beres2011a].
 Due to its low agricultural value, and propensity to form monocultures and out compete many native grasses, *B.
@@
 
-All data for experiments was analyzed using R Studio (R Studio version 2024.04.0+735, R 2024).
-
-### Weather data and NDVI analysis
-
-*Weather data.* We assessed the medium- and long-term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA).
-Data for each field site was averaged from three of the closest weather stations to that location.
-Data was plotted using R package 'ggplot' (version 3.4.4) [@ggplot].
-Data was then fit using a linear model using the 'lm' command using average yearly precipitation (inches) as the response variable and year as the predictor.
-
-*NDVI analysis.* We compared the relative greening throughout the growing season between wheat fields and adjacent *B.
-inermis* using the normalized difference vegetation index (NDVI).
-NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infrared light reflected by vegetation [@Pettorelli2005].
-NDVI data measurements were directly downloaded using Google Earth Engine (Google Inc.
-2023, Mountain View, CA, USA).
-Similar to our weather data analysis, NDVI data was modeled using the linear modeling package 'lm' in R to discern differences in NDVI measurments over time.
-
 ### Controlled *C. cinctus* infestation of *B. inermis*
@@
 inermis* stems were collected from each cage.
-Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae and live larvae, as well as very rare parasitism events, for each internode.
+Each stem was dissected using X-Acto® knives to collect data on infestation, dead larvae and live larvae, as well as very rare parasitism events, for each internode.
 
 ### Montana Field Survey
+
+*Weather data and NDVI* Field sites were chosen across two counties, Choteau and Judith Basin, within the 'Golden Triangle' of Montana, United States [fig.
+1](#figure1), areas that consistently experience high *C.
+cinctus* pressure.
+We assessed the medium- and long-term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA).
+Data for each field site was averaged from three of the closest weather stations to that location.
+
+To better understand how precipitation patterns vary throughout the growing season, we looked at local daily precipitation totals from weather stations at our sampling sites in Big Sandy and Moccasin, MT.
+Data was pulled online from historical WeatherUnderground weather station data [@WeatherUnderground].
+Precipitation readings were broken into two categories - dates before July 15 and after September 15, as well as between July 15 and September 15.
+These dates were chosen based on typical wheat harvest and historical periods of known *Bracon* spp activity.
+In the 'Golden Triangle', typical spring wheat harvest takes place in mid-July on average [@Wilson1923; @Zhu2021].
+
+We used the normalized difference vegetation index (NDVI) to compare the relative greening throughout the growing season between wheat fields and adjacent *B.
+inermis* ([fig.
+2](#figure2)).
+NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infrared light reflected by vegetation [@Pettorelli2005].
+NDVI data measurements were directly downloaded using Google Earth Engine (Google Inc.
+2023, Mountain View, CA, USA).
 
@@
 inermis* and adjacent wheat fields.
-Sites were chosen across 2 counties in north-central Montana, United States.
-(Chouteau, Judith Basin), which consistently experience high *C.
-cinctus* pressure.
 Samples were collected from wheat fields and adjacent *B.
@@
 
-*Statistical Analysis.* We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion of stems parasitized by *Bracon* spp.
-
 To better understand the densities of host *C.
@@
 inermis* and the adjacent cultivated wheat or barley.
+
+### Statistical Analysis
+
+All statistical analyses were done in R [@R2024] using R Studio (version 4.3.2).
+Data was plotted using R package 'ggplot' (version 3.4.4) [@ggplot].
+We analyzed long- and short-term weather trends using linear modeling and the 'lm' command.
+Weather data was analyzed using either average yearly precipitation (inches) or average yearly temperature (°C) as the response variable and year as the predictor.
+Similar to our weather data analysis, NDVI data was linear modeled to discern differences in NDVI measurements over time.
+
+We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion of stems parasitized by *Bracon* spp.
 
@@
 library(lme4)
-```
-
-### Historical Weather Analysis & NDVI
+library(gridExtra)
+```
+
+### Controlled *C. cinctus* Infestation of *B. inermis*
+
+```{r column.order}
+column.order <- c("distance", "collection", "p.inf", 
+                  "p.cut", "p.par", "poten")
+
+column.order2 <- c("location", "distance", "collection", "p.inf", 
+                  "p.cut", "p.par", "poten")
+
+round_based_on_size <- function(x) {
+  threshold <- 0.01
+  rounded_values <- ifelse(abs(x) > threshold, round(x, 2), round(x, 3))
+  return(rounded_values)
+}
+```
+
+```{r (pf.RAW) post farm n.inf data}
+pf.raw.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2022/Post\ Farm\ 2022.xlsx", sheet = "n.inf")
+pf.raw.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "n.inf")
+```
+
+```{r (pf.inf.23) post farm inf data}
+pf.inf.23 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf") %>%
+  rename(n.1 = 8, n.2 = 9, n.3 = 10, n.4 = 11,
+         n.5 = 12, n.6 = 13, n.7 = 14, n.8 = 15,
+         n.9 = 16, n.10 = 17)
+```
+
+```{r (pf.comb) post farm data combined}
+# 2022
+pf.2022.site <-
+pf.raw.2022 %>%
+  group_by(treatment, site) %>%
+  reframe(site, treatment,
+            p.inf = sum(iv, ir, ix)/sum(uv, ur, iv, ir, ix),
+            cut = sum(cut)/sum(total),
+            total.stems = sum(iv, ir, ix, uv, ur)) %>%
+  unique() %>%
+  pivot_longer(-c(treatment, total.stems, site),
+               names_to = "type",
+               values_to = "percent") %>%
+  mutate(year = "2022")
+
+# 2023
+pf.2023.site <-
+pf.raw.2023 %>%
+  group_by(treatment, site) %>%
+  reframe(treatment, site, 
+          t.inf = sum(n.iv, n.ir, n.ix),
+          p.inf = sum(n.iv, n.ir, n.ix)/sum(n.uv, n.ur, 
+                                            n.iv, n.ir, n.ix),
+          cut = sum(cut)/sum(total),
+          total.stems = sum(n.iv, n.ir, n.ix, n.uv, n.ur)) %>%
+  distinct() %>%
+  pivot_longer(-c(treatment, total.stems, site),
+               names_to = "type",
+               values_to = "percent") %>%
+  filter(!type == "t.inf") %>%
+  arrange(type) %>%
+  mutate(year = "2023")
+
+pf.comb <-
+  pf.2022.site %>%
+  rbind(pf.2023.site) %>%
+  group_by(treatment, year, type) %>%
+  mutate(count = n(),
+         average = mean(percent),
+         sd = sd(percent),
+         se = (sd/sqrt(count))) %>%
+  reframe(treatment, year, type,
+          mean = mean(percent),
+          sd = mean(sd),
+          se = mean(se)) %>%
+  unique()
+
+```
+
+```{r (pf.total) pf total stems}
+# 2022 totals
+pf.2022.total <-
+  pf.raw.2022 %>%
+  group_by(treatment) %>%
+  reframe(total = sum(uv, ur, iv, ir, ix)) %>%
+  mutate(year = "2022")
+
+# 2023 totals
+pf.2023.total <-
+  pf.raw.2023 %>%
+  group_by(treatment) %>%
+  reframe(total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
+  mutate(year = "2023")
+
+pf.total.comb <-
+  pf.2022.total %>%
+  rbind(pf.2023.total)
+
+```
+
+```{r pf_lm}
+raw.23 <-
+pf.raw.2023 %>%
+  dplyr::select(site, treatment,
+                n.iv, n.ir, n.ix, n.ur, n.uv,
+                cut) %>%
+  rename(iv = n.iv,
+         ir = n.ir,
+         ur = n.ur,
+         ix = n.ix,
+         uv = n.uv) %>%
+  mutate(year = 2023)
+
+raw.22 <-
+pf.raw.2022 %>%
+  dplyr::select(site, treatment, 
+                ix, ir, iv, ur, uv,
+                cut) %>%
+  mutate(year = 2022)
+
+pf_inf_lm <-
+raw.23 %>%
+  rbind(raw.22) %>%
+  group_by(site, treatment, year) %>%
+  reframe(total = sum(iv, ir, ix, ur, uv),
+          inf = sum(iv, ir, ix)) %>%
+  lm(inf ~ treatment + year, data = .) %>%
+  summary()
+
+pf_cut_lm <-
+raw.23 %>%
+  rbind(raw.22) %>%
+  group_by(site, treatment, year) %>%
+  reframe(total = sum(iv, ir, ix, ur, uv),
+          inf = sum(iv, ir, ix),
+          cut = sum(cut)) %>%
+  lm(cut ~ treatment + year, data = .) %>%
+  summary()
+```
+
+```{r pf_per, eval = F}
+raw.23 %>%
+  rbind(raw.22) %>%
+  group_by(treatment) %>%
+  reframe(total = sum(iv, ir, ix, ur, uv),
+          inf = sum(iv, ir, ix),
+          cut = sum(cut),
+          p.inf = inf/total,
+          p.cut = cut/inf)
+```
+
+*C.
+cinctus* heavily infested stems of *B.
+inermis* in the controlled test conditions.
+Averaged across both years, we observed 66.5% of stems infested for high treatments and 47.3% of stems for low treatments.
+We found strong evidence suggesting that there was a significant difference between infestation at high and low treatment levels when holding year constant (*r = `r round(pf_inf_lm$r.squared,3)`, P \< 0.05*, [Fig.
+3](#figure3)).
+
+```{r fig4_inf_cut_plot code}
+pf.comb.hl <-
+pf.comb %>%
+  filter(!treatment == "control")
+
+inf_cut_plot <-
+ggplot() +
+  facet_wrap(~treatment) +
+  geom_bar(stat = "identity",
+           position = position_dodge(),
+           data = pf.comb.hl %>%
+             filter(type == "p.inf"),
+           aes(year, mean, fill = "p.inf")) +
+  geom_bar(stat = "identity",
+           position = position_dodge(),
+           data = pf.comb.hl %>%
+             filter(type == "cut"),
+           aes(year, mean, fill = "cut")) +
+  geom_errorbar(data = pf.comb.hl,
+                aes(x = year,
+                    ymin = mean - se,
+                    ymax = mean + se),
+                position = position_dodge(0.9),
+                width = 0.2,
+                alpha = 0.5) +
+  theme_bw() +
+  labs(x = element_blank(),
+       y = "Percentage of stems (%)",
+       fill = element_blank()) +
+  scale_fill_brewer(palette = "Greys",
+                    labels = c("Cut", "Infested"))
+```
+
+Stem cutting by mature larvae, leading into winter, was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P < 0.05*).
+There was a dramatic decrease in survival observed from initial infestation to stem cutting.
+
+```{r figure_5_pf_inf_nodes}
+pf.y.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2022.xlsx", sheet = "inf")
+pf.y.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf")
+
+# DATA FOR THE PLOT
+pf.nodes.plot.data <-
+pf.y.2023 %>%
+  dplyr::select(-c(Name, 'Date Proc')) %>%
+  dplyr::select(treatment, 6:15) %>%
+  pivot_longer(-c(treatment),
+               names_to = "node",
+               values_to = "value") %>%
+  mutate(node = gsub("n_", "", node)) %>%
+  na.omit() %>%
+  mutate(year = "2023") %>%
+  rbind(pf.y.2022 %>%
+          dplyr::select(-c(site1, site2, site12, site3,
+                           Name, 'Date Proc')) %>%
+          dplyr::select(treatment, 
+                        2, 3, 4, 5, 6, 
+                        7, 8, 9, 10, 11) %>%
+          pivot_longer(-c(treatment),
+                       names_to = "node",
+                       values_to = "value") %>%
+          na.omit() %>%
+          mutate(year = "2022")) %>%
+  group_by(year, treatment, node) %>%
+  reframe(count = n()) %>%
+  group_by(year, treatment) %>%
+  mutate(per = count/sum(count)) %>%
+  mutate(node = as.numeric(node))
+
+# PLOT
+pf_nodes_plot <-
+  pf.nodes.plot.data %>%
+  ggplot() +
+  geom_bar(stat = 'identity',
+           aes(x = node, y = per, fill = treatment),
+           position = position_dodge(),
+           color = "black") +
+  theme_bw() +
+  scale_fill_brewer(palette = "Greys") +
+  scale_x_continuous(breaks = 1:10) +
+  labs(x = "Stem Internode",
+       y = "Percent of internodes infested") +
+  facet_wrap(~year)
+
+pf_nodes_over_4_table <-
+pf.nodes.plot.data %>%
+  group_by(year, node) %>%
+  reframe(count = sum(count)) %>%
+  group_by(year) %>%
+  mutate(per = count/sum(count)) %>%
+  filter(node > 4) %>%
+  group_by(year) %>%
+  reframe(sum = sum(per))
+```
+
+*C.
+cinctus* infested stems often contained larvae burrowing through multiple nodes.
+In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through, or 5 internodes occupied.
+In 2023, `r round(pf_nodes_over_4_table[2,2],3)*100`% of stems had at least 4 nodes infested [Fig.
+4](#figure4)
+
+### Montana Field Sites
+
+#### Historical Weather Analysis & NDVI
 
@@
 
+```{r bs_seasonal_plot}
+lochy_precip <-
+read_excel("weather/precip_scrape.xlsx", sheet = "big_sandy")
+
+line_table <-
+lochy_precip %>%
+  dplyr::select(date, prcp) %>%
+  mutate(month = month(date),
+         day = day(date),
+         year = as.factor(year(date))) %>%
+  mutate(new_date = make_date(year = 0,
+                              month = month,
+                              day = day)) %>%
+   mutate(category = case_when(new_date < "0000-07-15" ~ "pre",
+                              new_date > "0000-09-15" ~ "post",
+                              new_date >= "0000-07-15" & new_date <= "0000-09-15" ~ "mid")) %>%
+  mutate(model_category = if_else(category == "mid", 
+                                  "mid", "no"))
+
+line_table_average <-
+  line_table %>%
+  group_by(month, day) %>%
+  reframe(category, new_date,
+          prcp = mean(prcp))
+
+category_averages <-
+line_table_average %>%
+  na.omit() %>%
+  group_by(category) %>%
+  reframe(prcp = mean(prcp))
+
+bs_seasonal_lm <-  
+line_table %>%
+  lm(prcp ~ model_category + year, data = .) %>%
+  summary()
+
+mult = 3
+
+smooth_plot <-
+ggplot() +
+  geom_smooth(aes(x = new_date, y = prcp),
+              data = line_table_average,
+              se = FALSE,
+              method = 'loess',
+              color = "black") +
+  geom_rect(aes(xmin = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0528 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-07-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0528 * mult, yend = 0.0528 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0164 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-09-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0164 * mult, yend = 0.0164 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-10-31", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0281 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-10-31", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0281 * mult, yend = 0.0281 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  theme_bw()
+
+bs_seasonal_plot <-
+ggplot() +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "pre"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "post"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "mid"),
+            color = "black") +
+  geom_rect(aes(xmin = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0528 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-07-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0528 * mult, yend = 0.0528 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0164 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-09-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0164 * mult, yend = 0.0164 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-10-31", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0281 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-10-31", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0281 * mult, yend = 0.0281 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  labs(x = NULL,
+       y = "Average Daily Precipitation (in)",
+       color = NULL) +
+  scale_y_continuous(sec.axis = sec_axis(~ . * mult, 
+                                         name = "Average Seasonal Precpitation (in)")) +
+  scale_x_date(
+    date_breaks = "1 month",
+    date_labels = "%B") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r moc_seasonal_plot}
+moc_precip <-
+read_excel("weather/precip_scrape.xlsx", sheet = "moccasin")
+
+line_table <-
+moc_precip %>%
+  dplyr::select(date, prcp) %>%
+  mutate(month = month(date),
+         day = day(date),
+         year = as.factor(year(date))) %>%
+  mutate(new_date = make_date(year = 0,
+                              month = month,
+                              day = day)) %>%
+   mutate(category = case_when(new_date < "0000-07-15" ~ "pre",
+                              new_date > "0000-09-15" ~ "post",
+                              new_date >= "0000-07-15" & new_date <= "0000-09-15" ~ "mid")) %>%
+  mutate(model_category = if_else(category == "mid", 
+                                  "mid", "no"))
+
+line_table_average <-
+  line_table %>%
+  group_by(month, day) %>%
+  reframe(category, new_date,
+          prcp = mean(prcp))
+
+category_averages <-
+line_table_average %>%
+  na.omit() %>%
+  group_by(category) %>%
+  reframe(prcp = mean(prcp))
+
+moc_seasonal_lm <-  
+line_table %>%
+  lm(prcp ~ model_category + year, data = .) %>%
+  summary()
+
+mult = 3
+
+moc_seasonal_plot <-
+ggplot() +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "pre"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "post"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "mid"),
+            color = "black") +
+  geom_rect(aes(xmin = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0528 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-07-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0528 * mult, yend = 0.0528 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0164 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-09-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0164 * mult, yend = 0.0164 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-10-31", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0281 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-10-31", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0281 * mult, yend = 0.0281 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  labs(x = NULL,
+       y = "Average Daily Precipitation (in)",
+       color = NULL) +
+  scale_y_continuous(sec.axis = sec_axis(~ . * mult, 
+                                         name = "Average Seasonal Precpitation (in)")) +
+  scale_x_date(
+    date_breaks = "1 month",
+    date_labels = "%B") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+In both Big Sandy and Moccasin, MT, we observed a decrease in average yearly precipitation and increase in average yearly temperature.
 We observed a significant linear relationship (*r = `r round(both_prcp_lm$r.squared,3)`, P = `r round(both_prcp_lm$coefficients[2,4],3)`, estimate = `r round(both_prcp_lm$coefficients[2,1],3)`*) between average precipitation and year for both Big Sandy and Moccasin, Montana.
@@
 ([Fig.
-1](#figure1)).
+5](#figure5)).
 Additionally, we observed a significant positive linear relationship (*r = `r round(both_temp_lm$r.squared,3)`, P \< 0.05, estimate = `r round(both_temp_lm$coefficients[2,1],3)`*) between year and average yearly temperature for Moccasin and Big Sandy, MT.
@@
 ([Fig.
-2](#figure2)).
+6](#figure6)).
+Analyzing local, precipitation data over the past five years yielded information on how central and northern Montana experience relative dryness during the late-summer months.
+In Big Sandy, when controlling for year, we saw strong evidence of a significant difference (*p<0.05, F=2.067*, [fig.
+7](#figure7)) between daily rainfall in months prior July 15 and after September 15 compared to days within the July 15 to September 15 'late-summer' period.
+Analysis of Moccain precipitation data revealed that there is little evidence (*p=0.08, F=1.535*) to suggest a difference between the two date categories.
 
@@
 
-label_x <- as.Date("2023-07-20")
+label_x <- as.Date("2023-07-15")
 label_y <- 0.75
-line_x <- as.numeric(as.Date("2023-07-20"))
+line_x <- as.numeric(as.Date("2023-07-15"))
 
@@
              size = 0.75) +
-  geom_line(aes(x = date, y = ndvi, color = plant_type, linetype = plant_type),
+  geom_line(aes(x = date, y = ndvi, 
+                color = plant_type, linetype = plant_type),
             alpha = 0.5,
@@
   geom_label(aes(x = label_x,
-                 y = 0.75),
-             label = "Spring wheat harvest",
+                 y = 0.8),
+             label = "Average spring wheat harvest",
              hjust = -0.1,
              size = 3) +
-  geom_segment(aes(x = as.Date("2023-07-20")+5, 
-                   y = 0.75, 
-                   xend = as.Date("2023-07-20"), 
-                   yend = 0.75)) +
   theme_bw() +
   guides(color = FALSE) +
-  labs(x = "2023 Growing Season",
+  labs(x = NULL,
        y = "NDVI",
        linetype = NULL) +
-  scale_linetype_discrete(labels = c(expression(italic("B. inermis")),
-                                                      "Spring Wheat"))
+  scale_linetype_discrete(labels = c(expression(italic("B. inermis")),"Spring Wheat"))
 ```
@@
 inermis* and spring wheat.
-We saw a significant difference in the July NDVI (*`r round(ndvi_model$r.squared,3)`, P \< 0.05*).
+We saw a significant difference in the *B.
+inermis* post-July 15 NDVI (*`r round(ndvi_model$r.squared,3)`, P<0.05*) when compared to the post-July 15 NDVI of the adjacent wheat crop.
 The *B.
-inermis* NDVI remained relatively linear in it's downslope (BROME SLOPE POST JULY) compared to the wheat field (WHEAT FIELD POST JULY).
-([Fig.
-3](#figure3))
-
-Need to add map showing where NDVI measurements came from
-
-### Controlled *C. cinctus* Infestation of *B. inermis*
-
-```{r column.order}
-column.order <- c("distance", "collection", "p.inf", 
-                  "p.cut", "p.par", "poten")
-
-column.order2 <- c("location", "distance", "collection", "p.inf", 
-                  "p.cut", "p.par", "poten")
-
-round_based_on_size <- function(x) {
-  threshold <- 0.01
-  rounded_values <- ifelse(abs(x) > threshold, round(x, 2), round(x, 3))
-  return(rounded_values)
-}
-```
-
-```{r (pf.RAW) post farm n.inf data}
-pf.raw.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2022/Post\ Farm\ 2022.xlsx", sheet = "n.inf")
-pf.raw.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "n.inf")
-```
-
-```{r (pf.inf.23) post farm inf data}
-pf.inf.23 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf") %>%
-  rename(n.1 = 8, n.2 = 9, n.3 = 10, n.4 = 11,
-         n.5 = 12, n.6 = 13, n.7 = 14, n.8 = 15,
-         n.9 = 16, n.10 = 17)
-```
-
-```{r (pf.comb) post farm data combined}
-# 2022
-pf.2022.site <-
-pf.raw.2022 %>%
-  group_by(treatment, site) %>%
-  reframe(site, treatment,
-            p.inf = sum(iv, ir, ix)/sum(uv, ur, iv, ir, ix),
-            cut = sum(cut)/sum(total),
-            total.stems = sum(iv, ir, ix, uv, ur)) %>%
-  unique() %>%
-  pivot_longer(-c(treatment, total.stems, site),
-               names_to = "type",
-               values_to = "percent") %>%
-  mutate(year = "2022")
-
-# 2023
-pf.2023.site <-
-pf.raw.2023 %>%
-  group_by(treatment, site) %>%
-  reframe(treatment, site, 
-          t.inf = sum(n.iv, n.ir, n.ix),
-          p.inf = sum(n.iv, n.ir, n.ix)/sum(n.uv, n.ur, 
-                                            n.iv, n.ir, n.ix),
-          cut = sum(cut)/sum(total),
-          total.stems = sum(n.iv, n.ir, n.ix, n.uv, n.ur)) %>%
-  distinct() %>%
-  pivot_longer(-c(treatment, total.stems, site),
-               names_to = "type",
-               values_to = "percent") %>%
-  filter(!type == "t.inf") %>%
-  arrange(type) %>%
-  mutate(year = "2023")
-
-pf.comb <-
-  pf.2022.site %>%
-  rbind(pf.2023.site) %>%
-  group_by(treatment, year, type) %>%
-  mutate(count = n(),
-         average = mean(percent),
-         sd = sd(percent),
-         se = (sd/sqrt(count))) %>%
-  reframe(treatment, year, type,
-          mean = mean(percent),
-          sd = mean(sd),
-          se = mean(se)) %>%
-  unique()
-
-```
-
-```{r (pf.total) pf total stems}
-# 2022 totals
-pf.2022.total <-
-  pf.raw.2022 %>%
-  group_by(treatment) %>%
-  reframe(total = sum(uv, ur, iv, ir, ix)) %>%
-  mutate(year = "2022")
-
-# 2023 totals
-pf.2023.total <-
-  pf.raw.2023 %>%
-  group_by(treatment) %>%
-  reframe(total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
-  mutate(year = "2023")
-
-pf.total.comb <-
-  pf.2022.total %>%
-  rbind(pf.2023.total)
-
-```
-
-```{r pf_lm}
-raw.23 <-
-pf.raw.2023 %>%
-  dplyr::select(site, treatment,
-                n.iv, n.ir, n.ix, n.ur, n.uv,
-                cut) %>%
-  rename(iv = n.iv,
-         ir = n.ir,
-         ur = n.ur,
-         ix = n.ix,
-         uv = n.uv) %>%
-  mutate(year = 2023)
-
-raw.22 <-
-pf.raw.2022 %>%
-  dplyr::select(site, treatment, 
-                ix, ir, iv, ur, uv,
-                cut) %>%
-  mutate(year = 2022)
-
-pf_inf_lm <-
-raw.23 %>%
-  rbind(raw.22) %>%
-  group_by(site, treatment, year) %>%
-  reframe(total = sum(iv, ir, ix, ur, uv),
-          inf = sum(iv, ir, ix)) %>%
-  lm(inf ~ treatment + year, data = .) %>%
-  summary()
-
-pf_cut_lm <-
-raw.23 %>%
-  rbind(raw.22) %>%
-  group_by(site, treatment, year) %>%
-  reframe(total = sum(iv, ir, ix, ur, uv),
-          inf = sum(iv, ir, ix),
-          cut = sum(cut)) %>%
-  lm(cut ~ treatment + year, data = .) %>%
-  summary()
-```
-
-```{r pf_per, eval = F}
-raw.23 %>%
-  rbind(raw.22) %>%
-  group_by(treatment) %>%
-  reframe(total = sum(iv, ir, ix, ur, uv),
-          inf = sum(iv, ir, ix),
-          cut = sum(cut),
-          p.inf = inf/total,
-          p.cut = cut/inf)
-```
-
-Initially, *C.
-cinctus* heavily infested stems of *B.
-inermis* in the controlled test conditions.
-Averaged across both years, we observed 66.5% of stems infested for high treatments and 47.3% of stems for low treatments.
-We found strong evidence suggesting that there was a significant difference between infestation at high and low treatment levels when holding year constant (*r = `r round(pf_inf_lm$r.squared,3)`, P \< 0.05*, [Fig.
-3](#figure4)).
-
-```{r fig4_inf_cut_plot code}
-pf.comb.hl <-
-pf.comb %>%
-  filter(!treatment == "control")
-
-inf_cut_plot <-
-ggplot() +
-  facet_wrap(~treatment) +
-  geom_bar(stat = "identity",
-           position = position_dodge(),
-           data = pf.comb.hl %>%
-             filter(type == "p.inf"),
-           aes(year, mean, fill = "p.inf")) +
-  geom_bar(stat = "identity",
-           position = position_dodge(),
-           data = pf.comb.hl %>%
-             filter(type == "cut"),
-           aes(year, mean, fill = "cut")) +
-  geom_errorbar(data = pf.comb.hl,
-                aes(x = year,
-                    ymin = mean - se,
-                    ymax = mean + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  theme_bw() +
-  labs(x = element_blank(),
-       y = "Percentage of stems (%)",
-       fill = element_blank()) +
-  scale_fill_brewer(palette = "Greys",
-                    labels = c("Cut", "Infested"))
-```
-
-Stem cutting by mature larvae, leading into winter, was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P < 0.05*).
-There was a dramatic decrease in survival observed from initial infestation to stem cutting.
-
-```{r figure_5_pf_inf_nodes}
-pf.y.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2022.xlsx", sheet = "inf")
-pf.y.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf")
-
-# DATA FOR THE PLOT
-pf.nodes.plot.data <-
-pf.y.2023 %>%
-  dplyr::select(-c(Name, 'Date Proc')) %>%
-  dplyr::select(treatment, 6:15) %>%
-  pivot_longer(-c(treatment),
-               names_to = "node",
-               values_to = "value") %>%
-  mutate(node = gsub("n_", "", node)) %>%
-  na.omit() %>%
-  mutate(year = "2023") %>%
-  rbind(pf.y.2022 %>%
-          dplyr::select(-c(site1, site2, site12, site3,
-                           Name, 'Date Proc')) %>%
-          dplyr::select(treatment, 
-                        2, 3, 4, 5, 6, 
-                        7, 8, 9, 10, 11) %>%
-          pivot_longer(-c(treatment),
-                       names_to = "node",
-                       values_to = "value") %>%
-          na.omit() %>%
-          mutate(year = "2022")) %>%
-  group_by(year, treatment, node) %>%
-  reframe(count = n()) %>%
-  group_by(year, treatment) %>%
-  mutate(per = count/sum(count)) %>%
-  mutate(node = as.numeric(node))
-
-# PLOT
-pf_nodes_plot <-
-  pf.nodes.plot.data %>%
-  ggplot() +
-  geom_bar(stat = 'identity',
-           aes(x = node, y = per, fill = treatment),
-           position = position_dodge(),
-           color = "black") +
-  theme_bw() +
-  scale_fill_brewer(palette = "Greys") +
-  scale_x_continuous(breaks = 1:10) +
-  labs(x = "Stem Internode",
-       y = "Percent of internodes infested") +
-  facet_wrap(~year)
-
-pf_nodes_over_4_table <-
-pf.nodes.plot.data %>%
-  group_by(year, node) %>%
-  reframe(count = sum(count)) %>%
-  group_by(year) %>%
-  mutate(per = count/sum(count)) %>%
-  filter(node > 4) %>%
-  group_by(year) %>%
-  reframe(sum = sum(per))
-```
-
-*C.
-cinctus* infested stems often contained larvae burrowing through multiple nodes.
-In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through, or 5 internodes occupied.
-In 2023, `r round(pf_nodes_over_4_table[2,2],3)*100`% of stems had at least 4 nodes infested [Fig.
-5](#figure5)
-
-### Montana Field Sites
-
+inermis* NDVI remained relatively linear in it's downslope compared to the wheat field ([Fig.
+7](#figure7)).
+
+#### Field Site Samples
 ```{r stubble data}
@@
 Cutting by *C.
-cintus* was observed at an average rate of 5.59% (se = 0.02) in *B.
-inermis* and 47.1% (se = 0.04) in the adjacent cultivated wheat ([Fig.
-7](#figure7).
+cintus* was observed at an average rate of 5.59% in *B.
+inermis* and 47.1% in the adjacent cultivated wheat ([Fig.
+8](#figure8).
 
@@
 In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of burrowing injury at our second sampling date.
-In Moccasin, the corresponding figure was 66.5%.
+In Moccasin, the corresponding figure was 66.5% ([Fig.
+9](#figure9)).
 
@@
 
+```{r per.par}
+comb.stubble %>%
+  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
+  group_by(region) %>%
+  reframe(total.stems = sum(total.stems),
+          inf = sum(inf),
+          cut = sum(cut),
+          par = sum(c, e),
+          inf.per = inf/total.stems,
+          cut.per = cut/inf,
+          par.per = par/inf) %>%
+  dplyr::select(par.per)
+
+comb.stubble %>%
+  dplyr::select(-c(uninf, cut, c, e, total.stems))
+```
+
 ```{r infestation_per, eval = F}
@@
   arrange(desc(inf))
+
+plot.data %>%
+  filter(collection == "fall") %>%
+  group_by(region, collection, plant_type) %>%
+  reframe(per.par = sum(par)/sum(total.stems))
+
+comb.brome %>%
+  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
+  na.omit() %>%
+  group_by(region, collection) %>%
+  reframe(inf = sum(inf),
+          cut = sum(cut),
+          cut.per = cut/inf)
 ```
@@
 All other sample locations also showed positive ratios of parasitoids to cut stems [(Fig.
-8)](#figure8).
+10)](#figure10).
 
@@
 
-```{r montana map}
+## Discussion
+<!-- CONTROLLED INFESTATION -->
+Assessment of *C.
+cicntus* infestation and larval mortality within *B.
+inermis* from out controlled infestation survey showed high infestation coupled with high larval mortality prior to overwintering diapause.
+Even in cases of high infestation (70%+), larval survival to cutting topped out around around 15%, pointing towards *B.
+inermis'* ability to serve as a quality sink for *C.
+cinctus* larvae.
+These results confirm the assertions by Criddle (1992) and Seamans (1928) that *B.
+inermis* can serve as a quality *C.
+cinctus* trap crop.
+In addition to high larval mortality, stem dissection exhibited high numbers of stems with over four internodes burrowed through by *C.
+cinctus* larvae.
+Larval movement through stems can be quick in the case of susceptible cultivated cereal grains [@Criddle1923].
+However, *B.
+inermis*, given it's presumed antibiosis properties, leads to slower larval development and thus movement through the stem.
+We found several *B.inermis* stems with living larvae as late as late-September.
+
+<!-- WEATHER DATA -->
+As the climate of central and northern Montana continues to change, wheat stem sawfly infestation will continue to impact cereal crop yields.
+Precipitation and temperature data from our sampling sites underscores the importance of continuing to monitor and conserve areas of semi-natural grassland.
+50-year trends analyzed in this study show similar results to Pederson et al (2009) in which the dry-land agricultural environments found in central and northern Montana are likely to continue to become hotter and dryer on average.
+Over the past fifty years, precipitation in Big Sandy has decreased by an average of 0.05 inches per year.
+Given the trend continues, by 2050 the area could experience an inch less of rain per year.
+
+Both average precipitation and average temperatures vary widely between years as the region goes through many droughts and periods of relative high rainfall [@WeatherUnderground; @Pederson2009].
+Shown in our analysis of daily measurements over the past five years, precipitation significantly decreases during the period after wheat harvest, July 15 - September15, a period when second generation *Bracon* adults are searching for *C.
+cinctus* larval hosts.
+Mid-season senesence of many *C.
+cinctus* host plants can then leave *Bracon* parasitoids with little resources for
+
+Prolonged drought, like seen in 2021 in Big Sandy, MT, can cause a dearth of green host plants in these semi-arid environments that can be problimatic for second generation adult parasitoids.
+
+Given our analysis of NDVI at our sampling sites, it is clear that *B.
+inermis* is capable of continued late seasonal greening and may provide important resources for *Bracon* spp parasitoids in years of drought or crop failure.
+It is these buffer zones where parasitoids may be able to find *C.
+cinctus* larvae in numbers great enough to prevent the larger population crashes that can lead large localized increases in *C.
+cinctus* populations and damage [@Holmes1982].
+
+Analysis of larval *C.
+cinctus* infestation of *B.
+inermis* and adjacent winter wheat reveal similar results to our controlled *B.
+inermis* infestation in Bozeman.
+We found *B.
+inermis* infestation to be similar to that of neighboring wheat fields, making it likely that *C.
+cinctus* pressures *B.
+inermis* similarly to cultivated wheat.
+
+<div class="box">
+Still working on adding some citations and information discussing Rand & Peirce papers and how my findings are similar and different.
+I collected samples twice throughout the year so got a bigger picture of how parasitism changes for each generation.
+It is looking like parasitism in the fall collections is much higher in the brome compared to the summer collections.
+Interesting!
+</div>
+
+### Montana Field Site Survey
+
+Analysis of weather trends in Big Sandy and Moccasin, MT confirmed that the  patterns in central and northern Montana are changing towards warmer and drier conditions [@Pederson2009].
+We observed an average decrease in 0.058 inches of precipitation per year since 1975, while since 1950, temperatures have experienced an average yearly increase of 0.028°C.
+These changes in precipitation and temperature can have significant impacts on timing of crop production and harvest timing [@Zhu2021].
+Year-over-year populations of the *Bracon* spp parasitoids are particularly sensitive, as their bivoltine nature requires the second yearly generation to find a suitable *C.
+cinctus* host [@Nelson1953; @Runyon2002].
+NDVI analyses exhibited the prolonged senescence period for *B.
+inermis* when compared to adjacent cultivated wheat.
+*C.
+cinctus* larvae are cued via the thinning of the cell membranes [@Villacorta1971] during plant senescence to travel to the base of the stem and prepare for obligatory diapause [@Beres2011].
+Because *B.
+inermis* senescence occurs at slower rates compared to wheat or barley, *C.
+cinctus* may remain within the stem later in into the summer and fall, providing potential host larvae for second generation *Bracon* parasitoids.
+
+Previous research has shown *B.
+inermis* to be a quality host of both *C.
+cinctus* and *Bracon* spp parasitoids [@Criddle1922; @Peirce2021; @Rand2024], but little data is available on the season fluctuation of *B.
+inermis* utilization [@Criddle1922; @Peirce2021; @Rand2024].
+Because we collected stem samples twice during the growing season, we were able to compare levels of parasitism.
+We hoped to understand how parasitism rates changed from July (pre-wheat harvest) and September (post-wheat harvest) of each year.
+
+Our results showed that parasitoid presence increase significantly at all sites between pre- and post-harvest collections.
+Given the collection methods, September collection parasitism was inherently a combination of both pre-harvest and post-harvest activity.
+While we cannot ascertain *Bracon* preference for host stems, we can assess that the parasitoids utilize infested *B.
+inermis* stems post-harvest when *C.
+cinctus* is no longer available within wheat stems.
+
+Additionally, we observed a high number of stems exhibiting levels of *C.
+cinctus* damage that point to prolonged larval presence within *B.
+inermis* stems.
+In Big Sandy (76.2%) and Moccasin (66.5%), stems that exhibited four or more nodes burrowed suggests that *C.
+cinctus* larvae remain available within *B.
+inermis* stems past wheat harvest, when they are more limited in availability.
+Most importantly, in years of drought or increased temperatures where wheat senescence and harvest may occur earlier in the summer.
+
+We utilized parasitism to cut stem ratios to understand the relative impact of *B.
+inermis* host sources on overall field parasitism.
+Assessing overall numbers of insects can potentially diminish the apparent importance of *B.
+inermis*, as per square meter stem counts in cultivated wheat are denser than *B.
+inermis* counts within coulees and along roadsides *B.
+inermis*.
+Our findings of significantly greater ratios of parasitized stems to cut stems within *B.
+inermis* when compared to adjacent when indicates these field-side areas provide greater parasitoid source per square meter when compared to cultivated fields.
+This emphasizes the significance of accounting for habitat diversity to sustain parasitoid populations over successive years, as well as recognizing the crucial role of edge-row areas in maximizing parasitoid effectiveness in minimizing *C.
+cinctus* damage.
+
+### Old Discussion Text
+*C.
+cinctus* mortality was found to be high in controlled *B.
+inermis* infestation trials.
+Our results build upon other studies where *B.
+inermis* infestation confirmed previously reported high *C.
+cinctus* larval mortality prior to overwintering [@Shanower2004; @Mendoza2006; @Buteler2008].
+Of those infested stems, a large proportion exhibited larval damage in at least 5 internodes (62.5%).
+While the timing of stem collection and dissection did not align with precise measurement of living larvae (stems dissected in October), we can assert that stems with at least 4 nodes burrowed through would have a high probability of containing large, live larvae, and thus a viable potential parasitoid host, in late July and August of the same year.
+
+While total and per node *B inermis* infestation by *C.
+cinctus* was high, survival to overwintering (cutting) only occurred in 5.7% of *C.
+cinctus* infested stems.
+Our results indicate that *B.
+inermis* serves as a high-quality host for *C.
+cinctus* larvae during the spring and summer months.
+However, there was minimal survival of *C.
+cinctus* larvae beyond August and September, confirming that there will be little adult emergence from *B.
+inermis* the following season.
+
+Assessing the comparative abundance of *C.
+cinctus* and its associated parasitoids within *B.
+inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas serve as sink or source origins of the pest and its beneficial natural adversaries infiltrating agricultural crops.
+
+<!-- Overall, this work suggests that maintaining and supporting field-edge areas containing *B.
+inermis* provides an unmistakable benefit to *C.
+cinctus* control in cereal crop production.
+The characteristics of *B.
+inermis* make it uniquely beneficial in both as a sink for *C.
+cinctus* larvae, and as a source for *Bracon* adults.
+*B.
+inermis* is a common grass found throughout the golden triangle of Montana and other wheat cultivation areas [@Bolwahn2011].
+-->
+
+## Figures
+
+### Figure 1 {#figure1}
+```{r figure1, fig.cap="Figure 1: Location of Montana, USA field sites.
+Controlled *B.
+inermis* infestation site was in Bozeman, while field sites were located in centrally location Moccasin, MT and northern Big Sandy, MT."}
 world_map <- map_data("world")
@@
 
-montana_map <-
 ggplot() +
@@
   geom_polygon(data = counties, aes(x = long, y = lat, group = group),
-               fill = NA, color = "black", linewidth = 0.2) +
+               fill = NA, color = "black", linewidth = 0.1) +
   geom_polygon(data = states, aes(x = long, y = lat, group = group),
-               fill = NA, color = "black", linewidth = 0.9) +
+               fill = NA, color = "black", linewidth = 0.5) +
   geom_label(data = coordinates, aes(x = longs, y = lats), 
@@
   #       panel.grid.minor = element_blank()) +
-  theme(aspect.ratio = 1/1.8)
-
-```
-
-## Discussion
-
-Assessing the comparative abundance of *C.
-cinctus* and its associated parasitoids within *B.
-inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas serve as sink or source origins of the pest and its beneficial natural adversaries infiltrating agricultural crops.
-As the climate of central and northern Montana continues to change, wheat stem sawfly infestation will continue to impact cereal crop yields.
-In this study, we monitored how *B.
-inermis* may be potentially beneficial in ecologically buffering *Bracon* spp parasitoids by observing *C.
-cinctus* mortality within *B.
-inermis* and within adjacent cereal crops as well as *Bracon* spp prevalence.
-
-### Controlled Infestation
-
-Results from the controlled *B.
-inermis* infestation confirmed large amounts of *C.
-cinctus* larval mortality prior to overwintering [@Shanower2004; @Mendoza2006; @Buteler2008] with 53% of total collected stems exhibiting *C.
-cinctus* burrowing damage.
-Of those infested stems, a large proportion exhibited larval damage in at least 5 internodes (62.5%).
-While the timing of stem collection and dissection did not align with precise measurement of living larvae (stems dissected in October), we can assert that stems with at least 4 nodes burrowed through would have a high probability of containing large, live larvae, and thus a viable potential parasitoid host, in late July and August of the same year.
-
-While total and per node *B inermis* infestation by *C.
-cinctus* was high, survival to overwintering (cutting) only occurred in 5.7% of *C.
-cinctus* infested stems.
-Our results indicate that *B.
-inermis* serves as a high-quality host for *C.
-cinctus* larvae during the spring and summer months.
-However, there was minimal survival of *C.
-cinctus* larvae beyond August and September, confirming that there will be little adult emergence from *B.
-inermis* the following season.
-
-### Montana Field Site Survey
-
-Analysis of weather trends in Big Sandy and Moccasin, MT confirmed that the  patterns in central and northern Montana are changing towards warmer and drier conditions [@Pederson2009].
-We observed an average decrease in 0.058 inches of precipitation per year since 1975, while since 1950, temperatures have experienced an average yearly increase of 0.028°C.
-These changes in precipitation and temperature can have significant impacts on timing of crop production and harvest timing [@Zhu2021].
-Year-over-year populations of the *Bracon* spp parasitoids are particularly sensitive, as their bivoltine nature requires the second yearly generation to find a suitable *C.
-cinctus* host [@Nelson1953; @Runyon2002].
-NDVI analyses exhibited the prolonged senescence period for *B.
-inermis* when compared to adjacent cultivated wheat.
-*C.
-cinctus* larvae are cued via the thinning of the cell membranes [@Villacorta1971] during plant senescence to travel to the base of the stem and prepare for obligatory diapause [@Beres2011].
-Because *B.
-inermis* senescence occurs at slower rates compared to wheat or barley, *C.
-cinctus* may remain within the stem later in into the summer and fall, providing potential host larvae for second generation *Bracon* parasitoids.
-
-Previous research has shown *B.
-inermis* to be a quality host of both *C.
-cinctus* and *Bracon* spp parasitoids [@Criddle1922; @Peirce2021; @Rand2024], but little data is available on the season fluctuation of *B.
-inermis* utilization [@Criddle1922; @Peirce2021; @Rand2024].
-Because we collected stem samples twice during the growing season, we were able to compare levels of parasitism.
-We hoped to understand how parasitism rates changed from July (pre-wheat harvest) and September (post-wheat harvest) of each year.
-
-Our results showed that parasitoid presence increase significantly at all sites between pre- and post-harvest collections.
-Given the collection methods, September collection parasitism was inherently a combination of both pre-harvest and post-harvest activity.
-While we cannot ascertain *Bracon* preference for host stems, we can assess that the parasitoids utilize infested *B.
-inermis* stems post-harvest when *C.
-cinctus* is no longer available within wheat stems.
-
-Additionally, we observed a high number of stems exhibiting levels of *C.
-cinctus* damage that point to prolonged larval presence within *B.
-inermis* stems.
-In Big Sandy (76.2%) and Moccasin (66.5%), stems that exhibited four or more nodes burrowed suggests that *C.
-cinctus* larvae remain available within *B.
-inermis* stems past wheat harvest, when they are more limited in availability.
-Most importantly, in years of drought or increased temperatures where wheat senescence and harvest may occur earlier in the summer.
-
-We utilized parasitism to cut stem ratios to understand the relative impact of *B.
-inermis* host sources on overall field parasitism.
-Assessing overall numbers of insects can potentially diminish the apparent importance of *B.
-inermis*, as per square meter stem counts in cultivated wheat are denser than *B.
-inermis* counts within coulees and along roadsides *B.
-inermis*.
-Our findings of significantly greater ratios of parasitized stems to cut stems within *B.
-inermis* when compared to adjacent when indicates these field-side areas provide greater parasitoid source per square meter when compared to cultivated fields.
-This emphasizes the significance of accounting for habitat diversity to sustain parasitoid populations over successive years, as well as recognizing the crucial role of edge-row areas in maximizing parasitoid effectiveness in minimizing *C.
-cinctus* damage.
-
-<!-- Overall, this work suggests that maintaining and supporting field-edge areas containing *B.
-inermis* provides an unmistakable benefit to *C.
-cinctus* control in cereal crop production.
-The characteristics of *B.
-inermis* make it uniquely beneficial in both as a sink for *C.
-cinctus* larvae, and as a source for *Bracon* adults.
-*B.
-inermis* is a common grass found throughout the golden triangle of Montana and other wheat cultivation areas [@Bolwahn2011].
--->
-
-## Figures
-
-### Figure 1 {#figure1}
-```{r FIG1_both_prcp_plot, fig.cap = "Figure 1: Average yearly precipitation (in) for Big Sandy and Moccasin, Montana.
-Black trend line signifies averaged negative trend between both locations.
-Data gathered from NOAA and MSU-ARS.
-A significant linear relationship (r = 0.1, P = 0.033, estimate = -0.058) was observed between year and average yearly precipitation."}
-both_prcp_plot 
+  theme(aspect.ratio = 1/1.8) +
+  labs(title = "Montana, USA")
 ```
@@
 ### Figure 2 {#figure2}
-```{r both_temp_plot, fig.cap = "Figure 2: Average yearly temperature (°C) in Moccasin and Big Sandy, MT.
-Positive trendline slop suggests increasing average temperatures of the past 100 years.
-A significant linear relationship (*r =0.2447, P < 0.05, estimate = 0.028*) was observed between year and average yearly temperature."}
-both_temp_plot
+```{r bigsandy_ndvi, fig.cap="Figure 2: June 29, 2023 - Big Sandy, MT.
+Orange polygons denote location where NDVI was pulled using Google Earth Engine from both (upper left) Vida spring wheat and (central) coulee containing semi-natural grasses including majority *B.
+inermis*."}
+knitr::include_graphics("Images/BigSandy_NDVI_pic.jpg")
 ```
@@
 ### Figure 3 {#figure3}
-```{r bs_ndvi_plot, fig.cap = "Figure 3: Normalized difference vegetation index (NDVI) of *B.
-inermis* and adjacent spring wheat field from April 2023 to October 2023.
-Post harvest linear model indicates a significant difference (P values and stuff) when comparing the *B.
-inermis* post-harvest slope and the spring wheat post-harvest slope."}
-bs_ndvi_plot
-```
-
-![](Images/brome_images.jpg)
-
-### Figure 4 {#figure4}
-```{r inf_cut_plot, fig.cap = "Figure 4: Comparison of year and treatment group for controlled infestation of *B.
+```{r inf_cut_plot, fig.cap = "Figure 3: Comparison of year and treatment group for controlled infestation of *B.
 inermis*.
@@
 
-### Figure 5 {#figure5}
-```{r figure5_pf_nodes_plot, fig.cap="Figure 6: Ineternodes infested by *B.
+### Figure 4 {#figure4}
+```{r figure5_pf_nodes_plot, fig.cap="Figure 4: Ineternodes infested by *B.
 cephi* within controlled infestation plots in Bozeman, MT, USA.
@@
 
+### Figure 5 {#figure5}
+```{r FIG1_both_prcp_plot, fig.cap = "Figure 5: Average yearly precipitation (in) for Big Sandy and Moccasin, Montana.
+Data gathered from NOAA.
+Black trend line signifies averaged negative trend between both locations.
+Data gathered from NOAA and MSU-ARS.
+A significant linear relationship (r = 0.1, P = 0.033, estimate = -0.058) was observed between year and average yearly precipitation."}
+both_prcp_plot 
+```
 ### Figure 6 {#figure6}
-```{r figure6_brome_nodes, fig.cap="Figure 6: Internodes infested by *B. cephi* within collected *B. inermis* stems in Moccasin and Big Sandy, MT, USA."}
-brome.nodes
+```{r both_temp_plot, fig.cap = "Figure 6: Average yearly temperature (°C) in Moccasin and Big Sandy, MT.
+Data gathered from NOAA.
+Positive trendline slop suggests increasing average temperatures of the past 100 years.
+A significant linear relationship (*r =0.2447, P < 0.05, estimate = 0.028*) was observed between year and average yearly temperature."}
+both_temp_plot
 ```
@@
 ### Figure 7 {#figure7}
-```{r figure7_field_cut_plot, fig.cap="Figure 7: Stem cutting by *C.
+```{r bs_ndvi_plot, fig.cap = "Figure 7: Normalized difference vegetation index (NDVI) of *B.
+inermis* and adjacent spring wheat field from April 2023 to October 2023 in Big Sandy, MT.
+Post harvest linear model indicates a significant difference (P values and stuff) when comparing the *B.
+inermis* post-harvest slope and the spring wheat post-harvest slope."}
+bs_ndvi_plot
+```
+
+```{r moc_seasonal_pot_show, fig.cap="Figure 5: Average precipitation in Big Sandy, MT. Average daily precipitation"}
+moc_seasonal_plot
+```
+
+### Figure 8 {#figure8}
+```{r figure7_field_cut_plot, fig.cap="Figure 8: Stem cutting by *C.
 cinctus*.
@@
 ```
-
-### Figure 8 {#figure8}
-```{r figure8, fig.cap="Ratios of parasitized to cut stems in *B. inermis* and adjacent cultivated wheat. Dotted line represents where a 1:1 cut:parasitized ratio lies."}
+### Figure 9 {#figure9}
+```{r figure6_brome_nodes, fig.cap="Figure 9: Percentage of infested stems with internodes infested by *B. cephi* within collected *B. inermis* stems in Moccasin and Big Sandy, MT, USA."}
+brome.nodes
+```
+
+### Figure 10 {#figure10}
+```{r figure10, fig.cap="Ratios of parasitized:cut stems in *B. inermis* and adjacent cultivated wheat. Dotted line represents where a 1:1 cut:parasitized ratio lies."}
 ratio_plot
@@
 brome_par_plot
-```
-
-### Figure 10 {#figurex}
-```{r figure10, fig.cap="Location of Montana, USA field sites.
-Controlled *B.
-inermis* infestation site was in Bozeman, while field sites were located in centrally location Moccasin, MT and northern Big Sandy, MT."}
-montana_map +
-  labs(title = "Montana, USA")
-```
-
+
+```
+
+![](Images/brome_images.jpg)
 ## Citations

### 2024-06-13 qmd/projects/msu/wss/paper_1.qmd
update

@@
 ---
-title: "Ecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids"
+title: "Agroecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids"
 authors: "Jackson R. Strand, David K. Weaver"
-date: "May 17, 2024"
+date: "`r Sys.Date()`"
 editor_options: 
@@
 link-citations: true
-biblio-style: apalike
-cls: 
-  max-names: 1
+csl: entomological-society-of-america-2.csl
 ---
@@
 
+Wheat stem sawfly, *Cephus cinctus* Norton (WSS), causes significant damage in cereal crops throughout the Northern Great Plains of North America.
+*Bracon cephi* and *B.
+lissogaster* are native WSS parasitoids important in managing WSS outbreaks and damage.
+Smooth Bromegrass (*Bromus inermis*) grass has been found to be an effective parasitoid sink, and potential trap crop, when grown in areas surrounding wheat fields in Montana.
+As climate changes continue to alter central and northern Montana, WSS will continue to cause immense damage to dry-land cereal crop yields.
+We analyzed both local temperature and precipitation trends at our sample sites, and found statistically significant increases in temperature and decreases in precipitation over the past fifty years.
+Our study observes the utilization of smooth brome in providing year-over-year host refuge for Bracon spp parasitoids.
+We measured the WSS larval infestation and survival rate within controlled smooth brome plots, observing a maximum infestation of 66.5% and a maximum end-of-year survival of 5.7%.
+We also collected stems from sites in central and northern Montana where we measured the WSS infestation and parasitoid prevalence within cultivated wheat and adjacent smooth brome.
+Montana fields sites showed high levels of *C.
+cinctus* infestation within both *B.
+inermis* and adjacent wheat crop in Big Sandy (64.5% *B.
+inermis*, 65.7% adjacent wheat) and Moccasin (50.6%, 38.6%).
+We observed high year-end WSS larval mortality measured as stem cutting of infested stems with *B.
+inermis* when compared to adjacent wheat in Big Sandy (2.9% *B.
+inermis*, 46.5% adjacent wheat) and Moccasin (6.4%, 49.4%).
+Infested stems exhibited parasitism in both Big Sandy (7.11% *B.
+inermis*, 8.6% adjacent wheat) and Moccasin (6.11%, 2.09%).
+This research underlines the importance of providing a sustainable ecological buffer for WSS parasitoids as climate conditions continue to change.
+
 ## Introduction
@@
 
-The wheat stem sawfly (WSS), Hymenoptera *Cephus cinctus* Norton, is a grass-mining sawfly native to North America [@Lesieur2016].
-Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as the practice became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 [@Morrill1996].
-Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [@Holmes1956].
-Economic damage manifests in the form of both kernel weight reduction caused by tissue damage [@Beres2011] and stem toppling due to late-season larval girdling [@Holmes1960].
-
-Several parsitoid species attack *C.
-cinctus* within wild grass hosts [@Davis1955; @Morrill1998], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill1998].
+The wheat stem sawfly (WSS), Hymenoptera *Cephus cinctus* Norton, is a grass-mining sawfly native to North America [@Criddle1922; Ainslie1929; @Wallace1966; @Lesieur2016].
+Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as this resource became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 in Montana [@Morrill1996], and more recently in Colorado [@Cockrell2021].
+Climate warming has the potential to further expand the range of this pest both northward and southward [@Olfert2019; @Weaver2023].
+Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [Roemhild1954; @Holmes1956].
+Economic damage manifests in the form of both kernel weight reduction caused by tissue damage [@Beres2011a; Delaney2010] and stem toppling due to late-season larval girdling [@Holmes1960; @Nansen2005a; @McCullough2020].
+
+Several parasitoid or predatory species attack *C.
+cinctus* within wild grass hosts [@Davis1955; @Morrill1998, @Morrill2001], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill1998; @Runyon2001a; @Runyon2001b].
 These two host-specific ectoparasitoids are native to North America.
-Both parasitoids are bivoltine, with the first generation present in Montana from late June to late July, while the second generation is active in August and September [@Somsen1956].
+Both parasitoids are bivoltine, with the first generation present in Montana from late-June to late-July, while the second generation is active from mid-July to September [@Gahan1918; @Somsen1956; @Davis2013].
 Adult female *B.
 cephi* and *B.
-lissogaster* paralyze *C.
-cinctus* larvae within the host stem, paralyze the host larvae, and depositing 1 (*B.
-cephi*) to three (*B.
+lissogaster* inject a paralytic toxin into a *C.
+cinctus* larvae within the host stem, paralyzing the host larvae, and depositing from a single (*B.
+cephi*) to 1-3 (*B.
 lissogaster*) eggs.
@@
 lissogaster* have been shown to provide biological control via irreplaceable mortality of *C.
-cinctus* populations [@Buteler2015] and can significantly reduce *C.
-cinctus* damage and impact on crop yields [@Buteler2008; @Rand2020].
+cinctus* populations [@Buteler2015; @Peterson2011] and can significantly reduce *C.
+cinctus* damage and impact on crop yields [@Buteler2008; @Adhikari2019; @Bekkerman2018; @Rand2020].
 While *C.
-cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*)[@Wallace1966; @Cockrell2017], infestation levels can vary widely across hosts.
-Previous research has shown preference for host plants with larger stem diameter [@Ainslie1920; @Bhandari2020].
-
-Previous research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *B.
-inermis*, *Elymus trachycaulus*, and *Elytrigia intermedium* host *C.
-cinctus* and Bracon spp.
-parasitoids quite well [@Peirce2021; @Rand2024].
+cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*) [@Wallace1966; @Cockrell2017; @Achhami2020], infestation levels can vary widely across hosts.
+Previous research has shown preference for host plants with larger stem diameter [@Ainslie1920; @Buteler2009; @Bhandari2020].
+
+Recent research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *Bromus techtorum*, *Bromus inermis*, *Elymus trachycaulus*, and *Thinopyrum intermedium* host *C.
+cinctus* and most also host Bracon spp.
+parasitoids quite well [@Mendoza2006; @Peirce2021; @Rand2024].
 We find investigating *B.
-inermis* holds the most potential, as it has been shown to be a quality host for *C.
-cinctus* while also leading to high larval mortality [@Farstad1945].
-However, there is a lack of refined data relating to the total infestation, cutting, and parasitization within *B.
+inermis* holds the most potential, as it has been shown to be a quality initially-accepted host for *C.
+cinctus* oviposition while also leading to high larval mortality later on [@Seamans1928; @Farstad1945].
+However, there is a lack of detailed relating to the total infestation, cutting, and parasitism within *B.
 inermis* at multiple time points throughout the growing season.
-Quantificative assessment of potential 'reservoire' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B.
-inermis* along roadways and within coulles.
-
-Smooth brome (*Bromus inermis* Lyess) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Salesman2011].*B.
-inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Farstad1945].
+Quantitative assessment of potential 'reservoir' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B.
+inermis* along roadways and within coulees.
+
+Smooth brome (*Bromus inermis* Leyss) is a Eurasian cool-season rhizomatous grass introduced to the United States in 1884 [@Bolwahn2011].*B.
+inermis* was originally thought of as potentially useful as a trap crop in order to combat the growing WSS problem with cultural control practices [@Criddle1922; @Seamans1928; @Farstad1945; @Beres2011a].
 Due to its low agricultural value, and propensity to form monocultures and out compete many native grasses, *B.
-inermis* is categorized as invasive [@Carlson1985; @Willson2000; @Dillemuth2008].
+inermis* is categorized as invasive in some habitats [@Carlson1985; @Willson2000; @Dillemuth2008].
 However, *B.
-inermis* still holds immense ecological value in its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].
-
-Montana's golden triangle, an area of central and northern Montana with intense cereal grain production, is prone to drought and burning via wildfire [@Wilson1923].
+inermis* still holds considerable ecological and potential economic value due to its late senescence and WSS antibiosis properties [@Otfinowski2006; @Rand2024].
+
+Montana's golden triangle, an area of central and northern Montana with intense cereal grain production, is prone to drought and burning via wildfire [@Wilson1923; @Adhikari2019].
 As both *C.
@@
 As with many predator-prey interactions, the *C.
-cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession.
-Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinelidae and aphids [@Evans1999].
-
-We examined historical weather data from our field sites in central and northern Montana.
-We used this data to understand how changes in global climate trends may be impacting *B.
-cephis* and *B.
+cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession and through ineteraction with drought [@Holmes1963; @Holmes1982].
+Similar examples of these classical population dynamics patterns can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinellidae and aphids [@Evans1999], among others.
+
+We examined historical weather data from two field locations (10 sites overall) in central and northern Montana.
+We used this data to understand how changes in global climate trends may be impacting *C.
+cinctus*, *B.
+cephi* and *B.
+lissogaster* interact with *B.
 inermis* populations.
 In addition we analyzed normalized difference vegetative index (NDVI) at our field sites to understand how *B.
-inermis* senescence patterns compare to adjacent cultivated wheat.
+inermis* senescence patterns compare to adjacent cultivated wheat fields.
 
 In this study, we explored the ecological patterns between Montana populations of *C.
-cinctus* and its braconid parsitoids in relation to *B.
+cinctus* and its associated braconid parasitoids in relation to *B.
 inermis* host sites.
@@
 cinctus* within the *B.
-inermis* stems.
-The second study examined rates of cutting and parasitism at field sites in central and northern Montana.
-
-Using these studies, we addressed three questions, 1) what is the total cutting, and thus source rate, of *C.
-cinctus* from *B.
-inermis*, 2) can *C.
+inermis* stems in the absence of parasitoids.
+The second study examined rates of cutting and parasitism in wheat crops and their surroundings at field sites in central and northern Montana.
+
+Using these studies, we addressed three questions, 1) what is the total *C.
+cinctus*-caused stem cutting heading into overwinter - and thus the source rate, of the subsequent *C.
+cinctus* population in *B.
+inermis*?
+2) can *C.
 cinctus* larvae within *B.
-inermis* serve as second generation hosts for *B.
+inermis* serve as critical second generation hosts for *B.
 cephi* and *B.
-lissogaster*, and 3) how do *C.
+lissogaster*?
+and 3) how do annual *C.
 cinctus* and parasitoid populations in *B.
-inermis* compare with those in adjacent cultivated wheat fields.
+inermis* compare with those in the adjacent cultivated wheat fields?
 
 ## Materials and Methods
-
-### Weather data and NDVI analysis
-
-*Weather data.* We assessed the long and medium term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA).
-Data for each field site was averaged from three of the closest weather stations to that location.
-Data was plotted using R Studio (R Studio version 2024.04.0+735, R 2024) package 'ggplot' (version 3.4.4) [@ggplot].
-Data was then fit using a linear model using the 'lm' command using average yearly precipitation (inches) as the response variable and year as the predictor.
-
-*NDVI analysis.* We compared the relative greening throughout the growing season between wheat fields and adjacent *B.
-inermis* using the normalized difference vegetation index (NDVI).
-NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infared light reflected by vegetation [@Pettorelli2005].
-Data was downloaded using Google Earth Engine (Google Inc.
-2023, Mountain View, CA, USA).
 
@@
 inermis* were assessed using a 34 x 60 ft plot at the Arthur H.
-Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an extension research station of Montana State University in Bozeman, MT.
-The cage structure was built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
+Post Agronomy Farm (43°38'19.39"N, 116°14'28.86"W), an experimental farm of Montana State University in Bozeman, MT.
+There is a negligible population of *C.
+cinctus* and associated parasitoids at this location.
+Experimental cages  were built using 1-inch PVC piping with the netting made using 530$\mu$ Amber Lumite Screen (BioQuip$^\circledR$ Products, LLC).
 Twelve cages were built to dimensions of 6ft x 3ft x 3ft (L x W x H) with cage locations selected randomly based on the space available within the plot and arranged in sets of three.
@@
 inermis* stems were collected from each cage.
-Each stem was sliced open using X-Acto® knives to collect data on infestation, dead larvae, live larvae, and parasitism at each internode.
+Each stem was dissected using X-Acto® knives to collect data on infestation, dead larvae and live larvae, as well as very rare parasitism events, for each internode.
 
 ### Montana Field Survey
+
+*Weather data and NDVI* Field sites were chosen across two counties, Choteau and Judith Basin, within the 'Golden Triangle' of Montana, United States [fig.
+1](#figure1), areas that consistently experience high *C.
+cinctus* pressure.
+We assessed the medium- and long-term temperature and precipitation trends of our field sites using weather data from the National Oceanic and Atmospheric Administration (NOAA, Silver Spring, MA, USA).
+Data for each field site was averaged from three of the closest weather stations to that location.
+
+To better understand how precipitation patterns vary throughout the growing season, we looked at local daily precipitation totals from weather stations at our sampling sites in Big Sandy and Moccasin, MT.
+Data was pulled online from historical WeatherUnderground weather station data [@WeatherUnderground].
+Precipitation readings were broken into two categories - dates before July 15 and after September 15, as well as between July 15 and September 15.
+These dates were chosen based on typical wheat harvest and historical periods of known *Bracon* spp activity.
+In the 'Golden Triangle', typical spring wheat harvest takes place in mid-July on average [@Wilson1923; @Zhu2021].
+
+We used the normalized difference vegetation index (NDVI) to compare the relative greening throughout the growing season between wheat fields and adjacent *B.
+inermis* ([fig.
+2](#figure2)).
+NDVI is typically used to assess vegetation health and density, and is calculated from the visible and near-infrared light reflected by vegetation [@Pettorelli2005].
+NDVI data measurements were directly downloaded using Google Earth Engine (Google Inc.
+2023, Mountain View, CA, USA).
 
@@
 inermis* and adjacent wheat fields.
-Sites were chosen across 2 counties in north-central Montana, United States.
-(Chouteau, Judith Basin), which consistently experience high *C.
-cinctus* pressure.
 Samples were collected from wheat fields and adjacent *B.
-inermis* in early July and late August in 2021, 2022, and 2023 from sites in Big Sandy, Moccasin, and Amsterdam, MT, USA.
+inermis* in early July and late August in 2022, and 2023 from sites in Big Sandy and Moccasin MT, USA.
 Sampling sites were set up as 100$m^2$ polygons along the edge of adjoining wheat fields.
@@
 Wheat and *B.
-inermis* stems were then returned to Montana State University, Bozeman, Montana and stored in a 10°C cold wet storage until dissection.
-Stems were dissected lengthwise with a fine-bladed scalpel to determine presence or absence of *C.
-cinctus* larvae infestation, live eggs, dead eggs, dead larvae, live larvae, parasitism, and cutting.
+inermis* stems were then returned to Montana State University, Bozeman, Montana and stored in a 10°C cold, humid storage until dissection.
+Stems were dissected in the same manner as samples from our controlled experiment.
+We recorded presence or absence of *C.
+cinctus* larval infestation, live eggs, dead eggs, dead larvae, live larvae and parasitism, in addition to stem cutting.
 Sawfly larvae were identified based on descriptions in Criddle (1915) and Wallace and McNeal (1996).
 
-*Statistical Analysis.* We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion os stems parasitized by *Bracon* spp.
-'
-
-To better understand the density of *Bracon* spp.
-within the wheat and *B.
-inermis* sampling sites, we converted our units to the ratio of parasitoids to stem per unit area.
-Densities of stems are different when considering *B.
-inermis* and cultivated wheat or barley.
-Using this conversion, we attempted to better understand how the density of parasitoids (and WSS) is changed by looking at each kind of plant.
+To better understand the densities of host *C.
+cinctus* and the *Bracon* spp.
+within the wheat and adjacent *B.
+inermis* at each sampling site, we converted our units to the ratio of parasitoids to surviving hosts stem per unit area.
+Densities of stems are different when considering semi-natural, long-standing stands of *B.
+inermis* and the adjacent cultivated wheat or barley.
+
+### Statistical Analysis
+
+All statistical analyses were done in R [@R2024] using R Studio (version 4.3.2).
+Data was plotted using R package 'ggplot' (version 3.4.4) [@ggplot].
+We analyzed long- and short-term weather trends using linear modeling and the 'lm' command.
+Weather data was analyzed using either average yearly precipitation (inches) or average yearly temperature (°C) as the response variable and year as the predictor.
+Similar to our weather data analysis, NDVI data was linear modeled to discern differences in NDVI measurements over time.
+
+We used generalized linear mixed models with binomial errors (logit link) fit using bound optimization by quadratic approximation, with a maximum of 200,000 iterations, were run in the *lme4* package in R [@lme4] to examine the effects of location and year (fixed effects) on each of the three response variables: proportion of stems infested, proportion of stems cut, and proportion of stems parasitized by *Bracon* spp.
 
@@
 library(lme4)
-```
-
-### Historical Weather Analysis & NDVI
+library(gridExtra)
+```
+
+### Controlled *C. cinctus* Infestation of *B. inermis*
+
+```{r column.order}
+column.order <- c("distance", "collection", "p.inf", 
+                  "p.cut", "p.par", "poten")
+
+column.order2 <- c("location", "distance", "collection", "p.inf", 
+                  "p.cut", "p.par", "poten")
+
+round_based_on_size <- function(x) {
+  threshold <- 0.01
+  rounded_values <- ifelse(abs(x) > threshold, round(x, 2), round(x, 3))
+  return(rounded_values)
+}
+```
+
+```{r (pf.RAW) post farm n.inf data}
+pf.raw.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2022/Post\ Farm\ 2022.xlsx", sheet = "n.inf")
+pf.raw.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "n.inf")
+```
+
+```{r (pf.inf.23) post farm inf data}
+pf.inf.23 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf") %>%
+  rename(n.1 = 8, n.2 = 9, n.3 = 10, n.4 = 11,
+         n.5 = 12, n.6 = 13, n.7 = 14, n.8 = 15,
+         n.9 = 16, n.10 = 17)
+```
+
+```{r (pf.comb) post farm data combined}
+# 2022
+pf.2022.site <-
+pf.raw.2022 %>%
+  group_by(treatment, site) %>%
+  reframe(site, treatment,
+            p.inf = sum(iv, ir, ix)/sum(uv, ur, iv, ir, ix),
+            cut = sum(cut)/sum(total),
+            total.stems = sum(iv, ir, ix, uv, ur)) %>%
+  unique() %>%
+  pivot_longer(-c(treatment, total.stems, site),
+               names_to = "type",
+               values_to = "percent") %>%
+  mutate(year = "2022")
+
+# 2023
+pf.2023.site <-
+pf.raw.2023 %>%
+  group_by(treatment, site) %>%
+  reframe(treatment, site, 
+          t.inf = sum(n.iv, n.ir, n.ix),
+          p.inf = sum(n.iv, n.ir, n.ix)/sum(n.uv, n.ur, 
+                                            n.iv, n.ir, n.ix),
+          cut = sum(cut)/sum(total),
+          total.stems = sum(n.iv, n.ir, n.ix, n.uv, n.ur)) %>%
+  distinct() %>%
+  pivot_longer(-c(treatment, total.stems, site),
+               names_to = "type",
+               values_to = "percent") %>%
+  filter(!type == "t.inf") %>%
+  arrange(type) %>%
+  mutate(year = "2023")
+
+pf.comb <-
+  pf.2022.site %>%
+  rbind(pf.2023.site) %>%
+  group_by(treatment, year, type) %>%
+  mutate(count = n(),
+         average = mean(percent),
+         sd = sd(percent),
+         se = (sd/sqrt(count))) %>%
+  reframe(treatment, year, type,
+          mean = mean(percent),
+          sd = mean(sd),
+          se = mean(se)) %>%
+  unique()
+
+```
+
+```{r (pf.total) pf total stems}
+# 2022 totals
+pf.2022.total <-
+  pf.raw.2022 %>%
+  group_by(treatment) %>%
+  reframe(total = sum(uv, ur, iv, ir, ix)) %>%
+  mutate(year = "2022")
+
+# 2023 totals
+pf.2023.total <-
+  pf.raw.2023 %>%
+  group_by(treatment) %>%
+  reframe(total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
+  mutate(year = "2023")
+
+pf.total.comb <-
+  pf.2022.total %>%
+  rbind(pf.2023.total)
+
+```
+
+```{r pf_lm}
+raw.23 <-
+pf.raw.2023 %>%
+  dplyr::select(site, treatment,
+                n.iv, n.ir, n.ix, n.ur, n.uv,
+                cut) %>%
+  rename(iv = n.iv,
+         ir = n.ir,
+         ur = n.ur,
+         ix = n.ix,
+         uv = n.uv) %>%
+  mutate(year = 2023)
+
+raw.22 <-
+pf.raw.2022 %>%
+  dplyr::select(site, treatment, 
+                ix, ir, iv, ur, uv,
+                cut) %>%
+  mutate(year = 2022)
+
+pf_inf_lm <-
+raw.23 %>%
+  rbind(raw.22) %>%
+  group_by(site, treatment, year) %>%
+  reframe(total = sum(iv, ir, ix, ur, uv),
+          inf = sum(iv, ir, ix)) %>%
+  lm(inf ~ treatment + year, data = .) %>%
+  summary()
+
+pf_cut_lm <-
+raw.23 %>%
+  rbind(raw.22) %>%
+  group_by(site, treatment, year) %>%
+  reframe(total = sum(iv, ir, ix, ur, uv),
+          inf = sum(iv, ir, ix),
+          cut = sum(cut)) %>%
+  lm(cut ~ treatment + year, data = .) %>%
+  summary()
+```
+
+```{r pf_per, eval = F}
+raw.23 %>%
+  rbind(raw.22) %>%
+  group_by(treatment) %>%
+  reframe(total = sum(iv, ir, ix, ur, uv),
+          inf = sum(iv, ir, ix),
+          cut = sum(cut),
+          p.inf = inf/total,
+          p.cut = cut/inf)
+```
+
+*C.
+cinctus* heavily infested stems of *B.
+inermis* in the controlled test conditions.
+Averaged across both years, we observed 66.5% of stems infested for high treatments and 47.3% of stems for low treatments.
+We found strong evidence suggesting that there was a significant difference between infestation at high and low treatment levels when holding year constant (*r = `r round(pf_inf_lm$r.squared,3)`, P \< 0.05*, [Fig.
+3](#figure3)).
+
+```{r fig4_inf_cut_plot code}
+pf.comb.hl <-
+pf.comb %>%
+  filter(!treatment == "control")
+
+inf_cut_plot <-
+ggplot() +
+  facet_wrap(~treatment) +
+  geom_bar(stat = "identity",
+           position = position_dodge(),
+           data = pf.comb.hl %>%
+             filter(type == "p.inf"),
+           aes(year, mean, fill = "p.inf")) +
+  geom_bar(stat = "identity",
+           position = position_dodge(),
+           data = pf.comb.hl %>%
+             filter(type == "cut"),
+           aes(year, mean, fill = "cut")) +
+  geom_errorbar(data = pf.comb.hl,
+                aes(x = year,
+                    ymin = mean - se,
+                    ymax = mean + se),
+                position = position_dodge(0.9),
+                width = 0.2,
+                alpha = 0.5) +
+  theme_bw() +
+  labs(x = element_blank(),
+       y = "Percentage of stems (%)",
+       fill = element_blank()) +
+  scale_fill_brewer(palette = "Greys",
+                    labels = c("Cut", "Infested"))
+```
+
+Stem cutting by mature larvae, leading into winter, was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P < 0.05*).
+There was a dramatic decrease in survival observed from initial infestation to stem cutting.
+
+```{r figure_5_pf_inf_nodes}
+pf.y.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2022.xlsx", sheet = "inf")
+pf.y.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf")
+
+# DATA FOR THE PLOT
+pf.nodes.plot.data <-
+pf.y.2023 %>%
+  dplyr::select(-c(Name, 'Date Proc')) %>%
+  dplyr::select(treatment, 6:15) %>%
+  pivot_longer(-c(treatment),
+               names_to = "node",
+               values_to = "value") %>%
+  mutate(node = gsub("n_", "", node)) %>%
+  na.omit() %>%
+  mutate(year = "2023") %>%
+  rbind(pf.y.2022 %>%
+          dplyr::select(-c(site1, site2, site12, site3,
+                           Name, 'Date Proc')) %>%
+          dplyr::select(treatment, 
+                        2, 3, 4, 5, 6, 
+                        7, 8, 9, 10, 11) %>%
+          pivot_longer(-c(treatment),
+                       names_to = "node",
+                       values_to = "value") %>%
+          na.omit() %>%
+          mutate(year = "2022")) %>%
+  group_by(year, treatment, node) %>%
+  reframe(count = n()) %>%
+  group_by(year, treatment) %>%
+  mutate(per = count/sum(count)) %>%
+  mutate(node = as.numeric(node))
+
+# PLOT
+pf_nodes_plot <-
+  pf.nodes.plot.data %>%
+  ggplot() +
+  geom_bar(stat = 'identity',
+           aes(x = node, y = per, fill = treatment),
+           position = position_dodge(),
+           color = "black") +
+  theme_bw() +
+  scale_fill_brewer(palette = "Greys") +
+  scale_x_continuous(breaks = 1:10) +
+  labs(x = "Stem Internode",
+       y = "Percent of internodes infested") +
+  facet_wrap(~year)
+
+pf_nodes_over_4_table <-
+pf.nodes.plot.data %>%
+  group_by(year, node) %>%
+  reframe(count = sum(count)) %>%
+  group_by(year) %>%
+  mutate(per = count/sum(count)) %>%
+  filter(node > 4) %>%
+  group_by(year) %>%
+  reframe(sum = sum(per))
+```
+
+*C.
+cinctus* infested stems often contained larvae burrowing through multiple nodes.
+In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through, or 5 internodes occupied.
+In 2023, `r round(pf_nodes_over_4_table[2,2],3)*100`% of stems had at least 4 nodes infested [Fig.
+4](#figure4)
+
+### Montana Field Sites
+
+#### Historical Weather Analysis & NDVI
 
 ```{r moc_weather}
+
 # moccasin NOAA
@@
        x = NULL,
-       y = "Average Yearly Temperature (°C)",
-       title = "Average Yearly Temperature",
-       subtitle = "Moccasin, MT, USA")
-
-```
-
+       y = "Average Yearly Temperature (°C)")
+
+```
+
+```{r bs_seasonal_plot}
+lochy_precip <-
+read_excel("weather/precip_scrape.xlsx", sheet = "big_sandy")
+
+line_table <-
+lochy_precip %>%
+  dplyr::select(date, prcp) %>%
+  mutate(month = month(date),
+         day = day(date),
+         year = as.factor(year(date))) %>%
+  mutate(new_date = make_date(year = 0,
+                              month = month,
+                              day = day)) %>%
+   mutate(category = case_when(new_date < "0000-07-15" ~ "pre",
+                              new_date > "0000-09-15" ~ "post",
+                              new_date >= "0000-07-15" & new_date <= "0000-09-15" ~ "mid")) %>%
+  mutate(model_category = if_else(category == "mid", 
+                                  "mid", "no"))
+
+line_table_average <-
+  line_table %>%
+  group_by(month, day) %>%
+  reframe(category, new_date,
+          prcp = mean(prcp))
+
+category_averages <-
+line_table_average %>%
+  na.omit() %>%
+  group_by(category) %>%
+  reframe(prcp = mean(prcp))
+
+bs_seasonal_lm <-  
+line_table %>%
+  lm(prcp ~ model_category + year, data = .) %>%
+  summary()
+
+mult = 3
+
+smooth_plot <-
+ggplot() +
+  geom_smooth(aes(x = new_date, y = prcp),
+              data = line_table_average,
+              se = FALSE,
+              method = 'loess',
+              color = "black") +
+  geom_rect(aes(xmin = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0528 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-07-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0528 * mult, yend = 0.0528 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0164 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-09-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0164 * mult, yend = 0.0164 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-10-31", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0281 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-10-31", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0281 * mult, yend = 0.0281 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  theme_bw()
+
+bs_seasonal_plot <-
+ggplot() +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "pre"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "post"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "mid"),
+            color = "black") +
+  geom_rect(aes(xmin = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0528 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-07-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0528 * mult, yend = 0.0528 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0164 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-09-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0164 * mult, yend = 0.0164 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-10-31", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0281 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-10-31", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0281 * mult, yend = 0.0281 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  labs(x = NULL,
+       y = "Average Daily Precipitation (in)",
+       color = NULL) +
+  scale_y_continuous(sec.axis = sec_axis(~ . * mult, 
+                                         name = "Average Seasonal Precpitation (in)")) +
+  scale_x_date(
+    date_breaks = "1 month",
+    date_labels = "%B") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r moc_seasonal_plot}
+moc_precip <-
+read_excel("weather/precip_scrape.xlsx", sheet = "moccasin")
+
+line_table <-
+moc_precip %>%
+  dplyr::select(date, prcp) %>%
+  mutate(month = month(date),
+         day = day(date),
+         year = as.factor(year(date))) %>%
+  mutate(new_date = make_date(year = 0,
+                              month = month,
+                              day = day)) %>%
+   mutate(category = case_when(new_date < "0000-07-15" ~ "pre",
+                              new_date > "0000-09-15" ~ "post",
+                              new_date >= "0000-07-15" & new_date <= "0000-09-15" ~ "mid")) %>%
+  mutate(model_category = if_else(category == "mid", 
+                                  "mid", "no"))
+
+line_table_average <-
+  line_table %>%
+  group_by(month, day) %>%
+  reframe(category, new_date,
+          prcp = mean(prcp))
+
+category_averages <-
+line_table_average %>%
+  na.omit() %>%
+  group_by(category) %>%
+  reframe(prcp = mean(prcp))
+
+moc_seasonal_lm <-  
+line_table %>%
+  lm(prcp ~ model_category + year, data = .) %>%
+  summary()
+
+mult = 3
+
+moc_seasonal_plot <-
+ggplot() +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "pre"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "post"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "mid"),
+            color = "black") +
+  geom_rect(aes(xmin = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0528 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-05-01", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-07-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0528 * mult, yend = 0.0528 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0164 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-07-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-09-15", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0164 * mult, yend = 0.0164 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  geom_rect(aes(xmin = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                xmax = as.Date("0000-10-31", 
+                               format = "%Y-%m-%d"),
+                ymin = 0, 
+                ymax = 0.0281 * mult),
+            alpha = 0.1,
+            fill = "blue") +
+  geom_segment(aes(x = as.Date("0000-09-15", 
+                               format = "%Y-%m-%d"), 
+                   xend = as.Date("0000-10-31", 
+                                  format = "%Y-%m-%d"),
+                   y = 0.0281 * mult, yend = 0.0281 * mult),
+               color = "blue",
+               alpha = 0.5) +
+  labs(x = NULL,
+       y = "Average Daily Precipitation (in)",
+       color = NULL) +
+  scale_y_continuous(sec.axis = sec_axis(~ . * mult, 
+                                         name = "Average Seasonal Precpitation (in)")) +
+  scale_x_date(
+    date_breaks = "1 month",
+    date_labels = "%B") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+In both Big Sandy and Moccasin, MT, we observed a decrease in average yearly precipitation and increase in average yearly temperature.
 We observed a significant linear relationship (*r = `r round(both_prcp_lm$r.squared,3)`, P = `r round(both_prcp_lm$coefficients[2,4],3)`, estimate = `r round(both_prcp_lm$coefficients[2,1],3)`*) between average precipitation and year for both Big Sandy and Moccasin, Montana.
-This means that for each one year increase, we are seeing a decrease in 0.05 inches of precipitation.
+For each location and one-year increase, we observed a 0.05 inch decrease in precipitation.
 ([Fig.
-1](#figure1)).
+5](#figure5)).
 Additionally, we observed a significant positive linear relationship (*r = `r round(both_temp_lm$r.squared,3)`, P \< 0.05, estimate = `r round(both_temp_lm$coefficients[2,1],3)`*) between year and average yearly temperature for Moccasin and Big Sandy, MT.
-This means that each year, the average daily temperature has increased by `r round(both_temp_lm$coefficients[2,1],2)`°C.
+Thus, each year, the average daily temperature has increased by `r round(both_temp_lm$coefficients[2,1],2)`°C.
 ([Fig.
-2](#figure2)).
+6](#figure6)).
+Analyzing local, precipitation data over the past five years yielded information on how central and northern Montana experience relative dryness during the late-summer months.
+In Big Sandy, when controlling for year, we saw strong evidence of a significant difference (*p<0.05, F=2.067*, [fig.
+7](#figure7)) between daily rainfall in months prior July 15 and after September 15 compared to days within the July 15 to September 15 'late-summer' period.
+Analysis of Moccain precipitation data revealed that there is little evidence (*p=0.08, F=1.535*) to suggest a difference between the two date categories.
 
@@
   geom_label(aes(x = label_x,
-                 y = 0.75),
-             label = "Spring wheat harvest",
+                 y = 0.8),
+             label = "Average spring wheat harvest",
              hjust = -0.1,
              size = 3) +
-  geom_segment(aes(x = as.Date("2023-07-20")+5, 
-                   y = 0.75, 
-                   xend = as.Date("2023-07-20"), 
-                   yend = 0.75)) +
   theme_bw() +
   guides(color = FALSE) +
-  labs(x = "2023 Growing Season",
+  labs(x = NULL,
        y = "NDVI",
-       linetype = NULL,
-       title = "Seasonal NDVI Comparison",
-       subtitle = expression(paste(italic("B. inermis"),
-                                   " and adjacent spring wheat"))) +
-  scale_linetype_discrete(labels = c(expression(italic("B. inermis")),
-                                                      "Spring Wheat"))
+       linetype = NULL) +
+  scale_linetype_discrete(labels = c(expression(italic("B. inermis")),"Spring Wheat"))
 ```
@@
 inermis* and spring wheat.
-We saw a significant difference in the July NDVI (*`r round(ndvi_model$r.squared,3)`, P \< 0.05*).
+We saw a significant difference in the *B.
+inermis* post-July 15 NDVI (*`r round(ndvi_model$r.squared,3)`, P<0.05*) when compared to the post-July 15 NDVI of the adjacent wheat crop.
 The *B.
-inermis* NDVI remained relatively linear in it's downslope (BROME SLOPE POST JULY) compared to the wheat field (WHEAT FIELD POST JULY).
-([Fig.
-3](#figure3))
-
-Need to add map showing where NDVI measurments came from
-
-### Controlled *C. cinctus* Infestation of *B. inermis*
-
-```{r column.order}
-column.order <- c("distance", "collection", "p.inf", 
-                  "p.cut", "p.par", "poten")
-
-column.order2 <- c("location", "distance", "collection", "p.inf", 
-                  "p.cut", "p.par", "poten")
-
-round_based_on_size <- function(x) {
-  threshold <- 0.01
-  rounded_values <- ifelse(abs(x) > threshold, round(x, 2), round(x, 3))
-  return(rounded_values)
-}
-```
-
-```{r (pf.RAW) post farm n.inf data}
-pf.raw.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2022/Post\ Farm\ 2022.xlsx", sheet = "n.inf")
-pf.raw.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "n.inf")
-```
-
-```{r (pf.inf.23) post farm inf data}
-pf.inf.23 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf") %>%
-  rename(n.1 = 8, n.2 = 9, n.3 = 10, n.4 = 11,
-         n.5 = 12, n.6 = 13, n.7 = 14, n.8 = 15,
-         n.9 = 16, n.10 = 17)
-```
-
-```{r (pf.comb) post farm data combined}
-# 2022
-pf.2022.site <-
-pf.raw.2022 %>%
-  group_by(treatment, site) %>%
-  reframe(site, treatment,
-            p.inf = sum(iv, ir, ix)/sum(uv, ur, iv, ir, ix),
-            cut = sum(cut)/sum(total),
-            total.stems = sum(iv, ir, ix, uv, ur)) %>%
-  unique() %>%
-  pivot_longer(-c(treatment, total.stems, site),
-               names_to = "type",
-               values_to = "percent") %>%
-  mutate(year = "2022")
-
-# 2023
-pf.2023.site <-
-pf.raw.2023 %>%
-  group_by(treatment, site) %>%
-  reframe(treatment, site, 
-          t.inf = sum(n.iv, n.ir, n.ix),
-          p.inf = sum(n.iv, n.ir, n.ix)/sum(n.uv, n.ur, 
-                                            n.iv, n.ir, n.ix),
-          cut = sum(cut)/sum(total),
-          total.stems = sum(n.iv, n.ir, n.ix, n.uv, n.ur)) %>%
-  distinct() %>%
-  pivot_longer(-c(treatment, total.stems, site),
-               names_to = "type",
-               values_to = "percent") %>%
-  filter(!type == "t.inf") %>%
-  arrange(type) %>%
-  mutate(year = "2023")
-
-pf.comb <-
-  pf.2022.site %>%
-  rbind(pf.2023.site) %>%
-  group_by(treatment, year, type) %>%
-  mutate(count = n(),
-         average = mean(percent),
-         sd = sd(percent),
-         se = (sd/sqrt(count))) %>%
-  reframe(treatment, year, type,
-          mean = mean(percent),
-          sd = mean(sd),
-          se = mean(se)) %>%
-  unique()
-
-```
-
-```{r (pf.total) pf total stems}
-# 2022 totals
-pf.2022.total <-
-  pf.raw.2022 %>%
-  group_by(treatment) %>%
-  reframe(total = sum(uv, ur, iv, ir, ix)) %>%
-  mutate(year = "2022")
-
-# 2023 totals
-pf.2023.total <-
-  pf.raw.2023 %>%
-  group_by(treatment) %>%
-  reframe(total = sum(n.uv, n.ur, n.iv, n.ir, n.ix)) %>%
-  mutate(year = "2023")
-
-pf.total.comb <-
-  pf.2022.total %>%
-  rbind(pf.2023.total)
-
-```
-
-```{r pf_lm}
-raw.23 <-
-pf.raw.2023 %>%
-  dplyr::select(site, treatment,
-                n.iv, n.ir, n.ix, n.ur, n.uv,
-                cut) %>%
-  rename(iv = n.iv,
-         ir = n.ir,
-         ur = n.ur,
-         ix = n.ix,
-         uv = n.uv) %>%
-  mutate(year = 2023)
-
-raw.22 <-
-pf.raw.2022 %>%
-  dplyr::select(site, treatment, 
-                ix, ir, iv, ur, uv,
-                cut) %>%
-  mutate(year = 2022)
-
-pf_inf_lm <-
-raw.23 %>%
-  rbind(raw.22) %>%
-  group_by(site, treatment, year) %>%
-  reframe(total = sum(iv, ir, ix, ur, uv),
-          inf = sum(iv, ir, ix)) %>%
-  lm(inf ~ treatment + year, data = .) %>%
-  summary()
-
-pf_cut_lm <-
-raw.23 %>%
-  rbind(raw.22) %>%
-  group_by(site, treatment, year) %>%
-  reframe(total = sum(iv, ir, ix, ur, uv),
-          inf = sum(iv, ir, ix),
-          cut = sum(cut)) %>%
-  lm(cut ~ treatment + year, data = .) %>%
-  summary()
-```
-
-```{r pf_per, eval = F}
-raw.23 %>%
-  rbind(raw.22) %>%
-  group_by(treatment) %>%
-  reframe(total = sum(iv, ir, ix, ur, uv),
-          inf = sum(iv, ir, ix),
-          cut = sum(cut),
-          p.inf = inf/total,
-          p.cut = cut/inf)
-```
-
-*C.
-cinctus* showed high infestation of *B.
-inermis* in the controlled test conditions.
-Averaged across both years, we observed 66.5% of stems infested for high treatments and 47.3% of stems for low treatments.
-We found strong evidence suggesting that there was a significant difference between infestation at high and low treatment levels when holding year constant (*r = `r round(pf_inf_lm$r.squared,3)`, P \< 0.05*, [Fig.
-3](#figure4)).
-
-```{r fig4_inf_cut_plot code}
-inf_cut_plot <-
-ggplot() +
-  facet_wrap(~treatment) +
-  geom_bar(stat = "identity",
-           position = position_dodge(),
-           data = pf.comb %>%
-             filter(type == "p.inf"),
-           aes(year, mean, fill = "p.inf")) +
-  geom_bar(stat = "identity",
-           position = position_dodge(),
-           data = pf.comb %>%
-             filter(type == "cut"),
-           aes(year, mean, fill = "cut")) +
-  geom_errorbar(data = pf.comb,
-                aes(x = year,
-                    ymin = mean - se,
-                    ymax = mean + se),
-                position = position_dodge(0.9),
-                width = 0.2,
-                alpha = 0.5) +
-  theme_bw() +
-  labs(x = element_blank(),
-       y = "Percentage of stems (%)",
-       fill = element_blank()) +
-  scale_fill_brewer(palette = "Greys",
-                    labels = c("Cut", "Infested"))
-```
-
-Cutting was observed at 5.7% for the high treatments and 3.9% for the low, showing strong evidence for a difference in cutting between high and low treatment groups (*r = `r round(pf_cut_lm$r.squared,3)`, P < 0.05*).
-
-```{r figure_5_pf_inf_nodes}
-pf.y.2022 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2022.xlsx", sheet = "inf")
-pf.y.2023 <- read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/BWW/Field/Post\ Farm/2023/Post\ Farm\ 2023.xlsx", sheet = "inf")
-
-# DATA FOR THE PLOT
-pf.nodes.plot.data <-
-pf.y.2023 %>%
-  dplyr::select(-c(Name, 'Date Proc')) %>%
-  dplyr::select(treatment, 6:15) %>%
-  pivot_longer(-c(treatment),
-               names_to = "node",
-               values_to = "value") %>%
-  mutate(node = gsub("n_", "", node)) %>%
-  na.omit() %>%
-  mutate(year = "2023") %>%
-  rbind(pf.y.2022 %>%
-          dplyr::select(-c(site1, site2, site12, site3,
-                           Name, 'Date Proc')) %>%
-          dplyr::select(treatment, 
-                        2, 3, 4, 5, 6, 
-                        7, 8, 9, 10, 11) %>%
-          pivot_longer(-c(treatment),
-                       names_to = "node",
-                       values_to = "value") %>%
-          na.omit() %>%
-          mutate(year = "2022")) %>%
-  group_by(year, treatment, node) %>%
-  reframe(count = n()) %>%
-  group_by(year, treatment) %>%
-  mutate(per = count/sum(count)) %>%
-  mutate(node = as.numeric(node))
-
-# PLOT
-pf_nodes_plot <-
-  pf.nodes.plot.data %>%
-  ggplot() +
-  geom_bar(stat = 'identity',
-           aes(x = node, y = per, fill = treatment),
-           position = position_dodge(),
-           color = "black") +
-  theme_bw() +
-  scale_fill_brewer(palette = "Greys") +
-  scale_x_continuous(breaks = 1:10) +
-  labs(x = "Stem Internode",
-       y = "Percent of internodes infested",
-       title = expression(paste("Percentage of ",
-                                italic("B. inermis "),
-                                "Internodes Infested")),
-       subtitle = "Controlled infestation") +
-  facet_wrap(~year)
-
-pf_nodes_over_4_table <-
-pf.nodes.plot.data %>%
-  group_by(year, node) %>%
-  reframe(count = sum(count)) %>%
-  group_by(year) %>%
-  mutate(per = count/sum(count)) %>%
-  filter(node > 4) %>%
-  group_by(year) %>%
-  reframe(sum = sum(per))
-```
-
-*C.
-cinctus* infested stems often contained larvae burrowing through multiple nodes.
-In 2022, `r round(pf_nodes_over_4_table[1,2],3)*100`% of stems were observed to have at least 4 nodes burrowed through.
-In 2023, `r round(pf_nodes_over_4_table[2,2],3)*100`% of stems had at least 4 nodes infested [Fig.
-5](#figure5)
-
-### Montana Field Sites
-
+inermis* NDVI remained relatively linear in it's downslope compared to the wheat field ([Fig.
+7](#figure7)).
+
+#### Field Site Samples
 ```{r stubble data}
@@
 parasitoid presence in 6,148 wheat and grass stems across 10 research sites in 2022 and 2023.
-Infestation by *C.
-cinctus* within *B.
-inermis* varied between collection sites, with the greatest infestation taking place within our three Big Sandy, MT sampling sites (2023: 65.4%, 2022: 63.1%), while the lowest infestation observed was in Moccasin, MT (2023: 40.8%, 2022: 60.7%).
+Initial infestation of *B.
+inermis* by *C.
+cinctus* varied between collection sites, with the greatest infestation taking place within our three Big Sandy, MT sampling sites (2023: 65.4%, 2022: 63.1%), while the lowest infestation observed was in Moccasin, MT (2023: 40.8%, 2022: 60.7%).
 Across all sites in Big Sandy and Moccasin, we observed an average infestation of 57.5% within *B.
@@
 Cutting by *C.
-cintus* was observed at an average rate of 5.59% (se = 0.0164) in *B.
-inermis* and 47.1% (se = 0.0386) in the adjacent cultivated wheat ([fig.
-7](#figure7).
+cintus* was observed at an average rate of 5.59% in *B.
+inermis* and 47.1% in the adjacent cultivated wheat ([Fig.
+8](#figure8).
 
@@
 cinctus* host larvae within *B.
-inermis* late in the growing season, we calculated the number of stems that contained at least four or more nodes of infestation.
-In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of infestation.
-In Moccasin, the corresponding figure was 66.5%.
+inermis* late in the growing season, we calculated the number of stems that contained at least four or more nodes with burrowing injury.
+In Big Sandy, 76.2% of collected stems in the fall exhibited at least 4 nodes of burrowing injury at our second sampling date.
+In Moccasin, the corresponding figure was 66.5% ([Fig.
+9](#figure9)).
 
@@
   distinct()
+
+```
+
+```{r fig_9_brome_par_plot}
+brome_par_plot <-
+plot.data %>%
+  filter(
+         plant_type == "brome",
+         variable == "per.par") %>%
+  mutate(collection = factor(collection, levels = c("summer", "fall"))) %>%
+  ggplot() +
+  geom_bar(stat = "identity",
+           aes(x = sample_site, y = value, fill = collection),
+           position = position_dodge(),
+           width = 0.5,
+           color = "black") +
+  geom_errorbar(aes(x = sample_site,
+                    group = collection,
+                    ymin = value - se,
+                    ymax = value + se),
+                position = position_dodge(width = 0.5),
+                width = 0.15) +
+  theme_bw() +
+  labs(x = NULL) +
+  scale_fill_brewer(palette = "Greys") +
+  facet_wrap(year~region, scales = "free_x")
+```
+
+```{r brome_par_model}
+brome_par_model_data <-
+comb.brome %>%
+  rbind(comb.stubble) %>%
+  na.omit() %>%
+  filter(!location %in% c("chs", "b")) %>%
+  mutate(sample_site = case_when(
+    location %in% c("nn", "ss", "ns", "n") ~ "North Road",
+    location %in% c("kw", "ke", "k") ~ "Kolin Road",
+    location %in% c("sun", "har") ~ "Sunset & Harvest",
+    location == "g" ~ "Gardiner Road",
+    TRUE ~ NA_character_)) %>%
+  mutate(region = case_when(
+    location %in% c("nn", "ss", "ns", "n") ~ "Big Sandy",
+    location == "g" ~ "Big Sandy",
+    location %in% c("kw", "ke", "k") ~ "Moccasin",
+    location %in% c("sun", "har") ~ "Moccasin",
+    TRUE ~ NA_character_)) %>%
+  mutate(plant_type = if_else(distance == "brome", "brome", "wheat")) %>%
+  dplyr::select(year, region, sample_site, collection, plant_type,
+                total.stems, inf, uninf, cut, c, e) %>%
+  group_by(site, collection, year) %>%
+  mutate(par = sum(c,e)) %>% dplyr::select(-c(c,e)) %>%
+  group_by(year, collection, region, sample_site, plant_type, site) %>%
+  mutate(count = n()) %>%
+  ungroup() %>%
+  dplyr::select(-c(uninf)) %>%
+  group_by(year, collection, region, sample_site, plant_type, site)
+
+brome_par_model <-
+brome_par_model_data %>%
+  filter(plant_type == "brome") %>%
+  lm(par ~ year + region + collection, data = .) %>%
+  summary()
+
+```
+
+```{r per.par, eval = F}
+comb.stubble %>%
+  mutate(region = if_else(location %in% c("har", "sun", "k", "chs"), "moccasin", "big sandy")) %>%
+  group_by(region) %>%
+  reframe(total.stems = sum(total.stems),
+          inf = sum(inf),
+          cut = sum(cut),
+          par = sum(c, e),
+          inf.per = inf/total.stems,
+          cut.per = cut/inf,
+          par.per = par/inf) %>%
+  dplyr::select(par.per)
+
+comb.stubble %>%
+  dplyr::select(-c(uninf, cut, c, e, total.stems))
 ```
@@
 inermis* sink/source ecology.
-The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to 1 cut stem.
-This means that for every 1 square meter of *B.
-inermis*, we would observe 16 stems that produced a parasitoid for every 1 stem that produced an adult *C.
+The largest ratio observed was in Big Sandy, MT in the fall of 2022, where we observed a ratio of 16 parasitoids to a single *C.
+cinctus*-cut stem.
+Thus, for every square meter of *B.
+inermis*, we would observe 16 stems that produced a parasitoid for every 1 stem that was likely to produce an adult *C.
 cinctus*.
-All three other sample selections also showed positive ratios of parasitoids to cut stems [fig.
-8](#figure8).
+All other sample locations also showed positive ratios of parasitoids to cut stems [(Fig.
+10)](#figure10).
+
+To understand the seasonal trends of *Bracon* spp use of potential *C.
+cinctus* hosts within *B.
+inermis*, we collected stem samples twice each growing season.
+We collected one pre-harvest, in July, and a second time in September, after all adult *Bracon* spp would be dead.
+We observed a significant difference (*df = 96, r = 0.304, p < 0.05*) between parasitoid presence in infested stems collected in Big Sandy and Moccasin, MT.
 
@@
 ## Discussion
+<!-- CONTROLLED INFESTATION -->
+Assessment of *C.
+cicntus* infestation and larval mortality within *B.
+inermis* from out controlled infestation survey showed high infestation coupled with high larval mortality prior to overwintering diapause.
+Even in cases of high infestation (70%+), larval survival to cutting topped out around around 15%, pointing towards *B.
+inermis'* ability to serve as a quality sink for *C.
+cinctus* larvae.
+These results confirm the assertions by Criddle (1992) and Seamans (1928) that *B.
+inermis* can serve as a quality *C.
+cinctus* trap crop.
+In addition to high larval mortality, stem dissection exhibited high numbers of stems with over four internodes burrowed through by *C.
+cinctus* larvae.
+Larval movement through stems can be quick in the case of susceptible cultivated cereal grains [@Criddle1923].
+However, *B.
+inermis*, given it's presumed antibiosis properties, leads to slower larval development and thus movement through the stem.
+We found several *B.inermis* stems with living larvae as late as late-September.
+
+<!-- WEATHER DATA -->
+As the climate of central and northern Montana continues to change, wheat stem sawfly infestation will continue to impact cereal crop yields.
+Precipitation and temperature data from our sampling sites underscores the importance of continuing to monitor and conserve areas of semi-natural grassland.
+50-year trends analyzed in this study show similar results to Pederson et al (2009) in which the dry-land agricultural environments found in central and northern Montana are likely to continue to become hotter and dryer on average.
+Over the past fifty years, precipitation in Big Sandy has decreased by an average of 0.05 inches per year.
+Given the trend continues, by 2050 the area could experience an inch less of rain per year.
+
+Both average precipitation and average temperatures vary widely between years as the region goes through many droughts and periods of relative high rainfall [@WeatherUnderground; @Pederson2009].
+Shown in our analysis of daily measurements over the past five years, precipitation significantly decreases during the period after wheat harvest, July 15 - September15, a period when second generation *Bracon* adults are searching for *C.
+cinctus* larval hosts.
+Mid-season senesence of many *C.
+cinctus* host plants can then leave *Bracon* parasitoids with little resources for
+
+Prolonged drought, like seen in 2021 in Big Sandy, MT, can cause a dearth of green host plants in these semi-arid environments that can be problimatic for second generation adult parasitoids.
+
+Given our analysis of NDVI at our sampling sites, it is clear that *B.
+inermis* is capable of continued late seasonal greening and may provide important resources for *Bracon* spp parasitoids in years of drought or crop failure.
+It is these buffer zones where parasitoids may be able to find *C.
+cinctus* larvae in numbers great enough to prevent the larger population crashes that can lead large localized increases in *C.
+cinctus* populations and damage [@Holmes1982].
+
+Analysis of larval *C.
+cinctus* infestation of *B.
+inermis* and adjacent winter wheat reveal similar results to our controlled *B.
+inermis* infestation in Bozeman.
+We found *B.
+inermis* infestation to be similar to that of neighboring wheat fields, making it likely that *C.
+cinctus* pressures *B.
+inermis* similarly to cultivated wheat.
+
+<div class="box">
+Still working on adding some citations and information discussing Rand & Peirce papers and how my findings are similar and different.
+I collected samples twice throughout the year so got a bigger picture of how parasitism changes for each generation.
+It is looking like parasitism in the fall collections is much higher in the brome compared to the summer collections.
+Interesting!
+</div>
+
+### Montana Field Site Survey
+
+Analysis of weather trends in Big Sandy and Moccasin, MT confirmed that the  patterns in central and northern Montana are changing towards warmer and drier conditions [@Pederson2009].
+We observed an average decrease in 0.058 inches of precipitation per year since 1975, while since 1950, temperatures have experienced an average yearly increase of 0.028°C.
+These changes in precipitation and temperature can have significant impacts on timing of crop production and harvest timing [@Zhu2021].
+Year-over-year populations of the *Bracon* spp parasitoids are particularly sensitive, as their bivoltine nature requires the second yearly generation to find a suitable *C.
+cinctus* host [@Nelson1953; @Runyon2002].
+NDVI analyses exhibited the prolonged senescence period for *B.
+inermis* when compared to adjacent cultivated wheat.
+*C.
+cinctus* larvae are cued via the thinning of the cell membranes [@Villacorta1971] during plant senescence to travel to the base of the stem and prepare for obligatory diapause [@Beres2011].
+Because *B.
+inermis* senescence occurs at slower rates compared to wheat or barley, *C.
+cinctus* may remain within the stem later in into the summer and fall, providing potential host larvae for second generation *Bracon* parasitoids.
+
+Previous research has shown *B.
+inermis* to be a quality host of both *C.
+cinctus* and *Bracon* spp parasitoids [@Criddle1922; @Peirce2021; @Rand2024], but little data is available on the season fluctuation of *B.
+inermis* utilization [@Criddle1922; @Peirce2021; @Rand2024].
+Because we collected stem samples twice during the growing season, we were able to compare levels of parasitism.
+We hoped to understand how parasitism rates changed from July (pre-wheat harvest) and September (post-wheat harvest) of each year.
+
+Our results showed that parasitoid presence increase significantly at all sites between pre- and post-harvest collections.
+Given the collection methods, September collection parasitism was inherently a combination of both pre-harvest and post-harvest activity.
+While we cannot ascertain *Bracon* preference for host stems, we can assess that the parasitoids utilize infested *B.
+inermis* stems post-harvest when *C.
+cinctus* is no longer available within wheat stems.
+
+Additionally, we observed a high number of stems exhibiting levels of *C.
+cinctus* damage that point to prolonged larval presence within *B.
+inermis* stems.
+In Big Sandy (76.2%) and Moccasin (66.5%), stems that exhibited four or more nodes burrowed suggests that *C.
+cinctus* larvae remain available within *B.
+inermis* stems past wheat harvest, when they are more limited in availability.
+Most importantly, in years of drought or increased temperatures where wheat senescence and harvest may occur earlier in the summer.
+
+We utilized parasitism to cut stem ratios to understand the relative impact of *B.
+inermis* host sources on overall field parasitism.
+Assessing overall numbers of insects can potentially diminish the apparent importance of *B.
+inermis*, as per square meter stem counts in cultivated wheat are denser than *B.
+inermis* counts within coulees and along roadsides *B.
+inermis*.
+Our findings of significantly greater ratios of parasitized stems to cut stems within *B.
+inermis* when compared to adjacent when indicates these field-side areas provide greater parasitoid source per square meter when compared to cultivated fields.
+This emphasizes the significance of accounting for habitat diversity to sustain parasitoid populations over successive years, as well as recognizing the crucial role of edge-row areas in maximizing parasitoid effectiveness in minimizing *C.
+cinctus* damage.
+
+### Old Discussion Text
+*C.
+cinctus* mortality was found to be high in controlled *B.
+inermis* infestation trials.
+Our results build upon other studies where *B.
+inermis* infestation confirmed previously reported high *C.
+cinctus* larval mortality prior to overwintering [@Shanower2004; @Mendoza2006; @Buteler2008].
+Of those infested stems, a large proportion exhibited larval damage in at least 5 internodes (62.5%).
+While the timing of stem collection and dissection did not align with precise measurement of living larvae (stems dissected in October), we can assert that stems with at least 4 nodes burrowed through would have a high probability of containing large, live larvae, and thus a viable potential parasitoid host, in late July and August of the same year.
+
+While total and per node *B inermis* infestation by *C.
+cinctus* was high, survival to overwintering (cutting) only occurred in 5.7% of *C.
+cinctus* infested stems.
+Our results indicate that *B.
+inermis* serves as a high-quality host for *C.
+cinctus* larvae during the spring and summer months.
+However, there was minimal survival of *C.
+cinctus* larvae beyond August and September, confirming that there will be little adult emergence from *B.
+inermis* the following season.
 
 Assessing the comparative abundance of *C.
-cinctus* and its associated parasitiods within *B.
-inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas could potentially serve as origins for both pests and beneficial natural adversaries infiltrating agricultural crops.
-
-<!-- Big Sandy -->
-
-<!-- Turn: 3m x 17m = 51$m^2$ -->
-
-<!-- North Road North: 5m x 20m = 100$m^2$ -->
-
-<!-- North Road South: 5m x 20m = 200$m^2$ -->
-
-<!-- Each sample: 1$m^2$ -->
-
-<!-- Wheat field sampled area 4 rows, 10 meters between each row samples taken at 5 and 20 distance from the edge This means the area of the sampled area is 20x4m = 80$m^2$ Each sample we'll count as 1m x 0.5m = 0.5$m^2$ This means 160 possible samples from the sampling area.
+cinctus* and its associated parasitoids within *B.
+inermis* and adjacent cultivated cereal hosts is an important initial step in evaluating whether semi-natural environments within road-side or coulee areas serve as sink or source origins of the pest and its beneficial natural adversaries infiltrating agricultural crops.
+
+<!-- Overall, this work suggests that maintaining and supporting field-edge areas containing *B.
+inermis* provides an unmistakable benefit to *C.
+cinctus* control in cereal crop production.
+The characteristics of *B.
+inermis* make it uniquely beneficial in both as a sink for *C.
+cinctus* larvae, and as a source for *Bracon* adults.
+*B.
+inermis* is a common grass found throughout the golden triangle of Montana and other wheat cultivation areas [@Bolwahn2011].
 -->
 
-<!-- Big Sandy NN Brome: 100$m^2$ - number of Wheat: 80$m^2$ -->
-
-<!-- Just need the density per unit area. don't need the total number of p-toids in the entire sampling area. -->
-
-<!-- Just need to standardize it to area so that the sampling size of wheat is the same as brome. These were different in the field. Brome: 1$m^2$ wheat: 0.5$m^2$ -->
-
-<!-- Also need: - weather data to see if there is a drought causing crashes/spikes - NDVI data to show greening of brome compared to wheat late in the season -->
-
 ## Figures
@@
 ### Figure 1 {#figure1}
-```{r FIG1_both_prcp_plot, fig.cap = "Figure 1: Average yearly precipitation (in) for Big Sandy and Moccasin, Montana.
+```{r figure1, fig.cap="Figure 1: Location of Montana, USA field sites.
+Controlled *B.
+inermis* infestation site was in Bozeman, while field sites were located in centrally location Moccasin, MT and northern Big Sandy, MT."}
+world_map <- map_data("world")
+states <- map_data("state") %>%
+  filter(region == "montana")
+counties <- map_data("county") %>%
+  filter(region == "montana")
+
+area <- c("Big Sandy", "Moccasin", "Bozeman")
+lats <- c(48.177540, 47.098790, 45.678871)
+longs <- c(-110.114060, -109.882962, -111.060815)
+
+coordinates <- data.frame(area, lats, longs)
+
+ggplot() +
+  # geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
+  #              fill = "white", color = "black") +
+  geom_polygon(data = counties, aes(x = long, y = lat, group = group),
+               fill = NA, color = "black", linewidth = 0.1) +
+  geom_polygon(data = states, aes(x = long, y = lat, group = group),
+               fill = NA, color = "black", linewidth = 0.5) +
+  geom_label(data = coordinates, aes(x = longs, y = lats), 
+             color = "black", 
+             label = area,
+             size = 5,
+             vjust = 0,
+             hjust = -0.1,
+             label.padding = unit(0.25, "lines")) +
+  geom_point(data = coordinates, aes(x = longs, y = lats),
+             color = "red") +
+  coord_cartesian(xlim = c(-115.75, -104.5), ylim = c(44.5, 49)) +
+  theme_void() +
+  # theme(panel.grid.major = element_blank(),
+  #       panel.grid.minor = element_blank()) +
+  theme(aspect.ratio = 1/1.8) +
+  labs(title = "Montana, USA")
+```
+
+### Figure 2 {#figure2}
+```{r bigsandy_ndvi, fig.cap="Figure 2: June 29, 2023 - Big Sandy, MT.
+Orange polygons denote location where NDVI was pulled using Google Earth Engine from both (upper left) Vida spring wheat and (central) coulee containing semi-natural grasses including majority *B.
+inermis*."}
+knitr::include_graphics("Images/BigSandy_NDVI_pic2.jpg")
+```
+
+### Figure 3 {#figure3}
+```{r inf_cut_plot, fig.cap = "Figure 3: Comparison of year and treatment group for controlled infestation of *B.
+inermis*.
+Three treatment groups - high, low, and control (0) - were used.
+The controlled groups showed no sign of *C.
+cinctus* stem damage.
+We observed a significant difference in cutting between high and low treatment groups (*r = 0.592, P < 0.05*)."}
+inf_cut_plot
+```
+
+### Figure 4 {#figure4}
+```{r figure5_pf_nodes_plot, fig.cap="Figure 4: Ineternodes infested by *B.
+cephi* within controlled infestation plots in Bozeman, MT, USA.
+We found that 61.2% of stems exhibited infestation in at least 5 nodes"}
+pf_nodes_plot
+```
+
+### Figure 5 {#figure5}
+```{r FIG1_both_prcp_plot, fig.cap = "Figure 5: Average yearly precipitation (in) for Big Sandy and Moccasin, Montana.
+Data gathered from NOAA.
 Black trend line signifies averaged negative trend between both locations.
@@
 ```
-
-### Figure 2 {#figure2}
-```{r both_temp_plot, fig.cap = "Figure 2: Average yearly temperature (°C) in Moccasin and Big Sandy, MT.
+### Figure 6 {#figure6}
+```{r both_temp_plot, fig.cap = "Figure 6: Average yearly temperature (°C) in Moccasin and Big Sandy, MT.
+Data gathered from NOAA.
 Positive trendline slop suggests increasing average temperatures of the past 100 years.
@@
 
-### Figure 3 {#figure3}
-```{r bs_ndvi_plot, fig.cap = "Figure 3: Normalized difference vegetation index (NDVI) of *B.
-inermis* and adjacent spring wheat field from April 2023 to October 2023.
+### Figure 7 {#figure7}
+```{r bs_ndvi_plot, fig.cap = "Figure 7: Normalized difference vegetation index (NDVI) of *B.
+inermis* and adjacent spring wheat field from April 2023 to October 2023 in Big Sandy, MT.
 Post harvest linear model indicates a significant difference (P values and stuff) when comparing the *B.
@@
 
-### Figure 4 {#figure4}
-```{r inf_cut_plot, fig.align='left', fig.cap = "Figure 4: Comparison of year and treatment group for controlled infestation of *B.
-inermis*.
-We observed a significant difference in cutting between high and low treatment groups (*r = 0.592, P < 0.05*)."}
-inf_cut_plot
-```
-
-### Figure 5 {#figure5}
-```{r figure5_pf_nodes_plot, fig.align='left', fig.cap="Figure 6: Ineternodes infested by *B.
-cephi* within controlled infestation plots in Bozeman, MT, USA.
-We found that 61.2% of stems exhibited infestation in at least 5 nodes"}
-pf_nodes_plot
-```
-
-### Figure 6 {#figure6}
-```{r figure6_brome_nodes, fig.align='left', fig.cap="Figure 6: Internodes infested by *B. cephi* within collected *B. inermis* stems in Moccasin and Big Sandy, MT, USA."}
-brome.nodes
-```
-
-### Figure 7 {#figure7}
-```{r figure7_field_cut_plot, fig.align='left', fig.cap="Figure 7: Stem cutting by *C.
+```{r moc_seasonal_pot_show, fig.cap="Figure 5: Average precipitation in Big Sandy, MT. Average daily precipitation"}
+moc_seasonal_plot
+```
+
+### Figure 8 {#figure8}
+```{r figure7_field_cut_plot, fig.cap="Figure 8: Stem cutting by *C.
 cinctus*.
@@
 ```
-
-### Figure 8 {#figure8}
-```{r figure8, fig.align='left', fig.cap="Ratios of parasitized to cut stems in *B. inermis* and adjacent cultivated wheat. Dotted line represents where a 1:1 cut:parasitized ratio lies."}
+### Figure 9 {#figure9}
+```{r figure6_brome_nodes, fig.cap="Figure 9: Percentage of infested stems with internodes infested by *B. cephi* within collected *B. inermis* stems in Moccasin and Big Sandy, MT, USA."}
+brome.nodes
+```
+
+### Figure 10 {#figure10}
+```{r figure10, fig.cap="Ratios of parasitized:cut stems in *B. inermis* and adjacent cultivated wheat. Dotted line represents where a 1:1 cut:parasitized ratio lies."}
 ratio_plot
@@
 
+### Figure 9 {#figure9}
+```{r figure9, fig.align = 'left', fig.cap="Comparison of parasitoid presence in *B. inermis* between July and September collections."}
+brome_par_plot
+
+```
+
+![](Images/brome_images.jpg)
 ## Citations

### 2024-06-13 qmd/projects/msu/wss/weatherdata.qmd
update

@@
 ```
+
+```{r}
+library(readxl)
+lochy <-
+read_excel("qmd/projects/msu/wss/weather/lochy_weather_scrape.xlsx")
+
+line_table <-
+lochy %>%
+  dplyr::select(date, prcp) %>%
+  mutate(month = month(date),
+         day = day(date),
+         year = as.factor(year(date))) %>%
+  mutate(new_date = make_date(year = 0,
+                              month = month,
+                              day = day)) %>%
+   mutate(category = case_when(new_date < "0000-07-15" ~ "pre",
+                              new_date > "0000-09-15" ~ "post",
+                              new_date >= "0000-07-15" & new_date <= "0000-09-15" ~ "mid")) %>%
+  mutate(model_category = if_else(category == "mid", 
+                                  "mid", "no"))
+
+line_table_average <-
+  line_table %>%
+  group_by(month, day) %>%
+  reframe(category, new_date,
+          prcp = mean(prcp))
+
+line_table %>%
+  lm(prcp ~ model_category + year, data = .) %>%
+  summary()
+
+ggplot() +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "pre"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "post"),
+            color = "black",
+            linewidth = 0.25) +
+  geom_line(aes(x = new_date, y = prcp),
+            data = line_table_average %>%
+              filter(category == "mid"),
+            color = "black") +
+  geom_rect(aes(xmin = as.Date("0000-07-15", format = "%Y-%m-%d"),
+                xmax = as.Date("0000-09-15", format = "%Y-%m-%d"),
+                ymin = -0.01,
+                ymax = 0.5),
+            color = "Black",
+            fill = "Blue",
+            size = 0.3,
+            alpha = 0.05,
+            linetype = "solid") +
+  labs(x = NULL,
+       y = "Average Daily Precipitation (in)",
+       color = NULL) +
+  theme_bw()
+
+  # geom_vline(xintercept = as.Date("0000-07-15", 
+  #                                 format = "%Y-%m-%d"),
+  #            linetype = "dashed",
+  #            alpha = 0.5) +
+  # geom_vline(xintercept = as.Date("0000-09-15", 
+  #                                 format = "%Y-%m-%d"),
+  #            linetype = "dashed",
+  #            alpha = 0.5) +
+
+bar_table <-
+lochy %>%
+  dplyr::select(date, prcp) %>%
+  mutate(month = month(date),
+         day = day(date),
+         year = as.factor(year(date))) %>%
+  mutate(new_date = make_date(year = 0,
+                              month = month,
+                              day = day)) %>%
+  group_by(month) %>%
+  mutate(prcp = mean(prcp, na.rm = TRUE))
+```
